### 2009-11-13 src/Docs/Intro.tex
Reorganised Epitome - added introduction (ish) and moved perversity to an appendix

\section{In the beginning}

We really should right something informative here.

### 2009-11-18 src/Docs/Intro.tex
Intro: add a Recommended Reading section

@@
 
-We really should right something informative here.
+We really should write something informative here.
+
+\subsection{Recommended readings}
+
+The type-checker is a bidirectional
+one~\cite{turner:bidirectional_tc}. A remotely related type-checker
+has been described in the context of ETT~\cite{chapman:ett}.
+
+Normalization is achieved by evaluation~\cite{dybjer:nbe,
+  dybjer:dependent_types_work}. The implementation has been described
+in James Chapman's work~\cite{chapman:phd}. A graspable introduction
+to both normalization by evaluation and bidirectional type-checking
+\`a la Epigram can be found in Boutillier's
+report~\cite{boutillier:report}.
+
+The story for names has been told by McBride and
+McKinna~\cite{mcbride:free_variable}. \emph{Ite messa est}.
+
+For containers, there is a lot to say. So, I will not say anything for
+the moment.

### 2009-12-06 src/Docs/Intro.tex
Recommended readings for the Desc universe.

@@
 the moment.
+
+Concerning the |Desc| universe, Morris et al. \cite{morris:spf} wrote
+a clear article, covering that topic and much more. In particular,
+they show why and how |Box| and |mapBox| can automatically derived
+from |Desc|. Conor's work on Ornamemts \cite{mcbride:ornaments} builds
+up from a simplified |Desc| universe: some insights can be found there
+too. Finally, the authoritative source shall be cited: Dybjer et
+al. on induction-recursion \cite{dybjer:ir_axiom, dybjer:ir_algebra,
+  dybjer:iir}.

### 2010-01-14 src/Detritus/Language.tex
Docs: update Language.tex with latest changes

\begin{itemize}
\item \verb!*! for the universe of sets
\item \verb!#! for the universe of propositions
\item \verb|()| for the unit type
\item \verb!\ x -> t! for the $\lambda$-term $\lambda x.t$
\item \verb|f a| for $f$ applied to $a$ (no surprises here)
\item \verb!(x : S) -> T! for the $\Pi$-type $\Pi x:S. T$
\item \verb!S -> T! for the same $\Pi$-type if $T$ is independent of $x$
\item \verb!(x : S)(y : T) -> U! for the nested $\Pi$-type $\Pi x:S. \Pi y:T. U$
\item \verb!{| P}! for the type of proofs of $P$ (cf.\ set notation)
\item \verb!{x : S | P}! for the $\Sigma$-type $\Sigma x:S. P$
\item \verb!{x : S, y : T | P}! for the nested $\Sigma$-type $\Sigma x:S. \Sigma y:T. P$
\item \verb![a, b, c]! for the tuple $(a, (b, (c, ())))$ (think LISP lists)
\item \verb![a, c, c |]! for the tuple $(a, (b, c))$ (maybe?)
\item \verb!Enum! for the universe of enumerations
\item \verb!{}! for the empty enumerated type
\item \verb!{x, y, z}! for an enumerated type
\item \verb!0, 1, 2, ...! for \verb!zero!, \verb!suc zero!, \verb!suc (suc zero)!, et cetera
\item \verb!2 + tm! for \verb!suc (suc tm)!
\item \verb!@id! for a tag
\item \verb!F(..., ...)! for a fully applied operator
\end{itemize}

\question{Does this notation for $\Sigma$-types make sense, or should we go with the
original proposal of using an ampersand?}

\question{Perhaps we could use $[a, b, c, ]$ with a trailing comma to mean
$(a, (b, c))$? Using a vertical bar is problematic because it clashes with the
``such that'' interpretation.}

### 2010-01-16 src/Documentation/Introduction.tex
Hierarchy! No God! No Master! 8 Directories!
We have approx. 10.000 lines of code distributed in 38 files. It was time to organize that mess IMHO. 

The organization goes as follow:
 * ./Features: for features, hey. 
   Containing:
    - Features    - Desc    - Enum    - Equality    - FreeMonad    - IDesc    - INu
    - Labelled    - Nu      - Prop    - Quotient    - Sigma        - Uid
 * ./Evidences: the components of the "language of verifiable evidences"
   Containing:
    - Rules
    - Tm
    - Tactics
 * ./Cochon: the interface of the theorem prover
   Containing:
    - Cochon
    - DevLo

--- 2009-11-18 src/Docs/Intro.tex
@@
 
-\subsection{Recommended readings}
+\subsection{Recommended Reading}
 
@@
 the moment.
+
+Concerning the \verb|Desc| universe, Morris et al. \cite{morris:spf} wrote
+a clear article, covering that topic and much more. In particular,
+they show why and how \verb|Box| and \verb|mapBox| can automatically derived
+from \verb|Desc|. Conor's work on Ornamemts \cite{mcbride:ornaments} builds
+up from a simplified \verb|Desc| universe: some insights can be found there
+too. Finally, the authoritative source shall be cited: Dybjer et
+al. on induction-recursion \cite{dybjer:ir_axiom, dybjer:ir_algebra,
+  dybjer:iir}.

### 2010-03-07 papers/icfp-2010-desc/outline.tex
Desc@ICFP: add a map of the battlefield.

\documentclass{article}

\usepackage{verbatim}
\usepackage{fullpage}

%% Structure
\newenvironment{structure}{\footnotesize\verbatim}{\endverbatim}
%\newenvironment{structure}{\comment}{\endcomment}

\begin{document}

\section{The Type Theory}

\subsection{Base theory}

\begin{structure}
<- Presentation of the formalism
    <- Standard presentation
        -> No novelty here
    <- 3 judgments [equation]
        -> Context validity
        -> Typing judgements
        -> Equality judgements
    <- Invariants [equation]
        -> By induction on derivations
    <- Judgemental equality
        <- Presentation independant of particular implementation choice
        -> Model in Agda, intensional
        -> Used in Epigram, OTT
\end{structure}

\begin{structure}
<- Context validity [figure]
    <- Not much to be said
\end{structure}

\begin{structure}
<- Typing judgements [figure]
    <- Set in Set
        -> For simplicity of presentation
        -> Assume that a valid stratification can be inferred
            <- Implicit syntax, Pollack (?, or better reference)
        -> See later discussion
    <- Standard presentation of Pi and Sigma types
\end{structure}

\begin{structure}
<- Judgemental equality [figure]
    <- symmetry, reflexivity, and transitivity
    <- beta-rules for lambda and pair
    <- xi-rule for functions
    /> Assume transitivity, reflexity, and symmetry
    -> Agnostic in the notion of equality
        <- Doesn't rely on a ``propositional'' equality
\end{structure}

\begin{structure}
!!! Need Help !!!
<- Meta-theoretical properties
    <- Assuming a stratified discipline
        <- Harper-Pollack, Luo, Courant
    <> The point here is to reassert that dependent types are not evil, 
       there is no non-terminating type-checker, or such horrible lies <>
    -> Strongly normalizing
        -> Every program terminates
    -> Type-checking terminates
    ???
\end{structure}

\subsection{Finite sets}

\begin{structure}
<- Motivation
    <- Finite sets could be encoded with Unit and Bool
        /> Hinder the ability to name things
    -> For convenience
        <- Named elements
        <- Referring by name instead of code
        -> Types as coding presentation
            /> Also as coding representation!
\end{structure}

\begin{structure}
<- Implementation
    <- Tags
        -> Purely informational token
    <- EnumU
        -> Universe of finite sets
    <- EnumT e
        -> Elements of finite set e
\end{structure}

\begin{structure}
<- Equipment
    <- \spi operator
        <- Equivalent of Pi on finite sets
        <- First argument: (finite) domain
        <- Second argument: for each element of the domain, a co-domain
        -> Inhabitant of \spi: right-nested tuple of solutions
            <- Skip code for space reasons
    <- switch operator
        <- case analyses over x
        <- index into the \spi tuple to retrieve the corresponding result
\end{structure}

\begin{structure}
<- Equivalent to having a function space over finite sets
    /> Made non-obvious by low-level encodings
        <- General issue with codes
             -> Need to provide an attractive presentation to the user
    -> Types seem to obfuscate our reading
        <- Provide ``too much'' information
        /> False impression: information is actually waiting to be used more widely
        -> See next Section
\end{structure}

\subsection{Type propagation}

\begin{structure}
<- Bidirectional type-checking [ref. Turner,Pierce]
    -> Separating type-checking from type synthesis
    <- Type checking: push terms into types
        <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
    <- Type inference: pull types out of terms
        <- Example: |x : S l- x :<: S| gives us the type of x
\end{structure}

\begin{structure}
<- Formalization: type propagation
    <- Motivation
        -> High-level syntax
            -> exprIn: types are pushed in
                <- Subject to type *checking*
            -> exprEx: types are pulled from
                <- Subject to type *synthesis*
        -> Translated into our low-level type theory
        -> Presented as judgements
    -> Presentation mirors typing rule of [figure] 
        -> Ignore identitical judgements
\end{structure}

\begin{structure}
<- Type checking [figure]
    <- Push a type in an exprIn
    <- Result in a full term
    -> *Use* the type to build the term!
        -> Domain and co-domain propagation for Pi and Sigma
        -> Translation of 'tags into EnumTs
        -> Translation of ['tags ...] into EnumUs
        -> Finite function space into switch
\end{structure}

\begin{structure}
<- Type synthesis [figure]
    <- Pull a type out of an exprEx
    <- Result in a full term, together with its type
    -> Do *not* need to specify types
        -> Extracting a term from the context
        -> Function application
        -> Projections
\end{structure}

\begin{structure}
<- Summary
    -> Not a novel technique [refs?]
        /> Used as a boilerplate scrapper
    -> Make dealing with codes *practical*
        <- Example: Finite sets/finite function space
        -> We should not restrain ourself in using codes
            <- We know how to present them to the user
-> Will extend this machinery in further sections
\end{structure}

\section{A universe of simple data-types}

\begin{structure}
<- Why starting with simple data-types
    <- For pedagogical purposes
        <- Data-types as we know them every day
        /> Target dependent types
    -> Cut down version of Induction Recursion
        -> Presentation evolves independently as we go to dependent types
\end{structure}

\subsection{The power of $\Sigma$}

\begin{structure}
<- The duality of Sigma
    <- Sigma generalizes sum over arbitrary arities
        -> \Sigma A B == \Sigma_{x : A} B x
    <- Sigma generalizes product to have a dependant second component
        -> \Sigma A B == (x : A) \times (B x)
\end{structure}

\begin{structure}
<- Data-types in the simply-typed world
    -> "sums-of-product"
        <- Sum of constructors
        <- Product of arguments
<- Data-types in the dependently-typed world
    -> "sigmas-of-sigmas"
    /> Need ability to manipulate these sigmas
        -> Define a Code for data-types
        -> Together with a sigma-based Interpretation
\end{structure}

\subsection{A universe of descriptions}

\begin{structure}
<- Introduction to Universe construction
    <- Define a Code
        -> Name objects
    <- Define an Interpretation of codes into the type theory
        -> Give a semantics to objects
    -> Ability to manipulate code
    -> Ability to compute with these objects
\end{structure}

\begin{structure}
<- Justification of the code 
    <- [both figures]: cannot be read separately
    <- Mimic the standard grammar of data-types description
        -> Just as we already know it
        <- '\Sigma for making sigmas-of-sigmas
        <- 'indx for exhibiting the functoriality
            -> For recursive arguments
        <- '1 for end of description
\end{structure}

\subsection{Examples}

\begin{structure}
<- Nat
    <- Sum of zero, suc
    <- zero case: done
    <- suc case: leave open and done
    -> NatD Z = 1 + Z
\end{structure}

\begin{structure}
<- List
    <- Sum of nil, cons
    <- nil case: done
    <- cons case: product of X with leave open and done
    -> ListD X Z = 1 + X * Z
\end{structure}

\begin{structure}
<- Tree
    <- sum of leaf, node
    <- leaf case: done
    <- node case: product of X with two leave open and done
    -> TreeD X Z = 1 + X * Z * Z
\end{structure}

\begin{structure}
<- Fictive object [figure 'data Desc']
    -> Must be read as a type signature
    -> See further for its actual implementation
        <- Subject to our levitation exercise
\end{structure}

\subsection{Fix-point}

\begin{structure}
<- Build the fix-point of functors
    <- See examples: need to build their initial algebra
    -> Extend the type theory with Mu/Con [figure]
        <- Straightforward definition of a fixpoint
            <- Interpret D with (Mu D) as subobjects
\end{structure}

\begin{structure}
<- Extending type propagation
    <- Data-type declaration turns into definitions
        -> Straightforward translation to Desc
        -> Creation of a variable referring to the structure
    <- Push Mu to an applied name [figure]
        -> Direct integration into the type propagation machinery
    <- Labelled Mu
        /> Just mention the possibility of labelling, no details required
        -> For the user, objects have names rather than Mu of codes
    -> Coded presentation is practical
        <- The user never see a code
\end{structure}

\begin{structure}
<- Elimination on Mu
    <- We are used to foldD : \forall X. (desc D X -> X) -> mu D -> X
        /> Not dependent
        -> Cannot express some (which one again?) properties
    -> Develop a dependent induction
        <- Everywhere/All
        <- Induction
    -> *Generic*
\end{structure}

\section{Levitating the universe of descriptions}

\subsection{Implementing finite sets}

\begin{structure}
<- Recall typing rules of 1st section
    -> Make clear they were just promises
    -> Can be implemented now
        <- Simply List UId
\end{structure}

\begin{structure}
<- Consequences
    -> Type theory doesn't need to be extended with EnumU, NilE, and ConsE
        <- EnumU == Mu EnumUD
        <- NilE, ConsE are just tags
    -> Do not need a specific \spi eliminator
        <- \spi is an instance of the generic eliminator
            <- Code?
    -> Anything else remains the same (switch, EnumT, 0, 1+)
\end{structure}

\begin{structure}
<- Summary of the operation
    <- The content of the type theory is exactly the same
        <- before == after
    /> type naming scheme condenses
        <- Replace named constructors by codes in the universe of data-types
    -> Our next step is a similar move (in essence)
        /> Condenses the entire naming scheme of data-types
\end{structure}

\subsection{Implementing descriptions}

\begin{structure}
<- Realizing our promises
    <- We are going to implement Desc
    /> Desc is itself a data-type
        <- Same situation as EnumU
            <- We want to benefit from generic operations
        -> It ought to be encoded in itself
\end{structure}

\subsubsection{First try}

\begin{structure}
<- A partial implementation
    <- '1 and 'indx are easy
    <- 'sigma is partially doable
        /> lack the ability to do an higher-order inductive call
    -> Show partial code [figure]
\end{structure}

\subsubsection{Second try}

\begin{structure}
<- Extending the universe of description
    -> With higher-order induction
    <- Intuition: index elements in X by H, and go on reading
        -> indx is isomorph to hindx for H = 1
    /> Keep indx
        <- First order!
        -> Extensionally equal to hindx 1
        /> Practically, definitional equality on Sigma/Pi cannot cope with it
    -> Show DescD code
\end{structure}

\subsubsection{Final move}

\begin{structure}
<- Subtlety: translation of [ ... ]
    -> Let us do it manually
        -> Code with problem for the motive of switch
\end{structure}

\begin{structure}
<- The magician trick
    <- Our problem is to give a motive for switch
        /> We perfectly know what it ought to be: \_ -> DescD
    -> Solution: extend the type theory with a special purpose switchD
        -> Only extension required to the type theory!
        -> Hidden away to the user by the syntactic sugar
            -> Sufficient to ensure unavailability as a raw operator
            <- Another instance of type propagation
\end{structure}

\begin{structure}
<- Generic programming now!
    <- Desc is just data
        -> Can be manipulated
    <- Free induction scheme on Desc
        -> Ability to inspect data-types
        -> Ability to program on data-types
\end{structure}

\subsubsection{Desc, atomically}

\begin{structure}
<- Adding hindx have introduced some duplication
    <- indx == hindx 1
    -> We can factor out commonalities 
        /> Obtain an equivalent presentation
        /> Still embeddable (refer to the Agda model)
\end{structure}

\begin{structure}
<- Give new presentation [figure]
    <- hindx have introduced the notion of function space: 'Pi
    <- hindx and indx are both composed by a binary product and a left open term: 'x and 'id 
    -> Straightforward translation to the new system [equation]
\end{structure}

\subsection{The generic catamorphism}

\begin{structure}
<- Making cata
    ???
\end{structure}

\subsection{The generic Free Monad}

\begin{structure}
!!! EARLIER !!!
<- Tagged description
    <- Form TDesc = List (UId x Desc) [equation]
    <- Follow usual sums-of-product presentation of data-type
        <- Finite set of constructors
        <- Then whatever you want
    -> Any Desc data-type can be turned into this form
        -> No loss of expressive power
        /> Garantee a ``constructor form''
\end{structure}

\begin{structure}
<- A generic program: the free monad construction
    <- Recall free monad construction in Haskell
        -> Based on a functor F
    <- Note that the free monad construction is itself defined by a functor
        -> Extract it
    <- Encode it in the Desc world [equation]
        <- F is the Desc we start with
        <- The free monad functor is what we have just defined
        <- [\_]* : Desc -> Set -> Desc
           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
        -> Mu does the fix-point
\end{structure}

\begin{structure}
<- A generic program: monadic substitution [equation]
    <- subst : \forall T X Y. mu ([T]* X) -> (X -> mu ([T]* Y)) -> mu ([T]* Y)
    -> Using Fold

\end{structure}

\section{Indexing descriptions}

\begin{structure}
!!! Need Help !!!
<- Motivation
    <- Desc: expressivity of simply-typed data-types: inductive types
        <- Values do not influence types
    /> Example: Vectors
        <- Cannot be defined by just induction
            <- Vectors of all size need to be defined at the *same* time
            -> Defined as a *family* of types
                -> Index
        -> I -> IDesc I: Inductive family
    ???
\end{structure}

\subsection{From Desc to IDesc}

\begin{structure}
<- Labelling Id
    <- We had: data Desc : Set -> Set
    -> We want: data IDesc : (I -> Set) -> Set
        <- Indexed functor (?)
        -> It is sufficient to label Id
            <- Where the functor is built
\end{structure}

\begin{structure}
<- Also replacing '1 by 'const  [figure]
    <- For convenience
        <- 'const X equivalent to 'sigma X (\_ -> '1)
        /> Easier to abstract
            <- Extensionally same
            /> 'const more useful in practice
\end{structure}

\subsection{Examples}

\begin{structure}
<- Nat
    -> [equation]
    <- Non-indexed types lives in IDesc 1
        -> This applies to all previous examples
\end{structure}

\begin{structure}
<- Levitation [figure]
    <- Following Desc encoding
        /> Note: simple data-type
            -> Live in IDesc 1
    -> Behind the scene, relies on the special purpose switchD
\end{structure}

\begin{structure}
<- Untyped lambda terms [figure]
    ???
\end{structure}

\subsection{Constrained constructors}

\begin{structure}
<- Assuming a suitable notion of definitional equality for I
    ???
\end{structure}

\begin{structure}
<- Fin [figure]
    <- Presentation with equality constraints
    ???
\end{structure}

\begin{structure}
<- Vectors
    ???
\end{structure}

\begin{structure}
<- GADTS
    ???
\end{structure}

\begin{structure}
<- Brady optimization
    <- Source to source translation
    <- Able to remove some constraints
    -> Example: Fin [figure]
\end{structure}

\subsection{Free IMonad}

\begin{structure}
<- Untyped lambda terms
    ???
\end{structure}

\begin{structure}
<- IDesc
    ???
\end{structure}

\section{Discussion}

\begin{structure}
!!! Need Help !!!
<- Universe stratification
    <- Stratified agda model
        <- Fully stratified
        <- Proof of iso between host and embedding
    ???
\end{structure}

\end{document}

### 2010-03-07 papers/icfp-2010-desc/outline.tex
Desc@ICFP: explain the map

@@
 \begin{document}
+
+\begin{structure}
+How to read me:
+    '<-'  means "because"
+    '->'  means "therefore"
+    '/>'  means "but"
+
+The spatial tabbing encodes the scope. Things on the same column are
+linked by an implicit "and". Things under an argument justify it.
+
+I know, that's fucked up and you hate it. But that's much better to
+read that crap than reading my long frenglish sentences a la James
+Proust.
+\end{structure}
 

### 2010-03-08 papers/icfp-2010-desc/outline.tex
Desc@ICFP: precise the battlefield

@@
         -> Assume that a valid stratification can be inferred
-            <- Implicit syntax, Pollack (?, or better reference)
+            <- Harper-Pollack, Luo, Courant
         -> See later discussion
@@
     <- xi-rule for functions
-    /> Assume transitivity, reflexity, and symmetry
     -> Agnostic in the notion of equality
@@
     <- Assuming a stratified discipline
-        <- Harper-Pollack, Luo, Courant
     <> The point here is to reassert that dependent types are not evil, 
@@
 \begin{structure}
-<- Implementation
+<- Implementation [figure]      
     <- Tags
@@
     -> *Generic*
+    ???
 \end{structure}
@@
 <- Making cata
-    ???
+    <- Present the type signature
+    <- Starts with a call to generic induction
+        <- induction on Desc!
+        /> Show types at hand
+        -> Explain how to use inductive hypothesis
+    <- Implement the 'replace' function
+    -> Dependent-typeless catamorphism 
+\end{structure}
+
+\begin{structure}
+<- Deriving generic functions
+    <- Taking a Desc and computing a function
+        <- Desc comes equipped with an induction principle
+        -> Ability to compute more functions from it
+            -> More generic functions
+    <- Inspecting data-types
+        <- All described by a Desc code
+        -> Ability to explore the code
+            <- Desc equipped with an induction principle
+            -> Build new objects based on that structure
 \end{structure}
@@
         -> Using Fold
+    -> Consequences
+        <- We have free monad data-type
+            <- Term + variables
+        <- We have monad operations
+            <- Return / var
+            <- Substitution / bind
+\end{structure}
 
+\begin{structure}
+<- Deriving new data-structure and functions on them
+    <- Computing the Free Monad of a data-type
+        <- Derive new data-structure from previous one
+            <- It is just code
+        /> New data-structure comes with some equipment
+    <- Computing new functions on computed data-types
+        <- If data comes with structure, we ought to be able to capture it
+            <- Induction on Desc
+            -> Ability to compute over data
 \end{structure}

### 2010-03-08 papers/icfp-2010-desc/outline.tex
Desc@ICFP: more meat

@@
 <- Untyped lambda terms [figure]
-    ???
+    <- Indexed by Nat
+        <- De Bruijn "encoding"
+    <- Variable in Fin n
+        -> Correct by construction lambda term
+    /> Lack the monadic structure [Altenkirch & Reus ???]
+        -> Miss substitution operator
+        /> Will be fixed later
 \end{structure}
@@
 \begin{structure}
+!!! Need Help !!!
 <- Assuming a suitable notion of definitional equality for I
@@
     <- Presentation with equality constraints
+    <- Example of constrained constructor
+        -> Constraints are translated into equalities
     ???
@@
 <- Vectors
-    ???
+    Do we treat them in the end? 
+    What can we say here we haven't with Fin?
 \end{structure}
@@
 \begin{structure}
-<- GADTS
-    ???
+<- GADT-style data-type definition
+    <- We directly support it
+        <- Constraints turned into equalities
+            <- Fin example
+    <- Again, smooth transition from "today" data-types to advanced ones
+        <- Same grammar
+        /> More power
+    /> Equations in data-types are not a blessing
+        <- Complexify induction principle
+        <- Lose match between data-type definition and the elimination form
 \end{structure}
@@
 \begin{structure}
-<- Brady optimization
+!!! Need Help !!!
+<- Brady optimization: forcing
     <- Source to source translation
@@
     -> Example: Fin [figure]
+    ??? More technical detail needed
 \end{structure}
@@
 \begin{structure}
+<- Variation on a theme: free imonad construction
+    <- Recall existence of generic free monad construction
+    -> Present its generalization to IDesc [equation]
+        <- \I -> IDesc I as describing an indexed endofunctor
+        <- Free indexed monad construction
+    -> Still a suitable, generic notion of substitution
+        <- show type signature
+        <- show implementation?? (space! space!)
+\end{structure}
+
+\begin{structure}
 <- Untyped lambda terms
-    ???
+    <- Obvious instance of free monad
+        <- Variable: Return
+        <- App and Abst: indexed functor
+        <- Bind: substitution
+    -> Get a correct-by-construction lambda term
+        /> With substitution for free!
+        <- We should not pay for the structure which *is* here
 \end{structure}
@@
 <- IDesc
+    <- Another instance of free monad
+        <- Var: Return
+        <- Remaining: (not even indexed) functor
+    -> Madness just starts
+        <- map operation
+        <- [ subst sigma D ] X = [D] (\x -> [subst sigma x] X)
         ???

### 2010-03-09 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: merge outline into paper. Chunk paper in smaller pieces. (Base theory section)

--- 2010-03-08 papers/icfp-2010-desc/outline.tex
@@
-\documentclass{article}
-
-\usepackage{verbatim}
-\usepackage{fullpage}
-
-%% Structure
-\newenvironment{structure}{\footnotesize\verbatim}{\endverbatim}
-%\newenvironment{structure}{\comment}{\endcomment}
-
-\begin{document}
-
-\begin{structure}
-How to read me:
-    '<-'  means "because"
-    '->'  means "therefore"
-    '/>'  means "but"
-
-The spatial tabbing encodes the scope. Things on the same column are
-linked by an implicit "and". Things under an argument justify it.
-
-I know, that's fucked up and you hate it. But that's much better to
-read that crap than reading my long frenglish sentences a la James
-Proust.
-\end{structure}
-
 \section{The Type Theory}
@@
         -> Equality judgements
+\end{structure}
+
+\[
+\begin{array}{ll}
+\Gamma\vdash\Valid & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
+\Gamma\vdash \Bhab{t}{T} & \mbox{term \(t\) has type \(T\) in context \(\Gamma\)} \\
+\Gamma\vdash \Bhab{s \equiv t}{T} & \mbox{\(s\) and \(t\) are equal at type \(T\) in context \(\Gamma\)} \\
+\end{array}
+\]
+
+\begin{structure}
     <- Invariants [equation]
         -> By induction on derivations
+\end{structure}
+
+The systems of inference rules will be formulated to ensure that that the
+following implications always hold by induction on derivations.
+\[
+\begin{array}{l@{\;\;\Rightarrow\;\;}l}
+\Gamma\vdash \Bhab{t}{T}            & \Gamma\vdash\Valid \;\wedge\; \Gamma\vdash\Type{T} \\
+\Gamma\vdash s \equiv \Bhab{t}{T}   & \Gamma\vdash \Bhab{s}{T} \;\wedge\; \Gamma\vdash \Bhab{t}{T} \\
+\Gamma;\xS;\Delta \vdash J          & \Gamma\vdash \Bhab{s}{S} \;\Rightarrow\; 
+                                           \Gamma;\Delta[s/x] \vdash J[s/x] \\
+\end{array}
+\]
+
+\begin{structure}
     <- Judgemental equality
@@
 
+\begin{figure}
+
+\[
+%% Empty context validity
+\Axiom{\vdash \Valid}
+\qquad
+%% Extend context
+\Rule{\Gamma       \vdash \Type{S}}
+     {\Gamma ; \xS \vdash \Valid}\;x\not\in\Gamma
+\]
+
+\caption{Context validity}
+\label{fig:context-validity}
+\end{figure}
+
 \begin{structure}
@@
 
+\begin{figure}
+
+\input{figure_typing_judgements}
+
+\caption{Typing judgements}
+\label{fig:typing-judgements}
+
+\end{figure}
+
 \begin{structure}
@@
 \end{structure}
+
+\begin{figure}
+
+\input{figure_judgemental_equality}
+
+\caption{Judgemental equality}
+\label{fig:judgemental-equality}
+
+\end{figure}
 
@@
 
+\begin{figure}
+
+\input{figure_finite_sets}
+
+\caption{Typing rules for finite sets}
+\label{fig:typing-finite-set}
+
+\end{figure}
+
 \begin{structure}
@@
 \end{structure}
+
+\begin{figure}
+
+\[\stkl{
+%% spi
+\F{$\pi$} : \PITEL{\V{e}}{\EnumU}
+            \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set 
+\\
+%% switch
+\begin{array}{@{}ll}
+\F{switch} : & \PITEL{\V{e}}{\EnumU}
+               \PITEL{\V{P}}{\EnumT{e} \To \Set} \\
+             & \PITEL{\V{b}}{\spi{e}{P}}
+               \PITEL{\V{x}}{\EnumT{e}} \To P~x
+\end{array}
+}\]
+
+\caption{Operators on finite sets}
+\label{fig:operator-finite-set}
+
+\end{figure}
 
@@
 
+\begin{figure}
+
+\input{figure_type_checking}
+
+\caption{Type checking}
+\label{fig:type-checking}
+
+\end{figure}
+
 \begin{structure}
@@
 \end{structure}
+
+\begin{figure}
+
+\input{figure_type_synthesis}
+
+\caption{Type synthesis}
+\label{fig:type-synthesis}
+
+\end{figure}
 
@@
 \end{structure}
-
-\section{A universe of simple data-types}
-
-\begin{structure}
-<- Why starting with simple data-types
-    <- For pedagogical purposes
-        <- Data-types as we know them every day
-        /> Target dependent types
-    -> Cut down version of Induction Recursion
-        -> Presentation evolves independently as we go to dependent types
-\end{structure}
-
-\subsection{The power of $\Sigma$}
-
-\begin{structure}
-<- The duality of Sigma
-    <- Sigma generalizes sum over arbitrary arities
-        -> \Sigma A B == \Sigma_{x : A} B x
-    <- Sigma generalizes product to have a dependant second component
-        -> \Sigma A B == (x : A) \times (B x)
-\end{structure}
-
-\begin{structure}
-<- Data-types in the simply-typed world
-    -> "sums-of-product"
-        <- Sum of constructors
-        <- Product of arguments
-<- Data-types in the dependently-typed world
-    -> "sigmas-of-sigmas"
-    /> Need ability to manipulate these sigmas
-        -> Define a Code for data-types
-        -> Together with a sigma-based Interpretation
-\end{structure}
-
-\subsection{A universe of descriptions}
-
-\begin{structure}
-<- Introduction to Universe construction
-    <- Define a Code
-        -> Name objects
-    <- Define an Interpretation of codes into the type theory
-        -> Give a semantics to objects
-    -> Ability to manipulate code
-    -> Ability to compute with these objects
-\end{structure}
-
-\begin{structure}
-<- Justification of the code 
-    <- [both figures]: cannot be read separately
-    <- Mimic the standard grammar of data-types description
-        -> Just as we already know it
-        <- '\Sigma for making sigmas-of-sigmas
-        <- 'indx for exhibiting the functoriality
-            -> For recursive arguments
-        <- '1 for end of description
-\end{structure}
-
-\subsection{Examples}
-
-\begin{structure}
-<- Nat
-    <- Sum of zero, suc
-    <- zero case: done
-    <- suc case: leave open and done
-    -> NatD Z = 1 + Z
-\end{structure}
-
-\begin{structure}
-<- List
-    <- Sum of nil, cons
-    <- nil case: done
-    <- cons case: product of X with leave open and done
-    -> ListD X Z = 1 + X * Z
-\end{structure}
-
-\begin{structure}
-<- Tree
-    <- sum of leaf, node
-    <- leaf case: done
-    <- node case: product of X with two leave open and done
-    -> TreeD X Z = 1 + X * Z * Z
-\end{structure}
-
-\begin{structure}
-<- Fictive object [figure 'data Desc']
-    -> Must be read as a type signature
-    -> See further for its actual implementation
-        <- Subject to our levitation exercise
-\end{structure}
-
-\subsection{Fix-point}
-
-\begin{structure}
-<- Build the fix-point of functors
-    <- See examples: need to build their initial algebra
-    -> Extend the type theory with Mu/Con [figure]
-        <- Straightforward definition of a fixpoint
-            <- Interpret D with (Mu D) as subobjects
-\end{structure}
-
-\begin{structure}
-<- Extending type propagation
-    <- Data-type declaration turns into definitions
-        -> Straightforward translation to Desc
-        -> Creation of a variable referring to the structure
-    <- Push Mu to an applied name [figure]
-        -> Direct integration into the type propagation machinery
-    <- Labelled Mu
-        /> Just mention the possibility of labelling, no details required
-        -> For the user, objects have names rather than Mu of codes
-    -> Coded presentation is practical
-        <- The user never see a code
-\end{structure}
-
-\begin{structure}
-<- Elimination on Mu
-    <- We are used to foldD : \forall X. (desc D X -> X) -> mu D -> X
-        /> Not dependent
-        -> Cannot express some (which one again?) properties
-    -> Develop a dependent induction
-        <- Everywhere/All
-        <- Induction
-    -> *Generic*
-    ???
-\end{structure}
-
-\section{Levitating the universe of descriptions}
-
-\subsection{Implementing finite sets}
-
-\begin{structure}
-<- Recall typing rules of 1st section
-    -> Make clear they were just promises
-    -> Can be implemented now
-        <- Simply List UId
-\end{structure}
-
-\begin{structure}
-<- Consequences
-    -> Type theory doesn't need to be extended with EnumU, NilE, and ConsE
-        <- EnumU == Mu EnumUD
-        <- NilE, ConsE are just tags
-    -> Do not need a specific \spi eliminator
-        <- \spi is an instance of the generic eliminator
-            <- Code?
-    -> Anything else remains the same (switch, EnumT, 0, 1+)
-\end{structure}
-
-\begin{structure}
-<- Summary of the operation
-    <- The content of the type theory is exactly the same
-        <- before == after
-    /> type naming scheme condenses
-        <- Replace named constructors by codes in the universe of data-types
-    -> Our next step is a similar move (in essence)
-        /> Condenses the entire naming scheme of data-types
-\end{structure}
-
-\subsection{Implementing descriptions}
-
-\begin{structure}
-<- Realizing our promises
-    <- We are going to implement Desc
-    /> Desc is itself a data-type
-        <- Same situation as EnumU
-            <- We want to benefit from generic operations
-        -> It ought to be encoded in itself
-\end{structure}
-
-\subsubsection{First try}
-
-\begin{structure}
-<- A partial implementation
-    <- '1 and 'indx are easy
-    <- 'sigma is partially doable
-        /> lack the ability to do an higher-order inductive call
-    -> Show partial code [figure]
-\end{structure}
-
-\subsubsection{Second try}
-
-\begin{structure}
-<- Extending the universe of description
-    -> With higher-order induction
-    <- Intuition: index elements in X by H, and go on reading
-        -> indx is isomorph to hindx for H = 1
-    /> Keep indx
-        <- First order!
-        -> Extensionally equal to hindx 1
-        /> Practically, definitional equality on Sigma/Pi cannot cope with it
-    -> Show DescD code
-\end{structure}
-
-\subsubsection{Final move}
-
-\begin{structure}
-<- Subtlety: translation of [ ... ]
-    -> Let us do it manually
-        -> Code with problem for the motive of switch
-\end{structure}
-
-\begin{structure}
-<- The magician trick
-    <- Our problem is to give a motive for switch
-        /> We perfectly know what it ought to be: \_ -> DescD
-    -> Solution: extend the type theory with a special purpose switchD
-        -> Only extension required to the type theory!
-        -> Hidden away to the user by the syntactic sugar
-            -> Sufficient to ensure unavailability as a raw operator
-            <- Another instance of type propagation
-\end{structure}
-
-\begin{structure}
-<- Generic programming now!
-    <- Desc is just data
-        -> Can be manipulated
-    <- Free induction scheme on Desc
-        -> Ability to inspect data-types
-        -> Ability to program on data-types
-\end{structure}
-
-\subsubsection{Desc, atomically}
-
-\begin{structure}
-<- Adding hindx have introduced some duplication
-    <- indx == hindx 1
-    -> We can factor out commonalities 
-        /> Obtain an equivalent presentation
-        /> Still embeddable (refer to the Agda model)
-\end{structure}
-
-\begin{structure}
-<- Give new presentation [figure]
-    <- hindx have introduced the notion of function space: 'Pi
-    <- hindx and indx are both composed by a binary product and a left open term: 'x and 'id 
-    -> Straightforward translation to the new system [equation]
-\end{structure}
-
-\subsection{The generic catamorphism}
-
-\begin{structure}
-<- Making cata
-    <- Present the type signature
-    <- Starts with a call to generic induction
-        <- induction on Desc!
-        /> Show types at hand
-        -> Explain how to use inductive hypothesis
-    <- Implement the 'replace' function
-    -> Dependent-typeless catamorphism 
-\end{structure}
-
-\begin{structure}
-<- Deriving generic functions
-    <- Taking a Desc and computing a function
-        <- Desc comes equipped with an induction principle
-        -> Ability to compute more functions from it
-            -> More generic functions
-    <- Inspecting data-types
-        <- All described by a Desc code
-        -> Ability to explore the code
-            <- Desc equipped with an induction principle
-            -> Build new objects based on that structure
-\end{structure}
-
-\subsection{The generic Free Monad}
-
-\begin{structure}
-!!! EARLIER !!!
-<- Tagged description
-    <- Form TDesc = List (UId x Desc) [equation]
-    <- Follow usual sums-of-product presentation of data-type
-        <- Finite set of constructors
-        <- Then whatever you want
-    -> Any Desc data-type can be turned into this form
-        -> No loss of expressive power
-        /> Garantee a ``constructor form''
-\end{structure}
-
-\begin{structure}
-<- A generic program: the free monad construction
-    <- Recall free monad construction in Haskell
-        -> Based on a functor F
-    <- Note that the free monad construction is itself defined by a functor
-        -> Extract it
-    <- Encode it in the Desc world [equation]
-        <- F is the Desc we start with
-        <- The free monad functor is what we have just defined
-        <- [\_]* : Desc -> Set -> Desc
-           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
-        -> Mu does the fix-point
-\end{structure}
-
-\begin{structure}
-<- A generic program: monadic substitution [equation]
-    <- subst : \forall T X Y. mu ([T]* X) -> (X -> mu ([T]* Y)) -> mu ([T]* Y)
-        -> Using Fold
-    -> Consequences
-        <- We have free monad data-type
-            <- Term + variables
-        <- We have monad operations
-            <- Return / var
-            <- Substitution / bind
-\end{structure}
-
-\begin{structure}
-<- Deriving new data-structure and functions on them
-    <- Computing the Free Monad of a data-type
-        <- Derive new data-structure from previous one
-            <- It is just code
-        /> New data-structure comes with some equipment
-    <- Computing new functions on computed data-types
-        <- If data comes with structure, we ought to be able to capture it
-            <- Induction on Desc
-            -> Ability to compute over data
-\end{structure}
-
-\section{Indexing descriptions}
-
-\begin{structure}
-!!! Need Help !!!
-<- Motivation
-    <- Desc: expressivity of simply-typed data-types: inductive types
-        <- Values do not influence types
-    /> Example: Vectors
-        <- Cannot be defined by just induction
-            <- Vectors of all size need to be defined at the *same* time
-            -> Defined as a *family* of types
-                -> Index
-        -> I -> IDesc I: Inductive family
-    ???
-\end{structure}
-
-\subsection{From Desc to IDesc}
-
-\begin{structure}
-<- Labelling Id
-    <- We had: data Desc : Set -> Set
-    -> We want: data IDesc : (I -> Set) -> Set
-        <- Indexed functor (?)
-        -> It is sufficient to label Id
-            <- Where the functor is built
-\end{structure}
-
-\begin{structure}
-<- Also replacing '1 by 'const  [figure]
-    <- For convenience
-        <- 'const X equivalent to 'sigma X (\_ -> '1)
-        /> Easier to abstract
-            <- Extensionally same
-            /> 'const more useful in practice
-\end{structure}
-
-\subsection{Examples}
-
-\begin{structure}
-<- Nat
-    -> [equation]
-    <- Non-indexed types lives in IDesc 1
-        -> This applies to all previous examples
-\end{structure}
-
-\begin{structure}
-<- Levitation [figure]
-    <- Following Desc encoding
-        /> Note: simple data-type
-            -> Live in IDesc 1
-    -> Behind the scene, relies on the special purpose switchD
-\end{structure}
-
-\begin{structure}
-<- Untyped lambda terms [figure]
-    ???
-\end{structure}
-
-\subsection{Constrained constructors}
-
-\begin{structure}
-<- Assuming a suitable notion of definitional equality for I
-    ???
-\end{structure}
-
-\begin{structure}
-<- Fin [figure]
-    <- Presentation with equality constraints
-    ???
-\end{structure}
-
-\begin{structure}
-<- Vectors
-    ???
-\end{structure}
-
-\begin{structure}
-<- GADTS
-    ???
-\end{structure}
-
-\begin{structure}
-<- Brady optimization
-    <- Source to source translation
-    <- Able to remove some constraints
-    -> Example: Fin [figure]
-\end{structure}
-
-\subsection{Free IMonad}
-
-\begin{structure}
-<- Untyped lambda terms
-    ???
-\end{structure}
-
-\begin{structure}
-<- IDesc
-    ???
-\end{structure}
-
-\section{Discussion}
-
-\begin{structure}
-!!! Need Help !!!
-<- Universe stratification
-    <- Stratified agda model
-        <- Fully stratified
-        <- Proof of iso between host and embedding
-    ???
-\end{structure}
-
-\end{document}

### 2010-03-09 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: merge outline into paper. (Universe of simple data-types)

--- 2010-03-07 papers/icfp-2010-desc/outline.tex
@@
-\documentclass{article}
-
-\usepackage{verbatim}
-\usepackage{fullpage}
-
-%% Structure
-\newenvironment{structure}{\footnotesize\verbatim}{\endverbatim}
-%\newenvironment{structure}{\comment}{\endcomment}
-
-\begin{document}
-
-\begin{structure}
-How to read me:
-    '<-'  means "because"
-    '->'  means "therefore"
-    '/>'  means "but"
-
-The spatial tabbing encodes the scope. Things on the same column are
-linked by an implicit "and". Things under an argument justify it.
-
-I know, that's fucked up and you hate it. But that's much better to
-read that crap than reading my long frenglish sentences a la James
-Proust.
-\end{structure}
-
-\section{The Type Theory}
-
-\subsection{Base theory}
-
-\begin{structure}
-<- Presentation of the formalism
-    <- Standard presentation
-        -> No novelty here
-    <- 3 judgments [equation]
-        -> Context validity
-        -> Typing judgements
-        -> Equality judgements
-    <- Invariants [equation]
-        -> By induction on derivations
-    <- Judgemental equality
-        <- Presentation independant of particular implementation choice
-        -> Model in Agda, intensional
-        -> Used in Epigram, OTT
-\end{structure}
-
-\begin{structure}
-<- Context validity [figure]
-    <- Not much to be said
-\end{structure}
-
-\begin{structure}
-<- Typing judgements [figure]
-    <- Set in Set
-        -> For simplicity of presentation
-        -> Assume that a valid stratification can be inferred
-            <- Implicit syntax, Pollack (?, or better reference)
-        -> See later discussion
-    <- Standard presentation of Pi and Sigma types
-\end{structure}
-
-\begin{structure}
-<- Judgemental equality [figure]
-    <- symmetry, reflexivity, and transitivity
-    <- beta-rules for lambda and pair
-    <- xi-rule for functions
-    /> Assume transitivity, reflexity, and symmetry
-    -> Agnostic in the notion of equality
-        <- Doesn't rely on a ``propositional'' equality
-\end{structure}
-
-\begin{structure}
-!!! Need Help !!!
-<- Meta-theoretical properties
-    <- Assuming a stratified discipline
-        <- Harper-Pollack, Luo, Courant
-    <> The point here is to reassert that dependent types are not evil, 
-       there is no non-terminating type-checker, or such horrible lies <>
-    -> Strongly normalizing
-        -> Every program terminates
-    -> Type-checking terminates
-    ???
-\end{structure}
-
-\subsection{Finite sets}
-
-\begin{structure}
-<- Motivation
-    <- Finite sets could be encoded with Unit and Bool
-        /> Hinder the ability to name things
-    -> For convenience
-        <- Named elements
-        <- Referring by name instead of code
-        -> Types as coding presentation
-            /> Also as coding representation!
-\end{structure}
-
-\begin{structure}
-<- Implementation
-    <- Tags
-        -> Purely informational token
-    <- EnumU
-        -> Universe of finite sets
-    <- EnumT e
-        -> Elements of finite set e
-\end{structure}
-
-\begin{structure}
-<- Equipment
-    <- \spi operator
-        <- Equivalent of Pi on finite sets
-        <- First argument: (finite) domain
-        <- Second argument: for each element of the domain, a co-domain
-        -> Inhabitant of \spi: right-nested tuple of solutions
-            <- Skip code for space reasons
-    <- switch operator
-        <- case analyses over x
-        <- index into the \spi tuple to retrieve the corresponding result
-\end{structure}
-
-\begin{structure}
-<- Equivalent to having a function space over finite sets
-    /> Made non-obvious by low-level encodings
-        <- General issue with codes
-             -> Need to provide an attractive presentation to the user
-    -> Types seem to obfuscate our reading
-        <- Provide ``too much'' information
-        /> False impression: information is actually waiting to be used more widely
-        -> See next Section
-\end{structure}
-
-\subsection{Type propagation}
-
-\begin{structure}
-<- Bidirectional type-checking [ref. Turner,Pierce]
-    -> Separating type-checking from type synthesis
-    <- Type checking: push terms into types
-        <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
-    <- Type inference: pull types out of terms
-        <- Example: |x : S l- x :<: S| gives us the type of x
-\end{structure}
-
-\begin{structure}
-<- Formalization: type propagation
-    <- Motivation
-        -> High-level syntax
-            -> exprIn: types are pushed in
-                <- Subject to type *checking*
-            -> exprEx: types are pulled from
-                <- Subject to type *synthesis*
-        -> Translated into our low-level type theory
-        -> Presented as judgements
-    -> Presentation mirors typing rule of [figure] 
-        -> Ignore identitical judgements
-\end{structure}
-
-\begin{structure}
-<- Type checking [figure]
-    <- Push a type in an exprIn
-    <- Result in a full term
-    -> *Use* the type to build the term!
-        -> Domain and co-domain propagation for Pi and Sigma
-        -> Translation of 'tags into EnumTs
-        -> Translation of ['tags ...] into EnumUs
-        -> Finite function space into switch
-\end{structure}
-
-\begin{structure}
-<- Type synthesis [figure]
-    <- Pull a type out of an exprEx
-    <- Result in a full term, together with its type
-    -> Do *not* need to specify types
-        -> Extracting a term from the context
-        -> Function application
-        -> Projections
-\end{structure}
-
-\begin{structure}
-<- Summary
-    -> Not a novel technique [refs?]
-        /> Used as a boilerplate scrapper
-    -> Make dealing with codes *practical*
-        <- Example: Finite sets/finite function space
-        -> We should not restrain ourself in using codes
-            <- We know how to present them to the user
--> Will extend this machinery in further sections
-\end{structure}
-
 \section{A universe of simple data-types}
@@
 
+\[
+\stk{
+\data \Desc : \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+    \DUnit          & \Desc \\
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
+    \DIndx          & \Desc \To \Desc
+\end{array}
+}
+\]
+
+\[\stk{
+\descop{\_\:}{} : \Desc \To \Set \To \Set \\
+\begin{array}{@{}ll@{\:=\:\:}ll}
+\descop{\DUnit}{& X}        &  \Unit                                       \\
+\descop{\DSigma{S}{D}}{& X} &  \SIGMAS{\V{s} : S}{\descop{D~s}{X}}         \\
+\descop{\DIndx{D}}{& X}     &  \TIMES{X}{\descop{D}{X}}
+\end{array}
+}\]
+
 \subsection{Examples}
@@
 
+\[\stk{
+\NatD : \Desc \\
+\NatD \mapsto \DSigma{(\EnumT [ \NatZero, \NatSuc{} ])}
+                     {[ \DUnit \quad (\DIndx{\DUnit}) ]}
+}\]
+
 \begin{structure}
@@
 
+\[\stk{
+\ListD : \Set \To \Desc \\
+\ListD \: X \mapsto \DSigma{(\EnumT [ \ListNil, \ListCons ])}
+                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{\DUnit}}) ]}
+}\]
+
 \begin{structure}
@@
 \end{structure}
+
+\[\stk{
+\TreeD : \Set \To \Desc \\
+\TreeD \: X \mapsto \DSigma{(\EnumT [ \TreeLeaf, \TreeNode ])}
+                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
+}\]
 
@@
 \end{structure}
+
+\[
+\Rule{\Gamma \vdash \Bhab{D}{\Desc}}
+     {\Gamma \vdash \Bhab{\Mu{D}}{\Set}} \qquad
+\Rule{\Gamma \vdash \Bhab{D}{\Desc} \quad 
+      \Gamma \vdash \Bhab{x}{\descop{D}{(\Mu{D})}}}
+     {\Gamma \vdash \Bhab{\Con{x}}{\Mu{D}}}
+\]
 
@@
     -> *Generic*
-\end{structure}
-
-\section{Levitating the universe of descriptions}
-
-\subsection{Implementing finite sets}
-
-\begin{structure}
-<- Recall typing rules of 1st section
-    -> Make clear they were just promises
-    -> Can be implemented now
-        <- Simply List UId
-\end{structure}
-
-\begin{structure}
-<- Consequences
-    -> Type theory doesn't need to be extended with EnumU, NilE, and ConsE
-        <- EnumU == Mu EnumUD
-        <- NilE, ConsE are just tags
-    -> Do not need a specific \spi eliminator
-        <- \spi is an instance of the generic eliminator
-            <- Code?
-    -> Anything else remains the same (switch, EnumT, 0, 1+)
-\end{structure}
-
-\begin{structure}
-<- Summary of the operation
-    <- The content of the type theory is exactly the same
-        <- before == after
-    /> type naming scheme condenses
-        <- Replace named constructors by codes in the universe of data-types
-    -> Our next step is a similar move (in essence)
-        /> Condenses the entire naming scheme of data-types
-\end{structure}
-
-\subsection{Implementing descriptions}
-
-\begin{structure}
-<- Realizing our promises
-    <- We are going to implement Desc
-    /> Desc is itself a data-type
-        <- Same situation as EnumU
-            <- We want to benefit from generic operations
-        -> It ought to be encoded in itself
-\end{structure}
-
-\subsubsection{First try}
-
-\begin{structure}
-<- A partial implementation
-    <- '1 and 'indx are easy
-    <- 'sigma is partially doable
-        /> lack the ability to do an higher-order inductive call
-    -> Show partial code [figure]
-\end{structure}
-
-\subsubsection{Second try}
-
-\begin{structure}
-<- Extending the universe of description
-    -> With higher-order induction
-    <- Intuition: index elements in X by H, and go on reading
-        -> indx is isomorph to hindx for H = 1
-    /> Keep indx
-        <- First order!
-        -> Extensionally equal to hindx 1
-        /> Practically, definitional equality on Sigma/Pi cannot cope with it
-    -> Show DescD code
-\end{structure}
-
-\subsubsection{Final move}
-
-\begin{structure}
-<- Subtlety: translation of [ ... ]
-    -> Let us do it manually
-        -> Code with problem for the motive of switch
-\end{structure}
-
-\begin{structure}
-<- The magician trick
-    <- Our problem is to give a motive for switch
-        /> We perfectly know what it ought to be: \_ -> DescD
-    -> Solution: extend the type theory with a special purpose switchD
-        -> Only extension required to the type theory!
-        -> Hidden away to the user by the syntactic sugar
-            -> Sufficient to ensure unavailability as a raw operator
-            <- Another instance of type propagation
-\end{structure}
-
-\begin{structure}
-<- Generic programming now!
-    <- Desc is just data
-        -> Can be manipulated
-    <- Free induction scheme on Desc
-        -> Ability to inspect data-types
-        -> Ability to program on data-types
-\end{structure}
-
-\subsubsection{Desc, atomically}
-
-\begin{structure}
-<- Adding hindx have introduced some duplication
-    <- indx == hindx 1
-    -> We can factor out commonalities 
-        /> Obtain an equivalent presentation
-        /> Still embeddable (refer to the Agda model)
-\end{structure}
-
-\begin{structure}
-<- Give new presentation [figure]
-    <- hindx have introduced the notion of function space: 'Pi
-    <- hindx and indx are both composed by a binary product and a left open term: 'x and 'id 
-    -> Straightforward translation to the new system [equation]
-\end{structure}
-
-\subsection{The generic catamorphism}
-
-\begin{structure}
-<- Making cata
     ???
@@
 
-\subsection{The generic Free Monad}
+\[
+\F{cata} : \PITEL{D}{\Desc}
+           \PI{T}{\Set}
+           (\descop{D}{T}\To T) \To 
+           \Mu{D} \To T 
+\]
 
-\begin{structure}
-!!! EARLIER !!!
-<- Tagged description
-    <- Form TDesc = List (UId x Desc) [equation]
-    <- Follow usual sums-of-product presentation of data-type
-        <- Finite set of constructors
-        <- Then whatever you want
-    -> Any Desc data-type can be turned into this form
-        -> No loss of expressive power
-        /> Garantee a ``constructor form''
-\end{structure}
-
-\begin{structure}
-<- A generic program: the free monad construction
-    <- Recall free monad construction in Haskell
-        -> Based on a functor F
-    <- Note that the free monad construction is itself defined by a functor
-        -> Extract it
-    <- Encode it in the Desc world [equation]
-        <- F is the Desc we start with
-        <- The free monad functor is what we have just defined
-        <- [\_]* : Desc -> Set -> Desc
-           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
-        -> Mu does the fix-point
-\end{structure}
-
-\begin{structure}
-<- A generic program: monadic substitution [equation]
-    <- subst : \forall T X Y. mu ([T]* X) -> (X -> mu ([T]* Y)) -> mu ([T]* Y)
-    -> Using Fold
-
-\end{structure}
-
-\section{Indexing descriptions}
-
-\begin{structure}
-!!! Need Help !!!
-<- Motivation
-    <- Desc: expressivity of simply-typed data-types: inductive types
-        <- Values do not influence types
-    /> Example: Vectors
-        <- Cannot be defined by just induction
-            <- Vectors of all size need to be defined at the *same* time
-            -> Defined as a *family* of types
-                -> Index
-        -> I -> IDesc I: Inductive family
-    ???
-\end{structure}
-
-\subsection{From Desc to IDesc}
-
-\begin{structure}
-<- Labelling Id
-    <- We had: data Desc : Set -> Set
-    -> We want: data IDesc : (I -> Set) -> Set
-        <- Indexed functor (?)
-        -> It is sufficient to label Id
-            <- Where the functor is built
-\end{structure}
-
-\begin{structure}
-<- Also replacing '1 by 'const  [figure]
-    <- For convenience
-        <- 'const X equivalent to 'sigma X (\_ -> '1)
-        /> Easier to abstract
-            <- Extensionally same
-            /> 'const more useful in practice
-\end{structure}
-
-\subsection{Examples}
-
-\begin{structure}
-<- Nat
-    -> [equation]
-    <- Non-indexed types lives in IDesc 1
-        -> This applies to all previous examples
-\end{structure}
-
-\begin{structure}
-<- Levitation [figure]
-    <- Following Desc encoding
-        /> Note: simple data-type
-            -> Live in IDesc 1
-    -> Behind the scene, relies on the special purpose switchD
-\end{structure}
-
-\begin{structure}
-<- Untyped lambda terms [figure]
-    ???
-\end{structure}
-
-\subsection{Constrained constructors}
-
-\begin{structure}
-<- Assuming a suitable notion of definitional equality for I
-    ???
-\end{structure}
-
-\begin{structure}
-<- Fin [figure]
-    <- Presentation with equality constraints
-    ???
-\end{structure}
-
-\begin{structure}
-<- Vectors
-    ???
-\end{structure}
-
-\begin{structure}
-<- GADTS
-    ???
-\end{structure}
-
-\begin{structure}
-<- Brady optimization
-    <- Source to source translation
-    <- Able to remove some constraints
-    -> Example: Fin [figure]
-\end{structure}
-
-\subsection{Free IMonad}
-
-\begin{structure}
-<- Untyped lambda terms
-    ???
-\end{structure}
-
-\begin{structure}
-<- IDesc
-    ???
-\end{structure}
-
-\section{Discussion}
-
-\begin{structure}
-!!! Need Help !!!
-<- Universe stratification
-    <- Stratified agda model
-        <- Fully stratified
-        <- Proof of iso between host and embedding
-    ???
-\end{structure}
-
-\end{document}
+\[
+\begin{array}{lcl}
+\F{induction} & : & \PITEL{D}{\Desc}                \\
+              &   & \PITEL{P}{\Mu{D} \To \Set}      \\
+              &   & \PITEL{m}{\PI{xs}{\descop{D}{(\Mu{D})}} \All{D}{(\Mu{D})}{P}{xs} \To P (\Con{xs})} \\
+              &   & \PI{x}{\Mu{D}} P x
+\end{array}
+\]

### 2010-03-09 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: merge outline into paper. (Levitation of Desc)

--- 2010-03-08 papers/icfp-2010-desc/outline.tex
@@
-\documentclass{article}
-
-\usepackage{verbatim}
-\usepackage{fullpage}
-
-%% Structure
-\newenvironment{structure}{\footnotesize\verbatim}{\endverbatim}
-%\newenvironment{structure}{\comment}{\endcomment}
-
-\begin{document}
-
-\begin{structure}
-How to read me:
-    '<-'  means "because"
-    '->'  means "therefore"
-    '/>'  means "but"
-
-The spatial tabbing encodes the scope. Things on the same column are
-linked by an implicit "and". Things under an argument justify it.
-
-I know, that's fucked up and you hate it. But that's much better to
-read that crap than reading my long frenglish sentences a la James
-Proust.
-\end{structure}
-
-\section{The Type Theory}
-
-\subsection{Base theory}
-
-\begin{structure}
-<- Presentation of the formalism
-    <- Standard presentation
-        -> No novelty here
-    <- 3 judgments [equation]
-        -> Context validity
-        -> Typing judgements
-        -> Equality judgements
-    <- Invariants [equation]
-        -> By induction on derivations
-    <- Judgemental equality
-        <- Presentation independant of particular implementation choice
-        -> Model in Agda, intensional
-        -> Used in Epigram, OTT
-\end{structure}
-
-\begin{structure}
-<- Context validity [figure]
-    <- Not much to be said
-\end{structure}
-
-\begin{structure}
-<- Typing judgements [figure]
-    <- Set in Set
-        -> For simplicity of presentation
-        -> Assume that a valid stratification can be inferred
-            <- Harper-Pollack, Luo, Courant
-        -> See later discussion
-    <- Standard presentation of Pi and Sigma types
-\end{structure}
-
-\begin{structure}
-<- Judgemental equality [figure]
-    <- symmetry, reflexivity, and transitivity
-    <- beta-rules for lambda and pair
-    <- xi-rule for functions
-    -> Agnostic in the notion of equality
-        <- Doesn't rely on a ``propositional'' equality
-\end{structure}
-
-\begin{structure}
-!!! Need Help !!!
-<- Meta-theoretical properties
-    <- Assuming a stratified discipline
-    <> The point here is to reassert that dependent types are not evil, 
-       there is no non-terminating type-checker, or such horrible lies <>
-    -> Strongly normalizing
-        -> Every program terminates
-    -> Type-checking terminates
-    ???
-\end{structure}
-
-\subsection{Finite sets}
-
-\begin{structure}
-<- Motivation
-    <- Finite sets could be encoded with Unit and Bool
-        /> Hinder the ability to name things
-    -> For convenience
-        <- Named elements
-        <- Referring by name instead of code
-        -> Types as coding presentation
-            /> Also as coding representation!
-\end{structure}
-
-\begin{structure}
-<- Implementation [figure]      
-    <- Tags
-        -> Purely informational token
-    <- EnumU
-        -> Universe of finite sets
-    <- EnumT e
-        -> Elements of finite set e
-\end{structure}
-
-\begin{structure}
-<- Equipment
-    <- \spi operator
-        <- Equivalent of Pi on finite sets
-        <- First argument: (finite) domain
-        <- Second argument: for each element of the domain, a co-domain
-        -> Inhabitant of \spi: right-nested tuple of solutions
-            <- Skip code for space reasons
-    <- switch operator
-        <- case analyses over x
-        <- index into the \spi tuple to retrieve the corresponding result
-\end{structure}
-
-\begin{structure}
-<- Equivalent to having a function space over finite sets
-    /> Made non-obvious by low-level encodings
-        <- General issue with codes
-             -> Need to provide an attractive presentation to the user
-    -> Types seem to obfuscate our reading
-        <- Provide ``too much'' information
-        /> False impression: information is actually waiting to be used more widely
-        -> See next Section
-\end{structure}
-
-\subsection{Type propagation}
-
-\begin{structure}
-<- Bidirectional type-checking [ref. Turner,Pierce]
-    -> Separating type-checking from type synthesis
-    <- Type checking: push terms into types
-        <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
-    <- Type inference: pull types out of terms
-        <- Example: |x : S l- x :<: S| gives us the type of x
-\end{structure}
-
-\begin{structure}
-<- Formalization: type propagation
-    <- Motivation
-        -> High-level syntax
-            -> exprIn: types are pushed in
-                <- Subject to type *checking*
-            -> exprEx: types are pulled from
-                <- Subject to type *synthesis*
-        -> Translated into our low-level type theory
-        -> Presented as judgements
-    -> Presentation mirors typing rule of [figure] 
-        -> Ignore identitical judgements
-\end{structure}
-
-\begin{structure}
-<- Type checking [figure]
-    <- Push a type in an exprIn
-    <- Result in a full term
-    -> *Use* the type to build the term!
-        -> Domain and co-domain propagation for Pi and Sigma
-        -> Translation of 'tags into EnumTs
-        -> Translation of ['tags ...] into EnumUs
-        -> Finite function space into switch
-\end{structure}
-
-\begin{structure}
-<- Type synthesis [figure]
-    <- Pull a type out of an exprEx
-    <- Result in a full term, together with its type
-    -> Do *not* need to specify types
-        -> Extracting a term from the context
-        -> Function application
-        -> Projections
-\end{structure}
-
-\begin{structure}
-<- Summary
-    -> Not a novel technique [refs?]
-        /> Used as a boilerplate scrapper
-    -> Make dealing with codes *practical*
-        <- Example: Finite sets/finite function space
-        -> We should not restrain ourself in using codes
-            <- We know how to present them to the user
--> Will extend this machinery in further sections
-\end{structure}
-
-\section{A universe of simple data-types}
-
-\begin{structure}
-<- Why starting with simple data-types
-    <- For pedagogical purposes
-        <- Data-types as we know them every day
-        /> Target dependent types
-    -> Cut down version of Induction Recursion
-        -> Presentation evolves independently as we go to dependent types
-\end{structure}
-
-\subsection{The power of $\Sigma$}
-
-\begin{structure}
-<- The duality of Sigma
-    <- Sigma generalizes sum over arbitrary arities
-        -> \Sigma A B == \Sigma_{x : A} B x
-    <- Sigma generalizes product to have a dependant second component
-        -> \Sigma A B == (x : A) \times (B x)
-\end{structure}
-
-\begin{structure}
-<- Data-types in the simply-typed world
-    -> "sums-of-product"
-        <- Sum of constructors
-        <- Product of arguments
-<- Data-types in the dependently-typed world
-    -> "sigmas-of-sigmas"
-    /> Need ability to manipulate these sigmas
-        -> Define a Code for data-types
-        -> Together with a sigma-based Interpretation
-\end{structure}
-
-\subsection{A universe of descriptions}
-
-\begin{structure}
-<- Introduction to Universe construction
-    <- Define a Code
-        -> Name objects
-    <- Define an Interpretation of codes into the type theory
-        -> Give a semantics to objects
-    -> Ability to manipulate code
-    -> Ability to compute with these objects
-\end{structure}
-
-\begin{structure}
-<- Justification of the code 
-    <- [both figures]: cannot be read separately
-    <- Mimic the standard grammar of data-types description
-        -> Just as we already know it
-        <- '\Sigma for making sigmas-of-sigmas
-        <- 'indx for exhibiting the functoriality
-            -> For recursive arguments
-        <- '1 for end of description
-\end{structure}
-
-\subsection{Examples}
-
-\begin{structure}
-<- Nat
-    <- Sum of zero, suc
-    <- zero case: done
-    <- suc case: leave open and done
-    -> NatD Z = 1 + Z
-\end{structure}
-
-\begin{structure}
-<- List
-    <- Sum of nil, cons
-    <- nil case: done
-    <- cons case: product of X with leave open and done
-    -> ListD X Z = 1 + X * Z
-\end{structure}
-
-\begin{structure}
-<- Tree
-    <- sum of leaf, node
-    <- leaf case: done
-    <- node case: product of X with two leave open and done
-    -> TreeD X Z = 1 + X * Z * Z
-\end{structure}
-
-\begin{structure}
-<- Fictive object [figure 'data Desc']
-    -> Must be read as a type signature
-    -> See further for its actual implementation
-        <- Subject to our levitation exercise
-\end{structure}
-
-\subsection{Fix-point}
-
-\begin{structure}
-<- Build the fix-point of functors
-    <- See examples: need to build their initial algebra
-    -> Extend the type theory with Mu/Con [figure]
-        <- Straightforward definition of a fixpoint
-            <- Interpret D with (Mu D) as subobjects
-\end{structure}
-
-\begin{structure}
-<- Extending type propagation
-    <- Data-type declaration turns into definitions
-        -> Straightforward translation to Desc
-        -> Creation of a variable referring to the structure
-    <- Push Mu to an applied name [figure]
-        -> Direct integration into the type propagation machinery
-    <- Labelled Mu
-        /> Just mention the possibility of labelling, no details required
-        -> For the user, objects have names rather than Mu of codes
-    -> Coded presentation is practical
-        <- The user never see a code
-\end{structure}
-
-\begin{structure}
-<- Elimination on Mu
-    <- We are used to foldD : \forall X. (desc D X -> X) -> mu D -> X
-        /> Not dependent
-        -> Cannot express some (which one again?) properties
-    -> Develop a dependent induction
-        <- Everywhere/All
-        <- Induction
-    -> *Generic*
-    ???
-\end{structure}
-
 \section{Levitating the universe of descriptions}
@@
 \end{structure}
+
+\[\stk{
+\EnumU : \Set \\
+\EnumU \mapsto \Mu{(\List~\UId)}
+}\]
 
@@
 
+\[\stk{
+\DescD : \Desc \\
+\begin{array}{@{}ll}
+\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{} ]) \\
+                           & \left[\begin{array}{l}
+                                   \DUnit                                \\
+                                   \DSigma{\Set}{(\LAM{\V{X}} ???)} \\
+                                   \DIndx{\DUnit}                                  \\
+                                   \end{array}
+                             \right]
+\end{array}
+}\]
+
 \subsubsection{Second try}
@@
 \end{structure}
+
+\[
+\stk{
+\data \Desc : \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
+    \ldots          & \:\:\ldots \\
+    \DHindx         & \PI{H}{Set} \Desc \To \Desc
+\end{array}
+}
+\]
+
+\[\stk{
+\descop{\_\:}{} : \Desc \To \Set \To \Set \\
+\begin{array}{@{}l@{\:=\:\:}ll}
+\ldots                        &  \ldots \\
+\descop{\DHindx{H}{D}}{X}     &  \TIMES{(H \To X)}{\descop{D}{X}}
+\end{array}
+}\]
+
+\[\stk{
+\DescD : \Desc \\
+\begin{array}{@{}ll}
+\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]) \\
+                           & \left[\begin{array}{l}
+                                   \DUnit                                \\
+                                   \DSigma{\Set}{(\LAM{\V{X}} \DHindx{X}{\DUnit})} \\
+                                   \DIndx{\DUnit}                                  \\
+                                   \DSigma{\Set}{(\LAM{\_} \DIndx{\DUnit})}
+                                   \end{array}
+                             \right]
+\end{array}
+}
+\]
 
@@
 
+\[
+\stk{
+\data \Desc : \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+    \DId            & \Desc                                   \\
+    \DUnit          & \Desc                                   \\
+    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc         \\
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
+    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
+\end{array}
+}
+\]
+
+\[\stk{
+\descop{\_\:}{} : \Desc \To \Set \To \Set \\
+\begin{array}{@{}l@{\:=\:\:}ll}
+\descop{\DId}{X}          &  X                                           \\
+\descop{\DUnit}{X}        &  \Unit                                       \\
+\descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
+\descop{\DSigma{S}{D}}{X} &  \SIGMA{\V{s}}{S} \descop{D\: s}{X}                \\
+\descop{\DPi{S}{D}}{X}    &  \PI{\V{s}}{S} \descop{D\: s}{X}            
+\end{array}
+}\]
+
+\[\begin{array}{l@{\:\mapsto\:\:}l}
+\DIndx{D}         & \DProd{\DId}{D}                      \\
+\DHindx{H}{D}     & \DProd{(\DPi{H}{(\LAM{\_} \DId)})}{D}
+\end{array}
+\]
+
 \subsection{The generic catamorphism}
@@
 \end{structure}
+
+This \(\F{induction}\) operator is the natural dependent elimination
+principle, but we might also benefit from the traditional \emph{catamorphism}
+or `fold operator' which accompanies a (weakly) initial algebra. We should
+like to have
+\[\stk{
+\F{cata} : \PITEL{D}{\Desc}
+           \PI{T}{\Set}
+           (\descop{D}{T}\To T) \To 
+           \Mu{D} \To T \\
+\F{cata}\: D\: T\: f \mapsto
+  \F{induction}\: D\: (\LAM{\_}T)\: (\LAM{xs\:ts} f\: ?)
+}\]
+but what should \(?\) be? We have \(xs:\descop{D}{\Mu{D}}\)
+and \(ts:\All{D}{(\Mu{D})}{(\LAM{\_}T)}{xs}\), so surely we can construct
+an element of \(\descop{D}{T}\) by replacing each recursive component from
+\(xs\) with its counterpart from \(ts\).
+
+\[\stk{
+\F{replace} : \stk{\PITEL{D}{\Desc}
+                   \PITEL{X,Y}{\Set}\\
+                   \PI{xs}{\descop{D}{X}} 
+                   \All{D}{X}{(\LAM{\_}Y)}{xs} \To
+                   \descop{D}{Y}} \\
+\F{replace}\: \DUnit\:          X\: Y\: \Void\:          \Void          \mapsto \Void \\
+\F{replace}\: (\DSigma{S}{D})\: X\: Y\: \pair{s}{xs}{}\: ys             \mapsto
+    \pair{s}{\F{replace}\: {D~s}\: X\: Y\: xs\: ys}{}         \\
+\F{replace}\: (\DIndx{D})\:     X\: Y\: \pair{x}{xs}{}\: \pair{y}{ys}{} \mapsto
+    \pair{y}{\F{replace}\: D\: X\: Y\: xs\: ys}{}
+}\]
 
@@
 
+A \emph{tagged} description is given by an inhabitant of
+\[
+ \F{TagDesc} \mapsto \List{\TIMES{\UId}{\Desc}}
+\]
+
+Datatypes specified in the conventional `sum of products' style naturally give rise to tagged descriptions. Of course, every description can be dorced into this style with a singleton choice of tag.
+
 \begin{structure}
@@
 \end{structure}
+
+We may then implement the \emph{free monad} construction as a
+transformation on \(\F{TagDesc}\).
+\[\stk{
+\FreeMonad{\_} : \F{TagDesc} \To \Set \To \F{TagDesc} \\
+\FreeMonad{\pair{E}{D}{}}\:X \mapsto
+\pair{\ListCons{\:\DVar{}}{E}}{\pair{\DConst{X}}{D}{}}{}
+}\]
+Of course, we must equip the resulting datatypes with operations delivering a monadic interface.
+As usual, \(\LAM{\x}\DVar{\x}\) performs the r\^ole of `return', embedding variables into terms.
+We shall also need a generic \emph{substitution} operator.
 
@@
 \end{structure}
-
-\section{Indexing descriptions}
-
-\begin{structure}
-!!! Need Help !!!
-<- Motivation
-    <- Desc: expressivity of simply-typed data-types: inductive types
-        <- Values do not influence types
-    /> Example: Vectors
-        <- Cannot be defined by just induction
-            <- Vectors of all size need to be defined at the *same* time
-            -> Defined as a *family* of types
-                -> Index
-        -> I -> IDesc I: Inductive family
-    ???
-\end{structure}
-
-\subsection{From Desc to IDesc}
-
-\begin{structure}
-<- Labelling Id
-    <- We had: data Desc : Set -> Set
-    -> We want: data IDesc : (I -> Set) -> Set
-        <- Indexed functor (?)
-        -> It is sufficient to label Id
-            <- Where the functor is built
-\end{structure}
-
-\begin{structure}
-<- Also replacing '1 by 'const  [figure]
-    <- For convenience
-        <- 'const X equivalent to 'sigma X (\_ -> '1)
-        /> Easier to abstract
-            <- Extensionally same
-            /> 'const more useful in practice
-\end{structure}
-
-\subsection{Examples}
-
-\begin{structure}
-<- Nat
-    -> [equation]
-    <- Non-indexed types lives in IDesc 1
-        -> This applies to all previous examples
-\end{structure}
-
-\begin{structure}
-<- Levitation [figure]
-    <- Following Desc encoding
-        /> Note: simple data-type
-            -> Live in IDesc 1
-    -> Behind the scene, relies on the special purpose switchD
-\end{structure}
-
-\begin{structure}
-<- Untyped lambda terms [figure]
-    ???
-\end{structure}
-
-\subsection{Constrained constructors}
-
-\begin{structure}
-<- Assuming a suitable notion of definitional equality for I
-    ???
-\end{structure}
-
-\begin{structure}
-<- Fin [figure]
-    <- Presentation with equality constraints
-    ???
-\end{structure}
-
-\begin{structure}
-<- Vectors
-    ???
-\end{structure}
-
-\begin{structure}
-<- GADTS
-    ???
-\end{structure}
-
-\begin{structure}
-<- Brady optimization
-    <- Source to source translation
-    <- Able to remove some constraints
-    -> Example: Fin [figure]
-\end{structure}
-
-\subsection{Free IMonad}
-
-\begin{structure}
-<- Untyped lambda terms
-    ???
-\end{structure}
-
-\begin{structure}
-<- IDesc
-    ???
-\end{structure}
-
-\section{Discussion}
-
-\begin{structure}
-!!! Need Help !!!
-<- Universe stratification
-    <- Stratified agda model
-        <- Fully stratified
-        <- Proof of iso between host and embedding
-    ???
-\end{structure}
-
-\end{document}

### 2010-03-09 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: merge outline into paper. (IDesc)

--- 2010-03-08 papers/icfp-2010-desc/outline.tex
@@
-\documentclass{article}
-
-\usepackage{verbatim}
-\usepackage{fullpage}
-
-%% Structure
-\newenvironment{structure}{\footnotesize\verbatim}{\endverbatim}
-%\newenvironment{structure}{\comment}{\endcomment}
-
-\begin{document}
-
-\begin{structure}
-How to read me:
-    '<-'  means "because"
-    '->'  means "therefore"
-    '/>'  means "but"
-
-The spatial tabbing encodes the scope. Things on the same column are
-linked by an implicit "and". Things under an argument justify it.
-
-I know, that's fucked up and you hate it. But that's much better to
-read that crap than reading my long frenglish sentences a la James
-Proust.
-\end{structure}
-
-\section{The Type Theory}
-
-\subsection{Base theory}
-
-\begin{structure}
-<- Presentation of the formalism
-    <- Standard presentation
-        -> No novelty here
-    <- 3 judgments [equation]
-        -> Context validity
-        -> Typing judgements
-        -> Equality judgements
-    <- Invariants [equation]
-        -> By induction on derivations
-    <- Judgemental equality
-        <- Presentation independant of particular implementation choice
-        -> Model in Agda, intensional
-        -> Used in Epigram, OTT
-\end{structure}
-
-\begin{structure}
-<- Context validity [figure]
-    <- Not much to be said
-\end{structure}
-
-\begin{structure}
-<- Typing judgements [figure]
-    <- Set in Set
-        -> For simplicity of presentation
-        -> Assume that a valid stratification can be inferred
-            <- Harper-Pollack, Luo, Courant
-        -> See later discussion
-    <- Standard presentation of Pi and Sigma types
-\end{structure}
-
-\begin{structure}
-<- Judgemental equality [figure]
-    <- symmetry, reflexivity, and transitivity
-    <- beta-rules for lambda and pair
-    <- xi-rule for functions
-    -> Agnostic in the notion of equality
-        <- Doesn't rely on a ``propositional'' equality
-\end{structure}
-
-\begin{structure}
-!!! Need Help !!!
-<- Meta-theoretical properties
-    <- Assuming a stratified discipline
-    <> The point here is to reassert that dependent types are not evil, 
-       there is no non-terminating type-checker, or such horrible lies <>
-    -> Strongly normalizing
-        -> Every program terminates
-    -> Type-checking terminates
-    ???
-\end{structure}
-
-\subsection{Finite sets}
-
-\begin{structure}
-<- Motivation
-    <- Finite sets could be encoded with Unit and Bool
-        /> Hinder the ability to name things
-    -> For convenience
-        <- Named elements
-        <- Referring by name instead of code
-        -> Types as coding presentation
-            /> Also as coding representation!
-\end{structure}
-
-\begin{structure}
-<- Implementation [figure]      
-    <- Tags
-        -> Purely informational token
-    <- EnumU
-        -> Universe of finite sets
-    <- EnumT e
-        -> Elements of finite set e
-\end{structure}
-
-\begin{structure}
-<- Equipment
-    <- \spi operator
-        <- Equivalent of Pi on finite sets
-        <- First argument: (finite) domain
-        <- Second argument: for each element of the domain, a co-domain
-        -> Inhabitant of \spi: right-nested tuple of solutions
-            <- Skip code for space reasons
-    <- switch operator
-        <- case analyses over x
-        <- index into the \spi tuple to retrieve the corresponding result
-\end{structure}
-
-\begin{structure}
-<- Equivalent to having a function space over finite sets
-    /> Made non-obvious by low-level encodings
-        <- General issue with codes
-             -> Need to provide an attractive presentation to the user
-    -> Types seem to obfuscate our reading
-        <- Provide ``too much'' information
-        /> False impression: information is actually waiting to be used more widely
-        -> See next Section
-\end{structure}
-
-\subsection{Type propagation}
-
-\begin{structure}
-<- Bidirectional type-checking [ref. Turner,Pierce]
-    -> Separating type-checking from type synthesis
-    <- Type checking: push terms into types
-        <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
-    <- Type inference: pull types out of terms
-        <- Example: |x : S l- x :<: S| gives us the type of x
-\end{structure}
-
-\begin{structure}
-<- Formalization: type propagation
-    <- Motivation
-        -> High-level syntax
-            -> exprIn: types are pushed in
-                <- Subject to type *checking*
-            -> exprEx: types are pulled from
-                <- Subject to type *synthesis*
-        -> Translated into our low-level type theory
-        -> Presented as judgements
-    -> Presentation mirors typing rule of [figure] 
-        -> Ignore identitical judgements
-\end{structure}
-
-\begin{structure}
-<- Type checking [figure]
-    <- Push a type in an exprIn
-    <- Result in a full term
-    -> *Use* the type to build the term!
-        -> Domain and co-domain propagation for Pi and Sigma
-        -> Translation of 'tags into EnumTs
-        -> Translation of ['tags ...] into EnumUs
-        -> Finite function space into switch
-\end{structure}
-
-\begin{structure}
-<- Type synthesis [figure]
-    <- Pull a type out of an exprEx
-    <- Result in a full term, together with its type
-    -> Do *not* need to specify types
-        -> Extracting a term from the context
-        -> Function application
-        -> Projections
-\end{structure}
-
-\begin{structure}
-<- Summary
-    -> Not a novel technique [refs?]
-        /> Used as a boilerplate scrapper
-    -> Make dealing with codes *practical*
-        <- Example: Finite sets/finite function space
-        -> We should not restrain ourself in using codes
-            <- We know how to present them to the user
--> Will extend this machinery in further sections
-\end{structure}
-
-\section{A universe of simple data-types}
-
-\begin{structure}
-<- Why starting with simple data-types
-    <- For pedagogical purposes
-        <- Data-types as we know them every day
-        /> Target dependent types
-    -> Cut down version of Induction Recursion
-        -> Presentation evolves independently as we go to dependent types
-\end{structure}
-
-\subsection{The power of $\Sigma$}
-
-\begin{structure}
-<- The duality of Sigma
-    <- Sigma generalizes sum over arbitrary arities
-        -> \Sigma A B == \Sigma_{x : A} B x
-    <- Sigma generalizes product to have a dependant second component
-        -> \Sigma A B == (x : A) \times (B x)
-\end{structure}
-
-\begin{structure}
-<- Data-types in the simply-typed world
-    -> "sums-of-product"
-        <- Sum of constructors
-        <- Product of arguments
-<- Data-types in the dependently-typed world
-    -> "sigmas-of-sigmas"
-    /> Need ability to manipulate these sigmas
-        -> Define a Code for data-types
-        -> Together with a sigma-based Interpretation
-\end{structure}
-
-\subsection{A universe of descriptions}
-
-\begin{structure}
-<- Introduction to Universe construction
-    <- Define a Code
-        -> Name objects
-    <- Define an Interpretation of codes into the type theory
-        -> Give a semantics to objects
-    -> Ability to manipulate code
-    -> Ability to compute with these objects
-\end{structure}
-
-\begin{structure}
-<- Justification of the code 
-    <- [both figures]: cannot be read separately
-    <- Mimic the standard grammar of data-types description
-        -> Just as we already know it
-        <- '\Sigma for making sigmas-of-sigmas
-        <- 'indx for exhibiting the functoriality
-            -> For recursive arguments
-        <- '1 for end of description
-\end{structure}
-
-\subsection{Examples}
-
-\begin{structure}
-<- Nat
-    <- Sum of zero, suc
-    <- zero case: done
-    <- suc case: leave open and done
-    -> NatD Z = 1 + Z
-\end{structure}
-
-\begin{structure}
-<- List
-    <- Sum of nil, cons
-    <- nil case: done
-    <- cons case: product of X with leave open and done
-    -> ListD X Z = 1 + X * Z
-\end{structure}
-
-\begin{structure}
-<- Tree
-    <- sum of leaf, node
-    <- leaf case: done
-    <- node case: product of X with two leave open and done
-    -> TreeD X Z = 1 + X * Z * Z
-\end{structure}
-
-\begin{structure}
-<- Fictive object [figure 'data Desc']
-    -> Must be read as a type signature
-    -> See further for its actual implementation
-        <- Subject to our levitation exercise
-\end{structure}
-
-\subsection{Fix-point}
-
-\begin{structure}
-<- Build the fix-point of functors
-    <- See examples: need to build their initial algebra
-    -> Extend the type theory with Mu/Con [figure]
-        <- Straightforward definition of a fixpoint
-            <- Interpret D with (Mu D) as subobjects
-\end{structure}
-
-\begin{structure}
-<- Extending type propagation
-    <- Data-type declaration turns into definitions
-        -> Straightforward translation to Desc
-        -> Creation of a variable referring to the structure
-    <- Push Mu to an applied name [figure]
-        -> Direct integration into the type propagation machinery
-    <- Labelled Mu
-        /> Just mention the possibility of labelling, no details required
-        -> For the user, objects have names rather than Mu of codes
-    -> Coded presentation is practical
-        <- The user never see a code
-\end{structure}
-
-\begin{structure}
-<- Elimination on Mu
-    <- We are used to foldD : \forall X. (desc D X -> X) -> mu D -> X
-        /> Not dependent
-        -> Cannot express some (which one again?) properties
-    -> Develop a dependent induction
-        <- Everywhere/All
-        <- Induction
-    -> *Generic*
-    ???
-\end{structure}
-
-\section{Levitating the universe of descriptions}
-
-\subsection{Implementing finite sets}
-
-\begin{structure}
-<- Recall typing rules of 1st section
-    -> Make clear they were just promises
-    -> Can be implemented now
-        <- Simply List UId
-\end{structure}
-
-\begin{structure}
-<- Consequences
-    -> Type theory doesn't need to be extended with EnumU, NilE, and ConsE
-        <- EnumU == Mu EnumUD
-        <- NilE, ConsE are just tags
-    -> Do not need a specific \spi eliminator
-        <- \spi is an instance of the generic eliminator
-            <- Code?
-    -> Anything else remains the same (switch, EnumT, 0, 1+)
-\end{structure}
-
-\begin{structure}
-<- Summary of the operation
-    <- The content of the type theory is exactly the same
-        <- before == after
-    /> type naming scheme condenses
-        <- Replace named constructors by codes in the universe of data-types
-    -> Our next step is a similar move (in essence)
-        /> Condenses the entire naming scheme of data-types
-\end{structure}
-
-\subsection{Implementing descriptions}
-
-\begin{structure}
-<- Realizing our promises
-    <- We are going to implement Desc
-    /> Desc is itself a data-type
-        <- Same situation as EnumU
-            <- We want to benefit from generic operations
-        -> It ought to be encoded in itself
-\end{structure}
-
-\subsubsection{First try}
-
-\begin{structure}
-<- A partial implementation
-    <- '1 and 'indx are easy
-    <- 'sigma is partially doable
-        /> lack the ability to do an higher-order inductive call
-    -> Show partial code [figure]
-\end{structure}
-
-\subsubsection{Second try}
-
-\begin{structure}
-<- Extending the universe of description
-    -> With higher-order induction
-    <- Intuition: index elements in X by H, and go on reading
-        -> indx is isomorph to hindx for H = 1
-    /> Keep indx
-        <- First order!
-        -> Extensionally equal to hindx 1
-        /> Practically, definitional equality on Sigma/Pi cannot cope with it
-    -> Show DescD code
-\end{structure}
-
-\subsubsection{Final move}
-
-\begin{structure}
-<- Subtlety: translation of [ ... ]
-    -> Let us do it manually
-        -> Code with problem for the motive of switch
-\end{structure}
-
-\begin{structure}
-<- The magician trick
-    <- Our problem is to give a motive for switch
-        /> We perfectly know what it ought to be: \_ -> DescD
-    -> Solution: extend the type theory with a special purpose switchD
-        -> Only extension required to the type theory!
-        -> Hidden away to the user by the syntactic sugar
-            -> Sufficient to ensure unavailability as a raw operator
-            <- Another instance of type propagation
-\end{structure}
-
-\begin{structure}
-<- Generic programming now!
-    <- Desc is just data
-        -> Can be manipulated
-    <- Free induction scheme on Desc
-        -> Ability to inspect data-types
-        -> Ability to program on data-types
-\end{structure}
-
-\subsubsection{Desc, atomically}
-
-\begin{structure}
-<- Adding hindx have introduced some duplication
-    <- indx == hindx 1
-    -> We can factor out commonalities 
-        /> Obtain an equivalent presentation
-        /> Still embeddable (refer to the Agda model)
-\end{structure}
-
-\begin{structure}
-<- Give new presentation [figure]
-    <- hindx have introduced the notion of function space: 'Pi
-    <- hindx and indx are both composed by a binary product and a left open term: 'x and 'id 
-    -> Straightforward translation to the new system [equation]
-\end{structure}
-
-\subsection{The generic catamorphism}
-
-\begin{structure}
-<- Making cata
-    <- Present the type signature
-    <- Starts with a call to generic induction
-        <- induction on Desc!
-        /> Show types at hand
-        -> Explain how to use inductive hypothesis
-    <- Implement the 'replace' function
-    -> Dependent-typeless catamorphism 
-\end{structure}
-
-\begin{structure}
-<- Deriving generic functions
-    <- Taking a Desc and computing a function
-        <- Desc comes equipped with an induction principle
-        -> Ability to compute more functions from it
-            -> More generic functions
-    <- Inspecting data-types
-        <- All described by a Desc code
-        -> Ability to explore the code
-            <- Desc equipped with an induction principle
-            -> Build new objects based on that structure
-\end{structure}
-
-\subsection{The generic Free Monad}
-
-\begin{structure}
-!!! EARLIER !!!
-<- Tagged description
-    <- Form TDesc = List (UId x Desc) [equation]
-    <- Follow usual sums-of-product presentation of data-type
-        <- Finite set of constructors
-        <- Then whatever you want
-    -> Any Desc data-type can be turned into this form
-        -> No loss of expressive power
-        /> Garantee a ``constructor form''
-\end{structure}
-
-\begin{structure}
-<- A generic program: the free monad construction
-    <- Recall free monad construction in Haskell
-        -> Based on a functor F
-    <- Note that the free monad construction is itself defined by a functor
-        -> Extract it
-    <- Encode it in the Desc world [equation]
-        <- F is the Desc we start with
-        <- The free monad functor is what we have just defined
-        <- [\_]* : Desc -> Set -> Desc
-           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
-        -> Mu does the fix-point
-\end{structure}
-
-\begin{structure}
-<- A generic program: monadic substitution [equation]
-    <- subst : \forall T X Y. mu ([T]* X) -> (X -> mu ([T]* Y)) -> mu ([T]* Y)
-        -> Using Fold
-    -> Consequences
-        <- We have free monad data-type
-            <- Term + variables
-        <- We have monad operations
-            <- Return / var
-            <- Substitution / bind
-\end{structure}
-
-\begin{structure}
-<- Deriving new data-structure and functions on them
-    <- Computing the Free Monad of a data-type
-        <- Derive new data-structure from previous one
-            <- It is just code
-        /> New data-structure comes with some equipment
-    <- Computing new functions on computed data-types
-        <- If data comes with structure, we ought to be able to capture it
-            <- Induction on Desc
-            -> Ability to compute over data
-\end{structure}
-
 \section{Indexing descriptions}
@@
 
+\[
+\stk{
+\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+    \DVar{}         & I \To \IDesc{I}                                   \\
+    \DConst{}       & \Set \To \IDesc{I}                                \\
+    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
+    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
+\end{array}
+}
+\]
+
+\[\stk{
+\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To \Set \To \Set                  \\
+\begin{array}{@{}l@{\:=\:\:}ll}
+\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
+\idescop{\DConst{X}}{I}{P}    &  X                                                   \\
+\idescop{\DProd{D}{D'}}{I}{P} &  \TIMES{\idescop{D}{I}{P}}{\idescop{D'}{I}{P}}       \\
+\idescop{\DSigma{S}{D}}{I}{P} &  \SIGMA{\V{s}}{S} \idescop{D\: s}{I}{P}                    \\
+\idescop{\DPi{S}{D}}{I}{P}    &  \PI{\V{s}}{S} \idescop{D\: s}{I}{P}            
+\end{array}
+}\]
+
 \subsection{Examples}
@@
 
+\[\stk{
+\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
+\begin{array}{ll}
+\IDescD~I \mapsto \DSigma{}{} & (\EnumT [ \DVar{},
+\DConst{},
+                                                                                    \DProd{}{},
+                                          \DSigma{}{}, 
+                                          \DPi{}{} ]) \\
+                              & \left[\begin{array}{l}
+                                      \DConst{I}                  \\
+                                      \DConst{\Set}               \\
+                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
+                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
+                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
+                                   \end{array}
+                             \right]
+\end{array}
+}\]
+
 \begin{structure}
@@
 \end{structure}
+
+\[\stk{
+\LambdaTD : \Nat \To \IDesc{\Nat} \\
+\begin{array}{@{}ll}
+\LambdaTD\: n = \DSigma{}{} & (\EnumT [ \LambdaTVar, \LambdaTApp, \LambdaTLam ]) \\
+                            & \left[\begin{array}{l}
+                                  \DConst{\Fin{n}} \\
+                                  \DProd{\DVar{n}}{\DVar{n}} \\
+                                  \DVar{(\NatSuc{n})}
+                              \end{array}\right]
+\end{array}
+}\]
 
@@
 \end{structure}
+
+\[\stk{
+\FinD : \Nat \To \IDesc{\Nat} \\
+\begin{array}{@{}ll}
+\FinD\: n = \DSigma{}{} & (\EnumT [ \FinZero , \FinSuc{} ]) \\
+                        & \left[
+                          \begin{array}{l}
+                            \DSigma{\Nat}{\LAM{m} \DConst{(n \PropEq \NatSuc{m})}} \\
+                            \DSigma{\Nat}{\LAM{m} \DProd{\DVar{m}}{\DConst{(n \PropEq \NatSuc{m})}}}
+                          \end{array}
+                          \right]
+\end{array}
+}\]
 
@@
 
+\[\stk{
+\FinD : \Nat \To \IDesc{\Nat} \\
+\begin{array}{@{}llll}
+\FinD\: 0            & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
+\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT [ \FinZero , \FinSuc{} ]) \\
+                     &   &             & \left[
+                                         \begin{array}{l}
+                                         \DConst{\Unit} \\
+                                         \DVar{n}
+                                         \end{array}
+                                         \right]
+\end{array}
+}\]
+
 \subsection{Free IMonad}
@@
 
+\[\stk{
+\LambdaTFreeD : \Nat \To \IDesc{\Nat} \\
+\begin{array}{@{}ll}
+\LambdaTFreeD\: n = \DSigma{}{} & (\EnumT [ \LambdaTApp, \LambdaTLam ]) \\
+                                & \left[\begin{array}{l}
+                                  \DProd{\DVar{n}}{\DVar{n}} \\
+                                  \DVar{(\NatSuc{n})}
+                                  \end{array}\right]
+\end{array}
+}\]
+
+\[\stk{
+\LambdaTD : \Nat \To \IDesc{\Nat} \\
+\LambdaTD\: \mapsto \FreeMonad{[\LambdaTFreeD]}{\Fin{}}
+}\]
+
 \begin{structure}
@@
 
-\section{Discussion}
+\[\stk{
+\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
+\begin{array}{@{}ll}
+\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT [ \DConst{},
+                                              \DProd{}{},
+                                              \DSigma{}{}, 
+                                              \DPi{}{} ]) \\
+                                  & \left[\begin{array}{l}
+                                        \DConst{\Set}               \\
+                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
+                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
+                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
+                                    \end{array}\right]
+\end{array}
+}\]
 
-\begin{structure}
-!!! Need Help !!!
-<- Universe stratification
-    <- Stratified agda model
-        <- Fully stratified
-        <- Proof of iso between host and embedding
-    ???
-\end{structure}
-
-\end{document}
+\[\stk{
+\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
+\IDescD\: I \mapsto \FreeMonad{[\IDescFreeD]}{I}\: I
+}\]

### 2010-03-09 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: merge outline into paper. (Discussion)

--- 2010-03-07 papers/icfp-2010-desc/outline.tex
@@
-\documentclass{article}
-
-\usepackage{verbatim}
-\usepackage{fullpage}
-
-%% Structure
-\newenvironment{structure}{\footnotesize\verbatim}{\endverbatim}
-%\newenvironment{structure}{\comment}{\endcomment}
-
-\begin{document}
-
-\begin{structure}
-How to read me:
-    '<-'  means "because"
-    '->'  means "therefore"
-    '/>'  means "but"
-
-The spatial tabbing encodes the scope. Things on the same column are
-linked by an implicit "and". Things under an argument justify it.
-
-I know, that's fucked up and you hate it. But that's much better to
-read that crap than reading my long frenglish sentences a la James
-Proust.
-\end{structure}
-
-\section{The Type Theory}
-
-\subsection{Base theory}
-
-\begin{structure}
-<- Presentation of the formalism
-    <- Standard presentation
-        -> No novelty here
-    <- 3 judgments [equation]
-        -> Context validity
-        -> Typing judgements
-        -> Equality judgements
-    <- Invariants [equation]
-        -> By induction on derivations
-    <- Judgemental equality
-        <- Presentation independant of particular implementation choice
-        -> Model in Agda, intensional
-        -> Used in Epigram, OTT
-\end{structure}
-
-\begin{structure}
-<- Context validity [figure]
-    <- Not much to be said
-\end{structure}
-
-\begin{structure}
-<- Typing judgements [figure]
-    <- Set in Set
-        -> For simplicity of presentation
-        -> Assume that a valid stratification can be inferred
-            <- Implicit syntax, Pollack (?, or better reference)
-        -> See later discussion
-    <- Standard presentation of Pi and Sigma types
-\end{structure}
-
-\begin{structure}
-<- Judgemental equality [figure]
-    <- symmetry, reflexivity, and transitivity
-    <- beta-rules for lambda and pair
-    <- xi-rule for functions
-    /> Assume transitivity, reflexity, and symmetry
-    -> Agnostic in the notion of equality
-        <- Doesn't rely on a ``propositional'' equality
-\end{structure}
-
-\begin{structure}
-!!! Need Help !!!
-<- Meta-theoretical properties
-    <- Assuming a stratified discipline
-        <- Harper-Pollack, Luo, Courant
-    <> The point here is to reassert that dependent types are not evil, 
-       there is no non-terminating type-checker, or such horrible lies <>
-    -> Strongly normalizing
-        -> Every program terminates
-    -> Type-checking terminates
-    ???
-\end{structure}
-
-\subsection{Finite sets}
-
-\begin{structure}
-<- Motivation
-    <- Finite sets could be encoded with Unit and Bool
-        /> Hinder the ability to name things
-    -> For convenience
-        <- Named elements
-        <- Referring by name instead of code
-        -> Types as coding presentation
-            /> Also as coding representation!
-\end{structure}
-
-\begin{structure}
-<- Implementation
-    <- Tags
-        -> Purely informational token
-    <- EnumU
-        -> Universe of finite sets
-    <- EnumT e
-        -> Elements of finite set e
-\end{structure}
-
-\begin{structure}
-<- Equipment
-    <- \spi operator
-        <- Equivalent of Pi on finite sets
-        <- First argument: (finite) domain
-        <- Second argument: for each element of the domain, a co-domain
-        -> Inhabitant of \spi: right-nested tuple of solutions
-            <- Skip code for space reasons
-    <- switch operator
-        <- case analyses over x
-        <- index into the \spi tuple to retrieve the corresponding result
-\end{structure}
-
-\begin{structure}
-<- Equivalent to having a function space over finite sets
-    /> Made non-obvious by low-level encodings
-        <- General issue with codes
-             -> Need to provide an attractive presentation to the user
-    -> Types seem to obfuscate our reading
-        <- Provide ``too much'' information
-        /> False impression: information is actually waiting to be used more widely
-        -> See next Section
-\end{structure}
-
-\subsection{Type propagation}
-
-\begin{structure}
-<- Bidirectional type-checking [ref. Turner,Pierce]
-    -> Separating type-checking from type synthesis
-    <- Type checking: push terms into types
-        <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
-    <- Type inference: pull types out of terms
-        <- Example: |x : S l- x :<: S| gives us the type of x
-\end{structure}
-
-\begin{structure}
-<- Formalization: type propagation
-    <- Motivation
-        -> High-level syntax
-            -> exprIn: types are pushed in
-                <- Subject to type *checking*
-            -> exprEx: types are pulled from
-                <- Subject to type *synthesis*
-        -> Translated into our low-level type theory
-        -> Presented as judgements
-    -> Presentation mirors typing rule of [figure] 
-        -> Ignore identitical judgements
-\end{structure}
-
-\begin{structure}
-<- Type checking [figure]
-    <- Push a type in an exprIn
-    <- Result in a full term
-    -> *Use* the type to build the term!
-        -> Domain and co-domain propagation for Pi and Sigma
-        -> Translation of 'tags into EnumTs
-        -> Translation of ['tags ...] into EnumUs
-        -> Finite function space into switch
-\end{structure}
-
-\begin{structure}
-<- Type synthesis [figure]
-    <- Pull a type out of an exprEx
-    <- Result in a full term, together with its type
-    -> Do *not* need to specify types
-        -> Extracting a term from the context
-        -> Function application
-        -> Projections
-\end{structure}
-
-\begin{structure}
-<- Summary
-    -> Not a novel technique [refs?]
-        /> Used as a boilerplate scrapper
-    -> Make dealing with codes *practical*
-        <- Example: Finite sets/finite function space
-        -> We should not restrain ourself in using codes
-            <- We know how to present them to the user
--> Will extend this machinery in further sections
-\end{structure}
-
-\section{A universe of simple data-types}
-
-\begin{structure}
-<- Why starting with simple data-types
-    <- For pedagogical purposes
-        <- Data-types as we know them every day
-        /> Target dependent types
-    -> Cut down version of Induction Recursion
-        -> Presentation evolves independently as we go to dependent types
-\end{structure}
-
-\subsection{The power of $\Sigma$}
-
-\begin{structure}
-<- The duality of Sigma
-    <- Sigma generalizes sum over arbitrary arities
-        -> \Sigma A B == \Sigma_{x : A} B x
-    <- Sigma generalizes product to have a dependant second component
-        -> \Sigma A B == (x : A) \times (B x)
-\end{structure}
-
-\begin{structure}
-<- Data-types in the simply-typed world
-    -> "sums-of-product"
-        <- Sum of constructors
-        <- Product of arguments
-<- Data-types in the dependently-typed world
-    -> "sigmas-of-sigmas"
-    /> Need ability to manipulate these sigmas
-        -> Define a Code for data-types
-        -> Together with a sigma-based Interpretation
-\end{structure}
-
-\subsection{A universe of descriptions}
-
-\begin{structure}
-<- Introduction to Universe construction
-    <- Define a Code
-        -> Name objects
-    <- Define an Interpretation of codes into the type theory
-        -> Give a semantics to objects
-    -> Ability to manipulate code
-    -> Ability to compute with these objects
-\end{structure}
-
-\begin{structure}
-<- Justification of the code 
-    <- [both figures]: cannot be read separately
-    <- Mimic the standard grammar of data-types description
-        -> Just as we already know it
-        <- '\Sigma for making sigmas-of-sigmas
-        <- 'indx for exhibiting the functoriality
-            -> For recursive arguments
-        <- '1 for end of description
-\end{structure}
-
-\subsection{Examples}
-
-\begin{structure}
-<- Nat
-    <- Sum of zero, suc
-    <- zero case: done
-    <- suc case: leave open and done
-    -> NatD Z = 1 + Z
-\end{structure}
-
-\begin{structure}
-<- List
-    <- Sum of nil, cons
-    <- nil case: done
-    <- cons case: product of X with leave open and done
-    -> ListD X Z = 1 + X * Z
-\end{structure}
-
-\begin{structure}
-<- Tree
-    <- sum of leaf, node
-    <- leaf case: done
-    <- node case: product of X with two leave open and done
-    -> TreeD X Z = 1 + X * Z * Z
-\end{structure}
-
-\begin{structure}
-<- Fictive object [figure 'data Desc']
-    -> Must be read as a type signature
-    -> See further for its actual implementation
-        <- Subject to our levitation exercise
-\end{structure}
-
-\subsection{Fix-point}
-
-\begin{structure}
-<- Build the fix-point of functors
-    <- See examples: need to build their initial algebra
-    -> Extend the type theory with Mu/Con [figure]
-        <- Straightforward definition of a fixpoint
-            <- Interpret D with (Mu D) as subobjects
-\end{structure}
-
-\begin{structure}
-<- Extending type propagation
-    <- Data-type declaration turns into definitions
-        -> Straightforward translation to Desc
-        -> Creation of a variable referring to the structure
-    <- Push Mu to an applied name [figure]
-        -> Direct integration into the type propagation machinery
-    <- Labelled Mu
-        /> Just mention the possibility of labelling, no details required
-        -> For the user, objects have names rather than Mu of codes
-    -> Coded presentation is practical
-        <- The user never see a code
-\end{structure}
-
-\begin{structure}
-<- Elimination on Mu
-    <- We are used to foldD : \forall X. (desc D X -> X) -> mu D -> X
-        /> Not dependent
-        -> Cannot express some (which one again?) properties
-    -> Develop a dependent induction
-        <- Everywhere/All
-        <- Induction
-    -> *Generic*
-\end{structure}
-
-\section{Levitating the universe of descriptions}
-
-\subsection{Implementing finite sets}
-
-\begin{structure}
-<- Recall typing rules of 1st section
-    -> Make clear they were just promises
-    -> Can be implemented now
-        <- Simply List UId
-\end{structure}
-
-\begin{structure}
-<- Consequences
-    -> Type theory doesn't need to be extended with EnumU, NilE, and ConsE
-        <- EnumU == Mu EnumUD
-        <- NilE, ConsE are just tags
-    -> Do not need a specific \spi eliminator
-        <- \spi is an instance of the generic eliminator
-            <- Code?
-    -> Anything else remains the same (switch, EnumT, 0, 1+)
-\end{structure}
-
-\begin{structure}
-<- Summary of the operation
-    <- The content of the type theory is exactly the same
-        <- before == after
-    /> type naming scheme condenses
-        <- Replace named constructors by codes in the universe of data-types
-    -> Our next step is a similar move (in essence)
-        /> Condenses the entire naming scheme of data-types
-\end{structure}
-
-\subsection{Implementing descriptions}
-
-\begin{structure}
-<- Realizing our promises
-    <- We are going to implement Desc
-    /> Desc is itself a data-type
-        <- Same situation as EnumU
-            <- We want to benefit from generic operations
-        -> It ought to be encoded in itself
-\end{structure}
-
-\subsubsection{First try}
-
-\begin{structure}
-<- A partial implementation
-    <- '1 and 'indx are easy
-    <- 'sigma is partially doable
-        /> lack the ability to do an higher-order inductive call
-    -> Show partial code [figure]
-\end{structure}
-
-\subsubsection{Second try}
-
-\begin{structure}
-<- Extending the universe of description
-    -> With higher-order induction
-    <- Intuition: index elements in X by H, and go on reading
-        -> indx is isomorph to hindx for H = 1
-    /> Keep indx
-        <- First order!
-        -> Extensionally equal to hindx 1
-        /> Practically, definitional equality on Sigma/Pi cannot cope with it
-    -> Show DescD code
-\end{structure}
-
-\subsubsection{Final move}
-
-\begin{structure}
-<- Subtlety: translation of [ ... ]
-    -> Let us do it manually
-        -> Code with problem for the motive of switch
-\end{structure}
-
-\begin{structure}
-<- The magician trick
-    <- Our problem is to give a motive for switch
-        /> We perfectly know what it ought to be: \_ -> DescD
-    -> Solution: extend the type theory with a special purpose switchD
-        -> Only extension required to the type theory!
-        -> Hidden away to the user by the syntactic sugar
-            -> Sufficient to ensure unavailability as a raw operator
-            <- Another instance of type propagation
-\end{structure}
-
-\begin{structure}
-<- Generic programming now!
-    <- Desc is just data
-        -> Can be manipulated
-    <- Free induction scheme on Desc
-        -> Ability to inspect data-types
-        -> Ability to program on data-types
-\end{structure}
-
-\subsubsection{Desc, atomically}
-
-\begin{structure}
-<- Adding hindx have introduced some duplication
-    <- indx == hindx 1
-    -> We can factor out commonalities 
-        /> Obtain an equivalent presentation
-        /> Still embeddable (refer to the Agda model)
-\end{structure}
-
-\begin{structure}
-<- Give new presentation [figure]
-    <- hindx have introduced the notion of function space: 'Pi
-    <- hindx and indx are both composed by a binary product and a left open term: 'x and 'id 
-    -> Straightforward translation to the new system [equation]
-\end{structure}
-
-\subsection{The generic catamorphism}
-
-\begin{structure}
-<- Making cata
-    ???
-\end{structure}
-
-\subsection{The generic Free Monad}
-
-\begin{structure}
-!!! EARLIER !!!
-<- Tagged description
-    <- Form TDesc = List (UId x Desc) [equation]
-    <- Follow usual sums-of-product presentation of data-type
-        <- Finite set of constructors
-        <- Then whatever you want
-    -> Any Desc data-type can be turned into this form
-        -> No loss of expressive power
-        /> Garantee a ``constructor form''
-\end{structure}
-
-\begin{structure}
-<- A generic program: the free monad construction
-    <- Recall free monad construction in Haskell
-        -> Based on a functor F
-    <- Note that the free monad construction is itself defined by a functor
-        -> Extract it
-    <- Encode it in the Desc world [equation]
-        <- F is the Desc we start with
-        <- The free monad functor is what we have just defined
-        <- [\_]* : Desc -> Set -> Desc
-           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
-        -> Mu does the fix-point
-\end{structure}
-
-\begin{structure}
-<- A generic program: monadic substitution [equation]
-    <- subst : \forall T X Y. mu ([T]* X) -> (X -> mu ([T]* Y)) -> mu ([T]* Y)
-    -> Using Fold
-
-\end{structure}
-
-\section{Indexing descriptions}
-
-\begin{structure}
-!!! Need Help !!!
-<- Motivation
-    <- Desc: expressivity of simply-typed data-types: inductive types
-        <- Values do not influence types
-    /> Example: Vectors
-        <- Cannot be defined by just induction
-            <- Vectors of all size need to be defined at the *same* time
-            -> Defined as a *family* of types
-                -> Index
-        -> I -> IDesc I: Inductive family
-    ???
-\end{structure}
-
-\subsection{From Desc to IDesc}
-
-\begin{structure}
-<- Labelling Id
-    <- We had: data Desc : Set -> Set
-    -> We want: data IDesc : (I -> Set) -> Set
-        <- Indexed functor (?)
-        -> It is sufficient to label Id
-            <- Where the functor is built
-\end{structure}
-
-\begin{structure}
-<- Also replacing '1 by 'const  [figure]
-    <- For convenience
-        <- 'const X equivalent to 'sigma X (\_ -> '1)
-        /> Easier to abstract
-            <- Extensionally same
-            /> 'const more useful in practice
-\end{structure}
-
-\subsection{Examples}
-
-\begin{structure}
-<- Nat
-    -> [equation]
-    <- Non-indexed types lives in IDesc 1
-        -> This applies to all previous examples
-\end{structure}
-
-\begin{structure}
-<- Levitation [figure]
-    <- Following Desc encoding
-        /> Note: simple data-type
-            -> Live in IDesc 1
-    -> Behind the scene, relies on the special purpose switchD
-\end{structure}
-
-\begin{structure}
-<- Untyped lambda terms [figure]
-    ???
-\end{structure}
-
-\subsection{Constrained constructors}
-
-\begin{structure}
-<- Assuming a suitable notion of definitional equality for I
-    ???
-\end{structure}
-
-\begin{structure}
-<- Fin [figure]
-    <- Presentation with equality constraints
-    ???
-\end{structure}
-
-\begin{structure}
-<- Vectors
-    ???
-\end{structure}
-
-\begin{structure}
-<- GADTS
-    ???
-\end{structure}
-
-\begin{structure}
-<- Brady optimization
-    <- Source to source translation
-    <- Able to remove some constraints
-    -> Example: Fin [figure]
-\end{structure}
-
-\subsection{Free IMonad}
-
-\begin{structure}
-<- Untyped lambda terms
-    ???
-\end{structure}
-
-\begin{structure}
-<- IDesc
-    ???
-\end{structure}
-
 \section{Discussion}
@@
 \end{structure}
-
-\end{document}

### 2010-03-09 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: Section 'type theory' done. Good night.

@@
 \section{The Type Theory}
+
+Our work is deeply rooted in the realm of dependently-typed functional
+languages. In particular, we have developped this technology in our
+on-going work on Epigram~\cite{pigs:epigram}. However, in this paper,
+we strive to present our technology in the broadest setting
+possible. To this end, we are going to present a type theory from the
+ground-up.
+
+To the reader new to type theory, this first section will provide a
+gentle introduction to dependently-typed systems. To the reader
+familiar with type theory, she will recognize the basis of her
+favorite type theory, be it Agda or Coq for instance. This scene sets,
+we have a common ground for discussion.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Presentation of the formalism
@@
         -> Equality judgements
-\end{structure}
+\end{wstructure}
+
+In the following, we adopt a standard presentation for our type
+theory. It is defined by three forms of judgements: context validity
+judgements, typing judgements, and equality judgements. Formally,
+these judgments are presented as follow:
 
@@
 \begin{array}{ll}
-\Gamma\vdash\Valid & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
-\Gamma\vdash \Bhab{t}{T} & \mbox{term \(t\) has type \(T\) in context \(\Gamma\)} \\
-\Gamma\vdash \Bhab{s \equiv t}{T} & \mbox{\(s\) and \(t\) are equal at type \(T\) in context \(\Gamma\)} \\
+\Gamma \vdash \Valid                & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
+\Gamma \vdash \Bhab{t}{T}           & \mbox{term \(t\) has type \(T\) in context \(\Gamma\)} \\
+\Gamma \vdash \Bhab{s \equiv t}{T}  & \mbox{\(s\) and \(t\) are equal at type \(T\) in context \(\Gamma\)} \\
 \end{array}
@@
 
-\begin{structure}
+\begin{wstructure}
     <- Invariants [equation]
         -> By induction on derivations
-\end{structure}
+\end{wstructure}
 
-The systems of inference rules will be formulated to ensure that that the
-following implications always hold by induction on derivations.
+Each of these judgements is populated by a respective system of
+inferences rules. These rules are formulated so as to ensure that the
+following implications always hold by induction on derivations:
+
 \[
 \begin{array}{l@{\;\;\Rightarrow\;\;}l}
-\Gamma\vdash \Bhab{t}{T}            & \Gamma\vdash\Valid \;\wedge\; \Gamma\vdash\Type{T} \\
-\Gamma\vdash s \equiv \Bhab{t}{T}   & \Gamma\vdash \Bhab{s}{T} \;\wedge\; \Gamma\vdash \Bhab{t}{T} \\
-\Gamma;\xS;\Delta \vdash J          & \Gamma\vdash \Bhab{s}{S} \;\Rightarrow\; 
-                                           \Gamma;\Delta[s/x] \vdash J[s/x] \\
+\Gamma            \vdash \Bhab{t}{T}            
+    & \Gamma \vdash \Valid\; \wedge\; \Gamma\vdash\Type{T} \\
+\Gamma            \vdash s \equiv \Bhab{t}{T}   
+    & \Gamma \vdash \Bhab{s}{T} \;\wedge\; \Gamma\vdash \Bhab{t}{T} \\
+\Gamma ; \xS ; \Delta \vdash J                      
+    & \Gamma \vdash \Bhab{s}{S} \;\Rightarrow\; 
+          \Gamma ; \Delta[s/x] \vdash J[s/x] \\
 \end{array}
@@
 
-\begin{structure}
+We are not going to prove the validity of these invariants. They
+follow rather straightforwardly from the induction rules. For formal
+proofs, we refer the reader to standard presentations of type theory,
+such as Luo's seminal work \cite{luo:utt}.
+
+\begin{wstructure}
     <- Judgemental equality
@@
         -> Used in Epigram, OTT
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+Noticeably, we have adopted a judgemental presentation of
+equality. Therefore, we are not tied to a particular implementation
+choice. In particular, our system has been modeled in Agda, which
+features an intensional equality. On the other hand, it is used in
+Epigram, which equality has a slightly extensional
+flavor~\cite{altenkirch:ott}. We expect users of propositional or even
+fully extensional systems to also find their way through this
+presentation.
+
+\begin{wstructure}
 <- Context validity [figure]
     <- Not much to be said
-\end{structure}
+\end{wstructure}
 
@@
 
-\begin{structure}
+The rules ensuring Context Validity (Fig.\ref{fig:context-validity})
+are unsurprisingly common. They simply garantee the validity of the
+empty context, as well as the linear extension of a context by a
+well-typed variable.
+
+\begin{wstructure}
 <- Typing judgements [figure]
@@
     <- Standard presentation of Pi and Sigma types
-\end{structure}
+\end{wstructure}
+
+The typing judgements (Fig.\ref{fig:typing-judgements}) are just as
+familiar. We recognize the standard presentation of $\Pi$ and $\Sigma$
+types, respectively inhabited by lambda terms and dependent
+pairs. Naturally, there are rules for function application and
+projections of $\Sigma$-types. Equal types can be substituted, thanks
+to the conversion rule.
+
+For the sake of presentation, we postulate a $\Set$ in $\Set$
+rule. Having this rule makes our type theory inconsistent, by Girard's
+paradox~\cite{girard:set-in-set}. However, it has been
+shown~\cite{harper:implicit-universe, luo:utt,
+  courant:explicit-universe} that a valid stratification can be
+inferred, automatically or semi-automatically. In the remaining of our
+presentation, we will assume that such a stratification exists, even
+though we will keep it implicit. We shall discuss this assumption in
+Section~\ref{sec:discussion}.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Judgemental equality [figure]
@@
         <- Doesn't rely on a ``propositional'' equality
-\end{structure}
+        -> Key: wide applicability of our proposal
+\end{wstructure}
+
+Finally, we define the rules governing judgemental equality in
+Figure~\ref{fig:judgemental-equality}. As one would expect,
+judgemental equality is respects symmety, reflexivity, and
+transitivity. Further, it also encompasses the $β$-rules for function
+application and pairs. Finally, equality under lambda is dealt by the
+$ξ$-rule.
+
+Crucially, being judgemental, this presentation is agnostic in the
+notion of equality actually implemented. Indeed, our typing and
+equality judgements do not rely on a ``propositional'' equality. This
+freedom is a key point in favour of the wide applicability of our
+proposal. Obviously, when implementing our technology in an existing
+type-theory, some opportunities arise. We will present some of them
+along the course of the paper, clearly specifying which extensions we
+require.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 !!! Need Help !!!
@@
     ???
-\end{structure}
+\end{wstructure}
+
+This completes our presentation of the type theory. Assuming a
+stratified discpline of universe, the system we have described enjoy
+some very strong meta-theoretical properties. Unlike simply typed
+languages, such as Haskell, dependently-typed systems are
+\emph{strongly normalizing}: every program that type-checks
+terminates. Moreover, type-checking is decidable and can therefore be
+implemented by a terminating algorithm.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Motivation
@@
         /> Hinder the ability to name things
+    <- W-types considered harmful?
+        ???
     -> For convenience
@@
             /> Also as coding representation!
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+As a motivating example, we are now going to extend the type theory
+with a notion of finite set. One could argue that there is no need for
+such an extension: finite sets, just as any data-structure, can be
+encoded inside the type theory. A well-known example of such encoding
+is the Church encoding of natural numbers, which is isomorphic to
+finite sets.
+
+% Another example is Martin-Löf's W-types~\cite{}.
+
+However, using encodings is impractical. In the case of finite sets,
+for instance, we would like to name the elements of the sets. Then, we
+need to be able to manipulate these elements by their name, instead of
+their encoding. While we are able to give names to encodings, it is
+extremely tedious to map the encodings back to a name. Whereas these
+objects have a structure, the structure is lost during the encoding,
+when they become anonymous inhabitants of a $Π$ or $Σ$-type.
+
+In the simply-typed world, we are used to see types as a coding
+presentation -- presentation of invariants, presentation of
+properties. In the dependently-typed world, we also learn to use types
+as a coding representation: finite sets being good citizens, they
+ought to be democratically represented at the type level. As we will
+see, this gives us the ability to name and manipulate them (this is
+were the democracy analogy goes crazy, I think).
+
+\begin{wstructure}
 <- Implementation [figure]      
@@
         -> Elements of finite set e
-\end{structure}
+\end{wstructure}
+
+The implementation of finite sets is presented in
+Figure~\ref{fig:typing-finite-set}. It is composed by three
+components. First, we define tags as inhabitants of the $\UId$ type. A
+tag is solely an informative token, used for display purposes. Finite
+sets inhabits the $\EnumU$ type. Unfolding the definition, we get that
+a finite set is a list of tags. Finally, elements of a finite set $u$
+belong to the corresponding $\EnumT{u}$ type. Intuitively, it
+corresponds to an index, a number, pointing to an element of $u$.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Equipment
@@
         <- index into the \spi tuple to retrieve the corresponding result
-\end{structure}
+\end{wstructure}
 
-\begin{figure}
+The $\Enum$ type comes with further machinery. Indeed, to manipulate
+the $\EnumU$ and $\EnumT{}$ objects, we need to equip our type theory
+with an appropriate \emph{elimination principle}. A first step
+consists in defining the $\spi{}{}$ operator:
 
-\[\stkl{
+\[
 %% spi
-\F{$\pi$} : \PITEL{\V{e}}{\EnumU}
+\spi{}{} : \PITEL{\V{e}}{\EnumU}
            \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set 
-\\
+\]
+
+A closer look to its type should reveal a certain familiarity with its
+equivalent for sets, $\Pi$. Indeed, $\spi{}{}$ defines a \emph{finite
+  function space}, mapping elements of finite sets to objects in
+$\Set$. Intuitively, this operator build a right-nested tuple,
+effectively mapping an object of $P\:i$ for each $i$ the finite
+domain.
+
+Having understood the meaning of $\spi{}{}$, the definition of the
+$\F{switch}$ eliminator is not surprising:
+
+\[
 %% switch
@@
 \end{array}
-}\]
+\]
 
-\caption{Operators on finite sets}
-\label{fig:operator-finite-set}
+Again, there is a clear equivalent in the full-$\Set$ world: function
+application. The operational behavior of $\F{switch}$ is
+straightforward: $x$ is peeled off as we move deeper inside the nested
+tuple $b$. When $x$ equals $\Ze$, we simply return the value we are
+pointing to.
 
-\end{figure}
-
-\begin{structure}
+\begin{wstructure}
 <- Equivalent to having a function space over finite sets
@@
         -> See next Section
-\end{structure}
+\end{wstructure}
+
+As we have seen, finite sets are not just a way of making
+data. Through the $\spi{}{}$ and $\F{switch}$ operators, they also
+come with a notion of finite function space. However, we had to
+extract that intuition from the type, by a careful reading. At a first
+glance, this seems to contradict our argument in favor of types for
+coding representation. Here, we are overflown by low-level, very
+precise type information.
+
+However, our situation is significantly different from the one we
+faced with encoded data: while we were suffering from a crucial lack
+of information, we are here facing too much information, hence loosing
+focus. This is a general issue with the usage of codes, as they convey
+much more information than what the user is used to see.
+
+As we will see in the following section, there exists a cure to this
+problem. In a nutshell, instead of being overflown by typing
+information, we will put it at work, automatically. The consequence is
+that, in such system, working with codes is \emph{practical}: one
+should not be worried by information overload, but how to use it as
+much as possible. Having mastered this difficulty, we should not be
+afraid by code for practical purposes.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Bidirectional type-checking [ref. Turner,Pierce]
@@
         <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
-    <- Type inference: pull types out of terms
+    <- Type synthesis: pull types out of terms
         <- Example: |x : S l- x :<: S| gives us the type of x
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+Our proposal to tame the problem of codes is deeply rooted in the
+bidirectional presentation of type-checking of Pierce and
+Turner~\cite{pierce:bidirectional-tc}. In this formalism,
+type-checking is divided into two complementary phases. In the
+\emph{type inference} phase, types are \emph{pulled} out of terms. A
+typical example is a variable in the context:
+
+\[
+\Axiom{\Gamma ; \xS ; \Delta \Vdash \propag{\x}
+                                           {\pull{\x}{S}}}
+\]
+
+Because the variable and its type are stored in the context, whenever
+this variable is used, we can pull its type out of it.
+
+On the other hand, in the \emph{type checking} phase, terms are
+\emph{pushed} into types. We are handed a type together with a term,
+our task consists in checking that the term is indeed of this type. We
+are therefore allowed, and actually encouraged, to use the information
+provided by the type. Therefore, we can relax our requirements on the
+term. A typical example is the lambda:
+
+\[
+\Rule{\Gamma ; \xS \Vdash \propag{\push{t}{T\: x}}
+                                 {t'}}
+     {\Gamma \Vdash \propag{\push{\LAM{\x} t}{\PiTy{S}{T}}}
+                           {\PLAM{\x}{S} t'}} 
+\]
+
+In our formal presentation of the type theory, we were forced to
+annotate the lambda by the type of its domain. In a bidirectional
+setting, the domain is provided by the $\Pi$ type we are pushing the
+lambda into. Hence we do not need to annotate the lambda.
+
+\begin{wstructure}
 <- Formalization: type propagation
@@
         -> Ignore identitical judgements
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+We adapt and formalize this \emph{type propagation} mechanism to our
+richer type theory. First of all, we define a high-level syntax,
+composed by \emph{expressions}. Expressions divided into two syntactic
+categories: $\exprIn$ into which types are pushed in, and $\exprEx$
+from which types are pulled out. Following the bidirectional
+presentation, $\exprIn$ are subject to type \emph{checking}, while
+$\exprEx$ are subject to type \emph{synthesis}.
+
+The purpose of type propagation is to \emph{elaborate} an high-level
+expression into a term of the type theory, together with its
+type. This presentation builds on the experience gained in
+implementing Epigram 1~\cite{mcbride:view-from-the-left}. Being
+entirely syntax driven, we present our formalism as a system of
+syntax-directed inference rules. As the presentation largely mirrors
+the inference rules of the type theory, we will ignore the judgements
+which are identical. We refer our reader to the associated technical
+report~\cite{chapman:desc-tech-report} for the complete system of
+rules.
+
+\begin{wstructure}
+<- Type synthesis [figure]
+    <- Pull a type out of an exprEx
+    <- Result in a full term, together with its type
+    -> Do *not* need to specify types
+        -> Extracting a term from the context
+        -> Function application
+        -> Projections
+\end{wstructure}
+
+The type synthesis rules are presented
+Figure~\ref{fig:type-synthesis}. As illustrated by the judgement form,
+a type synthesis problem is provided an expression $\exprEx$ in a
+context, which is elaborated into a low-level term together with its
+type. These terms act as \emph{sources} for the type propagation
+machinery: they make types out of expressions. We have already studied
+the case of a variable in context. This also include function
+application as well as projections. Further down the line, these types
+might propagate to other terms, thanks to type checking.
+
+\begin{figure}
+\input{figure_type_synthesis}
+\caption{Type synthesis}
+\label{fig:type-synthesis}
+\end{figure}
+
+\begin{wstructure}
 <- Type checking [figure]
@@
         -> Finite function space into switch
-\end{structure}
+\end{wstructure}
+
+Dually to type synthesis, type checking judgements
+(Fig.\ref{fig:type-checking}) act as \emph{sinks}. From an expression
+$\exprIn$ and a type propagated down to it, they elaborate a low-level
+term. This is precisely the availability of the type we are pushed
+into that makes the power of this technique. Its benefit on lambda
+abstraction and pairs is well-known. Let us explore the consequences
+on our coding scheme for finite sets.
+
+A first consequence is that we can actually denote elements of finite
+sets \emph{by their name}. Indeed, such expression will be pushed into
+a $\EnumT{u}$, where $\Bhab{u}{\EnumU}$ contains all the names
+populating the finite set. It is therefore easy to elaborate a name
+into the corresponding index.
+
+A second consequence concerns the definition of a finite set. We can
+abstract away the details of code and use a familiar syntax
+instead. Hence, we use the syntax $\Enum\: [t_1 \ldots t_n]$ to denote
+the finite sets composed by the nammed elements $t_1$, \ldots,
+$t_n$. The translation to elements of $\EnumU$ is straightforward.
+
+Finally, and more importantly, we are able to present our user an
+abstraction of \emph{finite} function space. This work as follow. When
+an expression $t$ is pushed into a function which domain is a finite
+set, we instead elaborate it as a term in a small $\pi$-type. That is,
+a finite function. To realize our promise of a big $\Pi$-type, we
+elaborate a call to the $\F{switch}$ eliminator.
 
 \begin{figure}
-
 \input{figure_type_checking}
-
 \caption{Type checking}
 \label{fig:type-checking}
-
-\end{figure}
-
-\begin{structure}
-<- Type synthesis [figure]
-    <- Pull a type out of an exprEx
-    <- Result in a full term, together with its type
-    -> Do *not* need to specify types
-        -> Extracting a term from the context
-        -> Function application
-        -> Projections
-\end{structure}
-
-\begin{figure}
-
-\input{figure_type_synthesis}
-
-\caption{Type synthesis}
-\label{fig:type-synthesis}
-
 \end{figure}

### 2010-03-10 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: finish up summary of Type Theory section.

@@
 \subsection{Type propagation}
+\label{sec:type-propagation}
 
@@
 
+%%! This might deserve a more thorough explanation, with snippet of code.
+
 Finally, and more importantly, we are able to present our user an
@@
 
+The abstraction we have developed in this section are entirely
+transparent to the user. In the case of finite function space, the
+user has never to explicitly \emph{call} the $\F{switch}$
+eliminator. Nor will she \emph{be presented} the elaborated term, only
+the high-level . Bidirectional type-checking and type-directed
+pretty-printing seal the low-level type theory behind the high-level
+expressions.
+
 \begin{figure}
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Summary
@@
 -> Will extend this machinery in further sections
-\end{structure}
+\end{wstructure}
+
+In this section, we have developped a type propagation system, based
+on bidirectional type-checking. Using bidirectional type-checking as a
+boilerplate scrapper is a well-known technique~\cite{}. In our case,
+we have shown how to instrument bidirectionality to rationalize the
+expressivity of dependent types. We have illustrated our approach with
+finite sets. We have successfully abstracted away the low-level
+presentation of finite sets, offering a convenient syntax instead.
+
+This example teaches us that we should not be afraid by codes, as far
+as type information is available. We have shown how to rationalize
+this information in a formal presentation. Hence, we have proved that
+programming with such objects is practical. As we introduce more codes
+in our theory, we will show how to extend the framework we have
+developped so far.

### 2010-03-10 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: Code and interpretation face to face, for all Desc and IDesc definitions.
Desc@ICFP: Second section on Desc almost done. It is just too late, or too early to do initial algebras.

@@
 
-\begin{structure}
+\begin{wstructure}
 <- Why starting with simple data-types
@@
         -> Presentation evolves independently as we go to dependent types
-\end{structure}
+\end{wstructure}
+
+In this section, we describe an alternative implementation of
+inductive types, as we know them in simply-typed languages such as
+Haskell or OCaml. This choice is motivated by pedagogical
+considerations: by describing the layman data-types in our setting, we
+build on our intuition of data-types we are familiar with. However,
+our goal is and remains to discuss data-types as we find them in
+dependently-typed languages.
+
+Our proposal is based on induction-recursion~\cite{dybjer:general-ir,
+  dybjer:axiom-ir, dybjer:ir-initial-algebra, dybjer:iir}. However, in
+the context of simple data-types, the full power of
+induction-recursion is not required. Therefore, we will present a
+stripped-down version. As we extend the scope of our universe of
+data-types, our presentation will diverge from induction-recursion. We
+shall compare both approaches in Section~\ref{sec:discussion}
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- The duality of Sigma
@@
         -> \Sigma A B == (x : A) \times (B x)
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+In dependently-typed language, $\Sigma$-types can be read in two
+rather different ways. This duality is actually reflected in the
+notation we can find in the literature, depending on the sensibility
+of the author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
+$\Sigma_{x : A} (B x)$. This notation stresses the fact that
+$\Sigma$-types are a generalization of sums over arbitrary
+arities. When simply-typed languages have finite sums,
+dependently-typed languages have sums of indexed by any set.
+
+On the other hand, $\SigmaTy{A}{B}$ is sometimes written
+$\SIGMA{\V{x}}{A} (B x)$. Under this perspective, $\Sigma$-types can
+be read as a generalization of products, where the second component
+can depend on the first one. When simply-typed languages pack data
+into non-dependent tuples, dependently-types languages tolerate this
+non-dependent usage. However, they also give the rather novel ability
+for data to influence further data.
+
+\begin{wstructure}
 <- Data-types in the simply-typed world
@@
         -> Together with a sigma-based Interpretation
-\end{structure}
+\end{wstructure}
+
+In the simply-typed world, the essence of data-type definition can be
+sumed up by the term ``sum-of-product''. A data-type is defined by a
+finite sum, choice, of constructors, each of them composed by a
+product, a tuple, of arguments. Therefore, the grammar of data-types
+definition in simply-typed languages, such as Haskell or OCaml, follow
+this form.
+
+To model these data-types, we are simply left with capturing this
+grammar in a dependently-typed setting. Hence, the notion of
+``sum-of-product'' naturally translates into
+\emph{sigmas-of-sigmas}. However, beyond this intuition, it should be
+clear that a direct encoding of data-types through raw $\Sigma$-types
+is not a viable option. Indeed, just as for finite sets, encoding
+throw away information when we crucially need it. In the realm of
+data-types, things become even more tougher, as one could legitimately
+expect an induction principle, for instance. Anonymous in a see of
+$\Sigma$s, our data-types have little chance of survival.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Introduction to Universe construction
@@
     -> Ability to compute with these objects
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+%%! Check Martin-Lof reference
+
+For our sigma-of-sigmas to be known, we crucially need the ability to
+name them. To this end, we will use a standard technique in
+dependently-typed programming: we construct an ad-hoc universe. This
+technique dates back to Martin-L\"of definition of type
+theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
+programming technique~\cite{}. We refer the reader to Agda's
+tutorial~\cite{norell:agda-tutorial} for a pedagogical presentation of
+universe construction.
+
+The key idea behind universe construction is our ability to make names
+by defining new types. These names are called \emph{codes}. By
+defining a set of codes, we somehow define the syntax of a
+language. However, as such, a system of code is useless as it lacks a
+semantics. Instead of equiping the universe of codes with some
+computational behavior, we pragmatically chose to \emph{interpret}
+these codes back into the standard type theory. Hence, codes act only
+as labels, while the type theory provides the computational machinery.
+
+Codes being simple labels, we have the ability to inspect them, hence
+taking advantage of their structural information. Being able to
+inspect them, we are therefore free to compute over them: deriving new
+codes from previous ones, or even new functions on them. In several
+occasions, we will have the opportunity to witness the power of
+universes.
+
+\begin{wstructure}
 <- Justification of the code 
@@
         <- '1 for end of description
-\end{structure}
+\end{wstructure}
+
+Hence, we propose to embed inductive types as a universe in our
+dependent type theory, the universe of \emph{descriptions}
+(Figure~\ref{fig:desc_universe}). As expected, the code of this
+universe mimics the standard grammar of data-types definitions in
+simply-typed languages. Hence, we have a $\DSigma{}{}$ code,
+interpreted as a $\Sigma$-type, to build the
+sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
+contains no useful payload. The functoriality of the data-types is
+introduced by $\DIndx$. When we tie the knot with a fix-point, the
+hole left open by $\DIndx$ will be turned into a standard recursive
+argument. We notice that this functoriality appears in the type of
+$\descop{D}{}$ itself, for a given data-type definition $D$. This
+corresponds to the morphism part of the functor described by $D$.
+
+\begin{figure*}
 
 \[
+\begin{array}{ll}
 \stk{
@@
 }
-\]
-
-\[\stk{
+&
+\stk{
 \descop{\_\:}{} : \Desc \To \Set \To \Set \\
@@
 \end{array}
-}\]
+}
+\end{array}
+\]
+
+\caption{Universe of Descriptions}
+\label{fig:desc_universe}
+
+\end{figure*}
+
+To give some intuition on this universe of descriptions, we now turn
+to some examples. For obvious pedagogical reasons, we will manually
+build these descriptions. However, it should be clear that, in
+practice, these definitions can be automatically constructed an
+Haskell-like $\data$ definition.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Nat
@@
     -> NatD Z = 1 + Z
-\end{structure}
+\end{wstructure}
+
+Our first example is the natural numbers, or rather its carrier
+functor. Our code is presented in the high-level expression language
+of Section~\ref{sec:type-propagation}. The translation back to the raw
+terms is laborious but should not pose any difficulty. The code is the
+following:
 
@@
 
-\begin{structure}
+Let us explain its construction. First, we use $\DSigma{}{}$ to build
+a sum between $\NatZero$ and $\NatSuc{}$. In the $\NatZero$ case, we
+reach the end of the description: there is no useful payload in that
+case. In the $\NatSuc{}$ case, we left one hole open for the recursive
+argument, and we close the description.
+
+In a more synthetic notation, we have simply implemented the following
+functor:
+
+\[    \NatD\: Z = 1 + Z    \]
+
+\begin{wstructure}
 <- List
@@
     -> ListD X Z = 1 + X * Z
-\end{structure}
+\end{wstructure}
+
+With a small change to the definition of $\NatD$, we obtain the
+carrier of lists:
 
@@
 
-\begin{structure}
+The modification consists in turning the $\NatSuc{}$ into a proper
+$\ListCons$ taking an argument in X followed by an inductive
+argument. In this case, we use $\DSigma{}{}$ in its product
+interpretation: we pack an element of $X$ together with the recursive
+argument. Easily, one sees that this code actually defines the
+following functor:
+
+\[    \ListD\: X Z = 1 + X \times Z     \]
+
+\begin{wstructure}
 <- Tree
@@
     -> TreeD X Z = 1 + X * Z * Z
-\end{structure}
+\end{wstructure}
+
+Finally, we are not limited to one recursive argument. This is
+demonstated by our description of a binary tree functor below:
 
@@
 \TreeD : \Set \To \Desc \\
-\TreeD \: X \mapsto \DSigma{(\EnumT [ \TreeLeaf, \TreeNode ])}
-                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
+\begin{array}{@{}ll}
+\TreeD \: X \mapsto \DSigma{ & (\EnumT [ \TreeLeaf, \TreeNode ]) \\}
+                           { & [ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
+\end{array}
 }\]
 
-\begin{structure}
+Again, we are at one evolutionary step away from $\ListD$. However,
+instead of single appeal to the induction code, we call it twice. The
+interpretation of this code corresponds to the following functor:
+
+\[    \TreeD\: X Z = 1 + X \times Z \times Z     \]
+
+\begin{wstructure}
 <- Fictive object [figure 'data Desc']
@@
         <- Subject to our levitation exercise
-\end{structure}
+\end{wstructure}
+
+So far, for conveniency, we have taken as granted the existence of
+$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
+particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
+and $\DUnit$ as standard type constructors. Hence, we would expect the
+type theory to be extended by these constructors and their typing
+rules. Although it makes no conceptual difference, the code of
+description should rather be read as a \emph{promise}. We promise the
+existence of such objects, satisfying the typing rules. It will be the
+subject of Section~\ref{sec:desc-levitate} to fulfill these promises.
 

### 2010-03-10 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: finish section on Desc.

@@
 the context of simple data-types, the full power of
-induction-recursion is not required. Therefore, we will present a
+induction-recursion is not necessary. Therefore, we will present a
 stripped-down version. As we extend the scope of our universe of
@@
 particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
-and $\DUnit$ as standard type constructors. Hence, we would expect the
-type theory to be extended by these constructors and their typing
-rules. Although it makes no conceptual difference, the code of
-description should rather be read as a \emph{promise}. We promise the
-existence of such objects, satisfying the typing rules. It will be the
-subject of Section~\ref{sec:desc-levitate} to fulfill these promises.
+and $\DUnit$ as type formers, extending the basic type theory. Hence,
+in an implementation of this type theory, we would have extended it
+with these constructors and their typing rules. Although it makes no
+conceptual difference, the code of description should rather be read
+as a specification. We promise the existence of such objects,
+satisfying the typing rules. It will be the subject of
+Section~\ref{sec:desc-levitate} to fulfill this promise, by actually
+implementing the specification.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Build the fix-point of functors
@@
             <- Interpret D with (Mu D) as subobjects
-\end{structure}
+\end{wstructure}
+
+%%! Strictly positive types anyone?
+
+So far, we have used our universe of descriptions to build
+functors. To illustrate its usage, we have implemented the signature
+functors of natural numbers, lists, and binary trees. However, the
+class of functors expressible with descriptions enjoy another
+property: they all admit a fix-point.
+
+Let us constructively prove this. First of all, for an object $D$
+describing a signature functor, we define its fix-point by the type
+former $\Mu{D}$. Inhabitants of the fix-point are materialized by the
+structure map: $\Con{x}$ inhabits $\Mu{D}$ if $x$ belongs to the
+interpretation of $D$ with the carrier of the initial algebra:
 
@@
 
-\begin{structure}
-<- Extending type propagation
-    <- Data-type declaration turns into definitions
-        -> Straightforward translation to Desc
-        -> Creation of a variable referring to the structure
-    <- Push Mu to an applied name [figure]
-        -> Direct integration into the type propagation machinery
-    <- Labelled Mu
-        /> Just mention the possibility of labelling, no details required
-        -> For the user, objects have names rather than Mu of codes
-    -> Coded presentation is practical
-        <- The user never see a code
-\end{structure}
+For the less categorically inclined reader, we can simply read this
+construction as \emph{tying the knot}: the holes left open by the
+functor are filled by its own recursive definition. Hence, we have
+recovered the data-types we are usually able to define in simply-typed
+languages.
 
-\begin{structure}
+\begin{wstructure}
 <- Elimination on Mu
@@
     ???
-\end{structure}
+\end{wstructure}
+
+Together with the type formers defined above, we would legibly expect
+an elimination principle. Following the categorical intuition, we
+could be tempted to provide it as a catamorphism:
 
@@
            \PI{T}{\Set}
-           (\descop{D}{T}\To T) \To 
+           (\descop{D}{T} \To T) \To 
            \Mu{D} \To T 
@@
 
+Whereas this definition is amply sufficient in the simply-typed world,
+it comes short of its promises in the dependently-typed realm. Indeed,
+this type is absolutely non dependent. As this operator will be the
+corner stone of any inductive definition, we are better off sharping
+it before use.
+
+The first step consists in turning $T$ into a predicate
+$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
+descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
+\To T$ by a similar process. We start with turning the arrow into a
+dependenty arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
+further analysis.
+
+Then, we have to translate the notion of algebra in our dependent
+setting. This is achieved by the $\All{D}{X}{P}{x}$ predicate that,
+intuitively, states that $\Bhab{P}{X \To \Set}$ holds everywhere in
+the sub-structures of $x$. Based on that knowledge, we ought to be
+able to prove that $P$ holds for $\Con{xs}$. We close our elimination
+principle by taking an argument $x$ in $\Mu{D}$ and claiming that the
+motive $P$ is realized on $x$.
+
 \[
-\begin{array}{lcl}
-\F{induction} & : & \PITEL{D}{\Desc}                \\
-              &   & \PITEL{P}{\Mu{D} \To \Set}      \\
-              &   & \PITEL{m}{\PI{xs}{\descop{D}{(\Mu{D})}} \All{D}{(\Mu{D})}{P}{xs} \To P (\Con{xs})} \\
-              &   & \PI{x}{\Mu{D}} P x
+\begin{array}{lcll}
+\F{induction} & : & \multicolumn{2}{l}{\PITEL{D}{\Desc}}                   \\
+              &   & \multicolumn{2}{l}{\PITEL{P}{\Mu{D} \To \Set}}         \\
+              &   & \PITEL{m}{& \PI{xs}{\descop{D}{(\Mu{D})}}              \\
+              &   &           & \All{D}{(\Mu{D})}{P}{xs} \To P (\Con{xs})} \\
+              &   & \multicolumn{2}{l}{\PI{x}{\Mu{D}} P x}
 \end{array}
 \]
+
+At this stage, we are still left with implementing $\F{All}$ and the
+$\F{induction}$ principle. For space reason, we refer the reader to
+our technical report for the actual code. Having this dependent
+catamorphism, we are empowered with the ability to $\emph{compute}$
+over descriptions. Note that this induction principle is the first
+manifestation of a generic operation over descriptions. Thanks to
+$\F{induction}$, any data-type we define automatically comes equipped
+with an induction principle.
+
+\begin{wstructure}
+<- Extending type propagation
+    <- Data-type declaration turns into definitions
+        -> Straightforward translation to Desc
+        -> Creation of a variable referring to the structure
+    <- Labelled Mu
+        /> Just mention the possibility of labelling, no details required
+        -> For the user, objects have names rather than Mu of codes
+    <- Push Mu to an applied name [figure]
+        -> Direct integration into the type propagation machinery
+    -> Coded presentation is practical
+        <- The user never see a code
+\end{wstructure}
+
+In this section, we have introduced a universe of description. We have
+shown how to manipulate these codes to build data-types. We have also
+presented a fix-point construction and its associated dependent
+catamorphism. However, one could argue that programming with these
+objects is not practical.
+
+Whereas simply-typed languages feature a convenient grammar for
+defining data-types and manipulating them, our proposal could be seen
+as a step backward. In our setting, the user would have to write a
+description instead of a sugared data-type definition. Then, to
+recognize an element of $\Nat$ for instance, she would have to
+identify $\Mu{}$ of a description as indeed the fix-point of the
+functor signature of natural numbers. Finally, she would have to write
+codes instead of the natural data-type constructor.
+
+As for finite sets previously, these symptoms clearly indicate that we
+suffer from too much type information. Just as for finite, the cure
+lies in rationalizing this information. Let us first tackle the issue
+of data-type declaration: unsurprisingly, the grammar of data-types we
+are used can be straighforwardly translated to the
+``sigmas-of-sigmas'' paradigm. Hence, a data-type definition is turned
+into the corresponding description. Then, at the level of definitions,
+we associate the name of the data-type to the fix-point of the
+description. The induction principle as well as other generic
+operations follow for free.
+
+Doing so, we seal data-type definition behind a convenient
+abstraction. We take advantage of this abstraction barrier to tamper
+the second issue. Namely, our user would rather be presented a name
+rather that a fix-point of a signature functor. Rather surprisingly,
+it is sufficient to \emph{label} the $\Mu{}$ constructor with the
+user-provided name. When pretty-printing this object, we just expose
+the label, instead of the definition in full.
+
+Regaining the ability to write data-type constructor is simply a
+matter of type propagation. We therefore extends the type-checking
+framework of Figure~\ref{fig:type-checking} with the following
+inference rule:
+
+%% I'm kind of shafted (haha) by this:
+%%     - We have not yet introduced tagged descriptions
+%%     - Tagged descriptions are much more complex now
+%%     - "cases.i" is a syntax I just made up
+\[
+\Rule{\forall e_i \in \vec{e}, \propag{\push{\descop{cases.i}{\Mu{(\DSigma{(\Enum\: \CN{const})}{\CN{cases}})}}}{e_i}}
+                        {e'_i}}
+     {\propag{\push{C\: \vec{e}}{\Mu{(\DSigma{(\Enum\: \CN{const})}{\CN{cases}})}}}
+             {\Con{(\etag{C}, \widetilde{e'})}}}\;\etag{C} \in \CN{cases}
+\]
+
+Hence, in the high-level expression language, the user types a
+familiar type constructors, applied to a telescope of arguments. In
+the type propagation setting, this expression will be pushed to a
+$\Mu{}$ type of a tagged description. It is then straightforward to
+lookup the tag and turn the constructor form into the appropriate
+code.
+
+This concludes our presentation of the universe of descriptions. In
+this section, we have seen how data-types from the simply-typed world
+can be expressed in our dependent-type setting. By relying on a
+universe construction technique, we benefit from the ability to
+inspect and manipulate these codes. This opens some appealing
+opportunities for generic programming. Finally, we have demonstrated
+that programming with codes is practical. To this end, we have
+extended the type propagation machinery. In this setting, we can
+define, interact, and build data-types as transparently as in the
+simply-typed world.

### 2010-03-11 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: close section Levitation.

@@
 \subsection{Examples}
+\label{sec:desc-examples}
 
@@
 \begin{wstructure}
+<- Tagged description
+    <- Form TDesc = List (UId x Desc) [equation]
+    <- Follow usual sums-of-product presentation of data-type
+        <- Finite set of constructors
+        <- Then whatever you want
+    -> Any Desc data-type can be turned into this form
+        -> No loss of expressive power
+        /> Garantee a ``constructor form''
+\end{wstructure}
+
+From the examples above, we observe that data-types are defined by a
+$\DSigma$ which first argument is a finite set of constructor. The
+descriptions fitting into this pattern are called \emph{tagged}
+description. Formally, we have:
+
+\[
+ \TagDesc \mapsto \SIGMA{\V{E}}{\EnumU} \spi{\V{E}}{\LAM{\_} \Desc}
+\]
+
+Again, this is a clear reminiscence of the ``sum-of-products'' style:
+we have a finite sum of constructors. Of course, every description can
+be expressed into this style, by using a singleton as single
+constructor.
+
+\begin{wstructure}
 <- Fictive object [figure 'data Desc']
@@
 \subsection{Fix-point}
+\label{sec:desc-fix-point}
 
@@
 %%     - We have not yet introduced tagged descriptions
-%%     - Tagged descriptions are much more complex now
-%%     - "cases.i" is a syntax I just made up
+%%     - Tagged description in a Mu, that just doesn't work...
 \[
-\Rule{\forall e_i \in \vec{e}, \propag{\push{\descop{cases.i}{\Mu{(\DSigma{(\Enum\: \CN{const})}{\CN{cases}})}}}{e_i}}
+\Rule{\forall e_i \in \vec{e}, \propag{\push{e_i}
+                                            {\descop{f\: i}{(\Mu{(\DSigma{(\Enum\: b)}{f})})}}}
                                       {e'_i}}
-     {\propag{\push{C\: \vec{e}}{\Mu{(\DSigma{(\Enum\: \CN{const})}{\CN{cases}})}}}
-             {\Con{(\etag{C}, \widetilde{e'})}}}\;\etag{C} \in \CN{cases}
+     {\propag{\push{C\: \vec{e}}{\Mu{(\DSigma{(\Enum\: b)}{f})}}}
+             {\Con{(\etag{C}, \widetilde{e'})}}}\;\etag{C} \in b
 \]

### 2010-03-11 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: Code and interpretation face to face, for all Desc and IDesc definitions.
Desc@ICFP: Second section on Desc almost done. It is just too late, or too early to do initial algebras.
Desc@ICFP: close section Levitation.

@@
 \section{Levitating the universe of descriptions}
+\label{sec:desc-levitate}
+
+In this section, we will fulfill several promises. One promise we made
+was to effectively implement the code of the $\Desc$ universe. Another
+promise we tacitly made is to implement the type former of finite
+sets. Despite being a perilous pedagogical exercise for the authors,
+we hope to convey to the reader the dizzy feeling of levitation,
+without the falling.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Recall typing rules of 1st section
@@
         <- Simply List UId
-\end{structure}
+\end{wstructure}
+
+In Section~\ref{sec:finite-sets}, we have promised an implementation
+of finite sets. It consists in the objects specified in
+Figure~\ref{fig:typing-finite-set}. In particular, we will now
+implement the $\EnumU$ type former and its constructors. Let us recall
+their specification:
+
+\[\stkc{
+%% EnumU
+\Rule{\Gamma \vdash \Valid}
+     {\Gamma \vdash \Type{\EnumU}} 
+\\
+%% NilE
+\Rule{\Gamma \vdash \Valid}
+     {\Gamma \vdash \Bhab{\NilE}{\EnumU}} 
+\qquad
+%% ConsE
+\Rule{\Gamma \vdash \Bhab{t}{\UId} \quad
+      \Gamma \vdash \Bhab{e}{\EnumU}}
+     {\Gamma \vdash \Bhab{\ConsE{t}{e}}{\EnumU}}
+}\]
+
+The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
+coincidence. $\EnumU$ is indeed simply isomorph to a list, with
+$\UId$s as parameter. However, in Section~\ref{sec:desc-examples}, we
+have shown how to build lists in the universe of
+description. Therefore, the actual implementation of $\EnumU$ sums up
+to the following, trivial definition:
 
@@
 \EnumU : \Set \\
-\EnumU \mapsto \Mu{(\List~\UId)}
+\EnumU \mapsto \Mu{(\ListD~\UId)}
 }\]
 
-\begin{structure}
+\begin{wstructure}
 <- Consequences
@@
     -> Anything else remains the same (switch, EnumT, 0, 1+)
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+Let us examine the consequences of our act. First of all, we discover
+that the type theory does not actually need to be extended with the
+type former $\EnumU$, the constructors $\NilE$ and $\ConsE$. We have
+just defined $\EnumU$ above, by fix-point over the signature functor
+$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
+$\ListCons$ constructors of lists.
+
+Another interesting consequence concerns the $\spi{}{}$ operator: it
+can actually be implemented with the generic $\F{induction}$
+principle. From an implementation perspective, this means that we
+remove an operator, hard-coded in the host language. This operator is
+simply implemented in the dependently-typed target language, as any
+other function.
+
+Apart from these changes, we are left with implementing the other
+components of finite sets. This consists in $\EnumT$, $\Ze$, $\Su$, as
+well as the $\F{switch}$ operator. Note that the actual implementation
+of $\EnumU$ does not influence our implementation of $\EnumT$: be they
+hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects just
+behave similarly. We have witnessed this effect when carrying this
+operation in Epigram, moving from an hard-coded presentation to a
+self-hosted one. Absolutely no changes to the $\EnumT$ world was
+required.
+
+\begin{wstructure}
 <- Summary of the operation
@@
         /> Condenses the entire naming scheme of data-types
-\end{structure}
+\end{wstructure}
+
+In this section, we have replaced a low-level presentation of finite
+sets by a self-hosted one, expressed in the universe of
+descriptions. However, formally, the content of the type theory
+remains unchanged: objects which were present before the modification
+are still there. Conversely, we have not introduced any spurious
+objects in the type theory.
+
+If not on the content, this modification had an effect on the
+names. The type naming scheme of the type theory has condensed: named
+type formes ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are now
+replaced by codes and their fix-point in the universe of
+descriptions. Our next step is similar, in essence: we are going to
+condense the entire naming scheme of data-types \emph{in itself}.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Realizing our promises
@@
         -> It ought to be encoded in itself
-\end{structure}
+\end{wstructure}
 
-\subsubsection{First try}
+We shall now fulfill our promises. We are going to implement the codes
+of the universe of descriptions. Interestingly, in and by itself, a
+code is nothing but a data-type. We are in the same situation than
+with $\EnumU$: we ought to be able to describe the codes of $\Desc$ in
+$\Desc$ itself. Hence, this code would be a first-class citizen, born
+with the standard, generic equipment of data-types.
 
-\begin{structure}
+\subsubsection{First attempt}
+
+\begin{wstructure}
 <- A partial implementation
@@
     -> Show partial code [figure]
-\end{structure}
+\end{wstructure}
+
+Our first tentative is the following:
 
@@
                                    \DUnit                                \\
-                                   \DSigma{\Set}{(\LAM{\V{X}} ???)} \\
+                                   \DSigma{\Set}{(\LAM{\V{S}} ???)}      \\
                                    \DIndx{\DUnit}                        \\
@@
 
-\subsubsection{Second try}
+Let us explain how we have proceeded and the obstacle we face. Much as
+the data-types we have seen so far, $\DescD$ first consists in chosing
+a constructor. Constructors correspond to the code's cases: $\DUnit$,
+$\DSigma$, and $\DIndx$. The reader will notice that the tagged
+notation we have used so far for $\Desc$ constructors now fully make
+sense: these were actually the tags we are defining.
 
-\begin{structure}
+The case of $\DUnit$ is trivial: we immediately reach the end of the
+data-type. $\DUnit$ has no payload. The case of $\DIndx{}$ comes
+naturally: it only takes a recursive argument -- a description. The
+case of $\DSigma$ is problematic. Recall the definition of $\DSigma$:
+
+\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc      \]
+
+So, we first pack an element $S$ of $\Set$. Then, we would like to
+express a notion of recursive argument \emph{indexed} by $S$. Because
+our presentation is entirely first-order so far, we are not able to
+express this notion. We shall remedy to this situation.
+
+\subsubsection{Second attempt}
+
+\begin{wstructure}
 <- Extending the universe of description
@@
     -> Show DescD code
-\end{structure}
+\end{wstructure}
+
+In order to capture a notion of higher-order induction, we need to
+extend our universe of descriptions. This consists in adding a code
+$\DHindx$ that takes an indexing set $H$. The code and its
+interpretation are presented in
+Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx$ use the elements
+of $H$ to index as many recursive arguments.
+
+\begin{figure*}
 
 \[
+\begin{array}{ll}
 \stk{
@@
 \end{array}
-}\]
+}
+\end{array}
+\]
+
+\caption{Higher-order universe of descriptions}
+\label{fig:hindx_desc}
+
+\end{figure*}
+
+Note that $\DIndx$ is isomorph to $\DHindx{\Unit}$. However, we
+tolerate this duplication and do not replace $\DIndx$ by the more
+general $\DHindx$. Indeed, unlike its counterpart, $DIndx$ is
+first-order. Therefore, while both codes are \emph{extensionally} the
+same, in most practical implementation, they would be dealt with
+rather differently. While we can definitional equality to cope with
+first order objects, the functional presentation introduced by
+$\DHindx$ is unlikely to be amenable to a purely definitional
+treatment.
+
+Equipped with $\DHindx$, we can describe the extended code of our
+universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
+The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
+achieve the description of the $\DSigma$ case, by a simple appeal to
+the higher-order induction on $S$. The $\DHindx$ case consists in
+packing the $H$ in $\Set$ with a recursive arguement.
+
+In a first glance, we have achieved our goal. We have described the
+codes of the universe of description. Taking the fix-point of this
+object gives us $\Desc$, up to isomorphism. We have implemented our
+code of data-types as a levitating object inside itself. However, this
+levitation operation, just as any magic trick, relies on an invisible
+cable. Let us reveal it, hence finishing our implementation.
+
+\begin{figure}
 
@@
                                    \DUnit                                            \\
-                                   \DSigma{\Set}{(\LAM{\V{X}} \DHindx{X}{\DUnit})} \\
+                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{S}{\DUnit})}   \\
                                    \DIndx{\DUnit}                                    \\
-                                   \DSigma{\Set}{(\LAM{\_} \DIndx{\DUnit})}
+                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
                                    \end{array}
@@
 \end{array}
-}
-\]
+}\]
+
+\caption{Levitating description of $\Desc$}
+\label{fig:desc-levitate}
+
+\end{figure}
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Subtlety: translation of [ ... ]
@@
         -> Code with problem for the motive of switch
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+Our reader might be slightly confused to learn that the trick is
+visible in the definition of $\DescD$
+(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by a
+careful usage of type propagation. Indeed, let us elaborate this term
+down to the low-level type theory. The $\EnumT [ \ldots ]$ construct
+elaborates to a finite set $e$ in $\EnumU$, inhabited by the
+codes. Then, we can type-check the case definition, between square
+brackets $[ \ldots ]$. This term $f$ is pushed into the type
+$\EnumT{e} \To \Desc$. This corresponds to a finite function set we
+have implemented in Figure~\ref{fig:type-checking}.
+
+What happens if we unfold the definition? We ought to build the following term:
+
+\[
+\PLAM{x}{(\EnumT{e})} \switch{e}{\PLAM{\_} \Desc}{\pi^f}{x}
+\]
+
+But this is rather problematic! We are still in the process of
+constructing $\DescD$, and the motive of $\F{switch}$ is abruptely
+begging for an $\eta$-expanded $\Desc$. Despite our willingness, we
+will never be able to materialize such motive.
+
+\begin{wstructure}
 <- The magician trick
@@
             <- Another instance of type propagation
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+If we cannot make it happen, we simply do not make it happen. We
+perfectly know what the motive ought to be. So, straightforwardly, we
+extend the type theory with two special-purpose operators:
+
+\[
+\begin{array}{@{}ll}
+%% piD
+\green{\pi_D} : & \PITEL{\V{e}}{\EnumU} \To \Set \\
+%% switchD
+\F{switchD} : & \PITEL{\V{e}}{\EnumU}               
+                \PITEL{\V{b}}{\spiD{e}}
+                \PITEL{\V{x}}{\EnumT{e}} \To \Desc
+\end{array}
+\]
+
+The entire work of the magician stands here, in these two
+extensions. One could be worried by the availability of such
+operations in our type theory. Indeed, it would be inconvenient for
+the user to have to decide between these two variants. In fact, we
+give no choice to the user, hence no inconvenience. Just as the
+$\F{switch}$ operator was sealed off by the type propagation
+machinery, $\F{switchD}$ is not provided to the user. We simply extend
+the type-checking rules (Fig.~\ref{fig:type-checking}) with the
+following inference rule:
+
+\[
+\Rule{\Gamma \Vdash \propag{\push{t}{\spiD{e}}}
+                           {t'}}
+     {\Gamma \Vdash \begin{array}{@{}l} 
+                        \propag{\push{t}{\EnumT{e} \To \Desc}}
+                               {\\ \PLAM{x}{(\EnumT{e})} \switchD{e}{t'}{x}}
+                    \end{array}}\;\mbox{t is $[]$ or $[a,b]$}
+\]
+
+\begin{wstructure}
 <- Generic programming now!
@@
         -> Ability to program on data-types
-\end{structure}
+\end{wstructure}
+
+%%! Maybe a bit of perspective would be good
+%%    <- Everything stands up with the Mu
+%%    <- Induction principle defined over Mu-things
+
+This concludes our levitation work. Beyond the beauty of the exercise,
+there is also several practical outcomes. First of all, this exercise
+reveals that the $\Desc$ universe is just plain data. Just as any
+piece of data, it can therefore be inspected and manipulated. However,
+unlike any data, it is expressed in the $\Desc$ universe. As a
+consequence, it is equipped, for free, with an induction
+principle. So, our ability to inspect and program with $\Desc$ is not
+restricted to the meta-language: we now have all the necessary
+equipment in the target language to \emph{program} over data-types. In
+this setting, generic programming is just about \emph{programming}.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Adding hindx have introduced some duplication
@@
         /> Still embeddable (refer to the Agda model)
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+Before moving on some examples of generic programming, we have to
+carry out some maintenance work. The extension of $\Desc$ with
+$\DHindx$ have introduced some duplication of functionality. In order
+to remove this clutter, we will factor out these commonalities. This
+results in a new definition of the $\Desc$ universe. This
+reorganization is strictly equivalent to the previous one
+(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
+self-descriptive. We refer the reader to our companion technical
+report for details.
+
+\begin{wstructure}
 <- Give new presentation [figure]
@@
     -> Straightforward translation to the new system [equation]
-\end{structure}
+\end{wstructure}
+
+The factorized universe of description is presented in
+Figure~\ref{fig:factor-desc}. The modifications are justified as
+follow. First, $\DHindx$ have introduced a notion of function space:
+we capture this notion by the $\DPi$ constructor. Second, both
+$\DHindx$ and $\DIndx$ are composed by a product and the functorial
+argument: we capture these two notions with $\DProd{}{}$ and $\DId$,
+respectively.
+
+In this finer-grained presentation, we can define $\DIndx$ and
+$\DHindx$ as follow:
+
+\[\begin{array}{l@{\:\mapsto\:\:}l}
+\DIndx{D}         & \DProd{\DId}{D}                      \\
+\DHindx{H}{D}     & \DProd{(\DPi{H}{(\LAM{\_} \DId)})}{D}
+\end{array}
+\]
+
+Consequently, the examples developped above can be straightforwardly
+translated into this new presentation. In the following, we adopt this
+last version as our de facto universe of inductive types. In
+particular, we are going to demonstrate several instances of generic
+programming.
+
+%%! Shall we talk about the Type Theory being Desc Zero? or such story?
+
+\begin{figure*}
 
 \[
+\begin{array}{ll}
 \stk{
@@
 \end{array}
-}\]
-
-\[\begin{array}{l@{\:\mapsto\:\:}l}
-\DIndx{D}         & \DProd{\DId}{D}                      \\
-\DHindx{H}{D}     & \DProd{(\DPi{H}{(\LAM{\_} \DId)})}{D}
+}
 \end{array}
@@
 
+\caption{Factorized universe of descriptions}
+\label{fig:factor-desc}
+
+\end{figure*}
+
 \subsection{The generic catamorphism}
 
-\begin{structure}
+\begin{wstructure}
 <- Making cata
@@
     -> Dependent-typeless catamorphism 
-\end{structure}
+\end{wstructure}
 
-This \(\F{induction}\) operator is the natural dependent elimination
-principle, but we might also benefit from the traditional \emph{catamorphism}
-or `fold operator' which accompanies a (weakly) initial algebra. We should
-like to have
+In Section~\ref{sec:desc-fix-point}, we have argued for the
+implementation of a dependently-typed $\F{induction}$ principle,
+instead of the more traditional catamorphism. However, in some
+circumstances, the full-power of a dependent elimination is not
+necessary. In the following, we propose to derive the catamorphism
+from the generic induction principle.
+
+The catamorphism can be characterized as an induction on the
+description $D$, with a non-dependent motive targetting $T$. Given a
+node $xs$ and the induction hypotheses, the method ought to build an
+element of $T$. Provided we know how to make an element of
+$\descop{D}{T}$, this step will be performed by the algebra $f$. Let
+us take a look at this jigsaw:
+
+\newcommand{\cata}{\F{cata}}
+
 \[\stk{
-\F{cata} : \PITEL{D}{\Desc}
+\cata : \PITEL{D}{\Desc}
            \PI{T}{\Set}
-           (\descop{D}{T}\To T) \To 
+           (\descop{D}{T} \To T) \To 
            \Mu{D} \To T \\
-\F{cata}\: D\: T\: f \mapsto
-  \F{induction}\: D\: (\LAM{\_}T)\: (\LAM{xs\:ts} f\: ?)
+\cata\: D\: T\: f \mapsto
+  \F{induction}\: D\: (\LAM{\_}T)\: (\LAM{xs\:hs} f\: ???)
 }\]
-but what should \(?\) be? We have \(xs:\descop{D}{\Mu{D}}\)
-and \(ts:\All{D}{(\Mu{D})}{(\LAM{\_}T)}{xs}\), so surely we can construct
-an element of \(\descop{D}{T}\) by replacing each recursive component from
-\(xs\) with its counterpart from \(ts\).
+
+We are left with implementing the \(???\) program. Recall that we have
+\(\Bhab{xs}{\descop{D}{\Mu{D}}}\) and
+\(\Bhab{hs}{\All{D}{(\Mu{D})}{(\LAM{\_} T)}{xs}}\) at hand. Our goal
+is to make an element of \(\descop{D}{T}\). Intuitively, $xs$ is of
+the right shape, but its sub-elements are of the wrong type. On the
+other hand, for each sub-element of $xs$, $hs$ gives us the
+corresponding element in $T$. Hence the name ``induction
+hypotheses''. Therefore, to construct an element of \(\descop{D}{T}\),
+we replace the recursive components of \(xs\) by their counterpart in
+\(hs\):
+
+%%! Update to the factorized presentation
 
@@
 
-\begin{structure}
+Filling the \(???\) with \(\F{replace}\: D\: (\Mu{D})\: T\: xs\: hs\) closes the
+problem. In the type theory, we have built a generic catamorphism. Any
+data-type will now come equipped with this operation, for free.
+
+\begin{wstructure}
 <- Deriving generic functions
@@
             -> Build new objects based on that structure
-\end{structure}
+\end{wstructure}
+
+With this example, we have shown how we can derive a generic
+operation, the catamorphism, from a pre-existing generic operation,
+the induction principle. This has been made possible by our ability to
+manipulate descriptions as first-class objects: the catamorphism is,
+basically, a function mapping a $\Desc$ to a data-type specific
+operation. Having the ability to program on $\Desc$ is the key to
+define such function. The induction principle of $\Desc$, inherited by
+levitation, plays a central role here: without an induction principle,
+it would be impossible to program on $\Desc$.
+
+Moreover, the $\F{replace}$ function demonstrates the benefit of an
+approach based on universes. The data-types living in the universe of
+descriptions, we are able to \emph{inspect} them. As shown by
+$\F{replace}$, it is easy to explore these structures, as well as
+building new ones.
 
@@
 
-\begin{structure}
-!!! EARLIER !!!
-<- Tagged description
-    <- Form TDesc = List (UId x Desc) [equation]
-    <- Follow usual sums-of-product presentation of data-type
-        <- Finite set of constructors
-        <- Then whatever you want
-    -> Any Desc data-type can be turned into this form
-        -> No loss of expressive power
-        /> Garantee a ``constructor form''
-\end{structure}
-
-A \emph{tagged} description is given by an inhabitant of
-\[
- \F{TagDesc} \mapsto \List{\TIMES{\UId}{\Desc}}
-\]
-
-Datatypes specified in the conventional `sum of products' style naturally give rise to tagged descriptions. Of course, every description can be dorced into this style with a singleton choice of tag.
-
-\begin{structure}
+\begin{wstructure}
 <- A generic program: the free monad construction
@@
         -> Extract it
+\end{wstructure}
+
+In this section, we will turn to a more ambitious generic operation on
+data-type. Given a functor, represented as a tagged description, we
+build the free monad over this functor.
+
+\newcommand{\FMFreeMonad}{\D{FreeMonad}}
+\newcommand{\FMFreeMonadF}{\D{FreeMonad$_{\CN{Free}}$}}
+\newcommand{\FMVar}{\C{Var}}
+\newcommand{\FMComposite}{\C{Composite}}
+
+Let us recall the free monad construction. Given a functor $F$, the
+free monad over $F$ is defined by the following data-type:
+
+\[
+\stk{
+\data \FMFreeMonad : \PITEL{\V{F}}{\Set \To \Set} 
+                     \PI{\V{X}}{\Set} 
+                     \Set 
+\where \\
+\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+    \FMVar           & X \To \FMFreeMonad\: F\: X                            \\
+    \FMComposite     & F (\FMFreeMonad\: F\: X) \To \FMFreeMonad\: F\: X    
+\end{array}
+}
+\]
+
+%%! Recall monadic structure of this object
+
+Being an inductive type, this $\FMFreeMonad$ data-type is itself
+defined by a functor signature. It is given by:
+
+\[
+\FMFreeMonadF\: F\: X\: Z \mapsto X + F Z
+\]
+
+\begin{wstructure}
     <- Encode it in the Desc world [equation]
@@
         -> Mu does the fix-point
-\end{structure}
+\end{wstructure}
 
-We may then implement the \emph{free monad} construction as a
-transformation on \(\F{TagDesc}\).
+In our setting, the functor will be represented by a tagged
+description. The free monad construction will take such tagged
+description, a set $X$ of variables, and will compute the tagged
+description of the corresponding free monad.
+
+Implementing this function is surprisingly easy:
+
 \[\stk{
-\FreeMonad{\_} : \F{TagDesc} \To \Set \To \F{TagDesc} \\
+\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
 \FreeMonad{\pair{E}{D}{}}\:X \mapsto
-\pair{\ListCons{\:\DVar{}}{E}}{\pair{\DConst{X}}{D}{}}{}
+    \pair{\ListCons{\DVar{}}{E}}
+         {\pair{\DSigma{X}{\Unit}}{D}{}}{}
 }\]
-Of course, we must equip the resulting datatypes with operations delivering a monadic interface.
-As usual, \(\LAM{\x}\DVar{\x}\) performs the r\^ole of `return', embedding variables into terms.
-We shall also need a generic \emph{substitution} operator.
 
-\begin{structure}
+We simply add a constructor, $\DVar{}$, and define its argument to be
+a $\DSigma{X}{\Unit}$, that is an element of $X$. We keep $E$ and $D$
+as they were, hence leaving the functor untouched. The interpretation
+of this definition convince us that we have built the functor
+signature of the free monad. The fix-point operation ties the knot and
+gives us the full-blown free monad construction.
+
+\begin{wstructure}
 <- A generic program: monadic substitution [equation]
@@
         -> Using Fold
+\end{wstructure}
+
+Of course, we must equip the resulting datatypes with operations
+delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
+performs the r\^ole of ``return'', embedding variables into terms. The
+``bind'' operation corresponds to \emph{substitution}. We will now
+implement it, as a generic function.
+
+Our implementation will appeal to the $\cata$ function developped
+previously. So, let us write down the types, and fill as much
+arguments to $\cata$ as possible:
+
+\newcommand{\subst}{\F{subst}}
+\newcommand{\apply}{\F{apply}}
+
+%%! This is not quite Mu of a tagged description
+
+\[\stk{
+\begin{array}{@{}ll}
+\subst : & \PITEL{\V{D}}{\TagDesc}
+           \PI{\V{X}, \V{Y}}{\Set} \\
+         & \Mu{(\FreeMonad{D}{X})} \To
+           (\V{X} \To \Mu{(\FreeMonad{D}{Y})}) \To
+           \Mu{(\FreeMonad{D}{Y})} 
+\end{array} \\
+\subst\: D\: X\: Y\: x\: \sigma \mapsto
+  \cata\: (\FreeMonad{D}{X})\: (\Mu{(\FreeMonad{D}{Y})})\: ???\: x
+}\]
+
+Our task is therefore to implement the algebra of the
+catamorphism. Intuitively, its role is to catch appearance of
+$\DVar{x}$, to replace it by $\sigma x$. Otherwise, it should simply
+appeal to the induction hypothesis. This corresponds to the following
+definition:
+
+%%! Triple check. That doesn't work, right?
+
+\[\stk{
+\begin{array}{@{}ll}
+\apply : & \PITEL{\V{D}}{\TagDesc} 
+           \PI{\V{X}, \V{Y}}{\Set} \\
+         & (\V{X} \To \Mu{(\FreeMonad{D}{X})}) \To
+           \descop{\FreeMonad{D}{X}}{\Mu{(\FreeMonad{D}{Y})}} \To
+           \Mu{(\FreeMonad{D}{Y})}                            \\
+\end{array} \\
+\begin{array}{@{}l@{\:\:\mapsto\:}l}
+\apply\: \DVar{}\: X\: Y\: \sigma\: x  & \sigma x     \\
+\apply\: xs\:      X\: Y\: \sigma\: ys & ys
+\end{array}
+}\]
+
+\begin{wstructure}
     -> Consequences
@@
             <- Substitution / bind
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+Filling the $???$ subgoal with $\apply\: D\: X\: Y\: \sigma$ finishes
+the implementation. To sum up, we have implemented the free monad
+construction for an arbitrary tagged description. This gives our user
+the ability, for any data-type, to extend it with a notion of
+variable. Then, with have equipped this structure with the
+corresponding monadic operation, ``bind'' and ``return''. Whereas the
+``return'' is a trivial variable former, the ``bind'' operation
+corresponds to the substitution of variables by terms. We have shown
+how to implement it, for any free monad.
+
+\begin{wstructure}
 <- Deriving new data-structure and functions on them
@@
             -> Ability to compute over data
-\end{structure}
+\end{wstructure}
+
+With the free monad construction, we have seen two kind of generic
+operations. Firstly, we have derived a new data-structure from another
+one: we make the free monad from its underlying functor. To do so, we
+crucially rely on the fact that data-types are nothing but codes. We
+are therefore entitled to modify it and, in this case, extend
+it. Extending a data-type might give raise to a more
+structured object, as was the case here.
+
+So, secondly, we have equipped this new data-type with its inherent
+structure: the bind and return operations. We have been able to build
+them as generic functions. Here, we rely on the ability to compute
+over descriptions, thanks to the induction principle.

### 2010-03-11 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: close section Levitation.
Desc@ICFP: fix \cite{} which was preventing compilation on ancient version of pdflatex.

@@
 \subsection{Finite sets}
+\label{sec:finite-sets}
 
@@
 
-The implementation of finite sets is presented in
+The specification of finite sets is presented in
 Figure~\ref{fig:typing-finite-set}. It is composed by three
@@
 on bidirectional type-checking. Using bidirectional type-checking as a
-boilerplate scrapper is a well-known technique~\cite{}. In our case,
+boilerplate scrapper is a well-known technique~\cite{who?}. In our case,
 we have shown how to instrument bidirectionality to rationalize the

### 2010-03-11 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Code and interpretation face to face, for all Desc and IDesc definitions.
Desc@ICFP: finish IDesc section.

@@
     ???
-\end{structure}
+\end{wstructure}
+
+So far, we have explored the well-known realm of inductive types. We
+have built upon the experience gained in using simply-typed languages,
+such as Haskell or OCaml. In our dependent setting, we provided these
+data-types by the mean of $\Desc$, a universe of descriptions.
+
+While inductive types are the alpha and omega of the simply-typed
+languages, evolving in a dependent setting fosters new
+opportunities. The typical example is bounded lists, also known as
+vectors. A vector is a list decorated by its length. Having this
+information prevents hazardous operations, such as taking the head of
+an empty vector: the head function only takes vectors of length
+$\NatSuc{n}$, as enforced by its type. This is made possible by the
+very specificity of dependently-typed systems: terms -- the length --
+are allowed to influence types -- the vector type.
+
+However, this new class of objects cannot be defined by mere
+induction. In the case of vectors, for instance, we have to define the
+whole \emph{family} of vector in one go: vectors of all size needs to
+be defined at the same time. Where simply-typed languages have
+inductive types, we adopt inductive family of types as our basic
+grammar of data-types. To this end, we rely on \emph{indexing}. We are
+going to transform the $\Desc$ universe into an indexed $\IDesc$
+universe. We express inductive families by the $I \To \IDesc{I}$ type.
 
@@
 
-\begin{structure}
+\begin{wstructure}
 <- Labelling Id
@@
             <- Where the functor is built
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+In the previous section, we have presented the $\Desc$ universe as the
+grammar of functors in the category $\Set$. We have seen how to encode
+inductive types in this setting. To encode an inductive family indexed
+by $\Bhab{\V{I}}{\Set}$, we rely on functors in the category
+$\Set^I$. We call these \emph{indexed functors}. When $\Desc$ was
+defining a grammar a functor, we need $I \To \IDesc{I}$ to define a
+grammar of indexed functors:
+
+\[\stk{
+\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
+\;\;\ldots \\
+\\
+\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set    \\
+\ldots
+}\]
+
+This generalization can be achieved by a minor evolution of $\Desc$
+(Fig.~\ref{fig:factor-desc}). Indeed, the functorial nature of $\Desc$
+is entirely captured by the $\DId$ code. We get an indexed functor
+by turning $\DId$ into an index-dependent $\DVar$ code:
+
+\[\stk{
+\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+    \DVar{}         & I \To \IDesc{I}                                   \\
+    \ldots          & \ldots
+\end{array} \\
+\\
+\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set        \\
+\begin{array}{@{}l@{\:=\:\:}ll}
+\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
+\ldots                        &  \ldots
+\end{array}
+}\]
+
+\begin{wstructure}
 <- Also replacing '1 by 'const  [figure]
@@
             /> 'const more useful in practice
-\end{structure}
+\end{wstructure}
+
+The resulting code of $\IDesc$ is presented in
+Figure~\ref{fig:idesc}. The reader will notice that we have also
+replaced $\DUnit$ by a more general $\DConst$ code. Whereas $\DUnit$
+was interpreted as the unit set, $\DConst{X}$ is interpreted as $X$,
+for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
+$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
+succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
+its equivalent encoding. From a definitional perspective, we are
+giving more opportunities to the type-system, hence reducing the
+burden on the programmer in practice.
+
+\begin{figure*}
 
 \[
+\begin{array}{ll}
 \stk{
@@
 }
-\]
-
-\[\stk{
-\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To \Set \To \Set                  \\
+&
+\stk{
+\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set                  \\
 \begin{array}{@{}l@{\:=\:\:}ll}
@@
 \end{array}
-}\]
+}
+\end{array}
+\]
+
+\caption{Universe of indexed description}
+\label{fig:idesc}
+
+\end{figure*}
 
 \subsection{Examples}
+\label{sec:idesc-examples}
 
-\begin{structure}
+\begin{wstructure}
 <- Nat
@@
         -> This applies to all previous examples
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+In order to gain some intuition with $\IDesc$, let us re-implement the
+signature functor of natural numbers. We refer the reader to
+Section~\ref{sec:desc-examples} for its incarnation in the universe of
+descriptions. The code is the following:
+
+\[\stk{
+\NatD : \IDesc{\Unit} \\
+\NatD \mapsto \DSigma{(\EnumT [ \NatZero, \NatSuc{} ])}
+                     {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
+}\]
+
+Because $\Nat$ is just an inductive type, we make no use of the index:
+it is the trivial inhabitant of $\Unit$. Therefore, the recursive
+argument is materialized by $\DVar{\Void}$, where we were using $\DId$
+in the $\Desc$ presentation. This transformation applies to all
+inductive types: we have not lost in expressive power during the
+transition.
+
+\begin{wstructure}
 <- Levitation [figure]
@@
     -> Behind the scene, relies on the special purpose switchD
-\end{structure}
+\end{wstructure}
+
+A mandatory exercise consists in describing $\IDesc$ in itself. With
+the experience gained in Section~\ref{sec:desc-levitate}, this
+levitation does not pose any issue:
 
@@
 \IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
-\begin{array}{ll}
+\begin{array}{@{}ll}
 \IDescD~I \mapsto \DSigma{}{} & (\EnumT [ \DVar{},
@@
 
-\begin{structure}
+Note that this description also lives in $\IDesc{\Unit}$. Indeed,
+$\IDesc$ is merely an inductive type. As for the $\Desc$ levitation,
+the finite function space $[ \ldots ]$ elaborates into a special
+purpose $\F{switchID}$ operator.
+
+\begin{wstructure}
 <- Untyped lambda terms [figure]
@@
         /> Will be fixed later
-\end{structure}
+\end{wstructure}
+
+Our next example is, for once, building an inductive family. We
+present the implementation of untyped lambda terms, as a family
+indexed by natural numbers. The natural numbers play the role of De
+Bruijn indexes. The indexed description is the following:
 
@@
                                   \DConst{\Fin{n}} \\
-                                  \DProd{\DVar{n}}{\DVar{n}} \\
+                                  \DProd{\DVar{\NatSuc{n}}}{\DVar{n}} \\
                                   \DVar{(\NatSuc{n})}
@@
 
+A variable is any number between $0$ and $n$, hence forbidding the
+reference of out-of-scope binders. Making a lambda increases by one
+the possible index choice of the underlying term. A function
+application applies a term with $n+1$ binders, resulting in a term
+with $n$ binders.
+
+We have therefore given a representation of lambda terms as a
+data-type. Any inhabitant of this data-type is a valid lambda term,
+forbidding out-of-scope reference in variables. While we have the
+data-type, we can regret the absence of structure. As shown by
+Altenkirch and Reus~\cite{altenkirch:monadic-lambda}, the lambda terms
+actually enjoy a monadic structure. In this setting, the ``bind''
+corresponds to substitution. We shall remedy to this situation
+shortly.
+
 \subsection{Constrained constructors}
 
-\begin{structure}
+\begin{wstructure}
 !!! Need Help !!!
@@
     ???
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+However, before entering the realm of generic structures, let us
+further explore the universe of indexed descriptions. In the
+following, we ask $I$ to come equipped with a notion of definitional
+equality. Hence, we can legibly equates elements of $I$.
+
+\begin{wstructure}
 <- Fin [figure]
@@
     ???
-\end{structure}
+\end{wstructure}
+
+In our implementation of lambda terms, variables were elements of
+$\Fin{n}$, with $n$ the number of binders. Let us now implement
+$\Fin$. $\Fin$ is a typical example of an inductive family using
+\emph{constrained constructors}. The constructors $\FinZero$ and
+$\FinSuc$ are only defined for an index distinct from $\NatZero$:
+
+\[
+\Rule{\Bhab{n}{\Nat}}
+     {\Bhab{\FinZero}{\Fin{(\NatSuc{n})}}}
+\qquad
+\Rule{\Bhab{n}{\Nat} \quad 
+      \Bhab{i}{\Fin{n}}}
+     {\Bhab{\FinSuc{}}{\Fin{(\NatSuc{n})}}}
+\]
+
+In order to code this data-type, we have to appeal to equality:
 
@@
 
-\begin{structure}
+Intuitively, the constraint is captured as follow. First, we store an
+element $m$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
+stipulates that $m$ cannot be \emph{any} natural numbers: it must be
+``the index minus one''. This translates into the constraint $n
+\PropEq \NatSuc{m}$.
+
+\begin{wstructure}
 <- Vectors
@@
     What can we say here we haven't with Fin?
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+%%! I suspect we don't, space-wise
+
+\begin{wstructure}
 <- GADT-style data-type definition
@@
         <- Lose match between data-type definition and the elimination form
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+The $\Fin$ example is also a well-known example of Generalized
+Abstract Data-Type (GADT), in Haskell. It comes as no surprise that
+our universe of indexed description supports this restricted class of
+dependent data-types. As we have shown with $\Fin$, constrained
+constructors are straightforwardly translated into equality
+constraints.
+
+While we support the same grammar, indexed descriptions embraces a
+much wider class of data-types. Indeed, we pose no constraint apart
+from strict positivity. This suggests that, based on the familiar
+grammar of GADTs, we can simply lift off the restrictions and
+translates the user's input to a code in our universe of indexed
+description.
+
+However, one could, quite rightfully, be worried by the introduction
+of equations in the picture. In the context of programming, these
+equations will come back and haunt us, begging for proofs. Hence,
+whereas the user typed a constructor-constrained data-type, the
+elimination form will involve equations and their proofs. This
+introduces a loophole in our abstraction we should try to address.
+
+\begin{wstructure}
 !!! Need Help !!!
@@
     ??? More technical detail needed
-\end{structure}
+\end{wstructure}
+
+To get a clue of a solution, let us look back at $\Fin$. We note that
+the equation is introduced because we are \emph{storing} the index of
+the inductive family, through $m$. Hence the constraint. However,
+\emph{inductive families need not store their
+  indices}~\cite{brady:index-inductive-families}. Applying the
+\emph{forcing} optimization to our definition of $\Fin$ gives the
+following, equivalent definition:
 
@@
 
+Equations have simply disappeared! We should precise that forcing a
+description is not garanteed to remove all constraints. It is subject
+of future work to see if constraints can be entirely eradicated, or
+presented more conveniently to the user.
+
+Finally, it is worth mentioning that the forcing optimization is a
+\emph{source-to-source} transformation of the description.
+
+%%! Actual implementation as a generic function??!
+
 \subsection{Free IMonad}
 
-\begin{structure}
+\begin{wstructure}
 <- Variation on a theme: free imonad construction
@@
         <- \I -> IDesc I as describing an indexed endofunctor
-        <- Free indexed monad construction
+        <- Free monad construction
     -> Still a suitable, generic notion of substitution
@@
         <- show implementation?? (space! space!)
-\end{structure}
+\end{wstructure}
 
-\begin{structure}
+In this section, we develop some more structure for our indexed
+data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
+a free monad operation on simple descriptions. Building on this
+experience, we are going to present its equivalent in the indexed
+world. The processus is the same. Namely, given an indexed functor, we
+derive the indexed functor coding its free monad:
+
+\[\stk{
+\begin{array}{ll}
+\FreeIMonad{\_}{} : & \PITEL{\V{I}}{\Set}
+                      \PITEL{\V{X}}{\V{I} \To \Set} \\
+                    & \PITEL{\V{R}}{\V{I} \To \TagIDesc{I}} \To
+                      (\V{I} \To \TagIDesc{I})
+\end{array} \\
+\FreeIMonad{R}{I}{X} \mapsto
+    \LAM{i}
+    \pair{\ListCons{\DVar{}}{(\fst{(R\: i)})}}
+         {\pair{(\DConst{X\: i})}
+               {(\snd{(R\: i)})}{}}{}
+}\]
+
+Just as in the universe of descriptions, this construction comes with
+an obvious ``return'' and a substitution operation, the ``bind''. Its
+type signature is the following:
+
+\newcommand{\substI}{\F{substI}}
+
+\[
+\substI : \PITEL{\V{I}, \V{J}}{\Set}
+          \PITEL{\sigma}{\V{I} \To \IDesc{\V{J}}}
+          \PITEL{\V{D}}{\IDesc{I}} \To
+          \IDesc{J}
+\]
+
+\begin{wstructure}
 <- Untyped lambda terms
@@
         <- We should not pay for the structure which *is* here
-\end{structure}
+\end{wstructure}
+
+In Section~\ref{sec:idesc-examples}, we were missing the monadic
+structure of lambda terms. In particular, we wish we were given a
+substitution principle for free. Above, we have generically built such
+substitution operation for free monads. Can we be rewarded for our
+effort? Let us define the following $\LambdaTFreeD$ description,
+largely inspired by $\LambdaTD$:
 
@@
 
+In essence, we have forgotten variables, just keeping the nodes of
+lambda terms: abstraction and application. Our hope is that taking the
+free monad of this indexed functor gives us our lambda terms back:
+
+%%! FreeMonad takes a damned tagged description
+
 \[\stk{
 \LambdaTD : \Nat \To \IDesc{\Nat} \\
-\LambdaTD\: \mapsto \FreeMonad{[\LambdaTFreeD]}{\Fin{}}
+\LambdaTD\: \mapsto \FreeIMonad{[\LambdaTFreeD]}{\Nat}{\Fin{}}
 }\]
 
-\begin{structure}
+By a straightforward unfolding of definitions, the reader will
+convince herself that we recover the initial definition. However, this
+time, we have a substitution operation. In summary, we have done less
+work in defining $\LambdaTFreeD$ but we have pointed out its
+structure, so, rightfully, we have been rewarded in our effort. The
+lesson is that, when structure there is, one should be allowed to get
+it for free.
+
+\begin{wstructure}
 <- IDesc
@@
         ???
-\end{structure}
+\end{wstructure}
+
+Another instance of free monad is $\IDesc$ itself. Indeed, $\DVar$ is
+nothing but the ``return''. The remaining constructors are simply the
+freed functor, trivially indexed by $\Unit$. Hence, amusingly, we can
+define the $\IDesc$ freed functor by the following code:
 
@@
 
+Then, we get $\IDesc$ by building its free monad:
+
 \[\stk{
@@
 }\]
+
+%%! Should we stop here? Or show off some madness?

### 2010-03-11 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: transform my TeX comments into readable (haha) notes.

@@
 \end{wstructure}
+
+\note{ Need care: motivating motivation of indexing. }
 
@@
 
+\note{ I will need some tutorial on this ad-hoc notion of definitional
+  equality. Or what Conor's meant with that. }
+
 \begin{wstructure}
@@
 
+\note{ Anything else? }
+
 \begin{wstructure}
@@
 
-%%! I suspect we don't, space-wise
+\note{ I suspect we don't implement vectors, space-wise }
 
@@
 
+\note{Is that convincing? Worth saying? Anything else?}
+
 \begin{wstructure}
@@
 following, equivalent definition:
+
+\note{ I'm not familiar with this, I need to get into the paper. }
 
@@
 
-%%! Actual implementation as a generic function??!
+\note{ Actual implementation of forcing as a generic function??! }
 
@@
 
-%%! FreeMonad takes a damned tagged description
+\note{ FreeMonad takes a damned tagged description }
 
@@
 
-%%! Should we stop here? Or show off some madness?
+\note{ Should we stop here? Or show off some madness? }

### 2010-03-11 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: fix \cite{} which was preventing compilation on ancient version of pdflatex.
Desc@ICFP: transform my TeX comments into readable (haha) notes.
Desc@ICFP: more philosophical questions

@@
 
-%%! Check Martin-Lof reference
+\note{Check Martin-Lof reference}
 
@@
 theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
-programming technique~\cite{}. We refer the reader to Agda's
-tutorial~\cite{norell:agda-tutorial} for a pedagogical presentation of
-universe construction.
+programming technique~\cite{who?} \note{I would need some other
+  example of universe construction for practical purposes}. We refer
+the reader to Agda's tutorial~\cite{norell:agda-tutorial} for a
+pedagogical presentation of universe construction.
 
@@
 universes.
+
+\note{Is that really a motivating motivation for the usage of universe
+  construction? }
 
@@
 
+\note{We have talked about a tagged description based on $\List (\UId
+  \times \Desc)$. It makes the free monad easier to construct but,
+  imho, considerably clutters anything else. I've reverted back to a
+  more sigma-friendly presentation.}
+
 \[
@@
 constructor.
+
+\note{I think we need a function mapping a $\TagDesc$ to the
+  corresponding $\Desc$. This function should go here.}
 
@@
 
-%%! Strictly positive types anyone?
+\note{ Strictly positive types anyone? }
 
@@
 property: they all admit a fix-point.
+
+\note{ I'm afraid that this section is killed by my categorical gibberish. }
 
@@
 
+\note{Shall we give the code of $\All$?}
+
 \begin{wstructure}
@@
 
-%% I'm kind of shafted (haha) by this:
-%%     - We have not yet introduced tagged descriptions
-%%     - Tagged description in a Mu, that just doesn't work...
+\note{ This is rather drafty }
 \[

### 2010-03-11 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: finish IDesc section.
Desc@ICFP: transform my TeX comments into readable (haha) notes.
Desc@ICFP: more philosophical questions

@@
 
-%%! Maybe a bit of perspective would be good
-%%    <- Everything stands up with the Mu
-%%    <- Induction principle defined over Mu-things
+\note{
+Maybe a bit of perspective would be good: 
+Everything stands up with the Mu.
+Induction principle defined over Mu-things.
+}
 
@@
 Consequently, the examples developped above can be straightforwardly
-translated into this new presentation. In the following, we adopt this
-last version as our de facto universe of inductive types. In
-particular, we are going to demonstrate several instances of generic
-programming.
+translated into this new presentation. \note{We might want to re-do
+  the Nat example here, as we encode once again in $\IDesc$} In the
+following, we adopt this last version as our de facto universe of
+inductive types. In particular, we are going to demonstrate several
+instances of generic programming.
 
-%%! Shall we talk about the Type Theory being Desc Zero? or such story?
+\note{Shall we talk about the Type Theory being Desc Zero? or such story?}
 
@@
 
-%%! Update to the factorized presentation
+\note{Update to the factorized presentation}
 
@@
 \subsection{The generic Free Monad}
+\label{sec:desc-free-monad}
 
@@
 
-%%! Recall monadic structure of this object
+\note{Recall monadic structure of this object}
 
@@
 
-%%! This is not quite Mu of a tagged description
+\note{This is not quite Mu of a tagged description}
 
@@
 
-%%! Triple check. That doesn't work, right?
+\note{ Triple check. That doesn't work, right? }
 

### 2010-03-11 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: transform my TeX comments into readable (haha) notes.
Desc@ICFP: more philosophical questions

@@
 
+\note{Need some care here. Expansion would be good too. I wanted to
+  carry the intuition that we are not the bad guys with a
+  non-terminating type-checker.}
+
 \subsection{Finite sets}
@@
 
+\note{Shall we talk about W-types encoding?}
+
 However, using encodings is impractical. In the case of finite sets,
@@
 were the democracy analogy goes crazy, I think).
+
+\note{Did I got the coding presentation vs. coding representation
+  story right?}
 
@@
 
-%%! This might deserve a more thorough explanation, with snippet of code.
+\note{ This might deserve a more thorough explanation, with snippet of
+       code. }
 
@@
 expressions.
+
+\note{And suddenly, type-directed pretty printing comes into the
+  picture. Surely, it should ``just'' be the inverse of
+  elaboration. But is it? And how do we present it?}
 
@@
 on bidirectional type-checking. Using bidirectional type-checking as a
-boilerplate scrapper is a well-known technique~\cite{who?}. In our case,
-we have shown how to instrument bidirectionality to rationalize the
-expressivity of dependent types. We have illustrated our approach with
-finite sets. We have successfully abstracted away the low-level
-presentation of finite sets, offering a convenient syntax instead.
+boilerplate scrapper is a well-known technique~\cite{who?} \note{I
+  would need some citation of people using bidirectional presentation
+  of their type theory}. In our case, we have shown how to instrument
+bidirectionality to rationalize the expressivity of dependent
+types. We have illustrated our approach with finite sets. We have
+successfully abstracted away the low-level presentation of finite
+sets, offering a convenient syntax instead.
 

### 2010-03-12 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: ispell. Gosh.

@@
 <- The duality of Sigma
-    <- Sigma generalizes sum over arbitrary arities
+    <- Sigma generalises sum over arbitrary arities
         -> \Sigma A B == \Sigma_{x : A} B x
-    <- Sigma generalizes product to have a dependant second component
+    <- Sigma generalises product to have a dependant second component
         -> \Sigma A B == (x : A) \times (B x)
@@
 $\Sigma_{x : A} (B x)$. This notation stresses the fact that
-$\Sigma$-types are a generalization of sums over arbitrary
+$\Sigma$-types are a generalisation of sums over arbitrary
 arities. When simply-typed languages have finite sums,
@@
 $\SIGMA{\V{x}}{A} (B x)$. Under this perspective, $\Sigma$-types can
-be read as a generalization of products, where the second component
+be read as a generalisation of products, where the second component
 can depend on the first one. When simply-typed languages pack data
@@
 In the simply-typed world, the essence of data-type definition can be
-sumed up by the term ``sum-of-product''. A data-type is defined by a
+summed up by the term ``sum-of-product''. A data-type is defined by a
 finite sum, choice, of constructors, each of them composed by a
@@
 language. However, as such, a system of code is useless as it lacks a
-semantics. Instead of equiping the universe of codes with some
-computational behavior, we pragmatically chose to \emph{interpret}
+semantics. Instead of equipping the universe of codes with some
+computational behaviour, we pragmatically chose to \emph{interpret}
 these codes back into the standard type theory. Hence, codes act only
@@
 Finally, we are not limited to one recursive argument. This is
-demonstated by our description of a binary tree functor below:
+demonstrated by our description of a binary tree functor below:
 
@@
         -> No loss of expressive power
-        /> Garantee a ``constructor form''
+        /> Guarantee a ``constructor form''
 \end{wstructure}
@@
 
-So far, for conveniency, we have taken as granted the existence of
+So far, for convenience, we have taken as granted the existence of
 $\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
@@
 satisfying the typing rules. It will be the subject of
-Section~\ref{sec:desc-levitate} to fulfill this promise, by actually
+Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
 implementing the specification.
@@
     -> Extend the type theory with Mu/Con [figure]
-        <- Straightforward definition of a fixpoint
-            <- Interpret D with (Mu D) as subobjects
+        <- Straightforward definition of a fix-point
+            <- Interpret D with (Mu D) as sub-objects
 \end{wstructure}
@@
 describing a signature functor, we define its fix-point by the type
-former $\Mu{D}$. Inhabitants of the fix-point are materialized by the
+former $\Mu{D}$. Inhabitants of the fix-point are materialised by the
 structure map: $\Con{x}$ inhabits $\Mu{D}$ if $x$ belongs to the
@@
 \To T$ by a similar process. We start with turning the arrow into a
-dependenty arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
+dependently arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
 further analysis.
@@
 principle by taking an argument $x$ in $\Mu{D}$ and claiming that the
-motive $P$ is realized on $x$.
+motive $P$ is realised on $x$.
 
@@
 description instead of a sugared data-type definition. Then, to
-recognize an element of $\Nat$ for instance, she would have to
+recognise an element of $\Nat$ for instance, she would have to
 identify $\Mu{}$ of a description as indeed the fix-point of the
@@
 suffer from too much type information. Just as for finite, the cure
-lies in rationalizing this information. Let us first tackle the issue
+lies in rationalising this information. Let us first tackle the issue
 of data-type declaration: unsurprisingly, the grammar of data-types we
-are used can be straighforwardly translated to the
+are used can be straightforwardly translated to the
 ``sigmas-of-sigmas'' paradigm. Hence, a data-type definition is turned
@@
 
-\note{ This is rather drafty }
+\note{ This is rather drafty. I should explain what $\vec{e}$ and $\widetilde{e'}$ are. }
+
 \[
@@
 $\Mu{}$ type of a tagged description. It is then straightforward to
-lookup the tag and turn the constructor form into the appropriate
+look-up the tag and turn the constructor form into the appropriate
 code.

### 2010-03-12 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: ispell. Gosh.

@@
 
-In this section, we will fulfill several promises. One promise we made
+In this section, we will fulfil several promises. One promise we made
 was to effectively implement the code of the $\Desc$ universe. Another
@@
 The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
-coincidence. $\EnumU$ is indeed simply isomorph to a list, with
+coincidence. $\EnumU$ is indeed simply isomorphic to a list, with
 $\UId$s as parameter. However, in Section~\ref{sec:desc-examples}, we
@@
 names. The type naming scheme of the type theory has condensed: named
-type formes ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are now
+type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are now
 replaced by codes and their fix-point in the universe of
@@
 \begin{wstructure}
-<- Realizing our promises
+<- Realising our promises
     <- We are going to implement Desc
@@
 
-We shall now fulfill our promises. We are going to implement the codes
+We shall now fulfil our promises. We are going to implement the codes
 of the universe of descriptions. Interestingly, in and by itself, a
@@
 Let us explain how we have proceeded and the obstacle we face. Much as
-the data-types we have seen so far, $\DescD$ first consists in chosing
+the data-types we have seen so far, $\DescD$ first consists in choosing
 a constructor. Constructors correspond to the code's cases: $\DUnit$,
@@
 
-Note that $\DIndx$ is isomorph to $\DHindx{\Unit}$. However, we
+Note that $\DIndx$ is isomorphic to $\DHindx{\Unit}$. However, we
 tolerate this duplication and do not replace $\DIndx$ by the more
@@
 the higher-order induction on $S$. The $\DHindx$ case consists in
-packing the $H$ in $\Set$ with a recursive arguement.
+packing the $H$ in $\Set$ with a recursive argument.
 
@@
 But this is rather problematic! We are still in the process of
-constructing $\DescD$, and the motive of $\F{switch}$ is abruptely
+constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
 begging for an $\eta$-expanded $\Desc$. Despite our willingness, we
-will never be able to materialize such motive.
+will never be able to materialise such motive.
 
@@
 results in a new definition of the $\Desc$ universe. This
-reorganization is strictly equivalent to the previous one
+reorganisation is strictly equivalent to the previous one
 (Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
@@
 
-The factorized universe of description is presented in
+The factorised universe of description is presented in
 Figure~\ref{fig:factor-desc}. The modifications are justified as
@@
 
-Consequently, the examples developped above can be straightforwardly
+Consequently, the examples developed above can be straightforwardly
 translated into this new presentation. \note{We might want to re-do
@@
 
-\caption{Factorized universe of descriptions}
+\caption{Factorised universe of descriptions}
 \label{fig:factor-desc}
@@
 
-The catamorphism can be characterized as an induction on the
-description $D$, with a non-dependent motive targetting $T$. Given a
+The catamorphism can be characterised as an induction on the
+description $D$, with a non-dependent motive targeting $T$. Given a
 node $xs$ and the induction hypotheses, the method ought to build an
@@
 
-\note{Update to the factorized presentation}
+\note{Update to the factorised presentation}
 
@@
 
-Of course, we must equip the resulting datatypes with operations
+Of course, we must equip the resulting data-types with operations
 delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
@@
 
-Our implementation will appeal to the $\cata$ function developped
+Our implementation will appeal to the $\cata$ function developed
 previously. So, let us write down the types, and fill as much
@@
 
-Filling the $???$ subgoal with $\apply\: D\: X\: Y\: \sigma$ finishes
+Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$ finishes
 the implementation. To sum up, we have implemented the free monad

### 2010-03-12 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: more philosophical questions
Desc@ICFP: fix functor for lambda terms
Desc@ICFP: ispell. Gosh.

@@
 
-This generalization can be achieved by a minor evolution of $\Desc$
+This generalisation can be achieved by a minor evolution of $\Desc$
 (Fig.~\ref{fig:factor-desc}). Indeed, the functorial nature of $\Desc$
@@
 
+\note{Should we say something about IMu, fixed-points, and their
+  elimination before moving on? If so, where should we put it?}
+
 \subsection{Examples}
@@
 it is the trivial inhabitant of $\Unit$. Therefore, the recursive
-argument is materialized by $\DVar{\Void}$, where we were using $\DId$
+argument is materialised by $\DVar{\Void}$, where we were using $\DId$
 in the $\Desc$ presentation. This transformation applies to all
@@
 
-The $\Fin$ example is also a well-known example of Generalized
+The $\Fin$ example is also a well-known example of Generalised
 Abstract Data-Type (GADT), in Haskell. It comes as no surprise that
@@
 !!! Need Help !!!
-<- Brady optimization: forcing
+<- Brady optimisation: forcing
     <- Source to source translation
@@
   indices}~\cite{brady:index-inductive-families}. Applying the
-\emph{forcing} optimization to our definition of $\Fin$ gives the
+\emph{forcing} optimisation to our definition of $\Fin$ gives the
 following, equivalent definition:
@@
 Equations have simply disappeared! We should precise that forcing a
-description is not garanteed to remove all constraints. It is subject
+description is not guaranteed to remove all constraints. It is subject
 of future work to see if constraints can be entirely eradicated, or
@@
 
-Finally, it is worth mentioning that the forcing optimization is a
+Finally, it is worth mentioning that the forcing optimisation is a
 \emph{source-to-source} transformation of the description.
@@
     <- Recall existence of generic free monad construction
-    -> Present its generalization to IDesc [equation]
+    -> Present its generalisation to IDesc [equation]
         <- \I -> IDesc I as describing an indexed endofunctor
@@
                                 & \left[\begin{array}{l}
-                                  \DProd{\DVar{n}}{\DVar{n}} \\
+                                  \DProd{\DVar{\NatSuc{n}}}{\DVar{n}} \\
                                   \DVar{(\NatSuc{n})}

### 2010-03-12 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: ispell. Gosh.

@@
 Our work is deeply rooted in the realm of dependently-typed functional
-languages. In particular, we have developped this technology in our
+languages. In particular, we have developed this technology in our
 on-going work on Epigram~\cite{pigs:epigram}. However, in this paper,
@@
 gentle introduction to dependently-typed systems. To the reader
-familiar with type theory, she will recognize the basis of her
-favorite type theory, be it Agda or Coq for instance. This scene sets,
+familiar with type theory, she will recognise the basis of her
+favourite type theory, be it Agda or Coq for instance. This scene sets,
 we have a common ground for discussion.
@@
 judgements, typing judgements, and equality judgements. Formally,
-these judgments are presented as follow:
+these judgements are presented as follow:
 
@@
     <- Judgemental equality
-        <- Presentation independant of particular implementation choice
-        -> Model in Agda, intensional
+        <- Presentation independent of particular implementation choice
+        -> Model in Agda, intentional
         -> Used in Epigram, OTT
@@
 equality. Therefore, we are not tied to a particular implementation
-choice. In particular, our system has been modeled in Agda, which
-features an intensional equality. On the other hand, it is used in
+choice. In particular, our system has been modelled in Agda, which
+features an in-tensional equality. On the other hand, it is used in
 Epigram, which equality has a slightly extensional
-flavor~\cite{altenkirch:ott}. We expect users of propositional or even
-fully extensional systems to also find their way through this
-presentation.
+flavor~\cite{altenkirch:ott}. We expect users of fully extensional
+systems to also find their way through this presentation.
 
@@
 The rules ensuring Context Validity (Fig.\ref{fig:context-validity})
-are unsurprisingly common. They simply garantee the validity of the
+are unsurprisingly common. They simply guarantee the validity of the
 empty context, as well as the linear extension of a context by a
@@
 The typing judgements (Fig.\ref{fig:typing-judgements}) are just as
-familiar. We recognize the standard presentation of $\Pi$ and $\Sigma$
+familiar. We recognise the standard presentation of $\Pi$ and $\Sigma$
 types, respectively inhabited by lambda terms and dependent
@@
 Figure~\ref{fig:judgemental-equality}. As one would expect,
-judgemental equality is respects symmety, reflexivity, and
-transitivity. Further, it also encompasses the $β$-rules for function
-application and pairs. Finally, equality under lambda is dealt by the
-$ξ$-rule.
+judgemental equality respects symmetry, reflexivity, and
+transitivity. Further, it also encompasses the $\beta$-rules for
+function application and pairs. Finally, equality under lambda is
+dealt with by the $\xi$-rule.
 
@@
 type-theory, some opportunities arise. We will present some of them
-along the course of the paper, clearly specifying which extensions we
-require.
+along the course of the paper.
 
@@
        there is no non-terminating type-checker, or such horrible lies <>
-    -> Strongly normalizing
+    -> Strongly normalising
         -> Every program terminates
@@
 This completes our presentation of the type theory. Assuming a
-stratified discpline of universe, the system we have described enjoy
+stratified discipline of universe, the system we have described enjoy
 some very strong meta-theoretical properties. Unlike simply typed
 languages, such as Haskell, dependently-typed systems are
-\emph{strongly normalizing}: every program that type-checks
+\emph{strongly normalising}: every program that type-checks
 terminates. Moreover, type-checking is decidable and can therefore be
@@
 Again, there is a clear equivalent in the full-$\Set$ world: function
-application. The operational behavior of $\F{switch}$ is
+application. The operational behaviour of $\F{switch}$ is
 straightforward: $x$ is peeled off as we move deeper inside the nested
@@
 extract that intuition from the type, by a careful reading. At a first
-glance, this seems to contradict our argument in favor of types for
+glance, this seems to contradict our argument in favour of types for
 coding representation. Here, we are overflown by low-level, very
@@
 \begin{wstructure}
-<- Formalization: type propagation
+<- Formalisation: type propagation
     <- Motivation
@@
         -> Presented as judgements
-    -> Presentation mirors typing rule of [figure] 
-        -> Ignore identitical judgements
+    -> Presentation mirrors typing rule of [figure] 
+        -> Ignore identical judgements
 \end{wstructure}
 
-We adapt and formalize this \emph{type propagation} mechanism to our
+We adapt and formalise this \emph{type propagation} mechanism to our
 richer type theory. First of all, we define a high-level syntax,
@@
 $\exprEx$ are subject to type \emph{synthesis}.
+
+\note{Why don't we need a context validity judgement, again?}
 
@@
 instead. Hence, we use the syntax $\Enum\: [t_1 \ldots t_n]$ to denote
-the finite sets composed by the nammed elements $t_1$, \ldots,
+the finite sets composed by the named elements $t_1$, \ldots,
 $t_n$. The translation to elements of $\EnumU$ is straightforward.
@@
 set, we instead elaborate it as a term in a small $\pi$-type. That is,
-a finite function. To realize our promise of a big $\Pi$-type, we
+a finite function. To realise our promise of a big $\Pi$-type, we
 elaborate a call to the $\F{switch}$ eliminator.
@@
         <- Example: Finite sets/finite function space
-        -> We should not restrain ourself in using codes
+        -> We should not restrain our self in using codes
             <- We know how to present them to the user
@@
 
-In this section, we have developped a type propagation system, based
+In this section, we have developed a type propagation system, based
 on bidirectional type-checking. Using bidirectional type-checking as a
@@
   of their type theory}. In our case, we have shown how to instrument
-bidirectionality to rationalize the expressivity of dependent
+bi-directionality to rationalise the expressivity of dependent
 types. We have illustrated our approach with finite sets. We have
@@
 This example teaches us that we should not be afraid by codes, as far
-as type information is available. We have shown how to rationalize
+as type information is available. We have shown how to rationalise
 this information in a formal presentation. Hence, we have proved that
@@
 in our theory, we will show how to extend the framework we have
-developped so far.
+developed so far.

### 2010-03-13 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: merge Conor's comments in. (first pass)

@@
 
-Our proposal is based on induction-recursion~\cite{dybjer:general-ir,
-  dybjer:axiom-ir, dybjer:ir-initial-algebra, dybjer:iir}. However, in
-the context of simple data-types, the full power of
-induction-recursion is not necessary. Therefore, we will present a
-stripped-down version. As we extend the scope of our universe of
-data-types, our presentation will diverge from induction-recursion. We
-shall compare both approaches in Section~\ref{sec:discussion}
+Our proposal is inspired by, but less involved than, Dybjer and
+Setzer's closed formulation of
+induction-recursion~\cite{dybjer:general-ir, dybjer:axiom-ir,
+  dybjer:ir-initial-algebra, dybjer:iir}. In the context of simple
+data-types, the full power of induction-recursion is not
+necessary. Therefore, we adopt a stripped-down version of it. As we
+extend the scope of our universe of data-types, our presentation will
+diverge from induction-recursion. We shall compare both approaches in
+Section~\ref{sec:discussion}
 
@@
 of the author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
-$\Sigma_{x : A} (B x)$. This notation stresses the fact that
+$\Sigma_{x : A} (B\: x)$. This notation stresses the fact that
 $\Sigma$-types are a generalisation of sums over arbitrary
 arities. When simply-typed languages have finite sums,
-dependently-typed languages have sums of indexed by any set.
+dependently-typed languages have sums indexed by any set.
 
 On the other hand, $\SigmaTy{A}{B}$ is sometimes written
-$\SIGMA{\V{x}}{A} (B x)$. Under this perspective, $\Sigma$-types can
+$\SIGMA{\V{x}}{A} (B\: x)$. Under this perspective, $\Sigma$-types can
 be read as a generalisation of products, where the second component
@@
 summed up by the term ``sum-of-product''. A data-type is defined by a
-finite sum, choice, of constructors, each of them composed by a
-product, a tuple, of arguments. Therefore, the grammar of data-types
-definition in simply-typed languages, such as Haskell or OCaml, follow
-this form.
+finite sum -- a choice -- of constructors, each of them is composed by
+a product -- a tuple -- of arguments. Therefore, the grammar of
+data-types definition in simply-typed languages, such as Haskell or
+OCaml, follow this form.
 
@@
 theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
-programming technique~\cite{who?} \note{I would need some other
-  example of universe construction for practical purposes}. We refer
-the reader to Agda's tutorial~\cite{norell:agda-tutorial} for a
-pedagogical presentation of universe construction.
+programming technique~\cite{benke:universe-generic-prog, whoelse?}
+\note{I would need some other example of universe construction for
+  practical purposes}. We refer the reader to Agda's
+tutorial~\cite{norell:agda-tutorial} for a pedagogical presentation of
+universe construction.
 
@@
 $\descop{D}{}$ itself, for a given data-type definition $D$. This
-corresponds to the morphism part of the functor described by $D$.
+corresponds to the object part of the functor described by $D$.
 
@@
 build these descriptions. However, it should be clear that, in
-practice, these definitions can be automatically constructed an
+practice, these definitions can be automatically constructed from an
 Haskell-like $\data$ definition.
@@
 
-Our first example is the natural numbers, or rather its carrier
+Our first example is the natural numbers, or rather its signature
 functor. Our code is presented in the high-level expression language
@@
 
-\[    \NatD\: Z = 1 + Z    \]
+\[    \NatD\: Z \mapsto 1 + Z    \]
 
@@
 With a small change to the definition of $\NatD$, we obtain the
-carrier of lists:
+signature of lists:
 
@@
 
-The modification consists in turning the $\NatSuc{}$ into a proper
-$\ListCons$ taking an argument in X followed by an inductive
+The modification consists in turning the $\NatSuc{}$ constructor into
+a proper $\ListCons$ taking an argument in X followed by a recursive
 argument. In this case, we use $\DSigma{}{}$ in its product
@@
 
-\[    \ListD\: X Z = 1 + X \times Z     \]
+\[    \ListD\: X\: Z \mapsto 1 + X \times Z     \]
 
@@
 Finally, we are not limited to one recursive argument. This is
-demonstrated by our description of a binary tree functor below:
+demonstrated by our description of binary trees:
 
@@
 Again, we are at one evolutionary step away from $\ListD$. However,
-instead of single appeal to the induction code, we call it twice. The
+instead of a single call to the induction code, we call it twice. The
 interpretation of this code corresponds to the following functor:
 
-\[    \TreeD\: X Z = 1 + X \times Z \times Z     \]
+\[    \TreeD\: X\: Z \mapsto 1 + X \times Z \times Z     \]
 
@@
 From the examples above, we observe that data-types are defined by a
-$\DSigma$ which first argument is a finite set of constructor. The
+$\DSigma$ which first argument is a finite set of constructors. The
 descriptions fitting into this pattern are called \emph{tagged}
@@
 \[
- \TagDesc \mapsto \SIGMA{\V{E}}{\EnumU} \spi{\V{E}}{\LAM{\_} \Desc}
+ \TagDesc \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{(\LAM{\_} \Desc)})
 \]
@@
 we have a finite sum of constructors. Of course, every description can
-be expressed into this style, by using a singleton as single
-constructor.
+be expressed into this style, using a singleton as single constructor
+if necessary.
 
@@
 and $\DUnit$ as type formers, extending the basic type theory. Hence,
-in an implementation of this type theory, we would have extended it
-with these constructors and their typing rules. Although it makes no
+in an implementation of this type theory, we need to extended it with
+these constructors and their typing rules. Although it makes no
 conceptual difference, the code of description should rather be read
@@
 \To T$ by a similar process. We start with turning the arrow into a
-dependently arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
+dependent arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
 further analysis.
@@
 Then, we have to translate the notion of algebra in our dependent
-setting. This is achieved by the $\All{D}{X}{P}{x}$ predicate that,
+setting. This is achieved by the predicate $\All{D}{X}{P}{x}$ that,
 intuitively, states that $\Bhab{P}{X \To \Set}$ holds everywhere in
-the sub-structures of $x$. Based on that knowledge, we ought to be
-able to prove that $P$ holds for $\Con{xs}$. We close our elimination
-principle by taking an argument $x$ in $\Mu{D}$ and claiming that the
-motive $P$ is realised on $x$.
+$x$. Based on that knowledge, we ought to be able to prove that $P$
+holds for $\Con{xs}$. We close our elimination principle by taking an
+argument $x$ in $\Mu{D}$ and claiming that the motive $P$ is realised
+on $x$.
 
@@
 
-At this stage, we are still left with implementing $\F{All}$ and the
+At this stage, we are still left with implementing this
 $\F{induction}$ principle. For space reason, we refer the reader to
-our technical report for the actual code. Having this dependent
-catamorphism, we are empowered with the ability to $\emph{compute}$
-over descriptions. Note that this induction principle is the first
-manifestation of a generic operation over descriptions. Thanks to
-$\F{induction}$, any data-type we define automatically comes equipped
-with an induction principle.
+our technical report for the actual code, hence the proof of (weak)
+initiality. Having this dependent catamorphism, we are empowered with
+the ability to $\emph{compute}$ over descriptions. Note that this
+induction principle is the first manifestation of a generic operation
+over descriptions. Thanks to $\F{induction}$, any data-type we define
+automatically comes equipped with an induction principle.
 
@@
 As for finite sets previously, these symptoms clearly indicate that we
-suffer from too much type information. Just as for finite, the cure
-lies in rationalising this information. Let us first tackle the issue
-of data-type declaration: unsurprisingly, the grammar of data-types we
-are used can be straightforwardly translated to the
-``sigmas-of-sigmas'' paradigm. Hence, a data-type definition is turned
-into the corresponding description. Then, at the level of definitions,
-we associate the name of the data-type to the fix-point of the
+suffer from too much type information. Just as for finite sets, the
+cure lies in rationalising this information. Let us first tackle the
+issue of data-type declaration. The grammar of data-types we are used
+to can be straightforwardly translated to the ``sigmas-of-sigmas''
+paradigm. Hence, a data-type definition can be turned into the
+corresponding description. Then, at the level of definitions, we
+associate the name of the data-type to the fix-point of the
 description. The induction principle as well as other generic
@@
 the second issue. Namely, our user would rather be presented a name
-rather that a fix-point of a signature functor. Rather surprisingly,
-it is sufficient to \emph{label} the $\Mu{}$ constructor with the
+rather than a fix-point of a signature functor. Surprisingly, it is
+sufficient to \emph{label} the $\Mu{}$ constructor with the
 user-provided name. When pretty-printing this object, we just expose

### 2010-03-13 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: merge Conor's comments in. (first pass)

@@
 principle. From an implementation perspective, this means that we
-remove an operator, hard-coded in the host language. This operator is
+remove an operator hard-coded in the host language. This operator is
 simply implemented in the dependently-typed target language, as any
@@
 names. The type naming scheme of the type theory has condensed: named
-type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are now
-replaced by codes and their fix-point in the universe of
-descriptions. Our next step is similar, in essence: we are going to
+type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
+now replaced by codes and their fix-point in the universe of
+descriptions. In essence, our next step is similar: we are going to
 condense the entire naming scheme of data-types \emph{in itself}.
@@
 Let us explain how we have proceeded and the obstacle we face. Much as
-the data-types we have seen so far, $\DescD$ first consists in choosing
-a constructor. Constructors correspond to the code's cases: $\DUnit$,
+the data-types we have seen so far, we first have to choose a
+constructor. Constructors correspond to the code's cases: $\DUnit$,
 $\DSigma$, and $\DIndx$. The reader will notice that the tagged
 notation we have used so far for $\Desc$ constructors now fully make
-sense: these were actually the tags we are defining.
+sense: these were actually the tags we are defining. \note{This
+  sentence badly needs a rephrasing.}
 
@@
 tolerate this duplication and do not replace $\DIndx$ by the more
-general $\DHindx$. Indeed, unlike its counterpart, $DIndx$ is
+general $\DHindx$. Indeed, unlike its counterpart, $\DIndx$ is
 first-order. Therefore, while both codes are \emph{extensionally} the
-same, in most practical implementation, they would be dealt with
-rather differently. While we can definitional equality to cope with
-first order objects, the functional presentation introduced by
-$\DHindx$ is unlikely to be amenable to a purely definitional
-treatment.
+same, in most practical implementation they would be dealt with rather
+differently. While definitional equality can cope with first order
+objects, the functional presentation introduced by $\DHindx$ is
+unlikely to be amenable to a purely definitional treatment.
 
@@
 The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
-achieve the description of the $\DSigma$ case, by a simple appeal to
-the higher-order induction on $S$. The $\DHindx$ case consists in
-packing the $H$ in $\Set$ with a recursive argument.
+successfully describe the $\DSigma$ case, by a simple appeal to the
+higher-order induction on $S$. The $\DHindx$ case consists in packing
+the $H$ in $\Set$ with a recursive argument.
 
@@
 visible in the definition of $\DescD$
-(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by a
+(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by
 careful usage of type propagation. Indeed, let us elaborate this term
@@
 brackets $[ \ldots ]$. This term $f$ is pushed into the type
-$\EnumT{e} \To \Desc$. This corresponds to a finite function set we
-have implemented in Figure~\ref{fig:type-checking}.
+$\EnumT{e} \To \Desc$. This corresponds to a finite function
+definition we have implemented in Figure~\ref{fig:type-checking}.
 
@@
 \[
-\PLAM{x}{(\EnumT{e})} \switch{e}{\PLAM{\_} \Desc}{\pi^f}{x}
+\PLAM{x}{(\EnumT{e})} \switch{e}{\LAM{\_} \Mu{\DescD}}{\pi^f}{x}
 \]
@@
 If we cannot make it happen, we simply do not make it happen. We
-perfectly know what the motive ought to be. So, straightforwardly, we
-extend the type theory with two special-purpose operators:
+perfectly know what the motive ought to be. Consequently, we extend
+the type theory with a special-purpose operator:
 
@@
 \begin{array}{@{}ll}
-%% piD
-\green{\pi_D} : & \PITEL{\V{e}}{\EnumU} \To \Set \\
 %% switchD
 \F{switchD} : & \PITEL{\V{e}}{\EnumU}               
-                \PITEL{\V{b}}{\spiD{e}}
+                \PITEL{\V{b}}{\spi{e}{\LAM{\_} \Desc}}
                 \PITEL{\V{x}}{\EnumT{e}} \To \Desc
@@
 
-The entire work of the magician stands here, in these two
-extensions. One could be worried by the availability of such
-operations in our type theory. Indeed, it would be inconvenient for
-the user to have to decide between these two variants. In fact, we
-give no choice to the user, hence no inconvenience. Just as the
-$\F{switch}$ operator was sealed off by the type propagation
-machinery, $\F{switchD}$ is not provided to the user. We simply extend
-the type-checking rules (Fig.~\ref{fig:type-checking}) with the
-following inference rule:
+The entire work of the magician stands here, in this extension. One
+could be worried by the availability of such operations in our type
+theory. Indeed, it would be inconvenient for the user to have to
+decide between two variants. In fact, we give no choice to the user,
+hence no inconvenience. Just as the $\F{switch}$ operator was sealed
+off by the type propagation machinery, $\F{switchD}$ is not provided
+to the user. We simply extend the type-checking rules
+(Fig.~\ref{fig:type-checking}) with the following inference rule:
 
@@
 This concludes our levitation work. Beyond the beauty of the exercise,
-there is also several practical outcomes. First of all, this exercise
+there are several practical outcomes. First of all, this exercise
 reveals that the $\Desc$ universe is just plain data. Just as any
@@
 (Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
-self-descriptive. We refer the reader to our companion technical
+self-descriptive. We refer the reader to the companion technical
 report for details.
@@
     <- Inspecting data-types
-        <- All described by a Desc code
+        <- All described byu a Desc code
         -> Ability to explore the code
@@
 operation. Having the ability to program on $\Desc$ is the key to
-define such function. The induction principle of $\Desc$, inherited by
-levitation, plays a central role here: without an induction principle,
-it would be impossible to program on $\Desc$.
+define such function. The induction principle of $\Desc$ -- inherited
+by levitation -- plays a central role here: without an induction
+principle, it would be impossible to program on $\Desc$.
 
@@
 a $\DSigma{X}{\Unit}$, that is an element of $X$. We keep $E$ and $D$
-as they were, hence leaving the functor untouched. The interpretation
+as they were, hence leaving the functor unchanged. The interpretation
 of this definition convince us that we have built the functor
@@
 the ability, for any data-type, to extend it with a notion of
-variable. Then, with have equipped this structure with the
+variable. Then, we have equipped this structure with the
 corresponding monadic operation, ``bind'' and ``return''. Whereas the
@@
 crucially rely on the fact that data-types are nothing but codes. We
-are therefore entitled to modify it and, in this case, extend
-it. Extending a data-type might give raise to a more
-structured object, as was the case here.
+are therefore entitled to modify this code and, in this case, extend
+it. Extending a data-type might give rise to a more structured
+object, as was the case here.
 
@@
 structure: the bind and return operations. We have been able to build
-them as generic functions. Here, we rely on the ability to compute
+them as generic functions. Here, we rely on our ability to compute
 over descriptions, thanks to the induction principle.

### 2010-03-13 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: merge Conor's comments in. (first pass)

@@
 judgemental equality respects symmetry, reflexivity, and
-transitivity. Further, it also encompasses the $\beta$-rules for
-function application and pairs. Finally, equality under lambda is
-dealt with by the $\xi$-rule.
+transitivity. Further, it encompasses the $\beta$-rules for function
+application and pairs. Finally, it respects structural equality,
+including equality under lambda ($\xi$-rule).
 
@@
 type-theory, some opportunities arise. We will present some of them
-along the course of the paper.
+along the course of the paper. \note{It is just a specification. We
+  expect the propositional theory to provide a decidable machinery to
+  implement it.}
+
+\note{plus structural rules, including lambda ($\xi$-rule)}
 
@@
 objects have a structure, the structure is lost during the encoding,
-when they become anonymous inhabitants of a $Π$ or $Σ$-type.
+when they become anonymous inhabitants of a $\Pi$ or $\Sigma$-type.
 
@@
 
-The $\Enum$ type comes with further machinery. Indeed, to manipulate
-the $\EnumU$ and $\EnumT{}$ objects, we need to equip our type theory
-with an appropriate \emph{elimination principle}. A first step
-consists in defining the $\spi{}{}$ operator:
+The type of finite sets comes with further machinery. Indeed, to
+manipulate the $\EnumU$ and $\EnumT{}$ objects, we need to equip our
+type theory with an appropriate \emph{elimination principle}. A first
+step consists in defining the $\spi{}{}$ operator:
 
-\[
+\[\stk{
 %% spi
 \spi{}{} : \PITEL{\V{e}}{\EnumU}
-           \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set 
-\]
+           \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set \\
+\begin{array}{@{}ll@{\:\mapsto\:\:}l}
+\spi{\NilE}{& P}        & \Unit \\
+\spi{(\ConsE{t}{e})}{& P} & \TIMES{P\: \Ze}{\spi{e}{(\LAM{x} P\: (\Su{x}))}}
+\end{array}
+}\]
+
+\note{Needs to be defined.}
 
@@
 
-\[
+\[\stk{
 %% switch
@@
                \PITEL{\V{x}}{\EnumT{e}} \To P~x
+\end{array} \\
+\begin{array}{@{}l@{\:\mapsto\:\:}l}
+\switch{(\ConsE{t}{e})}{P}{b}{\Ze}      & \fst{b} \\
+\switch{(\ConsE{t}{e})}{P}{b}{(\Su{x})} & \switch{e}{(\LAM{x} P
+  (\Su{x}))}{(\snd{b})}{x}
 \end{array}
-\]
+}\]
 
@@
 come with a notion of finite function space. However, we had to
-extract that intuition from the type, by a careful reading. At a first
-glance, this seems to contradict our argument in favour of types for
-coding representation. Here, we are overflown by low-level, very
-precise type information.
+extract that intuition from the type, by a careful reading. This seems
+to contradict our argument in favour of types for coding
+representation. Here, we are overflown by low-level, very precise type
+information.
 
@@
 faced with encoded data: while we were suffering from a crucial lack
-of information, we are here facing too much information, hence loosing
+of information, we are now facing too much information, hence loosing
 focus. This is a general issue with the usage of codes, as they convey
-much more information than what the user is used to see.
+much more information than what the user is willing to see.
 
@@
 should not be worried by information overload, but how to use it as
-much as possible. Having mastered this difficulty, we should not be
-afraid by code for practical purposes.
+much as possible. Therefore, we should not be afraid by using code for
+practical purposes.
 
@@
 \emph{pushed} into types. We are handed a type together with a term,
-our task consists in checking that the term is indeed of this type. We
-are therefore allowed, and actually encouraged, to use the information
-provided by the type. Therefore, we can relax our requirements on the
-term. A typical example is the lambda:
+our task consists in checking that the term is indeed of this
+type. Doing so, we are allowed, and actually encouraged, to use the
+information provided by the type. Therefore, we can relax our
+requirements on the term. A typical example is the lambda abstraction:
 
 \[
-\Rule{\Gamma ; \xS \Vdash \propag{\push{t}{T\: x}}
+\Rule{\Gamma ; \xS \Vdash \propag{\push{t}{T}}
                                  {t'}}
-     {\Gamma \Vdash \propag{\push{\LAM{\x} t}{\PiTy{S}{T}}}
+     {\Gamma \Vdash \propag{\push{\LAM{\x} t}{\PIS{\xS}{T}}}
                            {\PLAM{\x}{S} t'}} 
@@
 In our formal presentation of the type theory, we were forced to
-annotate the lambda by the type of its domain. In a bidirectional
-setting, the domain is provided by the $\Pi$ type we are pushing the
-lambda into. Hence we do not need to annotate the lambda.
+annotate lambda abstractions by the type of the domain. In a
+bidirectional setting, the domain is provided by the $\Pi$ type we are
+pushing the abstraction into. Hence we do not need the annotation.
 
@@
 $\exprIn$ and a type propagated down to it, they elaborate a low-level
-term. This is precisely the availability of the type we are pushed
-into that makes the power of this technique. Its benefit on lambda
+term. The strength of this approach comes from the availability of the
+type to drive the elaboration of expressions. Its benefit on lambda
 abstraction and pairs is well-known. Let us explore the consequences
@@
 eliminator. Nor will she \emph{be presented} the elaborated term, only
-the high-level . Bidirectional type-checking and type-directed
+the high-level one. Bidirectional type-checking and type-directed
 pretty-printing seal the low-level type theory behind the high-level
@@
 
-This example teaches us that we should not be afraid by codes, as far
+This example teaches us that we should not be afraid by codes, as soon
 as type information is available. We have shown how to rationalise

### 2010-03-15 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: merge Conor's comments in. (first pass)
Desc@ICFP: fix substI type
Desc@ICFP: Move the lambda example out of the way. Replaced by Hutton's razor.

@@
 \[\stk{
-\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
+\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
 \;\;\ldots \\
@@
 \[\stk{
-\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
+\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
 \;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
@@
 \stk{
-\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
+\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
 \;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
@@
 
-\begin{wstructure}
-<- Untyped lambda terms [figure]
-    <- Indexed by Nat
-        <- De Bruijn "encoding"
-    <- Variable in Fin n
-        -> Correct by construction lambda term
-    /> Lack the monadic structure [Altenkirch & Reus ???]
-        -> Miss substitution operator
-        /> Will be fixed later
-\end{wstructure}
+\begin{structure}
+<- Hutton's razor
+    ???
+\end{structure}
 
-Our next example is, for once, building an inductive family. We
-present the implementation of untyped lambda terms, as a family
-indexed by natural numbers. The natural numbers play the role of De
-Bruijn indexes. The indexed description is the following:
-
-\[\stk{
-\LambdaTD : \Nat \To \IDesc{\Nat} \\
-\begin{array}{@{}ll}
-\LambdaTD\: n = \DSigma{}{} & (\EnumT [ \LambdaTVar, \LambdaTApp, \LambdaTLam ]) \\
-                            & \left[\begin{array}{l}
-                                  \DConst{\Fin{n}} \\
-                                  \DProd{\DVar{\NatSuc{n}}}{\DVar{n}} \\
-                                  \DVar{(\NatSuc{n})}
-                              \end{array}\right]
-\end{array}
-}\]
-
-A variable is any number between $0$ and $n$, hence forbidding the
-reference of out-of-scope binders. Making a lambda increases by one
-the possible index choice of the underlying term. A function
-application applies a term with $n+1$ binders, resulting in a term
-with $n$ binders.
-
-We have therefore given a representation of lambda terms as a
-data-type. Any inhabitant of this data-type is a valid lambda term,
-forbidding out-of-scope reference in variables. While we have the
-data-type, we can regret the absence of structure. As shown by
-Altenkirch and Reus~\cite{altenkirch:monadic-lambda}, the lambda terms
-actually enjoy a monadic structure. In this setting, the ``bind''
-corresponds to substitution. We shall remedy to this situation
-shortly.
+\note{Begging for a transtion here.}
 
@@
 \[
-\substI : \PITEL{\V{I}, \V{J}}{\Set}
-          \PITEL{\sigma}{\V{I} \To \IDesc{\V{J}}}
-          \PITEL{\V{D}}{\IDesc{I}} \To
-          \IDesc{J}
+\begin{array}{@{}ll}
+\substI : & \PITEL{\V{I}}{\Set}
+            \PITEL{\V{X}, \V{Y}}{\V{I} \To \Set}
+            \PITEL{\V{R}}{\V{I} \To \TagIDesc{\V{I}}} \\
+          & \PITEL{\V{\sigma}}{\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 
+                               \IMu{I}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}} \\
+          & \PITEL{\V{i}}{\V{I}}
+            \PITEL{\V{D}}{\IMu{I}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})}{\V{i}}} \To
+            \IMu{I}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}
+\end{array}
 \]
 
-\begin{wstructure}
-<- Untyped lambda terms
-    <- Obvious instance of free monad
-        <- Variable: Return
-        <- App and Abst: indexed functor
-        <- Bind: substitution
-    -> Get a correct-by-construction lambda term
-        /> With substitution for free!
-        <- We should not pay for the structure which *is* here
-\end{wstructure}
-
-In Section~\ref{sec:idesc-examples}, we were missing the monadic
-structure of lambda terms. In particular, we wish we were given a
-substitution principle for free. Above, we have generically built such
-substitution operation for free monads. Can we be rewarded for our
-effort? Let us define the following $\LambdaTFreeD$ description,
-largely inspired by $\LambdaTD$:
-
-\[\stk{
-\LambdaTFreeD : \Nat \To \IDesc{\Nat} \\
-\begin{array}{@{}ll}
-\LambdaTFreeD\: n = \DSigma{}{} & (\EnumT [ \LambdaTApp, \LambdaTLam ]) \\
-                                & \left[\begin{array}{l}
-                                  \DProd{\DVar{\NatSuc{n}}}{\DVar{n}} \\
-                                  \DVar{(\NatSuc{n})}
-                                  \end{array}\right]
-\end{array}
-}\]
-
-In essence, we have forgotten variables, just keeping the nodes of
-lambda terms: abstraction and application. Our hope is that taking the
-free monad of this indexed functor gives us our lambda terms back:
-
-\note{ FreeMonad takes a damned tagged description }
-
-\[\stk{
-\LambdaTD : \Nat \To \IDesc{\Nat} \\
-\LambdaTD\: \mapsto \FreeIMonad{[\LambdaTFreeD]}{\Nat}{\Fin{}}
-}\]
-
-By a straightforward unfolding of definitions, the reader will
-convince herself that we recover the initial definition. However, this
-time, we have a substitution operation. In summary, we have done less
-work in defining $\LambdaTFreeD$ but we have pointed out its
-structure, so, rightfully, we have been rewarded in our effort. The
-lesson is that, when structure there is, one should be allowed to get
-it for free.
+\begin{structure}
+<- Hutton's razor is a free monad
+    <- What is it?
+        -> It's an evaluator
+    ???
+\end{structure}
 

### 2010-03-15 papers/someconf-sometime-binding/paper.tex
Plop, the ICFP paper made a baby.

\begin{wstructure}
<- Untyped lambda terms [figure]
    <- Indexed by Nat
        <- De Bruijn "encoding"
    <- Variable in Fin n
        -> Correct by construction lambda term
    /> Lack the monadic structure [Altenkirch & Reus ???]
        -> Miss substitution operator
        /> Will be fixed later
\end{wstructure}

Our next example is, for once, building an inductive family. We
present the implementation of untyped lambda terms, as a family
indexed by natural numbers. The natural numbers play the role of De
Bruijn indexes. The indexed description is the following:

\[\stk{
\LambdaTD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\LambdaTD\: n = \DSigma{}{} & (\EnumT [ \LambdaTVar, \LambdaTApp, \LambdaTLam ]) \\
                            & \left[\begin{array}{l}
                                  \DConst{(\Fin{n})} \\
                                  \DProd{\DVar{n}}{\DVar{n}} \\
                                  \DVar{(\NatSuc{n})}
                              \end{array}\right]
\end{array}
}\]

A variable is any number between $0$ and $n$, hence forbidding the
reference of out-of-scope binders. Making a lambda increases by one
the possible index choice of the underlying term. A function
application applies a term with $n+1$ binders, resulting in a term
with $n$ binders.

We have therefore given a representation of lambda terms as a
data-type. Any inhabitant of this data-type is a valid lambda term,
forbidding out-of-scope reference in variables. While we have the
data-type, we can regret the absence of structure. As shown by
Altenkirch and Reus~\cite{altenkirch:monadic-lambda}, the lambda terms
actually enjoy a monadic structure. In this setting, the ``bind''
corresponds to substitution. We shall remedy to this situation
shortly.

\begin{wstructure}
<- Untyped lambda terms
    <- Obvious instance of free monad
        <- Variable: Return
        <- App and Abst: indexed functor
        <- Bind: substitution
    -> Get a correct-by-construction lambda term
        /> With substitution for free!
        <- We should not pay for the structure which *is* here
\end{wstructure}

In Section~\ref{sec:idesc-examples}, we were missing the monadic
structure of lambda terms. In particular, we wish we were given a
substitution principle for free. Above, we have generically built such
substitution operation for free monads. Can we be rewarded for our
effort? Let us define the following $\LambdaTFreeD$ description,
largely inspired by $\LambdaTD$:

\[\stk{
\LambdaTFreeD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\LambdaTFreeD\: n = \DSigma{}{} & (\EnumT [ \LambdaTApp, \LambdaTLam ]) \\
                                & \left[\begin{array}{l}
                                  \DProd{\DVar{n}}{\DVar{n}} \\
                                  \DVar{(\NatSuc{n})}
                                  \end{array}\right]
\end{array}
}\]

In essence, we have forgotten variables, just keeping the nodes of
lambda terms: abstraction and application. Our hope is that taking the
free monad of this indexed functor gives us our lambda terms back:

\note{ FreeMonad takes a damned tagged description }

\[\stk{
\LambdaTD : \Nat \To \IDesc{\Nat} \\
\LambdaTD\: \mapsto \FreeIMonad{[\LambdaTFreeD]}{\Nat}{\Fin{}}
}\]

By a straightforward unfolding of definitions, the reader will
convince herself that we recover the initial definition. However, this
time, we have a substitution operation. In summary, we have done less
work in defining $\LambdaTFreeD$ but we have pointed out its
structure, so, rightfully, we have been rewarded in our effort. The
lesson is that, when structure there is, one should be allowed to get
it for free.

### 2010-03-15 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: math code for All and all predicate.

@@
 Then, we have to translate the notion of algebra in our dependent
-setting. This is achieved by the predicate $\All{D}{X}{P}{x}$ that,
-intuitively, states that $\Bhab{P}{X \To \Set}$ holds everywhere in
-$x$. Based on that knowledge, we ought to be able to prove that $P$
-holds for $\Con{xs}$. We close our elimination principle by taking an
-argument $x$ in $\Mu{D}$ and claiming that the motive $P$ is realised
-on $x$.
+setting. This is achieved by the predicate $\All{D}{X}{P}{x}$
+(Fig.\ref{fig:all-predicates}) that, intuitively, states that
+$\Bhab{P}{X \To \Set}$ holds everywhere in $x$. Based on that
+knowledge, we ought to be able to prove that $P$ holds for
+$\Con{xs}$. We close our elimination principle by taking an argument
+$x$ in $\Mu{D}$ and claiming that the motive $P$ is realised on $x$.
 
@@
 
+\begin{figure*}
+
+\[
+\begin{array}{ll}
+%%
+\stk{
+\begin{array}{@{}ll}
+\All{}{}{}{} : & \PITEL{\V{D}}{\Desc}
+                 \PITEL{\V{X}}{\Set}
+                 \PITEL{\V{P}}{\V{X} \To \Set} \\
+               & \PI{\V{xs}}{\descop{D}{\Mu{X}}} 
+                 \Set 
+\end{array} \\
+\All{\DId}{X}{P}{x} \mapsto P\: x \\
+\All{(\DConst{Z})}{X}{P}{x} \mapsto Z \\
+\All{(\DProd{D}{D'})}{X}{P}{\pair{d}{d'}{}} \mapsto \TIMES{\All{D}{X}{P}{d}}{\All{D'}{X}{P}{d'}} \\
+\All{(\DSigma{S}{T})}{X}{P}{\pair{a}{b}{}} \mapsto \All{(T\: a)}{X}{P}{b} \\
+\All{(\DPi{S}{T})}{X}{P}{f} \mapsto \PI{\V{s}}{S} \All{(T\: a)}{X}{P}{(f\: s)}
+}
+&
+%%
+\stk{
+\begin{array}{@{}ll}
+\all{}{}{}{}{} : & \PITEL{\V{D}}{\Desc}
+                   \PITEL{\V{X}}{\Set}
+                   \PITEL{\V{P}}{\V{X} \To \Set} \\
+                 & \PITEL{\V{R}}{\PI{\V{x}}{X} P\: x}
+                   \PI{\V{xs}}{\descop{D}{\Mu{X}}} 
+                   \All{D}{X}{P}{xs} 
+\end{array} \\
+\all{\DId}{X}{P}{R}{x} = R x \\
+\all{(\DConst{Z})}{X}{P}{R}{x} = z \\
+\all{(\DProd{D}{D'})}{X}{P}{R}{\pair{d}{d'}{}} = \pair{\all{D}{X}{P}{R}{d}}{\all{D'}{X}{P}{R}{d'}}{} \\
+\all{(\DSigma{S}{T})}{X}{P}{R}{\pair{a}{b}{}} = \all{(T\: a)}{X}{P}{R}{b} \\
+\all{(\DPi{S}{T})}{X}{P}{R}{f} = \LAM{s} \all{(T\: s)}{X}{P}{R}{(f\: s)}
+\end{array}
+}
+\]
+
+\caption{Induction predicates}
+\label{fig:all-predicates}
+
+\end{figure*}
+
 At this stage, we are still left with implementing this
@@
 automatically comes equipped with an induction principle.
-
-\note{Shall we give the code of $\All$?}
 

### 2010-03-15 papers/icfp-2010-desc/paper_type_theory.tex
IDesc@ICFP: clarify elaboration/distillation

@@
 user has never to explicitly \emph{call} the $\F{switch}$
-eliminator. Nor will she \emph{be presented} the elaborated term, only
-the high-level one. Bidirectional type-checking and type-directed
-pretty-printing seal the low-level type theory behind the high-level
-expressions.
-
-\note{And suddenly, type-directed pretty printing comes into the
-  picture. Surely, it should ``just'' be the inverse of
-  elaboration. But is it? And how do we present it?}
+eliminator. For diagnosis purposes, a similar, reverse mechanism turns
+low-level terms into high-level expressions. Hence, this system, which
+presentation is out of the scope of this paper, seals the low-level
+type theory behind the high-level expressions.
 

### 2010-03-16 papers/icfp-2010-desc/paper_idesc.tex
IDesc@ICFP: update outline to accomodate the arrival of Hutton's razor. Includes tagged indexed descriptions.

@@
 \begin{structure}
-<- Hutton's razor
-    ???
+<- Tagged constructor choice
+    <- Index available when defining the data-type
+        -> Can influence the choice of constructors
+        -> Dependently-typed data-types
+            <- term (index) influence types
+    -> Two parts
+        <- Always on the menu
+            <- E : EnumU 
+            <- ED : E -> IDesc I
+        <- Index-dependent
+            <- F : I -> EnumU 
+            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
+        -> taggedIDesc I == Sigma E ED x Sigma F FD
+        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
 \end{structure}
 
-\note{Begging for a transtion here.}
+\begin{structure}
+<- Hutton's razor
+    <- Types
+        <- 'Nat
+        <- 'Bool
+    <- Term [figure]
+        <- val : Val a -> 'a  for Val : Ty -> Set, mapping to Nat and Bool
+        <- cond : 'Bool -> a -> a -> a
+        <- plus : 'Nat -> 'Nat -> 'Nat
+        <- le : 'Nat -> 'Nat -> 'Bool
+    -> evaluation: IMu TermD -> Val
+        -> it is a catamorphism
+            <- Look closer at the type
+        -> implementation [code]
+            <- Just define one reduction step
+            -> cata does the rest
+                /> cata is for free!
+    /> Closed term
+        <- only constants and operations on them
+        -> Extend Val with Var : Ty -> Set, mapping to EnumU
+            -> Open term
+            -> Language of well-typed terms
+                <- By construction
+        <- evaluator, with a context
+            -> First, close variables
+                -> Perform assignment
+                <- subst [code]
+            /> Then, perform subst everywhere in the term
+                -> Show type [code]
+                /> This is a bind!?
+                -> There is some more structure 
+                    -> We should try to get it
+\end{structure}
+
+\note{Begging for a transition here.}
 
@@
 <- Hutton's razor is a free monad
-    <- What is it?
-        -> It's an evaluator
-    ???
+    <- substI was our candidate bind
+        -> Massage the definition of expr to get it for free
+    <- Finding the functor
+        <- 'val is the return
+        <- The other components are the action
+        -> Updated tagged description [figure]
+    -> Compute the free monad Hutton * X
+        -> With X = Val: We get back our closed terms
+        -> With X = Val + Var vars: We get back our open terms
+    /> On open terms, we get a substitution
+        -> Apply substI on assgnmt
+            -> Have a well-typed language
+            -> Get a safe evaluator 
+                <- for well-typed terms 
+                <- in well-typed contexts
 \end{structure}

### 2010-03-16 papers/icfp-2010-desc/paper_idesc.tex
IDesc@ICFP: introduce tagged index description

@@
 
+\begin{structure}
+<- Tagged constructor choice
+    <- Index available when defining the data-type
+        -> Can influence the choice of constructors
+        -> Dependently-typed data-types
+            <- term (index) influence types
+    -> Two parts
+        <- Always on the menu
+            <- E : EnumU 
+            <- ED : E -> IDesc I
+        <- Index-dependent
+            <- F : I -> EnumU 
+            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
+        -> taggedIDesc I == Sigma E ED x Sigma F FD
+        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
+\end{structure}
+
+In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
+form for descriptions. This format follows the usual presentation of
+inductive types as sums of product. With indexed description, this
+definition can be generalized. Indeed, when defining an indexed
+data-type, we have access to this index. Therefore, we can use this
+index to influence the choice of constructors. This captures the
+essence of dependent data-types: a term -- the index -- has the
+ability to influence the data-type.
+
+For convenience, we divide a tagged indexed description in two parts:
+first, the constructors that do not depend on the index; then, the
+constructors that do. The non-dependent part mirrors the definition
+for non-indexed descriptions: we are provided a finite choice of
+constructors. The index-depend part simply indexes the choice of
+constructors by $I$. Hence, by inspecting the index, it is possible to
+enable or disable the constructors.
+
+\[\stk{
+ \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
+ \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})})  \\
+ \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}}
+}\]
+
+In the case of vectors, for instance, for the index $n = \NatZero$, we
+would only offer the constructor $\ListNil$. Similarly, for $n =
+\NatSuc{n'}$, we only offer the constructor $\ListCons{}{}$. We will
+see more instances of this pattern in the following.
+
 \note{Should we say something about IMu, fixed-points, and their
@@
 purpose $\F{switchID}$ operator.
-
-\begin{structure}
-<- Tagged constructor choice
-    <- Index available when defining the data-type
-        -> Can influence the choice of constructors
-        -> Dependently-typed data-types
-            <- term (index) influence types
-    -> Two parts
-        <- Always on the menu
-            <- E : EnumU 
-            <- ED : E -> IDesc I
-        <- Index-dependent
-            <- F : I -> EnumU 
-            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
-        -> taggedIDesc I == Sigma E ED x Sigma F FD
-        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
-\end{structure}
 
@@
 
+\subsubsection{Examples}
+
 \begin{structure}

### 2010-03-17 papers/icfp-2010-desc/paper_idesc.tex
IDesc@ICFP: done with Hutton's razor

--- 2010-03-15 papers/icfp-2010-desc/paper_idesc.tex
@@
 
+\begin{wstructure}
+<- Tagged constructor choice
+    <- Index available when defining the data-type
+        -> Can influence the choice of constructors
+        -> Dependently-typed data-types
+            <- term (index) influence types
+    -> Two parts
+        <- Always on the menu
+            <- E : EnumU 
+            <- ED : E -> IDesc I
+        <- Index-dependent
+            <- F : I -> EnumU 
+            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
+        -> taggedIDesc I == Sigma E ED x Sigma F FD
+        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
+\end{wstructure}
+
+In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
+form for descriptions. This format follows the usual presentation of
+inductive types as sums of product. With indexed description, this
+definition can be generalized. Indeed, when defining an indexed
+data-type, we have access to this index. Therefore, we can use this
+index to influence the choice of constructors. This captures the
+essence of dependent data-types: a term -- the index -- has the
+ability to influence the data-type.
+
+For convenience, we divide a tagged indexed description in two parts:
+first, the constructors that do not depend on the index; then, the
+constructors that do. The non-dependent part mirrors the definition
+for non-indexed descriptions: we are provided a finite choice of
+constructors. The index-depend part simply indexes the choice of
+constructors by $I$. Hence, by inspecting the index, it is possible to
+enable or disable the constructors.
+
+\[\stk{
+ \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
+ \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})})  \\
+ \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}}
+}\]
+
+In the case of vectors, for instance, for the index $n = \NatZero$, we
+would only offer the constructor $\ListNil$. Similarly, for $n =
+\NatSuc{n'}$, we only offer the constructor $\ListCons{}{}$. We will
+see more instances of this pattern in the following.
+
 \note{Should we say something about IMu, fixed-points, and their
@@
 
-\subsection{Examples}
+\subsubsection{Examples}
 \label{sec:idesc-examples}
+
+\paragraph{Natural numbers:}
 
@@
 transition.
+
+\paragraph{Indexed descriptions:}
 
@@
 
-\begin{structure}
+\paragraph{Typed expressions:}
+
+\begin{wstructure}
 <- Hutton's razor
-    ???
-\end{structure}
+    <- Types
+        <- 'Nat
+        <- 'Bool
+    <- Term [figure]
+        <- val : Val 'a -> 'a  for Val : Ty -> Set, mapping to Nat and Bool
+        <- cond : 'Bool -> a -> a -> a
+        <- plus : 'Nat -> 'Nat -> 'Nat
+        <- le : 'Nat -> 'Nat -> 'Bool
+\end{wstructure}
 
-\note{Begging for a transtion here.}
+%% Types
+\newcommand{\Ty}{\C{Ty}}
+\newcommand{\Ebool}{\etag{\CN{bool}}}
+\newcommand{\Enat}{\etag{\CN{nat}}}
+\newcommand{\Epolym}{\etag{\alpha}}
+
+%% Constructors
+\newcommand{\Eval}[1]{\etag{\CN{val}}~#1}
+\newcommand{\Econd}[3]{\etag{\CN{cond}}~#1~#2~#3}
+\newcommand{\Eplus}[2]{\etag{\CN{plus}}~#1~#2}
+\newcommand{\Ele}[2]{\etag{\CN{le}}~#1~#2}
+
+%% Index mapper (terminology?)
+\newcommand{\Val}[1]{\D{Val}~#1}
+\newcommand{\Var}[2]{\D{Var}_{#1}~#2}
+
+%% Hutton expressions
+\newcommand{\HExprD}{\C{ExprD}}
+\newcommand{\HExprAD}{\C{ExprAlwaysD}}
+\newcommand{\HExprID}{\C{ExprIndexedD}}
+\newcommand{\HExprVarD}{\C{ExprD}_{\C{Var}}}
+\newcommand{\HExprFreeD}{\C{ExprD}_{\C{Free}}}
+\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}_{\C{Free}}}
+
+So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
+not using any indexing. We are going to define a syntax for a small
+typed language. We consider two types, natural numbers and booleans:
+
+\[
+\Ty \mapsto \EnumT{[\Enat \quad \Ebool]}
+\]
+
+An expression of this language is either a value, a conditional
+expression, addition of numbers, or comparison of numbers. Informally,
+their type is the following:
+
+\[
+\begin{array}{l@{\::\:\:}l}
+\Econd{}{}{}     & \Ebool \To \Epolym \To \Epolym \To \Epolym  \\ 
+\Eplus{}{}       & \Enat \To \Enat \To \Enat                   \\
+\Ele{}{}         & \Enat \To \Enat \To \Ebool                  \\
+\Eval{}          & \Val{\Epolym} \To \Epolym
+\end{array}
+\]
+
+The notation $\Epolym$ represents a form of ``polymorphism'', meaning
+that $\alpha$ can be any type. \note{Polymorph? Will any ghost try to
+  haunt me with that?  A ghost under a mixed prefix?
+  Aaaaaaaaaaargh\ldots}
+
+The function $\Val{}$, used in the definition of $\Eval{}$, simply
+maps $\Epolym$ to the corresponding type in the host language. Hence,
+the argument of $\Eval{}$ are ensured to be of the expected type. We
+assume $\Nat$ and $\Bool$ represent natural numbers and booleans in
+the host language. Further, we also assume the existence of an
+addition and comparison operator in $\Nat$, respectively name
+$\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$ as follow:
+
+\[\stk{
+\Val{} : \Ty \To \Set \\
+\begin{array}{@{}l@{\:=\:\:}l}
+\Val{\Enat}   & \Nat \\
+\Val{\Ebool}  & \Bool
+\end{array}
+}\]
+
+In our universe of descriptions, this data-type is represented by a
+tagged indexed description. We use the index to carry the type: the
+resulting description is indexed by $\Ty$. We observe that some
+constructors are ``polymorph'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
+other hand, the $\Eplus$ and $\Ele$ constructors are
+index-dependent. $\Eplus$ is defined if and only if the result type --
+the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
+index is $\Ebool$. The actual code precisely follows this intuition,
+as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
+informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
+pattern-matching. Formally, this corresponds to a call to the
+$\switch{}{}{}{}$ eliminator.
+
+\begin{figure*}
+
+\[\stkc{
+\stk{
+\HExprD : \TagIDesc{\Ty} \\
+\HExprD \mapsto ( \HExprAD , \HExprID ) \\
+} \\
+\\
+\begin{array}{ll}
+\stk{
+\HExprAD : \ATagIDesc{\Ty} \\
+\HExprAD \mapsto \left[
+                 \begin{array}{l}
+                   \EnumT{[\Eval{}  \Econd{}{}{}]} \red{,} \\
+                   \LAM{ty}
+                   \left[ %%! Should be read
+                   \begin{array}{l}
+                   \DConst{(\Val{ty})} \\
+                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{ty}}{\DVar{ty}}}
+                   \end{array}
+                   \right]
+                 \end{array}
+                 \right] %%! Should be \red
+\\
+\\
+} &
+\stk{
+\HExprID : \ITagIDesc{\Ty} \\
+\HExprID \mapsto \left[
+                 \begin{array}{l}
+                   \LAM{ty} \case{ty}{\Enat \To \EnumT{[\Eplus{}{}]} \\ \Ebool \To \EnumT{[\Ele{}{}]}} \red{,} \\
+                   \LAM{ty}
+                   \case{ty}{
+                     \Enat \To \DProd{\DVar{nat}}{\DVar{nat}} \\
+                     \Ebool \To \DProd{\DVar{nat}}{\DVar{nat}}
+                   }
+                   \end{array}
+                   \right]
+}
+\end{array}
+}\]
+
+\caption{Syntax of typed expressions}
+\label{fig:hexpr-full}
+
+\end{figure*}
+
+\note{the big square brackets should be red}
+
+\begin{wstructure}
+    -> evaluation: IMu TermD -> Val
+        -> it is a catamorphism
+            <- Look closer at the type
+        -> implementation [code]
+            <- Just define one reduction step
+            -> cata does the rest
+                /> cata is for free!
+\end{wstructure}
+
+\newcommand{\evalH}{\F{eval}_{\green{\Downarrow}}}
+\newcommand{\evalOne}{\F{eval}_{\green{\downarrow}}}
+
+Having implemented the syntax, we would like to describe its
+semantics. To do so, we implement an evaluator. The type of the
+evaluator is:
+
+\[
+\evalH : \PI{\V{ty}}{\Ty} 
+         \IMu{\Ty}{\toIDesc{\HExprD}}{ty} \To
+         \Val{ty}
+\]
+
+The type of $\F{eval}$ is strikingly similar to a
+catamorphism. Indeed, implementing a single step of evaluation -- the
+algebra -- is sufficient, as $\F{cataI}$ gives, for free, the
+full evaluator. The implementation is as follow:
+
+\[\stk{
+\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}}{\Ty}{ty} \To {\Val{ty}} \\
+\begin{array}{@{}l@{\:=\:\:}l}
+\evalOne\: \_\: \pair{\Eval{}}{x}{}                                         & x \\
+\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{x}{\_}{}}{}}{}   & x \\
+\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{y}{}}{}}{}  & x \\
+\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{x}{y}{}}{}                       & \F{plusHost}\: x y \\
+\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{x}{y}{}}{}                        & \F{leHost}\: x y 
+\end{array} \\
+\\
+\evalH : \PI{\V{ty}}{\Ty} 
+           \IMu{\Ty}{\toIDesc{\HExprD}}{ty} \To
+           \Val{ty} \\
+\evalH\: ty\: term = \F{cataI}\: \Ty\: 
+                                 \toIDesc{\HExprD}\: 
+                                 \Val{}\: 
+                                 \evalOne\: 
+                                 ty\: 
+                                 term
+}\]
+
+\note{We have not defined \F{cataI}!}
+
+\begin{wstructure}
+    /> Closed term
+        <- only constants and operations on them
+        -> Extend Val with Var : Ty -> Set, mapping to EnumU
+            -> Open term
+            -> Language of well-typed terms
+                <- By construction
+\end{wstructure}
+
+Hence, we have defined the syntax of a typed language of arithmetic
+and boolean operations. We have given its semantics through an
+evaluation function. Provided a one step semantic of the language, the
+big step evaluation is granted without effort thanks to the generic
+catamorphism.
+
+However, so far, we are only able to define and manipulate
+\emph{closed} terms. By abstracting over $\Var{}$, it is possible to
+build and manipulate \emph{open} terms, that is terms with symbolic
+variables. On the model of $\Val{}$, we define $\Var{}{}$:
+
+\[\stk{
+\Var{}{} : \EnumU \To \Ty \To \Set \\
+\Var{dom}{\_} = \EnumT{dom}
+}\]
+
+Whereas $\Val{}$ was mapping the type to the corresponding host type,
+$\Var{}{}$ maps types to a finite set. The finite set -- the context
+-- contains values. A variable is therefore a $\Eval{}$ that contains
+a pointer to a particular element of the finite set -- an element of
+$\EnumT{dom}$. The extra argument to $\Var{}{}$ is the domain of this
+context.
+
+Consequently, replacing $\Val{ty}$ by $(\SUM{\Val{ty}}{\Var{dom}{ty}})$ in
+Figure~\ref{fig:hexpr-full} turns the language of closed term into a
+language of opened terms with constants. \note{Is it polite and
+  comprehensible to ask that?} This defines a new indexed description,
+called $\HExprVarD$.
+
+\begin{wstructure}
+        <- evaluator, with a context
+            -> First, close variables
+                -> Perform assignment
+                <- subst [code]
+\end{wstructure}
+
+\newcommand{\discharge}{\F{discharge}}
+
+Again, we would like to give a semantics to this extended language. We
+proceed in two steps: first, replace the variables by their value in
+the context; then, evaluate the resulting closed term. Thanks to
+$\evalH$, we are already able to solve the second problem. Let us
+focus on discharging variables from the context. Again, we can
+subdivide this problem: first, we need the ability to discharge a
+single variable from the context; then, we apply this $\discharge$
+function on every variables in the term.
+
+The $\discharge$ function is relative to the required type, the domain
+of the context, and a context containing values of the corresponding
+type. Its action is to map variables to their value in context, and
+directly return constant values. This corresponds to the following
+function:
+
+\[\stk{
+\begin{array}{@{}ll}
+\discharge : & \PITEL{\V{ty}}{\Ty}
+               \PITEL{\V{dom}}{\EnumU} \\
+             & \PI{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}} \\
+             & (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To
+               \IMu{\Ty}{\toIDesc{\HExprD}}{ty} 
+\end{array} \\
+\begin{array}{@{}l@{\:\mapsto\:\:}l}
+\discharge\: ty\: dom\: \gamma\: (\SumLeft x)  & \Con{\pair{\Eval{}}{x}{}} \\
+\discharge\: ty\: dom\: \gamma\: (\SumRight v) & \switch{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}{\gamma}{v}
+\end{array}
+}\]
+
+\begin{wstructure}
+            /> Then, perform subst everywhere in the term
+                -> Show type [code]
+                /> This is a bind!?
+                -> There is some more structure 
+                    -> We should try to get it
+\end{wstructure}
+
+Having implemented the local $\discharge$ operation, we are left with
+applying it over all variables of the term. The type of this operation
+is the following:
+
+\newcommand{\substH}{\F{substExpr}}
+
+\[
+\begin{array}{@{}ll}
+\substH  : & \PITEL{\V{ty}}{\Ty}
+            \PITEL{\V{dom}}{\EnumU} 
+            \PITEL{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}} \\
+          &  \PI{\V{\sigma}}{\PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom}{ty}) \To
+                             \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}} \\
+          & \IMu{\Ty}{\toIDesc{\HExprVarD}}{\V{ty}} \To
+            \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
+\end{array}
+\]
+
+Abstracting away the book-keeping introduced by the context, this
+definition looks familiar. Indeed, it is extremely similar to a
+monadic ``bind''. This is not surprising as we are defining a
+first-order syntax with variables: our data-type enjoys more structure
+than what we are currently provided. In particular, this is
+reminiscent to a free monad, where $\Eval{}$ is the ``return''
+introducing variables. Instead of implementing $\substH$ in this
+special case, we shall implement it in a generic setting, and express
+our example in that setting.
 
@@
 
-\begin{structure}
+\subsubsection{Examples}
+
+\paragraph{Typed expressions:}
+
+\begin{wstructure}
 <- Hutton's razor is a free monad
-    <- What is it?
-        -> It's an evaluator
-    ???
-\end{structure}
+    <- substI was our candidate bind
+        -> Massage the definition of expr to get it for free
+    <- Finding the functor
+        <- 'val is the return
+        <- The other components are the action
+        -> Updated tagged description [figure]
+    -> Compute the free monad Hutton * X
+        -> With X = Val: We get back our closed terms
+        -> With X = Val + Var vars: We get back our open terms
+\end{wstructure}
+
+In Section~\ref{sec:idesc-examples}, we had the intuition that our
+datatype $\HExprD$ and $\HExprVarD$ enjoys a monadic structure. We had
+identified the variable substitution operation as the ``bind'' of a
+free monad. The definition of $\substI$ above confirm our intuition:
+using $substI$, we should easily obtain $\substH$. To do so, we first
+have to massage the definition of our data-type, to exhibit its
+monadic structure.
+
+As previously mentioned, we identify $\Eval{}$ as the ``return'' of
+the free monad, while the other components are the action of the
+monad. As a result, the definition is similar to $\HExprD$ presented
+in Figure~\ref{fig:hexpr-full}, at the exception of $\HExprAD$ that is
+replaced by $\HExprAFreeD$:
+
+\[\stk{
+\HExprAFreeD : \ATagIDesc{\Ty} \\
+\HExprAFreeD \mapsto \left[
+                 \begin{array}{l}
+                   \EnumT{[\Econd{}{}{}]} \red{,} \\
+                   \LAM{ty}
+                   \left[ %%! Should be red
+                   \begin{array}{l}
+                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{ty}}{\DVar{ty}}}
+                   \end{array}
+                   \right]
+                 \end{array}
+                 \right] %%! Should be \red
+}\]
+
+The resulting data-type is called $\HExprFreeD$. By a simple unfolding
+of definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
+corresponds to the closed expression, $\HExprD$. Similarly,
+$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{ty})}$$
+corresponds to expressions with variables, $\HExprVarD$.
+
+\begin{wstructure}
+    /> On open terms, we get a substitution
+        -> Apply substI on assgnmt
+            -> Have a well-typed language
+            -> Get a safe evaluator 
+                <- for well-typed terms 
+                <- in well-typed contexts
+\end{wstructure}
+
+The evaluator for closed term we implemented in
+Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
+terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$ to values in
+$\Val{ty}$. We are therefore left with implementing $\substH$. We
+simply have to fill in the right arguments to $\substI$, the type
+guiding us:
+
+\[\stk{
+\begin{array}{@{}ll}
+\substH  : & \PITEL{\V{ty}}{\Ty}
+             \PITEL{\V{dom}}{\EnumU} 
+             \PITEL{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}}}{ty})}} \\
+           &  \PI{\V{\sigma}}{\PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom}{ty}) \To
+                              \IMu{\Ty}{(\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}})}{ty}} \\
+           & \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{} + \Var{dom}{}}}}{\V{ty}} \To
+             \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}}}{\V{ty}} \\
+\end{array} \\
+\substH\: ty\: dom\: \gamma\: \sigma\: term \mapsto 
+    \substI\: \Ty\: 
+              (\Val{} + \Var{dom}{})\: 
+              \Val{}\:
+              \HExprFreeD\:
+              \sigma\:
+              ty\:
+              term
+}\]
+
+\note{Not using $\discharge$, that's a shame.}
+
+Hence, we have completed our implementation of the evaluator for open
+terms. We started with a well-typed language of arithmetical
+expressions. We have seen how to take advantage of indexes to only
+present type-safe constructors. Then, we have implemented an evaluator
+for closed term, based on the generic catamorphism function. Having
+implemented an open term representation, we wanted a substitution
+operation, in order to close open terms in a context. Again, we have
+implemented this operation with the generic substitution
+operator. Hence, without much efforts, we have described the syntax of
+a well-typed language, together with its semantics.
 
@@
 \end{wstructure}
+
+\paragraph{Indexed description:}
 
@@
 \IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
-\IDescD\: I \mapsto \FreeMonad{[\IDescFreeD]}{I}\: I
+\IDescD\: I \mapsto \FreeIMonad{[\IDescFreeD]}{\Unit}{I}\: I
 }\]

### 2010-03-17 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix subst. Fix typo.
IDesc@ICFP: no 'beauty of the exercise'

@@
 
-This concludes our levitation work. Beyond the beauty of the exercise,
-there are several practical outcomes. First of all, this exercise
+This concludes our levitation work. Beyond its pedagogical value, this
+exercise has several practical outcomes. First of all, this exercise
 reveals that the $\Desc$ universe is just plain data. Just as any
@@
     \pair{\ListCons{\DVar{}}{E}}
-         {\pair{\DSigma{X}{\Unit}}{D}{}}{}
+         {\pair{\DSigma{X}{\DUnit}}{D}{}}{}
 }\]
@@
 We simply add a constructor, $\DVar{}$, and define its argument to be
-a $\DSigma{X}{\Unit}$, that is an element of $X$. We keep $E$ and $D$
+a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
 as they were, hence leaving the functor unchanged. The interpretation
@@
 
-\note{ Triple check. That doesn't work, right? }
-
 \[\stk{
@@
 \begin{array}{@{}l@{\:\:\mapsto\:}l}
-\apply\: \DVar{}\: X\: Y\: \sigma\: x  & \sigma x     \\
-\apply\: xs\:      X\: Y\: \sigma\: ys & ys
+\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}     & \sigma\: x                   \\
+\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
 \end{array}

### 2010-03-17 papers/icfp-2010-desc/paper_desc_levitation.tex
IDesc@ICFP: move Desc factorization to IDesc. Give a new reason for doing so.

@@
 
-\subsubsection{Desc, atomically}
-
-\begin{wstructure}
-<- Adding hindx have introduced some duplication
-    <- indx == hindx 1
-    -> We can factor out commonalities 
-        /> Obtain an equivalent presentation
-        /> Still embeddable (refer to the Agda model)
-\end{wstructure}
-
-Before moving on some examples of generic programming, we have to
-carry out some maintenance work. The extension of $\Desc$ with
-$\DHindx$ have introduced some duplication of functionality. In order
-to remove this clutter, we will factor out these commonalities. This
-results in a new definition of the $\Desc$ universe. This
-reorganisation is strictly equivalent to the previous one
-(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
-self-descriptive. We refer the reader to the companion technical
-report for details.
-
-\begin{wstructure}
-<- Give new presentation [figure]
-    <- hindx have introduced the notion of function space: 'Pi
-    <- hindx and indx are both composed by a binary product and a left open term: 'x and 'id 
-    -> Straightforward translation to the new system [equation]
-\end{wstructure}
-
-The factorised universe of description is presented in
-Figure~\ref{fig:factor-desc}. The modifications are justified as
-follow. First, $\DHindx$ have introduced a notion of function space:
-we capture this notion by the $\DPi$ constructor. Second, both
-$\DHindx$ and $\DIndx$ are composed by a product and the functorial
-argument: we capture these two notions with $\DProd{}{}$ and $\DId$,
-respectively.
-
-In this finer-grained presentation, we can define $\DIndx$ and
-$\DHindx$ as follow:
-
-\[\begin{array}{l@{\:\mapsto\:\:}l}
-\DIndx{D}         & \DProd{\DId}{D}                      \\
-\DHindx{H}{D}     & \DProd{(\DPi{H}{(\LAM{\_} \DId)})}{D}
-\end{array}
-\]
-
-Consequently, the examples developed above can be straightforwardly
-translated into this new presentation. \note{We might want to re-do
-  the Nat example here, as we encode once again in $\IDesc$} In the
-following, we adopt this last version as our de facto universe of
-inductive types. In particular, we are going to demonstrate several
-instances of generic programming.
-
-\note{Shall we talk about the Type Theory being Desc Zero? or such story?}
-
-\begin{figure*}
-
-\[
-\begin{array}{ll}
-\stk{
-\data \Desc : \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
-    \DId            & \Desc                                   \\
-    \DUnit          & \Desc                                   \\
-    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc         \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
-    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
-\end{array}
-}
-&
-\stk{
-\descop{\_\:}{} : \Desc \To \Set \To \Set \\
-\begin{array}{@{}l@{\:=\:\:}ll}
-\descop{\DId}{X}          &  X                                           \\
-\descop{\DUnit}{X}        &  \Unit                                       \\
-\descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
-\descop{\DSigma{S}{D}}{X} &  \SIGMA{\V{s}}{S} \descop{D\: s}{X}                \\
-\descop{\DPi{S}{D}}{X}    &  \PI{\V{s}}{S} \descop{D\: s}{X}            
-\end{array}
-}
-\end{array}
-\]
-
-\caption{Factorised universe of descriptions}
-\label{fig:factor-desc}
-
-\end{figure*}
-
 \subsection{The generic catamorphism}

### 2010-03-17 papers/icfp-2010-desc/paper_idesc.tex
IDesc@ICFP: better intro to Fin's constrained presentation
IDesc@ICFP: move Desc factorization to IDesc. Give a new reason for doing so.

@@
 universe. We express inductive families by the $I \To \IDesc{I}$ type.
+
+\subsection{Desc, atomically}
+
+\begin{wstructure}
+<- Adding hindx have introduced some duplication
+    <- indx == hindx 1
+    -> We can factor out commonalities 
+        /> Obtain an equivalent presentation
+        /> Still embeddable (refer to the Agda model)
+\end{wstructure}
+
+Before moving on indexed descriptions, we have to carry out some
+maintenance work on descriptions. We presented $\Desc$ as the grammar
+of inductive types. Hence, the codes closely follow this grammar. In
+the following, we adopt an alternative presentation. With $\DSigma$,
+we are already quoting a standard type-former. In the alternative
+presentation, we go further and entirely present our codes as
+quotation of standard type-formers. This presentation is shown in
+Figure~\ref{fig:type-former-desc}.
+
+This reorganisation is strictly equivalent to the previous one
+(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
+self-descriptive. We refer the reader to the companion technical
+report for details. In this finer-grained presentation, we can define
+$\DIndx$ and $\DHindx$ as follow:
+
+\[\begin{array}{l@{\:\mapsto\:\:}l}
+\DIndx{D}         & \DProd{\DId}{D}                      \\
+\DHindx{H}{D}     & \DProd{(\DPi{H}{(\LAM{\_} \DId)})}{D}
+\end{array}
+\]
+
+Consequently, the examples previously developed can be
+straightforwardly translated into this new presentation. \note{We
+  might want to re-do the Nat example here, as we encode once again in
+  $\IDesc$} In the following, we adopt this last version as our de
+facto universe of inductive types. In particular, we are going to
+evolve this presentation into an indexed one.
+
+\note{Shall we talk about the Type Theory being Desc Zero? or such story?}
+
+\begin{figure*}
+
+\[
+\begin{array}{ll}
+\stk{
+\data \Desc : \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+    \DId            & \Desc                                   \\
+    \DUnit          & \Desc                                   \\
+    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc         \\
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
+    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
+\end{array}
+}
+&
+\stk{
+\descop{\_\:}{} : \Desc \To \Set \To \Set \\
+\begin{array}{@{}l@{\:=\:\:}ll}
+\descop{\DId}{X}          &  X                                           \\
+\descop{\DUnit}{X}        &  \Unit                                       \\
+\descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
+\descop{\DSigma{S}{D}}{X} &  \SIGMA{\V{s}}{S} \descop{D\: s}{X}                \\
+\descop{\DPi{S}{D}}{X}    &  \PI{\V{s}}{S} \descop{D\: s}{X}            
+\end{array}
+}
+\end{array}
+\]
+
+\caption{Universe of descriptions based on Type-formers}
+\label{fig:type-former-desc}
+
+\end{figure*}
 
@@
 
-In order to code this data-type, we have to appeal to equality:
+One way to code constrained data-type is to appeal to equality. The
+constraints are therefore captured by equations in the data-type. In
+this case, we obtain the following definition:
 

### 2010-03-17 papers/icfp-2010-desc/paper_desc_levitation.tex
IDesc@ICFP: clarify switchD elaboration.

@@
 theory. Indeed, it would be inconvenient for the user to have to
-decide between two variants. In fact, we give no choice to the user,
-hence no inconvenience. Just as the $\F{switch}$ operator was sealed
-off by the type propagation machinery, $\F{switchD}$ is not provided
-to the user. We simply extend the type-checking rules
+decide between two variants. The solution to this problem stands, once
+again, in the type propagation system. Just as the $\F{switch}$
+operator, $\F{switchD}$ can be automatically elaborated. To this end,
+we simply extend the type-checking rules
 (Fig.~\ref{fig:type-checking}) with the following inference rule:

### 2010-03-17 papers/icfp-2010-desc/paper_type_theory.tex
IDesc@ICFP: clarify our functional-like presentation of eliminators.

@@
 manipulate the $\EnumU$ and $\EnumT{}$ objects, we need to equip our
-type theory with an appropriate \emph{elimination principle}. A first
-step consists in defining the $\spi{}{}$ operator:
+type theory with an appropriate \emph{elimination
+  principle}. Formally, as we add types for data -- such as finite
+sets here --, we need such primitive operators to compute with those
+data, and corresponding extension of the judgemental equality with
+their computational behavior. For compactness and readability, we
+shall present these operators as functional programs, much as we
+implement them in our Agda model. Hence, we first define the
+$\spi{}{}$ operator:
 

### 2010-03-17 papers/icfp-2010-desc/paper_idesc.tex
IDesc@ICFP: typos.

@@
 \emph{inductive families need not store their
-  indices}~\cite{brady:index-inductive-families}. Applying the
+  indices}~\cite{brady:index-inductive-families}\note{ I'm not
+  familiar with this, I need to get into the paper. }. Applying the
 \emph{forcing} optimisation to our definition of $\Fin$ gives the
 following, equivalent definition:
-
-\note{ I'm not familiar with this, I need to get into the paper. }
 
@@
 corresponds to the closed expression, $\HExprD$. Similarly,
-$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{ty})}$$
+$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{dom}{ty})}$$
 corresponds to expressions with variables, $\HExprVarD$.

### 2010-03-17 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: general review

@@
 
-In this section, we describe an alternative implementation of
-inductive types, as we know them in simply-typed languages such as
-Haskell or OCaml. This choice is motivated by pedagogical
-considerations: by describing the layman data-types in our setting, we
-build on our intuition of data-types we are familiar with. However,
-our goal is and remains to discuss data-types as we find them in
-dependently-typed languages.
+In this section, we describe an implementation of inductive types, as
+we know them in simply-typed languages such as Haskell or OCaml. This
+choice is motivated by pedagogical considerations: by describing the
+layman data-types in our setting, we build on our intuition of
+data-types we are familiar with. However, our goal is and remains to
+discuss data-types as we find them in dependently-typed languages.
 
@@
 diverge from induction-recursion. We shall compare both approaches in
-Section~\ref{sec:discussion}
+Section~\ref{sec:discussion}.
 
@@
 
-In dependently-typed language, $\Sigma$-types can be read in two
-rather different ways. This duality is actually reflected in the
-notation we can find in the literature, depending on the sensibility
-of the author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
+In dependently-typed language, $\Sigma$-types can be interpreted in
+two different ways. This duality is actually reflected in the notation
+we can find in the literature, depending on the sensibility of the
+author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
 $\Sigma_{x : A} (B\: x)$. This notation stresses the fact that
@@
 In the simply-typed world, the essence of data-type definition can be
-summed up by the term ``sum-of-product''. A data-type is defined by a
-finite sum -- a choice -- of constructors, each of them is composed by
-a product -- a tuple -- of arguments. Therefore, the grammar of
-data-types definition in simply-typed languages, such as Haskell or
-OCaml, follow this form.
+summed up by the expression ``sum-of-product''. A data-type is defined
+by a finite sum -- a choice -- of constructors, each of them is
+composed by a product -- a tuple -- of arguments. This form governs
+the grammar of data-type definition in languages such as Haskell or
+OCaml.
 
-To model these data-types, we are simply left with capturing this
-grammar in a dependently-typed setting. Hence, the notion of
-``sum-of-product'' naturally translates into
-\emph{sigmas-of-sigmas}. However, beyond this intuition, it should be
-clear that a direct encoding of data-types through raw $\Sigma$-types
-is not a viable option. Indeed, just as for finite sets, encoding
-throw away information when we crucially need it. In the realm of
-data-types, things become even more tougher, as one could legitimately
-expect an induction principle, for instance. Anonymous in a see of
-$\Sigma$s, our data-types have little chance of survival.
+To model these data-types, we simply need to capture this grammar in
+our dependently-typed setting. Here, the notion of ``sum-of-product''
+naturally translates into \emph{sigmas-of-sigmas}. However, beyond
+this intuition, it should be clear that a direct encoding of
+data-types through raw $\Sigma$-types is not a viable option. Indeed,
+just as for finite sets, encoding throw away information when we
+crucially need it. In the realm of data-types, things become even more
+tougher, as one could legitimately expect an induction principle, for
+instance. Anonymous in a see of $\Sigma$s, our data-types have little
+chance of survival. \note{Ridiculous tentative of poetry. I was
+  really tired when I wrote that.}
 
@@
 
-For our sigma-of-sigmas to be known, we crucially need the ability to
-name them. To this end, we will use a standard technique in
-dependently-typed programming: we construct an ad-hoc universe. This
-technique dates back to Martin-L\"of definition of type
+For our sigmas-of-sigmas to be known, we crucially need the ability to
+name them. \note{Pursuing this crappy analogy. Solving the type
+  presentation/representation debate would help solving that point.}
+To this end, we will use a standard technique in dependently-typed
+programming: we construct an ad-hoc universe. This technique dates
+back to Martin-L\"of definition of type
 theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
@@
 \note{I would need some other example of universe construction for
-  practical purposes}. We refer the reader to Agda's
+  practical purposes}. We refer the reader to the Agda
 tutorial~\cite{norell:agda-tutorial} for a pedagogical presentation of
-universe construction.
+universe construction. \note{Is that really a motivating motivation
+  for the usage of universe construction? }
 
@@
 these codes back into the standard type theory. Hence, codes act only
-as labels, while the type theory provides the computational machinery.
+as labels, while the type theory provides the computational
+machinery. \note{Is that a good description?}
 
 Codes being simple labels, we have the ability to inspect them, hence
-taking advantage of their structural information. Being able to
-inspect them, we are therefore free to compute over them: deriving new
-codes from previous ones, or even new functions on them. In several
+taking advantage of structural information. Being able to inspect
+them, we are therefore free to compute over them: deriving new codes
+from previous ones, or even new functions on them. In several
 occasions, we will have the opportunity to witness the power of
 universes.
-
-\note{Is that really a motivating motivation for the usage of universe
-  construction? }
 
@@
 
-Hence, we propose to embed inductive types as a universe in our
-dependent type theory, the universe of \emph{descriptions}
+We propose to embed inductive types as a universe in our dependent
+type theory, the universe of \emph{descriptions}
 (Figure~\ref{fig:desc_universe}). As expected, the code of this
@@
 
-Let us explain its construction. First, we use $\DSigma{}{}$ to build
-a sum between $\NatZero$ and $\NatSuc{}$. In the $\NatZero$ case, we
-reach the end of the description: there is no useful payload in that
-case. In the $\NatSuc{}$ case, we left one hole open for the recursive
-argument, and we close the description.
+Let us explain its construction. First, we use $\DSigma{}{}$ to give a
+choice between the $\NatZero$ and $\NatSuc{}$ constructors. In the
+$\NatZero$ case, we reach the end of the description: there is no
+useful payload in that case. In the $\NatSuc{}$ case, we left one hole
+open for the recursive argument, and we close the description.
 
-In a more synthetic notation, we have simply implemented the following
-functor:
+In a more synthetic notation, we have described the following functor:
 
@@
 interpretation: we pack an element of $X$ together with the recursive
-argument. Easily, one sees that this code actually defines the
+argument. Easily, one sees that this code actually describes the
 following functor:
@@
 description. Formally, we have:
-
 \note{We have talked about a tagged description based on $\List (\UId
@@
 we have a finite sum of constructors. Of course, every description can
-be expressed into this style, using a singleton as single constructor
-if necessary.
-
+be expressed in this style, using a singleton as single constructor if
+necessary.
 \note{I think we need a function mapping a $\TagDesc$ to the
@@
 class of functors expressible with descriptions enjoy another
-property: they all admit a fix-point.
+property: they all admit a fix-point. Indeed, descriptions are
+interpreted into \emph{polynomial functors}~\cite{who?}\note{who is
+  the relevant source for polynomial functors?}, as they are only
+composed of $\Sigma$-types, the unit type, and variables.
 
-\note{ I'm afraid that this section is killed by my categorical gibberish. }
-
-Let us constructively prove this. First of all, for an object $D$
-describing a signature functor, we define its fix-point by the type
-former $\Mu{D}$. Inhabitants of the fix-point are materialised by the
-structure map: $\Con{x}$ inhabits $\Mu{D}$ if $x$ belongs to the
-interpretation of $D$ with the carrier of the initial algebra:
+We construct the fix-point by \emph{tying the knot} of the
+interpretation: the holes left open by the functor are filled by its
+own recursive definition:
 
@@
 
-For the less categorically inclined reader, we can simply read this
-construction as \emph{tying the knot}: the holes left open by the
-functor are filled by its own recursive definition. Hence, we have
-recovered the data-types we are usually able to define in simply-typed
-languages.
+At this stage, we have recovered the data-types we are usually able to
+define in simply-typed languages.
 
@@
 Together with the type formers defined above, we would legibly expect
-an elimination principle. Following the categorical intuition, we
-could be tempted to provide it as a catamorphism:
+an elimination principle. Following a categorical intuition, we could
+be tempted to provide it as a catamorphism:
 
@@
 it comes short of its promises in the dependently-typed realm. Indeed,
-this type is absolutely non dependent. As this operator will be the
-corner stone of any inductive definition, we are better off sharping
-it before use.
+this type is not dependent. As this operator will be the corner stone
+of any inductive definition, we are better off sharping it before use.
 
@@
 $\F{induction}$ principle. For space reason, we refer the reader to
-our technical report for the actual code, hence the proof of (weak)
-initiality. Having this dependent catamorphism, we are empowered with
-the ability to $\emph{compute}$ over descriptions. Note that this
-induction principle is the first manifestation of a generic operation
-over descriptions. Thanks to $\F{induction}$, any data-type we define
-automatically comes equipped with an induction principle.
+our technical report for the actual code. Having this dependent
+catamorphism, we are empowered with the ability to $\emph{compute}$
+over descriptions. Note that this induction principle is the first
+manifestation of a generic operation over descriptions. Thanks to
+$\F{induction}$, any data-type we define automatically comes equipped
+with an induction principle.
+
+\subsection{Extending type propagation}
 
@@
 
-In this section, we have introduced a universe of description. We have
-shown how to manipulate these codes to build data-types. We have also
+So far, we have introduced a universe of description. We have shown
+how to manipulate these codes to build data-types. We have also
 presented a fix-point construction and its associated dependent
@@
 this section, we have seen how data-types from the simply-typed world
-can be expressed in our dependent-type setting. By relying on a
+can be expressed in our dependently-typed setting. By relying on a
 universe construction technique, we benefit from the ability to

### 2010-03-17 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: general review

@@
 was to effectively implement the code of the $\Desc$ universe. Another
-promise we tacitly made is to implement the type former of finite
+promise we tacitly was the existence of the type formers for finite
 sets. Despite being a perilous pedagogical exercise for the authors,
@@
 
-In Section~\ref{sec:finite-sets}, we have promised an implementation
-of finite sets. It consists in the objects specified in
-Figure~\ref{fig:typing-finite-set}. In particular, we will now
-implement the $\EnumU$ type former and its constructors. Let us recall
-their specification:
+In Section~\ref{sec:finite-sets}, we have given a specification of
+finite sets. It consists in the objects presented in
+Figure~\ref{fig:typing-finite-set}. We are going to implement the
+$\EnumU$ type former and its constructors. Let us recall their
+specification:
 
@@
 The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
-coincidence. $\EnumU$ is indeed simply isomorphic to a list, with
-$\UId$s as parameter. However, in Section~\ref{sec:desc-examples}, we
-have shown how to build lists in the universe of
-description. Therefore, the actual implementation of $\EnumU$ sums up
-to the following, trivial definition:
+coincidence. $\EnumU$ is indeed isomorphic to a list of $\UId$s. In
+Section~\ref{sec:desc-examples}, we have shown how to build lists in
+the universe of description. Therefore, the actual implementation of
+$\EnumU$ boils down to the following, trivial definition:
 
@@
 operation in Epigram, moving from an hard-coded presentation to a
-self-hosted one. Absolutely no changes to the $\EnumT$ world was
+self-hosted one. Absolutely no change to the $\EnumT$ objects was
 required.
@@
 are still there. Conversely, we have not introduced any spurious
-objects in the type theory.
+object.
 
@@
 
-We shall now fulfil our promises. We are going to implement the codes
-of the universe of descriptions. Interestingly, in and by itself, a
-code is nothing but a data-type. We are in the same situation than
-with $\EnumU$: we ought to be able to describe the codes of $\Desc$ in
-$\Desc$ itself. Hence, this code would be a first-class citizen, born
-with the standard, generic equipment of data-types.
+We shall now fulfil our promises for $\Desc$. We are going to
+implement the codes of the universe of descriptions. Interestingly, in
+and by itself, a code is nothing but a data-type. We are in the same
+situation than with $\EnumU$: we ought to be able to describe the
+codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
+first-class citizen, born with the standard, generic equipment of
+data-types.
 
@@
 the data-types we have seen so far, we first have to choose a
-constructor. Constructors correspond to the code's cases: $\DUnit$,
-$\DSigma$, and $\DIndx$. The reader will notice that the tagged
-notation we have used so far for $\Desc$ constructors now fully make
-sense: these were actually the tags we are defining. \note{This
+constructor. Constructors correspond to the code's constructors:
+$\DUnit$, $\DSigma$, and $\DIndx$. The reader will notice that the
+tagged notation we have used so far for $\Desc$ constructors now fully
+make sense: these were actually the tags we are defining. \note{This
   sentence badly needs a rephrasing.}
@@
 naturally: it only takes a recursive argument -- a description. The
-case of $\DSigma$ is problematic. Recall the definition of $\DSigma$:
+case of $\DSigma$ is problematic. Recall the specification of
+$\DSigma$:
 
@@
 object gives us $\Desc$, up to isomorphism. We have implemented our
-code of data-types as a levitating object inside itself. However, this
-levitation operation, just as any magic trick, relies on an invisible
-cable. Let us reveal it, hence finishing our implementation.
+code of data-types as an object levitating inside itself. However,
+this levitation operation, just as any magic trick, relies on an
+invisible cable. Let us reveal it, hence finishing our implementation.
 
@@
 (Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by
-careful usage of type propagation. Indeed, let us elaborate this term
-down to the low-level type theory. The $\EnumT [ \ldots ]$ construct
-elaborates to a finite set $e$ in $\EnumU$, inhabited by the
+careful usage of type propagation. Indeed, let us try to elaborate
+this term down to the low-level type theory. The $\EnumT [ \ldots ]$
+construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
 codes. Then, we can type-check the case definition, between square
@@
 $\EnumT{e} \To \Desc$. This corresponds to a finite function
-definition we have implemented in Figure~\ref{fig:type-checking}.
+definition, as formalised in Figure~\ref{fig:type-checking}.
 
@@
 \[
-\PLAM{x}{(\EnumT{e})} \switch{e}{\LAM{\_} \Mu{\DescD}}{\pi^f}{x}
+\PLAM{x}{(\EnumT{e})} \switch{e}{(\LAM{\_} \Mu{\DescD})}{\pi^f}{x}
 \]
 
-But this is rather problematic! We are still in the process of
+But this is quite problematic. We are still in the process of
 constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
-begging for an $\eta$-expanded $\Desc$. Despite our willingness, we
-will never be able to materialise such motive.
+begging for this very same $\DescD$. Despite our willingness, we
+cannot materialise such motive.
 
@@
 theory. Indeed, it would be inconvenient for the user to have to
-decide between two variants. The solution to this problem stands, once
-again, in the type propagation system. Just as the $\F{switch}$
+decide between two variants. The solution to this problem is, once
+again, to use the type propagation system. Just as the $\F{switch}$
 operator, $\F{switchD}$ can be automatically elaborated. To this end,
-we simply extend the type-checking rules
-(Fig.~\ref{fig:type-checking}) with the following inference rule:
+we extend the type-checking rules (Fig.~\ref{fig:type-checking}) with
+the following inference rule:
 
@@
                                {\\ \PLAM{x}{(\EnumT{e})} \switchD{e}{t'}{x}}
-                    \end{array}}\;\mbox{t is $[]$ or $[a,b]$}
+                    \end{array}
+     }\;\mbox{t is $[]$ or $[a,b]$}
 \]
@@
 reveals that the $\Desc$ universe is just plain data. Just as any
-piece of data, it can therefore be inspected and manipulated. However,
-unlike any data, it is expressed in the $\Desc$ universe. As a
+piece of data, it can therefore be inspected and
+manipulated. Moreover, it is expressed in the $\Desc$ universe. As a
 consequence, it is equipped, for free, with an induction
@@
 equipment in the target language to \emph{program} over data-types. In
-this setting, generic programming is just about \emph{programming}.
+this setting, \emph{generic programming is just about programming}.
 
@@
 node $xs$ and the induction hypotheses, the method ought to build an
-element of $T$. Provided we know how to make an element of
+element of $T$. Provided that we know how to make an element of
 $\descop{D}{T}$, this step will be performed by the algebra $f$. Let
@@
 basically, a function mapping a $\Desc$ to a data-type specific
-operation. Having the ability to program on $\Desc$ is the key to
-define such function. The induction principle of $\Desc$ -- inherited
-by levitation -- plays a central role here: without an induction
-principle, it would be impossible to program on $\Desc$.
+operation.
 
@@
 \newcommand{\FMFreeMonad}{\D{FreeMonad}}
-\newcommand{\FMFreeMonadF}{\D{FreeMonad$_{\CN{Free}}$}}
+\newcommand{\FMFreeMonadD}{\D{FreeMonadD}}
 \newcommand{\FMVar}{\C{Var}}
@@
 \[
-\FMFreeMonadF\: F\: X\: Z \mapsto X + F Z
+\FMFreeMonadD\: F\: X\: Z \mapsto X + F Z
 \]
@@
 description, a set $X$ of variables, and will compute the tagged
-description of the corresponding free monad.
-
-Implementing this function is surprisingly easy:
+description of the corresponding free monad. Implementing this
+function is surprisingly easy:
 
@@
 a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
-as they were, hence leaving the functor unchanged. The interpretation
-of this definition convince us that we have built the functor
-signature of the free monad. The fix-point operation ties the knot and
-gives us the full-blown free monad construction.
+as they were, hence leaving the functor unchanged. Unfolding the
+interpretation of this definition, we convince ourself that this
+corresponds to the functor $\FMFreeMonadD$. The fix-point operation
+ties the knot and gives us the full-blown free monad construction.
 
@@
 Our task is therefore to implement the algebra of the
-catamorphism. Intuitively, its role is to catch appearance of
-$\DVar{x}$, to replace it by $\sigma x$. Otherwise, it should simply
+catamorphism. Intuitively, its role is to catch appearances of
+$\DVar{x}$ and replace them by $\sigma x$. Otherwise, it should simply
 appeal to the induction hypothesis. This corresponds to the following
@@
 
-Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$ finishes
-the implementation. To sum up, we have implemented the free monad
-construction for an arbitrary tagged description. This gives our user
-the ability, for any data-type, to extend it with a notion of
-variable. Then, we have equipped this structure with the
-corresponding monadic operation, ``bind'' and ``return''. Whereas the
-``return'' is a trivial variable former, the ``bind'' operation
-corresponds to the substitution of variables by terms. We have shown
-how to implement it, for any free monad.
+Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$
+completes the implementation. To sum up, we have implemented the free
+monad construction for an arbitrary tagged description. This gives our
+user the ability, for any data-type, to extend it with a notion of
+variable. Then, we have equipped this structure with the corresponding
+monadic operation, ``bind'' and ``return''. Whereas the ``return'' is
+a trivial variable former, the ``bind'' operation corresponds to the
+substitution of variables by terms. We have shown how to implement it,
+for any free monad.
 

### 2010-03-17 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: general review

@@
 
-While inductive types are the alpha and omega of the simply-typed
+While inductive types are the alpha and omega of simply-typed
 languages, evolving in a dependent setting fosters new
@@
 the following, we adopt an alternative presentation. With $\DSigma$,
-we are already quoting a standard type-former. In the alternative
-presentation, we go further and entirely present our codes as
-quotation of standard type-formers. This presentation is shown in
-Figure~\ref{fig:type-former-desc}.
+we are actually \emph{quoting} a standard type-former, namely
+
+$$\Bhab{\Sigma}{\PI{\V{S}}{\Set} (S \To \Set) \To \Set}$$
+
+In the alternative presentation, we go further and present all our
+codes as quotations of standard type-formers. This presentation is
+shown in Figure~\ref{fig:type-former-desc}.
 
@@
 by $\Bhab{\V{I}}{\Set}$, we rely on functors in the category
-$\Set^I$. We call these \emph{indexed functors}. When $\Desc$ was
-defining a grammar a functor, we need $I \To \IDesc{I}$ to define a
-grammar of indexed functors:
+$\Set^I$. We call these \emph{indexed functors}. In the following, we
+implement $I \To \IDesc{I}$ as a grammar for indexed functors:
 
@@
 This generalisation can be achieved by a minor evolution of $\Desc$
-(Fig.~\ref{fig:factor-desc}). Indeed, the functorial nature of $\Desc$
+(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of $\Desc$
 is entirely captured by the $\DId$ code. We get an indexed functor
@@
 giving more opportunities to the type-system, hence reducing the
-burden on the programmer in practice.
+burden on the programmer.
 
@@
 
+\subsection{Tagged indexed descriptions}
+
 \begin{wstructure}
@@
 form for descriptions. This format follows the usual presentation of
-inductive types as sums of product. With indexed description, this
+inductive types as sum-of-product. With indexed description, this
 definition can be generalized. Indeed, when defining an indexed
@@
 \[\stk{
+ \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}}
  \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
  \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})})  \\
- \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}}
 }\]
 
-In the case of vectors, for instance, for the index $n = \NatZero$, we
-would only offer the constructor $\ListNil$. Similarly, for $n =
-\NatSuc{n'}$, we only offer the constructor $\ListCons{}{}$. We will
+In the case of vectors, for instance, for the index $\NatZero$, we
+would only propose the constructor $\ListNil$. Similarly, for
+$\NatSuc{n}$, we only propose the constructor $\ListCons{}{}$. We will
 see more instances of this pattern in the following.
@@
 the host language. Further, we also assume the existence of an
-addition and comparison operator in $\Nat$, respectively name
+addition and comparison operator in $\Nat$, respectively named
 $\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$ as follow:
@@
 catamorphism. Indeed, implementing a single step of evaluation -- the
-algebra -- is sufficient, as $\F{cataI}$ gives, for free, the
-full evaluator. The implementation is as follow:
+algebra -- is sufficient, as $\F{cataI}$ \note{We have not defined
+  \F{cataI}!} gives, for free, the full evaluator. The implementation
+is as follow:
 
@@
 
-\note{We have not defined \F{cataI}!}
-
 \begin{wstructure}
@@
 However, so far, we are only able to define and manipulate
-\emph{closed} terms. By abstracting over $\Var{}$, it is possible to
+\emph{closed} terms. By abstracting over $\Val{}$, it is possible to
 build and manipulate \emph{open} terms, that is terms with symbolic
@@
 $\Var{}{}$ maps types to a finite set. The finite set -- the context
--- contains values. A variable is therefore a $\Eval{}$ that contains
-a pointer to a particular element of the finite set -- an element of
-$\EnumT{dom}$. The extra argument to $\Var{}{}$ is the domain of this
-context.
+-- contains closed terms. A variable is therefore a $\Eval{}$ that
+contains a pointer to a particular element of the finite set -- an
+element of $\EnumT{dom}$. The extra argument to $\Var{}{}$ is the
+domain of this context.
 
@@
 Again, we would like to give a semantics to this extended language. We
-proceed in two steps: first, replace the variables by their value in
-the context; then, evaluate the resulting closed term. Thanks to
+proceed in two steps: first, we replace the variables by their value
+in the context; then, we evaluate the resulting closed term. Thanks to
 $\evalH$, we are already able to solve the second problem. Let us
@@
 first-order syntax with variables: our data-type enjoys more structure
-than what we are currently provided. In particular, this is
-reminiscent to a free monad, where $\Eval{}$ is the ``return''
-introducing variables. Instead of implementing $\substH$ in this
-special case, we shall implement it in a generic setting, and express
-our example in that setting.
+than what we are given. In particular, this is reminiscent to a free
+monad, where $\Eval{}$ is the ``return'' introducing
+variables. Instead of implementing $\substH$ in this special case, we
+shall implement it in a generic setting, and express our example in
+that setting. However, before entering the realm of generic
+structures, let us further explore the universe of indexed
+descriptions.
 
@@
 
-However, before entering the realm of generic structures, let us
-further explore the universe of indexed descriptions. In the
-following, we ask $I$ to come equipped with a notion of definitional
-equality. Hence, we can legibly equates elements of $I$.
+In the following, we ask $I$ to come equipped with a notion of
+definitional equality. Hence, we can legibly equates elements of $I$.
 
@@
 
-In our implementation of lambda terms, variables were elements of
-$\Fin{n}$, with $n$ the number of binders. Let us now implement
-$\Fin$. $\Fin$ is a typical example of an inductive family using
+Another common dependent data-type is $\Fin$, the finite sets of
+numbers. $\Fin$ is a typical example of an inductive family using
 \emph{constrained constructors}. The constructors $\FinZero$ and
@@
 The $\Fin$ example is also a well-known example of Generalised
-Abstract Data-Type (GADT), in Haskell. It comes as no surprise that
-our universe of indexed description supports this restricted class of
+Abstract Data-Type (GADT) in Haskell. It comes as no surprise that our
+universe of indexed description supports this restricted class of
 dependent data-types. As we have shown with $\Fin$, constrained
@@
 of future work to see if constraints can be entirely eradicated, or
-presented more conveniently to the user.
-
-Finally, it is worth mentioning that the forcing optimisation is a
-\emph{source-to-source} transformation of the description.
+presented more conveniently to the user. Finally, it is worth
+mentioning that the forcing optimisation is a \emph{source-to-source}
+transformation of the description.
 
@@
 
-In this section, we develop some more structure for our indexed
+In this section, we develop more structure for our indexed
 data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
@@
 In Section~\ref{sec:idesc-examples}, we had the intuition that our
-datatype $\HExprD$ and $\HExprVarD$ enjoys a monadic structure. We had
+datatypes $\HExprD$ and $\HExprVarD$ enjoy a monadic structure. We had
 identified the variable substitution operation as the ``bind'' of a
 free monad. The definition of $\substI$ above confirm our intuition:
-using $substI$, we should easily obtain $\substH$. To do so, we first
+using $\substI$, we should easily obtain $\substH$. To do so, we first
 have to massage the definition of our data-type, to exhibit its
@@
 of definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
-corresponds to the closed expression, $\HExprD$. Similarly,
+corresponds to the syntax of closed terms, $\HExprD$. Similarly,
 $$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{dom}{ty})}$$
@@
 Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
-terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$ to values in
+terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}\: ty$ to values in
 $\Val{ty}$. We are therefore left with implementing $\substH$. We
@@
 nothing but the ``return''. The remaining constructors are simply the
-freed functor, trivially indexed by $\Unit$. Hence, amusingly, we can
-define the $\IDesc$ freed functor by the following code:
+signature functor, trivially indexed by $\Unit$. Hence, we describe
+this signature functor by the following code:
 

### 2010-03-17 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: general review

@@
 choice. In particular, our system has been modelled in Agda, which
-features an in-tensional equality. On the other hand, it is used in
+features an intentional equality. On the other hand, it is used in
 Epigram, which equality has a slightly extensional
@@
 Finally, we define the rules governing judgemental equality in
-Figure~\ref{fig:judgemental-equality}. As one would expect,
+Figure~\ref{fig:judgemental-equality}. We implicitly assume that
 judgemental equality respects symmetry, reflexivity, and
-transitivity. Further, it encompasses the $\beta$-rules for function
-application and pairs. Finally, it respects structural equality,
-including equality under lambda ($\xi$-rule).
+transitivity. We capture the computational behavior of the language
+through the $\beta$-rules for function application and pairs. Finally,
+we implicitly assume that it respects purely syntactic and structural
+equality. This includes equality under lambda ($\xi$-rule).
 
@@
 freedom is a key point in favour of the wide applicability of our
-proposal. Obviously, when implementing our technology in an existing
+proposal. This judgemental presentation must be read as a
+\emph{specification}: our proposal works with any propositional
+equality satisfying this specification. Moreover, our lightweight
+requirements do not endanger decidability of equality-checking.
+Obviously, when implementing our technology in an existing
 type-theory, some opportunities arise. We will present some of them
-along the course of the paper. \note{It is just a specification. We
-  expect the propositional theory to provide a decidable machinery to
-  implement it.}
-
-\note{plus structural rules, including lambda ($\xi$-rule)}
+along the course of the paper.
 
@@
 The specification of finite sets is presented in
-Figure~\ref{fig:typing-finite-set}. It is composed by three
+Figure~\ref{fig:typing-finite-set}. It is composed of three
 components. First, we define tags as inhabitants of the $\UId$ type. A
-tag is solely an informative token, used for display purposes. Finite
-sets inhabits the $\EnumU$ type. Unfolding the definition, we get that
-a finite set is a list of tags. Finally, elements of a finite set $u$
-belong to the corresponding $\EnumT{u}$ type. Intuitively, it
-corresponds to an index, a number, pointing to an element of $u$.
+tag is solely an informative token, used for diagnostic
+purposes. Finite sets inhabits the $\EnumU$ type. Unfolding the
+definition, we get that a finite set is a list of tags. Finally,
+elements of a finite set $u$ belong to the corresponding $\EnumT{u}$
+type. Intuitively, it corresponds to an index -- a number -- pointing
+to an element of $u$.
 
@@
 }\]
-
-\note{Needs to be defined.}
 
@@
 Our proposal to tame the problem of codes is deeply rooted in the
-bidirectional presentation of type-checking of Pierce and
+bidirectional presentation of type-checking from Pierce and
 Turner~\cite{pierce:bidirectional-tc}. In this formalism,
@@
 richer type theory. First of all, we define a high-level syntax,
-composed by \emph{expressions}. Expressions divided into two syntactic
-categories: $\exprIn$ into which types are pushed in, and $\exprEx$
-from which types are pulled out. Following the bidirectional
+composed by \emph{expressions}. Expressions are divided into two
+syntactic categories: $\exprIn$ into which types are pushed in, and
+$\exprEx$ from which types are pulled out. Following the bidirectional
 presentation, $\exprIn$ are subject to type \emph{checking}, while
@@
 the inference rules of the type theory, we will ignore the judgements
-which are identical. We refer our reader to the associated technical
+that are identical. We refer our reader to the associated technical
 report~\cite{chapman:desc-tech-report} for the complete system of
@@
 the case of a variable in context. This also include function
-application as well as projections. Further down the line, these types
-might propagate to other terms, thanks to type checking.
+application as well as projections. Further down the line, these
+inferred types might propagate to other terms, thanks to type
+checking.
 
@@
 abstract away the details of code and use a familiar syntax
-instead. Hence, we use the syntax $\Enum\: [t_1 \ldots t_n]$ to denote
-the finite sets composed by the named elements $t_1$, \ldots,
-$t_n$. The translation to elements of $\EnumU$ is straightforward.
+instead. Hence, we use the syntax $[t_1 \ldots t_n]$ to denote the
+finite sets composed by the named elements $t_1$, \ldots, $t_n$. The
+translation to elements of $\EnumU$ is straightforward.
 
@@
 low-level terms into high-level expressions. Hence, this system, which
-presentation is out of the scope of this paper, seals the low-level
-type theory behind the high-level expressions.
+complete presentation is out of the scope of this paper, seals the
+low-level type theory behind the high-level expressions.
 
@@
 
-In this section, we have developed a type propagation system, based
-on bidirectional type-checking. Using bidirectional type-checking as a
+In this section, we have developed a type propagation system based on
+bidirectional type-checking. Using bidirectional type-checking as a
 boilerplate scrapper is a well-known technique~\cite{who?} \note{I
@@
 types. We have illustrated our approach with finite sets. We have
-successfully abstracted away the low-level presentation of finite
-sets, offering a convenient syntax instead.
+abstracted away the low-level presentation of finite sets, offering a
+convenient syntax instead.
 

### 2010-03-17 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: close some todos.
Desc@ICFP: more things todo, I'm afraid.
Desc@ICFP: fix dangling label
Desc@ICFP: clean-up \note{}s. Update todo.

@@
 \section{Indexing descriptions}
+\label{sec:indexing-desc}
 
@@
 grammar of data-types. To this end, we rely on \emph{indexing}. We are
-going to transform the $\Desc$ universe into an indexed $\IDesc$
-universe. We express inductive families by the $I \To \IDesc{I}$ type.
+going to transform the $\Desc$ universe into $\IDesc$, the universe of
+indexed descriptions. We express inductive families by the $I \To
+\IDesc{I}$ type.
 
@@
 
-\note{ Actual implementation of forcing as a generic function??! }
-
 \subsection{Free IMonad}

### 2010-03-17 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: more things todo, I'm afraid.
Desc@ICFP: clean-up \note{}s. Update todo.
Desc@ICFP: apply \toDesc on tagged description.
Desc@ICFP: uniform Fig.~\ref

@@
 \subsection{A universe of descriptions}
+\label{sec:desc-universe}
 
@@
 \end{wstructure}
-
-\note{Check Martin-Lof reference}
 
@@
 be expressed in this style, using a singleton as single constructor if
-necessary.
-\note{I think we need a function mapping a $\TagDesc$ to the
-  corresponding $\Desc$. This function should go here.}
+necessary. A tagged description can trivially be turned into a
+full-blown description. We note $\toDesc{TD}$ the description computed
+from the tagged description $TD$.
 
@@
 setting. This is achieved by the predicate $\All{D}{X}{P}{x}$
-(Fig.\ref{fig:all-predicates}) that, intuitively, states that
+(Fig.~\ref{fig:all-predicates}) that, intuitively, states that
 $\Bhab{P}{X \To \Set}$ holds everywhere in $x$. Based on that

### 2010-03-17 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: clean-up \note{}s. Update todo.
Desc@ICFP: uniform Fig.~\ref

@@
 
-The rules ensuring Context Validity (Fig.\ref{fig:context-validity})
+The rules ensuring Context Validity (Fig.~\ref{fig:context-validity})
 are unsurprisingly common. They simply guarantee the validity of the
@@
 
-The typing judgements (Fig.\ref{fig:typing-judgements}) are just as
+The typing judgements (Fig.~\ref{fig:typing-judgements}) are just as
 familiar. We recognise the standard presentation of $\Pi$ and $\Sigma$
@@
 Dually to type synthesis, type checking judgements
-(Fig.\ref{fig:type-checking}) act as \emph{sinks}. From an expression
+(Fig.~\ref{fig:type-checking}) act as \emph{sinks}. From an expression
 $\exprIn$ and a type propagated down to it, they elaborate a low-level
@@
 
-\note{ This might deserve a more thorough explanation, with snippet of
-       code. }
+\note{This might deserve a more thorough explanation, with snippet of
+  code.}
 

### 2010-03-17 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: clean-up \note{}s. Update todo.
Desc@ICFP: apply \toDesc on tagged description.
Desc@ICFP: uniform Fig.~\ref
Desc@ICFP: uniformize bind and return style

@@
 visible in the definition of $\DescD$
-(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by
+(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
 careful usage of type propagation. Indeed, let us try to elaborate
@@
 
-\note{Update to the factorised presentation}
-
 \[\stk{
@@
 delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
-performs the r\^ole of ``return'', embedding variables into terms. The
-``bind'' operation corresponds to \emph{substitution}. We will now
+performs the r\^ole of \return, embedding variables into terms. The
+\bind operation corresponds to \emph{substitution}. We will now
 implement it, as a generic function.
@@
            \PI{\V{X}, \V{Y}}{\Set} \\
-         & \Mu{(\FreeMonad{D}{X})} \To
-           (\V{X} \To \Mu{(\FreeMonad{D}{Y})}) \To
-           \Mu{(\FreeMonad{D}{Y})} 
+         & \Mu{(\toDesc{\FreeMonad{D}{X}})} \To
+           (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{Y}})}) \To
+           \Mu{(\toDesc{\FreeMonad{D}{Y}})} 
 \end{array} \\
 \subst\: D\: X\: Y\: x\: \sigma \mapsto
-  \cata\: (\FreeMonad{D}{X})\: (\Mu{(\FreeMonad{D}{Y})})\: ???\: x
+  \cata\: (\toDesc{\FreeMonad{D}{X}})\: 
+          (\Mu{(\toDesc{\FreeMonad{D}{Y}})})\: 
+          ???\: 
+          x
 }\]
@@
            \PI{\V{X}, \V{Y}}{\Set} \\
-         & (\V{X} \To \Mu{(\FreeMonad{D}{X})}) \To
-           \descop{\FreeMonad{D}{X}}{\Mu{(\FreeMonad{D}{Y})}} \To
-           \Mu{(\FreeMonad{D}{Y})}                            \\
+         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{X}})}) \To
+           \descop{\toDesc{\FreeMonad{D}{X}}}{\Mu{(\toDesc{\FreeMonad{D}{Y}})}} \To
+           \Mu{(\toDesc{\FreeMonad{D}{Y}})}
+\\
 \end{array} \\
@@
 variable. Then, we have equipped this structure with the corresponding
-monadic operation, ``bind'' and ``return''. Whereas the ``return'' is
-a trivial variable former, the ``bind'' operation corresponds to the
-substitution of variables by terms. We have shown how to implement it,
-for any free monad.
+monadic operation, \bind and \return. Whereas the \return is a trivial
+variable former, the \bind operation corresponds to the substitution
+of variables by terms. We have shown how to implement it, for any free
+monad.
 
@@
 So, secondly, we have equipped this new data-type with its inherent
-structure: the bind and return operations. We have been able to build
-them as generic functions. Here, we rely on our ability to compute
-over descriptions, thanks to the induction principle.
+structure: the \bind and \return operations. We have been able to
+build them as generic functions. Here, we rely on our ability to
+compute over descriptions, thanks to the induction principle.

### 2010-03-17 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix figure of tagged indexed description.
Desc@ICFP: uniformize bind and return style

@@
 \[\stk{
- \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}}
+ \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}} \\
  \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
- \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})})  \\
+ \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})}) 
 }\]
@@
 definition looks familiar. Indeed, it is extremely similar to a
-monadic ``bind''. This is not surprising as we are defining a
-first-order syntax with variables: our data-type enjoys more structure
-than what we are given. In particular, this is reminiscent to a free
-monad, where $\Eval{}$ is the ``return'' introducing
-variables. Instead of implementing $\substH$ in this special case, we
-shall implement it in a generic setting, and express our example in
-that setting. However, before entering the realm of generic
-structures, let us further explore the universe of indexed
-descriptions.
+monadic \bind. This is not surprising as we are defining a first-order
+syntax with variables: our data-type enjoys more structure than what
+we are given. In particular, this is reminiscent to a free monad,
+where $\Eval{}$ is the \return introducing variables. Instead of
+implementing $\substH$ in this special case, we shall implement it in
+a generic setting, and express our example in that setting. However,
+before entering the realm of generic structures, let us further
+explore the universe of indexed descriptions.
 
@@
 Just as in the universe of descriptions, this construction comes with
-an obvious ``return'' and a substitution operation, the ``bind''. Its
-type signature is the following:
+an obvious \return and a substitution operation, the \bind. Its type
+signature is the following:
 
@@
 datatypes $\HExprD$ and $\HExprVarD$ enjoy a monadic structure. We had
-identified the variable substitution operation as the ``bind'' of a
-free monad. The definition of $\substI$ above confirm our intuition:
-using $\substI$, we should easily obtain $\substH$. To do so, we first
-have to massage the definition of our data-type, to exhibit its
-monadic structure.
+identified the variable substitution operation as the \bind of a free
+monad. The definition of $\substI$ above confirm our intuition: using
+$\substI$, we should easily obtain $\substH$. To do so, we first have
+to massage the definition of our data-type, to exhibit its monadic
+structure.
 
-As previously mentioned, we identify $\Eval{}$ as the ``return'' of
-the free monad, while the other components are the action of the
-monad. As a result, the definition is similar to $\HExprD$ presented
-in Figure~\ref{fig:hexpr-full}, at the exception of $\HExprAD$ that is
+As previously mentioned, we identify $\Eval{}$ as the \return of the
+free monad, while the other components are the action of the monad. As
+a result, the definition is similar to $\HExprD$ presented in
+Figure~\ref{fig:hexpr-full}, at the exception of $\HExprAD$ that is
 replaced by $\HExprAFreeD$:
@@
 Another instance of free monad is $\IDesc$ itself. Indeed, $\DVar$ is
-nothing but the ``return''. The remaining constructors are simply the
+nothing but the \return. The remaining constructors are simply the
 signature functor, trivially indexed by $\Unit$. Hence, we describe

### 2010-03-17 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: uniformize spacing in arrays
Desc@ICFP: composed by -> composed of

@@
 \[
-\begin{array}{l@{\;\;\Rightarrow\;\;}l}
+\begin{array}{l@{\;\Rightarrow\;\;}l}
 \Gamma            \vdash \Bhab{t}{T}            
@@
 richer type theory. First of all, we define a high-level syntax,
-composed by \emph{expressions}. Expressions are divided into two
+composed of \emph{expressions}. Expressions are divided into two
 syntactic categories: $\exprIn$ into which types are pushed in, and
@@
 instead. Hence, we use the syntax $[t_1 \ldots t_n]$ to denote the
-finite sets composed by the named elements $t_1$, \ldots, $t_n$. The
+finite sets composed of the named elements $t_1$, \ldots, $t_n$. The
 translation to elements of $\EnumU$ is straightforward.

### 2010-03-17 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: uniformize spacing in arrays
Desc@ICFP: composed by -> composed of
Desc@ICFP: uniformize EnumT definitions
Desc@ICFP: uniformize sum-of-product and sigmas-of-sigmas

@@
 In the simply-typed world, the essence of data-type definition can be
-summed up by the expression ``sum-of-product''. A data-type is defined
+summed up by the expression \emph{sum-of-product}. A data-type is defined
 by a finite sum -- a choice -- of constructors, each of them is
-composed by a product -- a tuple -- of arguments. This form governs
+composed of a product -- a tuple -- of arguments. This form governs
 the grammar of data-type definition in languages such as Haskell or
@@
 To model these data-types, we simply need to capture this grammar in
-our dependently-typed setting. Here, the notion of ``sum-of-product''
+our dependently-typed setting. Here, the notion of sum-of-product
 naturally translates into \emph{sigmas-of-sigmas}. However, beyond
@@
 \data \Desc : \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
     \DUnit          & \Desc \\
@@
 \NatD : \Desc \\
-\NatD \mapsto \DSigma{(\EnumT [ \NatZero, \NatSuc{} ])}
+\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
                      {[ \DUnit \quad (\DIndx{\DUnit}) ]}
@@
 \ListD : \Set \To \Desc \\
-\ListD \: X \mapsto \DSigma{(\EnumT [ \ListNil, \ListCons ])}
+\ListD \: X \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
                            {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{\DUnit}}) ]}
@@
 \begin{array}{@{}ll}
-\TreeD \: X \mapsto \DSigma{ & (\EnumT [ \TreeLeaf, \TreeNode ]) \\}
+\TreeD \: X \mapsto \DSigma{ & (\EnumT{[ \TreeLeaf, \TreeNode ]}) \\}
                            { & [ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
@@
 
-Again, this is a clear reminiscence of the ``sum-of-products'' style:
+Again, this is a clear reminiscence of the sum-of-product style:
 we have a finite sum of constructors. Of course, every description can
@@
 issue of data-type declaration. The grammar of data-types we are used
-to can be straightforwardly translated to the ``sigmas-of-sigmas''
+to can be straightforwardly translated to the sigmas-of-sigmas
 paradigm. Hence, a data-type definition can be turned into the

### 2010-03-18 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: uniformize spacing in arrays
Desc@ICFP: uniformize EnumT definitions
Desc@ICFP: {Free} is exponent, instead of subscript
Desc@ICFP: Fin specified in Agda-like notation
Desc@ICFP: correct presentation of substitution for Hutton's razor. Introduce spurious alignment issues.

@@
 \data \Desc : \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
     \DId            & \Desc                                   \\
@@
 \data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
     \DVar{}         & I \To \IDesc{I}                                   \\
@@
 \NatD : \IDesc{\Unit} \\
-\NatD \mapsto \DSigma{(\EnumT [ \NatZero, \NatSuc{} ])}
+\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
                      {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
@@
 \begin{array}{@{}ll}
-\IDescD~I \mapsto \DSigma{}{} & (\EnumT [ \DVar{},
+\IDescD~I \mapsto \DSigma{}{} & (\EnumT{[ \DVar{},
                                           \DConst{},
@@
                                           \DSigma{}{}, 
-                                          \DPi{}{} ]) \\
+                                          \DPi{}{} ]}) \\
                               & \left[\begin{array}{l}
@@
 \newcommand{\HExprID}{\C{ExprIndexedD}}
-\newcommand{\HExprVarD}{\C{ExprD}_{\C{Var}}}
-\newcommand{\HExprFreeD}{\C{ExprD}_{\C{Free}}}
-\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}_{\C{Free}}}
+\newcommand{\HExprVarD}[1]{\C{ExprD}_{\C{Var},#1}}
+\newcommand{\HExprFreeD}{\C{ExprD}^{\C{Free}}}
+\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}^{\C{Free}}}
 
@@
 \[
-\Ty \mapsto \EnumT{[\Enat \quad \Ebool]}
+\Ty \mapsto \EnumT{[\Enat, \Ebool]}
 \]
@@
                  \begin{array}{l}
-                   \EnumT{[\Eval{}  \Econd{}{}{}]} \red{,} \\
+                   \EnumT{[\Eval{}, \Econd{}{}{}]} \red{,} \\
                    \LAM{ty}
@@
   comprehensible to ask that?} This defines a new indexed description,
-called $\HExprVarD$.
+called $\HExprVarD{dom}$.
 
@@
                \PITEL{\V{dom}}{\EnumU} \\
-             & \PI{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}} \\
+             & \PI{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})}} \\
              & (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To
@@
 \discharge\: ty\: dom\: \gamma\: (\SumLeft x)  & \Con{\pair{\Eval{}}{x}{}} \\
-\discharge\: ty\: dom\: \gamma\: (\SumRight v) & \switch{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}{\gamma}{v}
+\discharge\: ty\: dom\: \gamma\: (\SumRight v) & \switch{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})}{\gamma}{v}
 \end{array}
@@
 \newcommand{\substH}{\F{substExpr}}
+\newcommand{\domNat}{dom_{\CN{nat}}}
+\newcommand{\domBool}{dom_{\CN{bool}}}
+\newcommand{\gammaNat}{\gamma_{\CN{nat}}}
+\newcommand{\gammaBool}{\gamma_{\CN{bool}}}
 
@@
 \begin{array}{@{}ll}
-\substH  : & \PITEL{\V{ty}}{\Ty}
-            \PITEL{\V{dom}}{\EnumU} 
-            \PITEL{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}} \\
-          &  \PI{\V{\sigma}}{\PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom}{ty}) \To
-                             \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}} \\
-          & \IMu{\Ty}{\toIDesc{\HExprVarD}}{\V{ty}} \To
+\substH  : & \PITEL{\V{\domNat}}{\EnumU} 
+             \PITEL{\V{\gammaNat}}{\spi{\domNat}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domNat}}}{\Enat})}} \\
+           & \PITEL{\V{\domBool}}{\EnumU} 
+             \PITEL{\V{\gammaBool}}{\spi{\domBool}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domBool}}}{\Ebool})}} \\
+          &  \begin{array}{@{}ll}
+             \PI{\V{\sigma}}{& \PITEL{\V{\domNat}}{\EnumU} \\
+                             & \PITEL{\V{\gammaNat}}
+                                     {\spi{\domNat}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domNat}}}{\Enat})}} \\
+                             & \PITEL{\V{\domBool}}{\EnumU} \\
+                             & \PITEL{\V{\gammaBool}}
+                                     {\spi{\domBool}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domBool}}}{\Ebool})}} \\
+                             & \PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom_{ty}}{ty}) \To
+                               \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}}
+             \end{array}\\
+          & \PI{\V{ty}}{\Ty}
+            \IMu{\Ty}{\toIDesc{\HExprVarD{dom_{ty}}}}{\V{ty}} \To
             \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
@@
 \]
+
+Where $dom_{ty}$ is short for
+ $\case{ty}{\Enat  \To \domNat
+         \\ \Ebool \To \domBool}$
 
@@
 we are given. In particular, this is reminiscent to a free monad,
-where $\Eval{}$ is the \return introducing variables. Instead of
-implementing $\substH$ in this special case, we shall implement it in
-a generic setting, and express our example in that setting. However,
-before entering the realm of generic structures, let us further
-explore the universe of indexed descriptions.
+where $\Eval{}$ is the \return\ introducing variables. The substitution
+$\sigma$ is implemented from $\discharge$, by picking the domain and
+context corresponding to the type:
+
+\[
+\sigma\: \domNat\: \gammaNat\: \domBool\: \gammaBool\: ty\: var \mapsto 
+    \discharge\: ty\: dom_{ty}\: \gamma_{ty}\: var 
+\]
+
+Instead of implementing $\substH$ in this special case, we shall
+implement it in a generic setting, and express our example in that
+setting. However, before entering the realm of generic structures, let
+us further explore the universe of indexed descriptions.
 
@@
 \[
-\Rule{\Bhab{n}{\Nat}}
-     {\Bhab{\FinZero}{\Fin{(\NatSuc{n})}}}
-\qquad
-\Rule{\Bhab{n}{\Nat} \quad 
-      \Bhab{i}{\Fin{n}}}
-     {\Bhab{\FinSuc{}}{\Fin{(\NatSuc{n})}}}
+\stk{
+\data \Fin{} : \PI{\V{n}}{\Nat} \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
+    \FinZero_n      & \Fin{(\NatSuc{n})}   \\
+    \FinSuc{}_n     & \Fin{n} \To \Fin{(\NatSuc{n})}
+\end{array}
+}
 \]
@@
 \begin{array}{@{}ll}
-\FinD\: n = \DSigma{}{} & (\EnumT [ \FinZero , \FinSuc{} ]) \\
+\FinD\: n = \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                         & \left[
@@
 \FinD\: 0            & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
-\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT [ \FinZero , \FinSuc{} ]) \\
+\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                      &   &             & \left[
@@
 In Section~\ref{sec:idesc-examples}, we had the intuition that our
-datatypes $\HExprD$ and $\HExprVarD$ enjoy a monadic structure. We had
+datatypes $\HExprD$ and $\HExprVarD{dom}$ enjoy a monadic structure. We had
 identified the variable substitution operation as the \bind of a free
@@
 $$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{dom}{ty})}$$
-corresponds to expressions with variables, $\HExprVarD$.
+corresponds to expressions with variables, $\HExprVarD{dom}$.
 
@@
 \[\stk{
-\begin{array}{@{}ll}
-\substH  : & \PITEL{\V{ty}}{\Ty}
-             \PITEL{\V{dom}}{\EnumU} 
-             \PITEL{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}}}{ty})}} \\
-           &  \PI{\V{\sigma}}{\PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom}{ty}) \To
-                              \IMu{\Ty}{(\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}})}{ty}} \\
-           & \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{} + \Var{dom}{}}}}{\V{ty}} \To
-             \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}}}{\V{ty}} \\
-\end{array} \\
-\substH\: ty\: dom\: \gamma\: \sigma\: term \mapsto 
-    \substI\: \Ty\: 
-              (\Val{} + \Var{dom}{})\: 
-              \Val{}\:
-              \HExprFreeD\:
+\substH\: \domNat\: \domBool\: 
+            \gammaNat\: \gammaBool\:
             \sigma\: 
             ty\: 
+            term \mapsto  \\
+\;\;\ \begin{array}{ll}
+       \substI\: & \Ty\: 
+                  (\Val{} + \Var{dom_{ty}}{})\: 
+                  \Val{}\:
+                  \HExprFreeD\:  \\
+                &
+                  (\sigma\: \domNat\: \domBool\: \gammaNat\: \gammaBool)\:
+                  ty\:
                   term
+      \end{array}
 }\]
-
-\note{Not using $\discharge$, that's a shame.}
 
@@
 \begin{array}{@{}ll}
-\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT [ \DConst{},
+\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT{[ \DConst{},
                                               \DProd{}{},
                                               \DSigma{}{}, 
-                                              \DPi{}{} ]) \\
+                                              \DPi{}{} ]}) \\
                                   & \left[\begin{array}{l}

### 2010-03-18 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: Some minor fixes and notes

@@
 two different ways. This duality is actually reflected in the notation
-we can find in the literature, depending on the sensibility of the
-author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
-$\Sigma_{x : A} (B\: x)$. This notation stresses the fact that
-$\Sigma$-types are a generalisation of sums over arbitrary
-arities. When simply-typed languages have finite sums,
-dependently-typed languages have sums indexed by any set.
+we can find in the literature, depending on the sensibility of the author.
+The notation $\Sigma_{x : A} (B\: x)$ stresses the interpretation that
+$\Sigma$-types are a generalisation of sums over arbitrary arities.  Where
+simply-typed languages have finite sums, dependently-typed languages have
+sums indexed by any set.
 
-On the other hand, $\SigmaTy{A}{B}$ is sometimes written
-$\SIGMA{\V{x}}{A} (B\: x)$. Under this perspective, $\Sigma$-types can
-be read as a generalisation of products, where the second component
-can depend on the first one. When simply-typed languages pack data
-into non-dependent tuples, dependently-types languages tolerate this
-non-dependent usage. However, they also give the rather novel ability
-for data to influence further data.
+On the other hand, our choice of notation ($\SIGMA{\V{x}}{A} (B\: x)$)
+emphasises that $\Sigma$-types can be read as a generalisation of products,
+where the second component can depend on the first one. When simply-typed
+languages pack data into non-dependent tuples, dependently-types languages
+tolerate this non-dependent usage. However, they also give the rather novel
+ability for data to influence types of further data.
 
@@
 tougher, as one could legitimately expect an induction principle, for
-instance. Anonymous in a see of $\Sigma$s, our data-types have little
-chance of survival. \note{Ridiculous tentative of poetry. I was
-  really tired when I wrote that.}
+instance. Anonymous in a sea of $\Sigma$s, our data-types have little
+chance of survival. 
+\note{Ridiculous tentative of poetry. I was really tired when I wrote that.}
 
@@
 
-For our sigmas-of-sigmas to be known, we crucially need the ability to
-name them. \note{Pursuing this crappy analogy. Solving the type
-  presentation/representation debate would help solving that point.}
-To this end, we will use a standard technique in dependently-typed
-programming: we construct an ad-hoc universe. This technique dates
-back to Martin-L\"of definition of type
-theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
-programming technique~\cite{benke:universe-generic-prog, whoelse?}
+For the structure of our data-types to be known, we crucially need the
+ability to name them. \note{Pursuing this crappy analogy. Solving the type
+presentation/representation debate would help solving that point.} To this
+end, we will use a standard technique in dependently-typed programming: we
+construct an ad-hoc universe. This technique dates back to Martin-L\"of
+definition of type theory~\cite{martin-lof:itt}. Since then, it has been
+fruitful as a programming 
+technique~\cite{benke:universe-generic-prog,whoelse?}
 \note{I would need some other example of universe construction for
-  practical purposes}. We refer the reader to the Agda
-tutorial~\cite{norell:agda-tutorial} for a pedagogical presentation of
-universe construction. \note{Is that really a motivating motivation
-  for the usage of universe construction? }
+  practical purposes}. 
+We refer the reader to the Agda tutorial~\cite{norell:agda-tutorial} for a
+pedagogical presentation of universe construction. 
+\note{Is that really a motivating motivation for the usage of universe 
+construction?}
+\note{pwm: Surely ($\EnumU$, $\EnumT$) is a useful example of universes too.
+Even if it doesn't show off their full power?}
 
@@
 
-\begin{figure*}
+\begin{figure}
 
-\[
+\[\stk{
 \begin{array}{ll}
@@
 }
-&
+\vspace{0.2in}
+\\
 \stk{
@@
 \end{array}
-\]
+}\]
 
@@
 
-\end{figure*}
+\end{figure}
 
@@
 define in simply-typed languages.
+\note{pwm: Err, not quite --- not til we introduce $\DHindx{\!}{\!}$}
 
@@
 this type is not dependent. As this operator will be the corner stone
-of any inductive definition, we are better off sharping it before use.
+of any inductive definition, we are better off sharpening it before use.
 
@@
                  \PITEL{\V{P}}{\V{X} \To \Set} \\
-               & \PI{\V{xs}}{\descop{D}{\Mu{X}}} 
+               & \PI{\V{xs}}{\descop{D}{\V{X}}} 
                  \Set 
@@
                  & \PITEL{\V{R}}{\PI{\V{x}}{X} P\: x}
-                   \PI{\V{xs}}{\descop{D}{\Mu{X}}} 
+                   \PI{\V{xs}}{\descop{D}{\V{X}}} 
                    \All{D}{X}{P}{xs} 

### 2010-03-18 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: uniformize spacing in arrays
Desc@ICFP: uniformize EnumT definitions
Desc@ICFP: Some minor fixes and notes

@@
 type former $\EnumU$, the constructors $\NilE$ and $\ConsE$. We have
-just defined $\EnumU$ above, by fix-point over the signature functor
-$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
-$\ListCons$ constructors of lists.
+just defined $\EnumU$ above, by fix-point over the signature functor denoted
+by $\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and $\ListCons$
+constructors of lists.
 
@@
 implement the codes of the universe of descriptions. Interestingly, in
-and by itself, a code is nothing but a data-type. We are in the same
-situation than with $\EnumU$: we ought to be able to describe the
-codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
+and by itself, the system of codes, $\Desc$, is nothing but a data-type. We
+are in the same situation than with $\EnumU$: we ought to be able to describe
+the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
 first-class citizen, born with the standard, generic equipment of
@@
 \begin{array}{@{}ll}
-\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{} ])  \\
+\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{} ]})  \\
                            & \left[\begin{array}{l}
@@
     \ldots          & \:\:\ldots \\
-    \DHindx         & \PI{H}{Set} \Desc \To \Desc
+    \DHindx         & \PI{H}{\Set} \Desc \To \Desc
 \end{array}
 }
-&
+\vspace{0.2in}
+\\
 \stk{
@@
 \begin{array}{@{}ll}
-\DescD \mapsto \DSigma{}{} & (\EnumT [ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]) \\
+\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]}) \\
                            & \left[\begin{array}{l}
@@
 careful usage of type propagation. Indeed, let us try to elaborate
-this term down to the low-level type theory. The $\EnumT [ \ldots ]$
+this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
 construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
@@
 \where \\
-\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
+\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
     \FMVar           & X \To \FMFreeMonad\: F\: X                            \\
@@
 \end{array} \\
-\begin{array}{@{}l@{\:\:\mapsto\:}l}
+\begin{array}{@{}l@{\:\mapsto\:\:}l}
 \apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}   & \sigma\: x                   \\
@@
 build them as generic functions. Here, we rely on our ability to
-compute over descriptions, thanks to the induction principle.
+compute over descriptions.

### 2010-03-18 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Some minor fixes and notes

@@
 be defined at the same time. Where simply-typed languages have
-inductive types, we adopt inductive family of types as our basic
-grammar of data-types. To this end, we rely on \emph{indexing}. We are
-going to transform the $\Desc$ universe into $\IDesc$, the universe of
-indexed descriptions. We express inductive families by the $I \To
-\IDesc{I}$ type.
+inductive types, in dependently typed programming we the basic grammar of
+data-types should be that of inductive families of types. To this end, we
+rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
+$\IDesc$, the universe of indexed descriptions. We express inductive families
+by the $I \To \IDesc{I}$ type.  \note{pwm: Vec might not be the best
+motivating example here, since it doesn't appear in the paper (more on that
+later). Come to think of it, this is entirely too vague without spelling out
+a concrete example.}
 
@@
 shown in Figure~\ref{fig:type-former-desc}.
+\note{pwm: The exception being non dependent product -- we might want to
+explain why this is.}
 
@@
 
-\begin{figure*}
-
-\[
+\begin{figure}
+
+\[\stk{
 \begin{array}{ll}
@@
 by $\Bhab{\V{I}}{\Set}$, we rely on functors in the category
-$\Set^I$. We call these \emph{indexed functors}. In the following, we
+$\Set^I$. We call these \emph{indexed functors}. 
+\note{pwm: What's our story \emph{vis-\`{a}-vis} the 2 notions of indexed
+functor at play here: $(I\to\Set)\to\Set$ and $(I\to\Set)\to(I\to\Set)$?}  
+In the following, we
 implement $I \To \IDesc{I}$ as a grammar for indexed functors:
@@
 This generalisation can be achieved by a minor evolution of $\Desc$
-(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of $\Desc$
-is entirely captured by the $\DId$ code. We get an indexed functor
-by turning $\DId$ into an index-dependent $\DVar$ code:
+(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of the
+atomic $\Desc$ is entirely captured by the $\DId$ code. We get an indexed
+functor by turning $\DId$ into an index-dependent $\DVar$ code:
 
@@
 \\
-\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set        \\
+\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{I} \To (\V{I} \To \Set) \To \Set        \\
 \begin{array}{@{}l@{\:=\:\:}ll}
@@
 }
-&
+\vspace{0.2in}
+\\
 \stk{
-\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set                  \\
+\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{I} \To (\V{I} \To \Set) \To \Set                  \\
 \begin{array}{@{}l@{\:=\:\:}ll}
@@
 see more instances of this pattern in the following.
+\note{We're talking about vectors again, are we \emph{really} not going to
+define them?}
 
@@
 \begin{array}{@{}ll}
-\IDescD~I \mapsto \DSigma{}{} & (\EnumT{[ \DVar{},
+\IDescD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
                                           \DConst{},
@@
                                           \DSigma{}{}, 
-                                          \DPi{}{} ]}) \\
-                              & \left[\begin{array}{l}
+                                          \DPi{}{} \red{]}) \\
+                              & \bigRedBracket{\begin{array}{l}
                                       \DConst{I}                  \\
@@
                                    \end{array}
-                             \right]
+                             }
 \end{array}
@@
 Note that this description also lives in $\IDesc{\Unit}$. Indeed,
-$\IDesc$ is merely an inductive type. As for the $\Desc$ levitation,
-the finite function space $[ \ldots ]$ elaborates into a special
-purpose $\F{switchID}$ operator.
+for any $\V{I}$, $\IDesc~\V{I}$ is merely an inductive type. As for the
+$\Desc$ levitation, the finite function space $[ \ldots ]$ elaborates into a
+special purpose $\F{switchID}$ operator.
 
@@
   haunt me with that?  A ghost under a mixed prefix?
-  Aaaaaaaaaaargh\ldots}
+  Aaaaaaaaaaargh\ldots (Wha? - pwm)}
+\note{pwm: I don't really get the need for this $\Epolym$ notation. Why not
+just abstract over an elem of $\Ty£$?}
 
@@
 \HExprAD : \ATagIDesc{\Ty} \\
-\HExprAD \mapsto \left[
+\HExprAD \mapsto \bigRedBracket{
                  \begin{array}{l}
@@
                    \LAM{ty}
-                   \left[ %%! Should be read
+                   \bigRedBracket{
                    \begin{array}{l}
@@
                    \end{array}
-                   \right]
-                 \end{array}
-                 \right] %%! Should be \red
+                   }
+                 \end{array}
+                 }
 \\
@@
 \HExprID : \ITagIDesc{\Ty} \\
-\HExprID \mapsto \left[
+\HExprID \mapsto \bigRedBracket{
                  \begin{array}{l}
@@
                    \end{array}
-                   \right]
+                   }
 }
@@
 
-\note{the big square brackets should be red}
+\note{the big square brackets should be red (pwm: use \\bigRedBracket :)) }
 
@@
 }\]
+\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of IFunc
+floating around, $\idescop{\_}{}{}$ is defined using the 1st notion, and used it
+here as if it were the 2nd}
 
@@
 \FinD\: n = \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
-                        & \left[
+                        & \bigRedBracket{
                           \begin{array}{l}
@@
                           \end{array}
-                          \right]
-\end{array}
-}\]
-
-Intuitively, the constraint is captured as follow. First, we store an
+                          }
+\end{array}
+}\]
+
+Intuitively, the constraint is captured as follows. First, we store an
 element $m$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
@@
 \note{ I suspect we don't implement vectors, space-wise }
+\note{pwm: I suspect we don't get away with leaving them out}
 
@@
 much wider class of data-types. Indeed, we pose no constraint apart
-from strict positivity. This suggests that, based on the familiar
-grammar of GADTs, we can simply lift off the restrictions and
-translates the user's input to a code in our universe of indexed
-description.
+from strict positivity\note{pwm: Is this actually the 1st mention of this
+phrase? I think that should be rectified :)}. This suggests that, based on
+the familiar grammar of GADTs, we can simply lift off the restrictions and
+translates the user's input to a code in our universe of indexed description.
 
@@
 \FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
-                     &   &             & \left[
+                     &   &             & \bigRedBracket{
                                          \begin{array}{l}
@@
                                          \end{array}
-                                         \right]
+                                         }
 \end{array}
@@
 experience, we are going to present its equivalent in the indexed
-world. The processus is the same. Namely, given an indexed functor, we
+world. The process the same. Namely, given an indexed functor, we
 derive the indexed functor coding its free monad:
+\note{pwm: Whoa there. Maybe we should say something about IMonads in
+general before we get to this point?}
 
@@
 \HExprAFreeD : \ATagIDesc{\Ty} \\
-\HExprAFreeD \mapsto \left[
+\HExprAFreeD \mapsto \bigRedBracket{
                  \begin{array}{l}
-                   \EnumT{[\Econd{}{}{}]} \red{,} \\
+                   \EnumT{\red{[}\Econd{}{}{}\red{]}} \red{,} \\
                    \LAM{ty}
-                   \left[ %%! Should be red
+                   \bigRedBracket{
                    \begin{array}{l}
@@
                    \end{array}
-                   \right]
-                 \end{array}
-                 \right] %%! Should be \red
+                   }
+                 \end{array}
+                 }
 }\]
@@
 \begin{array}{@{}ll}
-\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT{[ \DConst{},
+\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DConst{},
                                               \DProd{}{},
                                               \DSigma{}{}, 
-                                              \DPi{}{} ]}) \\
-                                  & \left[\begin{array}{l}
+                                              \DPi{}{} \red{]}) \\
+                                  & \bigRedBracket{\begin{array}{l}
                                         \DConst{\Set}               \\
@@
                                         \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
-                                    \end{array}\right]
+                                    \end{array}}
 \end{array}
@@
 \IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
-\IDescD\: I \mapsto \FreeIMonad{[\IDescFreeD]}{\Unit}{I}\: I
+\IDescD\: I \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{]}}{\Unit}{I}\: I
 }\]

### 2010-03-18 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: 'bidirectional'
Desc@ICFP: Some minor fixes and notes

@@
         <- Presentation independent of particular implementation choice
-        -> Model in Agda, intentional
+        -> Model in Agda, intensional
         -> Used in Epigram, OTT
@@
 choice. In particular, our system has been modelled in Agda, which
-features an intentional equality. On the other hand, it is used in
+features an intensional equality. On the other hand, it is used in
 Epigram, which equality has a slightly extensional
@@
 
+\note{pwm: We should explain this new judgement form, specifically that
+$\propag{}{}$ means `elaborates to'.}
+
 \[
@@
   of their type theory}. In our case, we have shown how to instrument
-bi-directionality to rationalise the expressivity of dependent
+bidirectionality to rationalise the expressivity of dependent
 types. We have illustrated our approach with finite sets. We have
@@
 as type information is available. We have shown how to rationalise
-this information in a formal presentation. Hence, we have proved that
+this information in a formal presentation. Hence, we have shown that
 programming with such objects is practical. As we introduce more codes

### 2010-03-18 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: Single column figures, where possible.

@@
 type former $\EnumU$, the constructors $\NilE$ and $\ConsE$. We have
-just defined $\EnumU$ above, by fix-point over the signature functor denoted
-by $\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and $\ListCons$
-constructors of lists.
+just defined $\EnumU$ above, by fix-point over the signature functor
+$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
+$\ListCons$ constructors of lists.
 
@@
 implement the codes of the universe of descriptions. Interestingly, in
-and by itself, the system of codes, $\Desc$, is nothing but a data-type. We
-are in the same situation than with $\EnumU$: we ought to be able to describe
-the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
+and by itself, a code is nothing but a data-type. We are in the same
+situation than with $\EnumU$: we ought to be able to describe the
+codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
 first-class citizen, born with the standard, generic equipment of
@@
     \ldots          & \:\:\ldots \\
-    \DHindx         & \PI{H}{\Set} \Desc \To \Desc
+    \DHindx         & \PI{H}{Set} \Desc \To \Desc
 \end{array}
@@
 build them as generic functions. Here, we rely on our ability to
-compute over descriptions.
+compute over descriptions, thanks to the induction principle.

### 2010-03-18 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Single column figures, where possible.
Desc@ICFP: point-wise sum.

@@
 
-Consequently, replacing $\Val{ty}$ by $(\SUM{\Val{ty}}{\Var{dom}{ty}})$ in
-Figure~\ref{fig:hexpr-full} turns the language of closed term into a
-language of opened terms with constants. \note{Is it polite and
-  comprehensible to ask that?} This defines a new indexed description,
-called $\HExprVarD{dom}$.
+Consequently, replacing $\Val{ty}$ by
+$(\SUM{\Val{ty}}{\Var{dom}{ty}})$ in Figure~\ref{fig:hexpr-full} turns
+the language of closed term into a language of opened terms with
+constants. For readability, we will abbreviate $\LAM{ty}
+\SUM{\Val{ty}{\Var{dom}{ty}}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
+  it polite and comprehensible to ask that?} This defines a new
+indexed description, called $\HExprVarD{dom}$.
 
@@
                                      {\spi{\domBool}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domBool}}}{\Ebool})}} \\
-                             & \PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom_{ty}}{ty}) \To
+                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{ty}}{\Var{dom_{ty}}{ty}}) \To
                                \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}}
@@
 corresponds to the syntax of closed terms, $\HExprD$. Similarly,
-$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{dom}{ty})}$$
+$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \SUM{\Val{ty}}{\Var{dom}{ty}})}$$
 corresponds to expressions with variables, $\HExprVarD{dom}$.
@@
        \substI\: & \Ty\: 
-                  (\Val{} + \Var{dom_{ty}}{})\: 
+                  (\SUM{\Val{}}{\Var{dom_{ty}}{}})\: 
                   \Val{}\:

### 2010-03-19 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: Single column figures, where possible.
IDesc@ICFP: pass over \note{}s. Remove non documented \propag usage.

@@
 description. Formally, we have:
-\note{We have talked about a tagged description based on $\List (\UId
-  \times \Desc)$. It makes the free monad easier to construct but,
-  imho, considerably clutters anything else. I've reverted back to a
-  more sigma-friendly presentation.}
 
@@
 \]
-
-At this stage, we have recovered the data-types we are usually able to
-define in simply-typed languages.
-\note{pwm: Err, not quite --- not til we introduce $\DHindx{\!}{\!}$}
 

### 2010-03-19 papers/icfp-2010-desc/paper_type_theory.tex
IDesc@ICFP: pass over \note{}s. Remove non documented \propag usage.

@@
 
-\note{pwm: We should explain this new judgement form, specifically that
-$\propag{}{}$ means `elaborates to'.}
-
 \[
-\Axiom{\Gamma ; \xS ; \Delta \Vdash \propag{\x}
-                                           {\pull{\x}{S}}}
+\Rule{\Gamma ; \xS ; \Delta \vdash \Valid}
+     {\Gamma ; \xS ; \Delta \vdash \Bhab{x}{S}}
 \]
@@
 \[
-\Rule{\Gamma ; \xS \Vdash \propag{\push{t}{T}}
-                                 {t'}}
-     {\Gamma \Vdash \propag{\push{\LAM{\x} t}{\PIS{\xS}{T}}}
-                           {\PLAM{\x}{S} t'}} 
+\Rule{\Gamma       \vdash \Type{S} \quad
+      \Gamma ; \xS \vdash \Bhab{t}{T}}
+     {\Gamma \vdash \Bhab{\PLAM{\x}{S} t}{\PIS{\xS} T}}
 \]
 
-In our formal presentation of the type theory, we were forced to
-annotate lambda abstractions by the type of the domain. In a
-bidirectional setting, the domain is provided by the $\Pi$ type we are
-pushing the abstraction into. Hence we do not need the annotation.
+Here, we are forced to annotate lambda abstractions by the type of the
+domain. In a bidirectional setting, the domain is provided by the
+$\Pi$ type we are pushing the abstraction into. Hence we do not need
+the annotation.
 

### 2010-03-19 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: a tagged indexed description is a I -> IDesc I in becoming. Fix FreeIMonad construction. Fix usages of HExprFreeD.
Desc@ICFP: fix typos. Clean-up solved \notes.
Desc@ICFP: fix typo for evalOne

@@
 
-\note{the big square brackets should be red (pwm: use \\bigRedBracket :)) }
-
 \begin{wstructure}
@@
 \[\stk{
-\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}}{\Ty}{ty} \To {\Val{ty}} \\
+\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\: ty}{\Ty}{ty} \To {\Val{ty}} \\
 \begin{array}{@{}l@{\:=\:\:}l}
@@
 }\]
-\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of IFunc
-floating around, $\idescop{\_}{}{}$ is defined using the 1st notion, and used it
-here as if it were the 2nd}
+\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of
+  IFunc floating around, $\idescop{\_}{}{}$ is defined using the 1st
+  notion, and used it here as if it were the 2nd. pierre: in this
+  case, it was a typo. But you're right, there is two notions and we
+  probably want to clarify where these things live.}
 
@@
 constants. For readability, we will abbreviate $\LAM{ty}
-\SUM{\Val{ty}{\Var{dom}{ty}}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
+\SUM{\Val{ty}}{\Var{dom}{ty}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
   it polite and comprehensible to ask that?} This defines a new
@@
                       \PITEL{\V{X}}{\V{I} \To \Set} \\
-                    & \PITEL{\V{R}}{\V{I} \To \TagIDesc{I}} \To
-                      (\V{I} \To \TagIDesc{I})
+                    & \PITEL{\V{R}}{\TagIDesc{I}} \To
+                      \TagIDesc{I}
 \end{array} \\
-\FreeIMonad{R}{I}{X} \mapsto
-    \LAM{i}
-    \pair{\ListCons{\DVar{}}{(\fst{(R\: i)})}}
-         {\pair{(\DConst{X\: i})}
-               {(\snd{(R\: i)})}{}}{}
+\FreeIMonad{(E,F)}{I}{X} \mapsto
+    \pair{\pair{\ListCons{\DVar{}}{(\fst{E})}} 
+               {\LAM{i}
+                \pair{(\DConst{(X\: i)})}
+                     {((\snd{E})\: i)}{}}{}}
+         {F}{}
 }\]
@@
             \PITEL{\V{X}, \V{Y}}{\V{I} \To \Set}
-            \PITEL{\V{R}}{\V{I} \To \TagIDesc{\V{I}}} \\
+            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
           & \PITEL{\V{\sigma}}{\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 

### 2010-03-23 papers/icfp-2010-desc/paper_idesc.tex
IDesc@ICFP: fix typo in substI
Desc@ICFP: remove \Epolym, instead abstract over the type. pwm: is that presentation more convincing?

@@
 \newcommand{\Enat}{\etag{\CN{nat}}}
-\newcommand{\Epolym}{\etag{\alpha}}
 
@@
 \begin{array}{l@{\::\:\:}l}
-\Econd{}{}{}     & \Ebool \To \Epolym \To \Epolym \To \Epolym  \\ 
+\Econd{}{}{}     & \forall \Bhab{ty}{\Ty} . \Ebool \To ty \To ty \To ty  \\ 
 \Eplus{}{}       & \Enat \To \Enat \To \Enat                           \\
 \Ele{}{}         & \Enat \To \Enat \To \Ebool                          \\
-\Eval{}          & \Val{\Epolym} \To \Epolym
+\Eval{}          & \forall \Bhab{ty}{\Ty} . \Val{ty} \To ty
 \end{array}
@@
 
-The notation $\Epolym$ represents a form of ``polymorphism'', meaning
-that $\alpha$ can be any type. \note{Polymorph? Will any ghost try to
-  haunt me with that?  A ghost under a mixed prefix?
-  Aaaaaaaaaaargh\ldots (Wha? - pwm)}
-\note{pwm: I don't really get the need for this $\Epolym$ notation. Why not
-just abstract over an elem of $\Ty£$?}
-
 The function $\Val{}$, used in the definition of $\Eval{}$, simply
-maps $\Epolym$ to the corresponding type in the host language. Hence,
-the argument of $\Eval{}$ are ensured to be of the expected type. We
-assume $\Nat$ and $\Bool$ represent natural numbers and booleans in
-the host language. Further, we also assume the existence of an
-addition and comparison operator in $\Nat$, respectively named
-$\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$ as follow:
+maps an object type $ty$ to the corresponding type in the host
+language. Hence, the argument of $\Eval{}$ are ensured to be of the
+expected type. We assume $\Nat$ and $\Bool$ represent natural numbers
+and booleans in the host language. Further, we also assume the
+existence of an addition and comparison operator in $\Nat$,
+respectively named $\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$
+as follow:
 
@@
           & \PITEL{\V{\sigma}}{\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 
-                               \IMu{I}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}} \\
+                               \IMu{I}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}} \\
           & \PITEL{\V{i}}{\V{I}}
-            \PITEL{\V{D}}{\IMu{I}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})}{\V{i}}} \To
-            \IMu{I}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}
+            \PITEL{\V{D}}{\IMu{I}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{X}}})}{\V{i}}} \To
+            \IMu{I}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}
 \end{array}

### 2010-03-23 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Give Nat in type-former-desc.

@@
 Consequently, the examples previously developed can be
-straightforwardly translated into this new presentation. \note{We
-  might want to re-do the Nat example here, as we encode once again in
-  $\IDesc$} In the following, we adopt this last version as our de
+straightforwardly translated into this new presentation. For example,
+here is the new definition of $\NatD$:
+
+\[\stk{
+\NatD : \Desc \\
+\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
+                     {[ \DUnit \quad \DId ]}
+}\]
+
+In the following, we adopt this last version as our de
 facto universe of inductive types. In particular, we are going to

### 2010-03-23 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: justify introduction of \DProd

@@
 codes as quotations of standard type-formers. This presentation is
-shown in Figure~\ref{fig:type-former-desc}.
-\note{pwm: The exception being non dependent product -- we might want to
-explain why this is.}
+shown in Figure~\ref{fig:type-former-desc}. The reader will notice the
+introduction of $\DProd{}{}$ that overlaps with $\DSigma{}{}$. Unlike
+$\DSigma{}{}$, $\DProd{}{}$ is first-order, hence amenable to a
+definitional treatment. \note{We could replace $\DUnit$ by $\DConst$
+  here, instead of doing it in the following section. This is also a
+  story about being first-order, so that would make sense, isn't it?}
 

### 2010-03-23 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: evalOne: fix typo
Desc@ICFP: vectors have finally landed.

@@
 very specificity of dependently-typed systems: terms -- the length --
-are allowed to influence types -- the vector type.
+are allowed to influence types -- the vector type. \note{Reference
+  paper to cite about vectors in dependent types?}
 
@@
 
+\begin{wstructure}
+<- Vectors
+    Do we treat them in the end? 
+    What can we say here we haven't with Fin?
+\end{wstructure}
+
 In the case of vectors, for instance, for the index $\NatZero$, we
 would only propose the constructor $\ListNil$. Similarly, for
-$\NatSuc{n}$, we only propose the constructor $\ListCons{}{}$. We will
-see more instances of this pattern in the following.
-\note{We're talking about vectors again, are we \emph{really} not going to
-define them?}
+$\NatSuc{n}$, we only propose the constructor $\ListCons{}{}$. Hence,
+we obtain the following definition:
+
+\newcommand{\VecD}{\C{VecD}}
+\newcommand{\VecNil}{\etag{\CN{vnil}}}
+\newcommand{\VecCons}[2]{\etag{\CN{vcons}}\:#1\:#2}
+
+\[\stk{
+\VecD : \Set \To \TagIDesc{\Nat} \\
+\begin{array}{@{}ll}
+\VecD\: X = \bigRedBracket{ 
+            \begin{array}{l}
+                \pair{[]}{\_}{} \\
+                \bigRedBracket{
+                \begin{array}{l} 
+                  \LAM{n} \caseB{n}{\NatZero \To [\VecNil] \\
+                                   (\NatSuc n) \To [\VecCons{}{}]} \\
+                  \LAM{n} \caseB{n}{\NatZero \To [\DConst{Unit}] \\
+                                   (\NatSuc n) \To [\DSigma{X}{\DVar{n}}]}
+                \end{array}}
+            \end{array}}
+\end{array}
+}\]
+
+We will see more instances of this pattern in the following.
 
@@
 \evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{y}{}}{}}{}  & x \\
-\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{x}{y}{}}{}                       & \F{plusHost}\: x y \\
-\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{x}{y}{}}{}                        & \F{leHost}\: x y 
+\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{x}{y}{}}{}                       & \F{plusHost}\: x\: y \\
+\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{x}{y}{}}{}                        & \F{leHost}\: x\: y 
 \end{array} \\
@@
 \begin{wstructure}
-<- Vectors
-    Do we treat them in the end? 
-    What can we say here we haven't with Fin?
-\end{wstructure}
-
-\note{ I suspect we don't implement vectors, space-wise }
-\note{pwm: I suspect we don't get away with leaving them out}
-
-\begin{wstructure}
 <- GADT-style data-type definition

### 2010-03-23 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: introduce \DConst when moving Desc to a type former presentation. Transfered from IDesc and avoid redundancy.

@@
 codes as quotations of standard type-formers. This presentation is
-shown in Figure~\ref{fig:type-former-desc}. The reader will notice the
-introduction of $\DProd{}{}$ that overlaps with $\DSigma{}{}$. Unlike
-$\DSigma{}{}$, $\DProd{}{}$ is first-order, hence amenable to a
-definitional treatment. \note{We could replace $\DUnit$ by $\DConst$
-  here, instead of doing it in the following section. This is also a
-  story about being first-order, so that would make sense, isn't it?}
+shown in Figure~\ref{fig:type-former-desc}. The reader will notice
+that we replace $\DUnit$ by a more general $\DConst$ code. Whereas
+$\DUnit$ was interpreted as the unit set, $\DConst{X}$ is interpreted
+as $X$, for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
+$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
+succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
+its equivalent encoding. From a definitional perspective, we are
+giving more opportunities to the type-system, hence reducing the
+burden on the programmer. For the same reason, we introduce
+$\DProd{}{}$ that overlaps with $\DSigma{}{}$.
 
@@
     \DId            & \Desc                                    \\
-    \DUnit          & \Desc                                   \\
+    \DConst{}       & \Set \To \Desc                           \\
     \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
@@
 \descop{\DId}{X}          &  X                                           \\
-\descop{\DUnit}{X}        &  \Unit                                       \\
+\descop{\DConst{Z}}{X}    &  Z                                           \\
 \descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
@@
 The resulting code of $\IDesc$ is presented in
-Figure~\ref{fig:idesc}. The reader will notice that we have also
-replaced $\DUnit$ by a more general $\DConst$ code. Whereas $\DUnit$
-was interpreted as the unit set, $\DConst{X}$ is interpreted as $X$,
-for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
-$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
-succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
-its equivalent encoding. From a definitional perspective, we are
-giving more opportunities to the type-system, hence reducing the
-burden on the programmer.
+Figure~\ref{fig:idesc}.
 

### 2010-03-24 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix substExpr (\substH) type to fit the Agda model. I tried to come up with a nicer type signature, not sure it's a success tho.
Desc@ICFP: beginning of clarification of the indexed functor situation.

@@
 \begin{wstructure}
+[Outdated: type former presentation instead]
 <- Adding hindx have introduced some duplication
@@
         /> Still embeddable (refer to the Agda model)
+\end{wstructure}
+
+\begin{wstructure}
+<- Also replacing '1 by 'const  [figure]
+    <- For convenience
+        <- 'const X equivalent to 'sigma X (\_ -> '1)
+        /> Easier to abstract
+            <- Extensionally same
+            /> 'const more useful in practice
 \end{wstructure}
@@
 inductive types in this setting. To encode an inductive family indexed
-by $\Bhab{\V{I}}{\Set}$, we rely on functors in the category
-$\Set^I$. We call these \emph{indexed functors}. 
-\note{pwm: What's our story \emph{vis-\`{a}-vis} the 2 notions of indexed
-functor at play here: $(I\to\Set)\to\Set$ and $(I\to\Set)\to(I\to\Set)$?}  
-In the following, we
-implement $I \To \IDesc{I}$ as a grammar for indexed functors:
+by $\Bhab{\V{I}}{\Set}$, we rely on endofunctors on the category
+$\Set^I$. We call these \emph{indexed functors}. In the following, $I
+\To \IDesc{I}$ is a grammar for indexed functors: \note{pwm: What's
+  our story \emph{vis-\`{a}-vis} the 2 notions of indexed functor at
+  play here: $(I\to\Set)\to\Set$ and $(I\to\Set)\to(I\to\Set)$?}
 
@@
 This generalisation can be achieved by a minor evolution of $\Desc$
-(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of the
-atomic $\Desc$ is entirely captured by the $\DId$ code. We get an indexed
+(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of
+$\Desc$ is entirely captured by the $\DId$ code. We get an indexed
 functor by turning $\DId$ into an index-dependent $\DVar$ code:
@@
 
-\begin{wstructure}
-<- Also replacing '1 by 'const  [figure]
-    <- For convenience
-        <- 'const X equivalent to 'sigma X (\_ -> '1)
-        /> Easier to abstract
-            <- Extensionally same
-            /> 'const more useful in practice
-\end{wstructure}
-
 The resulting code of $\IDesc$ is presented in
-Figure~\ref{fig:idesc}.
+Figure~\ref{fig:idesc}. A fix-point consists of the fix-point of the
+\emph{family} of functors as a whole. Hence the following definition:
+
+\[\stkl{
+\Rule{\stk{ 
+          \Gamma \vdash \Bhab{I}{\Set} \\
+          \Gamma \vdash \Bhab{R}{I \To \IDesc{I}} \\
+          \Gamma \vdash \Bhab{i}{I}}}
+     {\Gamma \vdash \Bhab{\IMu{I}{R}{i}}{\Set}} \qquad
+\Rule{\stk{
+          \Gamma \vdash \Bhab{I}{\Set} \\
+          \Gamma \vdash \Bhab{R}{I \To \IDesc{I}} \\
+          \Gamma \vdash \Bhab{i}{I} \\
+          \Gamma \vdash \Bhab{x}{\idescop{R\: i}{I}{(\IMu{I}{R}{})}}}}
+     {\Gamma \vdash \Bhab{\Con{x}}{\IMu{I}{R}{i}}}
+}\]
 
@@
 \newcommand{\gammaBool}{\gamma_{\CN{bool}}}
+\newcommand{\GammaSpi}[2]{\Gamma_{\CN{ty}}~#1~#2}
 
@@
 \begin{array}{@{}ll}
-\substH  : & \PITEL{\V{\domNat}}{\EnumU} 
-             \PITEL{\V{\gammaNat}}{\spi{\domNat}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domNat}}}{\Enat})}} \\
-           & \PITEL{\V{\domBool}}{\EnumU} 
-             \PITEL{\V{\gammaBool}}{\spi{\domBool}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domBool}}}{\Ebool})}} \\
+\substH  : & \PITEL{\V{\dom}}{\EnumU} \\
+           & \PITEL{\V{\gammaNat}}{\GammaSpi{dom}{\Enat}} 
+             \PITEL{\V{\gammaBool}}{\GammaSpi{dom}{\Ebool}} \\
            & \begin{array}{@{}ll}
-             \PI{\V{\sigma}}{& \PITEL{\V{\domNat}}{\EnumU} \\
+             \PI{\V{\sigma}}{& \PITEL{\V{dom}}{\EnumU} \\
                              & \PITEL{\V{\gammaNat}}
-                                     {\spi{\domNat}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domNat}}}{\Enat})}} \\
-                             & \PITEL{\V{\domBool}}{\EnumU} \\
+                                     {\GammaSpi{dom}{\Enat}} \\
                              & \PITEL{\V{\gammaBool}}
-                                     {\spi{\domBool}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domBool}}}{\Ebool})}} \\
-                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{ty}}{\Var{dom_{ty}}{ty}}) \To
-                               \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}}
+                                     {\GammaSpi{dom}{\Ebool}} \\
+                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To \\
+                             & \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}}
              \end{array}\\
           & \PI{\V{ty}}{\Ty}
-            \IMu{\Ty}{\toIDesc{\HExprVarD{dom_{ty}}}}{\V{ty}} \To
-            \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
+            \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{\V{ty}} \To \\
+          & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
 \end{array}
@@
 
-Where $dom_{ty}$ is short for
- $\case{ty}{\Enat  \To \domNat
-         \\ \Ebool \To \domBool}$
+Where $\GammaSpi$ corresponds to a context, defined by:
+
+\[\stk{
+\GammaSpi{}{} : \EnumU \To \Ty \To \Set                                             \\
+\GammaSpi{dom}{ty} = \spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})} 
+}\]
 
@@
 \[
-\sigma\: \domNat\: \gammaNat\: \domBool\: \gammaBool\: ty\: var \mapsto 
-    \discharge\: ty\: dom_{ty}\: \gamma_{ty}\: var 
+\sigma\: dom\: \gammaNat\: \gammaBool\: ty\: var \mapsto 
+    \discharge\: ty\: dom\: \gamma_{ty}\: var 
 \]
+
+Where $\gamma_{ty}$ is short for
+ $\caseB{ty}{\Enat  \To \gammaNat \\
+             \Ebool \To \gammaBool}$
 
@@
 \[\stk{
-\substH\: \domNat\: \domBool\: 
+\substH\: dom\:
           \gammaNat\: \gammaBool\:
@@
        \substI\: & \Ty\: 
-                  (\SUM{\Val{}}{\Var{dom_{ty}}{}})\: 
+                  (\SUM{\Val{}}{\Var{dom}{}})\: 
                   \Val{}\:
@@
                 &
-                  (\sigma\: \domNat\: \domBool\: \gammaNat\: \gammaBool)\:
+                  (\sigma\: dom\: \gammaNat\: \gammaBool)\:
                   ty\:

### 2010-03-24 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: clearing the indexed functor business.

@@
 In the previous section, we have presented the $\Desc$ universe as the
-grammar of functors in the category $\Set$. We have seen how to encode
-inductive types in this setting. To encode an inductive family indexed
-by $\Bhab{\V{I}}{\Set}$, we rely on endofunctors on the category
-$\Set^I$. We call these \emph{indexed functors}. In the following, $I
-\To \IDesc{I}$ is a grammar for indexed functors: \note{pwm: What's
-  our story \emph{vis-\`{a}-vis} the 2 notions of indexed functor at
-  play here: $(I\to\Set)\to\Set$ and $(I\to\Set)\to(I\to\Set)$?}
+grammar of functors in the category $\Set$. We have seen how to code
+inductive types in this setting. To describe an inductive family
+indexed by $\Bhab{I}{\Set}$, we use endofunctors on the category
+$\Set^I$. We call these \emph{indexed functors}. \note{Predicate
+  transformer anyone?} In this setting, $I \To \IDesc{I}$ is our
+grammar for describing these functors. Hence, $\IDesc{}$ and its
+interpretation have the following types:
 
@@
 
-This generalisation can be achieved by a minor evolution of $\Desc$
-(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of
-$\Desc$ is entirely captured by the $\DId$ code. We get an indexed
-functor by turning $\DId$ into an index-dependent $\DVar$ code:
-
-\[\stk{
-\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \DVar{}         & I \To \IDesc{I}                                   \\
-    \ldots          & \ldots
-\end{array} \\
-\\
-\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{I} \To (\V{I} \To \Set) \To \Set        \\
-\begin{array}{@{}l@{\:=\:\:}ll}
-\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
-\ldots                        &  \ldots
-\end{array}
-}\]
-
-The resulting code of $\IDesc$ is presented in
-Figure~\ref{fig:idesc}. A fix-point consists of the fix-point of the
-\emph{family} of functors as a whole. Hence the following definition:
+Note that $I \To \IDesc{I}$ is isomorphic, up to interpretation, to an
+endofunctor on $\Set^I$. Indeed, $I \To \IDesc{I}$ is interpreted as
+$I \To (I \To \Set) \To \Set$, which is isomorphic to $(I \To \Set)
+\To (I \To \Set)$. Fix-point of inductive families is defined over
+these indexed functors, hence computing a fix-point of the entire
+\emph{family} of functors:
 
@@
 }\]
+
+However, we still have to define the actual grammar. This can be
+achieved by a minor evolution of $\Desc$
+(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of
+$\Desc$ is entirely captured by the $\DId$ code. We get an indexed
+functor by turning $\DId$ into an index-dependent $\DVar$ code. The
+resulting code of $\IDesc$ is presented in Figure~\ref{fig:idesc}.
+
+%% \[\stk{
+%% \data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
+%% \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
+%%     \DVar{}         & I \To \IDesc{I}                                   \\
+%%     \ldots          & \ldots
+%% \end{array} \\
+%% \\
+%% \idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{I} \To (\V{I} \To \Set) \To \Set        \\
+%% \begin{array}{@{}l@{\:=\:\:}ll}
+%% \idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
+%% \ldots                        &  \ldots
+%% \end{array}
+%% }\]
 
@@
 \begin{array}{@{}llll}
-\FinD\: 0            & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
+\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
 \FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\

### 2010-03-24 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix GADT discussion.

@@
 universe of indexed description supports this restricted class of
-dependent data-types. As we have shown with $\Fin$, constrained
-constructors are straightforwardly translated into equality
-constraints.
-
-While we support the same grammar, indexed descriptions embraces a
-much wider class of data-types. Indeed, we pose no constraint apart
-from strict positivity\note{pwm: Is this actually the 1st mention of this
-phrase? I think that should be rectified :)}. This suggests that, based on
-the familiar grammar of GADTs, we can simply lift off the restrictions and
-translates the user's input to a code in our universe of indexed description.
+inductive families. As we have seen with $\Fin$, constrained
+constructors are straightforwardly translated into equations.
 
 However, one could, quite rightfully, be worried by the introduction
-of equations in the picture. In the context of programming, these
-equations will come back and haunt us, begging for proofs. Hence,
-whereas the user typed a constructor-constrained data-type, the
-elimination form will involve equations and their proofs. This
-introduces a loophole in our abstraction we should try to address.
-
-\note{Is that convincing? Worth saying? Anything else?}
+of equalities in the picture. In a programming context, equalities
+haunt us, begging for proofs. Hence, whereas the user typed a
+constructor-constrained data-type, the elimination form will involve
+equalities and their proofs. As much as possible, one would rather
+avoid suffering from this burden. However, the equalities introduced
+here are rather special: they are \emph{equations} constraining
+variables to certain terms. Goguen et
+al.~\cite{goguen:pattern-matching} have shown, in the context of
+dependent pattern matching, how to tame these equations.
 
@@
 
-To get a clue of a solution, let us look back at $\Fin$. We note that
-the equation is introduced because we are \emph{storing} the index of
-the inductive family, through $m$. Hence the constraint. However,
-\emph{inductive families need not store their
-  indices}~\cite{brady:index-inductive-families}\note{ I'm not
-  familiar with this, I need to get into the paper. }. Applying the
-\emph{forcing} optimisation to our definition of $\Fin$ gives the
-following, equivalent definition:
+Further, we can actually remove these equations altogether. Let us
+look back at $\Fin$. We note that the equations are introduced because
+we are \emph{storing} the index of the inductive family, through
+$m$. Hence the constraints. However, \emph{inductive families need not
+  store their indices}~\cite{brady:index-inductive-families}. Applying
+the \emph{forcing} optimisation to our initial definition of $\Fin$
+gives the following, equivalent definition:
 
@@
 
-Equations have simply disappeared! We should precise that forcing a
+Equations have simply disappeared. We should precise that forcing a
 description is not guaranteed to remove all constraints. It is subject
-of future work to see if constraints can be entirely eradicated, or
+to future work to see if constraints can be entirely eradicated, or
 presented more conveniently to the user. Finally, it is worth

### 2010-03-25 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: Section 'type theory' done. Good night.
Desc@ICFP: Second section on Desc almost done. It is just too late, or too early to do initial algebras.
Desc@ICFP: get some sort of introduction.
Desc@ICFP: discuss stratification.

@@
 \section{Discussion}
+\label{sec:discussion}
 
-\begin{structure}
+\subsection{Universe stratification}
+
+\begin{wstructure}
 !!! Need Help !!!
@@
     ???
+\end{wstructure}
+
+As presented, our type theory suffers from a major weakness. Indeed,
+we are subject to Girard's paradox, as we assume that $\Set$ lives in
+$\Set$. We made that choice for presentational convenience, as
+universe stratification is orthogonal to our work. Nonetheless, our
+universes of description rather naturally leads itself to
+stratification. Unsurprisingly, $\IDesc{}$ at level $l$ is of type
+$\Set^{\blue{l+1}}$. Similarly, the interpretation of $\IDesc{}$ at
+level $l$ is an object of type $\Set^{\blue{l}}$:
+
+\[\stk{
+\data \IDesc{\!}^{\blue{l}} (\Bhab{\V{I}}{\Set^{\blue{l+1}}}) : \Set^{\blue{l+1}} \where \\
+\;\;\ldots \\
+\\
+\idescop{\_\:}{}{}^{\blue{l}} : \PI{\V{I}}{\Set^{\blue{l+1}}} \IDesc{{\!}^{\blue{l}}I} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
+\ldots
+}\]
+
+Consequently, we can implement the operations and examples developed
+above. We refer the reader to our Agda implementation, which take
+advantage of set polymorphism to implement the universe of indexed
+descriptions at any level. Further, we have coded $\IDesc{}$ in itself
+and have proved the isomorphism between the host and the embedded
+universes.
+
+\subsection{Related Work}
+
+\begin{structure}
+!!! Need Help !!!
+<- Comparison with Induction Recursion
+    ???
 \end{structure}
+
+\begin{wstructure}
+!!! Need Help !!!
+<- Related Work
+    Who? What?
+\end{wstructure}

### 2010-03-25 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: related work: bring references in paper.bib

@@
 <- Related Work
-    Who? What?
+    <- Generic in simply-typed functional languages
+        <- PolyP \cite{jansson:polyp}
+        <- Generic Haskell \cite{hinze:generic-haskell}
+        <- Scratch your boilerplate \cite{spj:syb}
+    <- Generic in dependent types
+        <- Universes for generic prog \cite{benke:universe-generic-prog}
+        <- Norell \cite{norell:msc-thesis}
+        <- Polytypic prog in Coq \cite{verbruggen:polytype-coq}
 \end{wstructure}

### 2010-03-29 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: conclusion done.
Desc@ICFP: related work almost done. Begging for Induction Recursion story.

--- 2010-03-25 papers/icfp-2010-desc/paper_discussion.tex
@@
 <- Related Work
-    Who? What?
+    <- Generic in simply-typed functional languages
+        <- PolyP \cite{jansson:polyp}
+        <- Generic Haskell \cite{hinze:generic-haskell}
+        <- Scratch your boilerplate \cite{spj:syb}
 \end{wstructure}
+
+Generic programming is a vast topic. We refer our reader to Garcia et
+al.~\cite{garcia:generic-comparative-study} for a broad overview of
+generic support in various languages. In the sole context of Haskell,
+there is a myriad of proposals. These approaches are presented and
+compared in Hinze et al.~\cite{hinze:generic-approach-comparative} and
+Rodriguez et al.~\cite{rodriguez:generic-libs-comparative}.
+
+In particular, our approach is similar in spirit with polytypic
+programming, as initiated by PolyP~\cite{jansson:polyp}. Indeed,
+generic functions, in our system, are built by induction on the
+pattern functor. Unlike PolyP, we do not have to resort to
+pre-processing code: our data-types are, natively, nothing but codes.
+
+Our approach also support the Generic
+Haskell~\cite{hinze:generic-haskell} model. This model, based on
+type-indexed data types, corresponds to computing new data-types from
+others. This is natural in our system, as data-types descriptions are
+first-class.
+
+Interestingly, our generic induction principle allows a programming
+style reminiscent to the Scrap Your Boilerplate~\cite{spj:syb} (SYB)
+approach. Like SYB, we provide a generic operator to manipulate and
+carry operations over data-types. However, unlike SYB, we do not need
+support from the compiler, as introduced by the $\CN{Typeable}$
+class. Indeed, in our setting, data-types are reflected in the
+language, hence our ability to manipulate them directly.
+
+\begin{wstructure}
+    <- Generic in dependent types
+        <- Norell \cite{norell:msc-thesis}
+        <- Polytypic prog in Coq \cite{verbruggen:polytype-coq}
+        <- Universes for generic prog \cite{benke:universe-generic-prog}
+\end{wstructure}
+
+The interest in generic programming for dependent types is not new
+either. Norell~\cite{norell:msc-thesis} have shown the benefit of
+polytypic programming in the setting of Alfa, a precursor of
+Agda. Similarly, Verbruggen et al.~\cite{verbruggen:polytype-prog-coq,
+  verbruggen:polytype-coq} have developed a framework for polytypic
+programming in the Coq theorem prover. However, these works aim at
+\emph{modelling} Generic Haskell in a dependently-typed setting, for
+the purpose of proving correctness properties of Haskell code. Our
+approach is different in that we aim at building a foundation for
+data-types in a dependently typed theory.
+
+Closer to us is the work by Benke et
+al.~\cite{benke:universe-generic-prog}. This seminal work introduced
+the usage of universes for developing generic programs. Moreover, our
+own universes are rather similar to theirs: our universe of
+descriptions is similar to their universe of iterated induction, and
+our universe of indexed descriptions is isomorphic to their universe of
+finitary indexed induction. This is not surprising, as we share the
+same source of inspiration, namely induction recursion.
+
+However, we differ in several ways. First, they adopt a generative
+perspective: each universe extends the base type theory with both type
+formers and elimination rules. Thanks to the levitation, we only rely
+on a generic induction and a specialised $\switchD{}{}{}$. Second, the
+authors develop several universes, each with its own strengths and
+weaknesses, but lack a formal, unifying universe. With indexed
+descriptions, we subsume simple descriptions, hence avoiding the move
+to and from several codings. Finally, the authors often resort to an
+extensional notion of equality, while we have given an
+equality-agnostic presentation. Moreover, our presentation is arranged
+so as to use definitional equality as much as possible. Hence, in
+practice, the developer is relieved from many trivial proof
+obligations.

### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: get some sort of introduction.
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)

@@
 \section{A universe of simple data-types}
+\label{sec:universe-desc}
 
@@
 we can find in the literature, depending on the sensibility of the author.
-The notation $\Sigma_{x : A} (B\: x)$ stresses the interpretation that
+The notation $\blue{\Sigma}_{\V{x} : \V{A}} (\V{B}\: \V{x})$ stresses the interpretation that
 $\Sigma$-types are a generalisation of sums over arbitrary arities.  Where
@@
 
-On the other hand, our choice of notation ($\SIGMA{\V{x}}{A} (B\: x)$)
+On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}} (\V{B}\:\V{x})$
 emphasises that $\Sigma$-types can be read as a generalisation of products,
@@
 argument. We notice that this functoriality appears in the type of
-$\descop{D}{}$ itself, for a given data-type definition $D$. This
-corresponds to the object part of the functor described by $D$.
+$\descop{\V{D}}{}$ itself, for a given data-type definition $\V{D}$. This
+corresponds to the object part of the functor described by $\V{D}$.
 
@@
     \DUnit          & \Desc \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
     \DIndx          & \Desc \To \Desc
@@
 \begin{array}{@{}ll@{\:=\:\:}ll}
-\descop{\DUnit}{& X}        &  \Unit                                       \\
-\descop{\DSigma{S}{D}}{& X} &  \SIGMAS{\V{s} : S}{\descop{D~s}{X}}         \\
-\descop{\DIndx{D}}{& X}     &  \TIMES{X}{\descop{D}{X}}
+\descop{\DUnit}{& \V{X}}        &  \Unit                                       \\
+\descop{\DSigma{\V{S}}{\V{D}}}{& \V{X}} &  \SIGMAS{\V{s} : \V{S}}{\descop{\V{D}~\V{s}}{\V{X}}}         \\
+\descop{\DIndx{\V{D}}}{& \V{X}}     &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
 \end{array}
@@
 
-\[    \NatD\: Z \mapsto 1 + Z    \]
+\[    \NatD\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{Z}    \]
 
@@
 \ListD : \Set \To \Desc \\
-\ListD \: X \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
-                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{\DUnit}}) ]}
+\ListD \: \V{X} \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
+                           {[ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) ]}
 }\]
@@
 The modification consists in turning the $\NatSuc{}$ constructor into
-a proper $\ListCons$ taking an argument in X followed by a recursive
+a proper $\ListCons$ taking an argument in $\V{X}$ followed by a recursive
 argument. In this case, we use $\DSigma{}{}$ in its product
-interpretation: we pack an element of $X$ together with the recursive
+interpretation: we pack an element of $\V{X}$ together with the recursive
 argument. Easily, one sees that this code actually describes the
@@
 
-\[    \ListD\: X\: Z \mapsto 1 + X \times Z     \]
+\[    \ListD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \D{\ensuremath{\times}} \V{Z}     \]
 
@@
 \begin{array}{@{}ll}
-\TreeD \: X \mapsto \DSigma{ & (\EnumT{[ \TreeLeaf, \TreeNode ]}) \\}
-                           { & [ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
+\TreeD \: \V{X} \mapsto \DSigma{ & (\EnumT{[ \TreeLeaf, \TreeNode ]}) \\}
+                           { & [ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
 \end{array}
@@
 
-\[    \TreeD\: X\: Z \mapsto 1 + X \times Z \times Z     \]
+\[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \Prod \V{Z} \Prod \V{Z}     \]
 
@@
 necessary. A tagged description can trivially be turned into a
-full-blown description. We note $\toDesc{TD}$ the description computed
-from the tagged description $TD$.
+full-blown description. We write $\toDesc{\V{TD}}$ for the description computed
+from the tagged description $\V{TD}$.
 
@@
 \[
-\Rule{\Gamma \vdash \Bhab{D}{\Desc}}
-     {\Gamma \vdash \Bhab{\Mu{D}}{\Set}} \qquad
-\Rule{\Gamma \vdash \Bhab{D}{\Desc} \quad 
-      \Gamma \vdash \Bhab{x}{\descop{D}{(\Mu{D})}}}
-     {\Gamma \vdash \Bhab{\Con{x}}{\Mu{D}}}
+\Rule{\Gamma \vdash \Bhab{\V{D}}{\Desc}}
+     {\Gamma \vdash \Bhab{\Mu{\V{D}}}{\Set}} \qquad
+\Rule{\Gamma \vdash \Bhab{\V{D}}{\Desc} \quad 
+      \Gamma \vdash \Bhab{\V{x}}{\descop{\V{D}}{(\Mu{\V{D}})}}}
+     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\Mu{\V{D}}}}
 \]
@@
 \[
-\F{cata} : \PITEL{D}{\Desc}
-           \PI{T}{\Set}
-           (\descop{D}{T} \To T) \To 
-           \Mu{D} \To T 
+\F{cata} : \PITEL{\V{D}}{\Desc}
+           \PI{\V{T}}{\Set}
+           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
+           \Mu{\V{D}} \To \V{T} 
 \]
@@
 
-The first step consists in turning $T$ into a predicate
-$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
-descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
-\To T$ by a similar process. We start with turning the arrow into a
-dependent arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
+The first step consists in turning $\V{T}$ into a predicate
+$\Bhab{\V{P}}{\Mu{\V{D}} \To \Set}$, hence introducing a dependency on the
+descriptions. Consequently, we strengthen the algebra $\descop{\V{D}}{\V{T}}
+\To \V{T}$ by a similar process. We start with turning the arrow into a
+dependent arrow, hence offering $\Bhab{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}$ to
 further analysis.
@@
 Then, we have to translate the notion of algebra in our dependent
-setting. This is achieved by the predicate $\All{D}{X}{P}{x}$
+setting. This is achieved by the predicate $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$
 (Fig.~\ref{fig:all-predicates}) that, intuitively, states that
-$\Bhab{P}{X \To \Set}$ holds everywhere in $x$. Based on that
-knowledge, we ought to be able to prove that $P$ holds for
-$\Con{xs}$. We close our elimination principle by taking an argument
-$x$ in $\Mu{D}$ and claiming that the motive $P$ is realised on $x$.
+$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Based on that
+knowledge, we ought to be able to prove that $\V{P}$ holds for
+$\Con{\V{xs}}$. We close our elimination principle by taking an argument
+$x$ in $\Mu{\V{D}}$ and claiming that the motive $\V{P}$ is realised on $\V{x}$.
 
@@
 \begin{array}{lcll}
-\F{induction} & : & \multicolumn{2}{l}{\PITEL{D}{\Desc}}                   \\
-              &   & \multicolumn{2}{l}{\PITEL{P}{\Mu{D} \To \Set}}         \\
-              &   & \PITEL{m}{& \PI{xs}{\descop{D}{(\Mu{D})}}              \\
-              &   &           & \All{D}{(\Mu{D})}{P}{xs} \To P (\Con{xs})} \\
-              &   & \multicolumn{2}{l}{\PI{x}{\Mu{D}} P x}
+\F{induction} & : & \multicolumn{2}{l}{\PITEL{\V{D}}{\Desc}}                   \\
+              &   & \multicolumn{2}{l}{\PITEL{\V{x}}{\Mu{\V{D}}}}                   \\
+              &   & \multicolumn{2}{l}{\PITEL{\V{P}}{\Mu{\V{D}} \To \Set}}         \\
+              &   & \PI{\V{m}}{& \PI{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}              \\
+              &   &           & \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{xs}} \To \V{P} (\Con{\V{xs}})} \\
+              &   & \V{P} \V{x}
 \end{array}
@@
                  \PITEL{\V{P}}{\V{X} \To \Set} \\
-               & \PI{\V{xs}}{\descop{D}{\V{X}}} 
+               & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                  \Set 
 \end{array} \\
-\All{\DId}{X}{P}{x} \mapsto P\: x \\
-\All{(\DConst{Z})}{X}{P}{x} \mapsto Z \\
-\All{(\DProd{D}{D'})}{X}{P}{\pair{d}{d'}{}} \mapsto \TIMES{\All{D}{X}{P}{d}}{\All{D'}{X}{P}{d'}} \\
-\All{(\DSigma{S}{T})}{X}{P}{\pair{a}{b}{}} \mapsto \All{(T\: a)}{X}{P}{b} \\
-\All{(\DPi{S}{T})}{X}{P}{f} \mapsto \PI{\V{s}}{S} \All{(T\: a)}{X}{P}{(f\: s)}
+\All{\DId}{\V{X}}{\V{P}}{\V{x}} \mapsto \V{P}\: \V{x} \\
+\All{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{z}} \mapsto \Unit \\
+\All{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\pair{\V{d}}{\V{d'}}{}} \mapsto \TIMES{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}{\All{\V{D'}}{\V{X}}{\V{P}}{\V{d'}}} \\
+\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} \mapsto \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
+\All{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{f}} \mapsto \PI{\V{\V{s}}}{\V{S}} \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{(\V{f}\: \V{s})}
 }
@@
                    \PITEL{\V{P}}{\V{X} \To \Set} \\
-                 & \PITEL{\V{R}}{\PI{\V{x}}{X} P\: x}
-                   \PI{\V{xs}}{\descop{D}{\V{X}}} 
-                   \All{D}{X}{P}{xs} 
+                 & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
+                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
+                   \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
 \end{array} \\
-\all{\DId}{X}{P}{R}{x} = R x \\
-\all{(\DConst{Z})}{X}{P}{R}{x} = z \\
-\all{(\DProd{D}{D'})}{X}{P}{R}{\pair{d}{d'}{}} = \pair{\all{D}{X}{P}{R}{d}}{\all{D'}{X}{P}{R}{d'}}{} \\
-\all{(\DSigma{S}{T})}{X}{P}{R}{\pair{a}{b}{}} = \all{(T\: a)}{X}{P}{R}{b} \\
-\all{(\DPi{S}{T})}{X}{P}{R}{f} = \LAM{s} \all{(T\: s)}{X}{P}{R}{(f\: s)}
+\all{\DId}{\V{X}}{\V{P}}{\V{R}}{\V{x}} = \V{R}\: \V{x} \\
+\all{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{R}}{\V{z}} = \Void \\
+\all{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{d}}{\V{d'}}{}} = \pair{\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{d}}}{\all{\V{D'}}{\V{X}}{\V{P}}{\V{R}}{\V{d'}}}{} \\
+\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} = \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
+\all{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\V{f}} = \LAM{\V{s}} \all{(\V{T}\: \V{s})}{\V{X}}{\V{P}}{\V{R}}{(\V{f}\: \V{s})}
 \end{array}
@@
 \[
-\Rule{\forall e_i \in \vec{e}, \propag{\push{e_i}
-                                            {\descop{f\: i}{(\Mu{(\DSigma{(\Enum\: b)}{f})})}}}
-                                      {e'_i}}
-     {\propag{\push{C\: \vec{e}}{\Mu{(\DSigma{(\Enum\: b)}{f})}}}
-             {\Con{(\etag{C}, \widetilde{e'})}}}\;\etag{C} \in b
+\Rule{\forall \V{e_i} \in \V{\vec{e}}, \propag{\push{\V{e_i}}
+                                            {\descop{\V{f}\: \V{i}}{(\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})})}}}
+                                      {\V{e'_i}}}
+     {\propag{\push{\V{C}\: \V{\vec{e}}}{\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})}}}
+             {\Con{(\etag{C}, \V{\widetilde{e'}})}}}\;\etag{t} \in \V{b}
 \]

### 2010-03-29 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix typos. Clean-up solved \notes.
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)

--- 2010-03-18 papers/icfp-2010-desc/paper_desc_levitation.tex
@@
 %% ConsE
-\Rule{\Gamma \vdash \Bhab{t}{\UId} \quad
-      \Gamma \vdash \Bhab{e}{\EnumU}}
-     {\Gamma \vdash \Bhab{\ConsE{t}{e}}{\EnumU}}
+\Rule{\Gamma \vdash \Bhab{\V{t}}{\UId} \quad
+      \Gamma \vdash \Bhab{\V{e}}{\EnumU}}
+     {\Gamma \vdash \Bhab{\ConsE{\V{t}}{\V{e}}}{\EnumU}}
 }\]
@@
 
-Our first tentative is the following:
+Our first tentative attempt is the following:
 
@@
                                    \DUnit                                \\
-                                   \DSigma{\Set}{(\LAM{\V{S}} ???)}      \\
+                                   \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
                                    \DIndx{\DUnit}                        \\
@@
 
-\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc      \]
+\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc      \]
 
-So, we first pack an element $S$ of $\Set$. Then, we would like to
-express a notion of recursive argument \emph{indexed} by $S$. Because
+So, we first pack an element $\V{S}$ of $\Set$. Then, we would like to
+express a notion of recursive argument \emph{indexed} by $\V{S}$. Because
 our presentation is entirely first-order so far, we are not able to
@@
 extend our universe of descriptions. This consists in adding a code
-$\DHindx$ that takes an indexing set $H$. The code and its
+$\DHindx$ that takes an indexing set $\V{H}$. The code and its
 interpretation are presented in
@@
     \ldots          & \:\:\ldots \\
-    \DHindx         & \PI{H}{\Set} \Desc \To \Desc
+    \DHindx         & \PI{\V{H}}{\Set} \Desc \To \Desc
 \end{array}
@@
 \ldots                        &  \ldots \\
-\descop{\DHindx{H}{D}}{X}     &  \TIMES{(H \To X)}{\descop{D}{X}}
+\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}}
 \end{array}
@@
 successfully describe the $\DSigma$ case, by a simple appeal to the
-higher-order induction on $S$. The $\DHindx$ case consists in packing
-the $H$ in $\Set$ with a recursive argument.
+higher-order induction on $\V{S}$. The $\DHindx$ case consists in packing
+the $\V{H}$ in $\Set$ with a recursive argument.
 
@@
                                    \DUnit                                            \\
-                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{S}{\DUnit})}   \\
+                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{\V{S}}{\DUnit})}   \\
                                    \DIndx{\DUnit}                                    \\
@@
 codes. Then, we can type-check the case definition, between square
-brackets $[ \ldots ]$. This term $f$ is pushed into the type
-$\EnumT{e} \To \Desc$. This corresponds to a finite function
+brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
+$\EnumT{\V{e}} \To \Desc$. This corresponds to a finite function
 definition, as formalised in Figure~\ref{fig:type-checking}.
@@
 \[
-\PLAM{x}{(\EnumT{e})} \switch{e}{(\LAM{\_} \Mu{\DescD})}{\pi^f}{x}
+\PLAM{\V{x}}{(\EnumT{\V{e}})} \switch{\V{e}}{(\LAM{\_} \Mu{\DescD})}{\V{\pi^f}}{\V{x}}
 \]
@@
 \F{switchD} : & \PITEL{\V{e}}{\EnumU}               
-                \PITEL{\V{b}}{\spi{e}{\LAM{\_} \Desc}}
-                \PITEL{\V{x}}{\EnumT{e}} \To \Desc
+                \PITEL{\V{b}}{\spi{\V{e}}{\LAM{\_} \Desc}}
+                \PITEL{\V{x}}{\EnumT{\V{e}}} \To \Desc
 \end{array}
@@
 \[
-\Rule{\Gamma \Vdash \propag{\push{t}{\spiD{e}}}
-                           {t'}}
+\Rule{\Gamma \Vdash \propag{\push{\V{t}}{\spiD{\V{e}}}}
+                           {\V{t'}}}
      {\Gamma \Vdash \begin{array}{@{}l}
-                        \propag{\push{t}{\EnumT{e} \To \Desc}}
-                               {\\ \PLAM{x}{(\EnumT{e})} \switchD{e}{t'}{x}}
+                        \propag{\push{\V{t}}{\EnumT{\V{e}} \To \Desc}}
+                               {\\ \PLAM{\V{x}}{(\EnumT{\V{e}})} \switchD{\V{e}}{\V{t'}}{\V{x}}}
                     \end{array}
-     }\;\mbox{t is $[]$ or $[a,b]$}
+     }\;\mbox{$\V{t}$ is $[]$ or $[\_,\_]$}
 \]
@@
 The catamorphism can be characterised as an induction on the
-description $D$, with a non-dependent motive targeting $T$. Given a
-node $xs$ and the induction hypotheses, the method ought to build an
-element of $T$. Provided that we know how to make an element of
-$\descop{D}{T}$, this step will be performed by the algebra $f$. Let
+description $\V{D}$, with a non-dependent motive targeting $\V{T}$. Given a
+node $\V{xs}$ and the induction hypotheses, the method ought to build an
+element of $\V{T}$. Provided that we know how to make an element of
+$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra $\V{f}$. Let
 us take a look at this jigsaw:
@@
 \[\stk{
-\cata : \PITEL{D}{\Desc}
-           \PI{T}{\Set}
-           (\descop{D}{T} \To T) \To 
-           \Mu{D} \To T \\
-\cata\: D\: T\: f \mapsto
-  \F{induction}\: D\: (\LAM{\_}T)\: (\LAM{xs\:hs} f\: ???)
+\cata : \PITEL{\V{D}}{\Desc}
+           \PI{\V{T}}{\Set}
+           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
+           \Mu{\V{D}} \To \V{T} \\
+\cata\: \V{D}\: \V{T}\: \V{f\: \V{x}} \mapsto
+  \F{induction}\: \V{D}\:\V{x}\: (\LAM{\_}\V{T})\: (\LAM{\V{xs}\:\V{hs}} \V{f}\: \SHED)
 }\]
 
-We are left with implementing the \(???\) program. Recall that we have
-\(\Bhab{xs}{\descop{D}{\Mu{D}}}\) and
-\(\Bhab{hs}{\All{D}{(\Mu{D})}{(\LAM{\_} T)}{xs}}\) at hand. Our goal
-is to make an element of \(\descop{D}{T}\). Intuitively, $xs$ is of
+We are left with implementing filling the hole. Recall that we have
+\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
+\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\) at hand. Our goal
+is to make an element of \(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of
 the right shape, but its sub-elements are of the wrong type. On the
-other hand, for each sub-element of $xs$, $hs$ gives us the
-corresponding element in $T$. Hence the name ``induction
-hypotheses''. Therefore, to construct an element of \(\descop{D}{T}\),
-we replace the recursive components of \(xs\) by their counterpart in
-\(hs\):
+other hand, for each sub-element of $\V{xs}$, $\V{hs}$ gives us the
+corresponding element in $\V{T}$. Hence the name ``induction
+hypotheses''. Therefore, to construct an element of \(\descop{\V{D}}{\V{T}}\),
+we replace the recursive components of \(\V{xs}\) by their counterpart in
+\(\V{hs}\):
 
 \[\stk{
-\F{replace} : \stk{\PITEL{D}{\Desc}
-                   \PITEL{X,Y}{\Set}\\
-                   \PI{xs}{\descop{D}{X}} 
-                   \All{D}{X}{(\LAM{\_}Y)}{xs} \To
-                   \descop{D}{Y}} \\
-\F{replace}\: \DUnit\:          X\: Y\: \Void\:          \Void          \mapsto \Void \\
-\F{replace}\: (\DSigma{S}{D})\: X\: Y\: \pair{s}{xs}{}\: ys             \mapsto
-    \pair{s}{\F{replace}\: {D~s}\: X\: Y\: xs\: ys}{}                                 \\
-\F{replace}\: (\DIndx{D})\:     X\: Y\: \pair{x}{xs}{}\: \pair{y}{ys}{} \mapsto
-    \pair{y}{\F{replace}\: D\: X\: Y\: xs\: ys}{}
+\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
+                   \PITEL{\V{X},\V{Y}}{\Set}\\
+                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
+                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
+                   \descop{\V{D}}{\V{Y}}} \\
+\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto \Void \\
+\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
+    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
+\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto
+    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}
 }\]
 
-Filling the \(???\) with \(\F{replace}\: D\: (\Mu{D})\: T\: xs\: hs\) closes the
+Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\: (\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the
 problem. In the type theory, we have built a generic catamorphism. Any
@@
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \FMVar           & X \To \FMFreeMonad\: F\: X                            \\
-    \FMComposite     & F (\FMFreeMonad\: F\: X) \To \FMFreeMonad\: F\: X    
+    \FMVar           & \V{X} \To \FMFreeMonad\: \V{F}\: \V{X}                            \\
+    \FMComposite     & \V{F} (\FMFreeMonad\: \V{F}\: \V{X}) \To \FMFreeMonad\: \V{F}\: \V{X}    
 \end{array}
@@
 \[
-\FMFreeMonadD\: F\: X\: Z \mapsto X + F Z
+\FMFreeMonadD\: \V{F}\: \V{X}\: \V{Z} \mapsto \V{X} \mathop{\D{+}} \V{F} \V{Z}
 \]
@@
 description. The free monad construction will take such tagged
-description, a set $X$ of variables, and will compute the tagged
+description, a set $\V{X}$ of variables, and will compute the tagged
 description of the corresponding free monad. Implementing this
@@
 \FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
-\FreeMonad{\pair{E}{D}{}}\:X \mapsto
-    \pair{\ListCons{\DVar{}}{E}}
-         {\pair{\DSigma{X}{\DUnit}}{D}{}}{}
+\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
+    \pair{\ListCons{\DVar{}}{\V{E}}}
+         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
 }\]
@@
 We simply add a constructor, $\DVar{}$, and define its argument to be
-a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
+a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep $\V{E}$ and $\V{D}$
 as they were, hence leaving the functor unchanged. Unfolding the
@@
 
-\note{This is not quite Mu of a tagged description}
-
 \[\stk{
@@
            \PI{\V{X}, \V{Y}}{\Set} \\
-         & \Mu{(\toDesc{\FreeMonad{D}{X}})} \To
-           (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{Y}})}) \To
-           \Mu{(\toDesc{\FreeMonad{D}{Y}})} 
+         & \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
+           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To
+           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
 \end{array} \\
-\subst\: D\: X\: Y\: x\: \sigma \mapsto
-  \cata\: (\toDesc{\FreeMonad{D}{X}})\: 
-          (\Mu{(\toDesc{\FreeMonad{D}{Y}})})\: 
-          ???\: 
-          x
+\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{x}\: \V{\sigma} \mapsto
+  \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
+          (\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})})\: 
+          \SHED\: 
+          \V{x}
 }\]
@@
 catamorphism. Intuitively, its role is to catch appearances of
-$\DVar{x}$ and replace them by $\sigma x$. Otherwise, it should simply
+$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. Otherwise, it should simply
 appeal to the induction hypothesis. This corresponds to the following
@@
            \PI{\V{X}, \V{Y}}{\Set} \\
-         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{X}})}) \To
-           \descop{\toDesc{\FreeMonad{D}{X}}}{\Mu{(\toDesc{\FreeMonad{D}{Y}})}} \To
-           \Mu{(\toDesc{\FreeMonad{D}{Y}})}
+         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To
+           \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
+           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
 \\
@@
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}   & \sigma\: x                   \\
-\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
+\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
+\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{ys}}{} & \Con{\pair{\etag{t}}{\V{ys}}{}}
 \end{array}
@@
 
-Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$
+Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}$
 completes the implementation. To sum up, we have implemented the free

### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)

@@
 an empty vector: the head function only takes vectors of length
-$\NatSuc{n}$, as enforced by its type. This is made possible by the
+$\NatSuc{\V{n}}$, as enforced by its type. This is made possible by the
 very specificity of dependently-typed systems: terms -- the length --
@@
 $\IDesc$, the universe of indexed descriptions. We express inductive families
-by the $I \To \IDesc{I}$ type.  \note{pwm: Vec might not be the best
+by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
 motivating example here, since it doesn't appear in the paper (more on that
@@
 
-$$\Bhab{\Sigma}{\PI{\V{S}}{\Set} (S \To \Set) \To \Set}$$
+$$\Bhab{\Sigma}{\PI{\V{S}}{\Set} \PI{\V{S}}{\Set} \Set}$$
 
@@
 that we replace $\DUnit$ by a more general $\DConst$ code. Whereas
-$\DUnit$ was interpreted as the unit set, $\DConst{X}$ is interpreted
-as $X$, for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
-$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
+$\DUnit$ was interpreted as the unit set, $\DConst{\V{X}}$ is interpreted
+as $\V{X}$, for any $\Bhab{\V{X}}{\Set}$. Extensionally, $\DConst{\V{X}}$ and
+$\DSigma{\V{X}}{\DUnit}$ are equivalent. However, $\DConst$ is more
 succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
@@
 \[\begin{array}{l@{\:\mapsto\:\:}l}
-\DIndx{D}         & \DProd{\DId}{D}                      \\
-\DHindx{H}{D}     & \DProd{(\DPi{H}{(\LAM{\_} \DId)})}{D}
+\DIndx{\V{D}}         & \DProd{\DId}{\V{D}}                      \\
+\DHindx{\V{H}}{\V{D}}     & \DProd{(\DPi{\V{H}}{(\LAM{\_} \DId)})}{\V{D}}
 \end{array}
@@
     \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
-    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
+    \DPi            & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc 
 \end{array}
@@
 \begin{array}{@{}l@{\:=\:\:}ll}
-\descop{\DId}{X}          &  X                                           \\
-\descop{\DConst{Z}}{X}    &  Z                                           \\
-\descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
-\descop{\DSigma{S}{D}}{X} &  \SIGMA{\V{s}}{S} \descop{D\: s}{X}          \\
-\descop{\DPi{S}{D}}{X}    &  \PI{\V{s}}{S} \descop{D\: s}{X}            
+\descop{\DId}{\V{X}}          &  \V{X}                                           \\
+\descop{\DConst{\V{Z}}}{\V{X}}    &  \V{Z}                                           \\
+\descop{\DProd{\V{D}}{\V{D'}}}{\V{X}} &  \TIMES{\descop{\V{D}}{\V{X}}}{\descop{\V{D}\V{'}}{\V{X}}}       \\
+\descop{\DSigma{\V{S}}{\V{D}}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}          \\
+\descop{\DPi{\V{S}}{\V{D}}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}            
 \end{array}
@@
 inductive types in this setting. To describe an inductive family
-indexed by $\Bhab{I}{\Set}$, we use endofunctors on the category
-$\Set^I$. We call these \emph{indexed functors}. \note{Predicate
-  transformer anyone?} In this setting, $I \To \IDesc{I}$ is our
+indexed by $\Bhab{\V{I}}{\Set}$, we use endofunctors on the category
+$\Set^{\V{I}}$. We call these \emph{indexed functors}. \note{Predicate
+  transformer anyone?} In this setting, $\V{I} \To \IDesc{\V{I}}$ is our
 grammar for describing these functors. Hence, $\IDesc{}$ and its
@@
 \\
-\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set    \\
+\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
 \ldots
@@
 
-Note that $I \To \IDesc{I}$ is isomorphic, up to interpretation, to an
-endofunctor on $\Set^I$. Indeed, $I \To \IDesc{I}$ is interpreted as
-$I \To (I \To \Set) \To \Set$, which is isomorphic to $(I \To \Set)
-\To (I \To \Set)$. Fix-point of inductive families is defined over
+Note that $I \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
+endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
+$\V{I} \To (\V{I} \To \Set) \To \Set$, which is isomorphic to $(\V{I} \To \Set)
+\To (\V{I} \To \Set)$. Fix-point of inductive families is defined over
 these indexed functors, hence computing a fix-point of the entire
@@
 \Rule{\stk{ 
-          \Gamma \vdash \Bhab{I}{\Set} \\
-          \Gamma \vdash \Bhab{R}{I \To \IDesc{I}} \\
-          \Gamma \vdash \Bhab{i}{I}}}
-     {\Gamma \vdash \Bhab{\IMu{I}{R}{i}}{\Set}} \qquad
+          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
+          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
+          \Gamma \vdash \Bhab{\V{i}}{\V{I}}}}
+     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\V{i}}}{\Set}} \qquad
 \Rule{\stk{
-          \Gamma \vdash \Bhab{I}{\Set} \\
-          \Gamma \vdash \Bhab{R}{I \To \IDesc{I}} \\
-          \Gamma \vdash \Bhab{i}{I} \\
-          \Gamma \vdash \Bhab{x}{\idescop{R\: i}{I}{(\IMu{I}{R}{})}}}}
-     {\Gamma \vdash \Bhab{\Con{x}}{\IMu{I}{R}{i}}}
+          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
+          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
+          \Gamma \vdash \Bhab{\V{i}}{\V{I}} \\
+          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{})}}}}
+     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}}
 }\]
@@
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \DVar{}         & I \To \IDesc{I}                                   \\
-    \DConst{}       & \Set \To \IDesc{I}                                \\
-    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
-    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
+    \DVar{}         & \V{I} \To \IDesc{\V{I}}                                   \\
+    \DConst{}       & \Set \To \IDesc{\V{I}}                                \\
+    \DProd{}{}      & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}           \\
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
+    \DPi            & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
 \end{array}
@@
 \stk{
-\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{I} \To (\V{I} \To \Set) \To \Set                  \\
+\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
 \begin{array}{@{}l@{\:=\:\:}ll}
-\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
-\idescop{\DConst{X}}{I}{P}    &  X                                                   \\
-\idescop{\DProd{D}{D'}}{I}{P} &  \TIMES{\idescop{D}{I}{P}}{\idescop{D'}{I}{P}}       \\
-\idescop{\DSigma{S}{D}}{I}{P} &  \SIGMA{\V{s}}{S} \idescop{D\: s}{I}{P}                    \\
-\idescop{\DPi{S}{D}}{I}{P}    &  \PI{\V{s}}{S} \idescop{D\: s}{I}{P}            
+\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
+\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
+\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
+\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
+\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
 \end{array}
@@
 \[\stk{
- \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}} \\
- \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
- \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})}) 
+ \TagIDesc{\V{I}}  \mapsto \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
+ \ATagIDesc{\V{I}} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
+ \ITagIDesc{\V{I}} \mapsto \\
+\qquad\SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
 }\]
@@
 \begin{array}{@{}ll}
-\VecD\: X = \bigRedBracket{ 
+\VecD\: \V{X} = \\ \quad\bigRedBracket{ 
             \begin{array}{l}
-                \pair{[]}{\_}{} \\
+                \pair{[]}{\_}{} 
+                \\\\
                 \bigRedBracket{
                 \begin{array}{l} 
-                  \LAM{n} \caseB{n}{\NatZero \To [\VecNil] \\
-                                   (\NatSuc n) \To [\VecCons{}{}]} \\
-                  \LAM{n} \caseB{n}{\NatZero \To [\DConst{Unit}] \\
-                                   (\NatSuc n) \To [\DSigma{X}{\DVar{n}}]}
+                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
+                                   (\NatSuc \V{n}) \mapsto [\VecCons{}{}]} \\
+                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
+                                   (\NatSuc \V{n}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
                 \end{array}}
@@
 \begin{array}{@{}ll}
-\IDescD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
+\IDescD~\V{I} \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
                                           \DConst{},
@@
                               & \bigRedBracket{\begin{array}{l}
-                                      \DConst{I}                  \\
+                                      \DConst{\V{I}}                  \\
                                       \DConst{\Set}               \\
                                       \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
-                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
-                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
+                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
+                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}
@@
 \begin{array}{l@{\::\:\:}l}
-\Econd{}{}{}     & \forall \Bhab{ty}{\Ty} . \Ebool \To ty \To ty \To ty  \\ 
+\Econd{}{}{}     & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
 \Eplus{}{}       & \Enat \To \Enat \To \Enat                           \\
 \Ele{}{}         & \Enat \To \Enat \To \Ebool                          \\
-\Eval{}          & \forall \Bhab{ty}{\Ty} . \Val{ty} \To ty
+\Eval{}          & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
 \end{array}
@@
                    \EnumT{[\Eval{}, \Econd{}{}{}]} \red{,} \\
-                   \LAM{ty}
+                   \LAM{\V{ty}}
                    \bigRedBracket{
                    \begin{array}{l}
-                   \DConst{(\Val{ty})} \\
-                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{ty}}{\DVar{ty}}}
+                   \DConst{(\Val{\V{ty}})} \\
+                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                    \end{array}
@@
                  \begin{array}{l}
-                   \LAM{ty} \case{ty}{\Enat \To \EnumT{[\Eplus{}{}]} \\ \Ebool \To \EnumT{[\Ele{}{}]}} \red{,} \\
-                   \LAM{ty}
-                   \case{ty}{
-                     \Enat \To \DProd{\DVar{nat}}{\DVar{nat}} \\
-                     \Ebool \To \DProd{\DVar{nat}}{\DVar{nat}}
-                   }
+                   \LAM{\V{t}y} \case{\V{t}y}{\Enat \To \EnumT{[\Eplus{\!}{\!}\!]} \\ \Ebool \To \EnumT{[\Ele{\!}{\!}\!]}} \red{,} \\
+                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                    \end{array}
@@
 \evalH : \PI{\V{ty}}{\Ty} 
-         \IMu{\Ty}{\toIDesc{\HExprD}}{ty} \To
-         \Val{ty}
+         \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
+         \Val{\V{ty}}
 \]
@@
 \[\stk{
-\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\: ty}{\Ty}{ty} \To {\Val{ty}} \\
+\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\: \V{ty}}{\Ty}{\V{ty}}
+\To {\Val{\V{ty}}} \\
 \begin{array}{@{}l@{\:=\:\:}l}
-\evalOne\: \_\: \pair{\Eval{}}{x}{}                                         & x \\
-\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{x}{\_}{}}{}}{}   & x \\
-\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{y}{}}{}}{}  & x \\
-\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{x}{y}{}}{}                       & \F{plusHost}\: x\: y \\
-\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{x}{y}{}}{}                        & \F{leHost}\: x\: y 
+\evalOne\: \_\: \pair{\Eval{}}{\V{x}}{}                                         & \V{x} \\
+\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
+\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{x} \\
+\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{\V{x}}{\V{y}}{}}{}                       & \F{plusHost}\: \V{x}\: \V{y} \\
+\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{\V{x}}{\V{y}}{}}{}                        & \F{leHost}\: \V{x}\: \V{y} 
 \end{array} \\
@@
 \evalH : \PI{\V{ty}}{\Ty} 
-           \IMu{\Ty}{\toIDesc{\HExprD}}{ty} \To
-           \Val{ty} \\
-\evalH\: ty\: term = \F{cataI}\: \Ty\: 
+           \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
+           \Val{\V{ty}} \\
+\evalH\: \V{ty}\: \V{term} = \F{cataI}\: \Ty\: 
                                  \toIDesc{\HExprD}\: 
@@
                                  \evalOne\: 
-                                 ty\: 
-                                 term
+                                 \V{ty}\: 
+                                 \V{term}
 }\]
@@
 \Var{}{} : \EnumU \To \Ty \To \Set \\
-\Var{dom}{\_} = \EnumT{dom}
+\Var{\V{dom}}{\_} = \EnumT{\V{dom}}
 }\]
@@
 contains a pointer to a particular element of the finite set -- an
-element of $\EnumT{dom}$. The extra argument to $\Var{}{}$ is the
+element of $\EnumT{\V{dom}}$. The extra argument to $\Var{}{}$ is the
 domain of this context.
 
-Consequently, replacing $\Val{ty}$ by
-$(\SUM{\Val{ty}}{\Var{dom}{ty}})$ in Figure~\ref{fig:hexpr-full} turns
+Consequently, replacing $\Val{\V{ty}}$ by
+$(\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})$ in Figure~\ref{fig:hexpr-full} turns
 the language of closed term into a language of opened terms with
-constants. For readability, we will abbreviate $\LAM{ty}
-\SUM{\Val{ty}}{\Var{dom}{ty}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
+constants. For readability, we will abbreviate $\LAM{\V{ty}}
+\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
   it polite and comprehensible to ask that?} This defines a new
@@
                \PITEL{\V{dom}}{\EnumU} \\
-             & \PI{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})}} \\
-             & (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To
-               \IMu{\Ty}{\toIDesc{\HExprD}}{ty} 
+             & \PI{\V{\gamma}}{\spi{\V{dom}}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})}} \\
+             & (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To
+               \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} 
 \end{array} \\
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\discharge\: ty\: dom\: \gamma\: (\SumLeft x)  & \Con{\pair{\Eval{}}{x}{}} \\
-\discharge\: ty\: dom\: \gamma\: (\SumRight v) & \switch{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})}{\gamma}{v}
-\end{array}
+\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumLeft \V{x})  & \Con{\pair{\Eval{}}{\V{x}}{}} \\
+\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumRight \V{v}) &
+\end{array}\\
+\qquad\switch{\V{dom}}{(\LAM{\_}
+\IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})}{\V{\gamma}}{\V{v}}
 }\]
@@
 \newcommand{\domBool}{dom_{\CN{bool}}}
-\newcommand{\gammaNat}{\gamma_{\CN{nat}}}
-\newcommand{\gammaBool}{\gamma_{\CN{bool}}}
-\newcommand{\GammaSpi}[2]{\Gamma_{\CN{ty}}~#1~#2}
+\newcommand{\gammaNat}{\V{\ensuremath{\gamma_{\CN{nat}}}}}
+\newcommand{\gammaBool}{\V{\ensuremath{\gamma_{\CN{bool}}}}}
+\newcommand{\GammaSpi}[2]{\F{\ensuremath{\Gamma_{\CN{ty}}}}~#1~#2}
 
@@
 \substH  : & \PITEL{\V{\dom}}{\EnumU} \\
-           & \PITEL{\V{\gammaNat}}{\GammaSpi{dom}{\Enat}} 
-             \PITEL{\V{\gammaBool}}{\GammaSpi{dom}{\Ebool}} \\
+           & \PITEL{\V{\gammaNat}}{\GammaSpi{\V{dom}}{\Enat}} 
+             \PITEL{\V{\gammaBool}}{\GammaSpi{\V{dom}}{\Ebool}} \\
            & \begin{array}{@{}ll}
@@
                              & \PITEL{\V{\gammaNat}}
-                                     {\GammaSpi{dom}{\Enat}} \\
+                                     {\GammaSpi{\V{dom}}{\Enat}} \\
                              & \PITEL{\V{\gammaBool}}
-                                     {\GammaSpi{dom}{\Ebool}} \\
-                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To \\
-                             & \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}}
+                                     {\GammaSpi{\V{dom}}{\Ebool}} \\
+                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To \\
+                             & \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}}}
              \end{array}\\
           & \PI{\V{ty}}{\Ty}
-            \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{\V{ty}} \To \\
+            \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}} \To \\
           & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
@@
 \GammaSpi{}{} : \EnumU \To \Ty \To \Set                                             \\
-\GammaSpi{dom}{ty} = \spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})} 
+\GammaSpi{\V{dom}}{\V{ty}} = \spi{\V{dom}}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})} 
 }\]
@@
 \[
-\sigma\: dom\: \gammaNat\: \gammaBool\: ty\: var \mapsto 
-    \discharge\: ty\: dom\: \gamma_{ty}\: var 
+\F{$\sigma$}\: \V{dom}\: \gammaNat\: \gammaBool\: \V{ty}\: \V{var} \mapsto
+    \F{$\discharge$}\: \V{ty}\: \V{dom}\: \F{$\gamma$}_{\V{ty}}\: \V{var} 
 \]
 
-Where $\gamma_{ty}$ is short for
- $\caseB{ty}{\Enat  \To \gammaNat \\
+Where $\F{$\gamma$}_{\V{ty}}$ is short for
+ $\caseB{\V{ty}}{\Enat  \To \gammaNat \\
                  \Ebool \To \gammaBool}$
@@
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \FinZero_n      & \Fin{(\NatSuc{n})}   \\
-    \FinSuc{}_n     & \Fin{n} \To \Fin{(\NatSuc{n})}
+    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
+    \FinSuc{}_{\V{n}}     & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
 \end{array}
@@
 \begin{array}{@{}ll}
-\FinD\: n = \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
+\FinD\: \V{n} = \DSigma{}{} \!\!\!\!\!\!& (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                         & \bigRedBracket{
                           \begin{array}{l}
-                            \DSigma{\Nat}{\LAM{m} \DConst{(n \PropEq \NatSuc{m})}} \\
-                            \DSigma{\Nat}{\LAM{m} \DProd{\DVar{m}}{\DConst{(n \PropEq \NatSuc{m})}}}
+                            \DSigma{\Nat}{\LAM{\V{m}} \DConst{(\V{n} \PropEq \NatSuc{\V{m}})}} \\
+                            \DSigma{\Nat}{\LAM{\V{m}} \DProd{\DVar{\V{m}}}{\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}
                           \end{array}
@@
 Intuitively, the constraint is captured as follows. First, we store an
-element $m$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
-stipulates that $m$ cannot be \emph{any} natural numbers: it must be
-``the index minus one''. This translates into the constraint $n
-\PropEq \NatSuc{m}$.
+element $\V{m}$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
+stipulates that $\V{m}$ cannot be \emph{any} natural numbers: it must be
+``the index minus one''. This translates into the constraint $\V{n}
+\PropEq \NatSuc{\V{m}}$.
 
@@
 we are \emph{storing} the index of the inductive family, through
-$m$. Hence the constraints. However, \emph{inductive families need not
+$\V{m}$. Hence the constraints. However, \emph{inductive families need not
   store their indices}~\cite{brady:index-inductive-families}. Applying
@@
 \begin{array}{@{}llll}
-\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
-\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
+\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
+\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{}{}\!\!\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                      &   &             & \bigRedBracket{
@@
                                          \DConst{\Unit} \\
-                                         \DVar{n}
+                                         \DVar{\V{n}}
                                          \end{array}
@@
 \begin{array}{ll}
-\FreeIMonad{\_}{} : & \PITEL{\V{I}}{\Set}
-                      \PITEL{\V{X}}{\V{I} \To \Set} \\
-                    & \PITEL{\V{R}}{\TagIDesc{I}} \To
-                      \TagIDesc{I}
+\FreeIMonad{\_}{} : & _{\PITEL{\V{I}}{\Set}}
+                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
+                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
+                      \TagIDesc{\V{I}}
 \end{array} \\
-\FreeIMonad{(E,F)}{I}{X} \mapsto
-    \pair{\pair{\ListCons{\DVar{}}{(\fst{E})}} 
-               {\LAM{i}
-                \pair{(\DConst{(X\: i)})}
-                     {((\snd{E})\: i)}{}}{}}
-         {F}{}
+\FreeIMonad{(\V{E},\V{F})}{\V{I}}{\V{R}} \mapsto \\ \;\;
+    \pair{\pair{\ListCons{\DVar{}}{(\fst{\V{E}})}} 
+               {\LAM{\V{i}}
+                \pair{(\DConst{(\V{R}\: \V{i})})}
+                     {((\snd{\V{E}})\: \V{i})}{}}{}}
+         {\V{F}}{}
 }\]
@@
           & \PITEL{\V{\sigma}}{\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 
-                               \IMu{I}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}} \\
+                               \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}} \\
           & \PITEL{\V{i}}{\V{I}}
@@
 In Section~\ref{sec:idesc-examples}, we had the intuition that our
-datatypes $\HExprD$ and $\HExprVarD{dom}$ enjoy a monadic structure. We had
+datatypes $\HExprD$ and $\HExprVarD{\V{dom}}$ enjoy a monadic structure. We had
 identified the variable substitution operation as the \bind of a free
@@
                    \EnumT{\red{[}\Econd{}{}{}\red{]}} \red{,} \\
-                   \LAM{ty}
+                   \LAM{\V{ty}}
                    \bigRedBracket{
                    \begin{array}{l}
-                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{ty}}{\DVar{ty}}}
+                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                    \end{array}
@@
 corresponds to the syntax of closed terms, $\HExprD$. Similarly,
-$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \SUM{\Val{ty}}{\Var{dom}{ty}})}$$
-corresponds to expressions with variables, $\HExprVarD{dom}$.
+$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{\V{ty}} \SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})}$$
+corresponds to expressions with variables, $\HExprVarD{\V{dom}}$.
 
@@
 Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
-terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}\: ty$ to values in
-$\Val{ty}$. We are therefore left with implementing $\substH$. We
+terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}\: \V{ty}$ to values in
+$\Val{\V{ty}}$. We are therefore left with implementing $\substH$. We
 simply have to fill in the right arguments to $\substI$, the type
@@
 \[\stk{
-\substH\: dom\:
+\substH\: \V{dom}\:
           \gammaNat\: \gammaBool\:
-          \sigma\: 
-          ty\: 
-          term \mapsto  \\
+          \V{\sigma}\: 
+          \V{ty}\: 
+          \V{term} \mapsto  \\
 \;\;\ \begin{array}{ll}
        \substI\: & \Ty\: 
-                  (\SUM{\Val{}}{\Var{dom}{}})\: 
+                  (\SUM{\Val{}}{\Var{\V{dom}}{}})\: 
                   \Val{}\:
@@
                 &
-                  (\sigma\: dom\: \gammaNat\: \gammaBool)\:
-                  ty\:
-                  term
+                  (\V{\sigma}\: \V{dom}\: \gammaNat\: \gammaBool)\:
+                  \V{ty}\:
+                  \V{term}
       \end{array}
@@
 \begin{array}{@{}ll}
-\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DConst{},
+\IDescFreeD~\V{I} \mapsto \DSigma{}{} & (\EnumT \red{[} \DConst{},
                                               \DProd{}{},
@@
                                         \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
-                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
-                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
+                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
+                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                     \end{array}}
@@
 \IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
-\IDescD\: I \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{]}}{\Unit}{I}\: I
+\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{]}}{\Unit}{\V{I}}\: \V{I}
 }\]

### 2010-03-29 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: get some sort of introduction.
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)

@@
 \section{The Type Theory}
+\label{sec:type-theory}
 
@@
 languages. In particular, we have developed this technology in our
-on-going work on Epigram~\cite{pigs:epigram}. However, in this paper,
+on-going work on Epigram. However, in this paper,
 we strive to present our technology in the broadest setting
@@
 \Gamma \vdash \Valid                & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
-\Gamma \vdash \Bhab{t}{T}           & \mbox{term \(t\) has type \(T\) in context \(\Gamma\)} \\
-\Gamma \vdash \Bhab{s \equiv t}{T}  & \mbox{\(s\) and \(t\) are equal at type \(T\) in context \(\Gamma\)} \\
+\Gamma \vdash \Bhab{\V{t}}{\V{T}}           & \mbox{term \(\V{t}\) has type \(\V{T}\) in context \(\Gamma\)} \\
+\Gamma \vdash \Bhab{\V{s} \equiv \V{t}}{\V{T}}  & \mbox{\(\V{s}\) and \(\V{t}\) are equal at type \(\V{T}\) in context \(\Gamma\)} \\
 \end{array}
@@
 \begin{array}{l@{\;\Rightarrow\;\;}l}
-\Gamma            \vdash \Bhab{t}{T}            
-    & \Gamma \vdash \Valid\; \wedge\; \Gamma\vdash\Type{T} \\
-\Gamma            \vdash s \equiv \Bhab{t}{T}   
-    & \Gamma \vdash \Bhab{s}{T} \;\wedge\; \Gamma\vdash \Bhab{t}{T} \\
-\Gamma ; \xS ; \Delta \vdash J                      
-    & \Gamma \vdash \Bhab{s}{S} \;\Rightarrow\; 
-          \Gamma ; \Delta[s/x] \vdash J[s/x] \\
+\Gamma            \vdash \Bhab{\V{t}}{\V{T}}            
+    & \Gamma \vdash \Valid\; \wedge\; \Gamma\vdash\Type{\V{T}} \\
+\Gamma            \vdash \V{s} \equiv \Bhab{\V{t}}{\V{T}}   
+    & \Gamma \vdash \Bhab{\V{s}}{\V{T}} \;\wedge\; \Gamma\vdash \Bhab{\V{t}}{\V{T}} \\
+\Gamma ; \xS ; \Delta \vdash \V{J}                      
+    & \Gamma \vdash \Bhab{\V{s}}{\V{S}} \;\Rightarrow\; 
+          \Gamma ; \Delta[\V{s}/\V{x}] \vdash \V{J}[\V{s}/\V{x}] \\
 \end{array}
@@
 \begin{wstructure}
-<- Context validity [figure]
+<- Context validity [figure no longer]
     <- Not much to be said
@@
 
-\begin{figure}
+The rules ensuring Context Validity 
+are unsurprisingly common. They simply guarantee the validity of the
+empty context, as well as the linear extension of a context by a
+well-typed variable.
 
@@
 %% Extend context
-\Rule{\Gamma       \vdash \Type{S}}
-     {\Gamma ; \xS \vdash \Valid}\;x\not\in\Gamma
+\Rule{\Gamma       \vdash \Type{\V{S}}}
+     {\Gamma ; \xS \vdash \Valid}\;\V{x}\not\in\Gamma
 \]
-
-\caption{Context validity}
-\label{fig:context-validity}
-\end{figure}
-
-The rules ensuring Context Validity (Fig.~\ref{fig:context-validity})
-are unsurprisingly common. They simply guarantee the validity of the
-empty context, as well as the linear extension of a context by a
-well-typed variable.
 
@@
 definition, we get that a finite set is a list of tags. Finally,
-elements of a finite set $u$ belong to the corresponding $\EnumT{u}$
+elements of a finite set $\V{u}$ belong to the corresponding $\EnumT{\V{u}}$
 type. Intuitively, it corresponds to an index -- a number -- pointing
-to an element of $u$.
+to an element of $\V{u}$.
 
@@
 \spi{}{} : \PITEL{\V{e}}{\EnumU}
-           \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set \\
+           \PITEL{\V{P}}{\EnumT{\V{e}} \To \Set} \To \Set \\
 \begin{array}{@{}ll@{\:\mapsto\:\:}l}
-\spi{\NilE}{& P}        & \Unit \\
-\spi{(\ConsE{t}{e})}{& P} & \TIMES{P\: \Ze}{\spi{e}{(\LAM{x} P\: (\Su{x}))}}
+\spi{\NilE}{& \V{P}}        & \Unit \\
+\spi{(\ConsE{\V{t}}{\V{e}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{e}}{(\LAM{\V{x}} \V{P}\: (\Su{\V{x}}))}}
 \end{array}
@@
 \F{switch} : & \PITEL{\V{e}}{\EnumU}
-               \PITEL{\V{P}}{\EnumT{e} \To \Set} \\
-             & \PITEL{\V{b}}{\spi{e}{P}}
-               \PITEL{\V{x}}{\EnumT{e}} \To P~x
+               \PITEL{\V{P}}{\EnumT{\V{e}} \To \Set} \\
+             & \PITEL{\V{b}}{\spi{\V{e}}{\V{P}}}
+               \PITEL{\V{x}}{\EnumT{\V{e}}} \To P~x
 \end{array} \\
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\switch{(\ConsE{t}{e})}{P}{b}{\Ze}      & \fst{b} \\
-\switch{(\ConsE{t}{e})}{P}{b}{(\Su{x})} & \switch{e}{(\LAM{x} P
-  (\Su{x}))}{(\snd{b})}{x}
+\switch{(\ConsE{\V{t}}{\V{e}})}{\V{P}}{\V{b}}{\Ze}      & \fst{\V{b}} \\
+\switch{(\ConsE{\V{t}}{\V{e}})}{\V{P}}{\V{b}}{(\Su{\V{x}})} & \switch{\V{e}}{(\LAM{\V{x}} \V{P}
+  (\Su{\V{x}}))}{(\snd{\V{b}})}{\V{x}}
 \end{array}
@@
 application. The operational behaviour of $\F{switch}$ is
-straightforward: $x$ is peeled off as we move deeper inside the nested
-tuple $b$. When $x$ equals $\Ze$, we simply return the value we are
+straightforward: $\V{x}$ is peeled off as we move deeper inside the nested
+tuple $\V{b}$. When $\V{x}$ equals $\Ze$, we simply return the value we are
 pointing to.
@@
 \Rule{\Gamma ; \xS ; \Delta \vdash \Valid}
-     {\Gamma ; \xS ; \Delta \vdash \Bhab{x}{S}}
+     {\Gamma ; \xS ; \Delta \vdash \Bhab{\V{x}}{\V{S}}}
 \]
@@
 \[
-\Rule{\Gamma       \vdash \Type{S} \quad
-      \Gamma ; \xS \vdash \Bhab{t}{T}}
-     {\Gamma \vdash \Bhab{\PLAM{\x}{S} t}{\PIS{\xS} T}}
+\Rule{\Gamma       \vdash \Type{\V{S}} \quad
+      \Gamma ; \xS \vdash \Bhab{\V{t}}{\V{T}}}
+     {\Gamma \vdash \Bhab{\PLAM{\x}{\V{S}} \V{t}}{\PIS{\xS} \V{T}}}
 \]
@@
 sets \emph{by their name}. Indeed, such expression will be pushed into
-a $\EnumT{u}$, where $\Bhab{u}{\EnumU}$ contains all the names
+a $\EnumT{\V{u}}$, where $\Bhab{\V{u}}{\EnumU}$ contains all the names
 populating the finite set. It is therefore easy to elaborate a name
@@
 abstraction of \emph{finite} function space. This work as follow. When
-an expression $t$ is pushed into a function which domain is a finite
+an expression $\V{t}$ is pushed into a function which domain is a finite
 set, we instead elaborate it as a term in a small $\pi$-type. That is,

### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: don't present Fin as a tagged constructor, but mention the possibility.

@@
                 \begin{array}{l} 
-                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
-                                   (\NatSuc \V{n}) \mapsto [\VecCons{}{}]} \\
-                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
-                                   (\NatSuc \V{n}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
+                  \LAM{n} \caseB{n}{\NatZero \To [\VecNil] \\
+                                   (\NatSuc n) \To [\VecCons{}{}]} \\
+                  \LAM{n} \caseB{n}{\NatZero \To [\DConst{Unit}] \\
+                                   (\NatSuc n) \To [\DSigma{X}{\DVar{n}}]}
                 \end{array}}
@@
 
-Equations have simply disappeared. We should precise that forcing a
-description is not guaranteed to remove all constraints. It is subject
-to future work to see if constraints can be entirely eradicated, or
-presented more conveniently to the user. Finally, it is worth
-mentioning that the forcing optimisation is a \emph{source-to-source}
-transformation of the description.
+Equations have simply disappeared. Note that it is straightforward to
+turn this definition into a tagged form. We should precise that
+forcing a description is not guaranteed to remove all constraints. It
+is subject to future work to see if constraints can be entirely
+eradicated, or presented more conveniently to the user. Finally, it is
+worth mentioning that the forcing optimisation is a
+\emph{source-to-source} transformation of the description.
 

### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: generic induction refers to Benke et al.

@@
 it comes short of its promises in the dependently-typed realm. Indeed,
-this type is not dependent. As this operator will be the corner stone
-of any inductive definition, we are better off sharpening it before use.
+this type is not dependent, being restricted to a notion of
+\emph{iteration}. As this operator will be the corner stone of any
+inductive definition, we are better off sharpening it before use.
 
-The first step consists in turning $\V{T}$ into a predicate
-$\Bhab{\V{P}}{\Mu{\V{D}} \To \Set}$, hence introducing a dependency on the
-descriptions. Consequently, we strengthen the algebra $\descop{\V{D}}{\V{T}}
-\To \V{T}$ by a similar process. We start with turning the arrow into a
-dependent arrow, hence offering $\Bhab{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}$ to
+The first step consists in turning $T$ into a predicate
+$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
+descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
+\To T$ by a similar process. We start with turning the arrow into a
+dependent arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
 further analysis.

### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: precise that Desc is not here to rivalize with Generic Haskell or related. It is just pedagogical.
Desc@ICFP: use 'pattern functor' instead of 'signature functor', which seems to be more used in the Generic prog. literature.

@@
 layman data-types in our setting, we build on our intuition of
-data-types we are familiar with. However, our goal is and remains to
-discuss data-types as we find them in dependently-typed languages.
+data-types we are familiar with. In particular, we do not aim at
+describing a model for generics in simply-typed functional
+language. Our goal is and remains to discuss data-types as we find
+them in dependently-typed languages.
 
@@
 
-Our first example is the natural numbers, or rather its signature
+Our first example is the natural numbers, or rather its pattern
 functor. Our code is presented in the high-level expression language
@@
 With a small change to the definition of $\NatD$, we obtain the
-signature of lists:
+pattern functor for lists:
 
@@
 So far, we have used our universe of descriptions to build
-functors. To illustrate its usage, we have implemented the signature
+functors. To illustrate its usage, we have implemented the pattern
 functors of natural numbers, lists, and binary trees. However, the
@@
 identify $\Mu{}$ of a description as indeed the fix-point of the
-functor signature of natural numbers. Finally, she would have to write
+pattern functor of natural numbers. Finally, she would have to write
 codes instead of the natural data-type constructor.
@@
 the second issue. Namely, our user would rather be presented a name
-rather than a fix-point of a signature functor. Surprisingly, it is
+rather than a fix-point of a pattern functor. Surprisingly, it is
 sufficient to \emph{label} the $\Mu{}$ constructor with the

### 2010-03-29 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: use 'pattern functor' instead of 'signature functor', which seems to be more used in the Generic prog. literature.

@@
 type former $\EnumU$, the constructors $\NilE$ and $\ConsE$. We have
-just defined $\EnumU$ above, by fix-point over the signature functor denoted
+just defined $\EnumU$ above, by fix-point over the pattern functor denoted
 by $\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and $\ListCons$
@@
 Being an inductive type, this $\FMFreeMonad$ data-type is itself
-defined by a functor signature. It is given by:
+defined by a pattern functor. It is given by:
 

### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: use 'pattern functor' instead of 'signature functor', which seems to be more used in the Generic prog. literature.

@@
 In order to gain some intuition with $\IDesc$, let us re-implement the
-signature functor of natural numbers. We refer the reader to
+pattern functor of natural numbers. We refer the reader to
 Section~\ref{sec:desc-examples} for its incarnation in the universe of
@@
 nothing but the \return. The remaining constructors are simply the
-signature functor, trivially indexed by $\Unit$. Hence, we describe
-this signature functor by the following code:
+pattern functor, trivially indexed by $\Unit$. Hence, we describe
+this functor by the following code:
 

### 2010-03-29 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: we use Lisp notation for EnumU.
Desc@ICFP: cite other users of bidirectional type-checking, oriented boilerplate scrapping.

@@
 is the Church encoding of natural numbers, which is isomorphic to
-finite sets.
-
-% Another example is Martin-Löf's W-types~\cite{}.
-
-\note{Shall we talk about W-types encoding?}
+finite sets. \note{Shall we talk about W-types encoding?}
 
@@
 \note{Did I got the coding presentation vs. coding representation
-  story right?}
+  story right? No.}
 
@@
 type. Intuitively, it corresponds to an index -- a number -- pointing
-to an element of $\V{u}$.
+to an element of $u$.
 
@@
 instead. Hence, we use the syntax $[t_1 \ldots t_n]$ to denote the
-finite sets composed of the named elements $t_1$, \ldots, $t_n$. The
-translation to elements of $\EnumU$ is straightforward.
+finite sets composed of the named elements $t_1$, \ldots, $t_n$. This
+syntax is reminiscent to Lisp notation for lists. The translation to
+elements of $\EnumU$ is straightforward.
 
@@
 bidirectional type-checking. Using bidirectional type-checking as a
-boilerplate scrapper is a well-known technique~\cite{who?} \note{I
-  would need some citation of people using bidirectional presentation
-  of their type theory}. In our case, we have shown how to instrument
-bidirectionality to rationalise the expressivity of dependent
-types. We have illustrated our approach with finite sets. We have
-abstracted away the low-level presentation of finite sets, offering a
-convenient syntax instead.
+boilerplate scrapper is a well-known
+technique~\cite{pierce:bidirectional-tc,
+  xi:bidirectional-tc-bound-array, chlipala:strict-bidirectional-tc}
+\note{Everybody ok with the citations?}. In our case, we have shown
+how to instrument bidirectionality to rationalise the expressivity of
+dependent types. We have illustrated our approach with finite sets. We
+have abstracted away the low-level presentation of finite sets,
+offering a convenient syntax instead.
 

### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: cite other users of universes.

@@
 fruitful as a programming 
-technique~\cite{benke:universe-generic-prog,whoelse?}
-\note{I would need some other example of universe construction for
-  practical purposes}. 
+technique~\cite{benke:universe-generic-prog, oury:power-of-pi}
+\note{Any idea of other papers?}. 
 We refer the reader to the Agda tutorial~\cite{norell:agda-tutorial} for a

### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: 'user' evokes an interactive setting. Go back to the less frightening 'developer' Joe.

@@
 defining data-types and manipulating them, our proposal could be seen
-as a step backward. In our setting, the user would have to write a
-description instead of a sugared data-type definition. Then, to
+as a step backward. In our setting, the developer would have to write
+a description instead of a sugared data-type definition. Then, to
 recognise an element of $\Nat$ for instance, she would have to
@@
 abstraction. We take advantage of this abstraction barrier to tamper
-the second issue. Namely, our user would rather be presented a name
-rather than a fix-point of a pattern functor. Surprisingly, it is
+the second issue. Namely, the developer would rather be presented a
+name rather than a fix-point of a pattern functor. Surprisingly, it is
 sufficient to \emph{label} the $\Mu{}$ constructor with the
-user-provided name. When pretty-printing this object, we just expose
+user-defined name. When pretty-printing this object, we just expose
 the label, instead of the definition in full.
@@
 
-Hence, in the high-level expression language, the user types a
+Hence, in the high-level expression language, the developer writes a
 familiar type constructors, applied to a telescope of arguments. In

### 2010-03-29 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: 'user' evokes an interactive setting. Go back to the less frightening 'developer' Joe.

@@
 could be worried by the availability of such operations in our type
-theory. Indeed, it would be inconvenient for the user to have to
+theory. Indeed, it would be inconvenient for the developer to have to
 decide between two variants. The solution to this problem is, once
@@
 completes the implementation. To sum up, we have implemented the free
-monad construction for an arbitrary tagged description. This gives our
-user the ability, for any data-type, to extend it with a notion of
-variable. Then, we have equipped this structure with the corresponding
-monadic operation, \bind and \return. Whereas the \return is a trivial
-variable former, the \bind operation corresponds to the substitution
-of variables by terms. We have shown how to implement it, for any free
-monad.
+monad construction for an arbitrary tagged description. This gives the
+developer the ability, for any data-type, to extend it with a notion
+of variable. Then, we have equipped this structure with the
+corresponding monadic operation, \bind and \return. Whereas the
+\return is a trivial variable former, the \bind operation corresponds
+to the substitution of variables by terms. We have shown how to
+implement it, for any free monad.
 

### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: 'user' evokes an interactive setting. Go back to the less frightening 'developer' Joe.

@@
 of equalities in the picture. In a programming context, equalities
-haunt us, begging for proofs. Hence, whereas the user typed a
+haunt us, begging for proofs. Hence, whereas the developer typed a
 constructor-constrained data-type, the elimination form will involve
@@
 is subject to future work to see if constraints can be entirely
-eradicated, or presented more conveniently to the user. Finally, it is
-worth mentioning that the forcing optimisation is a
+eradicated, or presented more conveniently to the developer. Finally,
+it is worth mentioning that the forcing optimisation is a
 \emph{source-to-source} transformation of the description.

### 2010-03-29 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: 'user' evokes an interactive setting. Go back to the less frightening 'developer' Joe.

@@
 focus. This is a general issue with the usage of codes, as they convey
-much more information than what the user is willing to see.
+much more information than what the developer is willing to see.
 
@@
 
-Finally, and more importantly, we are able to present our user an
+Finally, and more importantly, we are able to present the developer an
 abstraction of \emph{finite} function space. This work as follow. When
@@
 transparent to the user. In the case of finite function space, the
-user has never to explicitly \emph{call} the $\F{switch}$
+developer has never to explicitly \emph{call} the $\F{switch}$
 eliminator. For diagnosis purposes, a similar, reverse mechanism turns

### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: fix spacing of operators used in text.

@@
 universe mimics the standard grammar of data-types definitions in
-simply-typed languages. Hence, we have a $\DSigma{}{}$ code,
+simply-typed languages. Hence, we have a $\DSigma{\!}{\!}$ code,
 interpreted as a $\Sigma$-type, to build the
@@
 argument. We notice that this functoriality appears in the type of
-$\descop{\V{D}}{}$ itself, for a given data-type definition $\V{D}$. This
-corresponds to the object part of the functor described by $\V{D}$.
+$\descop{D}{}$ itself, for a given data-type definition $D$. This
+corresponds to the object part of the functor described by $D$.
 
@@
     \DUnit          & \Desc \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
     \DIndx          & \Desc \To \Desc
@@
 \NatD : \Desc \\
-\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
+\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
                      {[ \DUnit \quad (\DIndx{\DUnit}) ]}
@@
 
-Let us explain its construction. First, we use $\DSigma{}{}$ to give a
-choice between the $\NatZero$ and $\NatSuc{}$ constructors. In the
-$\NatZero$ case, we reach the end of the description: there is no
-useful payload in that case. In the $\NatSuc{}$ case, we left one hole
+Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
+give a choice between the $\NatZero$ and $\NatSuc{\!}$ constructors. In
+the $\NatZero$ case, we reach the end of the description: there is no
+useful payload in that case. In the $\NatSuc{\!}$ case, we left one hole
 open for the recursive argument, and we close the description.
@@
 \ListD : \Set \To \Desc \\
-\ListD \: \V{X} \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
-                           {[ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) ]}
+\ListD \: X \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
+                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{\DUnit}}) ]}
 }\]
@@
 The modification consists in turning the $\NatSuc{}$ constructor into
-a proper $\ListCons$ taking an argument in $\V{X}$ followed by a recursive
+a proper $\ListCons$ taking an argument in X followed by a recursive
 argument. In this case, we use $\DSigma{}{}$ in its product
@@
 From the examples above, we observe that data-types are defined by a
-$\DSigma$ which first argument is a finite set of constructors. The
-descriptions fitting into this pattern are called \emph{tagged}
-description. Formally, we have:
+$\DSigma{\!}{\!}$ which first argument is a finite set of
+constructors. The descriptions fitting into this pattern are called
+\emph{tagged} description. Formally, we have:
 
@@
 $\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
-particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
-and $\DUnit$ as type formers, extending the basic type theory. Hence,
-in an implementation of this type theory, we need to extended it with
-these constructors and their typing rules. Although it makes no
-conceptual difference, the code of description should rather be read
-as a specification. We promise the existence of such objects,
-satisfying the typing rules. It will be the subject of
+particular, we have considered its code, $\DSigma{\!}{\!}$,
+$\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
+theory. Hence, in an implementation of this type theory, we need to
+extended it with these constructors and their typing rules. Although
+it makes no conceptual difference, the code of description should
+rather be read as a specification. We promise the existence of such
+objects, satisfying the typing rules. It will be the subject of
 Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
@@
                  \PITEL{\V{P}}{\V{X} \To \Set} \\
-               & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
+               & \PI{\V{xs}}{\descop{D}{\V{X}}} 
                  \Set 
@@
                    \PITEL{\V{P}}{\V{X} \To \Set} \\
-                 & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
-                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
-                   \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
+                 & \PITEL{\V{R}}{\PI{\V{x}}{X} P\: x}
+                   \PI{\V{xs}}{\descop{D}{\V{X}}} 
+                   \All{D}{X}{P}{xs} 
 \end{array} \\

### 2010-03-29 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix spacing of operators used in text.

@@
 
-The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
+The names $\NilE$ and $\ConsE{\!}{\!}$ are absolutely not a
 coincidence. $\EnumU$ is indeed isomorphic to a list of $\UId$s. In
@@
 
-Another interesting consequence concerns the $\spi{}{}$ operator: it
-can actually be implemented with the generic $\F{induction}$
+Another interesting consequence concerns the $\spi{\!}{\!}$ operator:
+it can actually be implemented with the generic $\F{induction}$
 principle. From an implementation perspective, this means that we
@@
 \begin{array}{@{}ll}
-\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{} ]})  \\
+\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
                                & \left[\begin{array}{l}
@@
 constructor. Constructors correspond to the code's constructors:
-$\DUnit$, $\DSigma$, and $\DIndx$. The reader will notice that the
+$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice that the
 tagged notation we have used so far for $\Desc$ constructors now fully
@@
 The case of $\DUnit$ is trivial: we immediately reach the end of the
-data-type. $\DUnit$ has no payload. The case of $\DIndx{}$ comes
+data-type. $\DUnit$ has no payload. The case of $\DIndx{\!}$ comes
 naturally: it only takes a recursive argument -- a description. The
-case of $\DSigma$ is problematic. Recall the specification of
-$\DSigma$:
+case of $\DSigma{\!}{\!}$ is problematic. Recall the specification of
+$\DSigma{\!}{\!}$:
 
-\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc      \]
+\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc      \]
 
@@
 extend our universe of descriptions. This consists in adding a code
-$\DHindx$ that takes an indexing set $\V{H}$. The code and its
+$\DHindx$ that takes an indexing set $H$. The code and its
 interpretation are presented in
-Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx$ use the elements
-of $H$ to index as many recursive arguments.
+Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ use the
+elements of $H$ to index as many recursive arguments.
 
@@
     \ldots          & \:\:\ldots \\
-    \DHindx         & \PI{\V{H}}{\Set} \Desc \To \Desc
+    \DHindx         & \PI{H}{\Set} \Desc \To \Desc
 \end{array}
@@
 
-Note that $\DIndx$ is isomorphic to $\DHindx{\Unit}$. However, we
-tolerate this duplication and do not replace $\DIndx$ by the more
-general $\DHindx$. Indeed, unlike its counterpart, $\DIndx$ is
-first-order. Therefore, while both codes are \emph{extensionally} the
-same, in most practical implementation they would be dealt with rather
-differently. While definitional equality can cope with first order
-objects, the functional presentation introduced by $\DHindx$ is
-unlikely to be amenable to a purely definitional treatment.
+Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
+we tolerate this duplication and do not replace $\DIndx{\!}$ by the
+more general $\DHindx{\!}{\!}$. Indeed, unlike its counterpart,
+$\DIndx{\!}$ is first-order. Therefore, while both codes are
+\emph{extensionally} the same, in most practical implementation they
+would be dealt with rather differently. While definitional equality
+can cope with first order objects, the functional presentation
+introduced by $\DHindx{\!}{\!}$ is unlikely to be amenable to a purely
+definitional treatment.
 
@@
 successfully describe the $\DSigma$ case, by a simple appeal to the
-higher-order induction on $\V{S}$. The $\DHindx$ case consists in packing
-the $\V{H}$ in $\Set$ with a recursive argument.
+higher-order induction on $S$. The $\DHindx$ case consists in packing
+the $H$ in $\Set$ with a recursive argument.
 
@@
 \begin{array}{@{}ll}
-\DescD \mapsto \DSigma{}{} & (\EnumT{[ \DUnit, \DSigma{}{}, \DIndx{}, \DHindx{}{} ]}) \\
+\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!}, \DHindx{\!}{\!} ]}) \\
                            & \left[\begin{array}{l}
@@
 \FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
-\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
-    \pair{\ListCons{\DVar{}}{\V{E}}}
-         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
+\FreeMonad{\pair{E}{D}{}}\:X \mapsto
+    \pair{\ListCons{\DVar{}}{E}}
+         {\pair{\DSigma{X}{\DUnit}}{D}{}}{}
 }\]
@@
 We simply add a constructor, $\DVar{}$, and define its argument to be
-a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep $\V{E}$ and $\V{D}$
+a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
 as they were, hence leaving the functor unchanged. Unfolding the
@@
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
-\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{ys}}{} & \Con{\pair{\etag{t}}{\V{ys}}{}}
+\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}   & \sigma\: x                   \\
+\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
 \end{array}

### 2010-03-29 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)
Desc@ICFP: fix spacing of operators used in text.

@@
 universes of description rather naturally leads itself to
-stratification. Unsurprisingly, $\IDesc{}$ at level $l$ is of type
-$\Set^{\blue{l+1}}$. Similarly, the interpretation of $\IDesc{}$ at
+stratification. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of type
+$\Set^{\blue{l+1}}$. Similarly, the interpretation of $\IDesc{\!}$ at
 level $l$ is an object of type $\Set^{\blue{l}}$:
@@
 \\
-\idescop{\_\:}{}{}^{\blue{l}} : \PI{\V{I}}{\Set^{\blue{l+1}}} \IDesc{{\!}^{\blue{l}}I} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
+\idescop{\_\:}{}{}^{\blue{l}} : \PI{\V{I}}{\Set^{\blue{l+1}}} \IDesc{{\!}^{\blue{l}}\V{I}} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
 \ldots
@@
 advantage of set polymorphism to implement the universe of indexed
-descriptions at any level. Further, we have coded $\IDesc{}$ in itself
-and have proved the isomorphism between the host and the embedded
-universes.
+descriptions at any level. Further, we have coded $\IDesc{\!}$ in
+itself and have proved the isomorphism between the host and the
+embedded universes.
 
@@
 formers and elimination rules. Thanks to the levitation, we only rely
-on a generic induction and a specialised $\switchD{}{}{}$. Second, the
-authors develop several universes, each with its own strengths and
-weaknesses, but lack a formal, unifying universe. With indexed
-descriptions, we subsume simple descriptions, hence avoiding the move
-to and from several codings. Finally, the authors often resort to an
-extensional notion of equality, while we have given an
-equality-agnostic presentation. Moreover, our presentation is arranged
-so as to use definitional equality as much as possible. Hence, in
-practice, the developer is relieved from many trivial proof
-obligations.
+on a generic induction and a specialised
+$\switchD{\!}{\!}{\!}$. Second, the authors develop several universes,
+each with its own strengths and weaknesses, but lack a formal,
+unifying universe. With indexed descriptions, we subsume simple
+descriptions, hence avoiding the move to and from several
+codings. Finally, the authors often resort to an extensional notion of
+equality, while we have given an equality-agnostic
+presentation. Moreover, our presentation is arranged so as to use
+definitional equality as much as possible. Hence, in practice, the
+developer is relieved from many trivial proof obligations.

### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix spacing of operators used in text.

@@
 rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
-$\IDesc$, the universe of indexed descriptions. We express inductive families
+$\IDesc{\!}$, the universe of indexed descriptions. We express inductive families
 by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
@@
 of inductive types. Hence, the codes closely follow this grammar. In
-the following, we adopt an alternative presentation. With $\DSigma$,
-we are actually \emph{quoting} a standard type-former, namely
+the following, we adopt an alternative presentation. With
+$\DSigma{\!}{\!}$, we are actually \emph{quoting} a standard
+type-former, namely
 
@@
 shown in Figure~\ref{fig:type-former-desc}. The reader will notice
-that we replace $\DUnit$ by a more general $\DConst$ code. Whereas
-$\DUnit$ was interpreted as the unit set, $\DConst{\V{X}}$ is interpreted
-as $\V{X}$, for any $\Bhab{\V{X}}{\Set}$. Extensionally, $\DConst{\V{X}}$ and
-$\DSigma{\V{X}}{\DUnit}$ are equivalent. However, $\DConst$ is more
+that we replace $\DUnit$ by a more general $\DConst{\!}$ code. Whereas
+$\DUnit$ was interpreted as the unit set, $\DConst{X}$ is interpreted
+as $X$, for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
+$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
 succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
@@
 burden on the programmer. For the same reason, we introduce
-$\DProd{}{}$ that overlaps with $\DSigma{}{}$.
+$\DProd{\!}{\!}$ that overlaps with $\DSigma{\!}{\!}$.
 
@@
 report for details. In this finer-grained presentation, we can define
-$\DIndx$ and $\DHindx$ as follow:
+$\DIndx{\!}$ and $\DHindx{\!}{\!}$ as follow:
 
@@
 \NatD : \Desc \\
-\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
+\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
                      {[ \DUnit \quad \DId ]}
@@
     \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
-    \DPi            & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc 
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
+    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
 \end{array}
@@
   transformer anyone?} In this setting, $\V{I} \To \IDesc{\V{I}}$ is our
-grammar for describing these functors. Hence, $\IDesc{}$ and its
+grammar for describing these functors. Hence, $\IDesc{\!}$ and its
 interpretation have the following types:
@@
 \[\stk{
-\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
+\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
 \;\;\ldots \\
@@
 $\Desc$ is entirely captured by the $\DId$ code. We get an indexed
-functor by turning $\DId$ into an index-dependent $\DVar$ code. The
-resulting code of $\IDesc$ is presented in Figure~\ref{fig:idesc}.
+functor by turning $\DId$ into an index-dependent $\DVar{\!}$
+code. The resulting code of $\IDesc{\!}$ is presented in
+Figure~\ref{fig:idesc}.
 
@@
 \stk{
-\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
+\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \DVar{}         & \V{I} \To \IDesc{\V{I}}                                   \\
-    \DConst{}       & \Set \To \IDesc{\V{I}}                                \\
-    \DProd{}{}      & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}           \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
-    \DPi            & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
+    \DVar{}         & I \To \IDesc{I}                                   \\
+    \DConst{}       & \Set \To \IDesc{I}                                \\
+    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
+    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
+    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
 \end{array}
@@
 would only propose the constructor $\ListNil$. Similarly, for
-$\NatSuc{n}$, we only propose the constructor $\ListCons{}{}$. Hence,
+$\NatSuc{n}$, we only propose the constructor $\ListCons{\!}{\!}$. Hence,
 we obtain the following definition:
@@
 
-In order to gain some intuition with $\IDesc$, let us re-implement the
-pattern functor of natural numbers. We refer the reader to
+In order to gain some intuition with $\IDesc{\!}$, let us re-implement
+the pattern functor of natural numbers. We refer the reader to
 Section~\ref{sec:desc-examples} for its incarnation in the universe of
@@
 \NatD : \IDesc{\Unit} \\
-\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
+\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
                      {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
@@
 
-A mandatory exercise consists in describing $\IDesc$ in itself. With
+A mandatory exercise consists in describing $\IDesc{\!}$ in itself. With
 the experience gained in Section~\ref{sec:desc-levitate}, this
@@
 \begin{array}{@{}ll}
-\IDescD~\V{I} \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
+\IDescD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
                                           \DConst{},
@@
 Note that this description also lives in $\IDesc{\Unit}$. Indeed,
-for any $\V{I}$, $\IDesc~\V{I}$ is merely an inductive type. As for the
+for any $\V{I}$, $\IDesc{\V{I}}$ is merely an inductive type. As for the
 $\Desc$ levitation, the finite function space $[ \ldots ]$ elaborates into a
@@
 pattern-matching. Formally, this corresponds to a call to the
-$\switch{}{}{}{}$ eliminator.
+$\switch{\!}{\!}{\!}{\!}$ eliminator.
 
@@
 
-Another common dependent data-type is $\Fin$, the finite sets of
-numbers. $\Fin$ is a typical example of an inductive family using
+Another common dependent data-type is $\Fin{\!}$, the finite sets of
+numbers. $\Fin{\!}$ is a typical example of an inductive family using
 \emph{constrained constructors}. The constructors $\FinZero$ and
-$\FinSuc$ are only defined for an index distinct from $\NatZero$:
+$\FinSuc{\!}$ are only defined for an index distinct from $\NatZero$:
 
@@
 \stk{
-\data \Fin{} : \PI{\V{n}}{\Nat} \Set \where \\
+\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
-    \FinSuc{}_{\V{n}}     & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
+    \FinZero_n      & \Fin{(\NatSuc{n})}   \\
+    \FinSuc{}_n     & \Fin{n} \To \Fin{(\NatSuc{n})}
 \end{array}
@@
 \begin{array}{@{}ll}
-\FinD\: \V{n} = \DSigma{}{} \!\!\!\!\!\!& (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
+\FinD\: n = \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                         & \bigRedBracket{
                           \begin{array}{l}
-                            \DSigma{\Nat}{\LAM{\V{m}} \DConst{(\V{n} \PropEq \NatSuc{\V{m}})}} \\
-                            \DSigma{\Nat}{\LAM{\V{m}} \DProd{\DVar{\V{m}}}{\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}
+                            \DSigma{\Nat}{\LAM{m} \DConst{(n \PropEq \NatSuc{m})}} \\
+                            \DSigma{\Nat}{\LAM{m} \DProd{\DVar{m}}{\DConst{(n \PropEq \NatSuc{m})}}}
                           \end{array}
@@
 Intuitively, the constraint is captured as follows. First, we store an
-element $\V{m}$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
-stipulates that $\V{m}$ cannot be \emph{any} natural numbers: it must be
-``the index minus one''. This translates into the constraint $\V{n}
-\PropEq \NatSuc{\V{m}}$.
+element $m$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
+stipulates that $m$ cannot be \emph{any} natural numbers: it must be
+``the index minus one''. This translates into the constraint $n
+\PropEq \NatSuc{m}$.
 
@@
 
-The $\Fin$ example is also a well-known example of Generalised
+The $\Fin{\!}$ example is also a well-known example of Generalised
 Abstract Data-Type (GADT) in Haskell. It comes as no surprise that our
 universe of indexed description supports this restricted class of
-inductive families. As we have seen with $\Fin$, constrained
+inductive families. As we have seen with $\Fin{\!}$, constrained
 constructors are straightforwardly translated into equations.
@@
 Further, we can actually remove these equations altogether. Let us
-look back at $\Fin$. We note that the equations are introduced because
+look back at $\Fin{\!}$. We note that the equations are introduced because
 we are \emph{storing} the index of the inductive family, through
@@
   store their indices}~\cite{brady:index-inductive-families}. Applying
-the \emph{forcing} optimisation to our initial definition of $\Fin$
+the \emph{forcing} optimisation to our initial definition of $\Fin{\!}$
 gives the following, equivalent definition:
@@
 \begin{array}{@{}llll}
-\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
-\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{}{}\!\!\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
+\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
+\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                      &   &             & \bigRedBracket{
@@
                                          \DConst{\Unit} \\
-                                         \DVar{\V{n}}
+                                         \DVar{n}
                                          \end{array}
@@
 \end{array} \\
-\FreeIMonad{(\V{E},\V{F})}{\V{I}}{\V{R}} \mapsto \\ \;\;
-    \pair{\pair{\ListCons{\DVar{}}{(\fst{\V{E}})}} 
-               {\LAM{\V{i}}
-                \pair{(\DConst{(\V{R}\: \V{i})})}
-                     {((\snd{\V{E}})\: \V{i})}{}}{}}
-         {\V{F}}{}
+\FreeIMonad{(E,F)}{I}{X} \mapsto
+    \pair{\pair{\ListCons{\DVar{}}{(\fst{E})}} 
+               {\LAM{i}
+                \pair{(\DConst{(X\: i)})}
+                     {((\snd{E})\: i)}{}}{}}
+         {F}{}
 }\]
@@
 
-Another instance of free monad is $\IDesc$ itself. Indeed, $\DVar$ is
-nothing but the \return. The remaining constructors are simply the
-pattern functor, trivially indexed by $\Unit$. Hence, we describe
-this functor by the following code:
+Another instance of free monad is $\IDesc{\!}$ itself. Indeed,
+$\DVar{\!}$ is nothing but the \return. The remaining constructors are
+simply the pattern functor, trivially indexed by $\Unit$. Hence, we
+describe this functor by the following code:
 
@@
 \begin{array}{@{}ll}
-\IDescFreeD~\V{I} \mapsto \DSigma{}{} & (\EnumT \red{[} \DConst{},
+\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DConst{},
                                               \DProd{}{},

### 2010-03-29 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: fix spacing of operators used in text.

@@
 The type of finite sets comes with further machinery. Indeed, to
-manipulate the $\EnumU$ and $\EnumT{}$ objects, we need to equip our
+manipulate the $\EnumU$ and $\EnumT{\!}$ objects, we need to equip our
 type theory with an appropriate \emph{elimination
@@
 implement them in our Agda model. Hence, we first define the
-$\spi{}{}$ operator:
+$\spi{\!}{\!}$ operator:
 
@@
 %% spi
-\spi{}{} : \PITEL{\V{e}}{\EnumU}
+\spi{\!}{\!} : \PITEL{\V{e}}{\EnumU}
            \PITEL{\V{P}}{\EnumT{\V{e}} \To \Set} \To \Set \\
@@
 A closer look to its type should reveal a certain familiarity with its
-equivalent for sets, $\Pi$. Indeed, $\spi{}{}$ defines a \emph{finite
-  function space}, mapping elements of finite sets to objects in
-$\Set$. Intuitively, this operator build a right-nested tuple,
-effectively mapping an object of $P\:i$ for each $i$ the finite
+equivalent for sets, $\Pi$. Indeed, $\spi{\!}{\!}$ defines a
+\emph{finite function space}, mapping elements of finite sets to
+objects in $\Set$. Intuitively, this operator build a right-nested
+tuple, effectively mapping an object of $P\:i$ for each $i$ the finite
 domain.
 
-Having understood the meaning of $\spi{}{}$, the definition of the
+Having understood the meaning of $\spi{\!}{\!}$, the definition of the
 $\F{switch}$ eliminator is not surprising:
@@
 As we have seen, finite sets are not just a way of making
-data. Through the $\spi{}{}$ and $\F{switch}$ operators, they also
+data. Through the $\spi{\!}{\!}$ and $\F{switch}$ operators, they also
 come with a notion of finite function space. However, we had to

### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: clean-up some notes.
Desc@ICFP: merge Colors and Spaces.

@@
 contains no useful payload. The functoriality of the data-types is
-introduced by $\DIndx$. When we tie the knot with a fix-point, the
-hole left open by $\DIndx$ will be turned into a standard recursive
-argument. We notice that this functoriality appears in the type of
-$\descop{D}{}$ itself, for a given data-type definition $D$. This
-corresponds to the object part of the functor described by $D$.
+introduced by $\DIndx{\!}$. When we tie the knot with a fix-point, the
+hole left open by $\DIndx{\!}$ will be turned into a standard
+recursive argument. We notice that this functoriality appears in the
+type of $\descop{\V{D}}{\!}$ itself, for a given data-type definition
+$\V{D}$. This corresponds to the object part of the functor described
+by $\V{D}$.
 
@@
     \DUnit          & \Desc \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
-    \DIndx          & \Desc \To \Desc
+    \DSigma{\!}{\!} & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
+    \DIndx{\!}      & \Desc \To \Desc
 \end{array}
@@
 \ListD : \Set \To \Desc \\
-\ListD \: X \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
-                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{\DUnit}}) ]}
+\ListD \: \V{X} \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons{\!}{\!} ]})}
+                           {[ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) ]}
 }\]
 
-The modification consists in turning the $\NatSuc{}$ constructor into
-a proper $\ListCons$ taking an argument in X followed by a recursive
-argument. In this case, we use $\DSigma{}{}$ in its product
-interpretation: we pack an element of $\V{X}$ together with the recursive
-argument. Easily, one sees that this code actually describes the
-following functor:
+The modification consists in turning the $\NatSuc{\!}$ constructor
+into a proper $\ListCons{\!}{\!}$ taking an argument in $\V{X}$
+followed by a recursive argument. In this case, we use
+$\DSigma{\!}{\!}$ in its product interpretation: we pack an element of
+$\V{X}$ together with the recursive argument. Easily, one sees that
+this code actually describes the following functor:
 
@@
 
-\note{ Strictly positive types anyone? }
-
 So far, we have used our universe of descriptions to build
-functors. To illustrate its usage, we have implemented the pattern
-functors of natural numbers, lists, and binary trees. However, the
-class of functors expressible with descriptions enjoy another
-property: they all admit a fix-point. Indeed, descriptions are
-interpreted into \emph{polynomial functors}~\cite{who?}\note{who is
-  the relevant source for polynomial functors?}, as they are only
-composed of $\Sigma$-types, the unit type, and variables.
+functors. \note{ Strictly positive types anyone? } To illustrate its
+usage, we have implemented the pattern functors of natural numbers,
+lists, and binary trees. However, the class of functors expressible
+with descriptions enjoy another property: they all admit a
+fix-point. Indeed, descriptions are interpreted into \emph{polynomial
+  functors}~\cite{who?}\note{who is the relevant source for polynomial
+  functors?}, as they are only composed of $\Sigma$-types, the unit
+type, and variables.
 
@@
 
-The first step consists in turning $T$ into a predicate
-$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
-descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
-\To T$ by a similar process. We start with turning the arrow into a
-dependent arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
-further analysis.
+A full account of such generic induction schema can be found in Benke
+et al.~\cite{benke:universe-generic-prog,
+  dybjer:ir-initial-algebra}. In the following, we convey the
+intuition behind this schema by a step-by-step construction of the
+dependent eliminator from the catamorphism. The first step consists in
+turning $\V{T}$ into a predicate $\Bhab{\V{P}}{\Mu{\V{D}} \To \Set}$,
+hence introducing a dependency on the descriptions. Consequently, we
+strengthen the algebra $\descop{\V{D}}{\V{T}} \To \V{T}$ by a similar
+process. We start with turning the arrow into a dependent arrow, hence
+offering $\Bhab{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}$ to further
+analysis.
 
@@
 \begin{array}{@{}ll}
-\All{}{}{}{} : & \PITEL{\V{D}}{\Desc}
+\All{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                          \PITEL{\V{X}}{\Set}
                          \PITEL{\V{P}}{\V{X} \To \Set} \\
-               & \PI{\V{xs}}{\descop{D}{\V{X}}} 
+                       & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                          \Set 
@@
 \All{\DId}{\V{X}}{\V{P}}{\V{x}} \mapsto \V{P}\: \V{x} \\
-\All{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{z}} \mapsto \Unit \\
-\All{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\pair{\V{d}}{\V{d'}}{}} \mapsto \TIMES{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}{\All{\V{D'}}{\V{X}}{\V{P}}{\V{d'}}} \\
-\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} \mapsto \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
-\All{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{f}} \mapsto \PI{\V{\V{s}}}{\V{S}} \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{(\V{f}\: \V{s})}
+\All{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{z}} \mapsto 
+    \Unit \\
+\All{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\pair{\V{d}}{\V{d'}}{}} \mapsto 
+    \TIMES{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}{\All{\V{D'}}{\V{X}}{\V{P}}{\V{d'}}} \\
+\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} \mapsto 
+    \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
+\All{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{f}} \mapsto 
+    \PI{\V{\V{s}}}{\V{S}} \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{(\V{f}\: \V{s})}
 }
@@
 \begin{array}{@{}ll}
-\all{}{}{}{}{} : & \PITEL{\V{D}}{\Desc}
+\all{\!}{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                              \PITEL{\V{X}}{\Set}
                              \PITEL{\V{P}}{\V{X} \To \Set} \\
-                 & \PITEL{\V{R}}{\PI{\V{x}}{X} P\: x}
-                   \PI{\V{xs}}{\descop{D}{\V{X}}} 
-                   \All{D}{X}{P}{xs} 
+                           & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
+                             \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
+                             \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
 \end{array} \\
-\all{\DId}{\V{X}}{\V{P}}{\V{R}}{\V{x}} = \V{R}\: \V{x} \\
-\all{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{R}}{\V{z}} = \Void \\
-\all{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{d}}{\V{d'}}{}} = \pair{\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{d}}}{\all{\V{D'}}{\V{X}}{\V{P}}{\V{R}}{\V{d'}}}{} \\
-\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} = \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
-\all{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\V{f}} = \LAM{\V{s}} \all{(\V{T}\: \V{s})}{\V{X}}{\V{P}}{\V{R}}{(\V{f}\: \V{s})}
+\all{\DId}{\V{X}}{\V{P}}{\V{R}}{\V{x}} = 
+    \V{R}\: \V{x} \\
+\all{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{R}}{\V{z}} = 
+    \Void \\
+\all{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{d}}{\V{d'}}{}} =
+    \pair{\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{d}}}{\all{\V{D'}}{\V{X}}{\V{P}}{\V{R}}{\V{d'}}}{} \\
+\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} = 
+    \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
+\all{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\V{f}} = 
+    \LAM{\V{s}} \all{(\V{T}\: \V{s})}{\V{X}}{\V{P}}{\V{R}}{(\V{f}\: \V{s})}
 \end{array}
@@
 inference rule:
-
 \note{ This is rather drafty. I should explain what $\vec{e}$ and $\widetilde{e'}$ are. }

### 2010-03-29 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: merge Colors and Spaces.

@@
 
-\[    \DSigma{}{} : \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc      \]
+\[    \DSigma{\!}{\!} : \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc      \]
 
@@
 extend our universe of descriptions. This consists in adding a code
-$\DHindx$ that takes an indexing set $H$. The code and its
+$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. The code and its
 interpretation are presented in
@@
     \ldots          & \:\:\ldots \\
-    \DHindx         & \PI{H}{\Set} \Desc \To \Desc
+    \DHindx{\!}{\!} & \PI{\V{H}}{\Set} \Desc \To \Desc
 \end{array}
@@
 
-Equipped with $\DHindx$, we can describe the extended code of our
-universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
-The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
-successfully describe the $\DSigma$ case, by a simple appeal to the
-higher-order induction on $S$. The $\DHindx$ case consists in packing
-the $H$ in $\Set$ with a recursive argument.
+Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
+our universe of data-types, as shown in
+Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
+remains unchanged, as expected. We successfully describe the
+$\DSigma{\!}{\!}$ case, by a simple appeal to the higher-order
+induction on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing
+the $\V{H}$ in $\Set$ with a recursive argument.
 
@@
 \FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
-\FreeMonad{\pair{E}{D}{}}\:X \mapsto
-    \pair{\ListCons{\DVar{}}{E}}
-         {\pair{\DSigma{X}{\DUnit}}{D}{}}{}
+\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
+    \pair{\ListCons{\DVar{}}{\V{E}}}
+         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
 }\]
 
-We simply add a constructor, $\DVar{}$, and define its argument to be
-a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
-as they were, hence leaving the functor unchanged. Unfolding the
-interpretation of this definition, we convince ourself that this
-corresponds to the functor $\FMFreeMonadD$. The fix-point operation
-ties the knot and gives us the full-blown free monad construction.
+We simply add a constructor, $\DVar{\!}$, and define its argument to
+be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
+$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
+unchanged. Unfolding the interpretation of this definition, we
+convince ourself that this corresponds to the functor
+$\FMFreeMonadD$. The fix-point operation ties the knot and gives us
+the full-blown free monad construction.
 
@@
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}   & \sigma\: x                   \\
-\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
+\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
+\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{ys}}{} & \Con{\pair{\etag{t}}{\V{ys}}{}}
 \end{array}

### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: clean-up some notes.
Desc@ICFP: merge Colors and Spaces.

@@
 that we replace $\DUnit$ by a more general $\DConst{\!}$ code. Whereas
-$\DUnit$ was interpreted as the unit set, $\DConst{X}$ is interpreted
-as $X$, for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
-$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
-succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
-its equivalent encoding. From a definitional perspective, we are
-giving more opportunities to the type-system, hence reducing the
-burden on the programmer. For the same reason, we introduce
-$\DProd{\!}{\!}$ that overlaps with $\DSigma{\!}{\!}$.
+$\DUnit$ was interpreted as the unit set, $\DConst{\V{X}}$ is
+interpreted as $\V{X}$, for any $\Bhab{\V{X}}{\Set}$. Extensionally,
+$\DConst{\V{X}}$ and $\DSigma{\V{X}}{\DUnit}$ are equivalent. However,
+$\DConst{\!}$ is more succinct. More importantly, $\DConst$ is
+\emph{first-order}, unlike its equivalent encoding. From a
+definitional perspective, we are giving more opportunities to the
+type-system, hence reducing the burden on the programmer. For the same
+reason, we introduce $\DProd{\!}{\!}$ that overlaps with
+$\DSigma{\!}{\!}$.
 
@@
     \DId            & \Desc                                    \\
-    \DConst{}       & \Set \To \Desc                           \\
-    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
-    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
+    \DConst{\!}     & \Set \To \Desc                           \\
+    \DProd{\!}{\!}  & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
+    \DSigma{\!}{\!} & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
+    \DPi{\!}{\!}    & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc 
 \end{array}
@@
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \DVar{}         & I \To \IDesc{I}                                   \\
-    \DConst{}       & \Set \To \IDesc{I}                                \\
-    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
-    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
-    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
+    \DVar{\!}         & \V{I} \To \IDesc{\V{I}}                                       \\
+    \DConst{\!}       & \Set \To \IDesc{\V{I}}                                        \\
+    \DProd{\!}{\!}    & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}       \\
+    \DSigma{\!}{\!}   & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
+    \DPi{\!}{\!}      & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
 \end{array}
@@
                 \begin{array}{l} 
-                  \LAM{n} \caseB{n}{\NatZero \To [\VecNil] \\
-                                   (\NatSuc n) \To [\VecCons{}{}]} \\
-                  \LAM{n} \caseB{n}{\NatZero \To [\DConst{Unit}] \\
-                                   (\NatSuc n) \To [\DSigma{X}{\DVar{n}}]}
+                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
+                                   (\NatSuc{\V{n}}) \mapsto [\VecCons{\!}{\!}]} \\
+                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
+                                   (\NatSuc{\V{n}}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
                 \end{array}}
@@
 \begin{array}{@{}ll}
-\IDescD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
-                                          \DConst{},
-                                          \DProd{}{},
-                                          \DSigma{}{}, 
-                                          \DPi{}{} \red{]}) \\
+\IDescD~\V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[} 
+                                          \DVar{\!},
+                                          \DConst{\!},
+                                          \DProd{\!}{\!},
+                                          \DSigma{\!}{\!}, 
+                                          \DPi{\!}{\!}
+                                             \red{]}) \\
                               & \bigRedBracket{\begin{array}{l}
@@
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \FinZero_n      & \Fin{(\NatSuc{n})}   \\
-    \FinSuc{}_n     & \Fin{n} \To \Fin{(\NatSuc{n})}
+    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
+    \FinSuc{\!}_{\V{n}}   & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
 \end{array}
@@
 \begin{array}{@{}ll}
-\FinD\: n = \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
+\FinD\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                         & \bigRedBracket{
                           \begin{array}{l}
-                            \DSigma{\Nat}{\LAM{m} \DConst{(n \PropEq \NatSuc{m})}} \\
-                            \DSigma{\Nat}{\LAM{m} \DProd{\DVar{m}}{\DConst{(n \PropEq \NatSuc{m})}}}
+                            \DSigma{\Nat}{\LAM{\V{m}} \DConst{(\V{n} \PropEq \NatSuc{\V{m}})}} \\
+                            \DSigma{\Nat}{\LAM{\V{m}} \DProd{\DVar{\V{m}}}{\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}
                           \end{array}
@@
 Intuitively, the constraint is captured as follows. First, we store an
-element $m$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
-stipulates that $m$ cannot be \emph{any} natural numbers: it must be
-``the index minus one''. This translates into the constraint $n
-\PropEq \NatSuc{m}$.
+element $\V{m}$ of $\Nat$, thanks to a $\DSigma{\!}{\!}$. However, the
+constraint stipulates that $\V{m}$ cannot be \emph{any} natural
+numbers: it must be ``the index minus one''. This translates into the
+constraint $\V{n} \PropEq \NatSuc{\V{m}}$.
 
@@
 \begin{array}{@{}llll}
-\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
-\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
+\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
+\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                      &   &             & \bigRedBracket{
@@
 \end{array} \\
-\FreeIMonad{(E,F)}{I}{X} \mapsto
-    \pair{\pair{\ListCons{\DVar{}}{(\fst{E})}} 
-               {\LAM{i}
-                \pair{(\DConst{(X\: i)})}
-                     {((\snd{E})\: i)}{}}{}}
-         {F}{}
+\FreeIMonad{(\V{E},\V{F})}{\V{I}}{\V{R}} \mapsto \\ \;\;
+    \pair{\pair{\ListCons{\DVar{\!}}{(\fst{\V{E}})}} 
+               {\LAM{\V{i}}
+                \pair{(\DConst{(\V{R}\: \V{i})})}
+                     {((\snd{\V{E}})\: \V{i})}{}}{}}
+         {\V{F}}{}
 }\]
@@
 \begin{array}{@{}ll}
-\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DConst{},
-                                              \DProd{}{},
-                                              \DSigma{}{}, 
-                                              \DPi{}{} \red{]}) \\
+\IDescFreeD~\V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[}
+                                              \DConst{\!},
+                                              \DProd{\!}{\!},
+                                              \DSigma{\!}{\!}, 
+                                              \DPi{\!}{\!}
+                                                    \red{]}) \\
                                   & \bigRedBracket{\begin{array}{l}
@@
 }\]
-
-\note{ Should we stop here? Or show off some madness? }

### 2010-03-30 papers/icfp-2010-desc/paper_desc.tex
Some suggestions to fix some typos

@@
 data-types through raw $\Sigma$-types is not a viable option. Indeed,
-just as for finite sets, encoding throw away information when we
-crucially need it. In the realm of data-types, things become even more
+just as for finite sets, encoding throws away information when we
+crucially need it. In the realm of data-types, things become even
 tougher, as one could legitimately expect an induction principle, for
@@
 defining a set of codes, we somehow define the syntax of a
-language. However, as such, a system of code is useless as it lacks a
+language. However, as such, a system of codes is useless as it lacks a
 semantics. Instead of equipping the universe of codes with some
@@
 build these descriptions. However, it should be clear that, in
-practice, these definitions can be automatically constructed from an
+practice, these definitions can be automatically constructed from a
 Haskell-like $\data$ definition.
@@
 $\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
-particular, we have considered its code, $\DSigma{\!}{\!}$,
+particular, we have considered its codes, $\DSigma{\!}{\!}$,
 $\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
@@
 lists, and binary trees. However, the class of functors expressible
-with descriptions enjoy another property: they all admit a
+with descriptions enjoys another property: they all admit a
 fix-point. Indeed, descriptions are interpreted into \emph{polynomial
@@
 \end{array} \\
-\All{\DId}{\V{X}}{\V{P}}{\V{x}} \mapsto \V{P}\: \V{x} \\
-\All{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{z}} \mapsto 
+\All{\DUnit}{\V{X}}{\V{P}}{\Void} = 
     \Unit \\
-\All{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\pair{\V{d}}{\V{d'}}{}} \mapsto 
-    \TIMES{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}{\All{\V{D'}}{\V{X}}{\V{P}}{\V{d'}}} \\
-\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} \mapsto 
-    \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
-\All{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{f}} \mapsto 
-    \PI{\V{\V{s}}}{\V{S}} \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{(\V{f}\: \V{s})}
+\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} = 
+    \All{(\V{T}\: a)}{\V{X}}{\V{P}}{\V{b}} \\
+\All{(\DIndx{\V{D}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} =
+    \TIMES{\V{P}\: \V{a}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}} \\
+\All{(\DHindx{\V{H}}{\V{D}})}{\V{X}}{\V{P}}{\pair{\V{f}}{\V{b}}{}} =
+    \TIMES{(\PI{\V{i}}{\V{H}} \V{P}\: (\V{f}\: \V{i}))}
+          {\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}}
 }
@@
 \end{array} \\
-\all{\DId}{\V{X}}{\V{P}}{\V{R}}{\V{x}} = 
-    \V{R}\: \V{x} \\
-\all{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{R}}{\V{z}} = 
+\all{\DUnit}{\V{X}}{\V{P}}{\V{R}}{\Void} =
     \Void \\
-\all{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{d}}{\V{d'}}{}} =
-    \pair{\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{d}}}{\all{\V{D'}}{\V{X}}{\V{P}}{\V{R}}{\V{d'}}}{} \\
 \all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} =
-    \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
-\all{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\V{f}} = 
-    \LAM{\V{s}} \all{(\V{T}\: \V{s})}{\V{X}}{\V{P}}{\V{R}}{(\V{f}\: \V{s})}
+    \all{(\V{T}\: a)}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
+\all{(\DIndx{D})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} =
+    \pair{\V{R}\: \V{a}}
+         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{b}}}{} \\
+\all{(\DHindx{H}{D})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{f}}{\V{b}}{}} =
+    \pair{\LAM{i} \V{R}\: (\V{f}\: \V{i})}
+         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{b}}{}
 \end{array}
@@
 Doing so, we seal data-type definition behind a convenient
-abstraction. We take advantage of this abstraction barrier to tamper
+abstraction. We take advantage of this abstraction barrier to address
 the second issue. Namely, the developer would rather be presented a
-name rather than a fix-point of a pattern functor. Surprisingly, it is
+name than a fix-point of a pattern functor. Surprisingly, it is
 sufficient to \emph{label} the $\Mu{}$ constructor with the
@@
 
-Regaining the ability to write data-type constructor is simply a
-matter of type propagation. We therefore extends the type-checking
+Regaining the ability to write data-type constructors is simply a
+matter of type propagation. We therefore extend the type-checking
 framework of Figure~\ref{fig:type-checking} with the following
 inference rule:
-\note{ This is rather drafty. I should explain what $\vec{e}$ and $\widetilde{e'}$ are. }
 
 \[
-\Rule{\forall \V{e_i} \in \V{\vec{e}}, \propag{\push{\V{e_i}}
-                                            {\descop{\V{f}\: \V{i}}{(\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})})}}}
-                                      {\V{e'_i}}}
-     {\propag{\push{\V{C}\: \V{\vec{e}}}{\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})}}}
-             {\Con{(\etag{C}, \V{\widetilde{e'}})}}}\;\etag{t} \in \V{b}
+\Rule{\propag{\push{\widetilde{\V{e}}}
+                   {\descop{\V{f}\: \V{C}}{(\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})})}}}
+            {\widetilde{\V{e}}'}}
+     {\propag{\push{\V{C}\: \vec{\V{e}}}
+                   {\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})}}}
+             {\Con{(\etag{\V{C}}, \widetilde{\V{e}}')}}}\;\etag{\V{C}} \in \V{b}
 \]
 
+Where $\vec{\V{e}}$ denotes a telescope of arguments applied to the
+constructor $\V{C}$. For the sake of readability, we write function
+application in the small function space as $\V{f}\: \V{C}$. This
+corresponds to a call to $\F{switch}$. Also, we note
+$\widetilde{\V{e}}$ the right-nested tuple terminated by $\Void$
+corresponding to the curried telescope $\vec{\V{e}}$.
+
 Hence, in the high-level expression language, the developer writes a
-familiar type constructors, applied to a telescope of arguments. In
+familiar type constructor, applied to a telescope of arguments. In
 the type propagation setting, this expression will be pushed to a

### 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
Some suggestions to fix some typos

@@
 In Section~\ref{sec:finite-sets}, we have given a specification of
-finite sets. It consists in the objects presented in
+finite sets. It consists of the objects presented in
 Figure~\ref{fig:typing-finite-set}. We are going to implement the
@@
 
-The names $\NilE$ and $\ConsE{\!}{\!}$ are absolutely not a
-coincidence. $\EnumU$ is indeed isomorphic to a list of $\UId$s. In
+The names $\NilE$ and $\ConsE{\!}{\!}$ are not a coincidence. $\EnumU$
+is indeed isomorphic to a list of $\UId$s. In
 Section~\ref{sec:desc-examples}, we have shown how to build lists in
@@
 
-Let us examine the consequences of our act. First of all, we discover
-that the type theory does not actually need to be extended with the
-type former $\EnumU$, the constructors $\NilE$ and $\ConsE$. We have
-just defined $\EnumU$ above, by fix-point over the pattern functor denoted
-by $\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and $\ListCons$
-constructors of lists.
+Let us examine the consequences. First of all, we discover that the
+type theory does not actually need to be extended with the type former
+$\EnumU$, the constructors $\NilE$ and $\ConsE$. We have just defined
+$\EnumU$ above, by fix-point over the pattern functor denoted by
+$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
+$\ListCons$ constructors of lists.
 
@@
 of $\EnumU$ does not influence our implementation of $\EnumT$: be they
-hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects just
-behave similarly. We have witnessed this effect when carrying this
+hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects
+behave the same. We have witnessed this effect when carrying the
 operation in Epigram, moving from an hard-coded presentation to a
@@
 descriptions. However, formally, the content of the type theory
-remains unchanged: objects which were present before the modification
+remains unchanged: objects that were present before the modification
 are still there. Conversely, we have not introduced any spurious
@@
 
-Our first tentative attempt is the following:
+Our first attempt is the following:
 
@@
 interpretation are presented in
-Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ use the
+Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ uses the
 elements of $H$ to index as many recursive arguments.
@@
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \ldots          & \:\:\ldots \\
-    \DHindx{\!}{\!} & \PI{\V{H}}{\Set} \Desc \To \Desc
+    \DHindx{\!}{\!} & \PI{\V{H}}{\Set} \Desc \To \Desc \\
+    \ldots          & \:\:\ldots
 \end{array}
@@
 \begin{array}{@{}l@{\:=\:\:}ll}
-\ldots                        &  \ldots \\
-\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}}
+\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
+\ldots                                    &  \ldots 
 \end{array}
@@
 $\DIndx{\!}$ is first-order. Therefore, while both codes are
-\emph{extensionally} the same, in most practical implementation they
+\emph{extensionally} the same, in most practical implementations they
 would be dealt with rather differently. While definitional equality
 can cope with first order objects, the functional presentation
-introduced by $\DHindx{\!}{\!}$ is unlikely to be amenable to a purely
+introduced by $\DHindx{\!}{}$ is unlikely to be amenable to a purely
 definitional treatment.
@@
 
-In a first glance, we have achieved our goal. We have described the
+At a first glance, we have achieved our goal. We have described the
 codes of the universe of description. Taking the fix-point of this
@@
 brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
-$\EnumT{\V{e}} \To \Desc$. This corresponds to a finite function
+$\EnumT{\V{e}} \To \Mu{\DescD}$. This corresponds to a finite function
 definition, as formalised in Figure~\ref{fig:type-checking}.
@@
 
-We are left with implementing filling the hole. Recall that we have
+We are left with filling the hole. Recall that we have
 \(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
-\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\) at hand. Our goal
-is to make an element of \(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of
-the right shape, but its sub-elements are of the wrong type. On the
-other hand, for each sub-element of $\V{xs}$, $\V{hs}$ gives us the
-corresponding element in $\V{T}$. Hence the name ``induction
-hypotheses''. Therefore, to construct an element of \(\descop{\V{D}}{\V{T}}\),
-we replace the recursive components of \(\V{xs}\) by their counterpart in
-\(\V{hs}\):
+\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\)
+at hand. Our goal is to make an element of
+\(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of the right
+shape, but its sub-elements are of the wrong type. On the other hand,
+for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
+element in $\V{T}$. Hence the name ``induction
+hypotheses''. Therefore, to construct an element of
+\(\descop{\V{D}}{\V{T}}\), we replace the recursive components of
+\(\V{xs}\) by their counterpart in \(\V{hs}\):
 
@@
                    \descop{\V{D}}{\V{Y}}} \\
-\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto \Void \\
+\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
+    \Void                                                                                                         \\
 \F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
     \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
-\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto
-    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}
+\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto       \\
+\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                     \\
+\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{xs}}{}\: \pair{\V{g}}{\V{ys}}{} \mapsto   \\
+\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\{ \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
 }\]
@@
 data-type will now come equipped with this operation, for free.
+
+The astute reader will have been struck by the type of $\F{replace}$:
+it is \emph{almost} the morphism part -- sometimes called \emph{map}
+-- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
+the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
+is the dependent version of the map, which uses the induction
+hypotheses. For space reason, we will not present the non-dependent
+map. It can be found in the Agda model.
 
@@
 basically, a function mapping a $\Desc$ to a data-type specific
-operation.
+operation. This programming style is reminiscent to polytypic
+languages, such as PolyP~\cite{jansson:polyp}.
 
@@
 \stk{
-\data \FMFreeMonad : \PITEL{\V{F}}{\Set \To \Set} 
-                     \PI{\V{X}}{\Set} 
+\data \FMFreeMonad\: \PITEL{\V{F}}{\Set \To \Set} 
+                     \PITEL{\V{X}}{\Set} :
                      \Set 
@@
 unchanged. Unfolding the interpretation of this definition, we
-convince ourself that this corresponds to the functor
+convince ourselves that this corresponds to the functor
 $\FMFreeMonadD$. The fix-point operation ties the knot and gives us
@@
 Of course, we must equip the resulting data-types with operations
-delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
-performs the r\^ole of \return, embedding variables into terms. The
-\bind operation corresponds to \emph{substitution}. We will now
+delivering a monadic interface. As expected, \(\LAM{\x}\DVar{\x}\)
+plays the r\^ole of \return, embedding variables into terms. The
+\bind\ operation corresponds to \emph{substitution}. We will now
 implement it, as a generic function.
@@
 catamorphism. Intuitively, its role is to catch appearances of
-$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. Otherwise, it should simply
-appeal to the induction hypothesis. This corresponds to the following
-definition:
+$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
+corresponds to the following definition:
 
@@
 
-Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}$
-completes the implementation. To sum up, we have implemented the free
-monad construction for an arbitrary tagged description. This gives the
-developer the ability, for any data-type, to extend it with a notion
-of variable. Then, we have equipped this structure with the
-corresponding monadic operation, \bind and \return. Whereas the
-\return is a trivial variable former, the \bind operation corresponds
-to the substitution of variables by terms. We have shown how to
-implement it, for any free monad.
+Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
+\V{\sigma}$ completes the implementation. To sum up, we have
+implemented the free monad construction for an arbitrary tagged
+description. This gives the developer the ability, for any data-type,
+to extend it with a notion of variable. Then, we have equipped this
+structure with the corresponding monadic operation, \bind\ and
+\return. Whereas the \return\ is a trivial variable former, the
+\bind\ operation corresponds to the substitution of variables by
+terms. We have shown how to implement it, for any free monad. This
+programming style is reminiscent to the type-indexed data
+types~\cite{hinze:generic-haskell}, as implemented in Generic Haskell:
+from data-types, we built new data-types and equip them with some
+structure.
 
@@
 are therefore entitled to modify this code and, in this case, extend
-it. Extending a data-type might give rise to a more structured
-object, as was the case here.
-
-So, secondly, we have equipped this new data-type with its inherent
-structure: the \bind and \return operations. We have been able to
-build them as generic functions. Here, we rely on our ability to
-compute over descriptions.
+it. Extending a data-type might give rise to a more structured object,
+as was the case here.  So, secondly, we have equipped this new
+data-type with its inherent structure: the \bind\ and
+\return\ operations. We have been able to build them as generic
+functions.

### 2010-03-30 papers/icfp-2010-desc/paper_discussion.tex
Some suggestions to fix some typos

@@
 universe stratification is orthogonal to our work. Nonetheless, our
-universes of description rather naturally leads itself to
+universe of description rather naturally leads itself to
 stratification. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of type
@@
 generic functions, in our system, are built by induction on the
-pattern functor. Unlike PolyP, we do not have to resort to
-pre-processing code: our data-types are, natively, nothing but codes.
+pattern functor. Unlike PolyP, we do not have to resort to pre-process
+data-type definitions: our data-types are, natively, nothing but
+codes.
 
@@
 Haskell~\cite{hinze:generic-haskell} model. This model, based on
-type-indexed data types, corresponds to computing new data-types from
+type-indexed data types, enables computing new data-types from
 others. This is natural in our system, as data-types descriptions are
@@
 
-Interestingly, our generic induction principle allows a programming
-style reminiscent to the Scrap Your Boilerplate~\cite{spj:syb} (SYB)
-approach. Like SYB, we provide a generic operator to manipulate and
-carry operations over data-types. However, unlike SYB, we do not need
-support from the compiler, as introduced by the $\CN{Typeable}$
-class. Indeed, in our setting, data-types are reflected in the
-language, hence our ability to manipulate them directly.
+Another generic programming framework is Scrap Your
+Boilerplate~\cite{spj:syb} (SYB). Our proposal is different in various
+ways. The corner stone of SYB is the \emph{spine} view of data-type
+constructors. A data-type is a spine composed by a constructor applied
+to some arguments. Further, this spine is equipped with some
+combinators including, primarily, an iterator. In this setting,
+generic programs are written by composing these combinators. This
+relies on a $\CN{Typeable}$ type-class, allowing dynamic dispatch to
+data-type specific operations. As a result, SYB is not reflexive: it
+is restricted to data-types instanciating $\CN{Typeable}$. Moreover,
+it forbids type-indexed data types: it can only define generic
+functions.
 
@@
 programming in the Coq theorem prover. However, these works aim at
-\emph{modelling} Generic Haskell in a dependently-typed setting, for
-the purpose of proving correctness properties of Haskell code. Our
-approach is different in that we aim at building a foundation for
-data-types in a dependently typed theory.
+\emph{modelling} PolyP or Generic Haskell in a dependently-typed
+setting, for the purpose of proving correctness properties of Haskell
+code. Our approach is different in that we aim at building a
+foundation for data-types, in a dependently-typed theory, for a
+dependently-typed system.
 
@@
 al.~\cite{benke:universe-generic-prog}. This seminal work introduced
-the usage of universes for developing generic programs. Moreover, our
-own universes are rather similar to theirs: our universe of
-descriptions is similar to their universe of iterated induction, and
-our universe of indexed descriptions is isomorphic to their universe of
-finitary indexed induction. This is not surprising, as we share the
-same source of inspiration, namely induction recursion.
+the usage of universes for developing generic programs. Our own
+universes are rather similar to theirs: our universe of descriptions
+is similar to their universe of iterated induction, and our universe
+of indexed descriptions is isomorphic to their universe of finitary
+indexed induction. This is not surprising, as we share the same source
+of inspiration, namely induction recursion.
 
@@
 on a generic induction and a specialised
-$\switchD{\!}{\!}{\!}$. Second, the authors develop several universes,
-each with its own strengths and weaknesses, but lack a formal,
-unifying universe. With indexed descriptions, we subsume simple
-descriptions, hence avoiding the move to and from several
-codings. Finally, the authors often resort to an extensional notion of
-equality, while we have given an equality-agnostic
-presentation. Moreover, our presentation is arranged so as to use
-definitional equality as much as possible. Hence, in practice, the
-developer is relieved from many trivial proof obligations.
+$\switchD{\!}{\!}{\!}$. Second, the authors do not tackle the issue of
+\emph{programming} with such codes: it is not obvious, to us, how one
+can define, manipulate, and program over these coded data-types. With
+descriptions, we have shown how to abstract away codes and present a
+convenient and familiar presentation to the developer. Finally, the
+authors often resort to an extensional notion of equality, while we
+have given an equality-agnostic presentation. Beside, our presentation
+is arranged so as to use definitional equality as much as
+possible. Hence, in practice, the developer is relieved from many
+proof obligations.

### 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
Some suggestions to fix some typos

@@
 induction. In the case of vectors, for instance, we have to define the
-whole \emph{family} of vector in one go: vectors of all size needs to
+whole \emph{family} of vector in one go: vectors of all sizes need to
 be defined at the same time. Where simply-typed languages have
-inductive types, in dependently typed programming we the basic grammar of
+inductive types, in dependently typed programming the basic grammar of
 data-types should be that of inductive families of types. To this end, we
@@
 \subsection{Desc, atomically}
+\label{sec:idesc-atomic-desc}
 
@@
 
-Before moving on indexed descriptions, we have to carry out some
+Before moving on to indexed descriptions, we have to carry out some
 maintenance work on descriptions. We presented $\Desc$ as the grammar
@@
 \[\stk{
-\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
-\;\;\ldots \\
+\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \:\ldots \\
 \\
-\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
+\idescop{\_}{}{} : _{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
 \ldots
@@
 
-Note that $I \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
+Note that $\V{I} \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
 endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
-$\V{I} \To (\V{I} \To \Set) \To \Set$, which is isomorphic to $(\V{I} \To \Set)
-\To (\V{I} \To \Set)$. Fix-point of inductive families is defined over
+$\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
+\To \Set^{\V{I}}$. Fix-point of inductive families is defined over
 these indexed functors, hence computing a fix-point of the entire
@@
 \[\stkl{
-\Rule{\stk{ 
-          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
+\Rule{\Gamma \vdash \Bhab{\V{I}}{\Set} \qquad
+      \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}}}
+     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\!}}{\V{I}\To\Set}} \qquad
+\\:\
+\Rule{\begin{array}{l@{\qquad}l}
+          \Gamma \vdash \Bhab{\V{I}}{\Set} &
           \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
-          \Gamma \vdash \Bhab{\V{i}}{\V{I}}}}
-     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\V{i}}}{\Set}} \qquad
-\Rule{\stk{
-          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
-          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
-          \Gamma \vdash \Bhab{\V{i}}{\V{I}} \\
-          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{})}}}}
+          \Gamma \vdash \Bhab{\V{i}}{\V{I}} &
+          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\:\V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{\!})}}
+      \end{array}}
      {\Gamma \vdash \Bhab{\Con{\V{x}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}}
@@
 \stk{
-\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
+\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \\
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
@@
 
-\subsection{Tagged indexed descriptions}
-
-\begin{wstructure}
-<- Tagged constructor choice
-    <- Index available when defining the data-type
-        -> Can influence the choice of constructors
-        -> Dependently-typed data-types
-            <- term (index) influence types
-    -> Two parts
-        <- Always on the menu
-            <- E : EnumU 
-            <- ED : E -> IDesc I
-        <- Index-dependent
-            <- F : I -> EnumU 
-            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
-        -> taggedIDesc I == Sigma E ED x Sigma F FD
-        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
-\end{wstructure}
-
-In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
-form for descriptions. This format follows the usual presentation of
-inductive types as sum-of-product. With indexed description, this
-definition can be generalized. Indeed, when defining an indexed
-data-type, we have access to this index. Therefore, we can use this
-index to influence the choice of constructors. This captures the
-essence of dependent data-types: a term -- the index -- has the
-ability to influence the data-type.
-
-For convenience, we divide a tagged indexed description in two parts:
-first, the constructors that do not depend on the index; then, the
-constructors that do. The non-dependent part mirrors the definition
-for non-indexed descriptions: we are provided a finite choice of
-constructors. The index-depend part simply indexes the choice of
-constructors by $I$. Hence, by inspecting the index, it is possible to
-enable or disable the constructors.
-
-\[\stk{
- \TagIDesc{\V{I}}  \mapsto \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
- \ATagIDesc{\V{I}} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
- \ITagIDesc{\V{I}} \mapsto \\
-\qquad\SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
-}\]
-
-\begin{wstructure}
-<- Vectors
-    Do we treat them in the end? 
-    What can we say here we haven't with Fin?
-\end{wstructure}
-
-In the case of vectors, for instance, for the index $\NatZero$, we
-would only propose the constructor $\ListNil$. Similarly, for
-$\NatSuc{n}$, we only propose the constructor $\ListCons{\!}{\!}$. Hence,
-we obtain the following definition:
-
-\newcommand{\VecD}{\C{VecD}}
-\newcommand{\VecNil}{\etag{\CN{vnil}}}
-\newcommand{\VecCons}[2]{\etag{\CN{vcons}}\:#1\:#2}
-
-\[\stk{
-\VecD : \Set \To \TagIDesc{\Nat} \\
-\begin{array}{@{}ll}
-\VecD\: \V{X} = \\ \quad\bigRedBracket{ 
-            \begin{array}{l}
-                \pair{[]}{\_}{} 
-                \\\\
-                \bigRedBracket{
-                \begin{array}{l} 
-                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
-                                   (\NatSuc{\V{n}}) \mapsto [\VecCons{\!}{\!}]} \\
-                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
-                                   (\NatSuc{\V{n}}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
-                \end{array}}
-            \end{array}}
-\end{array}
-}\]
-
-We will see more instances of this pattern in the following.
+%\subsection{Tagged indexed descriptions}
+
+% This discussion seemed out of place to me, so i've moved it a bit later (pwm).
+
+%\begin{wstructure}
+%<- Tagged constructor choice
+%    <- Index available when defining the data-type
+%        -> Can influence the choice of constructors
+%        -> Dependently-typed data-types
+%            <- term (index) influence types
+%    -> Two parts
+%        <- Always on the menu
+%            <- E : EnumU 
+%            <- ED : E -> IDesc I
+%        <- Index-dependent
+%            <- F : I -> EnumU 
+%            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
+%        -> taggedIDesc I == Sigma E ED x Sigma F FD
+%        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
+%\end{wstructure}
+%
+%In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
+%form for descriptions. This format follows the usual presentation of
+%inductive types as sum-of-product. With indexed description, this
+%definition can be generalized. Indeed, when defining an indexed
+%data-type, we have access to this index. Therefore, we can use this
+%index to influence the choice of constructors. This captures the
+%essence of dependent data-types: a term -- the index -- has the
+%ability to influence the data-type.
+%
+%For convenience, we divide a tagged indexed description in two parts:
+%first, the constructors that do not depend on the index; then, the
+%constructors that do. The non-dependent part mirrors the definition
+%for non-indexed descriptions: we are provided a finite choice of
+%constructors. The index-depend part simply indexes the choice of
+%constructors by $I$. Hence, by inspecting the index, it is possible to
+%enable or disable the constructors. 
+%
+%\[\stk{
+% \TagIDesc{\V{I}}  \mapsto \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
+% \ATagIDesc{\V{I}} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
+% \ITagIDesc{\V{I}} \mapsto \\
+%\qquad\SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
+%}\]
+%
+%
+%\begin{wstructure}
+%<- Vectors
+%    Do we treat them in the end? 
+%    What can we say here we haven't with Fin?
+%\end{wstructure}
+%
+%In the case of vectors, for instance, for the index $\NatZero$, we
+%would only propose the constructor $\ListNil$. Similarly, for
+%$\NatSuc{n}$, we only propose the constructor $\ListCons{\!}{\!}$. Hence,
+%we obtain the following definition:
+%
+%\newcommand{\VecD}{\C{VecD}}
+%\newcommand{\VecNil}{\etag{\CN{vnil}}}
+%\newcommand{\VecCons}[2]{\etag{\CN{vcons}}\:#1\:#2}
+%
+%\[\stk{
+%\VecD : \Set \To \TagIDesc{\Nat} \\
+%\begin{array}{@{}ll}
+%\VecD\: \V{X} = \\ \quad\bigRedBracket{ 
+%            \begin{array}{l}
+%                \pair{[]}{\_}{} 
+%                \\\\
+%                \bigRedBracket{
+%                \begin{array}{l} 
+%                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
+%                                   (\NatSuc{\V{n}}) \mapsto [\VecCons{\!}{\!}]} \\
+%                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
+%                                   (\NatSuc{\V{n}}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
+%                \end{array}}
+%            \end{array}}
+%\end{array}
+%}\]
+%
+%We will see more instances of this pattern in the following.
 
@@
 
-\subsubsection{Examples}
+\subsection{Examples}
 \label{sec:idesc-examples}
@@
 In order to gain some intuition with $\IDesc{\!}$, let us re-implement
-the pattern functor of natural numbers. We refer the reader to
-Section~\ref{sec:desc-examples} for its incarnation in the universe of
-descriptions. The code is the following:
+the pattern functor of natural numbers. This follows the implementation of
+$\NatD$ from \ref{sec:idesc-atomic-desc}:
 
@@
 \NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
-                     {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
-}\]
-
-Because $\Nat$ is just an inductive type, we make no use of the index:
-it is the trivial inhabitant of $\Unit$. Therefore, the recursive
+                     {[ (\DConst{\Unit}) \quad (\DVar{\Void}) ]}
+}\]
+
+Because $\Nat$ is just an inductive type, $\NatD$ represents a
+$\Unit$-indexed functor. Therefore, the recursive
 argument is materialised by $\DVar{\Void}$, where we were using $\DId$
-in the $\Desc$ presentation. This transformation applies to all
-inductive types: we have not lost in expressive power during the
-transition.
+in the $\Desc$ presentation. Indeed, all inductive types can be
+translated into $\IDesc{\!}$ in exactly this way.
 
@@
 
-A mandatory exercise consists in describing $\IDesc{\!}$ in itself. With
-the experience gained in Section~\ref{sec:desc-levitate}, this
-levitation does not pose any issue:
+If we note that, for any $\V{I}$, $\IDesc{\V{I}}$ itself is merely an inductive 
+\emph{type} then we can certainly define its encoding in $\IDesc{\Unit}$:
 
@@
 
-Note that this description also lives in $\IDesc{\Unit}$. Indeed,
-for any $\V{I}$, $\IDesc{\V{I}}$ is merely an inductive type. As for the
-$\Desc$ levitation, the finite function space $[ \ldots ]$ elaborates into a
-special purpose $\F{switchID}$ operator.
+Using this description we can repeat the act of levitation 
+from section~\ref{sec:desc-levitate}. As before we rely on a special purpose 
+$\F{switchID}$ operator to build the finite function space
+$\bigRedBracket{\ldots}$.
+
+\paragraph{Vectors:}
+
+\newcommand{\VecD}{\C{VecD}}
+\newcommand{\VecNil}{\etag{\CN{vnil}}}
+\newcommand{\VecCons}[2]{\etag{\CN{vcons}}\:#1\:#2}
+
+So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
+not using any indexing. We remedy this by encoding the vectors:
+
+Recall that the constructors of $\D{Vec}$. The constructors $\VecNil$ and
+$\VecCons$ are only defined for an index $\NatZero$ and $\NatSuc$ 
+respectively.:
+
+\[
+\stk{
+\data \D{Vec}\: \PITEL{\V{X}}{\Set} : \PI{\V{n}}{\Nat} \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
+    \VecNil      & \D{Vec}\:\V{X}\:{\NatZero)}   \\
+    \VecCons     & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
+\end{array}
+}
+\]
+
+One way to code constrained data-type is to appeal to equality. The
+constraints are therefore captured by equations in the data-type. In
+this case, we obtain the following definition:
+
+\[\stk{
+\VecD : \Set \To \Nat \To \IDesc{\Nat} \\
+\begin{array}{@{}ll}
+\VecD\: \V{X}\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{[ \VecNil , \VecCons{\!}{\!} ]}) \\
+                        & \bigRedBracket{
+                          \begin{array}{l}
+                            \DConst{(\V{n} \PropEq \NatZero)} \\
+                            \DSigma{\Nat}{\LAM{\V{m}} \stk{\DProd{\DVar{\V{m}}}{\\\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}}
+                          \end{array}
+                          }
+\end{array}
+}\]
+
+If the $\VecNil$ tag is chosen, a proof must be provided that the index is equal
+to $\NatZero$. In the $\VecCons$ case, we first store an
+element $\V{m}$ of $\Nat$, with a $\DSigma{\!}{\!}$. However, the
+constraint stipulates that $\V{m}$ cannot be \emph{any} natural
+numbers: it must be ``the index minus one''. This translates into the
+constraint $\V{n} \PropEq \NatSuc{\V{m}}$.
+
+However, one could, quite rightfully, be worried by the introduction
+of equalities in the picture. In a programming context, equalities
+haunt us, begging for proofs. Hence, whereas the developer typed a
+constructor-constrained data-type, the elimination form will involve
+equalities and their proofs. As much as possible, one would rather
+avoid suffering from this burden. However, the equalities introduced
+here are rather special: they are \emph{equations} constraining
+variables to certain terms. \citet{goguen:pattern-matching} have
+shown, in the context of dependent pattern matching, how to tame these
+equations.
+
+\begin{wstructure}
+!!! Need Help !!!
+<- Brady optimisation: forcing
+    <- Source to source translation
+    <- Able to remove some constraints
+    -> Example: Fin [figure]
+    ??? More technical detail needed
+\end{wstructure}
+
+Further, we can actually remove these equations altogether. Let us
+look back at $\D{Vec}$. We note that the equations are introduced
+because we are \emph{storing} the index of the inductive family.
+However, \emph{inductive families need not store their
+  indices}~\cite{brady:index-inductive-families}.  By examining the
+incoming index, we can apply the \emph{forcing} and \emph{de-tagging}
+optimisations to our initial definition of $\D{Vec}$ gives the
+following, equivalent definition:
+
+\[\stk{
+\VecD \:\PITEL{\V{X}}{\Set} : \Nat \To \IDesc{\Nat} \\
+\begin{array}{@{}lll}
+\VecD\:\V{X}\: \NatZero     & = & \DConst{\Unit} \\
+\VecD\:\V{X}\: (\NatSuc{\V{n}}) & = & \DProd{\DConst{\V{X}}}{\DVar{\V{n}}}
+\end{array}
+
+}\]
+
+The equations (and constructors) have simply disappeared. Note that it is
+straightforward to turn this definition into a tagged form, where in each
+case, only one constructor choice is available to the developer. In another
+example, $\Fin{\!}$, we can force, but not detag, since both $\FinZero$ and 
+$\FinSuc{\!}$ both target $\NatSuc{\!}$:
+
+\[
+\stk{
+\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
+    \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
+    \FinSuc{\!}   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
+\end{array}
+\vspace{0.1in}
+\\
+\FinD : \Nat \To \IDesc{\Nat} \\
+\begin{array}{@{}llll}
+\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
+\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
+                     &   &             & \bigRedBracket{
+                                         \begin{array}{l}
+                                         \DConst{\Unit} \\
+                                         \DVar{\V{n}}
+                                         \end{array}
+                                         }
+\end{array}
+}\]
+
+We should
+precise that forcing a description is not guaranteed to remove all
+constraints. It is subject to future work to see if constraints can be
+entirely eradicated, or presented more conveniently to the developer.
+Finally, it is worth mentioning that these optimisations are 
+\emph{source-to-source} transformations on descriptions.
+
+\subsubsection{Tagged Indexed Descriptions}
+
+In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
+form for descriptions. This format follows the usual presentation of
+inductive types as sum-of-product. With indexed description, this
+definition can be generalized. Indeed, when defining an indexed
+data-type, we have access to this index. Therefore, we can use this
+index to influence the choice of constructors. This captures the
+essence of dependent data-types: a term -- the index -- has the
+ability to influence the data-type.
+
+For convenience, we divide a tagged indexed description in two parts:
+first, the constructors that do not depend on the index; then, the
+constructors that do. The non-dependent part mirrors the definition
+for non-indexed descriptions: we are provided a finite choice of
+constructors. The index-depend part simply indexes the choice of
+constructors by $\V{I}$. Hence, by inspecting the index, it is
+possible to enable or disable the constructors.
+
+\[\stk{
+ \TagIDesc{\V{I}}  \mapsto \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
+ \ATagIDesc{\V{I}} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
+ \ITagIDesc{\V{I}} \mapsto \\
+\qquad\SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
+}\]
+
+\begin{wstructure}
+<- Vectors
+    Do we treat them in the end? 
+    What can we say here we haven't with Fin?
+\end{wstructure}
+
+In the case of a tagged $\D{Vec}$, for instance, for the index $\NatZero$, we
+would only propose the constructor $\ListNil$. Similarly, for
+$\NatSuc{n}$, we would only propose the constructor $\ListCons{\!}{\!}$. We
+will see a concrete example of this in the following example.
 
@@
 %% Index mapper (terminology?)
-\newcommand{\Val}[1]{\D{Val}~#1}
-\newcommand{\Var}[2]{\D{Var}_{#1}~#2}
+\newcommand{\Val}[1]{\F{Val}~#1}
+\newcommand{\Var}[2]{\F{Var}_{#1}~#2}
 
@@
 \newcommand{\HExprD}{\C{ExprD}}
-\newcommand{\HExprAD}{\C{ExprAlwaysD}}
-\newcommand{\HExprID}{\C{ExprIndexedD}}
-\newcommand{\HExprVarD}[1]{\C{ExprD}_{\C{Var},#1}}
+\newcommand{\HExprAD}{\C{ExprAD}}
+\newcommand{\HExprID}{\C{ExprID}}
+\newcommand{\HExprVarD}[1]{\C{ExprD}_{\F{Var},#1}}
 \newcommand{\HExprFreeD}{\C{ExprD}^{\C{Free}}}
-\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}^{\C{Free}}}
+\newcommand{\HExprAFreeD}{\C{ExprAD}^{\C{Free}}}
 
@@
 respectively named $\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$
-as follow:
+as follows:
 
@@
 resulting description is indexed by $\Ty$. We observe that some
-constructors are ``polymorph'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
+constructors are ``polymorphic'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
 other hand, the $\Eplus$ and $\Ele$ constructors are
@@
 
-\begin{figure*}
-
-\[\stkc{
+\begin{figure}
+
+\[\stk{
 \stk{
@@
 \\
-\begin{array}{ll}
 \stk{
@@
                  \begin{array}{l}
-                   \EnumT{[\Eval{}, \Econd{}{}{}]} \red{,} \\
+                   {[\Eval{}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
                    \LAM{\V{ty}}
@@
 \\
-} &
+} 
+\\
 \stk{
@@
                  \begin{array}{l}
-                   \LAM{\V{t}y} \case{\V{t}y}{\Enat \To \EnumT{[\Eplus{\!}{\!}\!]} \\ \Ebool \To \EnumT{[\Ele{\!}{\!}\!]}} \red{,} \\
+                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To {[\Eplus{\!}{\!}\!\,]} \\ \Ebool \To {[\Ele{\!}{\!}\!\,]}} \red{,} \\
                    \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
@@
 }
-\end{array}
 }\]
@@
 
+Where we overload the notation $\toIDesc{\V{TID}}$ to denote the
+indexed description computed from the tagged indexed description
+$\V{TID}$.
+
 The type of $\F{eval}$ is strikingly similar to a
@@
   \F{cataI}!} gives, for free, the full evaluator. The implementation
-is as follow:
-
-\[\stk{
-\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\: \V{ty}}{\Ty}{\V{ty}}
+is as follows:
+
+\[\stk{
+\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\:\V{ty}}{\Ty}{\Val{\!}}
 \To {\Val{\V{ty}}} \\
@@
 \emph{closed} terms. By abstracting over $\Val{}$, it is possible to
-build and manipulate \emph{open} terms, that is terms with symbolic
+build and manipulate \emph{open} terms, that is, terms with symbolic
 variables. On the model of $\Val{}$, we define $\Var{}{}$:
@@
 constants. For readability, we will abbreviate $\LAM{\V{ty}}
-\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
+\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ into $\SUM{\Val{\!}}{\Var{\V{dom}}{\!}}$. \note{Is
   it polite and comprehensible to ask that?} This defines a new
-indexed description, called $\HExprVarD{dom}$.
+indexed description, called $\HExprVarD{\V{dom}}$.
 
@@
 \begin{array}{@{}ll}
-\substH  : & \PITEL{\V{\dom}}{\EnumU} \\
+\substH  : & \PITEL{\V{dom}}{\EnumU} \\
            & \PITEL{\V{\gammaNat}}{\GammaSpi{\V{dom}}{\Enat}} 
@@
                              & \PI{\V{ty}}{\Ty} (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To \\
-                             & \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}}}
+                             & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}}
              \end{array}\\
@@
 syntax with variables: our data-type enjoys more structure than what
-we are given. In particular, this is reminiscent to a free monad,
+we are given. In particular, this is reminiscent of a free monad,
 where $\Eval{}$ is the \return\ introducing variables. The substitution
-$\sigma$ is implemented from $\discharge$, by picking the domain and
+$\V{\sigma}$ is implemented from $\discharge$, by picking the domain and
 context corresponding to the type:
@@
 \end{array} \\
-\FreeIMonad{(\V{E},\V{F})}{\V{I}}{\V{R}} \mapsto \\ \;\;
+\FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto \\ \;\;
     \pair{\pair{\ListCons{\DVar{\!}}{(\fst{\V{E}})}} 
@@
 Just as in the universe of descriptions, this construction comes with
-an obvious \return and a substitution operation, the \bind. Its type
+an obvious \return\ and a substitution operation, the \bind. Its type
 signature is the following:
@@
 datatypes $\HExprD$ and $\HExprVarD{\V{dom}}$ enjoy a monadic structure. We had
-identified the variable substitution operation as the \bind of a free
-monad. The definition of $\substI$ above confirm our intuition: using
+identified the variable substitution operation as the \bind\ of a free
+monad. The definition of $\substI$ above confirms our intuition: using
 $\substI$, we should easily obtain $\substH$. To do so, we first have
@@
 
-As previously mentioned, we identify $\Eval{}$ as the \return of the
+As previously mentioned, we identify $\Eval{}$ as the \return\ of the
 free monad, while the other components are the action of the monad. As
 a result, the definition is similar to $\HExprD$ presented in
-Figure~\ref{fig:hexpr-full}, at the exception of $\HExprAD$ that is
-replaced by $\HExprAFreeD$:
+Figure~\ref{fig:hexpr-full}, replacing $\HExprAD$ with $\HExprAFreeD$:
 
@@
 corresponds to the syntax of closed terms, $\HExprD$. Similarly,
-$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{\V{ty}} \SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})}$$
+$\FreeIMonad{\HExprFreeD}{\Ty}{\SUM{\Val{\!}}{\Var{\V{dom}}{\!}}}$
 corresponds to expressions with variables, $\HExprVarD{\V{dom}}$.
@@
 
-The evaluator for closed term we implemented in
+The evaluator for closed terms we implemented in
 Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
@@
        \substI\: & \Ty\: 
-                  (\SUM{\Val{}}{\Var{\V{dom}}{}})\: 
+                  (\SUM{\Val{\!}}{\Var{\V{dom}}{\!}})\: 
                   \Val{}\:
@@
 terms. We started with a well-typed language of arithmetical
-expressions. We have seen how to take advantage of indexes to only
-present type-safe constructors. Then, we have implemented an evaluator
-for closed term, based on the generic catamorphism function. Having
-implemented an open term representation, we wanted a substitution
-operation, in order to close open terms in a context. Again, we have
-implemented this operation with the generic substitution
-operator. Hence, without much efforts, we have described the syntax of
-a well-typed language, together with its semantics.
+expressions. We have seen how to take advantage of indexes to define
+type-safe constructors. Then, we have implemented an evaluator for
+closed term, based on the generic catamorphism function. Having built
+an open term representation, we wanted a substitution operation, in
+order to close open terms in a context. Again, we have implemented
+this operation with the generic substitution operator. Hence, without
+much efforts, we have described the syntax of a well-typed language,
+together with its semantics.
 
@@
 
-\paragraph{Indexed description:}
+\paragraph{Indexed descriptions:}
 
@@
 \[\stk{
-\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
+\IDescFreeD : \ATagIDesc{\Unit} \\
 \begin{array}{@{}ll}
-\IDescFreeD~\V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[}
-                                              \DConst{\!},
+\IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
+                                [\DConst{\!},
                                  \DProd{\!}{\!},
                                  \DSigma{\!}{\!}, 
-                                              \DPi{\!}{\!}
-                                                    \red{]}) \\
-                                  & \bigRedBracket{\begin{array}{l}
+                                 \DPi{\!}{\!}] \red{,}\\
+                                  \LAM{\_}\bigRedBracket{\begin{array}{l}
                                         \DConst{\Set}               \\
@@
                                         \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
-                                    \end{array}}
+                                    \end{array}}\end{array}}
 \end{array}
@@
 \[\stk{
-\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
-\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{]}}{\Unit}{\V{I}}\: \V{I}
-}\]
+\IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
+\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}[]\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
+}\]

### 2010-03-30 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: clean-up some notes.
Desc@ICFP: merge Colors and Spaces.
Some suggestions to fix some typos

@@
 possible. To this end, we are going to present a type theory from the
-ground-up.
+ground up.
 
@@
 familiar with type theory, she will recognise the basis of her
-favourite type theory, be it Agda or Coq for instance. This scene sets,
+favourite type theory, be it Agda or Coq, for instance. This scene sets,
 we have a common ground for discussion.
@@
 judgements, typing judgements, and equality judgements. Formally,
-these judgements are presented as follow:
+these judgements are presented as follows:
 
@@
 features an intensional equality. On the other hand, it is used in
-Epigram, which equality has a slightly extensional
+Epigram, whose equality has a slightly extensional
 flavor~\cite{altenkirch:ott}. We expect users of fully extensional
@@
 implemented by a terminating algorithm.
-
 \note{Need some care here. Expansion would be good too. I wanted to
@@
 were the democracy analogy goes crazy, I think).
-
 \note{Did I got the coding presentation vs. coding representation
@@
 type. Intuitively, it corresponds to an index -- a number -- pointing
-to an element of $u$.
+to an element of $\V{u}$.
 
@@
 \emph{finite function space}, mapping elements of finite sets to
-objects in $\Set$. Intuitively, this operator build a right-nested
+objects in $\Set$. Intuitively, this operator builds a right-nested
 tuple, effectively mapping an object of $P\:i$ for each $i$ the finite
@@
 faced with encoded data: while we were suffering from a crucial lack
-of information, we are now facing too much information, hence loosing
+of information, we are now facing too much information, hence losing
 focus. This is a general issue with the usage of codes, as they convey
@@
 should not be worried by information overload, but how to use it as
-much as possible. Therefore, we should not be afraid by using code for
+much as possible. Therefore, we should not be afraid of using codes for
 practical purposes.
@@
 $\exprEx$ are subject to type \emph{synthesis}.
-
 \note{Why don't we need a context validity judgement, again?}
 
-The purpose of type propagation is to \emph{elaborate} an high-level
+The purpose of type propagation is to \emph{elaborate} a high-level
 expression into a term of the type theory, together with its
@@
 
-\note{This might deserve a more thorough explanation, with snippet of
-  code.}
-
 Finally, and more importantly, we are able to present the developer an
-abstraction of \emph{finite} function space. This work as follow. When
+abstraction of \emph{finite} function space. This works as follow. When
 an expression $\V{t}$ is pushed into a function which domain is a finite
@@
 
-The abstraction we have developed in this section are entirely
+The abstraction we have developed in this section is entirely
 transparent to the user. In the case of finite function space, the
-developer has never to explicitly \emph{call} the $\F{switch}$
+developer never has to explicitly \emph{call} the $\F{switch}$
 eliminator. For diagnosis purposes, a similar, reverse mechanism turns
-low-level terms into high-level expressions. Hence, this system, which
+low-level terms into high-level expressions. Hence, this system, whose
 complete presentation is out of the scope of this paper, seals the
@@
 
-This example teaches us that we should not be afraid by codes, as soon
+This example teaches us that we should not be afraid of codes, as soon
 as type information is available. We have shown how to rationalise

### 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix typos in Section 4, 5.4, 5.5, 6, and 7

@@
 was to effectively implement the code of the $\Desc$ universe. Another
-promise we tacitly was the existence of the type formers for finite
-sets. Despite being a perilous pedagogical exercise for the authors,
-we hope to convey to the reader the dizzy feeling of levitation,
-without the falling.
+promise we tacitly made was the existence of the type formers for
+finite sets. Despite being a perilous pedagogical exercise for the
+authors, we hope to convey to the reader the dizzy feeling of
+levitation, without the falling.
 
@@
 In Section~\ref{sec:finite-sets}, we have given a specification of
-finite sets. It consists of the objects presented in
+finite sets. It consists in the objects presented in
 Figure~\ref{fig:typing-finite-set}. We are going to implement the
@@
 We shall now fulfil our promises for $\Desc$. We are going to
-implement the codes of the universe of descriptions. Interestingly, in
-and by itself, the system of codes, $\Desc$, is nothing but a data-type. We
-are in the same situation than with $\EnumU$: we ought to be able to describe
+implement the codes of the universe of descriptions. In and by itself,
+the system of codes, $\Desc$, is nothing but a data-type. We are in
+the same situation than with $\EnumU$: we ought to be able to describe
 the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
@@
 constructor. Constructors correspond to the code's constructors:
-$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice that the
-tagged notation we have used so far for $\Desc$ constructors now fully
-make sense: these were actually the tags we are defining. \note{This
-  sentence badly needs a rephrasing.}
+$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
+that the tagged notation we have used for the $\Desc$ constructors now
+fully make sense: these were actually the tags we are
+defining. \note{This sentence badly needs a rephrasing.}
 
@@
 interpretation are presented in
-Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ uses the
+Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ use the
 elements of $H$ to index as many recursive arguments.
@@
 $\DIndx{\!}$ is first-order. Therefore, while both codes are
-\emph{extensionally} the same, in most practical implementations they
+\emph{extensionally} the same, in most practical implementation they
 would be dealt with rather differently. While definitional equality
@@
 Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
-remains unchanged, as expected. We successfully describe the
-$\DSigma{\!}{\!}$ case, by a simple appeal to the higher-order
+remain unchanged, as expected. We successfully describe the
+$\DSigma{\!}{}$ case, by a simple appeal to the higher-order
 induction on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing
@@
                    \descop{\V{D}}{\V{Y}}} \\
-\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
-    \Void                                                                                                         \\
+\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto \Void \\
 \F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
     \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
-\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto       \\
-\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                     \\
-\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{xs}}{}\: \pair{\V{g}}{\V{ys}}{} \mapsto   \\
-\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\{ \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
+\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto
+    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}
 }\]
@@
 data-type will now come equipped with this operation, for free.
-
-The astute reader will have been struck by the type of $\F{replace}$:
-it is \emph{almost} the morphism part -- sometimes called \emph{map}
--- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
-the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
-is the dependent version of the map, which uses the induction
-hypotheses. For space reason, we will not present the non-dependent
-map. It can be found in the Agda model.
 
@@
 unchanged. Unfolding the interpretation of this definition, we
-convince ourselves that this corresponds to the functor
+convince ourself that this corresponds to the functor
 $\FMFreeMonadD$. The fix-point operation ties the knot and gives us
@@
 
-With the free monad construction, we have seen two kind of generic
+With the free monad construction, we have seen two kinds of generic
 operations. Firstly, we have derived a new data-structure from another

### 2010-03-30 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: fix typos in Section 4, 5.4, 5.5, 6, and 7

@@
 universe stratification is orthogonal to our work. Nonetheless, our
-universe of description rather naturally leads itself to
+universes of description rather naturally leads itself to
 stratification. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of type
@@
 
-Consequently, we can implement the operations and examples developed
-above. We refer the reader to our Agda implementation, which take
-advantage of set polymorphism to implement the universe of indexed
-descriptions at any level. Further, we have coded $\IDesc{\!}$ in
-itself and have proved the isomorphism between the host and the
-embedded universes.
+The operations and examples developed in this paper stratify just as
+well. We refer the reader to our Agda model, which takes advantage of
+set polymorphism to implement the universe of indexed descriptions at
+any level. Further, we have coded $\IDesc{\!}$ in itself and have
+proved the isomorphism between the host and the embedded universes.
 
@@
 
-Another generic programming framework is Scrap Your
-Boilerplate~\cite{spj:syb} (SYB). Our proposal is different in various
-ways. The corner stone of SYB is the \emph{spine} view of data-type
-constructors. A data-type is a spine composed by a constructor applied
-to some arguments. Further, this spine is equipped with some
-combinators including, primarily, an iterator. In this setting,
-generic programs are written by composing these combinators. This
-relies on a $\CN{Typeable}$ type-class, allowing dynamic dispatch to
-data-type specific operations. As a result, SYB is not reflexive: it
-is restricted to data-types instanciating $\CN{Typeable}$. Moreover,
-it forbids type-indexed data types: it can only define generic
-functions.
+Interestingly, our generic induction principle allows a programming
+style reminiscent to the Scrap Your Boilerplate~\cite{spj:syb} (SYB)
+approach. Like SYB, we provide a generic operator to manipulate and
+carry operations over data-types. However, unlike SYB, we do not need
+support from the compiler, as required by the $\CN{Typeable}$
+class. Indeed, in our setting, data-types are reflected in the
+language, hence our ability to manipulate them directly.
 
@@
 The interest in generic programming for dependent types is not new
-either. Norell~\cite{norell:msc-thesis} have shown the benefit of
+either. Norell~\cite{norell:msc-thesis} have given a formalization of
 polytypic programming in the setting of Alfa, a precursor of

### 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix typos in Section 4, 5.4, 5.5, 6, and 7

@@
 induction. In the case of vectors, for instance, we have to define the
-whole \emph{family} of vector in one go: vectors of all sizes need to
+whole \emph{family} of vector in one go: vectors of all size needs to
 be defined at the same time. Where simply-typed languages have
-inductive types, in dependently typed programming the basic grammar of
+inductive types, in dependently typed programming we the basic grammar of
 data-types should be that of inductive families of types. To this end, we
@@
 
-Before moving on to indexed descriptions, we have to carry out some
+Before moving on indexed descriptions, we have to carry out some
 maintenance work on descriptions. We presented $\Desc$ as the grammar
@@
 \[\stk{
-\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \:\ldots \\
+\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
+\;\;\ldots \\
 \\
-\idescop{\_}{}{} : _{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
+\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
 \ldots
@@
 
-Note that $\V{I} \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
+Note that $I \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
 endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
-$\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
-\To \Set^{\V{I}}$. Fix-point of inductive families is defined over
+$\V{I} \To (\V{I} \To \Set) \To \Set$, which is isomorphic to $(\V{I} \To \Set)
+\To (\V{I} \To \Set)$. Fix-point of inductive families is defined over
 these indexed functors, hence computing a fix-point of the entire
@@
 \[\stkl{
-\Rule{\Gamma \vdash \Bhab{\V{I}}{\Set} \qquad
-      \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}}}
-     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\!}}{\V{I}\To\Set}} \qquad
-\\:\
-\Rule{\begin{array}{l@{\qquad}l}
-          \Gamma \vdash \Bhab{\V{I}}{\Set} &
+\Rule{\stk{ 
+          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
           \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
-          \Gamma \vdash \Bhab{\V{i}}{\V{I}} &
-          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\:\V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{\!})}}
-      \end{array}}
+          \Gamma \vdash \Bhab{\V{i}}{\V{I}}}}
+     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\V{i}}}{\Set}} \qquad
+\Rule{\stk{
+          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
+          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
+          \Gamma \vdash \Bhab{\V{i}}{\V{I}} \\
+          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{})}}}}
      {\Gamma \vdash \Bhab{\Con{\V{x}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}}
@@
 \stk{
-\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \\
+\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
@@
 here are rather special: they are \emph{equations} constraining
-variables to certain terms. \citet{goguen:pattern-matching} have
-shown, in the context of dependent pattern matching, how to tame these
-equations.
+variables to certain terms. Goguen et
+al.~\cite{goguen:pattern-matching} have shown, in the context of
+dependent pattern matching, how to tame these equations.
 
@@
 
-Further, we can actually remove these equations altogether. Let us
-look back at $\D{Vec}$. We note that the equations are introduced
-because we are \emph{storing} the index of the inductive family.
-However, \emph{inductive families need not store their
-  indices}~\cite{brady:index-inductive-families}.  By examining the
-incoming index, we can apply the \emph{forcing} and \emph{de-tagging}
-optimisations to our initial definition of $\D{Vec}$ gives the
-following, equivalent definition:
+Further, we can actually remove these equations altogether. Let us look back
+at $\D{Vec}$. We note that the equations are introduced because we are
+\emph{storing} the index of the inductive family.  However, \emph{inductive
+families need not store their indices}~\cite{brady:index-inductive-families}.
+By examining the in coming index, we can apply the \emph{forcing} and
+\emph{de-tagging} optimisations to our initial definition of $\D{Vec}$ gives
+the following, equivalent definition:
 
@@
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
-    \FinSuc{\!}   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
+    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
+    \FinSuc{\!}_{\V{n}}   & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
 \end{array}
@@
 constructors. The index-depend part simply indexes the choice of
-constructors by $\V{I}$. Hence, by inspecting the index, it is
-possible to enable or disable the constructors.
+constructors by $\V{I}$. Hence, by inspecting the index, it is possible to
+e
 
@@
 %% Index mapper (terminology?)
-\newcommand{\Val}[1]{\F{Val}~#1}
-\newcommand{\Var}[2]{\F{Var}_{#1}~#2}
+\newcommand{\Val}[1]{\D{Val}~#1}
+\newcommand{\Var}[2]{\D{Var}_{#1}~#2}
 
@@
 \newcommand{\HExprID}{\C{ExprID}}
-\newcommand{\HExprVarD}[1]{\C{ExprD}_{\F{Var},#1}}
+\newcommand{\HExprVarD}[1]{\C{ExprD}_{\C{Var},#1}}
 \newcommand{\HExprFreeD}{\C{ExprD}^{\C{Free}}}
-\newcommand{\HExprAFreeD}{\C{ExprAD}^{\C{Free}}}
-
-So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
-not using any indexing. We are going to define a syntax for a small
+\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}^{\C{Free}}}
+
+We are going to define a syntax for a small
 typed language. We consider two types, natural numbers and booleans:
@@
 respectively named $\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$
-as follows:
+as follow:
 
@@
 resulting description is indexed by $\Ty$. We observe that some
-constructors are ``polymorphic'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
+constructors are ``polymorph'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
 other hand, the $\Eplus$ and $\Ele$ constructors are
@@
                  \begin{array}{l}
-                   {[\Eval{}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
+                   \EnumT{[\Eval{}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
                    \LAM{\V{ty}}
@@
                  \begin{array}{l}
-                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To {[\Eplus{\!}{\!}\!\,]} \\ \Ebool \To {[\Ele{\!}{\!}\!\,]}} \red{,} \\
+                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To \EnumT{[\Eplus{\!}{\!}\!\,]} \\ \Ebool \To \EnumT{[\Ele{\!}{\!}\!\,]}} \red{,} \\
                    \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
@@
 
-Where we overload the notation $\toIDesc{\V{TID}}$ to denote the
-indexed description computed from the tagged indexed description
-$\V{TID}$.
-
 The type of $\F{eval}$ is strikingly similar to a
@@
   \F{cataI}!} gives, for free, the full evaluator. The implementation
-is as follows:
-
-\[\stk{
-\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\:\V{ty}}{\Ty}{\Val{\!}}
+is as follow:
+
+\[\stk{
+\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\: \V{ty}}{\Ty}{\V{ty}}
 \To {\Val{\V{ty}}} \\
@@
 \emph{closed} terms. By abstracting over $\Val{}$, it is possible to
-build and manipulate \emph{open} terms, that is, terms with symbolic
+build and manipulate \emph{open} terms, that is terms with symbolic
 variables. On the model of $\Val{}$, we define $\Var{}{}$:
@@
 constants. For readability, we will abbreviate $\LAM{\V{ty}}
-\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ into $\SUM{\Val{\!}}{\Var{\V{dom}}{\!}}$. \note{Is
+\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
   it polite and comprehensible to ask that?} This defines a new
@@
 \begin{array}{@{}ll}
-\substH  : & \PITEL{\V{dom}}{\EnumU} \\
+\substH  : & \PITEL{\V{\dom}}{\EnumU} \\
            & \PITEL{\V{\gammaNat}}{\GammaSpi{\V{dom}}{\Enat}} 
@@
 syntax with variables: our data-type enjoys more structure than what
-we are given. In particular, this is reminiscent of a free monad,
+we are given. In particular, this is reminiscent to a free monad,
 where $\Eval{}$ is the \return\ introducing variables. The substitution
@@
 
-Instead of implementing $\substH$ in this special case, we shall
-implement it in a generic setting, and express our example in that
-setting. However, before entering the realm of generic structures, let
-us further explore the universe of indexed descriptions.
-
-\subsection{Constrained constructors}
-
-\begin{wstructure}
-!!! Need Help !!!
-<- Assuming a suitable notion of definitional equality for I
-    ???
-\end{wstructure}
-
-In the following, we ask $I$ to come equipped with a notion of
-definitional equality. Hence, we can legibly equates elements of $I$.
-
-\note{ I will need some tutorial on this ad-hoc notion of definitional
-  equality. Or what Conor's meant with that. }
-
-\begin{wstructure}
-<- Fin [figure]
-    <- Presentation with equality constraints
-    <- Example of constrained constructor
-        -> Constraints are translated into equalities
-    ???
-\end{wstructure}
-
-Another common dependent data-type is $\Fin{\!}$, the finite sets of
-numbers. $\Fin{\!}$ is a typical example of an inductive family using
-\emph{constrained constructors}. The constructors $\FinZero$ and
-$\FinSuc{\!}$ are only defined for an index distinct from $\NatZero$:
-
-\[
-\stk{
-\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
-    \FinSuc{\!}_{\V{n}}   & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
-\end{array}
-}
-\]
-
-One way to code constrained data-type is to appeal to equality. The
-constraints are therefore captured by equations in the data-type. In
-this case, we obtain the following definition:
-
-\[\stk{
-\FinD : \Nat \To \IDesc{\Nat} \\
-\begin{array}{@{}ll}
-\FinD\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
-                        & \bigRedBracket{
-                          \begin{array}{l}
-                            \DSigma{\Nat}{\LAM{\V{m}} \DConst{(\V{n} \PropEq \NatSuc{\V{m}})}} \\
-                            \DSigma{\Nat}{\LAM{\V{m}} \DProd{\DVar{\V{m}}}{\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}
-                          \end{array}
-                          }
-\end{array}
-}\]
-
-Intuitively, the constraint is captured as follows. First, we store an
-element $\V{m}$ of $\Nat$, thanks to a $\DSigma{\!}{\!}$. However, the
-constraint stipulates that $\V{m}$ cannot be \emph{any} natural
-numbers: it must be ``the index minus one''. This translates into the
-constraint $\V{n} \PropEq \NatSuc{\V{m}}$.
-
-\note{ Anything else? }
-
-\begin{wstructure}
-<- GADT-style data-type definition
-    <- We directly support it
-        <- Constraints turned into equalities
-            <- Fin example
-    <- Again, smooth transition from "today" data-types to advanced ones
-        <- Same grammar
-        /> More power
-    /> Equations in data-types are not a blessing
-        <- Complexify induction principle
-        <- Lose match between data-type definition and the elimination form
-\end{wstructure}
-
-The $\Fin{\!}$ example is also a well-known example of Generalised
-Abstract Data-Type (GADT) in Haskell. It comes as no surprise that our
-universe of indexed description supports this restricted class of
-inductive families. As we have seen with $\Fin{\!}$, constrained
-constructors are straightforwardly translated into equations.
-
-However, one could, quite rightfully, be worried by the introduction
-of equalities in the picture. In a programming context, equalities
-haunt us, begging for proofs. Hence, whereas the developer typed a
-constructor-constrained data-type, the elimination form will involve
-equalities and their proofs. As much as possible, one would rather
-avoid suffering from this burden. However, the equalities introduced
-here are rather special: they are \emph{equations} constraining
-variables to certain terms. Goguen et
-al.~\cite{goguen:pattern-matching} have shown, in the context of
-dependent pattern matching, how to tame these equations.
-
-\begin{wstructure}
-!!! Need Help !!!
-<- Brady optimisation: forcing
-    <- Source to source translation
-    <- Able to remove some constraints
-    -> Example: Fin [figure]
-    ??? More technical detail needed
-\end{wstructure}
-
-Further, we can actually remove these equations altogether. Let us
-look back at $\Fin{\!}$. We note that the equations are introduced because
-we are \emph{storing} the index of the inductive family, through
-$\V{m}$. Hence the constraints. However, \emph{inductive families need not
-  store their indices}~\cite{brady:index-inductive-families}. Applying
-the \emph{forcing} optimisation to our initial definition of $\Fin{\!}$
-gives the following, equivalent definition:
-
-\[\stk{
-\FinD : \Nat \To \IDesc{\Nat} \\
-\begin{array}{@{}llll}
-\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
-\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
-                     &   &             & \bigRedBracket{
-                                         \begin{array}{l}
-                                         \DConst{\Unit} \\
-                                         \DVar{\V{n}}
-                                         \end{array}
-                                         }
-\end{array}
-}\]
-
-Equations have simply disappeared. Note that it is straightforward to
-turn this definition into a tagged form. We should precise that
-forcing a description is not guaranteed to remove all constraints. It
-is subject to future work to see if constraints can be entirely
-eradicated, or presented more conveniently to the developer. Finally,
-it is worth mentioning that the forcing optimisation is a
-\emph{source-to-source} transformation of the description.
+Instead of implementing $\substH$ in this special case, we can
+implement it in a generic setting, a \emph{free indexed-monad construction}
+that we will see in the next section.
 
@@
 In this section, we develop more structure for our indexed
-data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
-a free monad operation on simple descriptions. Building on this
-experience, we are going to present its equivalent in the indexed
-world. The process the same. Namely, given an indexed functor, we
-derive the indexed functor coding its free monad:
-\note{pwm: Whoa there. Maybe we should say something about IMonads in
-general before we get to this point?}
+data-types. In Section~\ref{sec:desc-free-monad}, we have built a free
+monad operation for simple descriptions. Building on this experience,
+we are going to present its equivalent in the indexed world. The
+process the same. Namely, given an indexed functor, we derive the
+indexed functor coding its free monad: \note{pwm: Whoa there. Maybe we
+  should say something about IMonads in general before we get to this
+  point?}
 
@@
 \end{array} \\
-\FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto \\ \;\;
+\FreeIMonad{(\V{E},\V{F})}{\V{I}}{\V{R}} \mapsto \\ \;\;
     \pair{\pair{\ListCons{\DVar{\!}}{(\fst{\V{E}})}} 
@@
 identified the variable substitution operation as the \bind\ of a free
-monad. The definition of $\substI$ above confirms our intuition: using
+monad. The definition of $\substI$ above confirm our intuition: using
 $\substI$, we should easily obtain $\substH$. To do so, we first have
@@
 a result, the definition is similar to $\HExprD$ presented in
-Figure~\ref{fig:hexpr-full}, replacing $\HExprAD$ with $\HExprAFreeD$:
+Figure~\ref{fig:hexpr-full}, at the exception of $\HExprAD$ that is
+replaced by $\HExprAFreeD$:
 
@@
 corresponds to the syntax of closed terms, $\HExprD$. Similarly,
-$\FreeIMonad{\HExprFreeD}{\Ty}{\SUM{\Val{\!}}{\Var{\V{dom}}{\!}}}$
+$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{\V{ty}} \SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})}$$
 corresponds to expressions with variables, $\HExprVarD{\V{dom}}$.
@@
 
-The evaluator for closed terms we implemented in
+The evaluator for closed term we implemented in
 Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
@@
        \substI\: & \Ty\: 
-                  (\SUM{\Val{\!}}{\Var{\V{dom}}{\!}})\: 
+                  (\SUM{\Val{}}{\Var{\V{dom}}{}})\: 
                   \Val{}\:
@@
 
-\paragraph{Indexed descriptions:}
+\paragraph{Indexed description:}
 
@@
 \[\stk{
-\IDescFreeD : \ATagIDesc{\Unit} \\
+\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
 \begin{array}{@{}ll}
-\IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
-                                [\DConst{\!},
+\IDescFreeD~\V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[}
+                                              \DConst{\!},
                                               \DProd{\!}{\!},
                                               \DSigma{\!}{\!}, 
-                                 \DPi{\!}{\!}] \red{,}\\
-                                  \LAM{\_}\bigRedBracket{\begin{array}{l}
+                                              \DPi{\!}{\!}
+                                                    \red{]}) \\
+                                  & \bigRedBracket{\begin{array}{l}
                                         \DConst{\Set}               \\
@@
                                         \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
-                                    \end{array}}\end{array}}
+                                    \end{array}}
 \end{array}
@@
 \[\stk{
-\IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
-\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}[]\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
-}\]
+\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
+\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{]}}{\Unit}{\V{I}}\: \V{I}
+}\]

### 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Structural changes to IDesc examples

@@
 In this section, we develop more structure for our indexed
-data-types. In Section~\ref{sec:desc-free-monad}, we have built a free
-monad operation for simple descriptions. Building on this experience,
-we are going to present its equivalent in the indexed world. The
-process the same. Namely, given an indexed functor, we derive the
-indexed functor coding its free monad: \note{pwm: Whoa there. Maybe we
-  should say something about IMonads in general before we get to this
-  point?}
+data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
+a free monad operation on simple descriptions. Building on this
+experience, we are going to present its equivalent in the indexed
+world. The process the same. Namely, given an indexed functor, we
+derive the indexed functor coding its free monad:
+\note{pwm: Whoa there. Maybe we should say something about IMonads in
+general before we get to this point?}
 
@@
 Just as in the universe of descriptions, this construction comes with
-an obvious \return\ and a substitution operation, the \bind. Its type
+an obvious \return and a substitution operation, the \bind. Its type
 signature is the following:
@@
 datatypes $\HExprD$ and $\HExprVarD{\V{dom}}$ enjoy a monadic structure. We had
-identified the variable substitution operation as the \bind\ of a free
+identified the variable substitution operation as the \bind of a free
 monad. The definition of $\substI$ above confirm our intuition: using
@@
 
-As previously mentioned, we identify $\Eval{}$ as the \return\ of the
+As previously mentioned, we identify $\Eval{}$ as the \return of the
 free monad, while the other components are the action of the monad. As
@@
 terms. We started with a well-typed language of arithmetical
-expressions. We have seen how to take advantage of indexes to define
-type-safe constructors. Then, we have implemented an evaluator for
-closed term, based on the generic catamorphism function. Having built
-an open term representation, we wanted a substitution operation, in
-order to close open terms in a context. Again, we have implemented
-this operation with the generic substitution operator. Hence, without
-much efforts, we have described the syntax of a well-typed language,
-together with its semantics.
+expressions. We have seen how to take advantage of indexes to only
+present type-safe constructors. Then, we have implemented an evaluator
+for closed term, based on the generic catamorphism function. Having
+implemented an open term representation, we wanted a substitution
+operation, in order to close open terms in a context. Again, we have
+implemented this operation with the generic substitution
+operator. Hence, without much efforts, we have described the syntax of
+a well-typed language, together with its semantics.
 

### 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: merge typos fix from Pedro

--- 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
@@
 We shall now fulfil our promises for $\Desc$. We are going to
-implement the codes of the universe of descriptions. Interestingly, in
-and by itself, the system of codes, $\Desc$, is nothing but a data-type. We
-are in the same situation than with $\EnumU$: we ought to be able to describe
+implement the codes of the universe of descriptions. In and by itself,
+the system of codes, $\Desc$, is nothing but a data-type. We are in
+the same situation as with $\EnumU$: we ought to be able to describe
 the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
@@
 constructor. Constructors correspond to the code's constructors:
-$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice that the
-tagged notation we have used so far for $\Desc$ constructors now fully
-make sense: these were actually the tags we are defining. \note{This
-  sentence badly needs a rephrasing.}
+$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
+that the tagged notation we have used for the $\Desc$ constructors now
+fully makes sense: these were actually the tags we are
+defining.
 
@@
 Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
-remains unchanged, as expected. We successfully describe the
-$\DSigma{\!}{\!}$ case, by a simple appeal to the higher-order
-induction on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing
-the $\V{H}$ in $\Set$ with a recursive argument.
+remain unchanged, as expected. We successfully describe the
+$\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
+on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing the $\V{H}$
+in $\Set$ with a recursive argument.
 

### 2010-03-30 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: merge typos fix from Pedro

--- 2010-03-30 papers/icfp-2010-desc/paper_discussion.tex
@@
 The interest in generic programming for dependent types is not new
-either. Norell~\cite{norell:msc-thesis} have shown the benefit of
+either. \citet{norell:msc-thesis} has given a formalization of
 polytypic programming in the setting of Alfa, a precursor of

### 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: merge typos fix from Pedro

--- 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
@@
 
-So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
-not using any indexing. We are going to define a syntax for a small
-typed language. We consider two types, natural numbers and booleans:
+We are going to define a syntax for a small typed language. We
+consider two types, natural numbers and booleans:
 
@@
 In this section, we develop more structure for our indexed
-data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
-a free monad operation on simple descriptions. Building on this
-experience, we are going to present its equivalent in the indexed
-world. The process the same. Namely, given an indexed functor, we
-derive the indexed functor coding its free monad:
-\note{pwm: Whoa there. Maybe we should say something about IMonads in
-general before we get to this point?}
+data-types. In Section~\ref{sec:desc-free-monad}, we have built a free
+monad operation for simple descriptions. Building on this experience,
+we are going to present its equivalent in the indexed world. The
+process is the same. Namely, given an indexed functor, we derive the
+indexed functor coding its free monad: \note{pwm: Whoa there. Maybe we
+  should say something about IMonads in general before we get to this
+  point?}
 

### 2010-03-30 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: use \citet instead of re-writing stuffs.

--- 2010-03-30 papers/icfp-2010-desc/paper_discussion.tex
@@
 
-Generic programming is a vast topic. We refer our reader to Garcia et
-al.~\cite{garcia:generic-comparative-study} for a broad overview of
+Generic programming is a vast topic. We refer our reader to
+\citet{garcia:generic-comparative-study} for a broad overview of
 generic support in various languages. In the sole context of Haskell,
 there is a myriad of proposals. These approaches are presented and
-compared in Hinze et al.~\cite{hinze:generic-approach-comparative} and
-Rodriguez et al.~\cite{rodriguez:generic-libs-comparative}.
+compared in \citet{hinze:generic-approach-comparative} and
+\citet{rodriguez:generic-libs-comparative}.
 
@@
 polytypic programming in the setting of Alfa, a precursor of
-Agda. Similarly, Verbruggen et al.~\cite{verbruggen:polytype-prog-coq,
+Agda. Similarly, \citet{verbruggen:polytype-prog-coq,
   verbruggen:polytype-coq} have developed a framework for polytypic
@@
 
-Closer to us is the work by Benke et
-al.~\cite{benke:universe-generic-prog}. This seminal work introduced
-the usage of universes for developing generic programs. Our own
-universes are rather similar to theirs: our universe of descriptions
-is similar to their universe of iterated induction, and our universe
-of indexed descriptions is isomorphic to their universe of finitary
-indexed induction. This is not surprising, as we share the same source
-of inspiration, namely induction recursion.
+Closer to us is the work by \citet{benke:universe-generic-prog}. This
+seminal work introduced the usage of universes for developing generic
+programs. Our own universes are rather similar to theirs: our universe
+of descriptions is similar to their universe of iterated induction,
+and our universe of indexed descriptions is isomorphic to their
+universe of finitary indexed induction. This is not surprising, as we
+share the same source of inspiration, namely induction recursion.
 

### 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: use \citet instead of re-writing stuffs.
Desc@ICFP: update TODO. Fix a typo.

--- 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
@@
 here are rather special: they are \emph{equations} constraining
-variables to certain terms. Goguen et
-al.~\cite{goguen:pattern-matching} have shown, in the context of
-dependent pattern matching, how to tame these equations.
+variables to certain terms. \citet{goguen:pattern-matching} have
+shown, in the context of dependent pattern matching, how to tame these
+equations.
 
@@
 
-Further, we can actually remove these equations altogether. Let us look back
-at $\D{Vec}$. We note that the equations are introduced because we are
-\emph{storing} the index of the inductive family.  However, \emph{inductive
-families need not store their indices}~\cite{brady:index-inductive-families}.
-By examining the in coming index, we can apply the \emph{forcing} and
-\emph{de-tagging} optimisations to our initial definition of $\D{Vec}$ gives
-the following, equivalent definition:
+Further, we can actually remove these equations altogether. Let us
+look back at $\D{Vec}$. We note that the equations are introduced
+because we are \emph{storing} the index of the inductive family.
+However, \emph{inductive families need not store their
+  indices}~\cite{brady:index-inductive-families}.  By examining the
+incoming index, we can apply the \emph{forcing} and \emph{de-tagging}
+optimisations to our initial definition of $\D{Vec}$ gives the
+following, equivalent definition:
 

### 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: say something about map vs. replace.

--- 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
@@
 
+The astute reader will have been struck by the type of $\F{replace}$:
+it is \emph{almost} the morphism part -- sometimes called \emph{map}
+-- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
+the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
+is the dependent version of the map, which uses the induction
+hypotheses. For space reason, we will not present the non-dependent
+map. It can be found in the Agda model.
+
 \begin{wstructure}

### 2010-03-31 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: write All and all using the first definition of Desc.
Desc@ICFP: fix elaboration of constructors in Mu

@@
 data-types through raw $\Sigma$-types is not a viable option. Indeed,
-just as for finite sets, encoding throws away information when we
-crucially need it. In the realm of data-types, things become even
+just as for finite sets, encoding throw away information when we
+crucially need it. In the realm of data-types, things become even more
 tougher, as one could legitimately expect an induction principle, for
@@
 defining a set of codes, we somehow define the syntax of a
-language. However, as such, a system of codes is useless as it lacks a
+language. However, as such, a system of code is useless as it lacks a
 semantics. Instead of equipping the universe of codes with some
@@
 build these descriptions. However, it should be clear that, in
-practice, these definitions can be automatically constructed from a
+practice, these definitions can be automatically constructed from an
 Haskell-like $\data$ definition.
@@
 $\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
-particular, we have considered its codes, $\DSigma{\!}{\!}$,
+particular, we have considered its code, $\DSigma{\!}{\!}$,
 $\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
@@
 lists, and binary trees. However, the class of functors expressible
-with descriptions enjoys another property: they all admit a
+with descriptions enjoy another property: they all admit a
 fix-point. Indeed, descriptions are interpreted into \emph{polynomial
@@
 Doing so, we seal data-type definition behind a convenient
-abstraction. We take advantage of this abstraction barrier to address
+abstraction. We take advantage of this abstraction barrier to tamper
 the second issue. Namely, the developer would rather be presented a
-name than a fix-point of a pattern functor. Surprisingly, it is
+name rather than a fix-point of a pattern functor. Surprisingly, it is
 sufficient to \emph{label} the $\Mu{}$ constructor with the
@@
 
-Regaining the ability to write data-type constructors is simply a
-matter of type propagation. We therefore extend the type-checking
+Regaining the ability to write data-type constructor is simply a
+matter of type propagation. We therefore extends the type-checking
 framework of Figure~\ref{fig:type-checking} with the following
@@
 Hence, in the high-level expression language, the developer writes a
-familiar type constructor, applied to a telescope of arguments. In
+familiar type constructors, applied to a telescope of arguments. In
 the type propagation setting, this expression will be pushed to a

### 2010-03-31 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: fix Generic Haskell related work.

@@
 
-Our approach also support the Generic
-Haskell~\cite{hinze:generic-haskell} model. This model, based on
-type-indexed data types, enables computing new data-types from
-others. This is natural in our system, as data-types descriptions are
-first-class.
+We share with Generic Haskell the \emph{type-indexed data type}
+approach~\cite{hinze:generic-haskell}, as exemplified by the free
+monad construction: from data-type, we can compute new data-types and
+equip them with their structure. Generic Haskell also features
+\emph{generic views}~\cite{holdermans:generic-view}, transparently
+transforming the structure of data-type definitions, to ease the
+implementation of generic functions over them. Views in
+dependently-typed system are widely
+used~\cite{mcbride.mckinna:view-from-the-left}. Unlike in Generic
+Haskell, we do not have to modify the compiler to develop views on
+data-types: we can massage description codes from inside the language
+and present them as a standard view. An example of such operation is
+the tagged descriptions, presenting data-types under a particular
+angle.
+
+Unlike Generic Haskell, we do not support polykinded
+programming~\cite{hinze:polytypic-polykinded}. Our descriptions are
+limited to endofunctor on $\Set$ and $\Set^I$. While it is possible to
+\emph{encode} higher-kinded data-types in our setting, we do not plan
+to adopt this strategy. Rather, as future work, we plan to extend our
+universe to capture higher-kinded definitions and generic functions
+over them. For the same reason, arity-generic
+programming~\cite{weirich:arity-generic} is out of reach of our
+current presentation.
 

### 2010-03-31 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: More bug fixes in the code

--- 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
@@
 
-We are going to define a syntax for a small typed language. We
-consider two types, natural numbers and booleans:
+We are going to define a syntax for a small
+typed language. We consider two types, natural numbers and booleans:
 
@@
   \F{cataI}!} gives, for free, the full evaluator. The implementation
-is as follows:
+is as follow:
 
@@
 we are going to present its equivalent in the indexed world. The
-process is the same. Namely, given an indexed functor, we derive the
+process the same. Namely, given an indexed functor, we derive the
 indexed functor coding its free monad: \note{pwm: Whoa there. Maybe we

### 2010-03-31 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: clarify the situation of SYB.

@@
 
-Interestingly, our generic induction principle allows a programming
-style reminiscent to the Scrap Your Boilerplate~\cite{spj:syb} (SYB)
-approach. Like SYB, we provide a generic operator to manipulate and
-carry operations over data-types. However, unlike SYB, we do not need
-support from the compiler, as required by the $\CN{Typeable}$
-class. Indeed, in our setting, data-types are reflected in the
-language, hence our ability to manipulate them directly.
+Another generic programming framework is Scrap Your
+Boilerplate~\cite{spj:syb} (SYB). Our proposal is different in various
+ways. The corner stone of SYB is the \emph{spine} view of data-type
+constructors. A data-type is a spine composed by a constructor applied
+to some arguments. Further, this spine is equipped with some
+combinators including, primarily, an iterator. In this setting,
+generic programs are written by composing these combinators. This
+relies on a $\CN{Typeable}$ type-class, allowing dynamic dispatch to
+data-type specific operations. As a result, SYB is not reflexive: it
+is restricted to data-types instanciating $\CN{Typeable}$. Moreover,
+it forbids type-indexed data types: it can only define generic
+functions.
 

### 2010-03-31 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: merge typos fixes

@@
 The interest in generic programming for dependent types is not new
-either. Norell~\cite{norell:msc-thesis} have given a formalization of
+either. \citet{norell:msc-thesis} has given a formalization of
 polytypic programming in the setting of Alfa, a precursor of

### 2010-03-31 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: explain where \overbar comes from
Desc@ICFP: demote \subsubsection{Indexed Tagged Description}
Desc@ICFP: introduce inductionI and AllI for indexed descriptions
Desc@ICFP: clean-up \note{}s

--- 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
@@
      {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\!}}{\V{I}\To\Set}} \qquad
-\\:\
+\\
 \Rule{\begin{array}{l@{\qquad}l}
@@
 Figure~\ref{fig:idesc}.
+
+The induction scheme on $\IDesc{\!}$ has the following type:
+
+\[
+\begin{array}{@{}ll}
+\F{inductionI} : & _{\Bhab{\V{I}}{\Set}}
+                   \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
+                   \PI{\V{P}}{\SIGMA{\V{i}}{\V{I}} \IMu{\V{I}}{\V{R}}{\!} \To \Set} \\
+                 & (      \PITEL{\V{i}}{\V{I}} 
+                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{\IMu{\V{I}}{\V{R}}{\!}}} \\
+                 & \   \idescop{\AllI{\V{I}}
+                                     {(\V{R}\: \V{i})}
+                                     {(\IMu{\V{I}}{\V{R}}{\!})}
+                                     {\V{xs}}}
+                               {}
+                               {\V{P}} \To
+                       \V{P}\: \pair{\V{i}}{\Con{\V{xs}}}{}) \\
+                 & \PITEL{\V{i}}{\V{I}}
+                   \PI{\V{x}}{\IMu{\V{I}}{\V{R}}{\V{i}}}
+                   \V{P}\: \pair{\V{i}}{\V{x}}{}
+\end{array}
+\]
+
+Where the operator $\AllI{\!}{\!}{\!}{\!}$ is of type:
+
+\[
+\AllI{\!}{\!}{\!}{\!} : _{\Bhab{\V{I}}{\Set}}
+                        \PITEL{\V{D}}{\IDesc{\V{I}}}
+                        \PI{\V{P}}{\V{I} \To \Set}
+                        \idescop{\V{D}}{\V{I}}{\V{P}} \To
+                        \IDesc{(\SIGMA{\V{i}}{\V{I}} \V{P})} 
+\]
+
+The implementation of these operators can be found in the companion
+technical report. As for descriptions, we can compute a generic
+catamorphism, $\F{cataI}$, from $\F{inductionI}$.
 
@@
 
-\note{Should we say something about IMu, fixed-points, and their
-  elimination before moving on? If so, where should we put it?}
-
 \subsection{Examples}
@@
 
-\subsubsection{Tagged Indexed Descriptions}
+\paragraph{Tagged Indexed Descriptions:}
 
@@
 catamorphism. Indeed, implementing a single step of evaluation -- the
-algebra -- is sufficient, as $\F{cataI}$ \note{We have not defined
-  \F{cataI}!} gives, for free, the full evaluator. The implementation
-is as follows:
+algebra -- is sufficient, as $\F{cataI}$ gives, for free, the full
+evaluator. The implementation is as follows:
 

### 2010-03-31 papers/icfp-2010-desc/paper_type_theory.tex
the basic type theory overhauled

@@
 
-Our work is deeply rooted in the realm of dependently-typed functional
-languages. In particular, we have developed this technology in our
-on-going work on Epigram. However, in this paper,
-we strive to present our technology in the broadest setting
-possible. To this end, we are going to present a type theory from the
-ground up.
-
-To the reader new to type theory, this first section will provide a
-gentle introduction to dependently-typed systems. To the reader
-familiar with type theory, she will recognise the basis of her
-favourite type theory, be it Agda or Coq, for instance. This scene sets,
-we have a common ground for discussion.
+One challenge in writing this paper is to extricate our account of datatypes from what else is new in Epigram 2.
+In fact, we demand relatively little from the setup, so we shall start with a `vanilla' theory and add just what we need.
+The reader accustomed to dependent types will recognise the basis of her favourite system; for those less familiar, we try to keep the presentation self-contained.
 
@@
 
-In the following, we adopt a standard presentation for our type
-theory. It is defined by three forms of judgements: context validity
-judgements, typing judgements, and equality judgements. Formally,
-these judgements are presented as follows:
+We adopt a traditional presentation for our type
+theory, with three mutually defined systems of judgments:
+\emph{context validity}, \emph{typing}, and \emph{equality},
+with the following forms:
 
@@
 \begin{array}{ll}
-\Gamma \vdash \Valid                & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
-\Gamma \vdash \Bhab{\V{t}}{\V{T}}           & \mbox{term \(\V{t}\) has type \(\V{T}\) in context \(\Gamma\)} \\
-\Gamma \vdash \Bhab{\V{s} \equiv \V{t}}{\V{T}}  & \mbox{\(\V{s}\) and \(\V{t}\) are equal at type \(\V{T}\) in context \(\Gamma\)} \\
+\G \vdash \Valid                & \mbox{\(\G\) is a valid context, giving types to variables} \\
+\G \vdash \Bhab{\M{t}}{\M{T}}           & \mbox{term \(\M{t}\) has type \(\M{T}\) in context \(\G\)} \\
+\G \vdash \Bhab{\M{s} \equiv \M{t}}{\M{T}}  & \mbox{\(\M{s}\) and \(\M{t}\) are equal at type \(\M{T}\) in context \(\G\)} \\
 \end{array}
@@
 
-Each of these judgements is populated by a respective system of
-inferences rules. These rules are formulated so as to ensure that the
-following implications always hold by induction on derivations:
-
+The rules are formulated to ensure that the
+following `sanity checks' hold by induction on derivations
 \[
 \begin{array}{l@{\;\Rightarrow\;\;}l}
-\Gamma            \vdash \Bhab{\V{t}}{\V{T}}            
-    & \Gamma \vdash \Valid\; \wedge\; \Gamma\vdash\Type{\V{T}} \\
-\Gamma            \vdash \V{s} \equiv \Bhab{\V{t}}{\V{T}}   
-    & \Gamma \vdash \Bhab{\V{s}}{\V{T}} \;\wedge\; \Gamma\vdash \Bhab{\V{t}}{\V{T}} \\
-\Gamma ; \xS ; \Delta \vdash \V{J}                      
-    & \Gamma \vdash \Bhab{\V{s}}{\V{S}} \;\Rightarrow\; 
-          \Gamma ; \Delta[\V{s}/\V{x}] \vdash \V{J}[\V{s}/\V{x}] \\
-\end{array}
+\G            \vdash \Bhab{\M{t}}{\M{T}}            
+    & \G \vdash \Valid\; \wedge\; \G\vdash\Type{\M{T}} \\
+\G            \vdash s \equiv \Bhab{\M{t}}{\M{T}}   
+    & \G \vdash \Bhab{s}{T} \;\wedge\; \G\vdash \Bhab{\M{t}}{\M{T}}
+\end{array} \]
+and that judgments \(\M{J}\) are preserved by well-typed instantiation.
+\[
+\G ; \xS ; \Delta \vdash \M{J}            \;\Rightarrow\;      
+    \G \vdash \Bhab{\M{s}}{\M{S}} \;\Rightarrow\; 
+          \G ; \Delta[\M{s}/\x] \vdash \M{J}[\M{s}/\x] 
 \]
 
-We are not going to prove the validity of these invariants. They
-follow rather straightforwardly from the induction rules. For formal
-proofs, we refer the reader to standard presentations of type theory,
-such as Luo's seminal work \cite{luo:utt}.
+%We are not going to prove the validity of these invariants. They
+%follow rather straightforwardly from the induction rules. For formal
+%proofs, we refer the reader to standard presentations of type theory,
+%such as Luo's seminal work \cite{luo:utt}.
 
@@
 
-Noticeably, we have adopted a judgemental presentation of
-equality. Therefore, we are not tied to a particular implementation
-choice. In particular, our system has been modelled in Agda, which
-features an intensional equality. On the other hand, it is used in
-Epigram, whose equality has a slightly extensional
-flavor~\cite{altenkirch:ott}. We expect users of fully extensional
-systems to also find their way through this presentation.
+We specify equality as a judgment, leaving open the details of its implementation, requiring only a congruence including ordinary computation (\(\beta\)-rules), decided, e.g., by testing \(\alpha\)-equivalence of \(\beta\)-normal forms~\cite{DBLP:journals/jfp/Adams06}.
+Coquand and Abel feature prominently in a literature of richer equalities, involving \(\eta\)-expansion, proof-irrelevance and other attractions~\cite{DBLP:journals/scp/Coquand96,DBLP:conf/tlca/AbelCP09}.
+Agda and Epigram 2 support such features, Coq currently does not, but they are surplus to requirements here.
+
+% Therefore, we are not tied to a particular implementation
+%choice. In particular, our system has been modelled in Agda, which
+%features an intensional equality. On the other hand, it is used in
+%Epigram, whose equality has a slightly extensional
+%flavor~\cite{altenkirch:ott}. We expect users of fully extensional
+%systems to also find their way through this presentation.
 
@@
 
-The rules ensuring Context Validity 
-are unsurprisingly common. They simply guarantee the validity of the
-empty context, as well as the linear extension of a context by a
-well-typed variable.
-
+Context validity ensures that variables inhabit well-formed
+sets.
 \[
@@
 %% Extend context
-\Rule{\Gamma       \vdash \Type{\V{S}}}
-     {\Gamma ; \xS \vdash \Valid}\;\V{x}\not\in\Gamma
+\Rule{\G       \vdash \Type{\M{S}}}
+     {\G ; \xS \vdash \Valid}\;\x\not\in\G
 \]
-
+%
 \begin{wstructure}
@@
 \end{wstructure}
-
-The typing judgements (Fig.~\ref{fig:typing-judgements}) are just as
-familiar. We recognise the standard presentation of $\Pi$ and $\Sigma$
-types, respectively inhabited by lambda terms and dependent
-pairs. Naturally, there are rules for function application and
-projections of $\Sigma$-types. Equal types can be substituted, thanks
-to the conversion rule.
-
-For the sake of presentation, we postulate a $\Set$ in $\Set$
-rule. Having this rule makes our type theory inconsistent, by Girard's
-paradox~\cite{girard:set-in-set}. However, it has been
-shown~\cite{harper:implicit-universe, luo:utt,
-  courant:explicit-universe} that a valid stratification can be
-inferred, automatically or semi-automatically. In the remaining of our
-presentation, we will assume that such a stratification exists, even
-though we will keep it implicit. We shall discuss this assumption in
-Section~\ref{sec:discussion}.
-
-\begin{figure}
-
+%
+The basic typing rules
+% (Fig.~\ref{fig:typing-judgements})
+for tuples and functions are also standard, save that we locally adopt
+\(\Set:\Set\), putting presentation before paradox~\cite{girard:set-in-set}.
+The usual remedies apply, \emph{stratifying}
+\(\Set\)~\cite{harper:implicit-universe, luo:utt,
+  courant:explicit-universe}.
 \input{figure_typing_judgements}
 
-\caption{Typing judgements}
-\label{fig:typing-judgements}
+\paragraph{Notation.} We subscript information needed for type synthesis
+but not type checking, e.g., the domain of a \(\LAMBINDER\)-abstraction,
+and suppress it informally where clear. Square brackets denote tuples,
+with a LISP-like right-nesting convention: \(\sqr{a\;b}\) abbreviates
+\(\pair{a}{\pair{b}{\void}{}}{}\).
 
-\end{figure}
+%We recognise the standard presentation of $\Pi$ and $\Sigma$
+%types, respectively inhabited by lambda terms and dependent
+%pairs. Naturally, there are rules for function application and
+%projections of $\Sigma$-types. Equal types can be substituted, thanks
+%to the conversion rule.
+
+%For the sake of presentation, we postulate a $\Set$ in $\Set$
+%rule. Having this rule makes our type theory inconsistent, by Girard's
+%paradox~\cite{girard:set-in-set}. However, it has been
+%shown~\cite{harper:implicit-universe, luo:utt,
+%  courant:explicit-universe} that a valid stratification can be
+%inferred, automatically or semi-automatically. In the remaining of our
+%presentation, we will assume that such a stratification exists, even
+%though we will keep it implicit. We shall discuss this assumption in
+%Section~\ref{sec:discussion}.
+
+%\begin{figure}
+%
+%\input{figure_typing_judgements}
+%
+%\caption{Typing judgements}
+%\label{fig:typing-judgements}
+%
+%\end{figure}
 
@@
 
-Finally, we define the rules governing judgemental equality in
-Figure~\ref{fig:judgemental-equality}. We implicitly assume that
-judgemental equality respects symmetry, reflexivity, and
-transitivity. We capture the computational behavior of the language
-through the $\beta$-rules for function application and pairs. Finally,
-we implicitly assume that it respects purely syntactic and structural
-equality. This includes equality under lambda ($\xi$-rule).
+The judgmental equality comprises these computational rules
+\input{figure_judgemental_equality} closed under reflexivity,
+symmetry, transitivity and structural congruence, even under binders.
+We omit the mundane rules which ensure these closure properties for
+reasons of space. Given a suitable stratification of \(\Set\), the
+computation rules yield a terminating evaluation procedure, ensuring
+the decidability of equality and thence type checking.
 
-Crucially, being judgemental, this presentation is agnostic in the
-notion of equality actually implemented. Indeed, our typing and
-equality judgements do not rely on a ``propositional'' equality. This
-freedom is a key point in favour of the wide applicability of our
-proposal. This judgemental presentation must be read as a
-\emph{specification}: our proposal works with any propositional
-equality satisfying this specification. Moreover, our lightweight
-requirements do not endanger decidability of equality-checking.
-Obviously, when implementing our technology in an existing
-type-theory, some opportunities arise. We will present some of them
-along the course of the paper.
+%Finally, we define the rules governing judgemental equality in
+%Figure~\ref{fig:judgemental-equality}. We implicitly assume that
+%judgemental equality respects symmetry, reflexivity, and
+%transitivity. We capture the computational behavior of the language
+%through the $\beta$-rules for function application and pairs. Finally,
+%we implicitly assume that it respects purely syntactic and structural
+%equality. This includes equality under lambda ($\xi$-rule).
 
-\begin{figure}
+%Crucially, being judgemental, this presentation is agnostic in the
+%notion of equality actually implemented. Indeed, our typing and
+%equality judgements do not rely on a ``propositional'' equality. This
+%freedom is a key point in favour of the wide applicability of our
+%proposal. This judgemental presentation must be read as a
+%\emph{specification}: our proposal works with any propositional
+%equality satisfying this specification. Moreover, our lightweight
+%requirements do not endanger decidability of equality-checking.
+%Obviously, when implementing our technology in an existing
+%type-theory, some opportunities arise. We will present some of them
+%along the course of the paper.
 
-\input{figure_judgemental_equality}
+%\begin{figure}
 
-\caption{Judgemental equality}
-\label{fig:judgemental-equality}
-
-\end{figure}
+%\input{figure_judgemental_equality}
+%
+%\caption{Judgemental equality}
+%\label{fig:judgemental-equality}
+%
+%\end{figure}
 
@@
 
-This completes our presentation of the type theory. Assuming a
-stratified discipline of universe, the system we have described enjoy
-some very strong meta-theoretical properties. Unlike simply typed
-languages, such as Haskell, dependently-typed systems are
-\emph{strongly normalising}: every program that type-checks
-terminates. Moreover, type-checking is decidable and can therefore be
-implemented by a terminating algorithm.
-\note{Need some care here. Expansion would be good too. I wanted to
-  carry the intuition that we are not the bad guys with a
-  non-terminating type-checker.}
+%This completes our presentation of the type theory. Assuming a
+%stratified discipline of universe, the system we have described enjoy
+%some very strong meta-theoretical properties. Unlike simply typed
+%languages, such as Haskell, dependently-typed systems are
+%\emph{strongly normalising}: every program that type-checks
+%terminates. Moreover, type-checking is decidable and can therefore be
+%implemented by a terminating algorithm.
+%\note{Need some care here. Expansion would be good too. I wanted to
+%  carry the intuition that we are not the bad guys with a
+%  non-terminating type-checker.}
 
@@
 \[
-\Rule{\Gamma ; \xS ; \Delta \vdash \Valid}
-     {\Gamma ; \xS ; \Delta \vdash \Bhab{\V{x}}{\V{S}}}
+\Rule{\G ; \xS ; \Delta \vdash \Valid}
+     {\G ; \xS ; \Delta \vdash \Bhab{\V{x}}{\V{S}}}
 \]
@@
 \[
-\Rule{\Gamma       \vdash \Type{\V{S}} \quad
-      \Gamma ; \xS \vdash \Bhab{\V{t}}{\V{T}}}
-     {\Gamma \vdash \Bhab{\PLAM{\x}{\V{S}} \V{t}}{\PIS{\xS} \V{T}}}
+\Rule{\G       \vdash \Type{\V{S}} \quad
+      \G ; \xS \vdash \Bhab{\V{t}}{\V{T}}}
+     {\G \vdash \Bhab{\PLAM{\x}{\V{S}} \V{t}}{\PIS{\xS} \V{T}}}
 \]

### 2010-03-31 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: proofread Universe of inductive types

--- 2010-03-30 papers/icfp-2010-desc/paper_desc.tex
@@
-\section{A universe of simple data-types}
+\section{A universe of inductive datatypes}
 \label{sec:universe-desc}
@@
 \begin{wstructure}
-<- Why starting with simple data-types
+<- Why starting with simple datatypes
     <- For pedagogical purposes
-        <- Data-types as we know them every day
+        <- datatypes as we know them every day
         /> Target dependent types
@@
 In this section, we describe an implementation of inductive types, as
-we know them in simply-typed languages such as Haskell or OCaml. This
-choice is motivated by pedagogical considerations: by describing the
-layman data-types in our setting, we build on our intuition of
-data-types we are familiar with. In particular, we do not aim at
-describing a model for generics in simply-typed functional
-language. Our goal is and remains to discuss data-types as we find
-them in dependently-typed languages.
+we know them in ML-like languages. This choice is motivated by
+pedagogical considerations: by describing the layman datatypes in our
+setting, we build on our intuition of datatypes we are familiar
+with. In particular, we do not aim at describing a model for generics
+in simply-typed functional language. Our goal is and remains to
+discuss datatypes as we find them in dependently-typed languages.
 
@@
 induction-recursion~\cite{dybjer:general-ir, dybjer:axiom-ir,
-  dybjer:ir-initial-algebra, dybjer:iir}. In the context of simple
-data-types, the full power of induction-recursion is not
-necessary. Therefore, we adopt a stripped-down version of it. As we
-extend the scope of our universe of data-types, our presentation will
-diverge from induction-recursion. We shall compare both approaches in
+  dybjer:ir-initial-algebra, dybjer:iir}. For inductive datatypes, the
+full power of induction-recursion is not necessary. Therefore, we
+adopt a stripped-down version of it. As we broaden the scope of our
+universe of datatypes, our presentation will diverge from
+induction-recursion. We shall compare both approaches in
 Section~\ref{sec:discussion}.
@@
 
-In dependently-typed language, $\Sigma$-types can be interpreted in
+In dependently-typed languages, $\Sigma$-types can be interpreted in
 two different ways. This duality is actually reflected in the notation
-we can find in the literature, depending on the sensibility of the author.
-The notation $\blue{\Sigma}_{\V{x} : \V{A}} (\V{B}\: \V{x})$ stresses the interpretation that
-$\Sigma$-types are a generalisation of sums over arbitrary arities.  Where
-simply-typed languages have finite sums, dependently-typed languages have
-sums indexed by any set.
+we can find in the literature, depending on the sensibility of the
+author.  The notation $\blue{\Sigma}_{\V{x} : \V{A}} (\V{B}\: \V{x})$
+stresses the interpretation that $\Sigma$-types are a generalisation
+of sums over arbitrary arities.  Where simply-typed languages have
+finite sums, dependently-typed languages have sums indexed by any set.
 
@@
 where the second component can depend on the first one. When simply-typed
-languages pack data into non-dependent tuples, dependently-types languages
+languages pack data into non-dependent tuples, dependently-typed languages
 tolerate this non-dependent usage. However, they also give the rather novel
@@
 \begin{wstructure}
-<- Data-types in the simply-typed world
+<- datatypes in the simply-typed world
     -> "sums-of-product"
@@
         <- Product of arguments
-<- Data-types in the dependently-typed world
+<- datatypes in the dependently-typed world
     -> "sigmas-of-sigmas"
     /> Need ability to manipulate these sigmas
-        -> Define a Code for data-types
+        -> Define a Code for datatypes
         -> Together with a sigma-based Interpretation
@@
 
-In the simply-typed world, the essence of data-type definition can be
-summed up by the expression \emph{sum-of-product}. A data-type is defined
-by a finite sum -- a choice -- of constructors, each of them is
-composed of a product -- a tuple -- of arguments. This form governs
-the grammar of data-type definition in languages such as Haskell or
-OCaml.
+In the simply-typed world, datatype definitions are often referred to
+as \emph{sum-of-product}. A datatype is defined by a finite sum -- a
+choice -- of constructors, each of them is composed of a product -- a
+tuple -- of arguments. This form governs the grammar of datatype
+definition in languages such as Haskell or ML.
 
-To model these data-types, we simply need to capture this grammar in
+To model these datatypes, we simply need to capture this grammar in
 our dependently-typed setting. Here, the notion of sum-of-product
-naturally translates into \emph{sigmas-of-sigmas}. However, beyond
-this intuition, it should be clear that a direct encoding of
-data-types through raw $\Sigma$-types is not a viable option. Indeed,
-just as for finite sets, encoding throws away information when we
-crucially need it. In the realm of data-types, things become even
-tougher, as one could legitimately expect an induction principle, for
-instance. Anonymous in a sea of $\Sigma$s, our data-types have little
-chance of survival. 
-\note{Ridiculous tentative of poetry. I was really tired when I wrote that.}
+naturally translates into \emph{sigmas-of-sigmas}. Beyond this
+intuition, it should be clear that a direct encoding of datatypes
+through raw $\Sigma$-types is not a viable option. Indeed, just as for
+finite sets, encoding throws away information when we crucially need
+it. Moreover, to program with these datatypes, one would need an
+induction principle. An encoding will be unable to deliver such
+operation~\cite{geuvers:induction-not-derivable}.
 
@@
 
-For the structure of our data-types to be known, we crucially need the
-ability to name them. \note{Pursuing this crappy analogy. Solving the type
-presentation/representation debate would help solving that point.} To this
-end, we will use a standard technique in dependently-typed programming: we
-construct an ad-hoc universe. This technique dates back to Martin-L\"of
-definition of type theory~\cite{martin-lof:itt}. Since then, it has been
-fruitful as a programming 
-technique~\cite{benke:universe-generic-prog, oury:power-of-pi}
-\note{Any idea of other papers?}. 
-We refer the reader to the Agda tutorial~\cite{norell:agda-tutorial} for a
-pedagogical presentation of universe construction. 
-\note{Is that really a motivating motivation for the usage of universe 
-construction?}
-\note{pwm: Surely ($\EnumU$, $\EnumT$) is a useful example of universes too.
-Even if it doesn't show off their full power?}
+While sigmas-of-sigmas form a convenient \emph{semantics} for
+datatypes, we have seen that it makes a poor \emph{syntax}. To
+decouple these two notions, we use a standard technique in
+dependently-typed programming: we construct an ad-hoc universe. This
+technique dates back to Martin-L\"of definition of type
+theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
+programming technique~\cite{benke:universe-generic-prog,
+  oury:power-of-pi} \note{Any idea of other papers?}.  We refer the
+reader to the Agda tutorial~\cite{norell:agda-tutorial} for a
+pedagogical presentation of universe construction.  \note{Is that
+  really a motivating motivation for the usage of universe
+  construction?}  \note{pwm: Surely ($\EnumU$, $\EnumT$) is a useful
+  example of universes too.  Even if it doesn't show off their full
+  power?}
 
@@
 by defining new types. These names are called \emph{codes}. By
-defining a set of codes, we somehow define the syntax of a
-language. However, as such, a system of codes is useless as it lacks a
-semantics. Instead of equipping the universe of codes with some
-computational behaviour, we pragmatically chose to \emph{interpret}
-these codes back into the standard type theory. Hence, codes act only
-as labels, while the type theory provides the computational
-machinery. \note{Is that a good description?}
+defining a set of codes, we define the syntax of a language. However,
+as such, a system of codes is useless as it lacks a semantics. Instead
+of equipping the universe of codes with some computational behaviour,
+we pragmatically chose to \emph{interpret} these codes back into the
+standard type theory. Hence, codes act only as labels, while the type
+theory provides the computational machinery. \note{Is that a good
+  description?}
 
 Codes being simple labels, we have the ability to inspect them, hence
-taking advantage of structural information. Being able to inspect
-them, we are therefore free to compute over them: deriving new codes
-from previous ones, or even new functions on them. In several
-occasions, we will have the opportunity to witness the power of
-universes.
+regaining structural information. Being able to inspect them, we are
+therefore free to compute over them: deriving new codes from previous
+ones, or even new functions on them. In several occasions, we will
+have the opportunity to witness the power of universes.
 
@@
     <- [both figures]: cannot be read separately
-    <- Mimic the standard grammar of data-types description
+    <- Mimic the standard grammar of datatypes description
         -> Just as we already know it
@@
 (Figure~\ref{fig:desc_universe}). As expected, the code of this
-universe mimics the standard grammar of data-types definitions in
+universe mimics the standard grammar of datatypes definitions in
 simply-typed languages. Hence, we have a $\DSigma{\!}{\!}$ code,
@@
 sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
-contains no useful payload. The functoriality of the data-types is
+contains no useful payload. The functoriality of the datatypes is
 introduced by $\DIndx{\!}$. When we tie the knot with a fix-point, the
@@
 recursive argument. We notice that this functoriality appears in the
-type of $\descop{\V{D}}{\!}$ itself, for a given data-type definition
+type of $\descop{\V{D}}{\!}$ itself, for a given datatype definition
 $\V{D}$. This corresponds to the object part of the functor described
@@
 build these descriptions. However, it should be clear that, in
-practice, these definitions can be automatically constructed from a
-Haskell-like $\data$ definition.
+practice, these definitions can be automatically constructed from an
+ML-like $\data$ definition.
 
@@
 Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
-give a choice between the $\NatZero$ and $\NatSuc{\!}$ constructors. In
-the $\NatZero$ case, we reach the end of the description: there is no
-useful payload in that case. In the $\NatSuc{\!}$ case, we left one hole
-open for the recursive argument, and we close the description.
+give a choice between the $\NatZero$ and $\NatSuc{\!}$
+constructors. In the $\NatZero$ case, we reach the end of the
+description: there is no useful payload in that case. In the
+$\NatSuc{\!}$ case, we leave one hole open for the recursive argument,
+and we close the description.
 
@@
     <- Form TDesc = List (UId x Desc) [equation]
-    <- Follow usual sums-of-product presentation of data-type
+    <- Follow usual sums-of-product presentation of datatype
         <- Finite set of constructors
         <- Then whatever you want
-    -> Any Desc data-type can be turned into this form
+    -> Any Desc datatype can be turned into this form
         -> No loss of expressive power
@@
 
-From the examples above, we observe that data-types are defined by a
+From the examples above, we observe that datatypes are defined by a
 $\DSigma{\!}{\!}$ which first argument is a finite set of
@@
 
-So far, for convenience, we have taken as granted the existence of
-$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
-particular, we have considered its codes, $\DSigma{\!}{\!}$,
-$\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
-theory. Hence, in an implementation of this type theory, we need to
-extended it with these constructors and their typing rules. Although
-it makes no conceptual difference, the code of description should
-rather be read as a specification. We promise the existence of such
-objects, satisfying the typing rules. It will be the subject of
+For convenience, we have taken for granted the existence of $\Desc$,
+as presented in Figure~\ref{fig:desc_universe}. In particular, we have
+considered its codes, $\DSigma{\!}{\!}$, $\DIndx{\!}$, and $\DUnit$ as
+type formers, extending the basic type theory. Hence, in an
+implementation of this type theory, we need to extended it with these
+constructors and their typing rules. Although it makes no conceptual
+difference, the code of description should rather be read as a
+specification. We promise the existence of such objects, satisfying
+the typing rules. It will be the subject of
 Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
@@
 inductive definition, we are better off sharpening it before use.
-
-A full account of such generic induction schema can be found in Benke
-et al.~\cite{benke:universe-generic-prog,
-  dybjer:ir-initial-algebra}. In the following, we convey the
-intuition behind this schema by a step-by-step construction of the
-dependent eliminator from the catamorphism. The first step consists in
-turning $\V{T}$ into a predicate $\Bhab{\V{P}}{\Mu{\V{D}} \To \Set}$,
-hence introducing a dependency on the descriptions. Consequently, we
-strengthen the algebra $\descop{\V{D}}{\V{T}} \To \V{T}$ by a similar
-process. We start with turning the arrow into a dependent arrow, hence
-offering $\Bhab{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}$ to further
-analysis.
-
-Then, we have to translate the notion of algebra in our dependent
-setting. This is achieved by the predicate $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$
-(Fig.~\ref{fig:all-predicates}) that, intuitively, states that
-$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Based on that
-knowledge, we ought to be able to prove that $\V{P}$ holds for
-$\Con{\V{xs}}$. We close our elimination principle by taking an argument
-$x$ in $\Mu{\V{D}}$ and claiming that the motive $\V{P}$ is realised on $\V{x}$.
+Following \citet{benke:universe-generic-prog}, we adopt the following
+definition:
 
@@
 \]
+
+Where, intuitively, $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$ states that
+$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its formal
+definition is presented in Figure~\ref{fig:all-predicates}.  We refer
+the reader to our technical report for the implementation of
+$\F{induction}$. Having this dependent catamorphism, we are empowered
+with the ability to $\emph{compute}$ over descriptions. Note that this
+induction principle is the first manifestation of a generic operation
+over descriptions. Thanks to $\F{induction}$, any datatype we define
+automatically comes equipped with an induction principle.
 
@@
 
-At this stage, we are still left with implementing this
-$\F{induction}$ principle. For space reason, we refer the reader to
-our technical report for the actual code. Having this dependent
-catamorphism, we are empowered with the ability to $\emph{compute}$
-over descriptions. Note that this induction principle is the first
-manifestation of a generic operation over descriptions. Thanks to
-$\F{induction}$, any data-type we define automatically comes equipped
-with an induction principle.
-
 \subsection{Extending type propagation}
@@
 <- Extending type propagation
-    <- Data-type declaration turns into definitions
+    <- datatype declaration turns into definitions
         -> Straightforward translation to Desc
@@
 
-So far, we have introduced a universe of description. We have shown
-how to manipulate these codes to build data-types. We have also
-presented a fix-point construction and its associated dependent
-catamorphism. However, one could argue that programming with these
+We have now enough machinery to build and manipulate inductive
+types. In a word, we ought be able to \emph{program} over these
+datatypes.  However, one could argue that programming with these
 objects is not practical.
@@
 Whereas simply-typed languages feature a convenient grammar for
-defining data-types and manipulating them, our proposal could be seen
+defining datatypes and manipulating them, our proposal could be seen
 as a step backward. In our setting, the developer would have to write
-a description instead of a sugared data-type definition. Then, to
+a description instead of a sugared datatype definition. Then, to
 recognise an element of $\Nat$ for instance, she would have to
-identify $\Mu{}$ of a description as indeed the fix-point of the
-pattern functor of natural numbers. Finally, she would have to write
-codes instead of the natural data-type constructor.
+identify $\Mu{}$ of some coded description as indeed the fix-point of
+the pattern functor of natural numbers. Finally, she would have to
+write codes instead of datatype constructors.
 
@@
 cure lies in rationalising this information. Let us first tackle the
-issue of data-type declaration. The grammar of data-types we are used
+issue of datatype declaration. The grammar of datatypes we are used
 to can be straightforwardly translated to the sigmas-of-sigmas
-paradigm. Hence, a data-type definition can be turned into the
+paradigm. Hence, a datatype definition can be turned into the
 corresponding description. Then, at the level of definitions, we
-associate the name of the data-type to the fix-point of the
+associate the name of the datatype to the fix-point of the
 description. The induction principle as well as other generic
@@
 
-Doing so, we seal data-type definition behind a convenient
+Doing so, we seal datatype definition behind a convenient
 abstraction. We take advantage of this abstraction barrier to address
@@
 
-Regaining the ability to write data-type constructors is simply a
+Regaining the ability to write datatype constructors is simply a
 matter of type propagation. We therefore extend the type-checking
@@
 Hence, in the high-level expression language, the developer writes a
-familiar type constructor, applied to a telescope of arguments. In
-the type propagation setting, this expression will be pushed to a
-$\Mu{}$ type of a tagged description. It is then straightforward to
-look-up the tag and turn the constructor form into the appropriate
-code.
+familiar type constructor, applied to a telescope of arguments. During
+type propagation, this expression will be pushed to a $\Mu{}$ type of
+a tagged description. It is then straightforward to look-up the tag
+and turn the constructor form into the appropriate code.
 
 This concludes our presentation of the universe of descriptions. In
-this section, we have seen how data-types from the simply-typed world
+this section, we have seen how datatypes from the simply-typed world
 can be expressed in our dependently-typed setting. By relying on a
@@
 extended the type propagation machinery. In this setting, we can
-define, interact, and build data-types as transparently as in the
+define, interact, and build datatypes as transparently as in the
 simply-typed world.

### 2010-03-31 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix replace, adding hindx case
Desc@ICFP: proofread Section 4 'Levitating Desc'. Next ICFP: in Esperanto.

--- 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
@@
 Apart from these changes, we are left with implementing the other
-components of finite sets. This consists in $\EnumT$, $\Ze$, $\Su$, as
+components of finite sets. This consists of $\EnumT$, $\Ze$, $\Su$, as
 well as the $\F{switch}$ operator. Note that the actual implementation
@@
     /> type naming scheme condenses
-        <- Replace named constructors by codes in the universe of data-types
+        <- Replace named constructors by codes in the universe of datatypes
     -> Our next step is a similar move (in essence)
-        /> Condenses the entire naming scheme of data-types
+        /> Condenses the entire naming scheme of datatypes
 \end{wstructure}
@@
 descriptions. In essence, our next step is similar: we are going to
-condense the entire naming scheme of data-types \emph{in itself}.
+condense the entire naming scheme of datatypes \emph{in itself}.
 
@@
     <- We are going to implement Desc
-    /> Desc is itself a data-type
+    /> Desc is itself a datatype
         <- Same situation as EnumU
@@
 implement the codes of the universe of descriptions. In and by itself,
-the system of codes, $\Desc$, is nothing but a data-type. We are in
+the system of codes, $\Desc$, is nothing but a datatype. We are in
 the same situation as with $\EnumU$: we ought to be able to describe
@@
 first-class citizen, born with the standard, generic equipment of
-data-types.
+datatypes.
 
@@
 Let us explain how we have proceeded and the obstacle we face. Much as
-the data-types we have seen so far, we first have to choose a
+the datatypes we have seen so far, we first have to choose a
 constructor. Constructors correspond to the code's constructors:
@@
 The case of $\DUnit$ is trivial: we immediately reach the end of the
-data-type. $\DUnit$ has no payload. The case of $\DIndx{\!}$ comes
+datatype. $\DUnit$ has no payload. The case of $\DIndx{\!}$ comes
 naturally: it only takes a recursive argument -- a description. The
@@
 Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
-our universe of data-types, as shown in
+our universe of datatypes, as shown in
 Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
@@
 object gives us $\Desc$, up to isomorphism. We have implemented our
-code of data-types as an object levitating inside itself. However,
+code of datatypes as an object levitating inside itself. However,
 this levitation operation, just as any magic trick, relies on an
@@
     <- Free induction scheme on Desc
-        -> Ability to inspect data-types
-        -> Ability to program on data-types
+        -> Ability to inspect datatypes
+        -> Ability to program on datatypes
 \end{wstructure}
@@
 restricted to the meta-language: we now have all the necessary
-equipment in the target language to \emph{program} over data-types. In
+equipment in the target language to \emph{program} over datatypes. In
 this setting, \emph{generic programming is just about programming}.
@@
 \F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{xs}}{}\: \pair{\V{g}}{\V{ys}}{} \mapsto   \\
-\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\{ \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
+\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
 }\]
@@
 problem. In the type theory, we have built a generic catamorphism. Any
-data-type will now come equipped with this operation, for free.
+datatype will now come equipped with this operation, for free.
 
@@
             -> More generic functions
-    <- Inspecting data-types
+    <- Inspecting datatypes
         <- All described byu a Desc code
@@
 manipulate descriptions as first-class objects: the catamorphism is,
-basically, a function mapping a $\Desc$ to a data-type specific
+basically, a function mapping a $\Desc$ to a datatype specific
 operation. This programming style is reminiscent to polytypic
@@
 Moreover, the $\F{replace}$ function demonstrates the benefit of an
-approach based on universes. The data-types living in the universe of
+approach based on universes. The datatypes living in the universe of
 descriptions, we are able to \emph{inspect} them. As shown by
@@
 In this section, we will turn to a more ambitious generic operation on
-data-type. Given a functor, represented as a tagged description, we
+datatype. Given a functor, represented as a tagged description, we
 build the free monad over this functor.
@@
 Let us recall the free monad construction. Given a functor $F$, the
-free monad over $F$ is defined by the following data-type:
+free monad over $F$ is defined by the following datatype:
 
@@
 
-Being an inductive type, this $\FMFreeMonad$ data-type is itself
+Being an inductive type, this $\FMFreeMonad$ datatype is itself
 defined by a pattern functor. It is given by:
@@
 
-Of course, we must equip the resulting data-types with operations
+Of course, we must equip the resulting datatypes with operations
 delivering a monadic interface. As expected, \(\LAM{\x}\DVar{\x}\)
@@
     -> Consequences
-        <- We have free monad data-type
+        <- We have free monad datatype
             <- Term + variables
@@
 implemented the free monad construction for an arbitrary tagged
-description. This gives the developer the ability, for any data-type,
+description. This gives the developer the ability, for any datatype,
 to extend it with a notion of variable. Then, we have equipped this
@@
 terms. We have shown how to implement it, for any free monad. This
-programming style is reminiscent to the type-indexed data
-types~\cite{hinze:generic-haskell}, as implemented in Generic Haskell:
-from data-types, we built new data-types and equip them with some
-structure.
+programming style is reminiscent to type-indexed
+datatypes~\cite{hinze:generic-haskell}, as implemented in Generic
+Haskell: from datatypes, we built new datatypes and equip them with
+their structure.
 
@@
 <- Deriving new data-structure and functions on them
-    <- Computing the Free Monad of a data-type
+    <- Computing the Free Monad of a datatype
         <- Derive new data-structure from previous one
@@
         /> New data-structure comes with some equipment
-    <- Computing new functions on computed data-types
+    <- Computing new functions on computed datatypes
         <- If data comes with structure, we ought to be able to capture it
@@
 
+%% Candidate for removal:
 With the free monad construction, we have seen two kinds of generic
@@
 one: we make the free monad from its underlying functor. To do so, we
-crucially rely on the fact that data-types are nothing but codes. We
+crucially rely on the fact that datatypes are nothing but codes. We
 are therefore entitled to modify this code and, in this case, extend
-it. Extending a data-type might give rise to a more structured object,
+it. Extending a datatype might give rise to a more structured object,
 as was the case here.  So, secondly, we have equipped this new
-data-type with its inherent structure: the \bind\ and
+datatype with its inherent structure: the \bind\ and
 \return\ operations. We have been able to build them as generic

### 2010-03-31 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: proofread Section 4 'Levitating Desc'. Next ICFP: in Esperanto.
Desc@ICFP: proofread Section 6 'Discussion'.

@@
 \[\stk{
-\data \IDesc{\!}^{\blue{l}} (\Bhab{\V{I}}{\Set^{\blue{l+1}}}) : \Set^{\blue{l+1}} \where \\
-\;\;\ldots \\
+\data \IDesc{\!}^{\blue{l}} (\Bhab{\V{I}}{\Set^{\blue{l+1}}}) : \Set^{\blue{l+1}} \where \ldots \\
 \\
@@
 
-In particular, our approach is similar in spirit with polytypic
+In particular, our approach is similar in spirit to polytypic
 programming, as initiated by PolyP~\cite{jansson:polyp}. Indeed,
 generic functions, in our system, are built by induction on the
-pattern functor. Unlike PolyP, we do not have to resort to pre-process
-data-type definitions: our data-types are, natively, nothing but
-codes.
+pattern functor. Unlike PolyP, we do not have to pre-process datatype
+definitions: our datatypes are, natively, nothing but codes.
 
@@
 approach~\cite{hinze:generic-haskell}, as exemplified by the free
-monad construction: from data-type, we can compute new data-types and
+monad construction: from datatype, we can compute new datatypes and
 equip them with their structure. Generic Haskell also features
 \emph{generic views}~\cite{holdermans:generic-view}, transparently
-transforming the structure of data-type definitions, to ease the
+transforming the structure of datatype definitions, to ease the
 implementation of generic functions over them. Views in
 dependently-typed system are widely
-used~\cite{mcbride.mckinna:view-from-the-left}. Unlike in Generic
+used~\cite{mcbride.mckinna:view-from-the-left}. Unlike Generic
 Haskell, we do not have to modify the compiler to develop views on
-data-types: we can massage description codes from inside the language
-and present them as a standard view. An example of such operation is
-the tagged descriptions, presenting data-types under a particular
-angle.
+datatypes: we can massage descriptions from inside the language. An
+example of such operation is the tagged descriptions, presenting
+datatypes under a sum-of-sigmas angle.
 
@@
 limited to endofunctor on $\Set$ and $\Set^I$. While it is possible to
-\emph{encode} higher-kinded data-types in our setting, we do not plan
+\emph{encode} higher-kinded datatypes in our setting, we do not plan
 to adopt this strategy. Rather, as future work, we plan to extend our
@@
 Boilerplate~\cite{spj:syb} (SYB). Our proposal is different in various
-ways. The corner stone of SYB is the \emph{spine} view of data-type
-constructors. A data-type is a spine composed by a constructor applied
-to some arguments. Further, this spine is equipped with some
+ways. The corner stone of SYB is the \emph{spine} view of datatype
+constructors. A piece of data is a spine composed by a constructor
+applied to some arguments. Further, this spine is equipped with some
 combinators including, primarily, an iterator. In this setting,
-generic programs are written by composing these combinators. This
-relies on a $\CN{Typeable}$ type-class, allowing dynamic dispatch to
-data-type specific operations. As a result, SYB is not reflexive: it
-is restricted to data-types instanciating $\CN{Typeable}$. Moreover,
-it forbids type-indexed data types: it can only define generic
-functions.
+generic programs are written by composing these combinators, building
+generic operations on spines. This relies on a $\CN{Typeable}$
+type-class, allowing dynamic dispatch to datatype-specific
+operations. As a result, SYB is not reflexive: it is restricted to
+datatypes instanciating $\CN{Typeable}$. Moreover, it is limited to
+building generic functions, hence type-indexed datatypes cannot be
+implemented in this framework.
 
@@
 
-The interest in generic programming for dependent types is not new
+The interest for generic programming in dependent types is not new
 either. \citet{norell:msc-thesis} has given a formalization of
@@
 code. Our approach is different in that we aim at building a
-foundation for data-types, in a dependently-typed theory, for a
+foundation for datatypes, in a dependently-typed theory, for a
 dependently-typed system.
@@
 universe of finitary indexed induction. This is not surprising, as we
-share the same source of inspiration, namely induction recursion.
+share the same source of inspiration, namely induction-recursion.
 
@@
 \emph{programming} with such codes: it is not obvious, to us, how one
-can define, manipulate, and program over these coded data-types. With
+can define, manipulate, and program over these coded datatypes. With
 descriptions, we have shown how to abstract away codes and present a

### 2010-03-31 papers/icfp-2010-desc/paper_type_theory.tex
sec 2.2 finite stuff

@@
 
-\subsection{Finite sets}
+\subsection{Finite enumerations of tags}
 \label{sec:finite-sets}
@@
 
-As a motivating example, we are now going to extend the type theory
-with a notion of finite set. One could argue that there is no need for
-such an extension: finite sets, just as any data-structure, can be
-encoded inside the type theory. A well-known example of such encoding
-is the Church encoding of natural numbers, which is isomorphic to
-finite sets. \note{Shall we talk about W-types encoding?}
+It is time for our first example of a \emph{universe}. You might
+want to offer a choice of named constructors in your datatypes, we shall
+equip you with sets of tags to choose from. Our plan
+is to implement (by extending the theory, or by encoding) the signature
+\[
+  \Type{\EnumU}\qquad \Type{\EnumT{(\Bhab{\M{E}}{\EnumU})}}
+\]
+where some value \(E:\EnumU\) in the `enumeration universe' describes
+a type of tag choices \(\EnumT{E}\). We shall need
+some tags---valid identifiers, marked to indicate that
+they are data, not variables scoped and substitutable---so we hardwire these
+rules:
+\[
+%% UId
+\Rule{\Gamma \vdash \Valid}
+     {\Gamma \vdash \Type{\UId}}
+\qquad
+%% Tag
+\Rule{\Gamma \vdash \Valid}
+     {\Gamma \vdash \Bhab{\Tag{\V{s}}}{\UId}}\;\V{s}\: \mbox{a valid identifier}
+\]
+Let us describe enumerations as lists of tags, with signature:
+\[
+\Bhab{\NilE}{\EnumU}\qquad
+\Bhab{\ConsE{(\Bhab{\M{t}}{\UId})}{(\Bhab{\M{E}}{\EnumU})}}{\EnumU}
+\]
+What are the \emph{values} in \(\EnumT{E}\)? Formally, we represent
+the choice of a tag as a numerical index into \(E\), via new rules:
+\[
+%% Ze
+\Rule{\Gamma \vdash \Valid}
+     {\Gamma \vdash \Bhab{\Ze}{\EnumT{(\ConsE{\M{t}}{\M{E}})}}} 
+\qquad
+%% Su
+\Rule{\Gamma \vdash \Bhab{\M{n}}{\EnumT{\M{E}}}}
+     {\Gamma \vdash \Bhab{\Su{\M{n}}}{\EnumT{(\ConsE{\M{t}}{\M{E}})}}}
+\]
+However, we expect that in practice, you might rather refer to these
+values \emph{by tag}, and we shall ensure that this is possible in due course.
 
-However, using encodings is impractical. In the case of finite sets,
-for instance, we would like to name the elements of the sets. Then, we
-need to be able to manipulate these elements by their name, instead of
-their encoding. While we are able to give names to encodings, it is
-extremely tedious to map the encodings back to a name. Whereas these
-objects have a structure, the structure is lost during the encoding,
-when they become anonymous inhabitants of a $\Pi$ or $\Sigma$-type.
+%As a motivating example, we are now going to extend the type theory
+%with a notion of finite set. One could argue that there is no need for
+%such an extension: finite sets, just as any data-structure, can be
+%encoded inside the type theory. A well-known example of such encoding
+%is the Church encoding of natural numbers, which is isomorphic to
+%finite sets. \note{Shall we talk about W-types encoding?}
 
-In the simply-typed world, we are used to see types as a coding
-presentation -- presentation of invariants, presentation of
-properties. In the dependently-typed world, we also learn to use types
-as a coding representation: finite sets being good citizens, they
-ought to be democratically represented at the type level. As we will
-see, this gives us the ability to name and manipulate them (this is
-were the democracy analogy goes crazy, I think).
-\note{Did I got the coding presentation vs. coding representation
-  story right? No.}
+%However, using encodings is impractical. In the case of finite sets,
+%for instance, we would like to name the elements of the sets. Then, we
+%need to be able to manipulate these elements by their name, instead of
+%their encoding. While we are able to give names to encodings, it is
+%extremely tedious to map the encodings back to a name. Whereas these
+%objects have a structure, the structure is lost during the encoding,
+%when they become anonymous inhabitants of a $\Pi$ or $\Sigma$-type.
+
+%In the simply-typed world, we are used to see types as a coding
+%presentation -- presentation of invariants, presentation of
+%properties. In the dependently-typed world, we also learn to use types
+%as a coding representation: finite sets being good citizens, they
+%ought to be democratically represented at the type level. As we will
+%see, this gives us the ability to name and manipulate them (this is
+%were the democracy analogy goes crazy, I think).
+%\note{Did I got the coding presentation vs. coding representation
+%  story right? No.}
 
@@
 
-The specification of finite sets is presented in
-Figure~\ref{fig:typing-finite-set}. It is composed of three
-components. First, we define tags as inhabitants of the $\UId$ type. A
-tag is solely an informative token, used for diagnostic
-purposes. Finite sets inhabits the $\EnumU$ type. Unfolding the
-definition, we get that a finite set is a list of tags. Finally,
-elements of a finite set $\V{u}$ belong to the corresponding $\EnumT{\V{u}}$
-type. Intuitively, it corresponds to an index -- a number -- pointing
-to an element of $\V{u}$.
+%The specification of finite sets is presented in
+%Figure~\ref{fig:typing-finite-set}. It is composed of three
+%components. First, we define tags as inhabitants of the $\UId$ type. A
+%tag is solely an informative token, used for diagnostic
+%purposes. Finite sets inhabits the $\EnumU$ type. Unfolding the
+%definition, we get that a finite set is a list of tags. Finally,
+%elements of a finite set $\V{u}$ belong to the corresponding $\EnumT{\V{u}}$
+%type. Intuitively, it corresponds to an index -- a number -- pointing
+%to an element of $\V{u}$.
 
-\begin{figure}
+%\begin{figure}
 
-\input{figure_finite_sets}
+%\input{figure_finite_sets}
 
-\caption{Typing rules for finite sets}
-\label{fig:typing-finite-set}
+%\caption{Typing rules for finite sets}
+%\label{fig:typing-finite-set}
 
-\end{figure}
+%\end{figure}
 
@@
 
-The type of finite sets comes with further machinery. Indeed, to
-manipulate the $\EnumU$ and $\EnumT{\!}$ objects, we need to equip our
-type theory with an appropriate \emph{elimination
-  principle}. Formally, as we add types for data -- such as finite
-sets here --, we need such primitive operators to compute with those
-data, and corresponding extension of the judgemental equality with
-their computational behavior. For compactness and readability, we
-shall present these operators as functional programs, much as we
-implement them in our Agda model. Hence, we first define the
-$\spi{\!}{\!}$ operator:
+Enumerations come with further machinery. Each \(\EnumT{E}\) needs an
+eliminator, allowing us to branch according to a tag choice. Formally,
+whenever we need such new computational facilities, we add primitive
+operators to the type theory and extend the judgmental equality with
+their computational behavior. However, for compactness and readability, we
+shall write these operators as functional programs (much as we
+model them in Agda).
 
+We first define the `small product' $\spi{\!}{\!}$ operator:
 \[\stk{
 %% spi
-\spi{\!}{\!} : \PITEL{\V{e}}{\EnumU}
-           \PITEL{\V{P}}{\EnumT{\V{e}} \To \Set} \To \Set \\
-\begin{array}{@{}ll@{\:\mapsto\:\:}l}
+\spi{}{}: \PITEL{\V{E}}{\EnumU}\PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To \Set \\
+\begin{array}{@{}l@{\:}l@{\:\mapsto\:\:}l}
 \spi{\NilE}{& \V{P}}        & \Unit \\
-\spi{(\ConsE{\V{t}}{\V{e}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{e}}{(\LAM{\V{x}} \V{P}\: (\Su{\V{x}}))}}
+\spi{(\ConsE{\V{t}}{\V{E}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{E}}{(\LAM{\V{x}} \V{P}\: (\Su{\V{x}}))}}
 \end{array}
 }\]
-
-A closer look to its type should reveal a certain familiarity with its
-equivalent for sets, $\Pi$. Indeed, $\spi{\!}{\!}$ defines a
-\emph{finite function space}, mapping elements of finite sets to
-objects in $\Set$. Intuitively, this operator builds a right-nested
-tuple, effectively mapping an object of $P\:i$ for each $i$ the finite
-domain.
-
-Having understood the meaning of $\spi{\!}{\!}$, the definition of the
-$\F{switch}$ eliminator is not surprising:
-
+This builds a right-nested
+tuple type, demanding an object of $P\:i$ for each $i$ in the given finite
+domain. We can see these tuples as `jump tables' tabulating dependently
+typed functions from the domain. We give this functional
+interpretation---the eliminator we need---by the
+$\F{switch}$ operator, which, unsurprisingly, iterates projection:
 \[\stk{
@@
 \begin{array}{@{}ll}
-\F{switch} : & \PITEL{\V{e}}{\EnumU}
-               \PITEL{\V{P}}{\EnumT{\V{e}} \To \Set} \\
-             & \PITEL{\V{b}}{\spi{\V{e}}{\V{P}}}
-               \PITEL{\V{x}}{\EnumT{\V{e}}} \To P~x
+\F{switch} :& \PITEL{\V{E}}{\EnumU}
+               \PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To\\
+             & \spi{\V{E}}{\V{P}} \To
+               \PITEL{\V{x}}{\EnumT{\V{E}}} \To \V{P}~\x
 \end{array} \\
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\switch{(\ConsE{\V{t}}{\V{e}})}{\V{P}}{\V{b}}{\Ze}      & \fst{\V{b}} \\
-\switch{(\ConsE{\V{t}}{\V{e}})}{\V{P}}{\V{b}}{(\Su{\V{x}})} & \switch{\V{e}}{(\LAM{\V{x}} \V{P}
+\switch{(\ConsE{\V{t}}{\V{E}})}{\V{P}}{\V{b}}{\Ze}      & \fst{\V{b}} \\
+\switch{(\ConsE{\V{t}}{\V{E}})}{\V{P}}{\V{b}}{(\Su{\V{x}})} & \switch{\V{E}}{(\LAM{\V{x}} \V{P}
   (\Su{\V{x}}))}{(\snd{\V{b}})}{\V{x}}
@@
 
-Again, there is a clear equivalent in the full-$\Set$ world: function
-application. The operational behaviour of $\F{switch}$ is
-straightforward: $\V{x}$ is peeled off as we move deeper inside the nested
-tuple $\V{b}$. When $\V{x}$ equals $\Ze$, we simply return the value we are
-pointing to.
+%Again, there is a clear equivalent in the full-$\Set$ world: function
+%application. The operational behaviour of $\F{switch}$ is
+%straightforward: $\V{x}$ is peeled off as we move deeper inside the nested
+%tuple $\V{b}$. When $\V{x}$ equals $\Ze$, we simply return the value we are
+%pointing to.
 
@@
 
-As we have seen, finite sets are not just a way of making
-data. Through the $\spi{\!}{\!}$ and $\F{switch}$ operators, they also
-come with a notion of finite function space. However, we had to
-extract that intuition from the type, by a careful reading. This seems
-to contradict our argument in favour of types for coding
-representation. Here, we are overflown by low-level, very precise type
-information.
+The $\spi{\!}{\!}$ and $\F{switch}$ operators deliver dependent
+elimination for finite enumerations, but are rather awkward to
+use directly. We do not write the range for a \(\LAMBINDER\)-abstraction, so
+it is galling to supply \(\V{P}\) for functions defined by $\F{switch}$.
+Let us therefore find a way to recover the tedious details of the
+encoding from types.
 
-However, our situation is significantly different from the one we
-faced with encoded data: while we were suffering from a crucial lack
-of information, we are now facing too much information, hence losing
-focus. This is a general issue with the usage of codes, as they convey
-much more information than what the developer is willing to see.
+%, they also
+%come with a notion of finite function space. However, we had to
+%extract that intuition from the type, by a careful reading. This seems
+%to contradict our argument in favour of types for coding
+%representation. Here, we are overflown by low-level, very precise type
+%information.
 
-As we will see in the following section, there exists a cure to this
-problem. In a nutshell, instead of being overflown by typing
-information, we will put it at work, automatically. The consequence is
-that, in such system, working with codes is \emph{practical}: one
-should not be worried by information overload, but how to use it as
-much as possible. Therefore, we should not be afraid of using codes for
-practical purposes.
+%However, our situation is significantly different from the one we
+%faced with encoded data: while we were suffering from a crucial lack
+%of information, we are now facing too much information, hence losing
+%focus. This is a general issue with the usage of codes, as they convey
+%much more information than what the developer is willing to see.
+
+%As we will see in the following section, there exists a cure to this
+%problem. In a nutshell, instead of being overflown by typing
+%information, we will put it at work, automatically. The consequence is
+%that, in such system, working with codes is \emph{practical}: one
+%should not be worried by information overload, but how to use it as
+%much as possible. Therefore, we should not be afraid of using codes for
+%practical purposes.
 

### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: WeightWatch section 3, A universe of inductive types

@@
 In this section, we describe an implementation of inductive types, as
-we know them in ML-like languages. This choice is motivated by
-pedagogical considerations: by describing the layman datatypes in our
-setting, we build on our intuition of datatypes we are familiar
-with. In particular, we do not aim at describing a model for generics
-in simply-typed functional language. Our goal is and remains to
-discuss datatypes as we find them in dependently-typed languages.
+we know them in ML-like languages. By working with datatypes we are
+familiar with, we rely on our intuition on these objects. However, our
+goal remains to discuss datatypes as we find them in dependently-typed
+languages.
 
-Our proposal is inspired by, but less involved than, Dybjer and
-Setzer's closed formulation of
-induction-recursion~\cite{dybjer:general-ir, dybjer:axiom-ir,
-  dybjer:ir-initial-algebra, dybjer:iir}. For inductive datatypes, the
+Our proposal is inspired by Dybjer and Setzer's closed formulation of
+induction-recursion~\cite{dybjer:axiom-ir}. For inductive datatypes, the
 full power of induction-recursion is not necessary. Therefore, we
-adopt a stripped-down version of it. As we broaden the scope of our
-universe of datatypes, our presentation will diverge from
-induction-recursion. We shall compare both approaches in
-Section~\ref{sec:discussion}.
+adopt a stripped-down version of it.
 
@@
 two different ways. This duality is actually reflected in the notation
-we can find in the literature, depending on the sensibility of the
-author.  The notation $\blue{\Sigma}_{\V{x} : \V{A}} (\V{B}\: \V{x})$
-stresses the interpretation that $\Sigma$-types are a generalisation
-of sums over arbitrary arities.  Where simply-typed languages have
-finite sums, dependently-typed languages have sums indexed by any set.
+we can find in the literature. The notation $\blue{\Sigma}_{\V{x} :
+  \V{A}} (\V{B}\: \V{x})$ stresses that $\Sigma$-types are a
+generalisation of sums over arbitrary arities.  Where simply-typed
+languages have finite sums, dependently-typed languages have sums
+indexed by any set.
 
-On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}} (\V{B}\:\V{x})$
-emphasises that $\Sigma$-types can be read as a generalisation of products,
-where the second component can depend on the first one. When simply-typed
-languages pack data into non-dependent tuples, dependently-typed languages
-tolerate this non-dependent usage. However, they also give the rather novel
-ability for data to influence types of further data.
+On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}}
+(\V{B}\:\V{x})$ emphasises that $\Sigma$-types are a generalisation of
+products, where the second component can depend on the first
+one. Where simply-typed languages pack data into non-dependent tuples,
+we can now build tuples where data influence further types.
 
@@
 
-In the simply-typed world, datatype definitions are often referred to
-as \emph{sum-of-product}. A datatype is defined by a finite sum -- a
-choice -- of constructors, each of them is composed of a product -- a
-tuple -- of arguments. This form governs the grammar of datatype
-definition in languages such as Haskell or ML.
+In ML-like languages, datatype definitions are called
+\emph{sum-of-product}. A datatype is defined by a finite sum of
+constructors, each composed by a product of arguments. To embrace
+these datatypes, we need to capture this grammar.
 
-To model these datatypes, we simply need to capture this grammar in
-our dependently-typed setting. Here, the notion of sum-of-product
-naturally translates into \emph{sigmas-of-sigmas}. Beyond this
-intuition, it should be clear that a direct encoding of datatypes
-through raw $\Sigma$-types is not a viable option. Indeed, just as for
-finite sets, encoding throws away information when we crucially need
-it. Moreover, to program with these datatypes, one would need an
-induction principle. An encoding will be unable to deliver such
+With dependent types, the notion of sum-of-product translates into
+\emph{sigmas-of-sigmas}. However, a direct encoding of datatypes
+through raw $\Sigma$-types is not a viable option. Indeed, to program
+with these datatypes, we need an induction principle. An encoding is
+unable to deliver such
 operation~\cite{geuvers:induction-not-derivable}.
@@
 datatypes, we have seen that it makes a poor \emph{syntax}. To
-decouple these two notions, we use a standard technique in
-dependently-typed programming: we construct an ad-hoc universe. This
-technique dates back to Martin-L\"of definition of type
-theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
-programming technique~\cite{benke:universe-generic-prog,
-  oury:power-of-pi} \note{Any idea of other papers?}.  We refer the
-reader to the Agda tutorial~\cite{norell:agda-tutorial} for a
-pedagogical presentation of universe construction.  \note{Is that
-  really a motivating motivation for the usage of universe
-  construction?}  \note{pwm: Surely ($\EnumU$, $\EnumT$) is a useful
-  example of universes too.  Even if it doesn't show off their full
-  power?}
+decouple these two notions, we construct an ad-hoc
+\emph{universe}. This technique originates in Martin-L\"of definition
+of type theory~\cite{martin-lof:itt}. It is a fruitful programming
+technique in dependently-typed
+languages~\cite{benke:universe-generic-prog, oury:power-of-pi}
+\note{Any idea of other papers?}.  \note{Is that really a motivating
+  motivation for the usage of universe construction?}  \note{pwm:
+  Surely ($\EnumU$, $\EnumT$) is a useful example of universes too.
+  Even if it doesn't show off their full power?}
 
@@
 by defining new types. These names are called \emph{codes}. By
-defining a set of codes, we define the syntax of a language. However,
-as such, a system of codes is useless as it lacks a semantics. Instead
-of equipping the universe of codes with some computational behaviour,
-we pragmatically chose to \emph{interpret} these codes back into the
-standard type theory. Hence, codes act only as labels, while the type
-theory provides the computational machinery. \note{Is that a good
-  description?}
+defining a set of codes, we define the syntax of a language. To give
+it a semantics, we \emph{interpret} these codes back into the type
+theory. Hence, codes act as labels, while the type theory provides the
+computational machinery. \note{Is that a good description?}
 
-Codes being simple labels, we have the ability to inspect them, hence
-regaining structural information. Being able to inspect them, we are
-therefore free to compute over them: deriving new codes from previous
-ones, or even new functions on them. In several occasions, we will
-have the opportunity to witness the power of universes.
+Codes being mere labels, we can inspect them, hence regaining
+structural information. We can also compute over them: deriving new
+codes from others, or even functions on them.
 
@@
 
-We propose to embed inductive types as a universe in our dependent
-type theory, the universe of \emph{descriptions}
-(Figure~\ref{fig:desc_universe}). As expected, the code of this
-universe mimics the standard grammar of datatypes definitions in
-simply-typed languages. Hence, we have a $\DSigma{\!}{\!}$ code,
-interpreted as a $\Sigma$-type, to build the
-sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
-contains no useful payload. The functoriality of the datatypes is
-introduced by $\DIndx{\!}$. When we tie the knot with a fix-point, the
-hole left open by $\DIndx{\!}$ will be turned into a standard
-recursive argument. We notice that this functoriality appears in the
-type of $\descop{\V{D}}{\!}$ itself, for a given datatype definition
-$\V{D}$. This corresponds to the object part of the functor described
-by $\V{D}$.
+To embed inductive types in our type theory, we define a universe of
+\emph{descriptions} (Figure~\ref{fig:desc_universe}). As expected, the
+code of this universe mimics the grammar of inductive types. To build
+sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
+$\Sigma$-type. Descriptions are terminated by $\DUnit$, containing no
+payload. To make recursive arguments, we introduce the $\DIndx{\!}$
+code. It interprets to a hole that will be closed by the fix-point. We
+can also read $\V{D}$ as describing an endofunctor on $\Set$, with
+$\descop{\V{D}}{\!}$ being the object part.
 
@@
 
-To give some intuition on this universe of descriptions, we now turn
-to some examples. For obvious pedagogical reasons, we will manually
-build these descriptions. However, it should be clear that, in
-practice, these definitions can be automatically constructed from an
-ML-like $\data$ definition.
+To give some intuition on this universe, we now turn to some
+examples. For obvious pedagogical reasons, we will manually build
+these descriptions. However, it should be clear that, in practice,
+these definitions are automatically computed from an ML-like $\data$
+definition.
 
@@
 of Section~\ref{sec:type-propagation}. The translation back to the raw
-terms is laborious but should not pose any difficulty. The code is the
-following:
+terms should not pose any difficulty. The code is the following:
 
@@
 Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
-give a choice between the $\NatZero$ and $\NatSuc{\!}$
-constructors. In the $\NatZero$ case, we reach the end of the
-description: there is no useful payload in that case. In the
-$\NatSuc{\!}$ case, we leave one hole open for the recursive argument,
-and we close the description.
-
-In a more synthetic notation, we have described the following functor:
+give a choice between the $\NatZero$ and $\NatSuc{}$ constructors. In
+the $\NatZero$ case, we reach the end of the description. In the
+$\NatSuc{\!}$ case, we attach a recursive argument and close the
+description. In a more synthetic notation, we have described the
+following functor:
 
@@
 
-With a small change to the definition of $\NatD$, we obtain the
-pattern functor for lists:
+With a small change to this definition, we obtain the pattern functor
+for lists:
 
@@
 
-The modification consists in turning the $\NatSuc{\!}$ constructor
-into a proper $\ListCons{\!}{\!}$ taking an argument in $\V{X}$
-followed by a recursive argument. In this case, we use
-$\DSigma{\!}{\!}$ in its product interpretation: we pack an element of
-$\V{X}$ together with the recursive argument. Easily, one sees that
-this code actually describes the following functor:
+The $\NatSuc{\!}$ constructor is turned into a proper
+$\ListCons{\!}{\!}$, taking an argument in $\V{X}$ followed by a
+recursive argument. Easily, one sees that this code describes the
+following functor:
 
@@
 
-Again, this is a clear reminiscence of the sum-of-product style:
-we have a finite sum of constructors. Of course, every description can
-be expressed in this style, using a singleton as single constructor if
-necessary. A tagged description can trivially be turned into a
-full-blown description. We write $\toDesc{\V{TD}}$ for the description computed
-from the tagged description $\V{TD}$.
+Again, this is a clear reminiscence of the sum-of-product style: we
+have a finite sum of constructors. Every description can be expressed
+in this style, using a singleton as a constructor if
+necessary. Converselly, a tagged description can be turned into a
+full-blown description. We denote $\toDesc{\V{TD}}$ the description
+computed from the tagged description $\V{TD}$.
 
@@
 For convenience, we have taken for granted the existence of $\Desc$,
-as presented in Figure~\ref{fig:desc_universe}. In particular, we have
-considered its codes, $\DSigma{\!}{\!}$, $\DIndx{\!}$, and $\DUnit$ as
-type formers, extending the basic type theory. Hence, in an
-implementation of this type theory, we need to extended it with these
-constructors and their typing rules. Although it makes no conceptual
-difference, the code of description should rather be read as a
-specification. We promise the existence of such objects, satisfying
-the typing rules. It will be the subject of
-Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
-implementing the specification.
+as presented in Figure~\ref{fig:desc_universe}. Although it makes no
+conceptual difference, the code of description should rather be read
+as a specification. We promise the existence of such objects,
+satisfying the typing rules. It will be the subject of
+Section~\ref{sec:desc-levitate} to actually implement the
+specification.
 
@@
 
-So far, we have used our universe of descriptions to build
-functors. \note{ Strictly positive types anyone? } To illustrate its
-usage, we have implemented the pattern functors of natural numbers,
-lists, and binary trees. However, the class of functors expressible
-with descriptions enjoys another property: they all admit a
-fix-point. Indeed, descriptions are interpreted into \emph{polynomial
-  functors}~\cite{who?}\note{who is the relevant source for polynomial
-  functors?}, as they are only composed of $\Sigma$-types, the unit
-type, and variables.
-
-We construct the fix-point by \emph{tying the knot} of the
+So far, we have used our universe of descriptions to build pattern
+functors. Being polynomial functors, they all admit a fix-point. We
+construct the fix-point by \emph{tying the knot} of the
 interpretation: the holes left open by the functor are filled by its
@@
 this type is not dependent, being restricted to a notion of
-\emph{iteration}. As this operator will be the corner stone of any
-inductive definition, we are better off sharpening it before use.
+\emph{iteration}. As this operator will be the corner stone of
+inductive definitions, we are better off sharpening it before use.
 Following \citet{benke:universe-generic-prog}, we adopt the following
@@
 \[
-\begin{array}{lcll}
-\F{induction} & : & \multicolumn{2}{l}{\PITEL{\V{D}}{\Desc}}                   \\
-              &   & \multicolumn{2}{l}{\PITEL{\V{x}}{\Mu{\V{D}}}}                   \\
-              &   & \multicolumn{2}{l}{\PITEL{\V{P}}{\Mu{\V{D}} \To \Set}}         \\
-              &   & \PI{\V{m}}{& \PI{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}              \\
-              &   &           & \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{xs}} \To \V{P} (\Con{\V{xs}})} \\
-              &   & \V{P} \V{x}
+\begin{array}{lcl}
+\F{induction} & : & \PITEL{\V{D}}{\Desc}
+                    \PI{\V{P}}{\Mu{\V{D}} \To \Set}         \\
+              &   & (\PI{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}              \\
+              &   & \ \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{xs}} \To \V{P} (\Con{\V{xs}})) \To \\
+              &   & \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x}
 \end{array}
@@
 Where, intuitively, $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$ states that
-$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its formal
+$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its
 definition is presented in Figure~\ref{fig:all-predicates}.  We refer
 the reader to our technical report for the implementation of
-$\F{induction}$. Having this dependent catamorphism, we are empowered
-with the ability to $\emph{compute}$ over descriptions. Note that this
-induction principle is the first manifestation of a generic operation
-over descriptions. Thanks to $\F{induction}$, any datatype we define
-automatically comes equipped with an induction principle.
+$\F{induction}$. Note that this induction principle is the first
+manifestation of a generic operation over descriptions. Any datatype
+we define automatically comes with an induction principle.
 
@@
 
-Whereas simply-typed languages feature a convenient grammar for
-defining datatypes and manipulating them, our proposal could be seen
-as a step backward. In our setting, the developer would have to write
-a description instead of a sugared datatype definition. Then, to
-recognise an element of $\Nat$ for instance, she would have to
-identify $\Mu{}$ of some coded description as indeed the fix-point of
-the pattern functor of natural numbers. Finally, she would have to
-write codes instead of datatype constructors.
-
-As for finite sets previously, these symptoms clearly indicate that we
-suffer from too much type information. Just as for finite sets, the
-cure lies in rationalising this information. Let us first tackle the
-issue of datatype declaration. The grammar of datatypes we are used
-to can be straightforwardly translated to the sigmas-of-sigmas
-paradigm. Hence, a datatype definition can be turned into the
-corresponding description. Then, at the level of definitions, we
-associate the name of the datatype to the fix-point of the
-description. The induction principle as well as other generic
-operations follow for free.
-
-Doing so, we seal datatype definition behind a convenient
-abstraction. We take advantage of this abstraction barrier to address
-the second issue. Namely, the developer would rather be presented a
-name than a fix-point of a pattern functor. Surprisingly, it is
-sufficient to \emph{label} the $\Mu{}$ constructor with the
-user-defined name. When pretty-printing this object, we just expose
-the label, instead of the definition in full.
-
-Regaining the ability to write datatype constructors is simply a
-matter of type propagation. We therefore extend the type-checking
-framework of Figure~\ref{fig:type-checking} with the following
-inference rule:
+Writing codes instead of datatype constructor is indeed unpractical.
+However, regaining this ability is simply a matter of type
+propagation. We extend the type-checking framework of
+Figure~\ref{fig:type-checking} with the following inference rule:
 
@@
 Where $\vec{\V{e}}$ denotes a telescope of arguments applied to the
-constructor $\V{C}$. For the sake of readability, we write function
-application in the small function space as $\V{f}\: \V{C}$. This
-corresponds to a call to $\F{switch}$. Also, we note
-$\widetilde{\V{e}}$ the right-nested tuple terminated by $\Void$
-corresponding to the curried telescope $\vec{\V{e}}$.
+constructor $\V{C}$. For readability, we write $\V{f}\: \V{C}$ instead
+of a call to $\F{switch}$. Also, we note $\widetilde{\V{e}}$ the
+right-nested tuple terminated by $\Void$ corresponding to the curried
+telescope $\vec{\V{e}}$.
 
@@
 this section, we have seen how datatypes from the simply-typed world
-can be expressed in our dependently-typed setting. By relying on a
-universe construction technique, we benefit from the ability to
-inspect and manipulate these codes. This opens some appealing
-opportunities for generic programming. Finally, we have demonstrated
-that programming with codes is practical. To this end, we have
-extended the type propagation machinery. In this setting, we can
-define, interact, and build datatypes as transparently as in the
-simply-typed world.
+can be expressed in a dependently-typed setting. By relying on a
+universe, we benefit from the ability to inspect and manipulate
+codes. We have shown that programming with codes is practical.

### 2010-04-01 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: WeightWatch section 4, Levitation

@@
 
-In this section, we will fulfil several promises. One promise we made
-was to effectively implement the code of the $\Desc$ universe. Another
-promise we tacitly made was the existence of the type formers for
-finite sets. Despite being a perilous pedagogical exercise for the
+In this section, we will fulfil our promises. First, we will build
+finite sets. Second, we will implement the code of the $\Desc$
+universe.  Despite being a perilous pedagogical exercise for the
 authors, we hope to convey to the reader the dizzy feeling of
@@
 In Section~\ref{sec:finite-sets}, we have given a specification of
-finite sets. It consists of the objects presented in
-Figure~\ref{fig:typing-finite-set}. We are going to implement the
-$\EnumU$ type former and its constructors. Let us recall their
-specification:
+finite sets. We are going to implement the $\EnumU$ type former and
+its constructors. Let us recall their specification:
 
@@
 The names $\NilE$ and $\ConsE{\!}{\!}$ are not a coincidence. $\EnumU$
-is indeed isomorphic to a list of $\UId$s. In
-Section~\ref{sec:desc-examples}, we have shown how to build lists in
-the universe of description. Therefore, the actual implementation of
-$\EnumU$ boils down to the following, trivial definition:
+is indeed isomorphic to a list of $\UId$s. Therefore, the actual
+implementation of $\EnumU$ boils down to the following definition:
 
@@
 
-Let us examine the consequences. First of all, we discover that the
-type theory does not actually need to be extended with the type former
-$\EnumU$, the constructors $\NilE$ and $\ConsE$. We have just defined
-$\EnumU$ above, by fix-point over the pattern functor denoted by
+Let us consider the consequences. First, we discover that the type
+theory does not need to be extended with the type former $\EnumU$, the
+constructors $\NilE$ and $\ConsE$. We have just defined $\EnumU$
+above, by fix-point over the pattern functor denoted by
 $\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
@@
 it can actually be implemented with the generic $\F{induction}$
-principle. From an implementation perspective, this means that we
-remove an operator hard-coded in the host language. This operator is
-simply implemented in the dependently-typed target language, as any
-other function.
+principle. From an implementation perspective, this means that this
+operator is simply implemented in the dependently-typed target
+language, as any other function.
 
-Apart from these changes, we are left with implementing the other
-components of finite sets. This consists of $\EnumT$, $\Ze$, $\Su$, as
-well as the $\F{switch}$ operator. Note that the actual implementation
-of $\EnumU$ does not influence our implementation of $\EnumT$: be they
-hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects
-behave the same. We have witnessed this effect when carrying the
-operation in Epigram, moving from an hard-coded presentation to a
-self-hosted one. Absolutely no change to the $\EnumT$ objects was
-required.
+%% Apart from these changes, we are left with implementing the other
+%% components of finite sets. This consists of $\EnumT$, $\Ze$, $\Su$, as
+%% well as the $\F{switch}$ operator. Note that the actual implementation
+%% of $\EnumU$ does not influence our implementation of $\EnumT$: be they
+%% hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects
+%% behave the same. We have witnessed this effect when carrying the
+%% operation in Epigram, moving from an hard-coded presentation to a
+%% self-hosted one. Absolutely no change to the $\EnumT$ objects was
+%% required.
 
@@
 Let us explain how we have proceeded and the obstacle we face. Much as
-the datatypes we have seen so far, we first have to choose a
-constructor. Constructors correspond to the code's constructors:
+the datatypes we have seen so far, we first choose a constructor among
 $\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
 that the tagged notation we have used for the $\Desc$ constructors now
-fully makes sense: these were actually the tags we are
-defining.
+fully makes sense: these were actually the tags we are defining.
 
-The case of $\DUnit$ is trivial: we immediately reach the end of the
-datatype. $\DUnit$ has no payload. The case of $\DIndx{\!}$ comes
-naturally: it only takes a recursive argument -- a description. The
-case of $\DSigma{\!}{\!}$ is problematic. Recall the specification of
+For $\DUnit$, we immediately reach the end of the description. For
+$\DIndx{\!}$, there is a single recursive argument. The case of
+$\DSigma{\!}{\!}$ is problematic. Recall the specification of
 $\DSigma{\!}{\!}$:
@@
 Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
-we tolerate this duplication and do not replace $\DIndx{\!}$ by the
-more general $\DHindx{\!}{\!}$. Indeed, unlike its counterpart,
-$\DIndx{\!}$ is first-order. Therefore, while both codes are
-\emph{extensionally} the same, in most practical implementations they
-would be dealt with rather differently. While definitional equality
-can cope with first order objects, the functional presentation
-introduced by $\DHindx{\!}{}$ is unlikely to be amenable to a purely
-definitional treatment.
+we tolerate this duplication. Indeed, unlike its counterpart,
+$\DIndx{\!}$ is first-order. Both codes are \emph{extensionally} the
+same but, in practice, they will be dealt with rather
+differently. While definitional equality can cope with first order
+objects, the functional presentation introduced by $\DHindx{\!}{}$ is
+unlikely to be amenable to a purely definitional treatment.
 
-Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
-our universe of datatypes, as shown in
-Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
-remain unchanged, as expected. We successfully describe the
+As shown in Figure~\ref{fig:desc-levitate}, we are now able to
+describe our universe of datatypes.  The $\DUnit$ and $\DIndx{\!}$
+cases remain unchanged, as expected. We successfully describe the
 $\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
@@
 $\EnumT{\V{e}} \To \Mu{\DescD}$. This corresponds to a finite function
-definition, as formalised in Figure~\ref{fig:type-checking}.
+definition, as presented in Figure~\ref{fig:type-checking}.
 
@@
 
-If we cannot make it happen, we simply do not make it happen. We
-perfectly know what the motive ought to be. Consequently, we extend
+However, we perfectly know what the motive is. Consequently, we extend
 the type theory with a special-purpose operator:
@@
 
-The entire work of the magician stands here, in this extension. One
-could be worried by the availability of such operations in our type
-theory. Indeed, it would be inconvenient for the developer to have to
-decide between two variants. The solution to this problem is, once
-again, to use the type propagation system. Just as the $\F{switch}$
-operator, $\F{switchD}$ can be automatically elaborated. To this end,
-we extend the type-checking rules (Fig.~\ref{fig:type-checking}) with
-the following inference rule:
+The entire work of the magician stands here, in this extension. Note
+that the developer do not have to decide between the two variants of
+$\F{switch}$. This is dealt with by the type propagation system
+(Fig.~\ref{fig:type-checking}), extended by the following inference
+rule:
 
@@
 This concludes our levitation work. Beyond its pedagogical value, this
-exercise has several practical outcomes. First of all, this exercise
-reveals that the $\Desc$ universe is just plain data. Just as any
-piece of data, it can therefore be inspected and
-manipulated. Moreover, it is expressed in the $\Desc$ universe. As a
-consequence, it is equipped, for free, with an induction
-principle. So, our ability to inspect and program with $\Desc$ is not
-restricted to the meta-language: we now have all the necessary
-equipment in the target language to \emph{program} over datatypes. In
-this setting, \emph{generic programming is just about programming}.
+exercise has several practical outcomes. First of all, it reveals that
+the $\Desc$ universe is just plain data. Just as any piece of data, it
+can therefore be inspected and manipulated. Moreover, it is expressed
+in the $\Desc$ universe. As a consequence, it is equipped, for free,
+with an induction principle. So, our ability to inspect and program
+with $\Desc$ is not restricted to the meta-language: we now have all
+the necessary equipment in the target language to \emph{program} over
+datatypes. In this setting, \emph{generic programming is just 
+  programming}.
 
@@
 
-In Section~\ref{sec:desc-fix-point}, we have argued for the
-implementation of a dependently-typed $\F{induction}$ principle,
-instead of the more traditional catamorphism. However, in some
-circumstances, the full-power of a dependent elimination is not
+In Section~\ref{sec:desc-fix-point}, we have implemented a dependent
+$\F{induction}$ principle, instead of the catamorphism. However, in
+some circumstances, the full-power of a dependent elimination is not
 necessary. In the following, we propose to derive the catamorphism
-from the generic induction principle.
+from the $\F{induction}$ principle.
 
-The catamorphism can be characterised as an induction on the
-description $\V{D}$, with a non-dependent motive targeting $\V{T}$. Given a
-node $\V{xs}$ and the induction hypotheses, the method ought to build an
-element of $\V{T}$. Provided that we know how to make an element of
-$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra $\V{f}$. Let
-us take a look at this jigsaw:
+The catamorphism is an induction on the description $\V{D}$, with a
+non-dependent motive targeting $\V{T}$. Given a node $\V{xs}$ and the
+induction hypotheses, the method ought to build an element of
+$\V{T}$. Provided that we know how to make an element of
+$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
+$\V{f}$. Let us take a look at this jigsaw:
 
@@
 for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
-element in $\V{T}$. Hence the name ``induction
-hypotheses''. Therefore, to construct an element of
+element in $\V{T}$.  Therefore, to construct an element of
 \(\descop{\V{D}}{\V{T}}\), we replace the recursive components of
@@
 
-Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\: (\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the
-problem. In the type theory, we have built a generic catamorphism. Any
-datatype will now come equipped with this operation, for free.
+Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
+(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the problem. In the
+type theory, we have built a generic catamorphism. Any datatype will
+now come equipped with this operation, for free.
 
-The astute reader will have been struck by the type of $\F{replace}$:
-it is \emph{almost} the morphism part -- sometimes called \emph{map}
--- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
-the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
-is the dependent version of the map, which uses the induction
-hypotheses. For space reason, we will not present the non-dependent
-map. It can be found in the Agda model.
+%% The astute reader will have been struck by the type of $\F{replace}$:
+%% it is \emph{almost} the morphism part -- sometimes called \emph{map}
+%% -- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
+%% the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
+%% is the dependent version of the map, which uses the induction
+%% hypotheses. For space reason, we will not present the non-dependent
+%% map. It can be found in the Agda model.
 
@@
 basically, a function mapping a $\Desc$ to a datatype specific
-operation. This programming style is reminiscent to polytypic
-languages, such as PolyP~\cite{jansson:polyp}.
+operation. This is a form of polytypic programming, as found in
+PolyP~\cite{jansson:polyp}.
 
-Moreover, the $\F{replace}$ function demonstrates the benefit of an
-approach based on universes. The datatypes living in the universe of
-descriptions, we are able to \emph{inspect} them. As shown by
-$\F{replace}$, it is easy to explore these structures, as well as
-building new ones.
+%% Moreover, the $\F{replace}$ function demonstrates the benefit of an
+%% approach based on universes. The datatypes living in the universe of
+%% descriptions, we are able to \emph{inspect} them. As shown by
+%% $\F{replace}$, it is easy to explore these structures, as well as
+%% building new ones.
 
@@
 
-\note{Recall monadic structure of this object}
-
 Being an inductive type, this $\FMFreeMonad$ datatype is itself
@@
 
-In our setting, the functor will be represented by a tagged
-description. The free monad construction will take such tagged
-description, a set $\V{X}$ of variables, and will compute the tagged
-description of the corresponding free monad. Implementing this
+In our setting, the free monad construction will take the functor as a
+tagged description, a set $\V{X}$ of variables, and will compute the
+tagged description of the corresponding free monad. Implementing this
 function is surprisingly easy:
@@
 
-Our task is therefore to implement the algebra of the
+We are left with implementing the algebra of the
 catamorphism. Intuitively, its role is to catch appearances of
@@
 structure with the corresponding monadic operation, \bind\ and
-\return. Whereas the \return\ is a trivial variable former, the
-\bind\ operation corresponds to the substitution of variables by
-terms. We have shown how to implement it, for any free monad. This
-programming style is reminiscent to type-indexed
-datatypes~\cite{hinze:generic-haskell}, as implemented in Generic
-Haskell: from datatypes, we built new datatypes and equip them with
-their structure.
+\return. We have shown how to implement \bind, for any free
+monad. This construction is an example of type-indexed
+datatype~\cite{hinze:generic-haskell}, as found in Generic Haskell:
+from a datatype, we built a new datatype and equip it with its
+structure.
 
@@
 %% Candidate for removal:
-With the free monad construction, we have seen two kinds of generic
-operations. Firstly, we have derived a new data-structure from another
-one: we make the free monad from its underlying functor. To do so, we
-crucially rely on the fact that datatypes are nothing but codes. We
-are therefore entitled to modify this code and, in this case, extend
-it. Extending a datatype might give rise to a more structured object,
-as was the case here.  So, secondly, we have equipped this new
-datatype with its inherent structure: the \bind\ and
-\return\ operations. We have been able to build them as generic
-functions.
+%% With the free monad construction, we have seen two kinds of generic
+%% operations. Firstly, we have derived a new data-structure from another
+%% one: we make the free monad from its underlying functor. To do so, we
+%% crucially rely on the fact that datatypes are nothing but codes. We
+%% are therefore entitled to modify this code and, in this case, extend
+%% it. Extending a datatype might give rise to a more structured object,
+%% as was the case here.  So, secondly, we have equipped this new
+%% datatype with its inherent structure: the \bind\ and
+%% \return\ operations. We have been able to build them as generic
+%% functions.

### 2010-04-01 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: WeightWatch section 5, Indexing

@@
 <- Motivation
-    <- Desc: expressivity of simply-typed data-types: inductive types
+    <- Desc: expressivity of simply-typed datatypes: inductive types
         <- Values do not influence types
@@
 So far, we have explored the well-known realm of inductive types. We
-have built upon the experience gained in using simply-typed languages,
-such as Haskell or OCaml. In our dependent setting, we provided these
-data-types by the mean of $\Desc$, a universe of descriptions.
-
-While inductive types are the alpha and omega of simply-typed
-languages, evolving in a dependent setting fosters new
-opportunities. The typical example is bounded lists, also known as
+have built upon our intuition of ML-like datatypes. In our dependent
+setting, we have provided these datatypes by the mean of $\Desc$, a
+universe of descriptions.
+
+Evolving in a dependent setting fosters new opportunities for
+datatypes. The typical example is bounded lists, also known as
 vectors. A vector is a list decorated by its length. Having this
@@
 an empty vector: the head function only takes vectors of length
-$\NatSuc{\V{n}}$, as enforced by its type. This is made possible by the
-very specificity of dependently-typed systems: terms -- the length --
-are allowed to influence types -- the vector type. \note{Reference
-  paper to cite about vectors in dependent types?}
-
-However, this new class of objects cannot be defined by mere
-induction. In the case of vectors, for instance, we have to define the
-whole \emph{family} of vector in one go: vectors of all sizes need to
-be defined at the same time. Where simply-typed languages have
-inductive types, in dependently typed programming the basic grammar of
-data-types should be that of inductive families of types. To this end, we
-rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
-$\IDesc{\!}$, the universe of indexed descriptions. We express inductive families
-by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
-motivating example here, since it doesn't appear in the paper (more on that
-later). Come to think of it, this is entirely too vague without spelling out
-a concrete example.}
-
-\subsection{Desc, atomically}
-\label{sec:idesc-atomic-desc}
-
-\begin{wstructure}
-[Outdated: type former presentation instead]
-<- Adding hindx have introduced some duplication
-    <- indx == hindx 1
-    -> We can factor out commonalities 
-        /> Obtain an equivalent presentation
-        /> Still embeddable (refer to the Agda model)
-\end{wstructure}
-
-\begin{wstructure}
-<- Also replacing '1 by 'const  [figure]
-    <- For convenience
-        <- 'const X equivalent to 'sigma X (\_ -> '1)
-        /> Easier to abstract
-            <- Extensionally same
-            /> 'const more useful in practice
-\end{wstructure}
-
-Before moving on to indexed descriptions, we have to carry out some
-maintenance work on descriptions. We presented $\Desc$ as the grammar
-of inductive types. Hence, the codes closely follow this grammar. In
-the following, we adopt an alternative presentation. With
-$\DSigma{\!}{\!}$, we are actually \emph{quoting} a standard
-type-former, namely
-
-$$\Bhab{\Sigma}{\PI{\V{S}}{\Set} \PI{\V{S}}{\Set} \Set}$$
-
-In the alternative presentation, we go further and present all our
-codes as quotations of standard type-formers. This presentation is
-shown in Figure~\ref{fig:type-former-desc}. The reader will notice
-that we replace $\DUnit$ by a more general $\DConst{\!}$ code. Whereas
-$\DUnit$ was interpreted as the unit set, $\DConst{\V{X}}$ is
-interpreted as $\V{X}$, for any $\Bhab{\V{X}}{\Set}$. Extensionally,
-$\DConst{\V{X}}$ and $\DSigma{\V{X}}{\DUnit}$ are equivalent. However,
-$\DConst{\!}$ is more succinct. More importantly, $\DConst$ is
-\emph{first-order}, unlike its equivalent encoding. From a
-definitional perspective, we are giving more opportunities to the
-type-system, hence reducing the burden on the programmer. For the same
-reason, we introduce $\DProd{\!}{\!}$ that overlaps with
-$\DSigma{\!}{\!}$.
-
-This reorganisation is strictly equivalent to the previous one
-(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
-self-descriptive. We refer the reader to the companion technical
-report for details. In this finer-grained presentation, we can define
-$\DIndx{\!}$ and $\DHindx{\!}{\!}$ as follow:
-
-\[\begin{array}{l@{\:\mapsto\:\:}l}
-\DIndx{\V{D}}         & \DProd{\DId}{\V{D}}                      \\
-\DHindx{\V{H}}{\V{D}}     & \DProd{(\DPi{\V{H}}{(\LAM{\_} \DId)})}{\V{D}}
-\end{array}
-\]
-
-Consequently, the examples previously developed can be
-straightforwardly translated into this new presentation. For example,
-here is the new definition of $\NatD$:
-
-\[\stk{
-\NatD : \Desc \\
-\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
-                     {[ \DUnit \quad \DId ]}
-}\]
-
-In the following, we adopt this last version as our de
-facto universe of inductive types. In particular, we are going to
-evolve this presentation into an indexed one.
-
-\note{Shall we talk about the Type Theory being Desc Zero? or such story?}
-
-\begin{figure}
-
-\[\stk{
-\begin{array}{ll}
-\stk{
-\data \Desc : \Set \where                                      \\
-\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \DId            & \Desc                                    \\
-    \DConst{\!}     & \Set \To \Desc                           \\
-    \DProd{\!}{\!}  & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
-    \DSigma{\!}{\!} & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
-    \DPi{\!}{\!}    & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc 
-\end{array}
-}
-\vspace{0.2in}
-\\
-\stk{
-\descop{\_\:}{} : \Desc \To \Set \To \Set \\
-\begin{array}{@{}l@{\:=\:\:}ll}
-\descop{\DId}{\V{X}}          &  \V{X}                                           \\
-\descop{\DConst{\V{Z}}}{\V{X}}    &  \V{Z}                                           \\
-\descop{\DProd{\V{D}}{\V{D'}}}{\V{X}} &  \TIMES{\descop{\V{D}}{\V{X}}}{\descop{\V{D}\V{'}}{\V{X}}}       \\
-\descop{\DSigma{\V{S}}{\V{D}}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}          \\
-\descop{\DPi{\V{S}}{\V{D}}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}            
-\end{array}
-}
-\end{array}
-}\]
-
-\caption{Universe of descriptions based on Type-formers}
-\label{fig:type-former-desc}
-
-\end{figure}
+$\NatSuc{\V{n}}$, as enforced by its type. This is made possible by
+the specificity of dependent types: a term -- the length -- can
+influence a type -- the vector type.
+
+However, these datatypes cannot be defined by mere induction. In the
+case of vectors, for instance, we have to define the whole
+\emph{family} of vectors in one go: vectors of all sizes need to be
+defined at the same time. In dependently-typed languages, the basic
+grammar of datatypes is that of inductive families. To capture this
+grammar, we rely on \emph{indexing}. We are going to transform the
+$\Desc$ universe into $\IDesc{\!}$, the universe of indexed
+descriptions. We model inductive families by the $\V{I} \To
+\IDesc{\V{I}}$ type.
+
+%% \subsection{Desc, atomically}
+%% \label{sec:idesc-atomic-desc}
+
+%% \begin{wstructure}
+%% [Outdated: type former presentation instead]
+%% <- Adding hindx have introduced some duplication
+%%     <- indx == hindx 1
+%%     -> We can factor out commonalities 
+%%         /> Obtain an equivalent presentation
+%%         /> Still embeddable (refer to the Agda model)
+%% \end{wstructure}
+
+%% \begin{wstructure}
+%% <- Also replacing '1 by 'const  [figure]
+%%     <- For convenience
+%%         <- 'const X equivalent to 'sigma X (\_ -> '1)
+%%         /> Easier to abstract
+%%             <- Extensionally same
+%%             /> 'const more useful in practice
+%% \end{wstructure}
+
+%% Before moving on to indexed descriptions, we have to carry out some
+%% maintenance work on descriptions. We presented $\Desc$ as the grammar
+%% of inductive types. Hence, the codes closely follow this grammar. In
+%% the following, we adopt an alternative presentation. With
+%% $\DSigma{\!}{\!}$, we are actually \emph{quoting} a standard
+%% type-former, namely
+
+%% $$\Bhab{\Sigma}{\PI{\V{S}}{\Set} \PI{\V{S}}{\Set} \Set}$$
+
+%% In the alternative presentation, we go further and present all our
+%% codes as quotations of standard type-formers. This presentation is
+%% shown in Figure~\ref{fig:type-former-desc}. The reader will notice
+%% that we replace $\DUnit$ by a more general $\DConst{\!}$ code. Whereas
+%% $\DUnit$ was interpreted as the unit set, $\DConst{\V{X}}$ is
+%% interpreted as $\V{X}$, for any $\Bhab{\V{X}}{\Set}$. Extensionally,
+%% $\DConst{\V{X}}$ and $\DSigma{\V{X}}{\DUnit}$ are equivalent. However,
+%% $\DConst{\!}$ is more succinct. More importantly, $\DConst$ is
+%% \emph{first-order}, unlike its equivalent encoding. From a
+%% definitional perspective, we are giving more opportunities to the
+%% type-system, hence reducing the burden on the programmer. For the same
+%% reason, we introduce $\DProd{\!}{\!}$ that overlaps with
+%% $\DSigma{\!}{\!}$.
+
+%% This reorganisation is strictly equivalent to the previous one
+%% (Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
+%% self-descriptive. We refer the reader to the companion technical
+%% report for details. In this finer-grained presentation, we can define
+%% $\DIndx{\!}$ and $\DHindx{\!}{\!}$ as follow:
+
+%% \[\begin{array}{l@{\:\mapsto\:\:}l}
+%% \DIndx{\V{D}}         & \DProd{\DId}{\V{D}}                      \\
+%% \DHindx{\V{H}}{\V{D}}     & \DProd{(\DPi{\V{H}}{(\LAM{\_} \DId)})}{\V{D}}
+%% \end{array}
+%% \]
+
+%% Consequently, the examples previously developed can be
+%% straightforwardly translated into this new presentation. For example,
+%% here is the new definition of $\NatD$:
+
+%% \[\stk{
+%% \NatD : \Desc \\
+%% \NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
+%%                      {[ \DUnit \quad \DId ]}
+%% }\]
+
+%% In the following, we adopt this last version as our de
+%% facto universe of inductive types. In particular, we are going to
+%% evolve this presentation into an indexed one.
+
+%% \note{Shall we talk about the Type Theory being Desc Zero? or such story?}
+
+%% \begin{figure}
+
+%% \[\stk{
+%% \begin{array}{ll}
+%% \stk{
+%% \data \Desc : \Set \where                                      \\
+%% \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
+%%     \DId            & \Desc                                    \\
+%%     \DConst{\!}     & \Set \To \Desc                           \\
+%%     \DProd{\!}{\!}  & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
+%%     \DSigma{\!}{\!} & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
+%%     \DPi{\!}{\!}    & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc 
+%% \end{array}
+%% }
+%% \vspace{0.2in}
+%% \\
+%% \stk{
+%% \descop{\_\:}{} : \Desc \To \Set \To \Set \\
+%% \begin{array}{@{}l@{\:=\:\:}ll}
+%% \descop{\DId}{\V{X}}          &  \V{X}                                           \\
+%% \descop{\DConst{\V{Z}}}{\V{X}}    &  \V{Z}                                           \\
+%% \descop{\DProd{\V{D}}{\V{D'}}}{\V{X}} &  \TIMES{\descop{\V{D}}{\V{X}}}{\descop{\V{D}\V{'}}{\V{X}}}       \\
+%% \descop{\DSigma{\V{S}}{\V{D}}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}          \\
+%% \descop{\DPi{\V{S}}{\V{D}}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}            
+%% \end{array}
+%% }
+%% \end{array}
+%% }\]
+
+%% \caption{Universe of descriptions based on Type-formers}
+%% \label{fig:type-former-desc}
+
+%% \end{figure}
 
@@
 $\Set^{\V{I}}$. We call these \emph{indexed functors}. \note{Predicate
-  transformer anyone?} In this setting, $\V{I} \To \IDesc{\V{I}}$ is our
+  transformer anyone?} $\V{I} \To \IDesc{\V{I}}$ is our
 grammar for describing these functors. Hence, $\IDesc{\!}$ and its
@@
 $\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
-\To \Set^{\V{I}}$. Fix-point of inductive families is defined over
+\To \Set^{\V{I}}$.The fix-point of inductive families is defined over
 these indexed functors, hence computing a fix-point of the entire
@@
 
-However, we still have to define the actual grammar. This can be
-achieved by a minor evolution of $\Desc$
-(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of
-$\Desc$ is entirely captured by the $\DId$ code. We get an indexed
-functor by turning $\DId$ into an index-dependent $\DVar{\!}$
-code. The resulting code of $\IDesc{\!}$ is presented in
-Figure~\ref{fig:idesc}.
-
-The induction scheme on $\IDesc{\!}$ has the following type:
+However, we still have to define the actual grammar. We obtain it by
+evolving $\Desc$ to cope with indexing. The step-by-step
+transformation is described in our technical report. The code of
+$\IDesc{\!}$ is presented in Figure~\ref{fig:idesc}. The induction
+scheme on $\IDesc{\!}$ has the following type:
 
@@
 %<- Tagged constructor choice
-%    <- Index available when defining the data-type
+%    <- Index available when defining the datatype
 %        -> Can influence the choice of constructors
-%        -> Dependently-typed data-types
+%        -> Dependently-typed datatypes
 %            <- term (index) influence types
@@
 %definition can be generalized. Indeed, when defining an indexed
-%data-type, we have access to this index. Therefore, we can use this
+%datatype, we have access to this index. Therefore, we can use this
 %index to influence the choice of constructors. This captures the
-%essence of dependent data-types: a term -- the index -- has the
-%ability to influence the data-type.
+%essence of dependent datatypes: a term -- the index -- has the
+%ability to influence the datatype.
 %
@@
 
-In order to gain some intuition with $\IDesc{\!}$, let us re-implement
-the pattern functor of natural numbers. This follows the implementation of
-$\NatD$ from \ref{sec:idesc-atomic-desc}:
+In order to gain some intuition of $\IDesc{\!}$, let us re-implement
+the pattern functor of natural numbers:
 
@@
 
-Because $\Nat$ is just an inductive type, $\NatD$ represents a
-$\Unit$-indexed functor. Therefore, the recursive
-argument is materialised by $\DVar{\Void}$, where we were using $\DId$
-in the $\Desc$ presentation. Indeed, all inductive types can be
-translated into $\IDesc{\!}$ in exactly this way.
+Because $\Nat$ is just an inductive type, $\NatD$ is a $\Unit$-indexed
+functor. Therefore, the recursive argument is materialised by
+$\DVar{\Void}$, where we were using $\DIndx$ in the previous
+presentation. This transformation generalizes to all inductive
+types. Moreover, we gain the ability to write mutually recursive
+inductive types.
 
@@
     <- Following Desc encoding
-        /> Note: simple data-type
+        /> Note: simple datatype
             -> Live in IDesc 1
@@
 
-If we note that, for any $\V{I}$, $\IDesc{\V{I}}$ itself is merely an inductive 
-\emph{type} then we can certainly define its encoding in $\IDesc{\Unit}$:
+Note that $\IDesc{\V{I}}$ itself is merely an inductive type. Hence,
+we can describe it in $\IDesc{\Unit}$:
 
@@
 
-Using this description we can repeat the act of levitation 
-from section~\ref{sec:desc-levitate}. As before we rely on a special purpose 
-$\F{switchID}$ operator to build the finite function space
-$\bigRedBracket{\ldots}$.
+Therefore, this universe is self-describing, hence can be
+levitated. As before, we rely on a special purpose $\F{switchID}$
+operator to build the finite function space $\bigRedBracket{\ldots}$.
 
@@
 So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
-not using any indexing. We remedy this by encoding the vectors:
-
-Recall that the constructors of $\D{Vec}$. The constructors $\VecNil$ and
-$\VecCons$ are only defined for an index $\NatZero$ and $\NatSuc$ 
-respectively.:
+not using any indexing. We remedy this by encoding the vectors. Recall
+that the constructors $\VecNil$ and $\VecCons$ are only defined for an
+index $\NatZero$ and $\NatSuc$ respectively.:
 
@@
 
-One way to code constrained data-type is to appeal to equality. The
-constraints are therefore captured by equations in the data-type. In
+One way to code constrained datatype is to appeal to equality. The
+constraints are therefore captured by equations in the datatype. In
 this case, we obtain the following definition:
@@
 
-If the $\VecNil$ tag is chosen, a proof must be provided that the index is equal
-to $\NatZero$. In the $\VecCons$ case, we first store an
-element $\V{m}$ of $\Nat$, with a $\DSigma{\!}{\!}$. However, the
-constraint stipulates that $\V{m}$ cannot be \emph{any} natural
-numbers: it must be ``the index minus one''. This translates into the
-constraint $\V{n} \PropEq \NatSuc{\V{m}}$.
+In the $\VecNil$ case, a proof must be provided that the index is
+equal to $\NatZero$. In the $\VecCons$ case, we first store an element
+$\V{m}$ of $\Nat$. However, the constraint stipulates that $\V{m}$
+cannot be \emph{any} natural numbers: it must be ``the index minus
+one''. This translates into the constraint $\V{n} \PropEq
+\NatSuc{\V{m}}$.
 
@@
 haunt us, begging for proofs. Hence, whereas the developer typed a
-constructor-constrained data-type, the elimination form will involve
-equalities and their proofs. As much as possible, one would rather
+constructor-constrained datatype, the elimination form will involve
+equalities and their proofs. As much as possible, one would like to
 avoid suffering from this burden. However, the equalities introduced
@@
 incoming index, we can apply the \emph{forcing} and \emph{de-tagging}
-optimisations to our initial definition of $\D{Vec}$ gives the
+optimisations to our initial definition of $\D{Vec}$. This gives the
 following, equivalent definition:
@@
 
-The equations (and constructors) have simply disappeared. Note that it is
-straightforward to turn this definition into a tagged form, where in each
-case, only one constructor choice is available to the developer. In another
-example, $\Fin{\!}$, we can force, but not detag, since both $\FinZero$ and 
-$\FinSuc{\!}$ both target $\NatSuc{\!}$:
+The equations (and constructors) have simply disappeared. A similar
+example is $\Fin{\!}$, specified by:
 
@@
 \end{array}
-\vspace{0.1in}
-\\
+}\]
+
+In this case, we can apply forcing, but not detagging, since both
+$\FinZero$ and $\FinSuc{\!}$ both target $\NatSuc{\!}$:
+
+\[\stk{
 \FinD : \Nat \To \IDesc{\Nat} \\
@@
 
-We should
-precise that forcing a description is not guaranteed to remove all
-constraints. It is subject to future work to see if constraints can be
-entirely eradicated, or presented more conveniently to the developer.
-Finally, it is worth mentioning that these optimisations are 
-\emph{source-to-source} transformations on descriptions.
+We should precise that forcing a description is not guaranteed to
+remove all constraints. It is subject to future work to see if
+constraints can be entirely eradicated, or presented more conveniently
+to the developer.  Finally, it is worth mentioning that these
+optimisations are \emph{source-to-source} transformations on
+descriptions.
 
@@
 
-In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
-form for descriptions. This format follows the usual presentation of
-inductive types as sum-of-product. With indexed description, this
-definition can be generalized. Indeed, when defining an indexed
-data-type, we have access to this index. Therefore, we can use this
-index to influence the choice of constructors. This captures the
-essence of dependent data-types: a term -- the index -- has the
-ability to influence the data-type.
+When defining an indexed datatype, we have access to its
+index. Therefore, we can use this index to influence the choice of
+constructors. This captures the essence of dependent datatypes: a
+term -- the index -- has the ability to influence the datatype. We
+define tagged indexed descriptions to capture this specificity.
 
@@
 constructors that do. The non-dependent part mirrors the definition
-for non-indexed descriptions: we are provided a finite choice of
-constructors. The index-depend part simply indexes the choice of
-constructors by $\V{I}$. Hence, by inspecting the index, it is
-possible to enable or disable the constructors.
+for non-indexed descriptions. The index-depend part simply indexes the
+choice of constructors by $\V{I}$. Hence, by inspecting the index, it
+is possible to enable or disable the constructors.
 
@@
 
-In the case of a tagged $\D{Vec}$, for instance, for the index $\NatZero$, we
-would only propose the constructor $\ListNil$. Similarly, for
-$\NatSuc{n}$, we would only propose the constructor $\ListCons{\!}{\!}$. We
-will see a concrete example of this in the following example.
+In the case of a tagged $\D{Vec}$, for instance, for the index
+$\NatZero$, we would only propose the constructor
+$\ListNil$. Similarly, for $\NatSuc{n}$, we would only propose the
+constructor $\ListCons{\!}{\!}$.
 
@@
 An expression of this language is either a value, a conditional
-expression, addition of numbers, or comparison of numbers. Informally,
-their type is the following:
+expression, an addition of numbers, or a comparison of
+numbers. Informally, their type is the following:
 
@@
 
-The function $\Val{}$, used in the definition of $\Eval{}$, simply
-maps an object type $ty$ to the corresponding type in the host
-language. Hence, the argument of $\Eval{}$ are ensured to be of the
-expected type. We assume $\Nat$ and $\Bool$ represent natural numbers
-and booleans in the host language. Further, we also assume the
-existence of an addition and comparison operator in $\Nat$,
-respectively named $\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$
-as follows:
+The function $\Val{\!}$, used in the definition of $\Eval{\!}$, simply
+maps a type $ty$ of the object language to the corresponding type in
+the host language. Hence, the arguments of $\Eval{}$ are ensured to be
+of the expected type. We assume that $\Nat$ and $\Bool$ represent
+natural numbers and booleans in the host language, equipped with an
+addition operation $\F{plusHost}$ and a comparison function
+$\F{leHost}$. We define $\Val{}$ as follows:
 
@@
 
-In our universe of descriptions, this data-type is represented by a
+In our universe of descriptions, this datatype is represented by a
 tagged indexed description. We use the index to carry the type: the
 resulting description is indexed by $\Ty$. We observe that some
-constructors are ``polymorphic'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
+constructors are always defined, namely $\Econd{}{}{}$ and $\Eval{}$. On the
 other hand, the $\Eplus$ and $\Ele$ constructors are
@@
 indexed description computed from the tagged indexed description
-$\V{TID}$.
-
-The type of $\F{eval}$ is strikingly similar to a
+$\V{TID}$. The type of $\evalH$ is strikingly similar to a
 catamorphism. Indeed, implementing a single step of evaluation -- the
@@
 }\]
-\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of
-  IFunc floating around, $\idescop{\_}{}{}$ is defined using the 1st
-  notion, and used it here as if it were the 2nd. pierre: in this
-  case, it was a typo. But you're right, there is two notions and we
-  probably want to clarify where these things live.}
 
@@
 Hence, we have defined the syntax of a typed language of arithmetic
-and boolean operations. We have given its semantics through an
+and boolean expressions. We have given its semantics through an
 evaluation function. Provided a one step semantic of the language, the
-big step evaluation is granted without effort thanks to the generic
+big step interpreter is granted without effort thanks to the generic
 catamorphism.
@@
 \emph{closed} terms. By abstracting over $\Val{}$, it is possible to
-build and manipulate \emph{open} terms, that is, terms with symbolic
+build and manipulate \emph{open} terms, that is, terms with
 variables. On the model of $\Val{}$, we define $\Var{}{}$:
@@
 contains a pointer to a particular element of the finite set -- an
-element of $\EnumT{\V{dom}}$. The extra argument to $\Var{}{}$ is the
-domain of this context.
+element of $\EnumT{\V{dom}}$.
 
@@
 in the context; then, we evaluate the resulting closed term. Thanks to
-$\evalH$, we are already able to solve the second problem. Let us
-focus on discharging variables from the context. Again, we can
-subdivide this problem: first, we need the ability to discharge a
-single variable from the context; then, we apply this $\discharge$
-function on every variables in the term.
-
-The $\discharge$ function is relative to the required type, the domain
-of the context, and a context containing values of the corresponding
-type. Its action is to map variables to their value in context, and
-directly return constant values. This corresponds to the following
-function:
+$\evalH$, the second problem is already solved. Let us focus on
+discharging variables from the context. Again, we can subdivide this
+problem: first, discharging a single variable from the context; then,
+applying this $\discharge$ function on every variables in the term.
+
+The $\discharge$ function is relative to the required type and a
+context of the right type. Its action is to map values to themself,
+and variables to their value in context. This corresponds to the
+following function:
 
@@
 
-Having implemented the local $\discharge$ operation, we are left with
-applying it over all variables of the term. The type of this operation
-is the following:
+We are now left with applying $\discharge$ over all variables of the
+term. The type of this operation is the following:
 
@@
 
-Abstracting away the book-keeping introduced by the context, this
-definition looks familiar. Indeed, it is extremely similar to a
-monadic \bind. This is not surprising as we are defining a first-order
-syntax with variables: our data-type enjoys more structure than what
-we are given. In particular, this is reminiscent of a free monad,
-where $\Eval{}$ is the \return\ introducing variables. The substitution
-$\V{\sigma}$ is implemented from $\discharge$, by picking the domain and
-context corresponding to the type:
+Abstracting away the book-keeping introduced by contexts, this
+definition looks familiar. It is similar to a monadic \bind. This is
+not surprising as we are defining a first-order syntax with variables:
+our datatype enjoys more structure than what we are given. We are
+facing a free monad, where $\Eval{}$ is the \return\ introducing
+variables. For convenience, we wrap $\discharge$ in a $\V{\sigma}$
+function that picks the context of the right type:
 
@@
 
-Instead of implementing $\substH$ in this special case, we can
-implement it in a generic setting, a \emph{free indexed-monad construction}
-that we will see in the next section.
+Instead of implementing $\substH$ in this special case, we are now
+going to implement the \emph{free indexed-monad construction}.
 
@@
 
-In this section, we develop more structure for our indexed
-data-types. In Section~\ref{sec:desc-free-monad}, we have built a free
-monad operation for simple descriptions. Building on this experience,
-we are going to present its equivalent in the indexed world. The
-process is the same. Namely, given an indexed functor, we derive the
-indexed functor coding its free monad: \note{pwm: Whoa there. Maybe we
-  should say something about IMonads in general before we get to this
-  point?}
+In Section~\ref{sec:desc-free-monad}, we have built a free monad
+operation for simple descriptions. The process is similar in the
+indexed world. Namely, given an indexed functor, we derive the indexed
+functor coding its free monad: \note{pwm: Whoa there. Maybe we should
+  say something about IMonads in general before we get to this point?}
 
@@
 
+We refer the reader to the Agda model for the actual implementation.
+
 \subsubsection{Examples}
@@
 In Section~\ref{sec:idesc-examples}, we had the intuition that our
-datatypes $\HExprD$ and $\HExprVarD{\V{dom}}$ enjoy a monadic structure. We had
-identified the variable substitution operation as the \bind\ of a free
-monad. The definition of $\substI$ above confirms our intuition: using
-$\substI$, we should easily obtain $\substH$. To do so, we first have
-to massage the definition of our data-type, to exhibit its monadic
-structure.
-
-As previously mentioned, we identify $\Eval{}$ as the \return\ of the
-free monad, while the other components are the action of the monad. As
-a result, the definition is similar to $\HExprD$ presented in
-Figure~\ref{fig:hexpr-full}, replacing $\HExprAD$ with $\HExprAFreeD$:
+datatypes $\HExprD$ and $\HExprVarD{\V{dom}}$ enjoy a monadic
+structure. We had identified the variable substitution operation as
+the \bind\ of a free monad. To exhibit its monadic structure, we first
+have to massage the definition of our datatype.
+
+As previously mentioned, we identify $\Eval{}$ with the \return\ of
+the free monad, while the other components are the action of the
+monad. As a result, the definition is similar to $\HExprD$ presented
+in Figure~\ref{fig:hexpr-full}, replacing $\HExprAD$ by
+$\HExprAFreeD$:
 
@@
 
-The resulting data-type is called $\HExprFreeD$. By a simple unfolding
-of definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
+We call this datatype $\HExprFreeD$. By a simple unfolding of
+definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
 corresponds to the syntax of closed terms, $\HExprD$. Similarly,
@@
 terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}\: \V{ty}$ to values in
-$\Val{\V{ty}}$. We are therefore left with implementing $\substH$. We
-simply have to fill in the right arguments to $\substI$, the type
-guiding us:
+$\Val{\V{ty}}$. We are left with implementing $\substH$. We simply
+have to fill in the right arguments to $\substI$, the type guiding us:
 
@@
 
-Hence, we have completed our implementation of the evaluator for open
-terms. We started with a well-typed language of arithmetical
-expressions. We have seen how to take advantage of indexes to define
-type-safe constructors. Then, we have implemented an evaluator for
-closed term, based on the generic catamorphism function. Having built
-an open term representation, we wanted a substitution operation, in
-order to close open terms in a context. Again, we have implemented
-this operation with the generic substitution operator. Hence, without
-much efforts, we have described the syntax of a well-typed language,
-together with its semantics.
+Hence completing our implementation of the open terms interpreter. We
+have defined a well-typed language of arithmetical expressions, taking
+advantage of indexing. Then, we have implemented an evaluator for
+closed term, based on the generic catamorphism function. Using the
+free monad construction, we have automatically derived the language of
+open terms. Using its monadic structure, we have implemented the
+interpreter for open terms in context. Hence, without much efforts, we
+have described the syntax of a well-typed language, together with its
+semantics.
 
@@
 $\DVar{\!}$ is nothing but the \return. The remaining constructors are
-simply the pattern functor, trivially indexed by $\Unit$. Hence, we
-describe this functor by the following code:
+the carrier functor, trivially indexed by $\Unit$. The carrier functor
+is described as follow:
 
@@
 }\]
+
+In this section, we have presented the universe of indexed
+description. It embraces indexed families of types and, as such,
+allows us to write dependent datatypes. Hence, we have presented
+several example of indexed datatypes. In this context, we have
+presented the free monad construction, together its monadic
+operations.

### 2010-04-01 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: WeightWatch section 6, Discussion

@@
 
-As presented, our type theory suffers from a major weakness. Indeed,
-we are subject to Girard's paradox, as we assume that $\Set$ lives in
-$\Set$. We made that choice for presentational convenience, as
-universe stratification is orthogonal to our work. Nonetheless, our
-universe of description rather naturally leads itself to
-stratification. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of type
-$\Set^{\blue{l+1}}$. Similarly, the interpretation of $\IDesc{\!}$ at
-level $l$ is an object of type $\Set^{\blue{l}}$:
+As such, our type theory suffers from an inconsistency. Indeed, the
+typing rule $\Bhab{\Set}{\Set}$ leads to Girard's paradox. We made
+that choice for presentational convenience, as universe stratification
+is orthogonal to our work. Nonetheless, our universe of description
+stratifies naturally. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of
+type $\Set^{\blue{l+1}}$. Similarly, the interpretation of
+$\IDesc{\!}$ at level $l$ is an object of type $\Set^{\blue{l}}$:
 
@@
 The operations and examples developed in this paper stratify just as
-well. We refer the reader to our Agda model, which takes advantage of
-set polymorphism to implement the universe of indexed descriptions at
-any level. Further, we have coded $\IDesc{\!}$ in itself and have
-proved the isomorphism between the host and the embedded universes.
+well. We refer the reader to our Agda model. Further, we have coded
+$\IDesc{\!}$ in itself and have proved the isomorphism between the
+host and the embedded universes.
 
@@
     ???
-\end{structure}
+\end{wstructure}
 
@@
 \citet{garcia:generic-comparative-study} for a broad overview of
-generic support in various languages. In the sole context of Haskell,
-there is a myriad of proposals. These approaches are presented and
-compared in \citet{hinze:generic-approach-comparative} and
+generic programming in various languages. In the sole context of
+Haskell, there is a myriad of proposals. These approaches are compared
+in \citet{hinze:generic-approach-comparative} and
 \citet{rodriguez:generic-libs-comparative}.
 
-In particular, our approach is similar in spirit to polytypic
-programming, as initiated by PolyP~\cite{jansson:polyp}. Indeed,
-generic functions, in our system, are built by induction on the
-pattern functor. Unlike PolyP, we do not have to pre-process datatype
-definitions: our datatypes are, natively, nothing but codes.
+Our approach is follow the polytypic programming style, as initiated
+by PolyP~\cite{jansson:polyp}. Indeed, we build generic functions by
+induction on pattern functors. Unlike PolyP, we do not have to resort
+to preprocessing: our datatypes are, natively, nothing but codes.
 
@@
 \emph{generic views}~\cite{holdermans:generic-view}, transparently
-transforming the structure of datatype definitions, to ease the
-implementation of generic functions over them. Views in
-dependently-typed system are widely
-used~\cite{mcbride.mckinna:view-from-the-left}. Unlike Generic
-Haskell, we do not have to modify the compiler to develop views on
-datatypes: we can massage descriptions from inside the language. An
-example of such operation is the tagged descriptions, presenting
-datatypes under a sum-of-sigmas angle.
+transforming the structure of datatype definitions. An example is the
+tagged descriptions, presenting datatypes under a sum-of-sigmas
+angle. Unlike Generic Haskell, we do not have to modify the compiler
+to obtain views on datatypes: we can massage descriptions from inside
+our language.
 
@@
 programming~\cite{hinze:polytypic-polykinded}. Our descriptions are
-limited to endofunctor on $\Set$ and $\Set^I$. While it is possible to
-\emph{encode} higher-kinded datatypes in our setting, we do not plan
-to adopt this strategy. Rather, as future work, we plan to extend our
-universe to capture higher-kinded definitions and generic functions
-over them. For the same reason, arity-generic
-programming~\cite{weirich:arity-generic} is out of reach of our
-current presentation.
+limited to endofunctors on $\Set$ and $\Set^I$. While we could
+\emph{encode} higher-kinded datatypes, we do not plan to adopt this
+strategy. As future work, we plan to extend our universe to capture
+higher-kinded definitions and generic functions over them. For the
+same reason, arity-generic programming~\cite{weirich:arity-generic} is
+out of reach of our current presentation.
 
-Another generic programming framework is Scrap Your
+Another generic programming paradigm is Scrap Your
 Boilerplate~\cite{spj:syb} (SYB). Our proposal is different in various
@@
 constructors. A piece of data is a spine composed by a constructor
-applied to some arguments. Further, this spine is equipped with some
-combinators including, primarily, an iterator. In this setting,
-generic programs are written by composing these combinators, building
-generic operations on spines. This relies on a $\CN{Typeable}$
+applied to some arguments. SYB provides a combinator library to write
+generic functions over spines. This relies on a $\CN{Typeable}$
 type-class, allowing dynamic dispatch to datatype-specific
@@
 
-The interest for generic programming in dependent types is not new
+Generic programming in dependent types is not new
 either. \citet{norell:msc-thesis} has given a formalization of
-polytypic programming in the setting of Alfa, a precursor of
-Agda. Similarly, \citet{verbruggen:polytype-prog-coq,
-  verbruggen:polytype-coq} have developed a framework for polytypic
-programming in the Coq theorem prover. However, these works aim at
-\emph{modelling} PolyP or Generic Haskell in a dependently-typed
-setting, for the purpose of proving correctness properties of Haskell
-code. Our approach is different in that we aim at building a
-foundation for datatypes, in a dependently-typed theory, for a
-dependently-typed system.
+polytypic programming in Alfa, a precursor of Agda. Similarly,
+\citet{verbruggen:polytype-prog-coq, verbruggen:polytype-coq} have
+developed a framework for polytypic programming in the Coq theorem
+prover. However, these works aim at \emph{modelling} PolyP or Generic
+Haskell in a dependently-typed setting for the purpose of proving
+correctness properties of Haskell code. Our approach is different in
+that we aim at building a foundation for datatypes, in a
+dependently-typed system, for a dependently-typed system.
 
-Closer to us is the work by \citet{benke:universe-generic-prog}. This
+Closer to us is the work of \citet{benke:universe-generic-prog}. This
 seminal work introduced the usage of universes for developing generic
-programs. Our own universes are rather similar to theirs: our universe
-of descriptions is similar to their universe of iterated induction,
-and our universe of indexed descriptions is isomorphic to their
-universe of finitary indexed induction. This is not surprising, as we
-share the same source of inspiration, namely induction-recursion.
+programs. Our universes share similarities to theirs: our universe of
+descriptions is similar to their universe of iterated induction, and
+our universe of indexed descriptions is equivalent to their universe
+of finitary indexed induction. This is not surprising, as we share the
+same source of inspiration, namely induction-recursion.
 
-However, we differ in several ways. First, they adopt a generative
-perspective: each universe extends the base type theory with both type
-formers and elimination rules. Thanks to the levitation, we only rely
-on a generic induction and a specialised
-$\switchD{\!}{\!}{\!}$. Second, the authors do not tackle the issue of
-\emph{programming} with such codes: it is not obvious, to us, how one
-can define, manipulate, and program over these coded datatypes. With
-descriptions, we have shown how to abstract away codes and present a
-convenient and familiar presentation to the developer. Finally, the
-authors often resort to an extensional notion of equality, while we
-have given an equality-agnostic presentation. Beside, our presentation
-is arranged so as to use definitional equality as much as
-possible. Hence, in practice, the developer is relieved from many
-proof obligations.
+However, we differ in several ways. First, there approach is
+generative: each universe extends the base type theory with both type
+formers and elimination rules. Thanks to levitation, we only rely on a
+generic induction and a specialised $\switchD{\!}{\!}{\!}$. Second,
+the authors do not tackle the issue of \emph{programming} with
+codes. We have shown how to abstract away codes and give a convenient
+presentation to the developer. The authors often resort to an
+extensional equality, while we have given an equality-agnostic
+presentation. Beside, our universes are arranged so as to use
+definitional equality as much as possible. Hence, in practice, the
+developer is relieved from many proof obligations.

### 2010-04-01 papers/icfp-2010-desc/paper_type_theory.tex
got the three line widows

@@
 
-The judgmental equality comprises these computational rules
-\input{figure_judgemental_equality} closed under reflexivity,
+The judgmental equality comprises the computational rules below,
+closed under reflexivity,
 symmetry, transitivity and structural congruence, even under binders.
 We omit the mundane rules which ensure these closure properties for
-reasons of space. Given a suitable stratification of \(\Set\), the
+reasons of space.
+\input{figure_judgemental_equality}
+Given a suitable stratification of \(\Set\), the
 computation rules yield a terminating evaluation procedure, ensuring

### 2010-04-01 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: update TODO list.

@@
 
-We share with Generic Haskell the \emph{type-indexed data type}
+We share with Generic Haskell the \emph{type-indexed datatype}
 approach~\cite{hinze:generic-haskell}, as exemplified by the free

### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: clean-up some \note{}s.

@@
 technique in dependently-typed
-languages~\cite{benke:universe-generic-prog, oury:power-of-pi}
-\note{Any idea of other papers?}.  \note{Is that really a motivating
-  motivation for the usage of universe construction?}  \note{pwm:
-  Surely ($\EnumU$, $\EnumT$) is a useful example of universes too.
-  Even if it doesn't show off their full power?}
+languages~\cite{benke:universe-generic-prog, oury:power-of-pi}.
 
@@
 theory. Hence, codes act as labels, while the type theory provides the
-computational machinery. \note{Is that a good description?}
+computational machinery.
 

### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: s/sum-of-product/sum-of-products/

@@
 
-In ML-like languages, datatype definitions are called
-\emph{sum-of-product}. A datatype is defined by a finite sum of
+In ML-like languages, datatypes are presented as a
+\emph{sum-of-products}. A datatype is defined by a finite sum of
 constructors, each composed by a product of arguments. To embrace
@@
 
-With dependent types, the notion of sum-of-product translates into
+With dependent types, the notion of sum-of-products translates into
 \emph{sigmas-of-sigmas}. However, a direct encoding of datatypes
@@
 
-Again, this is a clear reminiscence of the sum-of-product style: we
+Again, this is a clear reminiscence of the sum-of-products style: we
 have a finite sum of constructors. Every description can be expressed

### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: s/fix-point/fixpoint/

@@
 constructors, each composed by a product of arguments. To embrace
-these datatypes, we need to capture this grammar.
+these datatypes, we have to capture this grammar.
 
@@
 theory. Hence, codes act as labels, while the type theory provides the
-computational machinery.
-
-Codes being mere labels, we can inspect them, hence regaining
-structural information. We can also compute over them: deriving new
-codes from others, or even functions on them.
+computational machinery. Codes being mere labels, we can inspect them,
+hence regaining structural information. We can also compute over them:
+deriving new codes from others, or even functions on them.
 
@@
 payload. To make recursive arguments, we introduce the $\DIndx{\!}$
-code. It interprets to a hole that will be closed by the fix-point. We
+code. It interprets to a hole that will be closed by the fixpoint. We
 can also read $\V{D}$ as describing an endofunctor on $\Set$, with
@@
 
-\subsection{Fix-point}
+\subsection{Fixpoint}
 \label{sec:desc-fix-point}
@@
 \begin{wstructure}
-<- Build the fix-point of functors
+<- Build the fixpoint of functors
     <- See examples: need to build their initial algebra
     -> Extend the type theory with Mu/Con [figure]
-        <- Straightforward definition of a fix-point
+        <- Straightforward definition of a fixpoint
             <- Interpret D with (Mu D) as sub-objects
@@
 So far, we have used our universe of descriptions to build pattern
-functors. Being polynomial functors, they all admit a fix-point. We
-construct the fix-point by \emph{tying the knot} of the
+functors. Being polynomial functors, they all admit a fixpoint. We
+construct the fixpoint by \emph{tying the knot} of the
 interpretation: the holes left open by the functor are filled by its

### 2010-04-01 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: clean-up some \note{}s.
Desc@ICFP: s/fix-point/fixpoint/

@@
 constructors $\NilE$ and $\ConsE$. We have just defined $\EnumU$
-above, by fix-point over the pattern functor denoted by
+above, by fixpoint over the pattern functor denoted by
 $\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
@@
 type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
-now replaced by codes and their fix-point in the universe of
+now replaced by codes and their fixpoint in the universe of
 descriptions. In essence, our next step is similar: we are going to
@@
 At a first glance, we have achieved our goal. We have described the
-codes of the universe of description. Taking the fix-point of this
+codes of the universe of description. Taking the fixpoint of this
 object gives us $\Desc$, up to isomorphism. We have implemented our
@@
 \end{wstructure}
-
-\note{
-Maybe a bit of perspective would be good: 
-Everything stands up with the Mu.
-Induction principle defined over Mu-things.
-}
 
@@
            [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
-        -> Mu does the fix-point
+        -> Mu does the fixpoint
 \end{wstructure}
@@
 convince ourselves that this corresponds to the functor
-$\FMFreeMonadD$. The fix-point operation ties the knot and gives us
+$\FMFreeMonadD$. The fixpoint operation ties the knot and gives us
 the full-blown free monad construction.

### 2010-04-01 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: clean-up some \note{}s.
Desc@ICFP: s/fix-point/fixpoint/

@@
 \end{wstructure}
-
-\note{ Need care: motivating motivation of indexing. }
 
@@
 $\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
-\To \Set^{\V{I}}$.The fix-point of inductive families is defined over
-these indexed functors, hence computing a fix-point of the entire
+\To \Set^{\V{I}}$.The fixpoint of inductive families is defined over
+these indexed functors, hence computing a fixpoint of the entire
 \emph{family} of functors:

### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: fixing typos (mostly in section 3, universe of inductive types_

@@
 To give some intuition on this universe, we now turn to some
-examples. For obvious pedagogical reasons, we will manually build
-these descriptions. However, it should be clear that, in practice,
-these definitions are automatically computed from an ML-like $\data$
+examples. For obvious pedagogical reasons, we will manually build the
+descriptions. However, it should be clear that, in practice, these
+definitions are automatically computed from an Agda-like $\data$
 definition.
@@
 $\ListCons{\!}{\!}$, taking an argument in $\V{X}$ followed by a
-recursive argument. Easily, one sees that this code describes the
-following functor:
+recursive argument. This code describes the following functor:
 
@@
 
-Again, we are at one evolutionary step away from $\ListD$. However,
+Again, we are one evolutionary step away from $\ListD$. However,
 instead of a single call to the induction code, we call it twice. The
@@
 
-Again, this is a clear reminiscence of the sum-of-products style: we
-have a finite sum of constructors. Every description can be expressed
-in this style, using a singleton as a constructor if
-necessary. Converselly, a tagged description can be turned into a
+Again, this is a clear reminiscence of the sum-of-products style,
+characterized by a finite sum of constructors. Every description can
+be expressed in this style, using a singleton constructor if
+necessary. Conversely, a tagged description can be turned into a
 full-blown description. We denote $\toDesc{\V{TD}}$ the description
@@
 as presented in Figure~\ref{fig:desc_universe}. Although it makes no
-conceptual difference, the code of description should rather be read
-as a specification. We promise the existence of such objects,
-satisfying the typing rules. It will be the subject of
-Section~\ref{sec:desc-levitate} to actually implement the
+conceptual difference, this code should rather be read as a
+specification. We promise the existence of such objects, satisfying
+these typing rules. It will be the subject of
+Section~\ref{sec:desc-levitate} to actually implement this
 specification.

### 2010-04-01 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: proof-read section 4, levitation

@@
 
-In this section, we will fulfil our promises. First, we will build
-finite sets. Second, we will implement the code of the $\Desc$
-universe.  Despite being a perilous pedagogical exercise for the
-authors, we hope to convey to the reader the dizzy feeling of
-levitation, without the falling.
+In this section, we will fulfil our promises. We first implement
+finite sets. Then, we implement the codes of the $\Desc$ universe.
+Despite being a perilous pedagogical exercise for the authors, we hope
+to convey to the reader the dizzy feeling of levitation, without the
+falling.
 
@@
 theory does not need to be extended with the type former $\EnumU$, the
-constructors $\NilE$ and $\ConsE$. We have just defined $\EnumU$
-above, by fixpoint over the pattern functor denoted by
-$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
-$\ListCons$ constructors of lists.
+constructors $\NilE$ and $\ConsE$. $\EnumU$ is defined above, as the
+fixpoint of $\ListD~\UId$. $\NilE$ and $\ConsE$ are simply the type
+constructors of lists, $\ListNil$ and $\ListCons$.
 
@@
 principle. From an implementation perspective, this means that this
-operator is simply implemented in the dependently-typed target
-language, as any other function.
+operator is implemented in the dependently-typed target language, as
+any other function.
 
@@
 
-We shall now fulfil our promises for $\Desc$. We are going to
+We shall now fulfil our promise concerning $\Desc$. We are going to
 implement the codes of the universe of descriptions. In and by itself,
-the system of codes, $\Desc$, is nothing but a datatype. We are in
-the same situation as with $\EnumU$: we ought to be able to describe
-the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
-first-class citizen, born with the standard, generic equipment of
-datatypes.
+the codes, $\Desc$, is nothing but a datatype. We are in the same
+situation as with $\EnumU$: we ought to be able to describe the codes
+of $\Desc$ in $\Desc$ itself. Hence, this code would be a first-class
+citizen, born with the standard, generic equipment of datatypes.
 
@@
 For $\DUnit$, we immediately reach the end of the description. For
-$\DIndx{\!}$, there is a single recursive argument. The case of
+$\DIndx{\!}$, there is a single recursive argument. Describing
 $\DSigma{\!}{\!}$ is problematic. Recall the specification of
@@
 same but, in practice, they will be dealt with rather
-differently. While definitional equality can cope with first order
+differently. While definitional equality can cope with first-order
 objects, the functional presentation introduced by $\DHindx{\!}{}$ is
@@
 $\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
-on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing the $\V{H}$
+on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing an $\V{H}$
 in $\Set$ with a recursive argument.
@@
 codes of the universe of description. Taking the fixpoint of this
-object gives us $\Desc$, up to isomorphism. We have implemented our
-code of datatypes as an object levitating inside itself. However,
+object gives us $\Desc$, up to isomorphism. We have implemented the
+codes of datatypes as an object levitating inside itself. However,
 this levitation operation, just as any magic trick, relies on an
@@
 this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
-construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
-codes. Then, we can type-check the case definition, between square
+construct elaborates to a finite set $\V{E}$ in $\EnumU$, inhabited by
+the codes. Then, we can type-check the case definition, between square
 brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
-$\EnumT{\V{e}} \To \Mu{\DescD}$. This corresponds to a finite function
+$\EnumT{\V{E}} \To \Mu{\DescD}$. This corresponds to a finite function
 definition, as presented in Figure~\ref{fig:type-checking}.
-
-What happens if we unfold the definition? We ought to build the following term:
-
-\[
-\PLAM{\V{x}}{(\EnumT{\V{e}})} \switch{\V{e}}{(\LAM{\_} \Mu{\DescD})}{\V{\pi^f}}{\V{x}}
-\]
-
-But this is quite problematic. We are still in the process of
-constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
-begging for this very same $\DescD$. Despite our willingness, we
-cannot materialise such motive.
 
@@
 
-However, we perfectly know what the motive is. Consequently, we extend
-the type theory with a special-purpose operator:
+What happens if we unfold the definition? We ought to build the following term:
+
+\[
+\PLAM{\V{x}}{(\EnumT{\V{E}})} \switch{\V{E}}{(\LAM{\_} \Mu{\DescD})}{\V{\pi^f}}{\V{x}}
+\]
+
+But this is quite problematic. We are still in the process of
+constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
+begging for this very same $\DescD$. Despite our willingness, we
+cannot materialise such motive. However, we perfectly know what the
+motive is. Consequently, we extend the type theory with a
+special-purpose operator:
 
@@
 The entire work of the magician stands here, in this extension. Note
-that the developer do not have to decide between the two variants of
+that the developer do not have to chose between the two variants of
 $\F{switch}$. This is dealt with by the type propagation system
-(Fig.~\ref{fig:type-checking}), extended by the following inference
+(Fig.~\ref{fig:type-checking}), extended with the following inference
 rule:
@@
 exercise has several practical outcomes. First of all, it reveals that
-the $\Desc$ universe is just plain data. Just as any piece of data, it
-can therefore be inspected and manipulated. Moreover, it is expressed
-in the $\Desc$ universe. As a consequence, it is equipped, for free,
-with an induction principle. So, our ability to inspect and program
-with $\Desc$ is not restricted to the meta-language: we now have all
-the necessary equipment in the target language to \emph{program} over
+the $\Desc$ universe is just plain data. As any piece of data, it can
+therefore be inspected and manipulated. Moreover, it is expressed in
+the $\Desc$ universe. As a consequence, it is equipped, for free, with
+an induction principle. So, our ability to inspect and program with
+$\Desc$ is not restricted to the meta-language: we now have all the
+necessary equipment in the target language to \emph{program} over
 datatypes. In this setting, \emph{generic programming is just
@@
 
-The catamorphism is an induction on the description $\V{D}$, with a
-non-dependent motive targeting $\V{T}$. Given a node $\V{xs}$ and the
-induction hypotheses, the method ought to build an element of
+The catamorphism is defined by induction on the description $\V{D}$,
+with a non-dependent motive targeting $\V{T}$. Given a node $\V{xs}$
+and the induction hypotheses, the method ought to build an element of
 $\V{T}$. Provided that we know how to make an element of
@@
 structure with the corresponding monadic operation, \bind\ and
-\return. We have shown how to implement \bind, for any free
-monad. This construction is an example of type-indexed
+\return. This construction is an example of type-indexed
 datatype~\cite{hinze:generic-haskell}, as found in Generic Haskell:

### 2010-04-01 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: fix dangling bib reference.
2.3 redone

@@
 \begin{array}{@{}ll}
-\F{switch} :& \PITEL{\V{E}}{\EnumU}
-               \PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To\\
-             & \spi{\V{E}}{\V{P}} \To
+\F{switch} : \PITEL{\V{E}}{\EnumU}
+               \PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To
+              \spi{\V{E}}{\V{P}} \To
                \PITEL{\V{x}}{\EnumT{\V{E}}} \To \V{P}~\x
@@
 
-Our proposal to tame the problem of codes is deeply rooted in the
-bidirectional presentation of type-checking from Pierce and
-Turner~\cite{pierce:bidirectional-tc}. In this formalism,
-type-checking is divided into two complementary phases. In the
-\emph{type inference} phase, types are \emph{pulled} out of terms. A
+Our approach to tidying the coding cruft is deeply rooted in the
+bidirectional presentation of type checking from Pierce and
+Turner~\cite{pierce:bidirectional-tc}. They divide
+type inference into two communicating components. In
+\emph{type synthesis}, types are \emph{pulled} out of terms. A
 typical example is a variable in the context:
-
 \[
 \Rule{\G ; \xS ; \Delta \vdash \Valid}
-     {\G ; \xS ; \Delta \vdash \Bhab{\V{x}}{\V{S}}}
+     {\G ; \xS ; \Delta \vdash \Bhab{\V{x}}{\M{S}}}
 \]
+Because the context stores the type of the variable, we can extract the
+type whenever the variable is used.
 
-Because the variable and its type are stored in the context, whenever
-this variable is used, we can pull its type out of it.
-
-On the other hand, in the \emph{type checking} phase, terms are
-\emph{pushed} into types. We are handed a type together with a term,
-our task consists in checking that the term is indeed of this
-type. Doing so, we are allowed, and actually encouraged, to use the
-information provided by the type. Therefore, we can relax our
-requirements on the term. A typical example is the lambda abstraction:
-
+On the other hand, in the \emph{type checking} phase, types are
+\emph{pushed} into terms. We are handed a type together with a term,
+our task consists of checking that the type admits the term. In doing
+so, we can and should use the information
+provided by the type. Therefore, we can relax our requirements on the
+term. Consider \(\LAMBINDER\)-abstraction:
 \[
-\Rule{\G       \vdash \Type{\V{S}} \quad
-      \G ; \xS \vdash \Bhab{\V{t}}{\V{T}}}
-     {\G \vdash \Bhab{\PLAM{\x}{\V{S}} \V{t}}{\PIS{\xS} \V{T}}}
+\Rule{\G       \vdash \Type{\M{S}} \quad
+      \G ; \xS \vdash \Bhab{\M{t}}{\M{T}}}
+     {\G \vdash \Bhab{\PLAM{\x}{\M{S}} \M{t}}{\PIS{\xS} \M{T}}}
 \]
-
-Here, we are forced to annotate lambda abstractions by the type of the
-domain. In a bidirectional setting, the domain is provided by the
-$\Pi$ type we are pushing the abstraction into. Hence we do not need
-the annotation.
+The official rules require an annotation specifying the domain.
+However, in type \emph{checking}, the \(\Pi\)-type we push in determines
+the domain, so we can drop the annotation.
 
@@
 
-We adapt and formalise this \emph{type propagation} mechanism to our
-richer type theory. First of all, we define a high-level syntax,
-composed of \emph{expressions}. Expressions are divided into two
-syntactic categories: $\exprIn$ into which types are pushed in, and
-$\exprEx$ from which types are pulled out. Following the bidirectional
-presentation, $\exprIn$ are subject to type \emph{checking}, while
-$\exprEx$ are subject to type \emph{synthesis}.
-\note{Why don't we need a context validity judgement, again?}
+We adapt this idea, yielding a \emph{type
+propagation} system, whose purpose is to elaborate high-level
+\emph{expressions} into the terms of our underlying type theory, much
+as in the definition of Epigram
+1~\cite{mcbride.mckinna:view-from-the-left}.  We divide expressions
+into two syntactic categories: $\exprIn$ into which types are pushed,
+and $\exprEx$ from which types are extracted. In the
+bidirectional spirit, the $\exprIn$ are subject to type
+\emph{checking}, while the $\exprEx$---variables and elimination
+forms---admit type \emph{synthesis}. We embed $\exprEx$ into
+$\exprIn$, demanding that the synthesized type coincides with the type
+proposed. The other direction---only necessary to apply
+abstractions or project from pairs---takes a type annotation.
 
@@
 
-The type synthesis rules are presented
-Figure~\ref{fig:type-synthesis}. As illustrated by the judgement form,
-a type synthesis problem is provided an expression $\exprEx$ in a
-context, which is elaborated into a low-level term together with its
-type. These terms act as \emph{sources} for the type propagation
-machinery: they make types out of expressions. We have already studied
-the case of a variable in context. This also include function
-application as well as projections. Further down the line, these
-inferred types might propagate to other terms, thanks to type
-checking.
+Type synthesis is the \emph{source} of types. It follows the \(\exprEx\)
+syntax, delivering both the elaborated term and its type. Terms
+and expressions never mix: e.g., for application, we
+instantiate the range with the \emph{term} delivered by checking the
+argument \emph{expression}. Hardwired operators are checked as variables.
+\input{figure_type_synthesis}
 
-\begin{figure}
-\input{figure_type_synthesis}
-\caption{Type synthesis}
-\label{fig:type-synthesis}
-\end{figure}
+%\begin{figure}
+%\input{figure_type_synthesis}
+%\caption{Type synthesis}
+%\label{fig:type-synthesis}
+%\end{figure}
 
@@
 
-Dually to type synthesis, type checking judgements
-(Fig.~\ref{fig:type-checking}) act as \emph{sinks}. From an expression
-$\exprIn$ and a type propagated down to it, they elaborate a low-level
-term. The strength of this approach comes from the availability of the
-type to drive the elaboration of expressions. Its benefit on lambda
-abstraction and pairs is well-known. Let us explore the consequences
-on our coding scheme for finite sets.
-
-A first consequence is that we can actually denote elements of finite
-sets \emph{by their name}. Indeed, such expression will be pushed into
-a $\EnumT{\V{u}}$, where $\Bhab{\V{u}}{\EnumU}$ contains all the names
-populating the finite set. It is therefore easy to elaborate a name
-into the corresponding index.
-
-A second consequence concerns the definition of a finite set. We can
-abstract away the details of code and use a familiar syntax
-instead. Hence, we use the syntax $[t_1 \ldots t_n]$ to denote the
-finite sets composed of the named elements $t_1$, \ldots, $t_n$. This
-syntax is reminiscent to Lisp notation for lists. The translation to
-elements of $\EnumU$ is straightforward.
-
-Finally, and more importantly, we are able to present the developer an
-abstraction of \emph{finite} function space. This works as follow. When
-an expression $\V{t}$ is pushed into a function which domain is a finite
-set, we instead elaborate it as a term in a small $\pi$-type. That is,
-a finite function. To realise our promise of a big $\Pi$-type, we
-elaborate a call to the $\F{switch}$ eliminator.
-
-The abstraction we have developed in this section is entirely
-transparent to the user. In the case of finite function space, the
-developer never has to explicitly \emph{call} the $\F{switch}$
-eliminator. For diagnosis purposes, a similar, reverse mechanism turns
-low-level terms into high-level expressions. Hence, this system, whose
-complete presentation is out of the scope of this paper, seals the
-low-level type theory behind the high-level expressions.
+Dually, type checking judgments
+(Fig.~\ref{fig:type-checking}) are \emph{sinks} for types. From an
+$\exprIn$ and a type pushed into it, they elaborate a low-level
+term, extracting information from the type.
+Canonical set-formers are \emph{checked}: we could exploit \(\Set:\Set\)
+to give them synthesis rules, but this would prejudice our future
+stratification plans. Note that abstraction and pairing are free of
+annotation, as promised. Most of the propagation rules are unremarkably
+structural: we have omitted some mundane rules which just follow
+the pattern, e.g., for \(\UId\). It is for the finite enumerations that
+things get interesting.
 
@@
 \end{figure}
+
+Firstly, we present the codes for enumerations as right-nested tuples
+which, by our LISP convention, we write as unpunctuated lists of tags
+\(\sqr{\etag{t_0}\ldots\etag{t_n}}\).
+
+Secondly, we can denote an element \emph{by its
+name}: the type pushed in allows us to recover the numerical
+index. However, we retain the numerical forms to facilitate
+\emph{generic} operations and ensure that shadowing a tag is punished
+fittingly, not fatally.
+
+Finally, we enable you to write functions from finite sets as tuples.
+Any bracketed form, \(\void\) or \(\pair{\_}{\_}{}\), is accepted by the
+function space---the general product---if it is accepted by the
+small product. Propagation fills in the appeal to the $\F{switch}$,
+copying the range information.
+
+We note, but have no room to elaborate on the fact that the propagation
+is reversible. Indeed, our interactive development tools rely on the
+reverse transformation to relay intelligible information to the user.
+For any specific usage, the encoding is completely hidden by these
+translations. Only when programming generically with enumerations is it
+necessary to address the encoding, but in that case, it is a benefit!
 
@@
 
-In this section, we have developed a type propagation system based on
-bidirectional type-checking. Using bidirectional type-checking as a
-boilerplate scrapper is a well-known
-technique~\cite{pierce:bidirectional-tc,
-  xi:bidirectional-tc-bound-array, chlipala:strict-bidirectional-tc}
-\note{Everybody ok with the citations?}. In our case, we have shown
-how to instrument bidirectionality to rationalise the expressivity of
-dependent types. We have illustrated our approach with finite sets. We
-have abstracted away the low-level presentation of finite sets,
-offering a convenient syntax instead.
+%In this section, we have developed a type propagation system based on
+%bidirectional type-checking. Using bidirectional type-checking as a
+%boilerplate scrapper is a well-known
+%technique~\cite{pierce:bidirectional-tc,
+%  xi:bidirectional-tc-bound-array, chlipala:strict-bidirectional-tc}
+%\note{Everybody ok with the citations?}. In our case, we have shown
+%how to instrument bidirectionality to rationalise the expressivity of
+%dependent types. We have illustrated our approach with finite sets. We
+%have abstracted away the low-level presentation of finite sets,
+%offering a convenient syntax instead.
 
-This example teaches us that we should not be afraid of codes, as soon
-as type information is available. We have shown how to rationalise
-this information in a formal presentation. Hence, we have shown that
-programming with such objects is practical. As we introduce more codes
-in our theory, we will show how to extend the framework we have
-developed so far.
+%This example teaches us that we should not be afraid of codes, as soon
+%as type information is available. We have shown how to rationalise
+%this information in a formal presentation. Hence, we have shown that
+%programming with such objects is practical. As we introduce more codes
+%in our theory, we will show how to extend the framework we have
+%developed so far.
+
+The relatively civilised presentation we have achieved here is
+possible because our universe of enumerations specifies not only the
+\emph{representation} of the low-level values in each set as bounded
+numbers, but also the \emph{presentation} of these values as
+high-level tags. To encode only the former, we should merely need the
+\emph{size} of enumerations, but we extract more work from these types
+by making them more informative.  We have also, \emph{en passant},
+distinguished enumerations which have the same cardinality but
+describe distinct notions:
+\(\EnumT{\sqr{\etag{\CN{red}}\,\etag{\CN{blue}}}}\) is not
+\(\EnumT{\sqr{\etag{\CN{green}}\,\etag{\CN{orange}}}}\).

### 2010-04-01 papers/icfp-2010-desc/paper_type_theory.tex
merge TT

@@
 
-The purpose of type propagation is to \emph{elaborate} a high-level
-expression into a term of the type theory, together with its
-type. This presentation builds on the experience gained in
-implementing Epigram 1~\cite{mcbride:view-from-the-left}. Being
-entirely syntax driven, we present our formalism as a system of
-syntax-directed inference rules. As the presentation largely mirrors
-the inference rules of the type theory, we will ignore the judgements
-that are identical. We refer our reader to the associated technical
-report~\cite{chapman:desc-tech-report} for the complete system of
-rules.
+% As the presentation largely
+%mirrors the inference rules of the type theory, we will ignore the
+%judgments that are identical. We refer our reader to the associated
+%technical report~\cite{chapman:desc-tech-report} for the complete
+%system of rules.
 

### 2010-04-01 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: bring in AllI and allI

@@
 However, we still have to define the actual grammar. We obtain it by
-evolving $\Desc$ to cope with indexing. The step-by-step
-transformation is described in our technical report. The code of
-$\IDesc{\!}$ is presented in Figure~\ref{fig:idesc}. The induction
-scheme on $\IDesc{\!}$ has the following type:
-
-\[
+evolving $\Desc$ to cope with indexing. \note{Say something about the
+  transformation} The code of $\IDesc{\!}$ is presented in
+Figure~\ref{fig:idesc}. The induction scheme on $\IDesc{\!}$ has the
+following type:
+
+\[\stk{
 \begin{array}{@{}ll}
@@
                    \V{P}\: \pair{\V{i}}{\V{x}}{}
-\end{array}
+\end{array} \\
+\F{inductionI}\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: (\Con{\V{xs}}) =  \\
+\qquad
+    \V{ms}\: \V{i}\: \V{xs}\: (\allI{}
+                                    {\V{R}\: \V{i}}
+                                    {\IMu{\V{I}}{R}{}}
+                                    {\V{P}}
+                                    {(\LAM{\pair{\V{i}}{\V{xs}}{}} \F{inductionI}\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: \V{xs})}
+                                    {\V{xs}})
+}\]
+
+Where the operators $\AllI{\!}{\!}{\!}{\!}$ and
+$\allI{\!}{\!}{\!}{\!}{\!}{\!}$ are defined in
+Figure~\ref{fig:allI-predicates}. As for descriptions, we can compute
+a generic catamorphism, $\F{cataI}$, from $\F{inductionI}$.
+
+\begin{figure*}
+
+\[
+\begin{array}{ll}
+%%
+\stk{
+\begin{array}{@{}ll}
+\AllI{\!}{\!}{\!}{\!} : & _{\Bhab{\V{I}}{\Set}}
+                          \PITEL{\V{D}}{\IDesc{\V{I}}}
+                          \PI{\V{X}}{\V{I} \To \Set} \\
+                        & \idescop{\V{D}}{\V{I}}{\V{X}} \To
+                          \IDesc{(\TIMES{\V{I}}{\V{X}})}
+\end{array} \\
+\begin{array}{@{}ll@{\:=\:\:}l}
+\AllI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{x}} &
+    \DVar{\pair{\V{i}}{\V{x}}{}} \\
+\AllI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{k}} &
+    \DConst{K} \\
+\AllI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\pair{\V{d}}{\V{d'}}{}} &
+    \DProd{\AllI{}{\V{D}}{\V{X}}{\V{d}}}{\AllI{}{\V{D'}}{\V{X}}{\V{d'}}} \\
+\AllI{\:}{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\pair{\V{a}}{\V{b}}{}} &
+    \AllI{}{(\V{T}\: \V{a})}{\V{X}}{\V{b}} \\
+\AllI{\:}{(\DPi{\V{S}}{\V{T}})}{& \V{X}}{\V{f}} &
+    \DPi{\V{S}}{(\LAM{s} \AllI{}{(\V{T}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})})}
+\end{array}
+}
+&
+%%
+\stk{
+\begin{array}{@{}ll}
+\allI{\!}{\!}{\!}{\!}{\!}{\!} : & _{\Bhab{\V{I}}{\Set}}
+                                  \PITEL{\V{D}}{\IDesc{\V{I}}}
+                                  \PITEL{\V{X}}{\V{I} \To \Set} 
+                                  \PI{\V{R}}{\TIMES{\V{I}}{\V{X}} \To \Set} \\
+                                & (\PI{\V{x}}{\TIMES{\V{I}}{\V{X}}} \V{R}\: \V{x}) \To
+                                  \PI{\V{xs}}{\idescop{\V{D}}{}{\V{X}}} 
+                                  \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{\TIMES{\V{I}}{\V{X}}}{\V{R}}
+\end{array} \\
+\begin{array}{@{}ll@{\:=\:\:}l}
+\allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{R}}{\V{P}}{\V{x}} &
+    \V{P}\: \pair{\V{i}}{\V{x}}{} \\
+\allI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{R}}{\V{P}}{\V{k}} &
+    \V{k} \\
+\allI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\V{R}}{\V{P}}{(\V{d}, \V{d'})} &
+    \pair{\allI{}{\V{D}}{\V{X}}{\V{R}}{\V{P}}{\V{d}}}
+         {\allI{}{\V{D'}}{\V{X}}{\V{R}}{\V{P}}{\V{d'}}}{} \\
+\allI{\:}{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{R}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
+    \allI{}{(\V{T}\: \V{a})}{\V{X}}{\V{R}}{\V{P}}{\V{b}} \\
+\allI{\:}{(\DPi{\V{S}}{\V{T}})}{& \V{X}}{\V{R}}{\V{P}}{\V{f}} &
+    \allI{}{(\V{T}\: \V{a})}{\V{X}}{\V{R}}{\V{P}}{(\V{f}\: \V{a})}
+\end{array}
+\end{array}
+}
 \]
 
-Where the operator $\AllI{\!}{\!}{\!}{\!}$ is of type:
-
-\[
-\AllI{\!}{\!}{\!}{\!} : _{\Bhab{\V{I}}{\Set}}
-                        \PITEL{\V{D}}{\IDesc{\V{I}}}
-                        \PI{\V{P}}{\V{I} \To \Set}
-                        \idescop{\V{D}}{\V{I}}{\V{P}} \To
-                        \IDesc{(\SIGMA{\V{i}}{\V{I}} \V{P})} 
-\]
-
-The implementation of these operators can be found in the companion
-technical report. As for descriptions, we can compute a generic
-catamorphism, $\F{cataI}$, from $\F{inductionI}$.
+\caption{Indexed induction predicates}
+\label{fig:allI-predicates}
+
+\end{figure*}
 

### 2010-04-01 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: inline figures

@@
 extend our universe of descriptions. This consists in adding a code
-$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. The code and its
-interpretation are presented in
-Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ uses the
-elements of $H$ to index as many recursive arguments.
-
-\begin{figure}
+$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. Intuitively,
+$\DHindx{\!}{\!}$ uses the elements of $H$ to index as many recursive
+arguments:
 
@@
 
-\caption{Higher-order universe of descriptions}
-\label{fig:hindx_desc}
-
-\end{figure}
-
 Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
@@
 
-As shown in Figure~\ref{fig:desc-levitate}, we are now able to
-describe our universe of datatypes.  The $\DUnit$ and $\DIndx{\!}$
-cases remain unchanged, as expected. We successfully describe the
-$\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
-on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing an $\V{H}$
-in $\Set$ with a recursive argument.
-
-At a first glance, we have achieved our goal. We have described the
-codes of the universe of description. Taking the fixpoint of this
-object gives us $\Desc$, up to isomorphism. We have implemented the
-codes of datatypes as an object levitating inside itself. However,
-this levitation operation, just as any magic trick, relies on an
-invisible cable. Let us reveal it, hence finishing our implementation.
-
-\begin{figure}
+We are now able to describe our universe of datatypes:
 
@@
 
-\caption{Levitating description of $\Desc$}
-\label{fig:desc-levitate}
+The $\DUnit$ and $\DIndx{\!}$ cases remain unchanged, as expected. We
+successfully describe the $\DSigma{\!}{}$ case, by a simple appeal to
+the higher-order induction on $\V{S}$. The $\DHindx{\!}{\!}$ case
+consists in packing an $\V{H}$ in $\Set$ with a recursive argument.
 
-\end{figure}
+At a first glance, we have achieved our goal. We have described the
+codes of the universe of description. Taking the fixpoint of this
+object gives us $\Desc$, up to isomorphism. We have implemented the
+codes of datatypes as an object levitating inside itself. However,
+this levitation operation, just as any magic trick, relies on an
+invisible cable. Let us reveal it, hence finishing our implementation.
 
@@
 Our reader might be slightly confused to learn that the trick is
-visible in the definition of $\DescD$
-(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
-careful usage of type propagation. Indeed, let us try to elaborate
+visible in the definition of $\DescD$. Or rather, it is made invisible
+by careful usage of type propagation. Indeed, let us try to elaborate
 this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
@@
                            {\V{t'}}}
-     {\Gamma \Vdash \begin{array}{@{}l}
-                        \propag{\push{\V{t}}{\EnumT{\V{e}} \To \Desc}}
-                               {\\ \PLAM{\V{x}}{(\EnumT{\V{e}})} \switchD{\V{e}}{\V{t'}}{\V{x}}}
-                    \end{array}
+     {\Gamma \Vdash \propag{\push{\V{t}}{\EnumT{\V{e}} \To \Desc}}
+                           {\PLAM{\V{x}}{(\EnumT{\V{e}})} \switchD{\V{e}}{\V{t'}}{\V{x}}}
      }\;\mbox{$\V{t}$ is $[]$ or $[\_,\_]$}

### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: bring in AllI and allI
Desc@ICFP: align codes, whenever possible
Desc@ICFP: fix colors
start 3 run away by 10

@@
 In this section, we describe an implementation of inductive types, as
-we know them in ML-like languages. By working with datatypes we are
-familiar with, we rely on our intuition on these objects. However, our
-goal remains to discuss datatypes as we find them in dependently-typed
-languages.
-
-Our proposal is inspired by Dybjer and Setzer's closed formulation of
-induction-recursion~\cite{dybjer:axiom-ir}. For inductive datatypes, the
-full power of induction-recursion is not necessary. Therefore, we
-adopt a stripped-down version of it.
+we know them in ML-like languages. By working with familiar datatypes,
+we hope to focus on the delivery mechanism, warming up gently to the
+indexed datatypes we really want.  % Our encoding is inspired by
+Dybjer and Setzer's closed formulation of
+induction-recursion~\cite{dybjer:axiom-ir}, but without the
+`-recursion'.  An impredicative Church-style encoding of datatypes is
+not adequate for dependently typed programming, as although such
+encodings present data as non-dependent eliminators, they do not
+support dependent \emph{induction}~\cite{geuvers:induction-not-derivable}.
+Whilst the \(\LAMBINDER\)-calculus captures all that data can \emph{do},
+it cannot ultimately delimit all that data can \emph{be}.
 
@@
 
-In dependently-typed languages, $\Sigma$-types can be interpreted in
-two different ways. This duality is actually reflected in the notation
-we can find in the literature. The notation $\blue{\Sigma}_{\V{x} :
-  \V{A}} (\V{B}\: \V{x})$ stresses that $\Sigma$-types are a
-generalisation of sums over arbitrary arities.  Where simply-typed
-languages have finite sums, dependently-typed languages have sums
-indexed by any set.
+In dependently typed languages, $\Sigma$-types can be interpreted as
+two different generalisations. This duality is reflected in the
+notation we can find in the literature. The notation
+$\blue{\Sigma}_{\x : \M{A}} (\M{B}\: \x)$ stresses that
+$\Sigma$-types are `dependent sums', generalising of sums over
+arbitrary arities, where simply typed languages have finite sums.
 
-On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}}
-(\V{B}\:\V{x})$ emphasises that $\Sigma$-types are a generalisation of
-products, where the second component can depend on the first
-one. Where simply-typed languages pack data into non-dependent tuples,
-we can now build tuples where data influence further types.
+On the other hand, our choice of notation $\SIGMA{\x}{\M{A}}
+(\M{B}\:\x)$ emphasises that $\Sigma$-types generalise products,
+with the type of the second component depending on the value of the
+first, where simply typed languages do not express such relative validity.
 
@@
 \emph{sum-of-products}. A datatype is defined by a finite sum of
-constructors, each composed by a product of arguments. To embrace
+constructors, each carrying a product of arguments. To embrace
 these datatypes, we have to capture this grammar.
+With dependent types, the notion of sum-of-products translates into
+\emph{sigmas-of-sigmas}.
 
-With dependent types, the notion of sum-of-products translates into
-\emph{sigmas-of-sigmas}. However, a direct encoding of datatypes
-through raw $\Sigma$-types is not a viable option. Indeed, to program
-with these datatypes, we need an induction principle. An encoding is
-unable to deliver such
-operation~\cite{geuvers:induction-not-derivable}.
-
-\subsection{A universe of descriptions}
+\subsection{A universe of datatype descriptions}
 \label{sec:desc-universe}
@@
 
-While sigmas-of-sigmas form a convenient \emph{semantics} for
-datatypes, we have seen that it makes a poor \emph{syntax}. To
-decouple these two notions, we construct an ad-hoc
-\emph{universe}. This technique originates in Martin-L\"of definition
-of type theory~\cite{martin-lof:itt}. It is a fruitful programming
-technique in dependently-typed
-languages~\cite{benke:universe-generic-prog, oury:power-of-pi}.
+While sigmas-of-sigmas can give a \emph{semantics} for the
+sum-of-products structure in each node of the tree-like values in a
+datatype, we need to account somehow for the recursive structure which
+ties these nodes together. Not for the first time, we do this by
+constructing a \emph{universe}~\cite{martin-lof:itt}. Universes
+are ubiquitous in dependently typed
+programming~\cite{benke:universe-generic-prog, oury:power-of-pi},
+but here we seek to exploit them as the foundation of our notion
+of datatypes.
 
-The key idea behind universe construction is our ability to make names
-by defining new types. These names are called \emph{codes}. By
-defining a set of codes, we define the syntax of a language. To give
-it a semantics, we \emph{interpret} these codes back into the type
-theory. Hence, codes act as labels, while the type theory provides the
-computational machinery. Codes being mere labels, we can inspect them,
-hence regaining structural information. We can also compute over them:
-deriving new codes from others, or even functions on them.
+%The key idea behind universe construction is our ability to make names
+%by defining new types. These names are called \emph{codes}. By
+%defining a set of codes, we define the syntax of a language. To give
+%it a semantics, we \emph{interpret} these codes back into the type
+%theory. Hence, codes act as labels, while the type theory provides the
+%computational machinery. Codes being mere labels, we can inspect them,
+%hence regaining structural information. We can also compute over them:
+%deriving new codes from others, or even functions on them.
 
@@
 
-To embed inductive types in our type theory, we define a universe of
-\emph{descriptions} (Figure~\ref{fig:desc_universe}). As expected, the
-code of this universe mimics the grammar of inductive types. To build
-sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
-$\Sigma$-type. Descriptions are terminated by $\DUnit$, containing no
-payload. To make recursive arguments, we introduce the $\DIndx{\!}$
-code. It interprets to a hole that will be closed by the fixpoint. We
-can also read $\V{D}$ as describing an endofunctor on $\Set$, with
-$\descop{\V{D}}{\!}$ being the object part.
-
-\begin{figure}
-
+To add inductive types to our type theory, we build a universe of
+datatype \emph{descriptions} by implementing the signature below, with
+codes mimicking the grammar of datatype declarations. We
+can read a description $\V{D}:\Desc$ as an \emph{endofunctor}
+on $\Set$, with $\descop{\V{D}}{\!}$ its action on an object, \(\V{X}\), soon
+to be instantiated recursively.
 \[\stk{
-\begin{array}{ll}
-\stk{
-\data \Desc : \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \DUnit          & \Desc \\
-    \DSigma{\!}{\!} & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
-    \DIndx{\!}      & \Desc \To \Desc
+\begin{array}{l}
+\Desc : \Set\\
+\DUnit : \Desc\\
+\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
+\DIndx{(\Bhab{\V{D}}{\Desc})} : \Desc \\
 \end{array}
-}
-\vspace{0.2in}
-\\
-\stk{
-\descop{\_\:}{} : \Desc \To \Set \To \Set \\
-\begin{array}{@{}ll@{\:=\:\:}ll}
-\descop{\DUnit}{& \V{X}}        &  \Unit                                       \\
-\descop{\DSigma{\V{S}}{\V{D}}}{& \V{X}} &  \SIGMAS{\V{s} : \V{S}}{\descop{\V{D}~\V{s}}{\V{X}}}         \\
-\descop{\DIndx{\V{D}}}{& \V{X}}     &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
-\end{array}
-}
+\smallskip\\
+\begin{array}{l@{\V{X}\:\mapsto\:}l}     
+\multicolumn{2}{l}{\descop{\_\:}{} : \Desc \To \Set \To \Set} \\
+ \descop{\DUnit}{} &  \Unit \\
+ \descop{\DSigma{\V{S}}{\V{D}}}{} &
+     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{X}}}  \\
+\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
 \end{array}
 }\]
-
-\caption{Universe of Descriptions}
-\label{fig:desc_universe}
-
-\end{figure}
+Descriptions are sequential structures, terminated by $\DUnit$, indicating
+the empty tuple. To build
+sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
+$\Sigma$-type. To request a recursive component, we have $\DIndx{\V{D}}$,
+where \(\V{D}\) descibes the rest of the node.
 
@@
 
-\[
+\[\stk{
 \begin{array}{lcl}
@@
               &   & \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x} 
-\end{array}
-\]
+\end{array} \\
+\F{induction}\: \V{D}\: \V{P}\: \V{ms}\: (\Con{\V{xs}}) =  \\
+\qquad
+    \V{ms}\: \V{xs}\: (\all{D}{\Mu{D}}{P}
+                           {(\LAM{\V{x}} \F{induction}\: \V{D}\: \V{P}\: \V{ms}\: \V{x})}
+                           {\V{xs}})
+}\]
 
@@
 $\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its
-definition is presented in Figure~\ref{fig:all-predicates}.  We refer
-the reader to our technical report for the implementation of
-$\F{induction}$. Note that this induction principle is the first
-manifestation of a generic operation over descriptions. Any datatype
-we define automatically comes with an induction principle.
+definition is presented in Figure~\ref{fig:all-predicates}. Note that
+this induction principle is the first manifestation of a generic
+operation over descriptions. Any datatype we define automatically
+comes with an induction principle.
 
@@
 \end{array} \\
-\All{\DUnit}{\V{X}}{\V{P}}{\Void} = 
+\begin{array}{@{}l@{}l@{\:=\:\:}l}
+\All{\DUnit}{& \V{X}}{\V{P}}{\Void} &
     \Unit \\
-\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} = 
-    \All{(\V{T}\: a)}{\V{X}}{\V{P}}{\V{b}} \\
-\All{(\DIndx{\V{D}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} =
+\All{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
+    \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
+\All{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
     \TIMES{\V{P}\: \V{a}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}} \\
-\All{(\DHindx{\V{H}}{\V{D}})}{\V{X}}{\V{P}}{\pair{\V{f}}{\V{b}}{}} =
+\All{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{f}}{\V{b}}{}} &
     \TIMES{(\PI{\V{i}}{\V{H}} \V{P}\: (\V{f}\: \V{i}))}
           {\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}}
+\end{array}
 }
@@
 \end{array} \\
-\all{\DUnit}{\V{X}}{\V{P}}{\V{R}}{\Void} =
+\begin{array}{@{}l@{}l@{\:=\:\:}l}
+\all{\DUnit}{& \V{X}}{\V{P}}{\V{R}}{\Void} &
     \Void \\
-\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} =
-    \all{(\V{T}\: a)}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
-\all{(\DIndx{D})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} =
+\all{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} &
+    \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
+\all{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} &
     \pair{\V{R}\: \V{a}}
          {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{b}}}{} \\
-\all{(\DHindx{H}{D})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{f}}{\V{b}}{}} =
-    \pair{\LAM{i} \V{R}\: (\V{f}\: \V{i})}
-         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{b}}{}
+\all{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\V{R}}{\pair{\V{f}}{\V{b}}{}} &
+    \pair{\LAM{\V{i}} \V{R}\: (\V{f}\: \V{i})}
+         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{b}}}{}
+\end{array}
 \end{array}

### 2010-04-01 papers/icfp-2010-desc/paper_type_theory.tex
start 3 run away by 10

@@
 $\exprIn$ and a type pushed into it, they elaborate a low-level
-term, extracting information from the type.
-Canonical set-formers are \emph{checked}: we could exploit \(\Set:\Set\)
-to give them synthesis rules, but this would prejudice our future
-stratification plans. Note that abstraction and pairing are free of
-annotation, as promised. Most of the propagation rules are unremarkably
-structural: we have omitted some mundane rules which just follow
-the pattern, e.g., for \(\UId\). It is for the finite enumerations that
-things get interesting.
+term, extracting information from the type. Note that we inductively ensure the following `sanity checks':
+\[\stkc{
+\Gamma\Vdash\propag{e}{\pull{t}{T}} \Rightarrow
+  \Gamma\vdash t:T \\
+\Gamma\Vdash\push{\propag{e}{t}}{T} \Rightarrow
+  \Gamma\vdash t:T
+}\]
+
+Canonical set-formers are \emph{checked}: we could exploit
+\(\Set:\Set\) to give them synthesis rules, but this would prejudice
+our future stratification plans. Note that abstraction and pairing are
+free of annotation, as promised. Most of the propagation rules are
+unremarkably structural: we have omitted some mundane rules which just
+follow the pattern, e.g., for \(\UId\). It is for the finite
+enumerations that things get interesting.
 
@@
 
-Finally, we enable you to write functions from finite sets as tuples.
-Any bracketed form, \(\void\) or \(\pair{\_}{\_}{}\), is accepted by the
+Finally, we let you express functions from enumerations as tuples.
+Any tuple-form, \(\void\) or \(\pair{\_}{\_}{}\), is accepted by the
 function space---the general product---if it is accepted by the
-small product. Propagation fills in the appeal to the $\F{switch}$,
+small product. Propagation fills in the appeal to $\F{switch}$,
 copying the range information.
@@
 We note, but have no room to elaborate on the fact that the propagation
-is reversible. Indeed, our interactive development tools rely on the
-reverse transformation to relay intelligible information to the user.
-For any specific usage, the encoding is completely hidden by these
-translations. Only when programming generically with enumerations is it
-necessary to address the encoding, but in that case, it is a benefit!
+is reversible. Our interactive development tools rely on the
+reverse transformation to report intelligibly to the user.
+The encoding of any specific enumeration is thus hidden by these
+translations. Only in enumeration-generic programs is the
+encoding exposed, but in that case, it is a benefit!
 
@@
 
-The relatively civilised presentation we have achieved here is
+Our type propagation mechanism does no fancy unification or constraint
+solving, just copying.  The relative civilision it delivers is
 possible because our universe of enumerations specifies not only the

### 2010-04-01 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: align codes, whenever possible
Desc@ICFP: fix colors
Desc@ICFP: proofread section 5, 'Indexed description'

@@
 
-Evolving in a dependent setting fosters new opportunities for
+Working with dependent types fosters new opportunities for
 datatypes. The typical example is bounded lists, also known as
@@
 grammar of datatypes is that of inductive families. To capture this
-grammar, we rely on \emph{indexing}. We are going to transform the
-$\Desc$ universe into $\IDesc{\!}$, the universe of indexed
-descriptions. We model inductive families by the $\V{I} \To
-\IDesc{\V{I}}$ type.
+grammar, we rely on \emph{indexing}.
 
@@
 
-In the previous section, we have presented the $\Desc$ universe as the
+In the previous section, we have presented the $\Desc$ universe as a
 grammar of functors in the category $\Set$. We have seen how to code
@@
 $\Set^{\V{I}}$. We call these \emph{indexed functors}. \note{Predicate
-  transformer anyone?} $\V{I} \To \IDesc{\V{I}}$ is our
-grammar for describing these functors. Hence, $\IDesc{\!}$ and its
-interpretation have the following types:
+  transformer anyone?} $\V{I} \To \IDesc{\V{I}}$ is our grammar for
+describing these functors. Hence, $\IDesc{\!}$ and its interpretation
+have the following types:
 
@@
   transformation} The code of $\IDesc{\!}$ is presented in
-Figure~\ref{fig:idesc}. The induction scheme on $\IDesc{\!}$ has the
-following type:
+Figure~\ref{fig:idesc}. Induction on indexed descriptions is defined
+by:
 
@@
 \begin{array}{@{}ll}
-\F{inductionI} : & _{\Bhab{\V{I}}{\Set}}
+\F{inductionI} : & _{\PI{\V{I}}{\Set}}
                    \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
@@
 \F{inductionI}\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: (\Con{\V{xs}}) =  \\
-\qquad
+\quad
     \V{ms}\: \V{i}\: \V{xs}\: (\allI{}
@@
 
-Where the operators $\AllI{\!}{\!}{\!}{\!}$ and
-$\allI{\!}{\!}{\!}{\!}{\!}{\!}$ are defined in
+Where the operators $\AllI{}{\!}{\!}{}$ and
+$\allI{}{\!}{\!}{\!}{\!}{}$ are presented in
 Figure~\ref{fig:allI-predicates}. As for descriptions, we can compute
@@
 \begin{array}{@{}ll}
-\AllI{\!}{\!}{\!}{\!} : & _{\Bhab{\V{I}}{\Set}}
+\AllI{\!}{\!}{\!}{\!} : & _{\PI{\V{I}}{\Set}}
                           \PITEL{\V{D}}{\IDesc{\V{I}}}
@@
 \end{array} \\
-\begin{array}{@{}ll@{\:=\:\:}l}
+\begin{array}{@{}l@{}l@{\:=\:\:}l}
 \AllI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{x}} &
@@
 \AllI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{k}} &
-    \DConst{K} \\
+    \DConst{\V{K}} \\
 \AllI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\pair{\V{d}}{\V{d'}}{}} &
@@
 \AllI{\:}{(\DPi{\V{S}}{\V{T}})}{& \V{X}}{\V{f}} &
-    \DPi{\V{S}}{(\LAM{s} \AllI{}{(\V{T}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})})}
+    \DPi{\V{S}}{(\LAM{\V{s}} \AllI{}{(\V{T}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})})}
 \end{array}
@@
 \begin{array}{@{}ll}
-\allI{\!}{\!}{\!}{\!}{\!}{\!} : & _{\Bhab{\V{I}}{\Set}}
+\allI{\!}{\!}{\!}{\!}{\!}{\!} : & _{\PI{\V{I}}{\Set}}
                                   \PITEL{\V{D}}{\IDesc{\V{I}}}
@@
 \end{array} \\
-\begin{array}{@{}ll@{\:=\:\:}l}
+\begin{array}{@{}l@{}l@{\:=\:\:}l}
 \allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{R}}{\V{P}}{\V{x}} &
@@
 \begin{array}{@{}l@{\:=\:\:}ll}
-\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
+\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}\: \V{i}                                           \\
 \idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
@@
 functor. Therefore, the recursive argument is materialised by
-$\DVar{\Void}$, where we were using $\DIndx$ in the previous
+$\DVar{\Void}$, where we were using $\DIndx{\!}$ in the previous
 presentation. This transformation generalizes to all inductive
@@
 \begin{array}{@{}ll}
-\IDescD~\V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[} 
+\IDescD\: \V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[} 
                                           \DVar{\!},
@@
 not using any indexing. We remedy this by encoding the vectors. Recall
-that the constructors $\VecNil$ and $\VecCons$ are only defined for an
+that the constructors $\VecNil$ and $\VecCons{\!}{\!}$ are only defined for an
 index $\NatZero$ and $\NatSuc$ respectively.:
@@
     \VecNil          & \D{Vec}\:\V{X}\:{\NatZero)}   \\
-    \VecCons     & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
+    \VecCons{\!}{\!} & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
 \end{array}
@@
 
-One way to code constrained datatype is to appeal to equality. The
+One way to code constrained datatypes is to appeal to equality. The
 constraints are therefore captured by equations in the datatype. In
@@
 In the $\VecNil$ case, a proof must be provided that the index is
-equal to $\NatZero$. In the $\VecCons$ case, we first store an element
+equal to $\NatZero$. In the $\VecCons{\!}{\!}$ case, we first store an element
 $\V{m}$ of $\Nat$. However, the constraint stipulates that $\V{m}$
@@
 of equalities in the picture. In a programming context, equalities
-haunt us, begging for proofs. Hence, whereas the developer typed a
-constructor-constrained datatype, the elimination form will involve
-equalities and their proofs. As much as possible, one would like to
+haunt us, begging for proofs. As much as possible, one would like to
 avoid suffering from this burden. However, the equalities introduced
@@
 
-\paragraph{Tagged Indexed Descriptions:}
+\paragraph{Tagged indexed descriptions:}
 
@@
 
-For convenience, we divide a tagged indexed description in two parts:
-first, the constructors that do not depend on the index; then, the
-constructors that do. The non-dependent part mirrors the definition
-for non-indexed descriptions. The index-depend part simply indexes the
-choice of constructors by $\V{I}$. Hence, by inspecting the index, it
-is possible to enable or disable the constructors.
-
-\[\stk{
- \TagIDesc{\V{I}}  \mapsto \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
- \ATagIDesc{\V{I}} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
- \ITagIDesc{\V{I}} \mapsto \\
-\qquad\SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
-}\]
+We divide a tagged indexed description in two parts: first, the
+constructors that do not depend on the index; then, the constructors
+that do. The non-dependent part mirrors the definition for non-indexed
+descriptions. The index-depend part simply indexes the choice of
+constructors by $\V{I}$. Hence, by inspecting the index, it is
+possible to enable or disable constructors.
+
+\[
+\begin{array}{@{}l@{\:\mapsto\:\:}l}
+ \TagIDesc{\V{I}}  & \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
+ \ATagIDesc{\V{I}} & \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
+ \ITagIDesc{\V{I}} & 
+     \SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
+\end{array}
+\]
 
@@
 %% Constructors
-\newcommand{\Eval}[1]{\etag{\CN{val}}~#1}
-\newcommand{\Econd}[3]{\etag{\CN{cond}}~#1~#2~#3}
-\newcommand{\Eplus}[2]{\etag{\CN{plus}}~#1~#2}
-\newcommand{\Ele}[2]{\etag{\CN{le}}~#1~#2}
+\newcommand{\Eval}[1]{\etag{\CN{val}}\:#1}
+\newcommand{\Econd}[3]{\etag{\CN{cond}}\:#1\:#2\:#3}
+\newcommand{\Eplus}[2]{\etag{\CN{plus}}\:#1\:#2}
+\newcommand{\Ele}[2]{\etag{\CN{le}}\:#1\:#2}
 
 %% Index mapper (terminology?)
-\newcommand{\Val}[1]{\F{Val}~#1}
-\newcommand{\Var}[2]{\F{Var}_{#1}~#2}
+\newcommand{\Val}[1]{\F{Val}\:#1}
+\newcommand{\Var}[2]{\F{Var}_{#1} #2}
 
@@
 \begin{array}{l@{\::\:\:}l}
-\Econd{}{}{}     & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
-\Eplus{}{}       & \Enat \To \Enat \To \Enat                           \\
-\Ele{}{}         & \Enat \To \Enat \To \Ebool                          \\
-\Eval{}          & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
+\Econd{\!}{\!}{\!}   & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
+\Eplus{\!}{\!}       & \Enat \To \Enat \To \Enat                           \\
+\Ele{\!}{\!}         & \Enat \To \Enat \To \Ebool                          \\
+\Eval{\!}            & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
 \end{array}
@@
 maps a type $ty$ of the object language to the corresponding type in
-the host language. Hence, the arguments of $\Eval{}$ are ensured to be
-of the expected type. We assume that $\Nat$ and $\Bool$ represent
+the host language. Hence, the arguments of $\Eval{\!}$ are ensured to
+be of the expected type. We assume that $\Nat$ and $\Bool$ represent
 natural numbers and booleans in the host language, equipped with an
 addition operation $\F{plusHost}$ and a comparison function
-$\F{leHost}$. We define $\Val{}$ as follows:
-
-\[\stk{
-\Val{} : \Ty \To \Set \\
+$\F{leHost}$. We define $\Val{\!}$ as follows:
+
+\[\stk{
+\Val{\!} : \Ty \To \Set \\
 \begin{array}{@{}l@{\:=\:\:}l}
@@
 
-In our universe of descriptions, this datatype is represented by a
-tagged indexed description. We use the index to carry the type: the
-resulting description is indexed by $\Ty$. We observe that some
-constructors are always defined, namely $\Econd{}{}{}$ and $\Eval{}$. On the
-other hand, the $\Eplus$ and $\Ele$ constructors are
-index-dependent. $\Eplus$ is defined if and only if the result type --
-the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
-index is $\Ebool$. The actual code precisely follows this intuition,
-as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
-informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
+In our universe of descriptions, the syntax of this language is
+described by a tagged indexed description. We use the index to carry
+the type: the resulting description is indexed by $\Ty$. We observe
+that some constructors are always defined, namely $\Econd{\!}{\!}{\!}$
+and $\Eval{\!}$. On the other hand, the $\Eplus{\!}{\!}$ and
+$\Ele{\!}{\!}$ constructors are index-dependent. $\Eplus{\!}{\!}$ is
+defined if and only if the result type -- the index -- is $\Enat$,
+whereas $\Ele{\!}{\!}$ is defined if and only if the index is
+$\Ebool$. The actual code precisely follows this intuition, as shown
+in Figure~\ref{fig:hexpr-full}. For brevity, we use an informal
+$\case{\ldots}{\ldots}$ notation, simulating a definition by
 pattern-matching. Formally, this corresponds to a call to the
@@
                  \begin{array}{l}
-                   {[\Eval{}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
+                   {[\Eval{\!}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
                    \LAM{\V{ty}}
@@
            \To {\Val{\V{ty}}} \\
-\begin{array}{@{}l@{\:=\:\:}l}
-\evalOne\: \_\: \pair{\Eval{}}{\V{x}}{}                                         & \V{x} \\
-\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
-\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{x} \\
-\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{\V{x}}{\V{y}}{}}{}                       & \F{plusHost}\: \V{x}\: \V{y} \\
-\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{\V{x}}{\V{y}}{}}{}                        & \F{leHost}\: \V{x}\: \V{y} 
+\begin{array}{@{}l@{}c@{}l@{\:=\:\:}l}
+\evalOne\: & \_\: & \pair{\Eval{\!}}{\V{x}}{}                                             & \V{x} \\
+\evalOne\: & \_\: & \pair{\Econd{\!}{\!}{\!}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
+\evalOne\: & \_\: & \pair{\Econd{\!}{\!}{\!}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{x} \\
+\evalOne\: & \Enat\: & \pair{\Eplus{\!}{\!}}{\pair{\V{x}}{\V{y}}{}}{}                     & \F{plusHost}\: \V{x}\: \V{y} \\
+\evalOne\: & \Ebool\: & \pair{\Ele{\!}{\!}}{\pair{\V{x}}{\V{y}}{}}{}                      & \F{leHost}\: \V{x}\: \V{y} 
 \end{array} \\
@@
                                  \toIDesc{\HExprD}\: 
-                                 \Val{}\: 
+                                 \Val{\!}\: 
                                  \evalOne\: 
@@
 However, so far, we are only able to define and manipulate
-\emph{closed} terms. By abstracting over $\Val{}$, it is possible to
+\emph{closed} terms. By abstracting over $\Val{\!}$, it is possible to
 build and manipulate \emph{open} terms, that is, terms with
-variables. On the model of $\Val{}$, we define $\Var{}{}$:
+variables. Following $\Val{\!}$, we define $\Var{}{}$ by:
 
@@
 
-Whereas $\Val{}$ was mapping the type to the corresponding host type,
+Whereas $\Val{\!}$ was mapping the type to the corresponding host type,
 $\Var{}{}$ maps types to a finite set. The finite set -- the context
--- contains closed terms. A variable is therefore a $\Eval{}$ that
+-- contains closed terms. A variable is therefore a $\Eval{\!}$ that
 contains a pointer to a particular element of the finite set -- an
@@
 Consequently, replacing $\Val{\V{ty}}$ by
-$(\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})$ in Figure~\ref{fig:hexpr-full} turns
-the language of closed term into a language of opened terms with
-constants. For readability, we will abbreviate $\LAM{\V{ty}}
-\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ into $\SUM{\Val{\!}}{\Var{\V{dom}}{\!}}$. \note{Is
-  it polite and comprehensible to ask that?} This defines a new
-indexed description, called $\HExprVarD{\V{dom}}$.
+$(\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})$ in
+Figure~\ref{fig:hexpr-full} turns the language of closed terms into a
+language of opened terms with variables and constants. For
+readability, we abbreviate $\LAM{\V{ty}}
+\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ by
+$\SUM{\Val{\!}}{\Var{\V{dom}}{}}$. \note{Is it polite and
+  comprehensible to ask that?} This defines a new indexed description,
+called $\HExprVarD{\V{dom}}$.
 
@@
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumLeft \V{x})  & \Con{\pair{\Eval{}}{\V{x}}{}} \\
+\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumLeft \V{x})  & \Con{\pair{\Eval{\!}}{\V{x}}{}} \\
 \discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumRight \V{v}) &
@@
 \newcommand{\gammaBool}{\V{\ensuremath{\gamma_{\CN{bool}}}}}
-\newcommand{\GammaSpi}[2]{\F{\ensuremath{\Gamma_{\CN{ty}}}}~#1~#2}
+\newcommand{\GammaSpi}[2]{\F{\ensuremath{\Gamma_{\CN{ty}}}}\: #1\: #2}
 
@@
 
-Where $\GammaSpi$ corresponds to a context, defined by:
-
-\[\stk{
-\GammaSpi{}{} : \EnumU \To \Ty \To \Set                                             \\
+Where $\GammaSpi{\!}{\!}$ corresponds to a context, defined by:
+
+\[\stk{
+\GammaSpi{\!}{\!} : \EnumU \To \Ty \To \Set                                             \\
 \GammaSpi{\V{dom}}{\V{ty}} = \spi{\V{dom}}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})} 
@@
 our datatype enjoys more structure than what we are given. We are
-facing a free monad, where $\Eval{}$ is the \return\ introducing
+facing a free monad, where $\Eval{\!}$ is the \return\ introducing
 variables. For convenience, we wrap $\discharge$ in a $\V{\sigma}$
@@
 \begin{array}{ll}
-\FreeIMonad{\_}{} : & _{\PITEL{\V{I}}{\Set}}
+\FreeIMonad{\_}{} : & _{\PI{\V{I}}{\Set}}
                      \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
@@
 
-As previously mentioned, we identify $\Eval{}$ with the \return\ of
+As previously mentioned, we identify $\Eval{\!}$ with the \return\ of
 the free monad, while the other components are the action of the
@@
                  \begin{array}{l}
-                   \EnumT{\red{[}\Econd{}{}{}\red{]}} \red{,} \\
+                   \EnumT{\red{[}\Econd{\!}{\!}{\!}\red{]}} \red{,} \\
                    \LAM{\V{ty}}
@@
 We call this datatype $\HExprFreeD$. By a simple unfolding of
-definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
+definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{\!}}$
 corresponds to the syntax of closed terms, $\HExprD$. Similarly,
-$\FreeIMonad{\HExprFreeD}{\Ty}{\SUM{\Val{\!}}{\Var{\V{dom}}{\!}}}$
+$\FreeIMonad{\HExprFreeD}{\Ty}{(\SUM{\Val{\!}}{\Var{\V{dom}}{}})}$
 corresponds to expressions with variables, $\HExprVarD{\V{dom}}$.
@@
 Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
-terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}\: \V{ty}$ to values in
+terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{\!}}\: \V{ty}$ to values in
 $\Val{\V{ty}}$. We are left with implementing $\substH$. We simply
@@
        \substI\: & \Ty\: 
-                  (\SUM{\Val{\!}}{\Var{\V{dom}}{\!}})\: 
-                  \Val{}\:
+                  (\SUM{\Val{\!}}{\Var{\V{dom}}{}})\: 
+                  \Val{\!}\:
                   \HExprFreeD\:  \\

### 2010-04-01 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: suppress line break before equations: we are at 11.5 pages

@@
 to be instantiated recursively.
+%
 \[\stk{
@@
 }\]
+%
 Descriptions are sequential structures, terminated by $\DUnit$, indicating
@@
 terms should not pose any difficulty. The code is the following:
-
+%
 \[\stk{
@@
 }\]
-
+%
 Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
@@
 following functor:
-
+%
 \[    \NatD\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{Z}    \]
@@
 for lists:
-
+%
 \[\stk{
@@
 }\]
-
+%
 The $\NatSuc{\!}$ constructor is turned into a proper
@@
 recursive argument. This code describes the following functor:
-
+%
 \[    \ListD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \D{\ensuremath{\times}} \V{Z}     \]
@@
 demonstrated by our description of binary trees:
-
+%
 \[\stk{
@@
 }\]
-
+%
 Again, we are one evolutionary step away from $\ListD$. However,
@@
 interpretation of this code corresponds to the following functor:
-
+%
 \[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \Prod \V{Z} \Prod \V{Z}     \]
@@
 \emph{tagged} description. Formally, we have:
-
+%
 \[
@@
 \]
-
-Again, this is a clear reminiscence of the sum-of-products style,
+%
+This is a clear reminiscence of the sum-of-products style,
 characterized by a finite sum of constructors. Every description can
@@
 own recursive definition:
-
+%
 \[
@@
 be tempted to provide it as a catamorphism:
-
+%
 \[
@@
 \]
-
+%
 Whereas this definition is amply sufficient in the simply-typed world,
@@
 definition:
-
+%
 \[\stk{
@@
 }\]
-
+%
 Where, intuitively, $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$ states that
@@
 Figure~\ref{fig:type-checking} with the following inference rule:
-
+%
 \[
@@
 \]
-
+%
 Where $\vec{\V{e}}$ denotes a telescope of arguments applied to the

### 2010-04-01 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: suppress line break before equations: we are at 11.5 pages

@@
 with the following forms:
-
+%
 \[
@@
 following `sanity checks' hold by induction on derivations
+%
 \[
@@
 \end{array} \]
+%
 and that judgments \(\M{J}\) are preserved by well-typed instantiation.
+%
 \[
@@
 sets.
+%
 \[
@@
 is to implement (by extending the theory, or by encoding) the signature
+%
 \[
@@
 \]
+%
 where some value \(E:\EnumU\) in the `enumeration universe' describes
@@
 rules:
+%
 \[
@@
 \]
+%
 Let us describe enumerations as lists of tags, with signature:
+%
 \[
@@
 \]
+%
 What are the \emph{values} in \(\EnumT{E}\)? Formally, we represent
 the choice of a tag as a numerical index into \(E\), via new rules:
+%
 \[
@@
 \]
+%
 However, we expect that in practice, you might rather refer to these
@@
 We first define the `small product' $\spi{\!}{\!}$ operator:
+%
 \[\stk{
@@
 }\]
+%
 This builds a right-nested
@@
 $\F{switch}$ operator, which, unsurprisingly, iterates projection:
+%
 \[\stk{
@@
 typical example is a variable in the context:
+%
 \[
@@
 \]
+%
 Because the context stores the type of the variable, we can extract the
@@
 term. Consider \(\LAMBINDER\)-abstraction:
+%
 \[
@@
 \]
+%
 The official rules require an annotation specifying the domain.
@@
 term, extracting information from the type. Note that we inductively ensure the following `sanity checks':
+%
 \[\stkc{

### 2010-04-01 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: s/~/\:/ in equations

@@
               \spi{\V{E}}{\V{P}} \To
-               \PITEL{\V{x}}{\EnumT{\V{E}}} \To \V{P}~\x
+               \PITEL{\V{x}}{\EnumT{\V{E}}} \To \V{P}\: \x
 \end{array} \\
@@
 We adapt this idea, yielding a \emph{type
-propagation} system, whose purpose is to elaborate high-level
+propagation} system, whose purpose is to elaborate compact
 \emph{expressions} into the terms of our underlying type theory, much
@@
 Dually, type checking judgments
-(Fig.~\ref{fig:type-checking}) are \emph{sinks} for types. From an
+%(Fig.~\ref{fig:type-checking})
+are \emph{sinks} for types. From an
 $\exprIn$ and a type pushed into it, they elaborate a low-level
@@
 unremarkably structural: we have omitted some mundane rules which just
-follow the pattern, e.g., for \(\UId\). It is for the finite
-enumerations that things get interesting.
+follow the pattern, e.g., for \(\UId\).
 
-\begin{figure}
+%\begin{figure}
 \input{figure_type_checking}
-\caption{Type checking}
-\label{fig:type-checking}
-\end{figure}
+%\caption{Type checking}
+%\label{fig:type-checking}
+%\end{figure}
+
+However, we also add abbreviations. We write \(\spl{\M{f}}\),
+pronounced `uncurry \(\M{f}\)' for the function which takes a pair and
+feeds it to \(\M{f}\) one component at a time, letting us name
+them individually. Now, for the finite enumerations, we go to work.
 
@@
 \(\sqr{\etag{t_0}\ldots\etag{t_n}}\).
-
 Secondly, we can denote an element \emph{by its
 name}: the type pushed in allows us to recover the numerical
-index. However, we retain the numerical forms to facilitate
-\emph{generic} operations and ensure that shadowing a tag is punished
+index. We retain the numerical forms to facilitate
+\emph{generic} operations and ensure that shadowing is punished
 fittingly, not fatally.
-
-Finally, we let you express functions from enumerations as tuples.
+Finally, we express functions from enumerations as tuples.
 Any tuple-form, \(\void\) or \(\pair{\_}{\_}{}\), is accepted by the
-function space---the general product---if it is accepted by the
+function space---the generalised product---if it is accepted by the
 small product. Propagation fills in the appeal to $\F{switch}$,
@@
 
-We note, but have no room to elaborate on the fact that the propagation
-is reversible. Our interactive development tools rely on the
-reverse transformation to report intelligibly to the user.
+Our interactive development tools also perform the
+reverse transformation for intelligible output.
 The encoding of any specific enumeration is thus hidden by these
-translations. Only in enumeration-generic programs is the
-encoding exposed, but in that case, it is a benefit!
+translations. Only, and rightly, in enumeration-generic programs is the
+encoding exposed.
 
@@
 
-Our type propagation mechanism does no fancy unification or constraint
-solving, just copying.  The relative civilision it delivers is
-possible because our universe of enumerations specifies not only the
-\emph{representation} of the low-level values in each set as bounded
-numbers, but also the \emph{presentation} of these values as
-high-level tags. To encode only the former, we should merely need the
-\emph{size} of enumerations, but we extract more work from these types
-by making them more informative.  We have also, \emph{en passant},
-distinguished enumerations which have the same cardinality but
-describe distinct notions:
+Our type propagation mechanism does no constraint
+solving, just copying, so it just the thin end of the elaboration wedge.
+It can afford us this `assembly
+language' level of civilisation as \(\EnumU\) universe
+specifies not only the \emph{representation} of the
+low-level values in each set as bounded numbers, but also the
+\emph{presentation} of these values as high-level tags. To encode only
+the former, we should merely need the \emph{size} of enumerations, but
+we extract more work from these types by making them more informative.
+We have also, \emph{en passant}, distinguished enumerations which have
+the same cardinality but describe distinct notions:
 \(\EnumT{\sqr{\etag{\CN{red}}\,\etag{\CN{blue}}}}\) is not

### 2010-04-02 papers/icfp-2010-desc/paper_desc.tex
added split, tweaked the examples

@@
 codes mimicking the grammar of datatype declarations. We
-can read a description $\V{D}:\Desc$ as an \emph{endofunctor}
+can read a description $\V{D}:\Desc$ as a `pattern functor'
 on $\Set$, with $\descop{\V{D}}{\!}$ its action on an object, \(\V{X}\), soon
@@
 \smallskip\\
-\begin{array}{l@{\V{X}\:\mapsto\:}l}     
+\begin{array}{l@{\V{Z}\:\mapsto\:}l}     
 \multicolumn{2}{l}{\descop{\_\:}{} : \Desc \To \Set \To \Set} \\
@@
  \descop{\DSigma{\V{S}}{\V{D}}}{} &
-     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{X}}}  \\
-\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
+     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{Z}}}  \\
+\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{Z}}{\descop{\V{D}}{\V{Z}}}
 \end{array}
@@
 $\Sigma$-type. To request a recursive component, we have $\DIndx{\V{D}}$,
-where \(\V{D}\) descibes the rest of the node.
+where \(\V{D}\) describes the rest of the node.
 
-To give some intuition on this universe, we now turn to some
-examples. For obvious pedagogical reasons, we will manually build the
-descriptions. However, it should be clear that, in practice, these
-definitions are automatically computed from an Agda-like $\data$
-definition.
+You may have noticed that we are a little coy about this presentation,
+writing of `implementing a signature' without clarifying how. A viable
+approach would simply be to extend the theory with constants for the
+constructors and an operator for \(\descop{\V{D}}\). However, in
+Section~\ref{sec:desc-levitate}, you will see what we actually do.  In
+the meantime, let us first gain some intuition for its use by
+developing some examples.
 
@@
 
-Our first example is the natural numbers, or rather its pattern
-functor. Our code is presented in the high-level expression language
-of Section~\ref{sec:type-propagation}. The translation back to the raw
-terms should not pose any difficulty. The code is the following:
+We begin with the natural numbers, now working in the high-level
+expression language of Section~\ref{sec:type-propagation}, exploiting
+type propagation.
 %
@@
 \NatD : \Desc \\
-\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
-                     {[ \DUnit \quad (\DIndx{\DUnit}) ]}
+\NatD \mapsto \DSigma{\EnumT{\sqr{\NatZero\: \NatSuc{\!} }}}
+                     {\sqr{ \DUnit \quad (\DIndx{\DUnit}) }}
 }\]
@@
 Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
-give a choice between the $\NatZero$ and $\NatSuc{}$ constructors. In
-the $\NatZero$ case, we reach the end of the description. In the
-$\NatSuc{\!}$ case, we attach a recursive argument and close the
-description. In a more synthetic notation, we have described the
-following functor:
+give a choice between the $\NatZero$ and $\NatSuc{}$ constructors.
+What follows depends on this choice, so we write the function
+computing the rest of the description in tuple notation.  In the
+$\NatZero$ case, we reach the end of the description. In the
+$\NatSuc{\!}$ case, we attach one recursive argument and close the
+description. Translating the \(\Sigma\) to a binary sum, we have
+effectively described the functor:
 %
 \[    \NatD\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{Z}    \]
+Correspondingly, we can see the injections to the sum:
+\[
+\sqr{\NatZero}:\descop{\NatD}Z \qquad
+\sqr{\NatSuc\:z}:\descop{\NatD}Z \;\mbox{if}\;z:Z
+\]
 
@@
 \ListD : \Set \To \Desc \\
-\ListD \: \V{X} \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons{\!}{\!} ]})}
-                           {[ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) ]}
+\ListD \: \V{X} \mapsto
+ \DSigma{\EnumT{\sqr{ \ListNil\: \ListCons{\!}{\!} }}}
+         {\sqr{ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) }}
 }\]
@@
 Finally, we are not limited to one recursive argument. This is
-demonstrated by our description of binary trees:
+demonstrated by our description of node-labelled binary trees:
 %
@@
 \begin{array}{@{}ll}
-\TreeD \: \V{X} \mapsto \DSigma{ & (\EnumT{[ \TreeLeaf, \TreeNode ]}) \\}
-                           { & [ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
+\TreeD \: \V{X} \mapsto \DSigma{ & \EnumT{\sqr{ \TreeLeaf, \TreeNode }} \\}
+  { & \sqr{ \DUnit \quad
+       ( \DIndx{(\DSigma{\V{X}}{\LAM{\_}\DIndx{\DUnit}})})} }
 \end{array}
@@
 Again, we are one evolutionary step away from $\ListD$. However,
-instead of a single call to the induction code, we call it twice. The
+instead of a single call to the induction code, we add another. The
 interpretation of this code corresponds to the following functor:
 %
-\[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \Prod \V{Z} \Prod \V{Z}     \]
+\[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} 
+          \V{Z} \Prod \V{X}  \Prod \V{Z}     \]
 
@@
 From the examples above, we observe that datatypes are defined by a
-$\DSigma{\!}{\!}$ which first argument is a finite set of
-constructors. The descriptions fitting into this pattern are called
-\emph{tagged} description. Formally, we have:
-%
-\[
+$\DSigma{\!}{\!}$ whose first argument enumerates the constructors. We
+call codes fitting this pattern \emph{tagged} descriptions. Again,
+this is a clear reminder of the sum-of-products style. Every
+description can be forced into this style with a singleton constructor
+if necessary. We characterise tagged descriptions thus:
+\[\stk{
+ \TagDesc : \Set \\
  \TagDesc \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{(\LAM{\_} \Desc)})
-\]
-%
-This is a clear reminiscence of the sum-of-products style,
-characterized by a finite sum of constructors. Every description can
-be expressed in this style, using a singleton constructor if
-necessary. Conversely, a tagged description can be turned into a
-full-blown description. We denote $\toDesc{\V{TD}}$ the description
-computed from the tagged description $\V{TD}$.
+\smallskip\\
+\toDesc{\!} : \TagDesc\To\Desc \\
+\toDesc{\!} \mapsto
+\spl{\LAM{\V{E}}\LAM{\V{D}}
+\DSigma{\EnumT{\V{E}}}{(\F{switch}\:\V{E}\:(\LAM{\_}\Desc)\:\V{D})}}
+}\]
 
@@
 \end{wstructure}
-
-For convenience, we have taken for granted the existence of $\Desc$,
-as presented in Figure~\ref{fig:desc_universe}. Although it makes no
-conceptual difference, this code should rather be read as a
-specification. We promise the existence of such objects, satisfying
-these typing rules. It will be the subject of
-Section~\ref{sec:desc-levitate} to actually implement this
-specification.
 

### 2010-04-02 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: uniformize titles.

@@
-\section{A universe of inductive datatypes}
+\section{A Universe of Inductive Datatypes}
 \label{sec:universe-desc}
@@
 
-\subsection{A universe of datatype descriptions}
+\subsection{The universe of descriptions}
 \label{sec:desc-universe}
@@
 \sqr{\NatZero}:\descop{\NatD}Z \qquad
-\sqr{\NatSuc\:z}:\descop{\NatD}Z \;\mbox{if}\;z:Z
+\sqr{\NatSuc{z}}:\descop{\NatD}Z \;\mbox{if}\;z:Z
 \]
@@
 }\]
+It is not a great stretch to imagine that the traditional datatype
+declaration syntax might desugar to the \emph{definition} of a datatype
+via a tagged description.
 
@@
 
-So far, we have used our universe of descriptions to build pattern
-functors. Being polynomial functors, they all admit a fixpoint. We
-construct the fixpoint by \emph{tying the knot} of the
-interpretation: the holes left open by the functor are filled by its
-own recursive definition:
+So far, we have built pattern functors with our \(\Desc\) universe.
+Being polynomial functors, they all admit a least fixpoint, which we
+now construct by \emph{tying the knot}: the element type abstracted
+by the functor is now instantiated recursively:
 %
 \[
-\Rule{\Gamma \vdash \Bhab{\V{D}}{\Desc}}
-     {\Gamma \vdash \Bhab{\Mu{\V{D}}}{\Set}} \qquad
-\Rule{\Gamma \vdash \Bhab{\V{D}}{\Desc} \quad 
-      \Gamma \vdash \Bhab{\V{x}}{\descop{\V{D}}{(\Mu{\V{D}})}}}
-     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\Mu{\V{D}}}}
+\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc}}
+     {\Gamma \vdash \Bhab{\Mu{\M{D}}}{\Set}} \qquad
+\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc} \quad 
+      \Gamma \vdash \Bhab{\M{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}}
+     {\Gamma \vdash \Bhab{\Con{\M{d}}}{\Mu{\M{D}}}}
 \]
@@
 
-Together with the type formers defined above, we would legibly expect
-an elimination principle. Following a categorical intuition, we could
-be tempted to provide it as a catamorphism:
+We can now build datatypes and their elements, e.g.:
+\[\stk{
+\Nat\mapsto\Mu{(\toDesc{\pair{\sqr{\NatZero\: \NatSuc{\!} }}
+                     {\sqr{ \DUnit \quad (\DIndx{\DUnit})}}{}}:\Set
+\\
+\Con{\sqr{\NatZero}}:\Nat \qquad
+\Con{\sqr{\NatSuc{n}}}:\Nat \;\mbox{if}\;n:\Nat}
+}\]
+
+But how shall we compute with our data?  We should expect an
+elimination principle. Following a categorical intuition, we might
+provide the `fold', or `iterator', or `catamorphism':
 %
@@
 \]
-%
-Whereas this definition is amply sufficient in the simply-typed world,
-it comes short of its promises in the dependently-typed realm. Indeed,
-this type is not dependent, being restricted to a notion of
-\emph{iteration}. As this operator will be the corner stone of
-inductive definitions, we are better off sharpening it before use.
-Following \citet{benke:universe-generic-prog}, we adopt the following
-definition:
+
+However, iteration is inadequate for \emph{dependent} computation.
+We need \emph{induction} to write functions whose type depends on inductive
+data. Following \citet{benke:universe-generic-prog}, we adopt the following:
 %
 \[\stk{
-\begin{array}{lcl}
-\F{induction} & : & \PITEL{\V{D}}{\Desc}
+\F{ind} : \stk{ \PITEL{\V{D}}{\Desc}
                     \PI{\V{P}}{\Mu{\V{D}} \To \Set}         \\
-              &   & (\PI{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}              \\
-              &   & \ \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{xs}} \To \V{P} (\Con{\V{xs}})) \To \\
-              &   & \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x} 
-\end{array} \\
-\F{induction}\: \V{D}\: \V{P}\: \V{ms}\: (\Con{\V{xs}}) =  \\
-\qquad
-    \V{ms}\: \V{xs}\: (\all{D}{\Mu{D}}{P}
-                           {(\LAM{\V{x}} \F{induction}\: \V{D}\: \V{P}\: \V{ms}\: \V{x})}
-                           {\V{xs}})
+               (\PI{\V{d}}{\descop{\V{D}}{(\Mu{\V{D}})}}       
+                \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{d}} \To \V{P} (\Con{\V{d}})) \To \\
+               \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x} 
+} \\
+\F{ind}\, \V{D}\, \V{P}\, \V{m}\, (\Con{\V{d}}) = 
+    \V{m}\, \V{d}\, (\all{D}{(\Mu{D})}{\V{P}}
+                           {(\F{ind}\, \V{D}\, \V{P}\, \V{m})}
+                           {\V{d}})
 }\]
 %
-Where, intuitively, $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$ states that
-$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Its
-definition is presented in Figure~\ref{fig:all-predicates}. Note that
-this induction principle is the first manifestation of a generic
-operation over descriptions. Any datatype we define automatically
+Here, $\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}$ states that
+$\Bhab{\V{P}}{\V{X} \To \Set}$ holds for every subobject $\V{x}:\V{X}$ in \(\V{D}\), and \(\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}\) is a
+`dependent map', applying
+some \(\Bhab{\V{p}}{\PIS{\Bhab{\x}{\V{X}}}\V{P}\,\V{x}}\) to each \(\x\)
+contained in \(\V{d}\). The full
+definition (including an extra case, introduced shortly) is presented in
+Figure~\ref{fig:all-predicates}. Note that
+\(\F{ind}\) is our first generic
+operation over descriptions, albeit a hardwired operator.
+Any datatype we define automatically
 comes with an induction principle.
@@
     \Unit \\
-\All{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
-    \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
-\All{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
-    \TIMES{\V{P}\: \V{a}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}} \\
-\All{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{f}}{\V{b}}{}} &
-    \TIMES{(\PI{\V{i}}{\V{H}} \V{P}\: (\V{f}\: \V{i}))}
-          {\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}}
+\All{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{s}}{\V{d}}{}} &
+    \All{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{d}} \\
+\All{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{x}}{\V{d}}{}} &
+    \TIMES{\V{P}\: \V{x}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}} \\
+\All{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{f}}{\V{d}}{}} &
+    \TIMES{(\PI{\V{h}}{\V{H}} \V{P}\: (\V{f}\: \V{h}))}
+          {\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}
 \end{array}
@@
                              \PITEL{\V{P}}{\V{X} \To \Set} \\
-                           & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
+                           & \PITEL{\V{p}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
                              \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
@@
 \begin{array}{@{}l@{}l@{\:=\:\:}l}
-\all{\DUnit}{& \V{X}}{\V{P}}{\V{R}}{\Void} &
+\all{\DUnit}{& \V{X}}{\V{P}}{\V{p}}{\Void} &
     \Void \\
-\all{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} &
-    \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
-\all{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} &
-    \pair{\V{R}\: \V{a}}
-         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{b}}}{} \\
-\all{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\V{R}}{\pair{\V{f}}{\V{b}}{}} &
-    \pair{\LAM{\V{i}} \V{R}\: (\V{f}\: \V{i})}
-         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{b}}}{}
+\all{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{s}}{\V{d}}{}} &
+    \all{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{p}}{\V{d}} \\
+\all{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{x}}{\V{d}}{}} &
+    \pair{\V{p}\: \V{x}}
+         {\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}{} \\
+\all{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{f}}{\V{d}}{}} &
+    \pair{\LAM{\V{h}} \V{p}\: (\V{f}\: \V{h})}
+         {\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}{}
 \end{array}
@@
 
-\caption{Induction predicates}
+\caption{Defining and collecting inductive hypotheses}
 \label{fig:all-predicates}
@@
 We have now enough machinery to build and manipulate inductive
-types. In a word, we ought be able to \emph{program} over these
-datatypes.  However, one could argue that programming with these
-objects is not practical.
-
-Writing codes instead of datatype constructor is indeed unpractical.
-However, regaining this ability is simply a matter of type
-propagation. We extend the type-checking framework of
-Figure~\ref{fig:type-checking} with the following inference rule:
+types at a low level. Let us now apply cosmetic surgery
+to the syntactic overhead. We extend type checking of
+expressions:
 %
 \[
-\Rule{\propag{\push{\widetilde{\V{e}}}
-                   {\descop{\V{f}\: \V{C}}{(\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})})}}}
-            {\widetilde{\V{e}}'}}
-     {\propag{\push{\V{C}\: \vec{\V{e}}}
-                   {\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})}}}
-             {\Con{(\etag{\V{C}}, \widetilde{\V{e}}')}}}\;\etag{\V{C}} \in \V{b}
+\Rule{\Gamma\Vdash\propag{\push{\etag{c}}{\EnumT{\M{E}}}}{n} \quad
+\Gamma\Vdash  
+\propag{
+  \push{\sqr{\vec{t}}}
+   {\descop{\M{D}\: n}{(\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})})}}}
+        {\M{t'}}}
+     {\Gamma\Vdash\propag{\push{\etag{c}\: \vec{\M{t}}}
+                   {\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})}}}
+             {\Con{\pair{n}{\M{t'}}{}}}}
 \]
 %
-Where $\vec{\V{e}}$ denotes a telescope of arguments applied to the
-constructor $\V{C}$. For readability, we write $\V{f}\: \V{C}$ instead
-of a call to $\F{switch}$. Also, we note $\widetilde{\V{e}}$ the
-right-nested tuple terminated by $\Void$ corresponding to the curried
-telescope $\vec{\V{e}}$.
+Here $\etag{c}\:\vec{t}$ denotes a tag `applied' to a sequence
+of arguments, and \(\sqr{\vec{t}}\) that sequence's repackaging as
+a right-nested tuple. Now we can just write data directly.
+\[
+\NatZero:\Nat\qquad \NatSuc{n}:\Nat\;\mbox{if}\;n:\Nat
+\]
+Once again, the type explains the legible presentation, as well as
+the low-level representation.
 
-Hence, in the high-level expression language, the developer writes a
-familiar type constructor, applied to a telescope of arguments. During
-type propagation, this expression will be pushed to a $\Mu{}$ type of
-a tagged description. It is then straightforward to look-up the tag
-and turn the constructor form into the appropriate code.
+We may also simplify appeals to induction by type propagation, as we
+have done with functions from pairs and enumerations.
+\[
+\Rule{\Gamma\Vdash\stk{
+\propag{\push{\M{f}}
+             {\PI{\V{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}       
+                \All{\M{D}}{(\Mu{\M{D}})}{(\PLAM{\x}{\Mu{\M{D}}}\M{P})}{\V{d}} \To \M{P}[\Con{\V{d}}/\x]\\}}
+       {\M{f'}}}}
+{\Gamma\Vdash
+\propag{\push{\sind{\M{f}}}
+             {\PIS{\Bhab{\x}{\Mu{\M{D}}}}\M{P} }}
+       {\F{ind}\:\M{D}\:(\PLAM{\x}{\Mu{\M{D}}}\M{P})\:\M{f'}}}
+\]
+This abbreviation is no substitute for the dependent pattern
+matching to which we are entitled in a high-level language built on top
+of this theory~\cite{goguen:pattern-matching}, but it does at least make
+`assembly language' programming mercifully brief, if hieroglyphic.
+\[\stk{
+\F{plus}:\Nat\To\Nat\To\Nat \\
+\F{plus}\mapsto\sind{\spl{\sqr{(\LAM{\_}\LAM{\_}\LAM{\V{y}}\V{y})
+  \quad(\LAM{\_}\spl{\LAM{\V{h}}\LAM{\_}\LAM{\V{y}}
+    \NatSuc{(\V{h}\:\V{y})}})}}}
+}\]
 
-This concludes our presentation of the universe of descriptions. In
-this section, we have seen how datatypes from the simply-typed world
-can be expressed in a dependently-typed setting. By relying on a
-universe, we benefit from the ability to inspect and manipulate
-codes. We have shown that programming with codes is practical.
+This concludes our introduction to the universe of datatype descriptions.
+We have encoded sum-of-products datatypes from the simply-typed world
+as data and equipped them with computation. We have also made sure
+to hide the details by type propagation.

### 2010-04-02 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: suppress line break before equations: we are at 11.5 pages
Desc@ICFP: s/~/\:/ in equations
Desc@ICFP: uniformize titles.
Desc@ICFP: uniformize coding style

@@
-\section{Levitating the universe of descriptions}
+\section{Levitating the Universe of Descriptions}
 \label{sec:desc-levitate}
 
-In this section, we will fulfil our promises. We first implement
-finite sets. Then, we implement the codes of the $\Desc$ universe.
-Despite being a perilous pedagogical exercise for the authors, we hope
-to convey to the reader the dizzy feeling of levitation, without the
-falling.
+In this section, we will fulfil our promises and show how we implement
+the signatures, first for the enumerations, and then for the codes of
+the $\Desc$ universe.  Persuading this to perform was a perilous
+pedagogical peregrination for the protagonist.  Our method was indeed to
+hardwire constants implementing the signatures specified above, in the
+first instance, but then attempt to replace them, step by step, with
+\emph{definitions}: ``Is \(2+2\) still \(4\)?'', ``No, it's a loop!''.
+But we did find a way, so now we hope to convey to the reader the
+dizzy feeling of levitation, without the falling.
 
-\subsection{Implementing finite sets}
+\subsection{Implementing finite enumerations}
 
@@
 
-In Section~\ref{sec:finite-sets}, we have given a specification of
-finite sets. We are going to implement the $\EnumU$ type former and
-its constructors. Let us recall their specification:
-
-\[\stkc{
-%% EnumU
-\Rule{\Gamma \vdash \Valid}
-     {\Gamma \vdash \Type{\EnumU}} 
-\\
-%% NilE
-\Rule{\Gamma \vdash \Valid}
-     {\Gamma \vdash \Bhab{\NilE}{\EnumU}} 
+In Section~\ref{sec:finite-sets}, we specified the
+finite sets of tags. We are going to implement the $\EnumU$ type former and
+its constructors. Recall:
+%
+\[\Type{\EnumU}\qquad\Bhab{\NilE}{\EnumU}\qquad
+\Bhab{\ConsE{(\Bhab{\M{t}}{\UId})}{(\Bhab{\M{E}}{\EnumU})}}{\EnumU}
+\]
+%
+The $\NilE$ and $\ConsE{\!}{\!}$ constructors are just the `nil' and
+`cons' or ordinary lists, with elements from \(\UId\).
+Therefore, we
+implement:
+%
+\[
+\EnumU \mapsto \Mu{(\ListD\: \UId)}
 \qquad
-%% ConsE
-\Rule{\Gamma \vdash \Bhab{\V{t}}{\UId} \quad
-      \Gamma \vdash \Bhab{\V{e}}{\EnumU}}
-     {\Gamma \vdash \Bhab{\ConsE{\V{t}}{\V{e}}}{\EnumU}}
-}\]
-
-The names $\NilE$ and $\ConsE{\!}{\!}$ are not a coincidence. $\EnumU$
-is indeed isomorphic to a list of $\UId$s. Therefore, the actual
-implementation of $\EnumU$ boils down to the following definition:
-
-\[\stk{
-\EnumU : \Set \\
-\EnumU \mapsto \Mu{(\ListD~\UId)}
-}\]
+\NilE \mapsto \ListNil
+\qquad
+\ConsE{\M{t}}{\M{E}} \mapsto \ListCons{\M{t}}{\M{E}}
+\]
 
@@
 
-Let us consider the consequences. First, we discover that the type
-theory does not need to be extended with the type former $\EnumU$, the
-constructors $\NilE$ and $\ConsE$. $\EnumU$ is defined above, as the
-fixpoint of $\ListD~\UId$. $\NilE$ and $\ConsE$ are simply the type
-constructors of lists, $\ListNil$ and $\ListCons$.
+Let us consider the consequences. We discover that the type
+theory does not need to be extended with a special type former $\EnumU$,
+or special constructors $\NilE$ and $\ConsE$. 
+Moreover, the $\spi{\M{E}}{\M{P}}$ operator, computing tuple types
+of \(\M{P}\)s by recursion on \(\M{E}\) need not be hardwired: we can
+just use the generic $\F{ind}$ operator, as we would for any ordinary
+program.
 
-Another interesting consequence concerns the $\spi{\!}{\!}$ operator:
-it can actually be implemented with the generic $\F{induction}$
-principle. From an implementation perspective, this means that this
-operator is implemented in the dependently-typed target language, as
-any other function.
+Note, however, that the universe decoder \(\EnumT{\M{E}}\) \emph{is}
+hardwired, as are the primitive \(\Ze\) and \(\Su{-}\) constructs that
+we use for low-level values, and indeed the \(\F{switch}\) operator.
+We cannot dispose of data altogether! We have, however, gained
+the ordinariness of the enumeration \emph{codes}, and hence of
+programs which manipulate them. Our next step is similar: we are going to
+condense the entire naming scheme of datatypes \emph{into itself}.
 
@@
 
-In this section, we have replaced a low-level presentation of finite
-sets by a self-hosted one, expressed in the universe of
-descriptions. However, formally, the content of the type theory
-remains unchanged: objects that were present before the modification
-are still there. Conversely, we have not introduced any spurious
-object.
+%In this section, we have replaced a low-level presentation of finite
+%sets by a self-hosted one, expressed in the universe of
+%descriptions. However, formally, the content of the type theory
+%remains unchanged: objects that were present before the modification
+%are still there. Conversely, we have not introduced any spurious
+%object.
 
-If not on the content, this modification had an effect on the
-names. The type naming scheme of the type theory has condensed: named
-type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
-now replaced by codes and their fixpoint in the universe of
-descriptions. In essence, our next step is similar: we are going to
-condense the entire naming scheme of datatypes \emph{in itself}.
+%If not on the content, this modification had an effect on the
+%names. The type naming scheme of the type theory has condensed: named
+%type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
+%now replaced by codes and their fixpoint in the universe of
+%descriptions. In essence,
 
@@
 
-We shall now fulfil our promise concerning $\Desc$. We are going to
-implement the codes of the universe of descriptions. In and by itself,
+We shall now fulfil our implementation promises, delivering the
+codes of the universe of descriptions. In and by itself,
 the codes, $\Desc$, is nothing but a datatype. We are in the same
@@
 Our first attempt is the following:
-
+%
 \[\stk{
@@
 }\]
-
+%
 Let us explain how we have proceeded and the obstacle we face. Much as
@@
 $\DSigma{\!}{\!}$:
-
-\[    \DSigma{\!}{\!} : \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc      \]
-
+%
+\[
+      \DSigma{\!}{\!} : \PITEL{\V{S}}{\Set}
+                        \PI{\V{D}}{\V{S} \To \Desc} 
+                        \Desc      
+\]
+%
 So, we first pack an element $\V{S}$ of $\Set$. Then, we would like to
@@
 arguments:
-
+%
 \[\stk{
@@
 We are now able to describe our universe of datatypes:
-
+%
 \[\stk{
@@
 }\]
-
+%
 The $\DUnit$ and $\DIndx{\!}$ cases remain unchanged, as expected. We
@@
 What happens if we unfold the definition? We ought to build the following term:
-
+%
 \[
@@
 \]
-
+%
 But this is quite problematic. We are still in the process of
@@
 special-purpose operator:
-
+%
 \[
@@
 \]
-
+%
 The entire work of the magician stands here, in this extension. Note
@@
 rule:
-
+%
 \[
@@
 
+\newcommand{\cata}{\F{cata}}
+
 The catamorphism is defined by induction on the description $\V{D}$,
@@
 $\V{f}$. Let us take a look at this jigsaw:
-
-\newcommand{\cata}{\F{cata}}
-
+%
 \[\stk{
@@
 }\]
-
+%
 We are left with filling the hole. Recall that we have
@@
 \(\V{xs}\) by their counterpart in \(\V{hs}\):
-
+%
 \[\stk{
@@
     \Void                                                                                                         \\
-\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
-    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
-\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto       \\
-\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                     \\
-\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{xs}}{}\: \pair{\V{g}}{\V{ys}}{} \mapsto   \\
-\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
+\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}             \mapsto
+    \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                \\
+\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} \mapsto        \\
+\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                      \\
+\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d'}}{}\: \pair{\V{g}}{\V{d'}}{} \mapsto   \\
+\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
 }\]
-
+%
 Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
@@
 
-\subsection{The generic Free Monad}
+\subsection{The generic free monad}
 \label{sec:desc-free-monad}
@@
 free monad over $F$ is defined by the following datatype:
-
+%
 \[
@@
 \]
-
+%
 Being an inductive type, this $\FMFreeMonad$ datatype is itself
 defined by a pattern functor. It is given by:
-
+%
 \[
@@
 function is surprisingly easy:
-
+%
 \[\stk{
@@
 }\]
-
+%
 We simply add a constructor, $\DVar{\!}$, and define its argument to
@@
 
+\newcommand{\subst}{\F{subst}}
+\newcommand{\apply}{\F{apply}}
+
 Our implementation will appeal to the $\cata$ function developed
@@
 arguments to $\cata$ as possible:
-
-\newcommand{\subst}{\F{subst}}
-\newcommand{\apply}{\F{apply}}
-
+%
 \[\stk{
@@
            \PI{\V{X}, \V{Y}}{\Set} \\
-         & \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
-           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To
+         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To
+           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
            \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
 \end{array} \\
-\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{x}\: \V{\sigma} \mapsto
+\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{x} \mapsto
   \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
@@
 }\]
-
+%
 We are left with implementing the algebra of the
@@
 corresponds to the following definition:
-
+%
 \[\stk{
@@
 \apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
-\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{ys}}{} & \Con{\pair{\etag{t}}{\V{ys}}{}}
+\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{xs}}{} & \Con{\pair{\etag{t}}{\V{xs}}{}}
 \end{array}

### 2010-04-02 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: suppress line break before equations: we are at 11.5 pages
Desc@ICFP: uniformize titles.
Desc@ICFP: implement substI
Desc@ICFP: uniformize coding style

@@
-\section{Indexing descriptions}
+\section{A Universe of Inductive Families}
 \label{sec:indexing-desc}
@@
 
-\subsection{From Desc to IDesc}
+\subsection{The universe of indexed descriptions}
 
@@
 have the following types:
-
+%
 \[\stk{
@@
 \emph{family} of functors:
-
+%
 \[\stkl{
@@
 
+\newcommand{\inductionI}{\F{indI}}
+\newcommand{\cataI}{\F{cataI}}
+
 However, we still have to define the actual grammar. We obtain it by
@@
 by:
-
+%
 \[\stk{
 \begin{array}{@{}ll}
-\F{inductionI} : & _{\PI{\V{I}}{\Set}}
+\inductionI : & _{\PI{\V{I}}{\Set}}
                    \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
-                   \PI{\V{P}}{\SIGMA{\V{i}}{\V{I}} \IMu{\V{I}}{\V{R}}{\!} \To \Set} \\
+                   \PI{\V{P}}{\TIMES{\V{I}}{\IMu{\V{I}}{\V{R}}{\!}} \To \Set} \\
                  & (      \PITEL{\V{i}}{\V{I}} 
-                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{\IMu{\V{I}}{\V{R}}{\!}}} \\
-                 & \   \idescop{\AllI{\V{I}}
+                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{\!})}} \\
+                 & \   \idescop{\AllI{}
                                      {(\V{R}\: \V{i})}
@@
                                {\V{P}} \To
-                       \V{P}\: \pair{\V{i}}{\Con{\V{xs}}}{}) \\
+                       \V{P}\: \pair{\V{i}}{\Con{\V{xs}}}{}) \To \\
                  & \PITEL{\V{i}}{\V{I}}
@@
 \end{array} \\
-\F{inductionI}\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: (\Con{\V{xs}}) =  \\
-\quad
+\inductionI\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: (\Con{\V{xs}}) =  \\
+\qquad
     \V{ms}\: \V{i}\: \V{xs}\: (\allI{}
@@
                                     {\V{P}}
-                                    {(\LAM{\pair{\V{i}}{\V{xs}}{}} \F{inductionI}\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: \V{xs})}
+                                    {(\LAM{\pair{\V{i}}{\V{xs}}{}} \inductionI\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: \V{xs})}
                                     {\V{xs}})
 }\]
-
+%
 Where the operators $\AllI{}{\!}{\!}{}$ and
@@
 Figure~\ref{fig:allI-predicates}. As for descriptions, we can compute
-a generic catamorphism, $\F{cataI}$, from $\F{inductionI}$.
+a generic catamorphism, $\cataI$, from $\inductionI$.
 
@@
     \DProd{\AllI{}{\V{D}}{\V{X}}{\V{d}}}{\AllI{}{\V{D'}}{\V{X}}{\V{d'}}} \\
-\AllI{\:}{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\pair{\V{a}}{\V{b}}{}} &
-    \AllI{}{(\V{T}\: \V{a})}{\V{X}}{\V{b}} \\
-\AllI{\:}{(\DPi{\V{S}}{\V{T}})}{& \V{X}}{\V{f}} &
-    \DPi{\V{S}}{(\LAM{\V{s}} \AllI{}{(\V{T}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})})}
+\AllI{\:}{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\pair{\V{s}}{\V{d}}{}} &
+    \AllI{}{(\V{D}\: \V{s})}{\V{X}}{\V{d}} \\
+\AllI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{f}} &
+    \DPi{\V{S}}{(\LAM{\V{s}} \AllI{}{(\V{D}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})})}
 \end{array}
@@
                                   \PITEL{\V{X}}{\V{I} \To \Set} 
-                                  \PI{\V{R}}{\TIMES{\V{I}}{\V{X}} \To \Set} \\
-                                & (\PI{\V{x}}{\TIMES{\V{I}}{\V{X}}} \V{R}\: \V{x}) \To
-                                  \PI{\V{xs}}{\idescop{\V{D}}{}{\V{X}}} 
-                                  \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{\TIMES{\V{I}}{\V{X}}}{\V{R}}
+                                  \PI{\V{P}}{\TIMES{\V{I}}{\V{X}} \To \Set} \\
+                                & (\PI{\V{x}}{\TIMES{\V{I}}{\V{X}}} \V{P}\: \V{x}) \To
+                                  \PI{\V{xs}}{\idescop{\V{D}}{\V{I}}{\V{X}}} 
+                                  \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{\TIMES{\V{I}}{\V{X}}}{\V{P}}
 \end{array} \\
 \begin{array}{@{}l@{}l@{\:=\:\:}l}
-\allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{R}}{\V{P}}{\V{x}} &
-    \V{P}\: \pair{\V{i}}{\V{x}}{} \\
-\allI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{R}}{\V{P}}{\V{k}} &
+\allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{P}}{\V{p}}{\V{x}} &
+    \V{p}\: \pair{\V{i}}{\V{x}}{} \\
+\allI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{P}}{\V{p}}{\V{k}} &
     \V{k} \\
-\allI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\V{R}}{\V{P}}{(\V{d}, \V{d'})} &
-    \pair{\allI{}{\V{D}}{\V{X}}{\V{R}}{\V{P}}{\V{d}}}
-         {\allI{}{\V{D'}}{\V{X}}{\V{R}}{\V{P}}{\V{d'}}}{} \\
-\allI{\:}{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{R}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
-    \allI{}{(\V{T}\: \V{a})}{\V{X}}{\V{R}}{\V{P}}{\V{b}} \\
-\allI{\:}{(\DPi{\V{S}}{\V{T}})}{& \V{X}}{\V{R}}{\V{P}}{\V{f}} &
-    \allI{}{(\V{T}\: \V{a})}{\V{X}}{\V{R}}{\V{P}}{(\V{f}\: \V{a})}
+\allI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{d}}{\V{d'}}{}} &
+    \pair{\allI{}{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}
+         {\allI{}{\V{D'}}{\V{X}}{\V{P}}{\V{p}}{\V{d'}}}{} \\
+\allI{\:}{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{s}}{\V{d}}{}} &
+    \allI{}{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{p}}{\V{d}} \\
+\allI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\V{f}} &
+    \allI{}{(\V{D}\: \V{a})}{\V{X}}{\V{P}}{\V{p}}{(\V{f}\: \V{a})}
 \end{array}
@@
 \begin{array}{@{}l@{\:=\:\:}ll}
-\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}\: \V{i}                                           \\
-\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
-\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
-\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
-\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
+\idescop{\DVar{\V{i}}}{\V{I}}{\V{D}}      &  \V{X}\: \V{i}                                           \\
+\idescop{\DConst{\V{K}}}{\V{I}}{\V{X}}    &  \V{K}                                                   \\
+\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{X}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{X}}}{\idescop{\V{D'}}{\V{I}}{\V{X}}}       \\
+\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}                    \\
+\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}            
 \end{array}
@@
 the pattern functor of natural numbers:
-
+%
 \[\stk{
@@
 we can describe it in $\IDesc{\Unit}$:
-
+%
 \[\stk{
@@
 index $\NatZero$ and $\NatSuc$ respectively.:
-
+%
 \[
@@
 this case, we obtain the following definition:
-
+%
 \[\stk{
@@
 following, equivalent definition:
-
+%
 \[\stk{
@@
 }\]
-
+%
 The equations (and constructors) have simply disappeared. A similar
 example is $\Fin{\!}$, specified by:
-
+%
 \[
@@
 }\]
-
+%
 In this case, we can apply forcing, but not detagging, since both
 $\FinZero$ and $\FinSuc{\!}$ both target $\NatSuc{\!}$:
-
+%
 \[\stk{
@@
 possible to enable or disable constructors.
-
+%
 \[
@@
 consider two types, natural numbers and booleans:
-
+%
 \[
@@
 numbers. Informally, their type is the following:
-
+%
 \[
@@
 \]
-
+%
 The function $\Val{\!}$, used in the definition of $\Eval{\!}$, simply
@@
 $\F{leHost}$. We define $\Val{\!}$ as follows:
-
+%
 \[\stk{
@@
 evaluator is:
-
+%
 \[
@@
 \]
-
+%
 Where we overload the notation $\toIDesc{\V{TID}}$ to denote the
@@
 catamorphism. Indeed, implementing a single step of evaluation -- the
-algebra -- is sufficient, as $\F{cataI}$ gives, for free, the full
+algebra -- is sufficient, as $\cataI$ gives, for free, the full
 evaluator. The implementation is as follows:
-
+%
 \[\stk{
@@
 \evalOne\: & \_\: & \pair{\Econd{\!}{\!}{\!}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
-\evalOne\: & \_\: & \pair{\Econd{\!}{\!}{\!}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{x} \\
+\evalOne\: & \_\: & \pair{\Econd{\!}{\!}{\!}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{y} \\
 \evalOne\: & \Enat\: & \pair{\Eplus{\!}{\!}}{\pair{\V{x}}{\V{y}}{}}{}                     & \F{plusHost}\: \V{x}\: \V{y} \\
@@
            \Val{\V{ty}} \\
-\evalH\: \V{ty}\: \V{term} = \F{cataI}\: \Ty\: 
+\evalH\: \V{ty}\: \V{term} = \cataI_{\Ty}
                                  \toIDesc{\HExprD}\: 
@@
 variables. Following $\Val{\!}$, we define $\Var{}{}$ by:
-
+%
 \[\stk{
@@
 }\]
-
+%
 Whereas $\Val{\!}$ was mapping the type to the corresponding host type,
@@
 following function:
-
+%
 \[\stk{
@@
              & (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To
-               \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} 
+               \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}} 
 \end{array} \\
@@
 
-We are now left with applying $\discharge$ over all variables of the
-term. The type of this operation is the following:
-
 \newcommand{\substH}{\F{substExpr}}
@@
 
+We are now left with applying $\discharge$ over all variables of the
+term. The type of this operation is the following:
+%
 \[
@@
 \]
-
+%
 Where $\GammaSpi{\!}{\!}$ corresponds to a context, defined by:
-
+%
 \[\stk{
@@
 function that picks the context of the right type:
-
+%
 \[
@@
 \]
-
+%
 Where $\F{$\gamma$}_{\V{ty}}$ is short for
@@
 
-\subsection{Free IMonad}
+\subsection{Free indexed monad}
 
@@
   say something about IMonads in general before we get to this point?}
-
+%
 \[\stk{
@@
                {\LAM{\V{i}}
-                \pair{(\DConst{(\V{R}\: \V{i})})}
-                     {((\snd{\V{E}})\: \V{i})}{}}{}}
+                \pair{\DConst{(\V{R}\: \V{i})}}
+                     {(\snd{\V{E}})\: \V{i}}{}}{}}
          {\V{F}}{}
@@
 
+\newcommand{\substI}{\F{substI}}
+
 Just as in the universe of descriptions, this construction comes with
-an obvious \return\ and a substitution operation, the \bind. Its type
-signature is the following:
-
-\newcommand{\substI}{\F{substI}}
-
-\[
+an obvious \return\ and a substitution operation, the \bind. Its
+definition is the following:
+%
+\[\stk{
 \begin{array}{@{}ll}
-\substI : & \PITEL{\V{I}}{\Set}
-            \PITEL{\V{X}, \V{Y}}{\V{I} \To \Set}
-            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
-          & \PITEL{\V{\sigma}}{\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 
-                               \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}} \\
+\substI : & _{\PI{\V{I}}{\Set}}
+            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
+            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
+          & (\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 
+                               \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}) \To \\
           & \PITEL{\V{i}}{\V{I}}
@@
             \IMu{I}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}
-\end{array}
-\]
-
-We refer the reader to the Agda model for the actual implementation.
-
-\subsubsection{Examples}
+\end{array} \\
+\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} = \\
+\qquad    \cataI\: \toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}}\:
+                      (\IMu{}{\toIDesc{\FreeIMonad{\V{R}}{}{\V{Y}}}})\:
+                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
+                      \V{i}\:
+                      \V{t} 
+}\]
+% 
+Where  $\F{applyI}$ is defined as follow:
+%
+\[\stk{
+\begin{array}{@{}ll}
+\F{applyI} : & _{\PI{\V{I}}{\Set}}
+            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
+            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
+          & (\PI{\V{i}}{\V{I}} \V{X}\: \V{i} \To \IMu{\V{I}}{\toIDesc{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}}{\V{i}}) \To \\
+          & \PI{\V{i}}{\V{I}} 
+            \idescop{\toIDesc{(\FreeIMonad{\V{R}}
+                                         {\V{I}}
+                                         {\V{X}})}\: 
+                     \V{i}}
+                    {\V{I}}
+                    {\IMu{\V{I}}{\toIDesc{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}}} \To 
+           \IMu{\V{I}}{\toIDesc{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}}{\V{i}}
+\end{array} \\
+\begin{array}{@{}l@{\:\mapsto\:\:}l}
+\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
+\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\etag{t}\,}{\V{ys}}{} & \Con{\pair{\etag{t}}{\V{ys}}{}}
+\end{array}
+}\]
+
+Let us now consider two examples of free indexed monad.
 
@@
 $\HExprAFreeD$:
-
+%
 \[\stk{
@@
 have to fill in the right arguments to $\substI$, the type guiding us:
-
+%
 \[\stk{
@@
 \;\;\ \begin{array}{ll}
-       \substI\: & \Ty\: 
+       \substI_{\Ty}\: &  \HExprFreeD\: 
                          (\SUM{\Val{\!}}{\Var{\V{dom}}{}})\: 
-                  \Val{\!}\:
-                  \HExprFreeD\:  \\
+                         \Val{\!} \\
                       &
@@
 }\]
-
-Hence completing our implementation of the open terms interpreter. We
-have defined a well-typed language of arithmetical expressions, taking
-advantage of indexing. Then, we have implemented an evaluator for
-closed term, based on the generic catamorphism function. Using the
+%
+Hence completing our implementation of the open terms interpreter.
+
+We have defined a well-typed language of arithmetical expressions,
+taking advantage of indexing. Then, we have implemented an evaluator
+for closed term, based on the generic catamorphism function. Using the
 free monad construction, we have automatically derived the language of
@@
 is described as follow:
-
+%
 \[\stk{
@@
 }\]
-
+%
 Then, we get $\IDesc$ by building its free monad:
-
+%
 \[\stk{

### 2010-04-02 papers/icfp-2010-desc/paper_desc.tex
sec 3 done
Desc@ICFP: fix parenthesing typo
Desc@ICFP: uniformize coding style
4.1
Desc@ICFP: Desc code and interpretation back as a figure

@@
 To add inductive types to our type theory, we build a universe of
-datatype \emph{descriptions} by implementing the signature below, with
-codes mimicking the grammar of datatype declarations. We
-can read a description $\V{D}:\Desc$ as a `pattern functor'
-on $\Set$, with $\descop{\V{D}}{\!}$ its action on an object, \(\V{X}\), soon
-to be instantiated recursively.
-%
+datatype \emph{descriptions} by implementing the signature presented
+in Figure~\ref{fig:desc_universe}, with codes mimicking the grammar of
+datatype declarations. We can read a description $\V{D}:\Desc$ as a
+`pattern functor' on $\Set$, with $\descop{\V{D}}{\!}$ its action on
+an object, \(\V{X}\), soon to be instantiated recursively.
+
+Descriptions are sequential structures, terminated by $\DUnit$, indicating
+the empty tuple. To build
+sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
+$\Sigma$-type. To request a recursive component, we have $\DIndx{\V{D}}$,
+where \(\V{D}\) describes the rest of the node.
+
+\begin{figure}
 \[\stk{
@@
 \smallskip\\
-\begin{array}{l@{\V{Z}\:\mapsto\:}l}     
+\begin{array}{l@{\V{X}\:\mapsto\:}l}     
 \multicolumn{2}{l}{\descop{\_\:}{} : \Desc \To \Set \To \Set} \\
@@
  \descop{\DSigma{\V{S}}{\V{D}}}{} &
-     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{Z}}}  \\
-\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{Z}}{\descop{\V{D}}{\V{Z}}}
+     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{X}}}  \\
+\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
 \end{array}
 }\]
-%
-Descriptions are sequential structures, terminated by $\DUnit$, indicating
-the empty tuple. To build
-sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
-$\Sigma$-type. To request a recursive component, we have $\DIndx{\V{D}}$,
-where \(\V{D}\) describes the rest of the node.
+
+\caption{Universe of Descriptions}
+\label{fig:desc_universe}
+
+\end{figure}
 
@@
 \sqr{\NatZero}:\descop{\NatD}Z \qquad
-\sqr{\NatSuc{z}}:\descop{\NatD}Z \;\mbox{if}\;z:Z
+\sqr{\NatSuc{(\Bhab{z}{X})}}:\descop{\NatD}Z
 \]
@@
 
-\subsection{Fixpoint}
+\subsection{The least fixpoint}
 \label{sec:desc-fix-point}
@@
 \Nat\mapsto\Mu{(\toDesc{\pair{\sqr{\NatZero\: \NatSuc{\!} }}
-                     {\sqr{ \DUnit \quad (\DIndx{\DUnit})}}{}}:\Set
+                       {\sqr{ \DUnit \quad (\DIndx{\DUnit})}}{}})}:\Set
 \\
 \Con{\sqr{\NatZero}}:\Nat \qquad
-\Con{\sqr{\NatSuc{n}}}:\Nat \;\mbox{if}\;n:\Nat}
+\Con{\sqr{\NatSuc{(\Bhab{n}{\Nat})}}}:\Nat
 }\]
@@
 \F{ind}\, \V{D}\, \V{P}\, \V{m}\, (\Con{\V{d}}) = 
-    \V{m}\, \V{d}\, (\all{D}{(\Mu{D})}{\V{P}}
+    \V{m}\, \V{d}\, (\all{\V{D}}{(\Mu{\V{D}})}{\V{P}}
                            {(\F{ind}\, \V{D}\, \V{P}\, \V{m})}
@@
 comes with an induction principle.
+
+We note that the very same functors \(\descop{\M{D}}{\!}\) also admit greatest
+fixpoints, and we have indeed implemented coinductive types this way, but
+that is a story for another time.
 
@@
 \[
-\NatZero:\Nat\qquad \NatSuc{n}:\Nat\;\mbox{if}\;n:\Nat
+\NatZero:\Nat\qquad \NatSuc{(\Bhab{n}{\Nat})}:\Nat
 \]

### 2010-04-02 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix idesc typo.
Desc@ICFP: damned colors.

@@
                                     {\V{R}\: \V{i}}
-                                    {\IMu{\V{I}}{R}{}}
+                                    {\IMu{\V{I}}{\V{R}}{}}
                                     {\V{P}}
@@
 \idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
-\begin{array}{@{}l@{\:=\:\:}ll}
-\idescop{\DVar{\V{i}}}{\V{I}}{\V{D}}      &  \V{X}\: \V{i}                                           \\
-\idescop{\DConst{\V{K}}}{\V{I}}{\V{X}}    &  \V{K}                                                   \\
-\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{X}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{X}}}{\idescop{\V{D'}}{\V{I}}{\V{X}}}       \\
-\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}                    \\
-\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}            
+\begin{array}{@{}l@{\V{X}}@{\:=\:\:}ll}
+\idescop{\DVar{\V{i}}}{\V{I}}{&}      &  \V{X}\: \V{i}                                           \\
+\idescop{\DConst{\V{K}}}{\V{I}}{&}    &  \V{K}                                                   \\
+\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{&} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{X}}}{\idescop{\V{D'}}{\V{I}}{\V{X}}}       \\
+\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{&} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}                    \\
+\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{&}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}            
 \end{array}
@@
           & \PITEL{\V{i}}{\V{I}}
-            \PITEL{\V{D}}{\IMu{I}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{X}}})}{\V{i}}} \To
-            \IMu{I}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}
+            \PITEL{\V{D}}{\IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{X}}})}{\V{i}}} \To
+            \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}
 \end{array} \\

### 2010-04-02 papers/icfp-2010-desc/paper_desc_levitation.tex
4.1
Levitato!

@@
 Note, however, that the universe decoder \(\EnumT{\M{E}}\) \emph{is}
-hardwired, as are the primitive \(\Ze\) and \(\Su{-}\) constructs that
+hardwired, as are the primitive \(\Ze\) and \(\Su{\!}\) that
 we use for low-level values, and indeed the \(\F{switch}\) operator.
 We cannot dispose of data altogether! We have, however, gained
-the ordinariness of the enumeration \emph{codes}, and hence of
+the ordinariness of the enumeration \emph{codes}, and hence of generic
 programs which manipulate them. Our next step is similar: we are going to
@@
 
-We shall now fulfil our implementation promises, delivering the
-codes of the universe of descriptions. In and by itself,
-the codes, $\Desc$, is nothing but a datatype. We are in the same
-situation as with $\EnumU$: we ought to be able to describe the codes
-of $\Desc$ in $\Desc$ itself. Hence, this code would be a first-class
-citizen, born with the standard, generic equipment of datatypes.
+We shall now fulfil our implementation promises, encoding the universe
+of descriptions. In and of itself, the codes, $\Desc$, is nothing but
+a datatype. We are in the same situation as with $\EnumU$: we ought to
+be able to describe the codes of $\Desc$ in $\Desc$ itself. Hence,
+this code would be a first-class citizen, born with the standard,
+generic equipment of datatypes.
 
@@
 
-Our first attempt is the following:
+Our first attempt gets stuck quite quickly:
 %
@@
 \DescD : \Desc \\
-\begin{array}{@{}ll}
-\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
-                               & \left[\begin{array}{l}
-                                   \DUnit                                \\
+\DescD \mapsto
+  \tada{\stkm{\DUnit\\ \DSigma{\!}{\!}\\ \DIndx{\!}}}
+       {\stkm{ \DUnit                                \\
          \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
-                                   \DIndx{\DUnit}                        \\
-                                 \end{array}
-                                 \right]
-\end{array}
+         \DIndx{\DUnit}    }                    } \\
+%\begin{array}{@{}ll}
+%\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
+%                               & \left[\begin{array}{l}
+%                                   \DUnit                                \\
+%                                   \DSigma{\Set}{\LAM{\V{S}} \SHED}      \\
+%                                   \DIndx{\DUnit}                        \\
+%                                 \end{array}
+%                                 \right]
+%\end{array}
 }\]
 %
-Let us explain how we have proceeded and the obstacle we face. Much as
-the datatypes we have seen so far, we first choose a constructor among
+Let us explain where we stand. Much as
+we have done so far, we first offer a constructor choice from
 $\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
-that the tagged notation we have used for the $\Desc$ constructors now
+that the `tagged' notation we have used for the $\Desc$ constructors now
 fully makes sense: these were actually the tags we are defining.
-
 For $\DUnit$, we immediately reach the end of the description. For
@@
 \[
-      \DSigma{\!}{\!} : \PITEL{\V{S}}{\Set}
-                        \PI{\V{D}}{\V{S} \To \Desc} 
-                        \Desc      
+\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
 \]
 %
-So, we first pack an element $\V{S}$ of $\Set$. Then, we would like to
-express a notion of recursive argument \emph{indexed} by $\V{S}$. Because
-our presentation is entirely first-order so far, we are not able to
-express this notion. We shall remedy to this situation.
+So, we first pack a  $\Set$, $\V{S}$. We should then like
+a recursive argument \emph{indexed} by $\V{S}$, but
+that is an \emph{exponential}, and our presentation is entirely
+first-order so far, delivering only sums-of-products. To code our
+universe, we must first enlarge it!
 
@@
 
-In order to capture a notion of higher-order induction, we need to
-extend our universe of descriptions. This consists in adding a code
+In order to capture a notion of higher-order induction, we add a code
 $\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. Intuitively,
-$\DHindx{\!}{\!}$ uses the elements of $H$ to index as many recursive
-arguments:
+$\DHindx{\!}{\!}$ gives a recursive subobject for each element of $H$.
 %
 \[\stk{
-\begin{array}{ll}
-\stk{
-\data \Desc : \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \DHindx{\!}{\!} & \PI{\V{H}}{\Set} \Desc \To \Desc \\
-    \ldots          & \:\:\ldots
-\end{array}
-}
-\vspace{0.2in}
-\\
-\stk{
-\descop{\_\:}{} : \Desc \To \Set \To \Set \\
-\begin{array}{@{}l@{\:=\:\:}ll}
-\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
-\ldots                                    &  \ldots 
-\end{array}
-}
-\end{array}
+    \DHindx{(\Bhab{\M{H}}{\Set})}{(\Bhab{\M{D}}{\Desc})} : \Set \smallskip \\
+\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     \mapsto \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
 }\]
 
-Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
-we tolerate this duplication. Indeed, unlike its counterpart,
-$\DIndx{\!}$ is first-order. Both codes are \emph{extensionally} the
-same but, in practice, they will be dealt with rather
-differently. While definitional equality can cope with first-order
-objects, the functional presentation introduced by $\DHindx{\!}{}$ is
-unlikely to be amenable to a purely definitional treatment.
+Note that up to isomorphism, $\DIndx{\!}$ is subsumed by
+$\DHindx{\Unit}{\!}$. However, the apparent duplication has some
+value.  Unlike its counterpart, $\DIndx{\!}$ is first-order: we prefer
+not to demand dummy functions from \(\Unit\) in ordinary data,
+e.g. \(\NatSuc{(\LAM{\_}n)}\). It is na{\"\i}ve to imagine that up to
+isomorphism, any representation of data will do.  First-order
+representations are finitary by construction, and thus admit a richer,
+componentwise decidable equality than functions may in
+general possess.\footnote{E.g., extensionally, there is one inhabitant of
+\(\EnumT{\void}\To\Nat\); intensionally, there is a countable infinitude
+which it is not safe to collapse.}
 
@@
 \DescD : \Desc \\
-\begin{array}{@{}ll}
-\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!}, \DHindx{\!}{\!} ]}) \\
-                           & \left[\begin{array}{l}
+\DescD \mapsto\tada
+{\stkm{ \DUnit\\ \DSigma{\!}{\!}\\ \DIndx{\!}\\ \DHindx{\!}{\!}}}
+{\stkm{
 \DUnit                                            \\
-                                   \DSigma{\Set}{(\LAM{\V{S}} \DHindx{\V{S}}{\DUnit})}   \\
+                                   \DSigma{\Set}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
                                    \DIndx{\DUnit}                                    \\
-                                   \DSigma{\Set}{(\LAM{\V{H}} \DIndx{\DUnit})}
-                                   \end{array}
-                             \right]
-\end{array}
+                                   \DSigma{\Set}{\LAM{\_} \DIndx{\DUnit}}
+}}
 }\]
@@
 the higher-order induction on $\V{S}$. The $\DHindx{\!}{\!}$ case
-consists in packing an $\V{H}$ in $\Set$ with a recursive argument.
+consists in packing a $\Set$ with a recursive argument.
 
 At a first glance, we have achieved our goal. We have described the
-codes of the universe of description. Taking the fixpoint of this
-object gives us $\Desc$, up to isomorphism. We have implemented the
-codes of datatypes as an object levitating inside itself. However,
-this levitation operation, just as any magic trick, relies on an
-invisible cable. Let us reveal it, hence finishing our implementation.
+codes of the universe of descriptions. Taking the fixpoint of \(\descop{\DescD}\)
+gives us a datatype exactly like $\Desc$. Might we be
+so bold as to take \(\Desc \mapsto \Mu{\DescD}\) as the levitating
+definition? If we do, we shall come down with a bump! 
+To complete our levitation, just as in the magic trick, requires
+hidden assistance. Let us explain the problem and reveal the `invisible
+cable' which fixes it.
 
@@
 
-Our reader might be slightly confused to learn that the trick is
-visible in the definition of $\DescD$. Or rather, it is made invisible
-by careful usage of type propagation. Indeed, let us try to elaborate
-this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
-construct elaborates to a finite set $\V{E}$ in $\EnumU$, inhabited by
-the codes. Then, we can type-check the case definition, between square
-brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
-$\EnumT{\V{E}} \To \Mu{\DescD}$. This corresponds to a finite function
-definition, as presented in Figure~\ref{fig:type-checking}.
+The definition \(\Desc \mapsto \Mu{\DescD}\) is circular,
+but the offensive recursion is concealed by a prestidigitation.
+Expanding be \(\toDesc{-}\) and propagating types as in
+Figure~\ref{fig:type-checking} reveals the
+awful truth:
+\[
+\Desc\mapsto
+\Mu{\stk{(\DSigma{\EnumT{\sqr{ \DUnit\: \DSigma{\!}{\!}\: \DIndx{\!}\: 
+    \DHindx{\!}{\!}}}\\}
+    {\;\F{switch}\:\sqr{ \DUnit\: \DSigma{\!}{\!}\: \DIndx{\!}\: 
+    \DHindx{\!}{\!}}\:(\LAM{\_}\Desc)\\
+\;\sqr{\stkm{
+\DUnit                                            \\
+                                   \DSigma{\Set}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
+                                   \DIndx{\DUnit}                                    \\
+                                   \DSigma{\Set}{\LAM{\_} \DIndx{\DUnit}}
+}}})}}
+\]
+The recursion shows up only because we must specify the return type
+of the general-purpose \(\F{switch}\), and it is computing a \(\Desc\)!
+Although type propagation
+allows us to hide this detail \emph{when defining a function}, we cannot
+readily suppress this information and check types
+when \(\F{switch}\) is fully applied.
 
@@
 
-What happens if we unfold the definition? We ought to build the following term:
+%What happens if we unfold the definition? We ought to build the following term:
+%
+%\[
+%\PLAM{\V{x}}{(\EnumT{\V{E}})} \switch{\V{E}}{(\LAM{\_} %\Mu{\DescD})}{\V{\pi^f}}{\V{x}}
+%\]
+%
+%But this is quite problematic. We are still in the process of
+%constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
+%begging for this very same $\DescD$. Despite our willingness, we
+%cannot materialise such motive. However, we perfectly know what the
+%motive is.
+
+We are too close to give up now. If only we did not need to
+supply that return type, especially when we know what it must be.
+We eliminate the recursion by \emph{specializing} \(\F{switch}\):
 %
 \[
-\PLAM{\V{x}}{(\EnumT{\V{E}})} \switch{\V{E}}{(\LAM{\_} \Mu{\DescD})}{\V{\pi^f}}{\V{x}}
+\F{switchD} : \PITEL{\V{E}}{\EnumU}   \To
+                (\spi{\V{E}}{\LAM{\_} \Desc}) \To
+                \EnumT{\V{E}} \To \Desc
 \]
 %
-But this is quite problematic. We are still in the process of
-constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
-begging for this very same $\DescD$. Despite our willingness, we
-cannot materialise such motive. However, we perfectly know what the
-motive is. Consequently, we extend the type theory with a
-special-purpose operator:
+The magician's art rests here, in this extension. We conceal it
+behind a type propagation rule for \(\F{switchD}\) which we apply
+with higher priority than for \(\F{switch}\) in general.
 %
 \[
-\begin{array}{@{}ll}
-%% switchD
-\F{switchD} : & \PITEL{\V{e}}{\EnumU}               
-                \PITEL{\V{b}}{\spi{\V{e}}{\LAM{\_} \Desc}}
-                \PITEL{\V{x}}{\EnumT{\V{e}}} \To \Desc
-\end{array}
+\Rule{\Gamma \Vdash
+  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{(\PLAM{x}{\EnumT{E}}\Desc)}}}
+                           {\M{t'}}}
+     {\Gamma \Vdash
+\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc}}
+ {\switchD{\M{E}}{\M{t'}}}}
 \]
-%
-The entire work of the magician stands here, in this extension. Note
-that the developer do not have to chose between the two variants of
-$\F{switch}$. This is dealt with by the type propagation system
-(Fig.~\ref{fig:type-checking}), extended with the following inference
-rule:
-%
-\[
-\Rule{\Gamma \Vdash \propag{\push{\V{t}}{\spiD{\V{e}}}}
-                           {\V{t'}}}
-     {\Gamma \Vdash \propag{\push{\V{t}}{\EnumT{\V{e}} \To \Desc}}
-                           {\PLAM{\V{x}}{(\EnumT{\V{e}})} \switchD{\V{e}}{\V{t'}}{\V{x}}}
-     }\;\mbox{$\V{t}$ is $[]$ or $[\_,\_]$}
-\]
+As a consequence, our definition above now propagates without
+introducing recursion. Of course, by pasting together
+the declaration of \(\Desc\) and its internal copy, we have made
+it appear in its own type. Hardwaired as a \emph{fait accompli},
+this creates no regress, although one must assume the definition
+to recheck it.
 
@@
 
-This concludes our levitation work. Beyond its pedagogical value, this
+We have levitated \(\Desc\). Beyond its pedagogical value, this
 exercise has several practical outcomes. First of all, it reveals that
@@
 an induction principle. So, our ability to inspect and program with
-$\Desc$ is not restricted to the meta-language: we now have all the
-necessary equipment in the target language to \emph{program} over
-datatypes. In this setting, \emph{generic programming is just
+$\Desc$ is not restricted to a meta-language: we now have all the
+necessary equipment in the theory to \emph{program} over
+datatypes. \emph{Generic programming is just
   programming}.

### 2010-04-02 papers/icfp-2010-desc/paper_desc_levitation.tex
last of 4

@@
 
-In Section~\ref{sec:desc-fix-point}, we have implemented a dependent
-$\F{induction}$ principle, instead of the catamorphism. However, in
-some circumstances, the full-power of a dependent elimination is not
-necessary. In the following, we propose to derive the catamorphism
-from the $\F{induction}$ principle.
+In Section~\ref{sec:desc-fix-point}, we hardwired a dependent
+$\F{ind}$unction principle, instead of the catamorphism. However, in
+some circumstances, the full power of a dependent elimination is not
+necessary. Let us now derive the catamorphism
+from $\F{ind}$ principle.
 
@@
 The catamorphism is defined by induction on the description $\V{D}$,
-with a non-dependent motive targeting $\V{T}$. Given a node $\V{xs}$
+with a readily propagated non-dependent return type $\V{T}$.
+Given a node $\V{xs}$
 and the induction hypotheses, the method ought to build an element of
@@
            \Mu{\V{D}} \To \V{T} \\
-\cata\: \V{D}\: \V{T}\: \V{f\: \V{x}} \mapsto
-  \F{induction}\: \V{D}\:\V{x}\: (\LAM{\_}\V{T})\: (\LAM{\V{xs}\:\V{hs}} \V{f}\: \SHED)
+\cata\: \V{D}\: \V{T}\: \V{f} \mapsto
+  \sind \LAM{\V{xs}}\LAM{\V{hs}} \V{f}\: \SHED
 }\]
@@
 element in $\V{T}$.  Therefore, to construct an element of
-\(\descop{\V{D}}{\V{T}}\), we replace the recursive components of
-\(\V{xs}\) by their counterpart in \(\V{hs}\):
+\(\descop{\V{D}}{\V{T}}\), we must replace the recursive components of
+\(\V{xs}\) by their counterparts from \(\V{hs}\). Let us write a
+program to do that---please forgive us if we lapse to a pattern matching
+notation, for readability's sake.
 %
@@
 basically, a function mapping a $\Desc$ to a datatype specific
-operation. This is a form of polytypic programming, as found in
+operation. This is a form of polytypic programming, as we learned from
 PolyP~\cite{jansson:polyp}.
@@
 \FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
-    \pair{\ListCons{\DVar{}}{\V{E}}}
+    \pair{\pair{\DVar{}}{\V{E}}{}}
          {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
@@
 be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
-$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
+$\V{E}$ and $\V{D}$ as they were, hence leaving the other constructors
 unchanged. Unfolding the interpretation of this definition, we
@@
            \PI{\V{X}, \V{Y}}{\Set} \\
-         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To
-           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
-           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
+         & (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To
+           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
+           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
 \end{array} \\
-\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{x} \mapsto
-  \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
-          (\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})})\: 
-          \SHED\: 
-          \V{x}
+\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
+  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
+          (\Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})})\: 
+          \SHED
 }\]
@@
 \apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
-\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\etag{t}\,}{\V{xs}}{} & \Con{\pair{\etag{t}}{\V{xs}}{}}
+\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
 \end{array}
@@
 datatype~\cite{hinze:generic-haskell}, as found in Generic Haskell:
-from a datatype, we built a new datatype and equip it with its
+from a datatype, we build a new datatype and equip it with its
 structure.

### 2010-04-02 papers/icfp-2010-desc/paper_idesc.tex
EnumU fixes
sec 5 start

@@
 \[\stk{
-\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \:\ldots \\
-\\
+\IDesc{(\Bhab{\V{I}}{\Set})} : \Set \smallskip \\
 \idescop{\_}{}{} : _{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
-\ldots
-}\]
-
-Note that $\V{I} \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
-endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
+}\]
+
+Given these components, we may interpret a \emph{function}
+$\V{R}:\V{I} \To \IDesc{\V{I}}$ is interpreted as a function
 $\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
-\To \Set^{\V{I}}$.The fixpoint of inductive families is defined over
+\To \Set^{\V{I}}$, the type of endofunctors on \(\Set^{\V{I}}\).
+Inductive families are fixpoints defined over
 these indexed functors, hence computing a fixpoint of the entire
@@
                    \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
-                   \PI{\V{P}}{\TIMES{\V{I}}{\IMu{\V{I}}{\V{R}}{\!}} \To \Set} \\
+                   \PI{\V{P}}{(\SIGMAS{\Bhab{\V{i}}{\V{I}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}) \To \Set} \\
                  & (      \PITEL{\V{i}}{\V{I}} 
@@
 \end{array} \\
-\inductionI\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: (\Con{\V{xs}}) =  \\
+\inductionI\: \V{R}\: \V{P}\: \V{m}\: \V{i}\: (\Con{\V{xs}}) =  \\
 \qquad
-    \V{ms}\: \V{i}\: \V{xs}\: (\allI{}
+    \V{m}\: \V{i}\: \V{xs}\: (\allI{}
                                     {\V{R}\: \V{i}}
@@
                                     {\V{P}}
-                                    {(\LAM{\pair{\V{i}}{\V{xs}}{}} \inductionI\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: \V{xs})}
+                                    {(\spl{\LAM{\V{i}}\LAM{\V{xs}} \inductionI\: \V{R}\: \V{P}\: \V{m}})}
                                     {\V{xs}})
@@
 
-\[\stk{
-\begin{array}{ll}
-\stk{
-\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \DVar{\!}         & \V{I} \To \IDesc{\V{I}}                                       \\
-    \DConst{\!}       & \Set \To \IDesc{\V{I}}                                        \\
-    \DProd{\!}{\!}    & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}       \\
-    \DSigma{\!}{\!}   & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
-    \DPi{\!}{\!}      & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
-\end{array}
-}
-\vspace{0.2in}
-\\
-\stk{
+\[\stk{\begin{array}{@{}ll}
+\IDesc{(\Bhab{\M{I}}{\Set})} &: \Set \\
+\DVar{(\Bhab{\M{i}}{\M{I}})} &: \IDesc{\M{I}} \\
+\DConst{(\Bhab{\M{A}}{\Set})} &:\IDesc{\M{I}}       \\
+\DProd{(\Bhab{\M{D}}{\IDesc{\M{I}}})}{(\Bhab{\M{D}}{\IDesc{\M{I}}})}
+  & :\IDesc{\M{I}}       \\
+\DSigma{(\Bhab{\M{S}}{\Set})}{(\Bhab{\M{D}}{\M{S}\To\IDesc{\M{I}}})}
+& : \IDesc{\M{I}}  \\
+\DPi{(\Bhab{\M{S}}{\Set})}{(\Bhab{\M{D}}{\M{S}\To\IDesc{\M{I}}})}
+& : \IDesc{\M{I}}  \\
+\end{array}\smallskip \\
 \idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
-\begin{array}{@{}l@{\V{X}}@{\:=\:\:}ll}
+\begin{array}{@{}l@{\V{X}}@{\:\mapsto\:\:}ll}
 \idescop{\DVar{\V{i}}}{\V{I}}{&}      &  \V{X}\: \V{i}                                           \\
@@
 }
-\end{array}
-}\]
-
-\caption{Universe of indexed description}
+\]
+
+\caption{Universe of indexed descriptions}
 \label{fig:idesc}
@@
 \NatD : \IDesc{\Unit} \\
-\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
-                     {[ (\DConst{\Unit}) \quad (\DVar{\Void}) ]}
+\NatD \mapsto \DSigma{(\EnumT{[ \NatZero\: \NatSuc{\!} ]})}
+                     {\red{[} \DConst{\Unit} \quad \DVar{\Void} \red{]}}
 }\]
@@
 \begin{array}{@{}ll}
-\IDescD\: \V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[} 
-                                          \DVar{\!},
-                                          \DConst{\!},
-                                          \DProd{\!}{\!},
-                                          \DSigma{\!}{\!}, 
+\IDescD\: \V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \sqr{ 
+                                          \DVar{\!}\:
+                                          \DConst{\!}\:
+                                          \DProd{\!}{\!}\:
+                                          \DSigma{\!}{\!}\: 
                                           \DPi{\!}{\!}
-                                             \red{]}) \\
+                                             }) \\
                               & \bigRedBracket{\begin{array}{l}
@@
 
-\newcommand{\VecD}{\C{VecD}}
+\newcommand{\VecD}{\F{VecD}}
 \newcommand{\VecNil}{\etag{\CN{vnil}}}
@@
 \begin{array}{@{}ll}
-\VecD\: \V{X}\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{[ \VecNil , \VecCons{\!}{\!} ]}) \\
+\VecD\: \V{X}\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{\sqr{ \VecNil\:\VecCons{\!}{\!} }}) \\
                         & \bigRedBracket{
@@
 \begin{array}{@{}llll}
-\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
-\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
+\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{\Void})}{\Void} } \\
+\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{\sqr{ \FinZero \: \FinSuc{\!} }}) \\
                      &   &             & \bigRedBracket{
@@
 \[
-\Ty \mapsto \EnumT{[\Enat, \Ebool]}
+\Ty \mapsto \EnumT{\sqr{\Enat\: \Ebool}}
 \]
@@
                  \begin{array}{l}
-                   {[\Eval{\!}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
+                   {\sqr{\Eval{\!}\: \Econd{\!}{\!}{\!}\!\,}} \red{,} \\
                    \LAM{\V{ty}}
@@
                  \begin{array}{l}
-                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To {[\Eplus{\!}{\!}\!\,]} \\ \Ebool \To {[\Ele{\!}{\!}\!\,]}} \red{,} \\
+                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To {\sqr{\Eplus{\!}{\!}\!\,}} \\ \Ebool \To {\sqr{\Ele{\!}{\!}\!\,}}} \red{,} \\
                    \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
@@
                  \begin{array}{l}
-                   \EnumT{\red{[}\Econd{\!}{\!}{\!}\red{]}} \red{,} \\
+                   \EnumT{\sqr{\Econd{\!}{\!}{\!}}} \red{,} \\
                    \LAM{\V{ty}}
@@
 \IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
-                                [\DConst{\!},
-                                 \DProd{\!}{\!},
-                                 \DSigma{\!}{\!}, 
-                                 \DPi{\!}{\!}] \red{,}\\
+                                \sqr{\DConst{\!}\:
+                                 \DProd{\!}{\!}\:
+                                 \DSigma{\!}{\!}\: 
+                                 \DPi{\!}{\!}} \red{,}\\
                                   \LAM{\_}\bigRedBracket{\begin{array}{l}
@@
 \IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
-\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}[]\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
+\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}\sqr{}\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
 }\]

### 2010-04-02 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: replace \TIMES by the proper \SIGMA
sec 5 adjustments

@@
                    \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
-                   \PI{\V{P}}{(\SIGMAS{\Bhab{\V{i}}{\V{I}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}) \To \Set} \\
+                   \PI{\V{P}}{(\SIGMA{\V{i}}{\V{I}}{\IMu{\V{I}}{\V{R}}{\V{i}}}) \To \Set} \\
                  & (      \PITEL{\V{i}}{\V{I}} 
@@
                         & \idescop{\V{D}}{\V{I}}{\V{X}} \To
-                          \IDesc{(\TIMES{\V{I}}{\V{X}})}
+                          \IDesc{(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}})}
 \end{array} \\
@@
                                   \PITEL{\V{X}}{\V{I} \To \Set} 
-                                  \PI{\V{P}}{\TIMES{\V{I}}{\V{X}} \To \Set} \\
-                                & (\PI{\V{x}}{\TIMES{\V{I}}{\V{X}}} \V{P}\: \V{x}) \To
+                                  \PI{\V{P}}{(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}) \To \Set} \\
+                                & (\PI{\V{x}}{\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}} \V{P}\: \V{x}) \To
                                   \PI{\V{xs}}{\idescop{\V{D}}{\V{I}}{\V{X}}} 
-                                  \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{\TIMES{\V{I}}{\V{X}}}{\V{P}}
+                                  \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{}{\V{P}}
 \end{array} \\
@@
 levitated. As before, we rely on a special purpose $\F{switchID}$
-operator to build the finite function space $\bigRedBracket{\ldots}$.
+operator to build the finite function $\bigRedBracket{\ldots}$
+without mentioning \(\IDesc{\!}\).
 
@@
 
-So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
+So far, the examples we have seen live in $\IDesc{\Unit}$, hence are
 not using any indexing. We remedy this by encoding the vectors. Recall
@@
 one''. This translates into the constraint $\V{n} \PropEq
-\NatSuc{\V{m}}$.
-
-However, one could, quite rightfully, be worried by the introduction
-of equalities in the picture. In a programming context, equalities
-haunt us, begging for proofs. As much as possible, one would like to
-avoid suffering from this burden. However, the equalities introduced
-here are rather special: they are \emph{equations} constraining
-variables to certain terms. \citet{goguen:pattern-matching} have
-shown, in the context of dependent pattern matching, how to tame these
-equations.
+\NatSuc{\V{m}}$, given a suitable presentation of propositional equality.
+
+We have been careful to keep our setup agnostic with respect to
+notions of propositional equality. Any will do, according to your
+convictions, or for vectors, none---equality for \(\Nat\) is definable by
+recursion---and many variations are popular. The traditional
+homogeneous identity type used in Coq is not adequate to support
+dependent pattern matching, but its heterogeneous variant, allowing
+equations between elements of arbitrary types, is sufficient to
+allow the translation of structurally recursive pattern matching
+programs to \(\F{indI}\)~\cite{goguen:pattern-matching}. Our present
+inclination is towards the extensional equality proposed by
+\citet{altenkirch:ott}, which also sustains the translation.
 
@@
 
-Further, we can actually remove these equations altogether. Let us
+However, sometimes, we can actually remove these equations altogether. Let us
 look back at $\D{Vec}$. We note that the equations are introduced
@@
 \begin{array}{@{}lll}
-\VecD\:\V{X}\: \NatZero     & = & \DConst{\Unit} \\
-\VecD\:\V{X}\: (\NatSuc{\V{n}}) & = & \DProd{\DConst{\V{X}}}{\DVar{\V{n}}}
+\VecD\:\V{X}\: \NatZero     & \mapsto & \DConst{\Unit} \\
+\VecD\:\V{X}\: (\NatSuc{\V{n}}) & \mapsto & \DProd{\DConst{\V{X}}}{\DVar{\V{n}}}
 \end{array}
@@
 \begin{array}{@{}llll}
-\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{\Void})}{\Void} } \\
-\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{\sqr{ \FinZero \: \FinSuc{\!} }}) \\
+\FinD\: \NatZero     & \mapsto & \multicolumn{2}{l}{\DSigma{(\EnumT{\Void})}{\Void} } \\
+\FinD\: (\NatSuc{\V{n}}) & \mapsto & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{\sqr{ \FinZero \: \FinSuc{\!} }}) \\
                      &   &             & \bigRedBracket{
@@
 constructor $\ListCons{\!}{\!}$.
+
+We use the notation $\toIDesc{\V{TID}}$ to denote the
+indexed description computed from the tagged indexed description
+$\V{TID}$. Its expansion is similar to the definition of \(\toDesc{\!}\)
+but more involved.
 
@@
 %
-Where we overload the notation $\toIDesc{\V{TID}}$ to denote the
-indexed description computed from the tagged indexed description
-$\V{TID}$. The type of $\evalH$ is strikingly similar to a
+The type of $\evalH$ is strikingly similar to a
 catamorphism. Indeed, implementing a single step of evaluation -- the
@@
 \F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
-\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\etag{t}\,}{\V{ys}}{} & \Con{\pair{\etag{t}}{\V{ys}}{}}
+\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
 \end{array}

### 2010-04-02 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Expr using []

@@
 that some constructors are always defined, namely $\Econd{\!}{\!}{\!}$
-and $\Eval{\!}$. On the other hand, the $\Eplus{\!}{\!}$ and
-$\Ele{\!}{\!}$ constructors are index-dependent. $\Eplus{\!}{\!}$ is
+and $\Eval{\!}$. On the other hand, the $\Eplus{\!}{}$ and
+$\Ele{}{\!}$ constructors are index-dependent. $\Eplus{\!}{}$ is
 defined if and only if the result type -- the index -- is $\Enat$,
-whereas $\Ele{\!}{\!}$ is defined if and only if the index is
+whereas $\Ele{}{\!}$ is defined if and only if the index is
 $\Ebool$. The actual code precisely follows this intuition, as shown
-in Figure~\ref{fig:hexpr-full}. For brevity, we use an informal
-$\case{\ldots}{\ldots}$ notation, simulating a definition by
-pattern-matching. Formally, this corresponds to a call to the
-$\switch{\!}{\!}{\!}{\!}$ eliminator.
+in Figure~\ref{fig:hexpr-full}.
 
@@
                  \begin{array}{l}
-                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To {\sqr{\Eplus{\!}{\!}\!\,}} \\ \Ebool \To {\sqr{\Ele{\!}{\!}\!\,}}} \red{,} \\
+                   \sqr{\sqr{\Eplus{\!}{\!}} \quad \sqr{\Ele{\!}{\!}}} \red{,} \\
                    \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 

### 2010-04-02 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: suppress line break before equations: we are at 11.5 pages
Desc@ICFP: uniformize titles.
Desc@ICFP: uniformize coding style
strat chat

@@
 is orthogonal to our work. Nonetheless, our universe of description
-stratifies naturally. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of
-type $\Set^{\blue{l+1}}$. Similarly, the interpretation of
+stratifies naturally. $\IDesc{\!}$ is self-encoding only in a
+level-polymorphic sense.  Unsurprisingly, $\IDesc{\!}$ at level $l$ is
+of type $\Set^{\blue{l+1}}$. Similarly, the interpretation of
 $\IDesc{\!}$ at level $l$ is an object of type $\Set^{\blue{l}}$:
+%
+\[\stk{\begin{array}{@{}ll}
+\IDesc{\!}^{\blue{l}} (\Bhab{\M{I}}{\Set^{\blue{l+1}}}) & :\Set^{\blue{l+1}} \\
+\DVar{(\Bhab{\M{i}}{\M{I}})} &: \IDesc{\!}^{\blue{l}}{\M{I}} \\
+\DConst{(\Bhab{\M{A}}{\Set^{\blue{l}}})} &:\IDesc{\!}^{\blue{l}}{\M{I}}       \\
+\DProd{(\Bhab{\M{D}}{\IDesc{\!}^{\blue{l}}{\M{I}}})}{(\Bhab{\M{D}}{\IDesc{\!}^{\blue{l}}{\M{I}}})}
+  & :\IDesc{\!}^{\blue{l}}{\M{I}}       \\
+\DSigma{(\Bhab{\M{S}}{\Set^{\blue{l}}})}{(\Bhab{\M{D}}{\M{S}\To\IDesc{\!}^{\blue{l}}{\M{I}}})}
+& : \IDesc{\!}^{\blue{l}}{\M{I}}  \\
+\DPi{(\Bhab{\M{S}}{\Set^{\blue{l}}})}{(\Bhab{\M{D}}{\M{S}\To\IDesc{\!}^{\blue{l}}{\M{I}}})}
+& : \IDesc{\!}^{\blue{l}}{\M{I}}  \\
+\end{array}\smallskip \\
+\idescop{\_\:}{}{}^{\blue{l}} : _{\PI{\V{I}}{\Set^{\blue{l+1}}}} \IDesc{{\!}^{\blue{l}}\V{I}} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
+\ldots
+}
+\]
+Crucially, the types of data stored in an
+\(\IDesc{\!}^{\blue{l}}{\M{I}}\) all live no higher---we may store an
+\(I\) and a \(\Set^{\blue{l}}\) in a \(\Set^{\blue{l+1}}\).  The code
+for \(\IDesc{\!}^{\blue{l}}{\M{I}}\) is an element of
+\(\IDesc{\!}^{\blue{l+1}}{\Unit}\), so there is a spiral, not a cycle.
+We have checked the construction using Agda's universe polymorphism,
+coding $\IDesc{\!}$ in itself and have proving the isomorphism
+between the host and the embedded universes.
 
-\[\stk{
-\data \IDesc{\!}^{\blue{l}} (\Bhab{\V{I}}{\Set^{\blue{l+1}}}) : \Set^{\blue{l+1}} \where \ldots \\
-\\
-\idescop{\_\:}{}{}^{\blue{l}} : \PI{\V{I}}{\Set^{\blue{l+1}}} \IDesc{{\!}^{\blue{l}}\V{I}} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
-\ldots
-}\]
-
-The operations and examples developed in this paper stratify just as
-well. We refer the reader to our Agda model. Further, we have coded
-$\IDesc{\!}$ in itself and have proved the isomorphism between the
-host and the embedded universes.
-
-\subsection{Related Work}
+\subsection{Related work}
 

### 2010-04-02 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix alignment of \subst
Desc@ICFP: fix alignment of \replace
Desc@ICFP: ispell

@@
 supply that return type, especially when we know what it must be.
-We eliminate the recursion by \emph{specializing} \(\F{switch}\):
+We eliminate the recursion by \emph{specialising} \(\F{switch}\):
 %
@@
 the declaration of \(\Desc\) and its internal copy, we have made
-it appear in its own type. Hardwaired as a \emph{fait accompli},
+it appear in its own type. Hardwired as a \emph{fait accompli},
 this creates no regress, although one must assume the definition
@@
 \subst : & \PITEL{\V{D}}{\TagDesc}
-           \PI{\V{X}, \V{Y}}{\Set} \\
-         & (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To
-           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
+           \PI{\V{X}, \V{Y}}{\Set} 
+           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
+         & \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
            \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
@@
 \apply : & \PITEL{\V{D}}{\TagDesc} 
-           \PI{\V{X}, \V{Y}}{\Set} \\
-         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To
-           \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
+           \PI{\V{X}, \V{Y}}{\Set} 
+           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To \\
+         & \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
            \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}

### 2010-04-02 papers/icfp-2010-desc/paper_type_theory.tex
added split, tweaked the examples
Desc@ICFP: restore type-checking as a floating figure
Desc@ICFP: type-synthesis floating
Desc@ICFP: ispell

@@
 forms---admit type \emph{synthesis}. We embed $\exprEx$ into
-$\exprIn$, demanding that the synthesized type coincides with the type
+$\exprIn$, demanding that the synthesised type coincides with the type
 proposed. The other direction---only necessary to apply
@@
 
-Type synthesis is the \emph{source} of types. It follows the \(\exprEx\)
-syntax, delivering both the elaborated term and its type. Terms
-and expressions never mix: e.g., for application, we
-instantiate the range with the \emph{term} delivered by checking the
-argument \emph{expression}. Hardwired operators are checked as variables.
+Type synthesis (Fig.~\ref{fig:type-synthesis}) is the \emph{source} of
+types. It follows the \(\exprEx\) syntax, delivering both the
+elaborated term and its type. Terms and expressions never mix: e.g.,
+for application, we instantiate the range with the \emph{term}
+delivered by checking the argument \emph{expression}. Hardwired
+operators are checked as variables.
+
+\begin{figure}
 \input{figure_type_synthesis}
-
-%\begin{figure}
-%\input{figure_type_synthesis}
-%\caption{Type synthesis}
-%\label{fig:type-synthesis}
-%\end{figure}
+\caption{Type synthesis}
+\label{fig:type-synthesis}
+\end{figure}
 
@@
 
-Dually, type checking judgments
-%(Fig.~\ref{fig:type-checking})
+Dually, type checking judgments (Fig.~\ref{fig:type-checking})
 are \emph{sinks} for types. From an
@@
 
-%\begin{figure}
+\begin{figure}
 \input{figure_type_checking}
-%\caption{Type checking}
-%\label{fig:type-checking}
-%\end{figure}
+\caption{Type checking}
+\label{fig:type-checking}
+\end{figure}
 

### 2010-04-02 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: ispell
bonkers substitution remark

@@
 $\DVar{\Void}$, where we were using $\DIndx{\!}$ in the previous
-presentation. This transformation generalizes to all inductive
+presentation. This transformation generalises to all inductive
 types. Moreover, we gain the ability to write mutually recursive
@@
 
+The fact that indexed descriptions are closed under substitution
+is potentially of considerable utility, if we can exploit this fact:
+\[
+\idescop{\sigma D}{J}{X} = \idescop{D}{I}{\LAM{i}\idescop{\sigma i}{J}{X}}
+\quad \mbox{where}\;\sigma:I\To\IDesc{J}
+\]
+By observing that a description can be decomposed via substitution, we
+split its meaning into a superstructure of substructures, e.g. a
+`database containing salaries', ready for traversal operations
+preserving the former and targeting the latter.
+
 In this section, we have presented the universe of indexed
@@
 several example of indexed datatypes. In this context, we have
-presented the free monad construction, together its monadic
+presented the free monad construction, together with its monadic
 operations.

### 2010-04-03 papers/icfp-2010-desc/paper_desc.tex
EnumU fixes
Desc@ICFP: definitely fix spacing of operators.

@@
 datatype declarations. We can read a description $\V{D}:\Desc$ as a
-`pattern functor' on $\Set$, with $\descop{\V{D}}{\!}$ its action on
+`pattern functor' on $\Set$, with $\SYMBdescop{\V{D}}$ its action on
 an object, \(\V{X}\), soon to be instantiated recursively.
@@
 the empty tuple. To build
-sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
+sigmas-of-sigmas, we define a $\SYMBDSigma$ code, interpreted as a
 $\Sigma$-type. To request a recursive component, we have $\DIndx{\V{D}}$,
@@
 \NatD : \Desc \\
-\NatD \mapsto \DSigma{\EnumT{\sqr{\NatZero\: \NatSuc{\!} }}}
+\NatD \mapsto \DSigma{\EnumT{\sqr{\NatZero\: \SYMBNatSuc }}}
                      {\sqr{ \DUnit \quad (\DIndx{\DUnit}) }}
@@
 %
-Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
-give a choice between the $\NatZero$ and $\NatSuc{}$ constructors.
+Let us explain its construction. First, we use $\SYMBDSigma$ to
+give a choice between the $\NatZero$ and $\SYMBNatSuc$ constructors.
 What follows depends on this choice, so we write the function
@@
 $\NatZero$ case, we reach the end of the description. In the
-$\NatSuc{\!}$ case, we attach one recursive argument and close the
+$\SYMBNatSuc$ case, we attach one recursive argument and close the
 description. Translating the \(\Sigma\) to a binary sum, we have
@@
 \sqr{\NatZero}:\descop{\NatD}Z \qquad
-\sqr{\NatSuc{(\Bhab{z}{X})}}:\descop{\NatD}Z
+\sqr{\NatSuc{(\Bhab{z}{Z})}}:\descop{\NatD}Z
 \]
@@
 \ListD \: \V{X} \mapsto
- \DSigma{\EnumT{\sqr{ \ListNil\: \ListCons{\!}{\!} }}}
+ \DSigma{\EnumT{\sqr{ \ListNil\: \SYMBListCons }}}
          {\sqr{ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) }}
@@
 %
-The $\NatSuc{\!}$ constructor is turned into a proper
-$\ListCons{\!}{\!}$, taking an argument in $\V{X}$ followed by a
+The $\SYMBNatSuc$ constructor is turned into a proper
+$\SYMBListCons$, taking an argument in $\V{X}$ followed by a
 recursive argument. This code describes the following functor:
@@
 \begin{array}{@{}ll}
-\TreeD \: \V{X} \mapsto \DSigma{ & \EnumT{\sqr{ \TreeLeaf, \TreeNode }} \\}
+\TreeD \: \V{X} \mapsto \DSigma{ & \EnumT{\sqr{ \TreeLeaf\: \TreeNode }} \\}
   { & \sqr{ \DUnit \quad
@@
 From the examples above, we observe that datatypes are defined by a
-$\DSigma{\!}{\!}$ whose first argument enumerates the constructors. We
+$\SYMBDSigma$ whose first argument enumerates the constructors. We
 call codes fitting this pattern \emph{tagged} descriptions. Again,
@@
 \smallskip\\
-\toDesc{\!} : \TagDesc\To\Desc \\
-\toDesc{\!} \mapsto
+\SYMBtoDesc : \TagDesc\To\Desc \\
+\SYMBtoDesc \mapsto
 \spl{\LAM{\V{E}}\LAM{\V{D}}
@@
 \[\stk{
-\Nat\mapsto\Mu{(\toDesc{\pair{\sqr{\NatZero\: \NatSuc{\!} }}
+\Nat\mapsto\Mu{(\toDesc{\pair{\sqr{\NatZero\: \SYMBNatSuc }}
                        {\sqr{ \DUnit \quad (\DIndx{\DUnit})}}{}})}:\Set
@@
 
-We note that the very same functors \(\descop{\M{D}}{\!}\) also admit greatest
+We note that the very same functors \(\SYMBdescop{\M{D}}\) also admit greatest
 fixpoints, and we have indeed implemented coinductive types this way, but
@@
 \begin{array}{@{}ll}
-\All{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
+\SYMBAll : & \PITEL{\V{D}}{\Desc}
              \PITEL{\V{X}}{\Set}
@@
 \begin{array}{@{}ll}
-\all{\!}{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
+\SYMBall : & \PITEL{\V{D}}{\Desc}
              \PITEL{\V{X}}{\Set}

### 2010-04-03 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: spacing of operator
Desc@ICFP: definitely fix spacing of operators.

@@
 %
-The $\NilE$ and $\ConsE{\!}{\!}$ constructors are just the `nil' and
+The $\NilE$ and $\SYMBConsE$ constructors are just the `nil' and
 `cons' or ordinary lists, with elements from \(\UId\).
@@
 theory does not need to be extended with a special type former $\EnumU$,
-or special constructors $\NilE$ and $\ConsE$. 
+or special constructors $\NilE$ and $\SYMBConsE$. 
 Moreover, the $\spi{\M{E}}{\M{P}}$ operator, computing tuple types
@@
 Note, however, that the universe decoder \(\EnumT{\M{E}}\) \emph{is}
-hardwired, as are the primitive \(\Ze\) and \(\Su{\!}\) that
-we use for low-level values, and indeed the \(\F{switch}\) operator.
+hardwired, as are the primitive \(\Ze\) and \(\Su\) that
+we use for low-level values, and indeed the \(\SYMBswitch\) operator.
 We cannot dispose of data altogether! We have, however, gained
@@
 \DescD \mapsto
-  \tada{\stkm{\DUnit\\ \DSigma{\!}{\!}\\ \DIndx{\!}}}
+  \tada{\stkm{\DUnit\\ \SYMBDSigma\\ \SYMBDIndx}}
        {\stkm{ \DUnit                                \\
@@
 we have done so far, we first offer a constructor choice from
-$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
+$\DUnit$, $\SYMBDSigma$, and $\SYMBDIndx$. The reader will notice
 that the `tagged' notation we have used for the $\Desc$ constructors now
@@
 For $\DUnit$, we immediately reach the end of the description. For
-$\DIndx{\!}$, there is a single recursive argument. Describing
-$\DSigma{\!}{\!}$ is problematic. Recall the specification of
-$\DSigma{\!}{\!}$:
+$\SYMBDIndx$, there is a single recursive argument. Describing
+$\SYMBDSigma$ is problematic. Recall the specification of
+$\SYMBDSigma$:
 %
@@
 In order to capture a notion of higher-order induction, we add a code
-$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. Intuitively,
-$\DHindx{\!}{\!}$ gives a recursive subobject for each element of $H$.
+$\SYMBDHindx$ that takes an indexing set $\V{H}$. Intuitively,
+$\SYMBDHindx$ gives a recursive subobject for each element of $H$.
 %
@@
 
-Note that up to isomorphism, $\DIndx{\!}$ is subsumed by
+Note that up to isomorphism, $\SYMBDIndx$ is subsumed by
 $\DHindx{\Unit}{\!}$. However, the apparent duplication has some
-value.  Unlike its counterpart, $\DIndx{\!}$ is first-order: we prefer
+value.  Unlike its counterpart, $\SYMBDIndx$ is first-order: we prefer
 not to demand dummy functions from \(\Unit\) in ordinary data,
@@
 \DescD \mapsto\tada
-{\stkm{ \DUnit\\ \DSigma{\!}{\!}\\ \DIndx{\!}\\ \DHindx{\!}{\!}}}
+{\stkm{ \DUnit\\ \SYMBDSigma\\ \SYMBDIndx\\ \SYMBDHindx}}
 {\stkm{
@@
 %
-The $\DUnit$ and $\DIndx{\!}$ cases remain unchanged, as expected. We
-successfully describe the $\DSigma{\!}{}$ case, by a simple appeal to
-the higher-order induction on $\V{S}$. The $\DHindx{\!}{\!}$ case
+The $\DUnit$ and $\SYMBDIndx$ cases remain unchanged, as expected. We
+successfully describe the $\SYMBDSigma$ case, by a simple appeal to
+the higher-order induction on $\V{S}$. The $\SYMBDHindx$ case
 consists in packing a $\Set$ with a recursive argument.
@@
 \Desc\mapsto
-\Mu{\stk{(\DSigma{\EnumT{\sqr{ \DUnit\: \DSigma{\!}{\!}\: \DIndx{\!}\: 
-    \DHindx{\!}{\!}}}\\}
-    {\;\F{switch}\:\sqr{ \DUnit\: \DSigma{\!}{\!}\: \DIndx{\!}\: 
-    \DHindx{\!}{\!}}\:(\LAM{\_}\Desc)\\
+\Mu{\stk{(\DSigma{\EnumT{\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}}\\}
+    {\;\SYMBswitch\:\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}\:(\LAM{\_}\Desc)\\
 \;\sqr{\stkm{
@@
 %
-We simply add a constructor, $\DVar{\!}$, and define its argument to
+We simply add a constructor, $\SYMBDVar$, and define its argument to
 be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
@@
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
+\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
 \apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}

### 2010-04-03 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: definitely fix spacing of operators.

--- 2010-04-01 papers/icfp-2010-desc/paper_discussion.tex
@@
 is orthogonal to our work. Nonetheless, our universe of description
-stratifies naturally. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of
-type $\Set^{\blue{l+1}}$. Similarly, the interpretation of
-$\IDesc{\!}$ at level $l$ is an object of type $\Set^{\blue{l}}$:
+stratifies naturally. $\SYMBIDesc$ is self-encoding only in a
+level-polymorphic sense.  Unsurprisingly, $\SYMBIDesc$ at level $l$ is
+of type $\Set^{\blue{l+1}}$. Similarly, the interpretation of
+$\SYMBIDesc$ at level $l$ is an object of type $\Set^{\blue{l}}$:
+%
+\[\stk{\begin{array}{@{}ll}
+\SYMBIDesc^{\blue{l}} (\Bhab{\M{I}}{\Set^{\blue{l+1}}}) & :\Set^{\blue{l+1}} \\
+\DVar{(\Bhab{\M{i}}{\M{I}})} &: \SYMBIDesc^{\blue{l}}{\M{I}} \\
+\DConst{(\Bhab{\M{A}}{\Set^{\blue{l}}})} &:\SYMBIDesc^{\blue{l}}{\M{I}}       \\
+\DProd{(\Bhab{\M{D}}{\SYMBIDesc^{\blue{l}}{\M{I}}})}{(\Bhab{\M{D}}{\SYMBIDesc^{\blue{l}}{\M{I}}})}
+  & :\SYMBIDesc^{\blue{l}}{\M{I}}       \\
+\DSigma{(\Bhab{\M{S}}{\Set^{\blue{l}}})}{(\Bhab{\M{D}}{\M{S}\To\SYMBIDesc^{\blue{l}}{\M{I}}})}
+& : \SYMBIDesc^{\blue{l}}{\M{I}}  \\
+\DPi{(\Bhab{\M{S}}{\Set^{\blue{l}}})}{(\Bhab{\M{D}}{\M{S}\To\SYMBIDesc^{\blue{l}}{\M{I}}})}
+& : \SYMBIDesc^{\blue{l}}{\M{I}}  \\
+\end{array}\smallskip \\
+\idescop{\_\:}{}{}^{\blue{l}} : _{\PI{\V{I}}{\Set^{\blue{l+1}}}} \SYMBIDesc^{\blue{l}}\:\V{I} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
+\ldots
+}
+\]
+Crucially, the types of data stored in an
+\(\SYMBIDesc^{\blue{l}}{\M{I}}\) all live no higher---we may store an
+\(I\) and a \(\Set^{\blue{l}}\) in a \(\Set^{\blue{l+1}}\).  The code
+for \(\SYMBIDesc^{\blue{l}}{\M{I}}\) is an element of
+\(\SYMBIDesc^{\blue{l+1}}{\Unit}\), so there is a spiral, not a cycle.
+We have checked the construction using Agda's universe polymorphism,
+coding $\SYMBIDesc$ in itself and have proving the isomorphism
+between the host and the embedded universes.
 
-\[\stk{
-\data \IDesc{\!}^{\blue{l}} (\Bhab{\V{I}}{\Set^{\blue{l+1}}}) : \Set^{\blue{l+1}} \where \ldots \\
-\\
-\idescop{\_\:}{}{}^{\blue{l}} : \PI{\V{I}}{\Set^{\blue{l+1}}} \IDesc{{\!}^{\blue{l}}\V{I}} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
-\ldots
-}\]
-
-The operations and examples developed in this paper stratify just as
-well. We refer the reader to our Agda model. Further, we have coded
-$\IDesc{\!}$ in itself and have proved the isomorphism between the
-host and the embedded universes.
-
-\subsection{Related Work}
+\subsection{Related work}
 
@@
 formers and elimination rules. Thanks to levitation, we only rely on a
-generic induction and a specialised $\switchD{\!}{\!}{\!}$. Second,
+generic induction and a specialised $\SYMBswitchD$. Second,
 the authors do not tackle the issue of \emph{programming} with

### 2010-04-03 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: definitely fix spacing of operators.

@@
   transformer anyone?} $\V{I} \To \IDesc{\V{I}}$ is our grammar for
-describing these functors. Hence, $\IDesc{\!}$ and its interpretation
+describing these functors. Hence, $\SYMBIDesc$ and its interpretation
 have the following types:
@@
       \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}}}
-     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\!}}{\V{I}\To\Set}} \qquad
+     {\Gamma \vdash \Bhab{\SYMBIMu_{\V{I}}{\V{R}}}{\V{I}\To\Set}} \qquad
 \\
@@
           \Gamma \vdash \Bhab{\V{i}}{\V{I}} &
-          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\:\V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{\!})}}
+          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\:\V{i}}{\V{I}}{(\SYMBIMu_{\V{I}}{\V{R}})}}
       \end{array}}
@@
 evolving $\Desc$ to cope with indexing. \note{Say something about the
-  transformation} The code of $\IDesc{\!}$ is presented in
+  transformation} The code of $\SYMBIDesc$ is presented in
 Figure~\ref{fig:idesc}. Induction on indexed descriptions is defined
@@
                  & (      \PITEL{\V{i}}{\V{I}} 
-                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{\!})}} \\
+                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\SYMBIMu_{\V{I}}{\V{R}})}} \\
                  & \   \idescop{\AllI{}
                                      {(\V{R}\: \V{i})}
-                                     {(\IMu{\V{I}}{\V{R}}{\!})}
+                                     {(\SYMBIMu_{\V{I}}{\V{R}})}
                                      {\V{xs}}}
@@
 %
-Where the operators $\AllI{}{\!}{\!}{}$ and
-$\allI{}{\!}{\!}{\!}{\!}{}$ are presented in
-Figure~\ref{fig:allI-predicates}. As for descriptions, we can compute
-a generic catamorphism, $\cataI$, from $\inductionI$.
+Where the operators $\SYMBAllI$ and $\SYMBallI$ are
+presented in Figure~\ref{fig:allI-predicates}. As for descriptions, we
+can compute a generic catamorphism, $\cataI$, from $\inductionI$.
 
@@
 \begin{array}{@{}ll}
-\AllI{\!}{\!}{\!}{\!} : & _{\PI{\V{I}}{\Set}}
+\SYMBAllI : & _{\PI{\V{I}}{\Set}}
               \PITEL{\V{D}}{\IDesc{\V{I}}}
@@
 \begin{array}{@{}ll}
-\allI{\!}{\!}{\!}{\!}{\!}{\!} : & _{\PI{\V{I}}{\Set}}
+\SYMBallI : & _{\PI{\V{I}}{\Set}}
               \PITEL{\V{D}}{\IDesc{\V{I}}}
@@
 
-In order to gain some intuition of $\IDesc{\!}$, let us re-implement
+In order to gain some intuition of $\SYMBIDesc$, let us re-implement
 the pattern functor of natural numbers:
@@
 \NatD : \IDesc{\Unit} \\
-\NatD \mapsto \DSigma{(\EnumT{[ \NatZero\: \NatSuc{\!} ]})}
+\NatD \mapsto \DSigma{(\EnumT{[ \NatZero\: \SYMBNatSuc ]})}
                      {\red{[} \DConst{\Unit} \quad \DVar{\Void} \red{]}}
@@
 functor. Therefore, the recursive argument is materialised by
-$\DVar{\Void}$, where we were using $\DIndx{\!}$ in the previous
+$\DVar{\Void}$, where we were using $\SYMBDIndx$ in the previous
 presentation. This transformation generalises to all inductive
@@
 \begin{array}{@{}ll}
-\IDescD\: \V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \sqr{ 
-                                          \DVar{\!}\:
-                                          \DConst{\!}\:
-                                          \DProd{\!}{\!}\:
-                                          \DSigma{\!}{\!}\: 
-                                          \DPi{\!}{\!}
+\IDescD\: \V{I} \mapsto \SYMBDSigma & (\EnumT \sqr{ 
+                                          \SYMBDVar\:\:
+                                          \SYMBDConst\:\:
+                                          \DProd\:\:
+                                          \SYMBDSigma\:\:
+                                          \SYMBDPi
                                              }) \\
@@
                                       \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
-                                   \end{array}
-                             }
+                                   \end{array}}
 \end{array}
@@
 operator to build the finite function $\bigRedBracket{\ldots}$
-without mentioning \(\IDesc{\!}\).
+without mentioning \(\SYMBIDesc\).
 
@@
 \newcommand{\VecNil}{\etag{\CN{vnil}}}
-\newcommand{\VecCons}[2]{\etag{\CN{vcons}}\:#1\:#2}
+\newcommand{\SYMBVecCons}{\etag{\CN{vcons}}\xspace}
+\newcommand{\VecCons}[2]{\SYMBVecCons\:#1\:#2}
 
@@
 not using any indexing. We remedy this by encoding the vectors. Recall
-that the constructors $\VecNil$ and $\VecCons{\!}{\!}$ are only defined for an
+that the constructors $\VecNil$ and $\SYMBVecCons$ are only defined for an
 index $\NatZero$ and $\NatSuc$ respectively.:
@@
     \VecNil          & \D{Vec}\:\V{X}\:{\NatZero)}   \\
-    \VecCons{\!}{\!} & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
+    \SYMBVecCons & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
 \end{array}
@@
 \VecD : \Set \To \Nat \To \IDesc{\Nat} \\
-\begin{array}{@{}ll}
-\VecD\: \V{X}\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{\sqr{ \VecNil\:\VecCons{\!}{\!} }}) \\
+\begin{array}{@{}l@{}l}
+\VecD\: \V{X}\: \V{n} = \SYMBDSigma & (\EnumT{\sqr{ \VecNil\:\SYMBVecCons }}) \\
                         & \bigRedBracket{
@@
 In the $\VecNil$ case, a proof must be provided that the index is
-equal to $\NatZero$. In the $\VecCons{\!}{\!}$ case, we first store an element
+equal to $\NatZero$. In the $\SYMBVecCons$ case, we first store an element
 $\V{m}$ of $\Nat$. However, the constraint stipulates that $\V{m}$
@@
 The equations (and constructors) have simply disappeared. A similar
-example is $\Fin{\!}$, specified by:
+example is $\SYMBFin$, specified by:
 %
@@
 \stk{
-\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
+\data \SYMBFin : \PI{\V{n}}{\Nat} \Set \where \\
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
     \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
-    \FinSuc{\!}   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
+    \SYMBFinSuc   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
 \end{array}
@@
 In this case, we can apply forcing, but not detagging, since both
-$\FinZero$ and $\FinSuc{\!}$ both target $\NatSuc{\!}$:
+$\FinZero$ and $\SYMBFinSuc$ both target $\SYMBNatSuc$:
 %
@@
 \FinD : \Nat \To \IDesc{\Nat} \\
-\begin{array}{@{}llll}
+\begin{array}{@{}lll@{}l}
 \FinD\: \NatZero         & \mapsto & \multicolumn{2}{l}{\DSigma{(\EnumT{\Void})}{\Void} } \\
-\FinD\: (\NatSuc{\V{n}}) & \mapsto & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{\sqr{ \FinZero \: \FinSuc{\!} }}) \\
+\FinD\: (\NatSuc{\V{n}}) & \mapsto & \SYMBDSigma & (\EnumT{\sqr{ \FinZero \: \SYMBFinSuc }}) \\
                          &         &             & \bigRedBracket{
@@
                                                        \DVar{\V{n}}
-                                         \end{array}
-                                         }
+                                                   \end{array}}
 \end{array}
@@
 $\ListNil$. Similarly, for $\NatSuc{n}$, we would only propose the
-constructor $\ListCons{\!}{\!}$.
+constructor $\SYMBListCons$.
 
@@
 indexed description computed from the tagged indexed description
-$\V{TID}$. Its expansion is similar to the definition of \(\toDesc{\!}\)
+$\V{TID}$. Its expansion is similar to the definition of \(\SYMBtoDesc\)
 but more involved.
@@
 %% Constructors
-\newcommand{\Eval}[1]{\etag{\CN{val}}\:#1}
-\newcommand{\Econd}[3]{\etag{\CN{cond}}\:#1\:#2\:#3}
-\newcommand{\Eplus}[2]{\etag{\CN{plus}}\:#1\:#2}
-\newcommand{\Ele}[2]{\etag{\CN{le}}\:#1\:#2}
+\newcommand{\SYMBEval}{\etag{\CN{val}}\xspace}
+\newcommand{\Eval}[1]{\SYMBEval\:#1}
+\newcommand{\SYMBEcond}{\etag{\CN{cond}}\xspace}
+\newcommand{\Econd}[3]{\SYMBEcond\:#1\:#2\:#3}
+\newcommand{\SYMBEplus}{\etag{\CN{plus}}\xspace}
+\newcommand{\Eplus}[2]{\SYMBEplus\:#1\:#2}
+\newcommand{\SYMBEle}{\etag{\CN{le}}\xspace}
+\newcommand{\Ele}[2]{\SYMBEle\:#1\:#2}
 
 %% Index mapper (terminology?)
-\newcommand{\Val}[1]{\F{Val}\:#1}
-\newcommand{\Var}[2]{\F{Var}_{#1} #2}
+\newcommand{\SYMBVal}{\F{Val}\xspace}
+\newcommand{\Val}[1]{\SYMBVal\:#1}
+\newcommand{\SYMBVar}{\F{Var}\xspace}
+\newcommand{\Var}[2]{\SYMBVar_{#1} #2}
 
@@
 \begin{array}{l@{\::\:\:}l}
-\Econd{\!}{\!}{\!}   & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
-\Eplus{\!}{\!}       & \Enat \To \Enat \To \Enat                           \\
-\Ele{\!}{\!}         & \Enat \To \Enat \To \Ebool                          \\
-\Eval{\!}            & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
+\SYMBEcond           & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
+\SYMBEplus           & \Enat \To \Enat \To \Enat                           \\
+\SYMBEle             & \Enat \To \Enat \To \Ebool                          \\
+\SYMBEval            & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
 \end{array}
@@
 %
-The function $\Val{\!}$, used in the definition of $\Eval{\!}$, simply
+The function $\SYMBVal$, used in the definition of $\SYMBEval$, simply
 maps a type $ty$ of the object language to the corresponding type in
-the host language. Hence, the arguments of $\Eval{\!}$ are ensured to
+the host language. Hence, the arguments of $\SYMBEval$ are ensured to
 be of the expected type. We assume that $\Nat$ and $\Bool$ represent
@@
 addition operation $\F{plusHost}$ and a comparison function
-$\F{leHost}$. We define $\Val{\!}$ as follows:
-%
-\[\stk{
-\Val{\!} : \Ty \To \Set \\
+$\F{leHost}$. We define $\SYMBVal$ as follows:
+%
+\[\stk{
+\SYMBVal : \Ty \To \Set \\
 \begin{array}{@{}l@{\:=\:\:}l}
@@
 the type: the resulting description is indexed by $\Ty$. We observe
-that some constructors are always defined, namely $\Econd{\!}{\!}{\!}$
-and $\Eval{\!}$. On the other hand, the $\Eplus{\!}{}$ and
-$\Ele{}{\!}$ constructors are index-dependent. $\Eplus{\!}{}$ is
+that some constructors are always defined, namely $\SYMBEcond$
+and $\SYMBEval$. On the other hand, the $\SYMBEplus$ and
+$\SYMBEle$ constructors are index-dependent. $\SYMBEplus$ is
 defined if and only if the result type -- the index -- is $\Enat$,
-whereas $\Ele{}{\!}$ is defined if and only if the index is
+whereas $\SYMBEle$ is defined if and only if the index is
 $\Ebool$. The actual code precisely follows this intuition, as shown
@@
                  \begin{array}{l}
-                   {\sqr{\Eval{\!}\: \Econd{\!}{\!}{\!}\!\,}} \red{,} \\
+                   {\sqr{\SYMBEval\: \SYMBEcond \,}} \red{,} \\
                    \LAM{\V{ty}}
@@
                  \begin{array}{l}
-                   \sqr{\sqr{\Eplus{\!}{\!}} \quad \sqr{\Ele{\!}{\!}}} \red{,} \\
+                   \sqr{\sqr{\SYMBEplus} \quad \sqr{\SYMBEle}} \red{,} \\
                    \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
@@
 \[\stk{
-\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\:\V{ty}}{\Ty}{\Val{\!}}
+\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\:\V{ty}}{\Ty}{\SYMBVal}
            \To {\Val{\V{ty}}} \\
 \begin{array}{@{}l@{}c@{}l@{\:=\:\:}l}
-\evalOne\: & \_\: & \pair{\Eval{\!}}{\V{x}}{}                                             & \V{x} \\
-\evalOne\: & \_\: & \pair{\Econd{\!}{\!}{\!}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
-\evalOne\: & \_\: & \pair{\Econd{\!}{\!}{\!}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{y} \\
-\evalOne\: & \Enat\: & \pair{\Eplus{\!}{\!}}{\pair{\V{x}}{\V{y}}{}}{}                     & \F{plusHost}\: \V{x}\: \V{y} \\
-\evalOne\: & \Ebool\: & \pair{\Ele{\!}{\!}}{\pair{\V{x}}{\V{y}}{}}{}                      & \F{leHost}\: \V{x}\: \V{y} 
+\evalOne\: & \_\: & \pair{\SYMBEval}{\V{x}}{}                                             & \V{x} \\
+\evalOne\: & \_\: & \pair{\SYMBEcond}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
+\evalOne\: & \_\: & \pair{\SYMBEcond}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{y} \\
+\evalOne\: & \Enat\: & \pair{\SYMBEplus}{\pair{\V{x}}{\V{y}}{}}{}                     & \F{plusHost}\: \V{x}\: \V{y} \\
+\evalOne\: & \Ebool\: & \pair{\SYMBEle}{\pair{\V{x}}{\V{y}}{}}{}                      & \F{leHost}\: \V{x}\: \V{y} 
 \end{array} \\
@@
                                  \toIDesc{\HExprD}\: 
-                                 \Val{\!}\: 
+                                 \SYMBVal\: 
                                  \evalOne\: 
@@
 However, so far, we are only able to define and manipulate
-\emph{closed} terms. By abstracting over $\Val{\!}$, it is possible to
+\emph{closed} terms. By abstracting over $\SYMBVal$, it is possible to
 build and manipulate \emph{open} terms, that is, terms with
-variables. Following $\Val{\!}$, we define $\Var{}{}$ by:
+variables. Following $\SYMBVal$, we define $\SYMBVar$ by:
 %
@@
 %
-Whereas $\Val{\!}$ was mapping the type to the corresponding host type,
-$\Var{}{}$ maps types to a finite set. The finite set -- the context
--- contains closed terms. A variable is therefore a $\Eval{\!}$ that
+Whereas $\SYMBVal$ was mapping the type to the corresponding host type,
+$\SYMBVar$ maps types to a finite set. The finite set -- the context
+-- contains closed terms. A variable is therefore a $\SYMBEval$ that
 contains a pointer to a particular element of the finite set -- an
@@
 \SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ by
-$\SUM{\Val{\!}}{\Var{\V{dom}}{}}$. \note{Is it polite and
+$\SUM{\SYMBVal}{\Var{\V{dom}}{}}$. \note{Is it polite and
   comprehensible to ask that?} This defines a new indexed description,
@@
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumLeft \V{x})  & \Con{\pair{\Eval{\!}}{\V{x}}{}} \\
+\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumLeft \V{x})  & \Con{\pair{\SYMBEval}{\V{x}}{}} \\
 \discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumRight \V{v}) &
@@
 \newcommand{\gammaBool}{\V{\ensuremath{\gamma_{\CN{bool}}}}}
-\newcommand{\GammaSpi}[2]{\F{\ensuremath{\Gamma_{\CN{ty}}}}\: #1\: #2}
+\newcommand{\SYMBGammaSpi}{\F{\ensuremath{\Gamma_{\CN{ty}}}}\xspace}
+\newcommand{\GammaSpi}[2]{\SYMBGammaSpi\: #1\: #2}
 
@@
 %
-Where $\GammaSpi{\!}{\!}$ corresponds to a context, defined by:
-%
-\[\stk{
-\GammaSpi{\!}{\!} : \EnumU \To \Ty \To \Set                                             \\
+Where $\SYMBGammaSpi$ corresponds to a context, defined by:
+%
+\[\stk{
+\SYMBGammaSpi : \EnumU \To \Ty \To \Set                                             \\
 \GammaSpi{\V{dom}}{\V{ty}} = \spi{\V{dom}}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})} 
@@
 our datatype enjoys more structure than what we are given. We are
-facing a free monad, where $\Eval{\!}$ is the \return\ introducing
+facing a free monad, where $\SYMBEval$ is the \return\ introducing
 variables. For convenience, we wrap $\discharge$ in a $\V{\sigma}$
@@
 \FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto \\ \;\;
-    \pair{\pair{\ListCons{\DVar{\!}}{(\fst{\V{E}})}} 
+    \pair{\pair{\ListCons{\SYMBDVar}{(\fst{\V{E}})}} 
                {\LAM{\V{i}}
@@
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\DVar{\!}}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
+\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
 \F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
@@
 
-As previously mentioned, we identify $\Eval{\!}$ with the \return\ of
+As previously mentioned, we identify $\SYMBEval$ with the \return\ of
 the free monad, while the other components are the action of the
@@
                  \begin{array}{l}
-                   \EnumT{\sqr{\Econd{\!}{\!}{\!}}} \red{,} \\
+                   \EnumT{\sqr{\SYMBEcond}} \red{,} \\
                    \LAM{\V{ty}}
@@
 We call this datatype $\HExprFreeD$. By a simple unfolding of
-definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{\!}}$
+definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}$
 corresponds to the syntax of closed terms, $\HExprD$. Similarly,
-$\FreeIMonad{\HExprFreeD}{\Ty}{(\SUM{\Val{\!}}{\Var{\V{dom}}{}})}$
+$\FreeIMonad{\HExprFreeD}{\Ty}{(\SUM{\SYMBVal}{\Var{\V{dom}}{}})}$
 corresponds to expressions with variables, $\HExprVarD{\V{dom}}$.
@@
 Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
-terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{\!}}\: \V{ty}$ to values in
+terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}\: \V{ty}$ to values in
 $\Val{\V{ty}}$. We are left with implementing $\substH$. We simply
@@
        \substI_{\Ty}\: &  \HExprFreeD\: 
-                         (\SUM{\Val{\!}}{\Var{\V{dom}}{}})\: 
-                         \Val{\!} \\
+                         (\SUM{\SYMBVal}{\Var{\V{dom}}{}})\: 
+                         \SYMBVal \\
                       &
@@
 
-Another instance of free monad is $\IDesc{\!}$ itself. Indeed,
-$\DVar{\!}$ is nothing but the \return. The remaining constructors are
+Another instance of free monad is $\SYMBIDesc$ itself. Indeed,
+$\SYMBDVar$ is nothing but the \return. The remaining constructors are
 the carrier functor, trivially indexed by $\Unit$. The carrier functor
@@
 \IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
-                                \sqr{\DConst{\!}\:
-                                 \DProd{\!}{\!}\:
-                                 \DSigma{\!}{\!}\: 
-                                 \DPi{\!}{\!}} \red{,}\\
+                                \sqr{\SYMBDConst\:
+                                     \DProd\:
+                                     \SYMBDSigma\: 
+                                     \SYMBDPi} \red{,}\\
                                   \LAM{\_}\bigRedBracket{\begin{array}{l}

### 2010-04-03 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: definitely fix spacing of operators.

@@
 
-We first define the `small product' $\spi{\!}{\!}$ operator:
+We first define the `small product' $\SYMBspi$ operator:
 %
@@
 This builds a right-nested
-tuple type, demanding an object of $P\:i$ for each $i$ in the given finite
+tuple type, demanding an object of $\V{P}\:\V{i}$ for each $\V{i}$ in the given finite
 domain. We can see these tuples as `jump tables' tabulating dependently
@@
 interpretation---the eliminator we need---by the
-$\F{switch}$ operator, which, unsurprisingly, iterates projection:
+$\SYMBswitch$ operator, which, unsurprisingly, iterates projection:
 %
@@
 \begin{array}{@{}ll}
-\F{switch} : \PITEL{\V{E}}{\EnumU}
+\SYMBswitch : \PITEL{\V{E}}{\EnumU}
                \PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To
@@
 
-The $\spi{\!}{\!}$ and $\F{switch}$ operators deliver dependent
+The $\SYMBspi$ and $\SYMBswitch$ operators deliver dependent
 elimination for finite enumerations, but are rather awkward to
 use directly. We do not write the range for a \(\LAMBINDER\)-abstraction, so
-it is galling to supply \(\V{P}\) for functions defined by $\F{switch}$.
+it is galling to supply \(\V{P}\) for functions defined by $\SYMBswitch$.
 Let us therefore find a way to recover the tedious details of the
@@
 function space---the generalised product---if it is accepted by the
-small product. Propagation fills in the appeal to $\F{switch}$,
+small product. Propagation fills in the appeal to $\SYMBswitch$,
 copying the range information.

### 2010-04-07 papers/ydtm-rev/ydtm-rev.tex
ydtm-rev: kickoff

\documentclass{fundam}

\usepackage{url,fontlock,alltt,latexsym}
\input{macros.ltx}
\input{library.ltx}
\NatPackage

\begin{document}
\title{Why Dependent Types Matter}
\author{Thorsten Altenkirch, Conor McBride and James McKinna}
\maketitle

\begin{abstract}
  We exhibit the rationale behind the design of Epigram, a dependently
  typed programming language and interactive program development
  system, using refinements of a well known program---merge sort---as a
  running example. We discuss its relationship with other proposals
  to introduce aspects of dependent types into functional programming
  languages and sketch some topics for further work in this area.
\end{abstract}

\section{Introduction}

Types matter. That's what they're for---to classify data with respect
to criteria which matter: how they should be stored in memory, whether
they can be safely passed as inputs to a given operation, even who is
allowed to see them. Dependent types are types expressed in terms of
data, explicitly relating their inhabitants to that data. As such,
they enable you to express more of what matters about data.  While
conventional type systems allow us to validate our programs with
respect to a fixed set of criteria, dependent types are much more
flexible, they realize a continuum of precision from the basic
assertions we are used to expect from types up to a complete
specification of the program's behaviour. It is the programmer's
choice to what degree he wants to exploit the expressiveness of such a
powerful type discipline. While the price for formally certified
software may be high, it is good to know that we can pay it in
installments and that we are free to decide how far we want to go.
Dependent types reduce certification to type checking, hence they
provide a means to convince others that the assertions we make about
our programs are correct. Dependently typed programs are, by their
nature, proof carrying code \cite{necula1996,hamid03:safpcc}.

%Dependent
%types are better at mattering on your behalf, and that is why we hope
%they might matter to you.

Functional programmers have started to incorporate many aspects of
dependent types into novel type systems using \emph{generalized
  algebraic data types} and \emph{singleton types}. Indeed, we share
Sheard's vision \cite{sheard-04} of closing the \emph{semantic gap}
between programs and their properties.
While Sheard's language $\Omega$mega approaches this goal by an
evolutionary step from current functional languages like Haskell, we
are proposing a more radical departure with Epigram, exploiting what
we have learnt from proof development tools like LEGO and COQ.

Epigram is a full dependently typed programming language defined by
McBride and McKinna \cite{ConorJames:vfl}, drawing on experience with
the LEGO system. McBride has implemented a prototype which is
available together with basic documentation
\cite{epigram-man,epigram-afp} from the Epigram homepage.%
\footnote{Currently {\tt http://sneezy.cs.nott.ac.uk/epigram/}.}
The prototype implements most of
the features discussed in this article, and we are continuing to
develop it to close the remaining gaps, improve performance and add
new features. Brady has implemented a compiler
\cite{edwin:thesis,edwin.conor.james:family.indices} for the Epigram
language, providing important technology for producing efficient code
from dependently typed programs.

\newcommand{\general}{\mathrm{g}\RW{eneral}}
\newcommand{\nil}{\DC{nil}}
\newcommand{\cons}{\mbox{\(\;\DC{:\!:}\;\)}}
\newcommand{\tup}{\;\times\;}

\begin{figure}[htb]
\[\AR{
\RW{data}\;\;
\Axiom{\TC{Nat}\hab\Type}
\;\;
\RW{where}
\;\;
\Axiom{\DC{0}\hab\TC{Nat}}
\;\;
\Rule{\vn\hab\TC{Nat}}
     {\DC{1+}\;\vn\hab\TC{Nat}}
\medskip \\
%
\RW{data}\;\;
\Axiom{\TC{Order}\hab\Type}
\;\;
\RW{where}
\;\;
\Axiom{\DC{le},\DC{ge}\hab\TC{Order}}
\medskip \\
%
\RW{data}\;\;
\Rule{\vX\hab\Type}
     {\TC{List}\;\vX\hab\Type}
\;\;
\RW{where}
\;\;
\Axiom{\DC{nil}\hab\TC{List}\;\vX}
\;\;
\Rule{\vx\hab\vX \quad
      \vxs\hab\TC{List}\;\vX}
     {\vx\cons\vxs\hab\TC{List}\;\vX}
\medskip \\
%
\RW{let}\;\;
\Rule{\vx,\vy\hab\TC{Nat}}
     {\FN{order}\;\vx\;\vy\hab\TC{Order}}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}c@{\;}l}
\FN{order} & \vx & \vy &\BY\RW{rec}\;\vx \\
\;\;\FN{order} & \DC{0} & \vy & \cq\DC{le} \\
\;\;\FN{order} & (\DC{1+}\vx\Pp) & \DC{0} & \cq\DC{ge} \\
\;\;\FN{order} & (\DC{1+}\vx\Pp) & (\DC{1+}\vy\Pp) &
        \cq\FN{order}\;\vx\Pp\;\vy\Pp \\ 
\end{array}
\medskip \\
%
\RW{let}\;\;\AR{
\Rule{\vxs,\vys\hab\TC{List}\:\TC{Nat}}
     {\FN{merge}\;\vxs\;\vys\hab\TC{List}\;\TC{Nat}}
\smallskip \\
\begin{array}{@{}l@{\;}c@{\;}c@{\;}l}
\FN{merge}&\vxs&\vys&\BY\RW{rec}\;\vxs \\
\;\;  \FN{merge} & \nil & \vys & \cq \vys \\
\;\;  \FN{merge} & (\vx \cons \vxs\Pp) & \vys & \BY \RW{rec}\;\vys \\
\;\;  \;\;\FN{merge} & (\vx \cons \vxs\Pp) & \nil & \cq \vxs \\
\;\;  \;\;\FN{merge} & (\vx \cons \vxs\Pp) & (\vy \cons \vys\Pp) &
    \begin{array}[t]{|c@{\;}l}
      \FN{order}\;\vx\;\vy \\
      \DC{le} & \cq \vx \cons \FN{merge}\;\vxs\Pp\;\vys \\
      \DC{ge} & \cq \vy \cons \FN{merge}\;\vxs\;\vys\Pp \\
    \end{array}
\end{array}
}
\medskip\\
%
\RW{let}\;\;
\Rule{\vxs\hab\TC{List}\;\vX}
     {\FN{deal}\;\vxs\hab\TC{List}\;\vX \tup \TC{List}\;\vX}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}l}
  \FN{deal} & \vxs & \BY \RW{rec}\;\vx \\
\;\;  \FN{deal} & \nil & \cq (\nil;\;\nil) \\
\;\;  \FN{deal} & (\vx \cons \nil) & \cq (\vx \cons \nil;\;\nil) \\
\;\;  \FN{deal} & (\vy \cons \vz \cons \vxs) &
    \begin{array}[t]{|c@{\;}l}
      \FN{deal}\;\vxs \\
      (\vys;\;\vzs) & \cq (\vy\cons\vys;\;\vz\cons\vzs)
    \end{array}
\end{array}
\medskip\\
%
\RW{let}\;\;\AR{
\Rule{\vxs\hab\TC{List}\;\TC{Nat}}
     {\FN{sort}\;\vxs\hab\TC{List}\;\TC{Nat}}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}l}
  \FN{sort} & \vxs &\AR{
    \BY\general \\
    \begin{array}[t]{|c@{\;}l}
      \FN{deal}\;\vxs \\
      (\vys;\;\nil) & \cq \vys \\
      (\vys;\;\vz\cons\vzs) & \cq \FN{merge}\;(\FN{sort}\;\vys)
                                    \;(\FN{sort}\;(\vz\cons\vzs))
    \end{array}}
\end{array}
}
}\]
\caption{Merge-sort, generally\label{fig:mergesortgenerally}}
\end{figure}

In this article we exhibit the rationale behind Epigram's design,
using refinements of a well known program---merge sort---as a running
example.  Our starting point is the implementation shown in Figure
\ref{fig:mergesortgenerally}: it is written in Epigram, but it could have
been written in any functional language. We start by
revisiting the question of totality versus partiality in section
\ref{sec:partial}, showing how $\FN{sort}$ can be made
structurally recursive. Section \ref{sec:static} continues by
addressing the problem of how to maintain static invariants which is
illustrated by implementing a sized $\FN{sort}$. In section
\ref{sec:evidence} we how to use dependent types to maintain static
invariants about dynamic data, which is illustrated by implementing a
version of $\FN{sort}$ which certifies that its output is in order.
We look behind the curtains of the Epigram system in section
\ref{sec:tools} and discuss how dependent types support an extensible
system of programming patterns which include, but are not restricted to,
constructor case analysis and constructor guarded recursion; we also
discuss the design of the
interactive aspect of the Epigram programming environment.  Finally,
we describe areas of further work in \ref{sec:further} and
summarize our conclusions in section \ref{sec:concl}.

Before embarking on this programme let's quickly describe Epigram's
syntax (which may look unusual to functional programmers who have
grown up with languages like ML or Haskell), taking our
implementation of merge sort as an example.  Epigram uses a
two-dimensional syntax for declarations, based on natural deduction
rules, a choice which pays off once type
dependencies become more intricate.  E.g. the declaration of the constructor
$\DC{1+}$ for the datatype%
\footnote{While the declaration of $\TC{Nat}$ provides a convenient interface to the
type of natural numbers, there is no need to implement them using
a unary representation internally. Moreover, we shall also exploit the
syntactic
convenience of using the usual decimal notation to refer to elements of $\TC{Nat}$.}
$\Nat$ is equivalent to writing
$\DC{1+}\hab \TC{Nat}\To\TC{Nat}$.
The conclusion of a declaration visualizes its use---this is important
because parameters of any type can be implicit in Epigram. For example,
when declaring $\DC{nil}$ we do not declare $\vX$---Epigram figures out
from its usage that it must inhabit $\Type$, the type of types and internally
completes the declaration
\[ 
\Rule{\vX\hab\Type} {\DC{nil}_{\vX} \hab\TC{List}\;\vX}
\]
When presented with
$\DC{nil}$, Epigram uses the well known technology established by
Damas and Milner \cite{damas.milner:principal} to figure out the value
of this implicit parameter.  We can also make an implicit parameter
explicit by writing it as a subscript, eg., 
$\DC{nil}_{\TC{Nat}} \hab\TC{List}\;\TC{Nat}$.

Epigram programs are tree-structured: each node has a `left-hand side'
indicating a \emph{programming problem} and a `right hand side'
indicating either how to solve it outright by returning a value, $\TO
t$ ($\TO$ is pronounced `return'), or how to reduce it to subproblems
by deploying a programming
pattern indicated by the $\BY$ symbol ($\BY$ is pronounced `by').
Programming patterns include
structural recursion, like $\RW{rec}\;\vx$ in $\FN{order}$, $\general$
recursion and also case analysis. We suppress nonempty case analyses
for the sake of brevity---they can be recovered by the standard algorithm
\cite{augustsson:compiling.patterns}. If, as in the
$\FN{merge}$ function, we need to analyse the result of an
intermediate computation, we bring it to the left with the $|cdots$
construct ($|$ is pronounced `with').%
 \footnote{The current prototype doesn't yet support
  the suppression of $\RW{case}$ and it doesn't implement the $|$ notation.
  As a consequence its code is more verbose.} 
Here, $\FN{order}$ decides the
$\leq$ relation, returning a value in the enumeration $\TC{Order}$.

We have given a complete program, but Epigram can also typecheck and
evaluate incomplete programs with unfinished sections sitting in
\emph{sheds}, $\SHEDC{\cdots}$, where the typechecker is forbidden to
tread. Programs can be developed interactively, with the machine
showing the available context and the required type, wherever the
cursor may be. Moreover, it is Epigram which generates the left-hand
sides of programs from type information, each time a problem is
simplified with $\BY$ on the right.

\section{Related Work}

Dependent types are a central feature of Martin-L\"of's Type
Theory \cite{martinloef:intuitionistic,nordstrom:programming}, integrating constructive
logic and strongly typed functional programming. Type Theory and its
impredicative extension, the Calculus of Constructions, inspired many
type-based proof systems, such as NUPRL \cite{nuprl-book}, LEGO
\cite{luo.pollack:legomanual} and the widely used COQ
\cite{CoqManualV8}. Magnusson's ALF system \cite{magnusson.nordstrom:alf}
%\cite{magnusson94implementation} 
was not only the first system to
implement inductive families \cite{dybjer:families} and pattern
matching for dependent types \cite{coquand:patternmatching} and it
also pioneered the interactive style of type-driven program and proof
development which inspired Epigram.

Xi and Pfenning's DML (for Dependent ML) \cite{DML99} was an
impressive experiment in exploiting dependent types for a real
functional programming language. DML, like other \emph{Applied Type
  Systems} \cite{ATStypes03}, separates the world of indexing
expressions and programs, thereby 
keeping types unaffected from potentially badly behaved programs.  In contrast to
DML, Augustsson's implementation of the Cayenne language
\cite{augustsson:cayenne}, which also inspired the AGDA proof system
\cite{coquand99structured}, uses full dependency and doesn't
differentiate between static and dynamic types.

Nested types provide a poor man's approach to many indexed
data structures such as square matrices or well scoped $\lambda$-terms,
e.g. see \cite{Hin00Man}. McBride suggested a more general approach
\cite{conor:faking}, exploiting that Haskell's class system provides a
static logic programming language. Having realized the power of
indexed data structures, Cheney and Hinze \cite{ChH03Pha} proposed
to extend the type system to introduce a rich language of index
expressions leading to \emph{generalized
  algebraic datatypes} (GADTs), which are basically inductive families
in a programming language setting \cite{sheard-04,peyton-jones-wobbly}.

%Sheard's $\Omega$mega
%\cite{sheard-04} and \emph{wobbly types} \cite{peyton-jones-wobbly}
%are based on GADTs.

%While Cayenne
%was an important, and mostly successful experiment, Epigram goes beyond
%Cayenne in a number if important points, which: Epigram introduces a notion of
%totality which leads to a better behaved symbolic evaluator, it provides 
%support for inductively defined families and it provides a type-driven
%interactive development system, which seems essential to deal with the
%complexity of dependently type programming. We will discuss all these 
%issues in more detail below.

\section{Why we are Partial to Totality}
\label{sec:partial}

A popular myth, still to be found in learned articles and referee reports on
grant applications, is that dependent types and general recursion do not
mix. This is a misunderstanding, but it's an understandable one.
Let us examine the facts, beginning with the typing rule for application:

\[
\Rule{\vf\hab\forall\vx:\vS\TO\vT[\vx]\quad\vs\hab\vS}
     {\vf\;\vs\hab\vT[\vs]}
\]

It's clear from the premises that, as ever, to check an application we
need to compare the function domain and the argument type. It's also
clear from the rule's conclusion that these types may contain
expressions. If computation is to preserve typings, then
$\vf\;(\red{2}\green{+}\red{2})$ should have the same type as
$\vf\;\red{4}$, so $\vT[\red{2}\green{+}\red{2}]$ must \emph{be} the
same type as $\vT[\red{4}]$. To decide typechecking, we therefore need
to decide some kind of equivalence \emph{up to computation}. There are
various approaches to this problem.

\textbf{The totalitarian approach.}$\quad$
%
Some proof systems based on intensional type theory, including Coq and
Lego, forbid general recursion. As all computations terminate,
equality of types is just syntactic equality of their normal forms.
Decidability of typechecking is a consequence, but it's not the
primary motivation for this choice. As a proof method, general
recursion is wholly bogus---its type, $\forall\vP\TO(\vP\to\vP)\To\vP$
is a blatant lie. General recursion, non-exhaustive
patterns and other such non-total programming features compromise the
logical soundness of a proof system.
Trust is more important than termination in proof checking.

Of course, even in the absence of general recursion, it's possible to
write programs which take a long time---e.g. checking all the basic
configurations of four-colouring problems. That doesn't make dependent
typechecking necessarily intractable: the complexity of the programs
in types is entirely controlled by the programmer---the more you say,
the more you pay, but the more also you can be repaid in terms of
genericity, or precision, or brevity. Georges Gonthier's proof of the
four colour theorem \cite{gonthier:four} is made \emph{tractable} by
type-level computation, because it lets him avoid generating and
checking a separate proof for each configuration---the latter approach
would have involved at least as much work for the computer and a great
deal more work for Georges!

\textbf{The libertarian approach.}$\quad$
%
It's reasonable to allow arbitrary recursion in type-level programs,
provided you have some sort of cut-off mechanism which interrupts
loops when they happen in practice. This is the approach taken by the
Agda proof system, Cayenne and by Haskell with
`undecidable instances'---Haskell's overloading resolution amounts
to executing a kind of `compile-time Prolog'. Agda restores logical
soundness by a separate termination check, performed after typechecking.
The basic point is that you include recursive programs in types at your
own risk: mostly they're benign and typechecking behaves sensibly.

The legendary `loopiness' of dependent typechecking stems from the
particular way the libertarian approach was implemented in Cayenne. It's
perfectly reasonable to implement recursion via fixpoints in the
value-only run-times of functional programming languages, but Lennart
Augustsson's attempt to lift this to the open terms used in dependent
typechecking had an unintended consequence---when even a structurally
recursive function is stuck on a non-constructor input, you can still
expand the fixpoint, potentially putting the system into a
spin: this is intolerable, but it's not inevitable, as the other
`libertarian' systems have shown.

The pragmatic advantage of libertarianism is that we don't have to
care why a program works in order to start playing with it---it
seems a shame to ban certain programs at run-time just to protect
ourselves at compile-time. However, it also seems a shame to forsake
the certainties which totalitarianism supports.

\textbf{The mixed economy.}$\quad$
%
As both of the above have their merits, it seems
sensible to support both, as long as programs are clear about where
they stand.  Dependent types can be used to express schemes of
recursion in the form of induction principles, such as
constructor-guarded recursion for a given inductive datatype. In
Epigram, programs invoke recursion schemes explicitly---each recursive
call must be translatable to an `inductive hypothesis' arising from
such a scheme. Any program which employs only constructor-guarded
recursion, or some fancier scheme derived therefrom, is guaranteed
to be total.

However, general recursion also takes the form of an induction
principle (with a particularly generous inductive hypothesis). We can
make general recursion available, but only by explicit appeal to this
principle. If we suppress the compile-time computational behaviour of
general recursion, we can preserve decidability of typechecking, leaving
computation by explicitly total programs activated.

This opens a range of possibilities, as shown by the implementation of
merge-sort shown in figure \ref{fig:mergesortgenerally}. The
$\FN{order}$ing of the natural numbers with respect to $\le$ is a
one-step constructor guarded recursion. To $\FN{merge}$ two sorted
lists, we need a combination of one-step recursions on each
argument---for each successive element of the first list, we step
along the second list until its rightful place appears. The
$\FN{deal}$ing out of a list into two lists of (roughly) half the
length here exploits a two-step recursion, but this still fits within
the constructor-guarded scheme indicated by the keyword $\RW{rec}$.
However, $\FN{sort}$ performs a peculiar recursion via
$\FN{deal}$---it's not obvious yet how to justify this, so for now we give
up and use $\general$.\footnote{To see how subtle the justification
can be, try swapping the case analysis on $\FN{deal}\;\vxs$ so the
patterns are $(\nil;\;\vys)$ and $(\vz\cons\vzs;\;\vys)$.}

This approach rules nothing out, but it still allows us to observe
guaranteed totality where we can see the explanation. The notational
overhead is not large and could be reduced still further if we were to
install an Agda-style termination checker, inferring simple
explanations when the user omits them.  We could go even further,
pushing the total-versus-general distinction into types by treating
general recursion as an effect which we lock away safely in a monad.
For practical purposes, we should need a better notation for monadic
programming in the \emph{functional} style.  Both of these are active
topics of research.

\subsection{Totality is Good for more than the Soul}

The warm fuzzy feeling you get when you've persuaded your program to live
in a total programming language should not be underestimated. It's a
strong static guarantee---you can say that you've written a \emph{function}
without having to pretend that $\bot$ is a value.
But warm fuzzy feelings don't pay the rent: what are the practical benefits
of virtue?

Randy Pollack has often said to us `the point of writing a proof in a
strongly normalizing calculus is that you don't need to normalize it'.
When you have an expression of a given type in a total language, you
can guarantee that it will compute to a value: if you don't care what
that value is---as is usually the case with a proof---you have no need
to perform the computation. Now we know that we can integrate proofs
of logical properties into our programs at \emph{no run-time cost}.

This is particularly important with proofs of equations. Equality is
defined as follows:

\newcommand{\EQ}{\;\blue{=}\;}
\[
\RW{data}\;\;
\Rule{\vs\hab\vS\quad\vt\hab\vT}
     {\vs\EQ\vt\hab\Type}
\;\;\RW{where}\;\;
\Axiom{\DC{refl}\hab\vt\EQ\vt}
\]

An equation between types induces a coercion from one
to the other which is trivial if the proof is $\DC{refl}$.

\[
\RW{let}\;\;
\Rule{\vQ\hab\vS=\vT\quad \vs\hab\vS}
     {\{\vQ\}\vs\hab\vT}
\qquad
\{\DC{refl}\}\vt \cq\vt
\]

In a partial setting, we need to run $\vQ$ to check that it's
$\DC{refl}$, because trusting a false equation (like
$\TC{Nat}=\TC{Nat}\to\TC{Nat}$) induces a run-time type error.  When
$\vQ$ is total, the compiler can erase
$\{\vQ\}$. 
Contrast this with the proposal to represent type equations 
% current vogue for representing 
in Haskell by isomorphisms, e.g. \cite{BaSw02},  --- even though good programmers
always try to ensure that these functions turn out at run-time to be
functorial liftings of $\texttt{id}$, there is no way to guarantee
this to the compiler, so the isomorphisms must actually be executed.

Moreover, an optimising compiler can exploit totality in various
useful ways. The evaluation strategy of a total program is irrelevant,
so they can be run as strictly or lazily as a heuristic analysis might
suggest. Optimisations like replacing $\texttt{foldr}$ with $\texttt{foldl}$,
which only work with finite values, can be applied safely. Specialisation
by partial evaluation is untroubled by $\bot$. Further the explicit marking
of a program as structurally recursive is a clear invitation to apply
fusion techniques.

All in all, there is no bliss to be had from ignorance of totality;
there is no disadvantage to wisdom.

\subsection{Defusing General Recursion}

A recursive function which happens to be total will generally exploit some
sort of structure, but perhaps not the `native' structure of its inductive
arguments. The totality of the function can be made clear if that structure,
whatever it is, can be brought into the open and expressed inductively.

A typical `divide and conquer' recursion often becomes structural by
introducing an intermediate data structure which represents the
division of the input, built by a process of insertion, and collapsed
structurally by `conquering'. This intermediate data structure thus
corresponds to the control structure built by the original recursion,
which can be reconstructed by fusing the building with the collapse.

As David Turner observed \cite{turner:strong.functional}, defusing
quick-sort exposes the binary search tree structure.  The standard
example of a non-structural program is actually tree-sort---Rod
Burstall's first example of a structural program
\cite{burstall:induction}!

We play the same game with merge-sort in figure
\ref{fig:mergesortstructurally}. The `divide' phase deals out the
input to each sort into two inputs for sub-sorts (roughly) half the
size; the `conquer' phase merges the two sorted outputs into one
(roughly) twice the size. If we build a tree representing the sorting
processes, we find that each node deals its inputs fairly to its
subnodes, with the leaves having none or one.

\newcommand{\pev}{\DC{p0}}
\newcommand{\poo}{\DC{p1}}

\begin{figure}[htb]
\[\AR{
\RW{data}\;\;
\Axiom{\TC{Parity}\hab\Type}
\;\;
\RW{where}
\;\;
\Axiom{\pev,\poo\hab\TC{Parity}}
\medskip\\
%
\RW{data}\;\;
\Rule{\vX\hab\Type}
     {\TC{DealT}\;\vX\hab\Type}
\;\;
\RW{where}
\;\;
\Axiom{\DC{empT}\hab\TC{DealT}\;\vX}
\;\;
\Rule{\vx\hab\vX}
     {\DC{leafT}\;\vx\hab\TC{DealT}\;\vX}
\;\;
\Rule{\vp\hab\TC{Parity}\quad\vl,\vr\hab\TC{DealT}\;\vX}
     {\DC{nodeT}\;\vp\;\vl\;\vr \hab \TC{DealT}\;\vX}
\medskip\\
%
\RW{let}\;\;
\Rule{\vx\hab\vX\quad\vt\hab\TC{DealT}\;\vX}
     {\FN{insertT}\;\vx\;\vt\hab\TC{DealT}\;\vX}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}c@{\;}l}
  \FN{insertT} & \vx & \vt & \BY \RW{rec}\;\vt \\
\;\;  \FN{insertT} & \vx & \DC{empT} & \cq \DC{leafT}\;\vx \\
\;\;   \FN{insertT} & \vx & (\DC{leafT}\;\vy) & \cq
    \DC{nodeT}\;\pev\; (\DC{leafT}\;\vy)\;(\DC{leafT}\;\vx)\\
\;\;   \FN{insertT} & \vx & (\DC{nodeT}\;\pev\;\vl\;\vr) & \cq
    \DC{nodeT}\;\poo\;(\FN{insertT}\;\vx\;\vl)\;\vr \\
\;\;   \FN{insertT} & \vx & (\DC{nodeT}\;\poo\;\vl\;\vr) & \cq
    \DC{nodeT}\;\pev\;\vl\;(\FN{insertT}\;\vx\;\vr)
\end{array}
\medskip\\
%
\RW{let}\;\;
\Rule{\vxs\hab\TC{List}\;\vX}
     {\FN{dealT}\;\vxs\hab\TC{DealT}\;\vX}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}l}
  \FN{dealT} & \vxs & \BY \RW{rec}\;\vxs \\
\;\;  \FN{dealT} & \nil & \cq \DC{empT}\\
\;\;  \FN{dealT} & (\vx\cons\vxs) & \cq\FN{insertT}\;\vx\;(\FN{dealT}\:\vxs)
\end{array}
\medskip\\
%
\RW{let}\;\;
\Rule{\vt\hab\TC{DealT}\;\TC{Nat}}
     {\FN{mergeT}\;\vt\hab\TC{List}\;\TC{Nat}}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}l}
\FN{mergeT} & \vt & \BY \RW{rec}\;\vt \\
\;\;\FN{mergeT} & \DC{empT} & \cq \nil \\
\;\;\FN{mergeT} & (\DC{leafT}\;\vx) & \cq \vx\cons\nil \\
\;\;\FN{mergeT} & (\DC{nodeT}\;\vp\;\vl\;\vr) & \cq
  \FN{merge}\;(\FN{mergeT}\;\vl)\;(\FN{mergeT}\;\vr)
\end{array}
\medskip\\
%
\RW{let}\;\;
\Rule{\vxs\hab\TC{List}\;\TC{Nat}}
     {\FN{sort}\;\vxs\hab\TC{List}\;\TC{Nat}}
\qquad
  \FN{sort} \; \cq \FN{mergeT}\cdot\FN{dealT}
}\]
\caption{Merge-sort, structurally
    (with $\FN{merge}$ as before)\label{fig:mergesortstructurally}}
\end{figure}

Correspondingly, a `dealing' is a binary tree with leaves of weight
zero or one, and nodes off balance by at most one---if we keep a
parity bit at each node, we shall know into which subnode the next
element should be dealt. In effect, we defuse the general recursion as
a composition of folds, with `divide' replacing $[\nil,(\!\cons\!)]$
by $[\DC{empT},\FN{insertT}]$ and `conquer' ($\FN{mergeT}$) replacing
$[\DC{empT},\DC{leafT},\DC{nodeT}]$ by
$[\nil,(\!\cons\nil),\lambda\vp\TO\FN{merge}]$.

Of course, there is no panacea: there are all sorts of ways to write
programs which conceal the structures by which they operate. A
dependent type system provides a rich language of datatypes in which
to expose these structures---other examples include evaluation for the
simply typed $\lambda$-calculus with primitive recursion on
$\TC{Nat}$, which gives a denotational semantics to both types and
terms, and first-order unification, which combines recursion on the
number of available variables with recursion on terms over those
variables.

If you care about totality, it's often easier to write a new program
which works with the relevant structure than to write a proof which
finds the structure which a $\general$-program is hiding. The best way
to tidy up the mess is not to make it in the first place, if you can
possibly avoid it.

\section{Maintaining Invariants by Static Indexing}
\label{sec:static}

An important aspect of many recent and innovative type systems is the
idea of \emph{indexing} datatypes in order to express and enforce
structural invariants. There are various ways this can be achieved: in
Epigram, we define \emph{datatype families} in the style of the Alf
system~\cite{dybjer:families}.  A good introductory example is given
by the \emph{vectors}---lists indexed with their length.

\newcommand{\VEC}{\TC{Vec}}
\newcommand{\vnil}{\DC{vnil}}
\newcommand{\vcons}{\DC{vcons}}
\[
\RW{data}\;\;
\Rule{\vn\hab\Nat\quad\vX\hab\Type}
     {\VEC\;\vn\;\vX\hab\Type}
\;\;
\RW{where}
\;\;
\Axiom{\vnil\hab\VEC\;\DC{0}\;\vX}
\;\;
\Rule{\vx\hab\vX\quad
      \vxs\hab\VEC\;\vn\;\vX}
     {\vcons\;\vx\;\vxs\hab\VEC\;(\DC{1+}\vn)\;\vX}
\]

Case analysis on inductive families~\cite{coquand:patternmatching}
involves \emph{unifying} the type of the scrutinee with the type of
each possible constructor pattern---those patterns for which
constructors clash are rejected as impossible, as in this notorious
example:

\[
\RW{let}\;\;
\Rule{\vxs\hab\VEC\;(\DC{1+}\vn)\;\vX}
     {\FN{vtail}\;\vxs\hab\VEC\;\vn\;\vX}
\qquad
\FN{vtail}\;(\vcons\;\vx\;\vxs) \;\cq\vxs
\]

Vectors admit operations which enforce and
maintain length invariants, such as this `vectorized application':

\newcommand{\vapp}{\;^{\FN{@}}\;}
\[
\RW{let}\;\;
\Rule{\vfs\hab\VEC\;\vn\;(\vS\to\vT)\quad
      \vsss\hab\VEC\;\vn\;\vS}
     {\vfs\vapp\vsss\hab\VEC\;\vn\;\vT}
\qquad
\AR{
\vfs\vapp\vsss\;\BY\RW{rec}\;\vfs \\
\;\;\begin{array}{l@{}c@{}l@{\;}l}
       \vnil & \vapp & \vnil & \cq \vnil \\
       \vcons\;\vf\;\vfs\Pp & \vapp & \vcons\;\vs\;\vsss\Pp & \cq
         \vcons\;(\vf\;\vs)\;(\vfs\Pp\vapp\vsss\Pp)
    \end{array}
}
\]

Sometimes, we need some operations on the indices in order to express
the type of an operation on indexed data. Concatenating vectors is a simple
example

\newcommand{\plus}{\;\FN{+}\;}
\newcommand{\vconc}{\;\FN{+\!+}\;}
\[\begin{array}{l@{\qquad}l}
\RW{let}\;\;
\Rule{\vm,\vn\hab\TC{Nat}}
     {\vm\plus\vn\hab\TC{Nat}}
&
\AR{
\vm\plus\vn\;\BY\RW{rec}\;\vm \\
\;\;\begin{array}{l@{}c@{}l@{\;}l}
       \DC{0} & \plus & \vn & \cq \vn \\
       (\DC{1+}\;\vm\Pp) & \plus & \vn & \cq\DC{1+}\;(\vm\Pp\plus\vn)
    \end{array}
}
\medskip\\
\RW{let}\;\;
\Rule{\vxs\hab\VEC\;\vm\;\vX\qquad\vys\hab\VEC\;\vn\;\vX}
     {\vxs\vconc\vys\hab\VEC\;(\vm\plus\vn)\;\vX}
&
\AR{
\vxs\plus\vys\;\BY\RW{rec}\;\vxs \\
\;\;\begin{array}{l@{}c@{}l@{\;}l}
       \vnil & \vconc & \vys & \cq \vys \\
       \vcons\;\vx\;\vxs\Pp & \vconc & \vys &
         \cq\vcons\;\vx\;(\vxs\Pp\vconc\vys)
    \end{array}
}
\end{array}\]

Note the importance of the index unification in the above example---it's
the instantiation of the first argument's length with $\DC{0}$ or
$(\DC{1+}\;\vm\Pp)$ which enables the length of the concatenation to
compute down to the length of the vectors we actually return in each case.

The fact that the length is some kind of data allows us to write
generic operations by computation over it. This example computes a
constant vector of the required size.

\[
\RW{let}\;\;
\Rule{\vx\hab\vX}
     {\FN{vec}_{\vn}\;\vx\hab\VEC\;\vn\;\vX}
\qquad
\AR{
\FN{vec}_{\vn}\;\vx\;\BY\RW{rec}\;\vn \\
\;\;\begin{array}{l@{\;}l@{\;}l}
      \FN{vec}_{\DC{0}} & \vx & \cq \vnil \\
      \FN{vec}_{(\DC{1+}\;\vn\Pp)} & \vx &
         \cq \vcons\;\vx\;(\FN{vec}_{\vn\Pp}\;\vx)
    \end{array}
}
\]

We write $\FN{vec}$'s length argument as a subscript to indicate that
it is usually to be left implicit when the $\FN{vec}$ function is
used.  For example, we can map a function $\vf$ across a vector $\vxs$
just by writing $\FN{vec}\:\vf\vapp\vxs$, because the type of
$\!\vapp\!$ will require the output of $\FN{vec}$ to have the same
length as $\vxs$.  The technology we have inherited from Damas and
Milner is now being used to infer \emph{value} parameters as well as
types.  The behaviours of $\FN{vec}$ and $\!\vapp\!$ combine
conveniently to give us `vectorized applicative programming' with size
invariants quietly maintained: transposition shows this in action.

\[
\RW{let}\;\;
\Rule{\VV{xij}\hab\VEC\;\vi\;(\VEC\;\vj\;\vX)}
     {\FN{xpose}\;\VV{xij}\hab\VEC\;\vj\;(\VEC\;\vi\;\vX)}
\qquad
\AR{
  \FN{xpose}\;\VV{xij}\;\BY\RW{rec}\;\VV{xij}\\
  \begin{array}{l@{\;}c@{\;}l}
    \FN{xpose} & \vnil & \cq \FN{vec}\;\vnil \\
    \FN{xpose} & (\vcons\;\VV{xj}\;\VV{xi}\Pp\!\vj) & \cq
      \FN{vec}\;\vcons\vapp\VV{xj}\vapp\FN{xpose}\;\VV{xi}\Pp\!\vj
  \end{array}
}
\]

\subsection{Static Indexing and Proofs}

In our definition of $\!\vconc\!$, we were lucky---the computation on
the vectors was in harmony with the computation on the numbers. We are
not always so lucky---if we try to reverse a vector by the usual
accumulating algorithm, we kick against the computational behaviour of
$\!\plus\!$ and the obvious program does not typecheck.

\[
\RW{let}\;\;
\Rule{\vxs\hab\VEC\;\vm\;\vX\qquad\vys\hab\VEC\;\vn\;\vX}
     {\FN{vrevacc}\;\vxs\;\vys\hab\VEC\;(\vm\plus\vn)\;\vX}
\qquad
\AR{
\FN{vrevacc}\;\vxs\;\vys\;\BY\RW{rec}\;\vxs \\
\;\;\begin{array}{l@{\;}c@{\;}l@{\;}l}
       \FN{vrevacc} & \vnil & \vys & \cq \vys \\
       \FN{vrevacc} & (\vcons\;\vx\;\vxs\Pp) & \vys &
         \cq\brownBG{$\FN{vrevacc}\;\vxs\Pp\;(\vcons\;\vx\;\vys)$}
    \end{array}
}
\]

The trouble is that the shaded expression has length
$\vm\Pp\plus(\DC{1+}\;\vn)$, and we require a length of
$\DC{1+}\;(\vm\Pp\plus\vn)$, where $\vxs\Pp$ and $\vys$ have lengths
$\vm\Pp$ and $\vn$ respectively. The fact that these two lengths are
the same does not follow directly from applying the computational
behaviour of $\!\plus\!$, rather it's an algebraic property for which
we can offer an inductive explanation.

\newcommand{\CONG}[1]{\left[#1\right>}
\newcommand{\GNOC}[1]{\left<#1\right]}

\[
\RW{let}\;\;\AR{
\Axiom{\FN{plusSuc}\;\vm\;\vn\hab
       \vm\plus(\DC{1+}\;\vn) \EQ \DC{1+}\;(\vm\plus\vn)}
\smallskip\\
\FN{plusSuc}\;\vm\;\vn\;\BY\RW{rec}\;\vm \\
\;\;\begin{array}{l@{\;}c@{\;}c@{\;}l}
      \FN{plusSuc} & \DC{0} & \vn & \cq\DC{refl} \\
      \FN{plusSuc} & (\DC{1+}\;\vm\Pp) & \vn & \cq
         \CONG{\FN{plusSuc}\;\vm\Pp\;\vn} \\
    \end{array}
}
\]

We write $\CONG{\vq}$ for the proof of an equation
$\vp[\vs]\EQ\vp[\vt]$ where $\vq\hab\vs\EQ\vt$ and
$\GNOC{\vq}$ for the symmetric proof of $\vp[\vt]\EQ\vp[\vs]$
Once we have this proof, we can fix our accumulating reverse:

\[
\RW{let}\;\;\AR{
\Rule{\vxs\hab\VEC\;\vm\;\vX\qquad\vys\hab\VEC\;\vn\;\vX}
     {\FN{vrevacc}_{\vm\;\vn}\;\vxs\;\vys\hab\VEC\;(\vm\plus\vn)\;\vX}
\smallskip\\
\FN{vrevacc}\;\vxs\;\vys\;\BY\RW{rec}\;\vxs \\
\;\;\begin{array}{l@{\;}c@{\;}l@{\;}l}
       \FN{vrevacc} & \vnil & \vys & \cq \vys \\
       \FN{vrevacc}_{(\DC{1+}\:\vm\Pp)\;\vn} & (\vcons\;\vx\;\vxs\Pp) & \vys &
         \cq\{\CONG{\FN{plusSuc}\;\vm\Pp\;\vn}\}\;
            \FN{vrevacc}\;\vxs\Pp\;(\vcons\;\vx\;\vys)
    \end{array}
}
\]

It is perhaps not surprising that to finish the job, we need another
lemma (whose proof is an unremarkable induction):

\[\AR{
\RW{let}\;\;
\Axiom{\FN{plusZero}\;\vn\hab\vn\plus\DC{0} \EQ \vn} \qquad \cdots
\medskip\\
\RW{let}\;\;
\Rule{\vxs\hab\VEC\;\vn\;\vX}
     {\FN{vrev}_{\vn}\;\vxs\hab\VEC\;\vn\;\vX}
\qquad
\FN{vrev}_{\vn}\;\vxs\;\cq \{\CONG{\FN{plusZero}\:\vn}\}\;
                            \FN{vrevacc}\;\vxs\;\vnil
}\]

As we have already mentioned, these proofs are erased at run-time---their
only r\^ole is to make the typechecker aware of more than it can figure
out by computation alone. Perhaps it's undesirable to `repair' programs with
proofs---our $\FN{vrev}$ is certainly more trouble to write than the
reversal of an unsized list, but this is just the work which has to
be done if you want to know that length is preserved. If we don't
care, we don't have to work so hard---we could just return a vector of
\emph{some} length, rather than the \emph{same} length:

\newcommand{\Gp}{\green{'}}
\[
\RW{let}\;\;\AR{
\Rule{\vxs\hab\VEC\;\vm\;\vX\qquad\vys\hab\VEC\;\vn\;\vX}
     {\FN{vrevacc}\Gp\;\vxs\;\vys\hab\exists_{\vl}\TO\VEC\;\vl\;\vX}
\smallskip\\
\FN{vrevacc}\Gp\;\vxs\;\vys\;\BY\RW{rec}\;\vxs \\
\;\;\begin{array}{l@{\;}c@{\;}l@{\;}l}
       \FN{vrevacc}\Gp & \vnil & \vys & \cq \vys \\
       \FN{vrevacc}\Gp & (\vcons\;\vx\;\vxs\Pp) & \vys &
           \cq \FN{vrevacc}\Gp\;\vxs\Pp\;(\vcons\;\vx\;\vys)
    \end{array}
}
\]

The notion of `some length' is expressed via an \emph{implicit
  existential} $\exists_{\vl}\TO\VEC\;\vl\;\vX$. The vector is packed
(inferring the witness) and unpacked (discarding the witness)
automatically. It allows us to recover a less dependent type by hiding
an index---here we recover ordinary lists.  We can write the old
programs with the old precision just as easily as before.

As with totality, we have enough language to negotiate a pragmatic
compromise, adopting more precise methods where the gain is worth the
work.  
%To reject this language is to presuppose the outcome of this
%negotiation in all cases---we are not capable of such courage.  
The
more prudent course, perhaps, is to try to make the unremarkable
proofs as cheap as possible. Xi and Pfenning's approach of equipping
the typechecker with decision procedures for standard problem classes
is a great help in practice: DML has no difficulty discharging the
elementary properties of $\!\plus\!$ we required above. We should certainly
emulate this functionality.

\subsection{Sized Merge-Sort}

We can provide a more substantial example of `the pragmatics of precision' by
rolling out size invariants across our development of merge-sort. We shall
replace the lists by vectors, and we shall seek to ensure that sorting
preserves the length of its input.

How will sizing affect $\FN{merge}$? The output length should be the
sum of the input lengths.

\[
\RW{let}\;\;\AR{
\Rule{\vxs\hab\VEC\;\vm\;\Nat\quad
      \vys\hab\VEC\;\vn\;\Nat}
     {\FN{merge}_{\vm\;\vn}\;\vxs\;\vys\hab\VEC\;(\vm\plus\vn)\;\Nat}
\smallskip \\
\begin{array}{@{}l@{\;}c@{\;}c@{\;}l}
\FN{merge}&\vxs&\vys&\BY\RW{rec}\;\vxs \\
\;\;  \FN{merge} & \vnil & \vys & \cq \vys \\
\;\;  \FN{merge} & (\vcons\;\vx\; \vxs\Pp) & \vys & \BY \RW{rec}\;\vys \\
\;\;  \;\;\FN{merge}_{(\DC{1+}\:\vm\Pp)\;\DC{0}}
        & (\vcons\;\vx\; \vxs\Pp) & \vnil &
        \cq \{\GNOC{\FN{plusZero}\;(\DC{1+}\:\vm\Pp)}\}\;\vxs \\
\;\;  \;\;\FN{merge}_{(\DC{1+}\:\vm\Pp)\;(\DC{1+}\:\vn\Pp)} &
      (\vcons\;\vx\; \vxs\Pp) & (\vcons\;\vy\; \vys\Pp)\\
  \multicolumn{4}{l}{\qquad
    \begin{array}[t]{|c@{\;}l}
      \FN{order}\;\vx\;\vy \\
      \DC{le} & \cq \vcons\;\vx \;(\FN{merge}\;\vxs\Pp\;\vys) \\
      \DC{ge} & \cq \vcons\;\vy \;
      (\{\GNOC{\FN{plusSuc}\;\vm\Pp\;\vn}\}\;
       \FN{merge}\;\vxs\;\vys\Pp) \\
    \end{array}
  }
\end{array}
}
\]

%Note that we need to use our lemmas the other way around when we are
%\emph{using} constructors on the right---in $\FN{vrev}$, we were putting
%constructors on the right.

We shall also need to add sizes to our intermediate $\TC{DealT}$ data
structure. The sizes for $\DC{empT}$ and $\DC{leafT}$ are obvious
enough, but what about $\DC{nodeT}$?  A useful clue is provided by the
\emph{balancing} invariant which our program preserves---the size of
the left subtree is either equal to that of the right subtree or just
one more, depending on the parity bit.  Let's write that down (we use
decimals to abbreviate numerical constants):

\newcommand{\pfog}[1]{\green{\hat{#1}}}
\[\AR{
\RW{let}\;\;\Rule{\vp\hab\TC{Parity}}
                 {\pfog{\vp}\hab\Nat}\quad
\AR{\pfog{\pev}\;\cq\DC{0}\\
    \pfog{\poo}\;\cq\DC{1}\\
   }
\medskip\\
\RW{data}\;\;
\Rule{\vn\hab\Nat\quad\vX\hab\Type}
     {\TC{DealT}\;\vn\;\vX\hab\Type}
\;\;
\RW{where}
\;\;
\AR{
\Axiom{\DC{empT}\hab\TC{DealT}\;\DC{0}\;\vX}
\;\;
\Rule{\vx\hab\vX}
     {\DC{leafT}\;\vx\hab\TC{DealT}\;\DC{1}\;\vX}
\\
\Rule{\vp\hab\TC{Parity}\quad\vl\hab\TC{DealT}\;(\pfog{\vp}\plus\vn)\;\vX
                        \quad\vr\hab\TC{DealT}\;\vn\;\vX}
     {\DC{nodeT}_{\vn}\;\vp\;\vl\;\vr \hab
       \TC{DealT}\;((\pfog{\vp}\plus\vn)\plus\vn)\;\vX}
}
}\]

There is more than one way to write down the size of a
$\DC{nodeT}$.\footnote{Indeed, the above does not ensure that
the subtrees of a node are nonempty---this can be done by replacing
$\vn$ with $(\DC{1+}\;\vn\Pp)$ in the type of $\DC{nodeT}$.} The
choice we make here is motivated by the $\FN{mergeT}$ operation, which now
has a more informative type:

\[
\RW{let}\;\;\AR{
\Rule{\vt\hab\TC{DealT}\;\vn\;\Nat}
     {\FN{mergeT}\;\vt\hab\VEC\;\vn\;\Nat}
\smallskip \\
\begin{array}{@{}l@{\;}c@{\;}l}
\FN{mergeT} & \vt & \BY \RW{rec}\;\vt \\
\;\;\FN{mergeT} & \DC{empT} & \cq \vnil \\
\;\;\FN{mergeT} & (\DC{leafT}\;\vx) & \cq \vcons\;\vx\;\nil \\
\;\;\FN{mergeT} & (\DC{nodeT}\;\vp\;\vl\;\vr) & \cq
  \FN{merge}\;(\FN{mergeT}\;\vl)\;(\FN{mergeT}\;\vr)
\end{array}
}
\]

We simply chose the return type for $\DC{nodeT}$ which made the old code
for $\FN{mergeT}$ go through as it stood, given the new type of
$\FN{merge}$! Of course, we shall pay when we come to write
$\FN{insertT}$---we could shift the burden the other way by taking the
size of a $\DC{nodeT}$ to be
$\pfog{\vp}\plus\vn\;\green{\ast}\;\DC{2}$.

\[
\RW{let}\;\;\AR{
\Rule{\vx\hab\vX\quad\vt\hab\TC{DealT}\;\vn\;\vX}
     {\FN{insertT}\;\vx\;\vt\hab\TC{DealT}\;(\DC{1+}\;\vn)\;\vX}
\smallskip \\
\begin{array}{@{}l@{\;}c@{\;}c@{\;}l}
  \FN{insertT} & \vx & \vt & \BY \RW{rec}\;\vt \\
\;\;  \FN{insertT} & \vx & \DC{empT} & \cq \DC{leafT}\;\vx \\
\;\;   \FN{insertT} & \vx & (\DC{leafT}\;\vy) & \cq
    \DC{nodeT}\;\pev\; (\DC{leafT}\;\vy)\;(\DC{leafT}\;\vx)\\
\;\;   \FN{insertT} & \vx & (\DC{nodeT}\;\pev\;\vl\;\vr) & \cq
    \DC{nodeT}\;\poo\;(\FN{insertT}\;\vx\;\vl)\;\vr \\
\;\;   \FN{insertT} & \vx & (\DC{nodeT}_{\vn}\;\poo\;\vl\;\vr) & \cq
    \{\CONG{\FN{plusSuc}\;\vn\;\vn}\}\;
    \DC{nodeT}_{(\DC{1+}\:\vn)}\;\pev\;\vl\;(\FN{insertT}\;\vx\;\vr)
\end{array}
}
\]

The damage is not too bad---we just have to appeal to algebraic properties of $\plus$,
to show that the constructed tree of size $(\DC{1+}\vn)\plus\vn$ fits the constraint 
$\DC{1+}(\vn\plus\vn)$.
%when we insert on the right, we still want
%the $\DC{1+}$ on the left!
This leaves $\FN{dealT}$ and $\FN{sort}$ with new types, but basically
the same code:

\[\AR{
\RW{let}\;\;
\Rule{\vxs\hab\VEC\;\vn\;\vX}
     {\FN{dealT}\;\vxs\hab\TC{DealT}\;\vn\;\vX}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}l}
  \FN{dealT} & \vxs & \BY \RW{rec}\;\vxs \\
\;\;  \FN{dealT} & \vnil & \cq \DC{empT}\\
\;\;  \FN{dealT} & (\vcons\;\vx\;\vxs) &
      \cq\FN{insertT}\;\vx\;(\FN{dealT}\:\vxs)
\end{array}
\medskip\\
%
\RW{let}\;\;
\Rule{\vxs\hab\VEC\;\vn\;\Nat}
     {\FN{sort}\;\vxs\hab\VEC\;\vn\;\Nat}
\qquad
  \FN{sort} \; \cq \FN{mergeT}\cdot\FN{dealT}
}\]

It seems appropriate at this point to emphasize the importance of feedback
from the typechecker when doing a development like this. It's the typechecker
which tells you which lemmas you need and where you need to insert them,
and it's the constraints which arise during typechecking which tell you
how to engineer a data structure's indices so that its operations typecheck
where possible. The computational coincidences between the indices
we encounter are really a matter of care, not luck.

\subsection{A Question of Phase}

In our sorting example, the natural numbers $\TC{Nat}$ are playing two
separate parts---dynamically, they represent the data being sorted;
statically, they give sizes to the data structures involved in the
process. All the case analysis happens over the indexed datatypes,
rather than the indices themselves, so there is no need for sizes at
run-time. It would appear that, in this example at least, the data on
which types depend is entirely static. Although we need something like
the natural numbers within the language of types, should it be the
natural numbers? Perhaps it would be simpler if, as well as the natural
numbers, there was something else exactly like them.

It's not as ridiculous as it sounds: it enables us to keep the
workaday term language out of types, allowing dependency only on the
static things.  This keeps the type/term distinction in alignment with
the phase distinction, separating the static $\forall$ from
the dynamic $\to$, each with their own distinct notions of
abstraction, application and now of datatype.

Of course, static datatypes are not quite enough: our $\TC{DealT}$
datatype family relies on operations such as $\plus$. We need static
functions over static data, together with the associated partial
evaluation technology. We need to find design criteria for this
emerging static programming language. Will all our static data
structures be monomorphic? Do we believe that static data will never
need to be indexed itself? It would be bold to imagine that `yes' is
the answer to these questions.  At what point will we stop extending
the static language with a replica of the dynamic language?

We're beginning to see more and more of the same phenomema showing up
on either side of the phase distinction; we're even beginning to see
the potential emergence of a phase \emph{hierarchy}. Occam's razor
suggests that we should understand `data' and `function' once,
regardless of phase, since the phase distinction no longer
distinguishes where these notions arise.

But Occam's razor is a subjective instrument, so we cannot presume
that others will come to the same judgment as ourselves. We can,
however, examine the impact of the underlying presumption that `the
data on which types depend is entirely static'.

\section{Evidence is About Data; Evidence is Data}
\label{sec:evidence}

Type systems without dependency on dynamic data tend to satisfy the
\emph{replacement} property---any subexpression of a well typed
expression can be replaced by an alternative subexpression of the same
type in the same scope, and the whole will remain well typed. For
example, in Java or Haskell, you can always swap the $\texttt{then}$
and $\texttt{else}$ branches of conditionals and nothing will go
wrong---nothing of any static significance, anyway. The simplifying
assumption is that within any given type, one value is as good as
another. These type systems have no means to express the way that
different data mean different things, and should be treated accordingly
in different ways. That is why dependent types matter.

More specifically, we have seen how the apparatus of dependent types can
be used to maintain the length invariant in our sorting algorithm, but
that the length can essentially be regarded as a prior static notion
which the code must \emph{respect} dynamically.  What if we wanted to
guarantee the the output of our sorting algorithm is in order? The
order is not a prior static notion---that is why we need a sorting
algorithm---the order is \emph{established} by run-time testing of the
dynamic data. Can we observe this fact statically? As things stand,
$\TC{Order}$ does not matter: we can swap around the $\DC{le}$ and
$\DC{ge}$ outputs of our $\FN{order}$ test without affecting the well
typedness of $\FN{merge}$. How can we make $\TC{Order}$ matter? By
making $\TC{Order}$ a dependent type!

\subsection{Evidence of Ordering}

We replace the uninformative type $\TC{Order}$ by an inductive family with
$\TC{Order}\;\vx\;\vy$ expressing that  $\vx$ and $\vy$ can be ordered
and each possibility can be established with \emph{evidence}.
%There are two ways a given $\vx$ and $\vy$ can be ordered,
%and each possibility can be established with \emph{evidence}.

\newcommand{\LE}{\mathrel{\blue{\le}}}
\[\AR{
\RW{data}\;\;
\Rule{\vx,\vy\hab\Nat}
     {\vx\LE\vy\hab\Type}
\;\;\RW{where}\;\;
\orange{\cdots}
\\
\RW{data}\;\;
\Rule{\vx,\vy\hab\Nat}
     {\TC{Order}\;\vx\;\vy\hab\Type}
\;\;\RW{where}\;\;
\Rule{\VV{xley}\hab\vx\LE\vy}
     {\DC{le}\;\VV{xley}\hab\TC{Order}\;\vx\;\vy}
\;\;
\Rule{\VV{ylex}\hab\vy\LE\vx}
     {\DC{ge}\;\VV{ylex}\hab\TC{Order}\;\vx\;\vy}
}\]

\newcommand{\HOLE}{\yellowBG{[]}}
We shall give $\LE$ its constructors shortly. First, let us see what
happens to the $\FN{order}$ test---for a start, its type now tells us
what its output says about its input! We provide empty sheds $\HOLE$
which will be completed once we have decided how to represent $\LE$ proofs.
The freedom to delay implementation decisions while type checking
the rest of the code is an essential feature of Epigram which we 
further elaborate in the next section.

%We have left $\HOLE$s standing
%for the $\LE$ proofs.

\[
\RW{let}\;\;
\Axiom{\FN{order}\;\vx\;\vy\hab\TC{Order}\;\vx\;\vy}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}c@{\;}l}
\FN{order} & \vx & \vy &\BY\RW{rec}\;\vx \\
\;\;\FN{order} & \DC{0} & \vy & \cq \DC{le}\;\HOLE \\
\;\;\FN{order} & (\DC{1+}\vx\Pp) & \DC{0} & \cq \DC{ge}\;\HOLE \\
\;\;\FN{order} & (\DC{1+}\vx\Pp) & (\DC{1+}\vy\Pp) &
    \begin{array}[t]{|c@{\;}l}
      \FN{order}\;\vx\Pp\;\vy\Pp \\
      \DC{le}\;\VV{x\Pp\!ley\Pp} & \cq \DC{le}\;\HOLE \\
      \DC{ge}\;\VV{y\Pp\!lex\Pp} & \cq \DC{ge}\;\HOLE \\
    \end{array}
\end{array}
\]

This program is almost as before, except that we cannot pass the recursive
call back directly---it is the ordering of $\vx\Pp$ and $\vy\Pp$, but
we need to order their successors. We can treat this partial program as
a clue to good constructors for $\LE$. What should we wish them to
be? Well, our first $\HOLE$ requires a proof that $\DC{0}\LE\vy$, so let
us have

\[
\Axiom{\DC{le0}\hab\DC{0}\LE\vy}
\]

Actually, that will also satisfy our second $\HOLE$, which needs a
proof of $\DC{0}\LE(\DC{1+}\vx\Pp)$. Our third $\HOLE$ requires us to
establish $(\DC{1+}\vx\Pp)\LE (\DC{1+}\vy\Pp)$, given
$\VV{x\Pp\!ley\Pp}\hab\vx\Pp\LE\vy\Pp$, and the fourth $\HOLE$ is
similar, so let us have

\[
\Rule{\VV{xley}\hab\vx\LE\vy}
     {\DC{leS}\;\VV{xley}\hab(\DC{1+}\vx\Pp)\LE (\DC{1+}\vy\Pp)}
\]

The two seem like a reasonable definition of $\LE$, and they certainly
enable us to fill in our $\HOLE$s.

\[
\RW{let}\;\;
\Axiom{\FN{order}\;\vx\;\vy\hab\TC{Order}\;\vx\;\vy}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}c@{\;}l}
\FN{order} & \vx & \vy &\BY\RW{rec}\;\vx \\
\;\;\FN{order} & \DC{0} & \vy & \cq \DC{le}\;\DC{le0} \\
\;\;\FN{order} & (\DC{1+}\vx\Pp) & \DC{0} & \cq \DC{ge}\;\DC{le0} \\
\;\;\FN{order} & (\DC{1+}\vx\Pp) & (\DC{1+}\vy\Pp) &
    \begin{array}[t]{|c@{\;}l}
      \FN{order}\;\vx\Pp\;\vy\Pp \\
      \DC{le}\;\VV{x\Pp\!ley\Pp} & \cq
             \DC{le}\;(\DC{leS}\;\VV{x\Pp\!ley\Pp}) \\
      \DC{ge}\;\VV{y\Pp\!lex\Pp} & \cq
             \DC{ge}\;(\DC{leS}\;\VV{y\Pp\!lex\Pp}) \\
    \end{array}
\end{array}
\]

What has happened here? We started with a program which \emph{did} the
right thing but did not \emph{say} so. It should be no surprise that
when we try to make this program say what it does, we learn how to say
the right thing.

However, some of you may be wondering whether it is
worth saying the right thing, if it means spending heap on this data
structure of evidence and losing the tail-call optimisation into the
bargain. Fortunately, our $\LE$ type has the property of being
\emph{content-free}~\cite{edwin.conor.james:family.indices}: just as with
$\EQ$, for any given indices $\vx$ and $\vy$, $\vx\LE\vy$ contains at
most one value, so the evidence can be erased at run-time and the
tail-call restored. Moreover, it is no accident that $\LE$ is
content-free: our method of packing up the cases arising
ensured that $\DC{le0}$ and $\DC{leS}$ covered distinct indices, and
that the indices of any recursive proofs were determined in turn.

Just to be sure, let us check that $\LE$ is a total ordering---given
$\FN{order}$, we just need

\[\AR{
\RW{let}\;\;
\Axiom{\FN{leRefl}_{\vx}\hab\vx\LE\vx}
\qquad
  \AR{
    \FN{leRefl}_{\vx}\;\BY\RW{rec}\;\vx \\
    \;\;\begin{array}{@{}l@{\;}l}
       \FN{leRefl}_{\DC{0}} & \cq \DC{le0} \\
       \FN{leRefl}_{(\DC{1+}\:\vx\Pp)} & \cq \DC{leS}\;\FN{leRefl}_{\vx\Pp}
    \end{array}
  }
\medskip \\
\RW{let}\;\;
\Rule{\VV{xley}\hab\vx\LE\vy\quad\VV{ylez}\hab\vy\LE\vz}
     {\FN{leTrans}\;\VV{xley}\;\VV{ylez}\hab\vx\LE\vz}
\qquad
  \begin{array}[t]{@{}l@{\;}c@{\;}c@{\;}l}
    \FN{leTrans}     & \VV{xley} & \VV{ylez} & \BY\RW{rec}\;\VV{xley} \\
    \;\;\FN{leTrans} & \DC{le0}  & \VV{ylez} & \cq \DC{le0} \\
    \;\;\FN{leTrans} & (\DC{leS}\;\VV{xley}\Pp)
                     & (\DC{leS}\;\VV{ylez}\Pp) 
                     & \cq \DC{leS}\;
                       (\FN{leTrans}\;\VV{xley}\Pp\;\VV{ylez}\Pp)\\
 \end{array}
\medskip \\
\RW{let}\;\;
\Rule{\VV{xley}\hab\vx\LE\vy\quad\VV{ylex}\hab\vy\LE\vx}
     {\FN{leASym}\;\VV{xley}\;\VV{ylex}\hab\vx\EQ\vy}
\qquad
  \begin{array}[t]{@{}l@{\;}c@{\;}c@{\;}l}
    \FN{leASym}     & \VV{xley} & \VV{ylex} & \BY\RW{rec}\;\VV{xley} \\
    \;\;\FN{leASym} & \DC{le0}  & \DC{le0} & \cq \DC{refl} \\
    \;\;\FN{leASym} & (\DC{leS}\;\VV{xley}\Pp)
                    & (\DC{leS}\;\VV{ylex}\Pp) 
                    & \cq \CONG{\FN{leASym}\;\VV{xley}\Pp\;\VV{ylex}\Pp}\\
 \end{array}
}\]

As it happens, these properties of $\LE$ are not necessary in order to
implement sorting. To be sure that a list is sorted, you need at least to
have checked that each adjacent pair is in order---that it's \emph{locally
sorted}. It's not hard to see that a list can always be locally sorted
with respect to any binary relation which always holds one way around or
the other (when inserting a new element, if it fits nowhere before the
end, then it must fit at the end). Of course, knowing that it is
also a partial order enhances what you can deduce from a locally sorted
list.

\subsection{Locally Sorted Lists}

How shall we define locally sorted lists? We shall clearly have to index
lists with some sort of interval, but there are several ways we might
do it: one bound or two? open bounds or closed bounds? As with the design
of the sized $\TC{DealT}$, we should take care to ensure that the decision
leads to operations which are as cleanly defined as possible. For pedagogical
purposes, we shall sort lists rather than vectors---sizing and sorting are
independent refinements of the list structure. We do not verify the fact
that the resulting list is a permutation of the input here.

\textbf{One bound or two?}$\quad$ In order to do a `cons', we shall certainly
need to know that the head and the tail are suitably ordered, so the tail
will require a lower bound. Meanwhile, $\FN{merge}$ makes no restrictions
between the bounds of its inputs---only between the input bounds and the
output bounds. That suggests that we can get away with a lower bound for
this example. Of course, if we wanted to \emph{concatenate} sorted lists
(in an algorithm based on pivoting, say), we should need upper bounds too.

\textbf{Open or closed?}$\quad$ It is perhaps a little tricky to give a
precise lower bound for the empty list---we could make a bound `elements
lifted with $\red{\infty}$' and lift $\LE$ accordingly:

\[
  \RW{data}\;\;
  \Rule{\vb\hab\TC{Lifted}\;\Nat}
       {\TC{CList}\;\vb\hab\Type}
  \;\;\RW{where}\;\;
  \Axiom{\DC{cnil}\hab\TC{CList}\;\red{\infty}} \;\;
  \Rule{\vx\hab\Nat\quad \VV{xley}\hab\vx\LE\vy\quad
        \vxs\hab\TC{CList}\;\vy}
       {\DC{ccons}\;\vx\;\VV{xley}\;\vxs\hab\TC{CList}\;(\DC{lift}\;\vx)}
\]

The lifting is not a big issue here---we only get away without
$\red{-\infty}$ because $\Nat$ stops at $\DC{0}$. The advantage of this
definition is its precision: each list is assigned its tightest bound.
The disadvantage of this definition is also its precision---when we
are making a $\TC{CList}$, we must either specify its lower bound
(eg., $\FN{sort}$ is bounded by $\FN{min}$) and satisfy that specification,
or say `don't care' with an existential. But we do care! The recursive calls
in $\FN{merge}$ can't have any old lower bound---the lower bound of the tail
must be at least the head!

The fact that we can \emph{assign} closed bounds to sorted lists does
not necessarily make them a good choice for a type, because we also
need to \emph{prescribe} bounds. We can usually think of an open bound,
even if it is not the best one. Let us try:

\[
  \RW{data}\;\;
  \Rule{\vb\hab\Nat}
       {\TC{OList}\;\vb\hab\Type}
  \;\;\RW{where}\;\;
  \Axiom{\DC{onil}\hab\TC{OList}\;\vb} \;\;
  \Rule{\vx\hab\Nat\quad \VV{blex}\hab\vb\LE\vx\quad
        \vxs\hab\TC{OList}\;\vx}
       {\DC{ocons}\;\vx\;\VV{blex}\;\vxs\hab\TC{OList}\;\vb}
\]

Here, we have that $\DC{onil}$ trivially satisfies any prescribed bound,
whilst for $\DC{ocons}$, the head must exceed the prescribed lower bound
and bound the tail in turn. Any old sorted list can certainly be
represented as an element of $\TC{OList}\;\DC{0}$. Meanwhile, if \emph{both}
inputs to $\FN{merge}$ share a lower bound, then the output certainly
shares it too.

\[
\RW{let}\;\;\AR{
\Rule{\vxs,\vys\hab\TC{OList}\:\vb}
     {\FN{merge}\;\vxs\;\vys\hab\TC{OList}\;\vb}
\smallskip \\
\begin{array}{@{}l@{\;}c@{\;}c@{\;}l}
\FN{merge}&\vxs&\vys&\BY\RW{rec}\;\vxs \\
\;\;  \FN{merge} & \DC{onil} & \vys & \cq \vys \\
\;\;  \FN{merge} & (\DC{ocons}\;\vx\;\VV{blex}\;\vxs\Pp)
                 & \vys & \BY \RW{rec}\;\vys \\
\;\;  \;\;\FN{merge} & (\DC{ocons}\;\vx\;\VV{blex}\;\vxs\Pp) & \DC{onil} & \cq \vxs \\
\;\;  \;\;\FN{merge} & (\DC{ocons}\;\vx\;\VV{blex}\;\vxs\Pp)
                     & (\DC{ocons}\;\vy\;\VV{bley}\;\vys\Pp) \\
\multicolumn{4}{l}{\;\; \;\; \quad
    \begin{array}[t]{|c@{\;}l}
      \FN{order}\;\vx\;\vy \\
      \DC{le}\;\VV{xley} & \cq \DC{ocons}\;\vx\;\VV{blex}\;
              (\FN{merge}\;\vxs\Pp\;(\DC{ocons}\;\vy\;\VV{xley}\;\vys\Pp)) \\
      \DC{ge}\;\VV{ylex} & \cq \DC{ocons}\;\vy\;\VV{bley}\;
              (\FN{merge}\;(\DC{ocons}\;\vx\;\VV{ylex}\;\vxs\Pp)\;\vys\Pp) \\
    \end{array}
}
\end{array}
}
\]

Each input list already satisfies the bound required for the output, so
the $\DC{onil}$ cases are trivial. When we have two $\DC{ocons}$es, we
know both heads satisfy the lower bound, but whichever we pick must
bound the recursive $\FN{merge}$. Hence, we had better pick the smaller
one---the evidence we get from $\FN{order}$ is exactly what we need to
show that the list which keeps its head satisfies the newer tighter bound.

Now we can flatten a $\TC{DealT}$ into a sorted list. This gives us a
new back end which we can compose with the old $\FN{dealT}$ to get a
sort which produces sorted output:

\[\AR{
\RW{let}\;\;
\Rule{\vt\hab\TC{DealT}\;\TC{Nat}}
     {\FN{mergeT}\;\vt\hab\TC{OList}\;\DC{0}}
\qquad
\begin{array}[t]{@{}l@{\;}c@{\;}l}
\FN{mergeT} & \vt & \BY \RW{rec}\;\vt \\
\;\;\FN{mergeT} & \DC{empT} & \cq \DC{onil} \\
\;\;\FN{mergeT} & (\DC{leafT}\;\vx)
                & \cq \DC{ocons}\;\vx\;\DC{le0}\;\DC{onil} \\
\;\;\FN{mergeT} & (\DC{nodeT}\;\vp\;\vl\;\vr) & \cq
  \FN{merge}\;(\FN{mergeT}\;\vl)\;(\FN{mergeT}\;\vr)
\end{array}
\medskip\\
%
\RW{let}\;\;
\Rule{\vxs\hab\TC{List}\;\TC{Nat}}
     {\FN{sort}\;\vxs\hab\TC{OList}\;\DC{0}}
\qquad
  \FN{sort} \; \cq \FN{mergeT}\cdot\FN{dealT}
}\]

\textbf{Remark.}$\quad$ In the above definitions of $\FN{merge}$ and
$\FN{mergeT}$, all of the $\le$ proofs which we supply in the lists
we build are either by $\DC{le0}$ or by direct appeal to a hypothesis
in scope. The proofs which we uncover by case analysis are only used
in this way. It seems reasonable to consider suppressing all of them
by default from the explicit syntax of the program. Implicit hypotheses
could be kept in the context and searched whenever an implicit proof is
required, in much the way that Haskell handles the implicit dictionaries
when unpacking and packing existential types. Of course, a `manual
override' is still necessary---not all proofs are so immediate.

\subsection{Programming with Evidence}

The idea that one datatype can represent evidence about the values in
another is alien to mainstream functional programming languages, but its
absence is beginning to cause pain. A recent experiment in `dynamically
typed' generic programming---the `Scrap Your Boilerplate' library
of traversal operators by Ralf L\"ammel and Simon Peyton
Jones~\cite{laemmel.peytonjones:boilerplate}---is a case in point.
The library relies on a `type safe cast' operator, effectively comparing
types at run time by comparing their encodings as data:

\begin{alltt}
  cast :: (Typeable a, Typeable b) => a -> Maybe b
  cast x = r
         where
           r = if typeOf x == typeOf (get r)
               then Just (unsafeCoerce x)
               else Nothing

           get :: Maybe a -> a
           get x = undefined
\end{alltt}

Here, the Boolean test $\texttt{typeOf x == typeOf (get r)}$ serves
the purpose of comparing type $\texttt{a}$ with type $\texttt{b}$, but
the Haskell typechecker cannot interpret the value $\texttt{True}$ as
a reason to unify types. In the absence of evidence, the programmers
resort to coercion. \emph{We} trust them, of course, but this degree
of trust should not be necessary. This is an example
where static dependency on dynamic data is the solution, not the
problem.

We now work in a setting where we expect operations which test their
input in some way to have a dependent type which explains what the
output reveals about the input---$\FN{order}$ is a simple
example. Without some kind of movement between terms and types this is
impossible---you can exploit valid data (eg., writing a type-preserving
evaluator for a typed expression language) , but you cannot \emph{validate}
it dynamically (eg., by writing a typechecker).

Traditional dependent types achieve this movement directly---the
argument of a function can appear in the type of its result, and often
will, if the result is evidence of the argument's properties. An
alternative, adopted for DML's numerical indices, is to push the other
way with \emph{singleton types}---types of dynamic data constrained to
be equal to the corresponding static data. We would have something like
this:

\[
  \FN{order}\hab\forall\vx,\vy:\Nat\TO
    \TC{Only}\;\vx \To \TC{Only}\;\vy \To \TC{Order}\;\vx\;\vy
\]

Here every piece of data about which we seek evidence must be
abstracted twice---the type of evidence depends on the static copy,
but the testing itself is performed on the dynamic copy. In the
context of DML, this is a sensible separation---erasing the indices is
intended to yield a well typed SML program. In a broader context,
where we might need to represent properties of any kind of dynamic
data, this approach seems unlikely to scale. Effectively, we may need
to replace each type $\vT$ by the existential pairing of its static
and dynamic copies $\exists\vt:\vT\TO\TC{Only}\;\vt$ throughout our
programs, in order to have data at all the levels where it is used.

Singleton types thus provide a convenient way to retro-fit a more
sophisticated type system to an existing compiler architecture.
With just one notion of data at all levels, dependent types
provide a convenient way for programmers to exploit the potential
of working with data as evidence. Our typechecker example in
\cite{ConorJames:vfl} not only generates enough evidence about
the types being inferred and compared to feed the tagless interpreter,
it generates evidence about the program being checked---it is the
first typechecker in the literature which is statically guaranteed to
check its input!

%=======
%\subsection{Programming with Evidence}

%The idea that one datatype can represent evidence about the values in
%another is alien to mainstream functional programming languages, but its
%absence is beginning to cause pain. A recent experiment in `dynamically
%typed' generic programming---the `Scrap Your Boilerplate' library
%of traversal operators by Ralf L\"ammel and Simon Peyton
%Jones~\cite{laemmel.peytonjones:boilerplate}---is a case in point.
%The library relies on a `type safe cast' operator, effectively comparing
%types at run time by comparing their encodings as data:

%\begin{alltt}
%  cast :: (Typeable a, Typeable b) => a -> Maybe b
%  cast x = r
%         where
%           r = if typeOf x == typeOf (get r)
%               then Just (unsafeCoerce x)
%               else Nothing

%           get :: Maybe a -> a
%           get x = undefined
%\end{alltt}

%Here, the Boolean test $\texttt{typeOf x == typeOf (get r)}$ serves
%the purpose of comparing type $\texttt{a}$ with type $\texttt{b}$, but
%the Haskell typechecker cannot interpret the value $\texttt{True}$ as
%a reason to unify types. In the absence of evidence, the programmers
%resort to coercion. \emph{We} trust them, of course, but this degree
%of trust should not be necessary. This is an example
%where static dependency on dynamic data is the solution, not the
%problem.

%We now work in a setting where we expect operations which test their
%input in some way to have a dependent type which explains what the
%output reveals about the input---$\FN{order}$ is a simple
%example. Without some kind of movement between terms and types this is
%impossible---you can exploit valid data (eg., writing a type-preserving
%evaluator for a typed expression language) , but you cannot \emph{validate}
%it dynamically (eg., by writing a typechecker).

%Traditional dependent types achieve this movement directly---the
%argument of a function can appear in the type of its result, and often
%will, if the result is evidence of the argument's properties. An
%alternative, adopted for DML's numerical indices, is to push the other
%way with \emph{singleton types}---types of dynamic data constrained to
%be equal to the corresponding static data. We would have something like
%this:

%\[
%  \FN{order}\hab\forall\vx,\vy:\Nat\TO
%    \TC{Only}\;\vx \To \TC{Only}\;\vy \To \TC{Order}\;\vx\;\vy
%\]

%Here every piece of data about which we seek evidence must be
%abstracted twice---the type of evidence depends on the static copy,
%but the testing itself is performed on the dynamic copy. In the
%context of DML, this is a sensible separation---erasing the indices is
%intended to yield a well typed SML program. In a broader context,
%where we might need to represent properties of any kind of dynamic
%data, this approach seems unlikely to scale. Effectively, we may need
%to replace each type $\vT$ by the existential pairing of its static
%and dynamic copies $\exists\vt:\vT\TO\TC{Only}\;\vt$ throughout our
%programs, in order to have data at all the levels where it is used.

%Singleton types thus provide a convenient way to retro-fit a more
%sophisticated type system to an existing compiler architecture.
%With just one notion of data at all levels, dependent types
%provide a convenient way for programmers to exploit the potential
%of working with data as evidence. Our typechecker example in
%\cite{conor.james:viewfromleft} not only generates enough evidence about
%the types being inferred and compared to feed the tagless interpreter,
%it generates evidence about the program being checked---it is the
%first typechecker in the literature which is statically guaranteed to
%check its input!

\section{The Tools of the Trade}
\label{sec:tools}

Programming is a complex task which can be made easier for people to
do with the help of computers. The conventional cycle of programming
with a text editor then compiling in `batch mode' is a welcome
shortening of the feedback loop since the days of punched cards, but it
clearly under-uses the technology available today. Any typed
programming language can benefit from the capacity---but not
necessarily the compulsion---to invoke the typechecker incrementally
on incomplete subprograms whilst they are under development. The more
powerful the type system, the more pressing this need becomes---it just
gets harder to do it in your head, especially when types contain
\emph{computations}, for which computers are inherently useful.

Moreover, a type acts as a partial specification of a program, and can
thus be used to narrow the search space for correct programs, even if
only as a key for searching a library. The choice of a program's type
is inherently suggestive of the programming strategies with which that
type is naturally equipped---constructors and case analysis for
datatypes, abstraction and application for functions, and so on.  It
is a tragic waste if types play only a passive r\^ole in
programming, providing a basis for error reporting. Our technology should
enable programmers to exploit the clues which types provide.

% For example,
%Haskell's \emph{ad hoc}
%polymorphism~\cite{wadler.blott:less.ad.hoc,%
%peytonjones.jones.meijer:type.classes} allows standard components to
%be constructed automatically in a type-directed way---the programmer can
%indicate the general idea and leave the machine to do the tedious `plumbing'.

\subsection{Dependent Types Also Matter Behind the Scenes}

The key innovation of Epigram is its use of types to express
programming tasks and programming patterns. For example, the pattern
of primitive recursion on a datatype is expressed by an \emph{induction
principle}, like
\[
\AR{\forall\vP:\Nat\to\Type\;\TO \\
       \vP\;\DC{0} \To \;\;
       (\forall \vn\Pp:\Nat\TO\;\vP\;\vn\Pp\to \vP\;(\DC{1+}\;\vn\Pp)) \To \\
       \forall \vn:\Nat\TO\;\vP\;\vn}
\]
%
The conventional type of primitive recursion may be shorter, but it is less
informative.
\[
\forall\vP\TO\vP\To(\Nat\to\vP\to\vP)\To\Nat\To\vP
\]
%
This type does not connect the arguments to the tasks they serve,
whereas the induction principle is a dependent type which explains
that a recursive computation for $\vn$ needs a method for $\DC{0}$ and
a method for $(\DC{1+\;\vn\Pp})$.
Correspondingly, the system transforms a programming problem such as
\[
  \RW{let}\;\;
  \Rule{\vx,\vy\hab\Nat}
       {\vx\plus\vy\hab\Nat}
\qquad\qquad\mbox{into a `computability proof' goal}\qquad\qquad
  \orange{?}\hab\forall\vx,\vy\hab\Nat\TO\;\LBTY{\vx\plus\vy:\Nat}
\]
We read $\LBTY{\vx\plus\vy:\Nat}$ as `$\vx\plus\vy$ is computable'.
If we attack this goal with induction, we acquire subgoals like these:
\[\AR{
  \orange{?}\hab\forall\vy\hb\Nat\TO\;\LBTY{\DC{0}\plus\vy:\Nat} \\
  \orange{?}\hab
    \AR{\forall\vx\Pp\hb\Nat\TO \\
        (\forall\vy\hb\Nat\TO\;\LBTY{\vx\Pp\plus\vy:\Nat})\To \\
        \forall\vy\hb\Nat\TO\;\LBTY{(\DC{1+}\;\vx\Pp)\plus\vy:\Nat}
       }
}\]
These goals tell us exactly the `left-hand sides' for the subprograms
which the recursive strategy requires.

By using dependent types to represent programming problems and
programming patterns, we have acquired an interactive programming
environment for the price of an interactive proof system. The basic
constructs of the Epigram language give a `programming' presentation
to the basic tactics of the system. The $\BY$ construct is just
McBride's `elimination with a motive' tactic~\cite{conor:motive} which
synthesizes an appropriate `$\vP$' parameter for any induction-like
rule. The $|\VV{scrutinee}$ construct is just `cut'.
%
The details of the process by which Epigram code is \emph{elaborated}
into the underlying type theory---a variation of Luo's
UTT~\cite{luo:typetheoryforcs}---are given in
\cite{ConorJames:vfl}. Edwin Brady's compiler for UTT,
which erases extraneous information at run time, is presented
in \cite{edwin:thesis}.

The point is this: UTT is not our programming language---UTT is our
language for describing how programming works.  Epigram has no
hard-wired construct for constructor case analysis or
constructor-guarded recursion---these are just programming patterns
specified by type and supplied as standard with every datatype you
define. However, UTT types are Epigram types, so you are free to
extend our language by specifying and implementing your own
patterns. Our typechecking example in \cite{ConorJames:vfl} is a
derived case analysis principle for expressions, exposing not their
syntax, but their types or type errors. By making programming patterns
first-class citizens via dependent types, we raise the level of
abstraction available to programmers and provide interactive support for its
deployment at a single stroke.

\subsection{Programming Interactively}

The interface to Epigram is inspired by the Alf
proof editor~\cite{magnusson.nordstrom:alf}, which introduced
a type-directed structure editor for incomplete programs.
In Epigram's  concrete syntax, a \emph{shed} $\SHEDC{\texttt{raw
text}}$ may stand in for any subexpression. The elaborator is not
permitted inside a \emph{shed}, so the text it contains may be edited
freely.  There are two basic editing moves---removing the brackets to
admit the elaborator (which will process as far as any nested sheds)
and placing brackets around an elaborated subexpression to `undo' it
and return it to a shed. Correspondingly, the full spectrum of
interactivity is supported: an entire program can be written (or, more
to the point, reloaded) inside a shed, then elaborated in `batch
mode'; or a single syntactic construct can be elaborated, with sheds
for subexpressions. The advantages of structure editing are available,
but the disadvantages are not compulsory.

The interactive development of a program is a kind of dialogue.  The
system poses the problems---the left-hand sides of programs.  We
supply the solutions by filling in the right-hand sides, either by
directly giving the program's output $\TO\;\vt$, or by invoking a
programming pattern which reduces the problem to subproblems which are
then posed in turn. There is a direct mapping from sheds
$\SHEDC{\cdots}$ in source code to metavariables, $\orange{?x}$ in the
underlying proof state---when you elaborate a shed, your code triggers
a refinement of the proof state, and the resulting refinement to the
source code is read off from the types of the subgoals. Nothing is
hidden---the proof state is recoverable from the source code simply by
re-elaborating it in a single step.

Our approach to metavariables basically follows McBride's OLEG
system~\cite{conor-phd}---metavariables represent not only the missing
contents of sheds, but also all the unknowns arising from implicit
quantification. The latter are resolved, where possible, by solving
the unification constraints which arise during typechecking---we
follow Dale Miller's `mixed prefix'
approach~\cite{miller:mixed}. Epigram will not guess the type of your
program, but it will infer the bits of your program which the type
determines. The Damas-Milner approach to type
inference~\cite{damas.milner:principal} is alive and well and working
harder than ever, even though we have dispensed with the shackles on
programming which allow it to be complete.

\section{Further Work}
\label{sec:further}

We have hardly started. Exploiting
the expressivity of dependent types in a practicable way involves
a wide range of challenges in the development of the theory, the
design of language, the engineering of tools and the pragmatics of
programming. In this section, we summarize just a few of them.

\textbf{First-Class Modules.}$\quad$ No programming language can succeed
without strong support for the large-scale engineering of
systems. Dependent type systems already allow us to express record
types which pack up data structures, operations over them---and also
proofs of the properties of those operations. \emph{Manifest} record
types, which specify the values of some of their fields, can be used
to express sharing between records, and between the the inputs and
outputs of record-transforming operations~\cite{pollack:records}.
Epigram's first-class notion of programming pattern allows an abstract
datatype to offer admissible notions of pattern matching which hide
the actual data representation but are guaranteed to be faithful to
it---we have Wadler's \emph{views} for free~\cite{wadler:views}.

We now need a practical theory of subtyping to deliver a suitable form
of inheritance, and a convenient high-level syntax for working with
records. Our elaboration mechanism naturally lends itself to the approach
of \emph{coercive subtyping}, where subsumptions in source code elaborate to
explicit coercion functions---typically projections---in the underlying
theory~\cite{luo:coercive}.

\textbf{Universe Polymorphism.}$\quad$ What is the type of types, and how
do we quantify over them safely and consistently? In
\cite{ConorJames:vfl}, we follow the \emph{predicative} fragment of Luo's
Extended Calculus of Constructions~\cite{luo:ecc},
installing a cumulative hierarchy of universes
$\Type_0,\Type_1,\ldots$ each of which both inhabits and embeds in the 
next, so that $\Type_i:\Type_{i+1}$ holds and $\vT:\Type_i$ implies
$\vT:\Type_{i+1}$. As Harper and Pollack have
shown~\cite{harper.pollack:universes}, the user need never
write a universe level explicitly---the machine can maintain a graph
of relative level constraints and protest if any construction induces
a cycle.

This much is certainly safe, and it allows every type to find its own
particular level---this is called \emph{typical ambiguity}. The
trouble is that, as things stand, there is no satisfactory way to
define datatype constructors which operate at \emph{every}
level---this is called \emph{universe polymorphism}. A simple example
shows up if we have a list $\VV{Ts}$ of element types, and we want to
construct the corresponding list of list types
\[
  \FN{map}\;\TC{List}\;\VV{Ts}
\]
The types \emph{in} $\VV{Ts}$ live one level below the type \emph{of}
$\VV{Ts}$, so we need $\TC{List}$ to operate at \emph{both}
levels. There has been some very promising theoretical work in this
area~\cite{paul.zhaohui:coercions.universes,courant02tphols} but
again, a clear and convenient design has yet to emerge. In the
interim, we have adopted the cheap but inconsistent fudge of taking
$\Type:\Type$.

\textbf{Generics and Reflection.}$\quad$ Any function $\vT\hab\vU\to\Type$
naturally \emph{reflects} a sublanguage or \emph{universe} of types---those
given by $\vT\;\vu$. We can think of $\vu$ as the `name' of a type in the
universe.
If $\vU$ happens to be a datatype, we can write generic programs which work
for all the types named by a value $\vu$. Perhaps $\vU$ is the type of
regular expressions and $\vT$ computes for each regular expression the
type of its words, yielding \emph{regular expression
types}~\cite{pierce.vouillon.hosoya:regular.expression.types}.
Perhaps $\vU$ represents a collection of datatypes with a decidable equality
\[
  \FN{eq}\hab\forall\vu:\vU;\,\vx,\vy:\vT\;\vu\;\TO\;
    \vx\EQ\vy\;\blue{\vee}\;\vx\;\blue{\neq}\;\vy
\]
Peter Morris has recently implemented exactly such a generic equality
in Epigram for the universe of regular types~\cite{Morris:bctcs05}.
Dependently typed generic programming is a lively research area
\cite{alti:wcgp02,benke.dybjer.jansson:universes}, inspired by the pioneering
work on generics in
Haskell~\cite{backhouse.jansson.jeuring.meertens:generic,%
              hinze.jeuring:tidts,genHaskell}.

Alternatively, perhaps our universe $\vU$ represents a class of decidable
propositions, equipped with a decision procedure
\[
  \FN{decide}\hab\forall\vu:\vU\;\TO\;
    \vT\;\vu\;\blue{\vee}\;\vT\;\vu\To\blue{\bot}
\]
Such a procedure could be used to extend the elaborator's capacity to solve
simple proof obligations and equational constraints automatically, following
the lead of DML~\cite{DML99}, but without wiring a particular constraint
domain into the language design and the compiler.
However, to make this work conveniently, we need language support for the
declaration of universes $(\vU,\vT)$ where $\vU$ reflects a given class of
types and $\vT$ is invertible by construction, enabling the elaborator to
infer the appropriate $\vu:\vU$ when invoking a generic operation.

\textbf{Observational Type Theory.}$\quad$ The type of equality proofs
used in an intensional theory like Epigram's underlying Type Theory,
is inconvenient when working with infinite objects like function types
or lazy lists (codata). We plan to overcome this restriction without
affecting the decidability of type checking by implementing an
\emph{Observational Type Theory} based on \cite{alti:lics99,mh:phd}.

\textbf{Monadic interfaces to the real world.}$\quad$ A dependently typed
language offers the opportunity to develop the idea of monadic IO further.
Such a monadic interface comes in two guises: a static, denotational semantics
which can be used to reason about the programs and an operational semantics,
which is employed at runtime. This approach is not only relevant for IO,
based on \cite{venanzio-general} we can develop a \emph{partiality monad}
which allows us not only to implement but also reason about genuinely partial
programs like interpreters or programs on the computable reals.

\textbf{Refactoring.}$\quad$ Epigram's
\emph{refinement} style of editing supports the process of working from
types to programs, but it does not help with the inevitable iterations
of the design cycle, as our plans evolve. Indeed, it is quite normal
to build up the index structure for our in layers, as we did for our
$\FN{sort}$ example. Our interactive editing technology should support
these process also, allowing us to experiment at pushing different
type refinements through our programs. We should be able to try out
the options which gave rise to the `open lower bound' choice for
sorted lists.

More generally, we can seek to emulate existing tools
for \emph{refactoring} evolving the design of data structures and
programs~\cite{kent-refactoring}, now in the context of a system which
supports incomplete objects. There is plenty of scope to develop tools
which really reflect the way most programmers work, iteratively improving
program \emph{attempts}---good ideas often come a bit at a time.

\section{Conclusions}
\label{sec:concl}

This much is clear: many programmers are already finding practical
uses for the approximants to dependent types which mainstream
functional languages (especially Haskell) admit, by hook or by crook.
From arrows and functional reactive
programming~\cite{hughes:afp-arrows,henrik:dynamic-optimisation},
through extensible records~\cite{oleg.ralf.keean:hlists},
database programming~\cite{bringert.hockersten:haskellDB}  and
dynamic web scripting~\cite{thiemann:programmabletypes} to code
generation~\cite{baars.swierstra:selfinspecting}, people are `faking it'
any way they can to great effect.
Each little step along the road to dependent types makes the task
a little easier, the code a little neater and the next improvement a
little closer: the arrival of GADTs in Haskell is a joy and a
relief~\cite{peyton-jones-wobbly}.

Now is the time to recognize the direction of these developments and
pursue it by design, not drift. In the long term, there is a great
deal more chaos and confusion to be feared from fumbling through
today's jungle of type class Prolog, singletons and proxies than from
a dependent type system whose core rules have been well studied and
fit on the back of an envelope. The Epigram project is our attempt to
plunder the proof systems based on Type Theory for the technology they
can offer programmers. It is also a platform for radical
experimentation without industrial inertia---an attempt to discover in
which ways dependent types might affect the assumptions upon which
mainstream functional language designs have been predicated. We are
having a lot of fun, and there is plenty of fun left for many more
researchers, but we do not expect to change the mainstream
overnight. What we can hope to do is contribute a resource of
experiments---successful or otherwise---to this design process.

More technically, what we have tried to demonstrate here is that the
distinctions term/type, dynamic/static, explicit/inferred are no longer
naturally aligned to each other in a type system which recognizes the
relationships between values. We have decoupled these dichotomies and
found a language which enables us to explore the continuum of
pragmatism and precision and find new sweet spots within it.
Of course this continuum also contains opportunities for remarkable
ugliness and convolution---one can never legislate against bad design---but
that is no reason to toss away its opportunities. Often, by
bringing out the ideas which lie behind good designs, by expressing
the things which matter, dependent types make data and programs fit better.

\bibliographystyle{alpha}
\bibliography{handy,obsdec,alti,local}

\end{document}

### 2010-04-24 papers/containers-for-dummies/containers.tex
Containers for Dummies: first import, just infrastructure.

\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{verbatim}
\usepackage{url}
\usepackage{xspace}
\usepackage{ifthen}

\newcommand{\DRAFT}{}

%% Structure
\ifthenelse{\isundefined{\DRAFT}}
           {\newenvironment{structure}
                           {\comment}
                           {\endcomment}}
           {\newenvironment{structure}
                           {\footnotesize\verbatim}
                           {\endverbatim}}

%% Written bits of Structure
\newenvironment{wstructure}{\comment}{\endcomment}

%% Comments
\ifthenelse{\isundefined{\DRAFT}}
           {\newcommand{\note}[1]{}}
           {\setlength{\marginparwidth}{0.7in}
            \newcommand{\note}[1]{\-\marginpar[\raggedright\footnotesize #1]%
                                 {\raggedright\footnotesize #1}}}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Draft: Containers for Dummies}
\author{The MSP Doctoral Crew
        \footnote{Correspondent author: 
                  \texttt{dagand@cis.strath.ac.uk}}}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Content
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
The idea of writing this document arised during a MSP meeting, when I
(Pierre) was begging for some explanation of Michael Abbott's PhD
thesis~\cite{abbott:phd-thesis}. Neil answered that I should develop
my own understanding of the Container story instead. Hank acknowledged
that resources on Containers were rather technical and going in many
directions, mostly because each member of the Container Mafia have his
personal take on it.

It was suggested to write a document that would serve as an
introduction as well as a reference on Containers. Being an
introduction, this document should allow newcomers to become familiar
with our technology, thanks to a plethora of examples and
exercices. Being a reference, it should be an exhaustive overview of
the Container landscape, with clear definitions and theorems.
\end{abstract}

\begin{structure}
Some braindump:

* I have several pages of definitions from Hank. That could make a
  first skeleton.

* As suggested by Hank, we should first focus on the table of content,
  the content will follow.

* We will inevitably end up talking to the Programmer, the Type
Theorist and the Category Theorist. That would be good to organize the
document to allow people to build on their intuition and then move on
what they are less familiar with. We could (graphically) signal when
some paragraph goes into technicalities of a particular field.
\end{structure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{abbrvnat}
\bibliography{containers}

\end{document}

### 2010-06-30 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: Fix typos.

@@
 $\blue{\Sigma}_{\x : \M{A}} (\M{B}\: \x)$ stresses that
-$\Sigma$-types are `dependent sums', generalising of sums over
+$\Sigma$-types are `dependent sums', generalising sums over
 arbitrary arities, where simply typed languages have finite sums.

### 2010-06-30 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: Fix typos.

@@
 but the offensive recursion is concealed by a prestidigitation.
-Expanding be \(\toDesc{-}\) and propagating types as in
+Expanding \(\toDesc{-}\) and propagating types as in
 Figure~\ref{fig:type-checking} reveals the
@@
 \qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                      \\
-\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d'}}{}\: \pair{\V{g}}{\V{d'}}{} \mapsto   \\
+\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}\: \pair{\V{g}}{\V{d'}}{} \mapsto   \\
 \qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
@@
 In this section, we will turn to a more ambitious generic operation on
-datatype. Given a functor, represented as a tagged description, we
+datatypes. Given a functor, represented as a tagged description, we
 build the free monad over this functor.
@@
 plays the r\^ole of \return, embedding variables into terms. The
-\bind\ operation corresponds to \emph{substitution}. We will now
+\bind operation corresponds to \emph{substitution}. We will now
 implement it, as a generic function.
@@
 Our implementation will appeal to the $\cata$ function developed
-previously. So, let us write down the types, and fill as much
+previously. So, let us write down the types, and fill as many
 arguments to $\cata$ as possible:
@@
            \PI{\V{X}, \V{Y}}{\Set} 
-           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})}) \To \\
+           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To \\
          & \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
@@
 to extend it with a notion of variable. Then, we have equipped this
-structure with the corresponding monadic operation, \bind\ and
+structure with the corresponding monadic operations, \bind and
 \return. This construction is an example of type-indexed

### 2010-06-30 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: Fix typos.

@@
 We have checked the construction using Agda's universe polymorphism,
-coding $\SYMBIDesc$ in itself and have proving the isomorphism
+coding $\SYMBIDesc$ in itself and have proved the isomorphism
 between the host and the embedded universes.
@@
 
-Our approach is follow the polytypic programming style, as initiated
+Our approach follows the polytypic programming style, as initiated
 by PolyP~\cite{jansson:polyp}. Indeed, we build generic functions by
@@
 approach~\cite{hinze:generic-haskell}, as exemplified by the free
-monad construction: from datatype, we can compute new datatypes and
+monad construction: from a datatype, we can compute new datatypes and
 equip them with their structure. Generic Haskell also features
@@
 operations. As a result, SYB is not reflexive: it is restricted to
-datatypes instanciating $\CN{Typeable}$. Moreover, it is limited to
+datatypes instantiating $\CN{Typeable}$. Moreover, it is limited to
 building generic functions, hence type-indexed datatypes cannot be
@@
 
-However, we differ in several ways. First, there approach is
+However, we differ in several ways. First, their approach is
 generative: each universe extends the base type theory with both type

### 2010-06-30 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix error in AllI/allI definition
Desc@ICFP: Minor bugs in the code Vec and AllI
Desc@ICFP: Fix typos.

@@
 \AllI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{k}} &
-    \DConst{\V{K}} \\
+    \DConst{\Unit} \\
 \AllI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\pair{\V{d}}{\V{d'}}{}} &
@@
 \allI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{P}}{\V{p}}{\V{k}} &
-    \V{k} \\
+    \void \\
 \allI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{d}}{\V{d'}}{}} &
@@
 \allI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\V{f}} &
-    \allI{}{(\V{D}\: \V{a})}{\V{X}}{\V{P}}{\V{p}}{(\V{f}\: \V{a})}
+    \LAM{\V{a}}\allI{}{(\V{D}\: \V{a})}{\V{X}}{\V{P}}{\V{p}}{(\V{f}\: \V{a})}
 \end{array}
@@
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \VecNil          & \D{Vec}\:\V{X}\:{\NatZero)}   \\
+    \VecNil          & \D{Vec}\:\V{X}\:{\NatZero}   \\
     \SYMBVecCons & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
@@
                             \DConst{(\V{n} \PropEq \NatZero)} \\
-                            \DSigma{\Nat}{\LAM{\V{m}} \stk{\DProd{\DVar{\V{m}}}{\\\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}}
+                            \DSigma{\Nat}{\LAM{\V{m}}
+\stk{\DProd{\DConst{\V{X}}}{\DProd{\DVar{\V{m}}}{\\\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}}}
                           \end{array}
@@
 
-We should precise that forcing a description is not guaranteed to
-remove all constraints. It is subject to future work to see if
-constraints can be entirely eradicated, or presented more conveniently
-to the developer.  Finally, it is worth mentioning that these
-optimisations are \emph{source-to-source} transformations on
-descriptions.
+Note that forcing a description is not guaranteed to remove all
+constraints. It is subject to future work to see if constraints can be
+entirely eradicated, or presented more conveniently to the developer.
+Finally, it is worth mentioning that these optimisations are
+\emph{source-to-source} transformations on descriptions.
 
@@
 that do. The non-dependent part mirrors the definition for non-indexed
-descriptions. The index-depend part simply indexes the choice of
+descriptions. The index-dependent part simply indexes the choice of
 constructors by $\V{I}$. Hence, by inspecting the index, it is
@@
 and boolean expressions. We have given its semantics through an
-evaluation function. Provided a one step semantic of the language, the
+evaluation function. Provided a one step semantics of the language, the
 big step interpreter is granted without effort thanks to the generic
@@
 our datatype enjoys more structure than what we are given. We are
-facing a free monad, where $\SYMBEval$ is the \return\ introducing
+facing a free monad, where $\SYMBEval$ is the \return introducing
 variables. For convenience, we wrap $\discharge$ in a $\V{\sigma}$
@@
 Just as in the universe of descriptions, this construction comes with
-an obvious \return\ and a substitution operation, the \bind. Its
+an obvious \return and a substitution operation, the \bind. Its
 definition is the following:
@@
 \substI : & _{\PI{\V{I}}{\Set}}
-            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
-            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
+            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
+            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
           & (\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 
                                \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}) \To \\
-          & \PITEL{\V{i}}{\V{I}}
-            \PITEL{\V{D}}{\IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{X}}})}{\V{i}}} \To
+          & \PI{\V{i}}{\V{I}}
+            \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{X}}})}{\V{i}} \To
             \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}
@@
 
-Let us now consider two examples of free indexed monad.
+Let us now consider two examples of free indexed monads.
 
@@
 structure. We had identified the variable substitution operation as
-the \bind\ of a free monad. To exhibit its monadic structure, we first
+the \bind of a free monad. To exhibit its monadic structure, we first
 have to massage the definition of our datatype.
 
-As previously mentioned, we identify $\SYMBEval$ with the \return\ of
-the free monad, while the other components are the action of the
+As previously mentioned, we identify $\SYMBEval$ with the \return of
+the free monad, while the other components are the actions of the
 monad. As a result, the definition is similar to $\HExprD$ presented
@@
 open terms. Using its monadic structure, we have implemented the
-interpreter for open terms in context. Hence, without much efforts, we
+interpreter for open terms in context. Hence, without much effort, we
 have described the syntax of a well-typed language, together with its

### 2010-06-30 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix FreeMonad pseudo-code.

@@
 
-Let us recall the free monad construction. Given a functor $F$, the
-free monad over $F$ is defined by the following datatype:
+Let us recall the free monad construction in, say, Haskell. Given a
+functor \texttt{f}, the free monad over \texttt{f} is defined by the
+following datatype:
 %
-\[
-\stk{
-\data \FMFreeMonad\: \PITEL{\V{F}}{\Set \To \Set} 
-                     \PITEL{\V{X}}{\Set} :
-                     \Set 
-\where \\
-\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-    \FMVar           & \V{X} \To \FMFreeMonad\: \V{F}\: \V{X}                            \\
-    \FMComposite     & \V{F} (\FMFreeMonad\: \V{F}\: \V{X}) \To \FMFreeMonad\: \V{F}\: \V{X}    
-\end{array}
-}
-\]
+\begin{code}
+data FreeMonad f x
+    = Var x
+    | Composite (f (FreeMonad f x))
+\end{code}
 %

### 2010-07-05 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix Hutton's example
Instead of fiddling with two separate finite domains for each types,
we really define a notion of context, environment, and variable
pointing into that.

Addressing the question of a reviewer, the model contains the case of
pairs, mixing any odd type.

A consequence is that we talk about equality. But that should be ok.

@@
 %% Types
-\newcommand{\Ty}{\C{Ty}}
+\newcommand{\Ty}{\D{Ty}}
 \newcommand{\Ebool}{\etag{\CN{bool}}}
@@
 \newcommand{\SYMBVar}{\F{Var}\xspace}
-\newcommand{\Var}[2]{\SYMBVar_{#1} #2}
+\newcommand{\Var}[2]{\SYMBVar\: #1\: #2}
 
@@
 
+\newcommand{\Ctxt}{\D{Context}}
+\newcommand{\SYMBCtxtEmpty}{\C{[]}\xspace}
+\newcommand{\CtxtEmpty}{\SYMBCtxtEmpty}
+\newcommand{\SYMBCtxtSnoc}{\C{snoc}\xspace}
+\newcommand{\CtxtSnoc}[2]{\SYMBCtxtSnoc\:#1\:#2}
+\newcommand{\SYMBEnv}{\F{Env}}
+\newcommand{\Env}[1]{\SYMBEnv\: #1}
+\newcommand{\SYMBlookup}{\F{lookup}}
+\newcommand{\lookup}[4]{\SYMBlookup\: #1\: #2\: #3\: #4}
+
 However, so far, we are only able to define and manipulate
 \emph{closed} terms. By abstracting over $\SYMBVal$, it is possible to
-build and manipulate \emph{open} terms, that is, terms with
-variables. Following $\SYMBVal$, we define $\SYMBVar$ by:
-%
-\[\stk{
-\Var{}{} : \EnumU \To \Ty \To \Set \\
-\Var{\V{dom}}{\_} = \EnumT{\V{dom}}
-}\]
-%
-Whereas $\SYMBVal$ was mapping the type to the corresponding host type,
-$\SYMBVar$ maps types to a finite set. The finite set -- the context
--- contains closed terms. A variable is therefore a $\SYMBEval$ that
-contains a pointer to a particular element of the finite set -- an
-element of $\EnumT{\V{dom}}$.
+build and manipulate \emph{open} terms, that is, terms in a
+context. The context is represented by a snoc-list of types:
+%
+\[
+\stk{
+\data \Ctxt : \Set \where \\
+\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
+    \SYMBCtxtEmpty  & \Ctxt   \\
+    \SYMBCtxtSnoc   & \Ctxt \To \Ty \To \Ctxt
+\end{array}
+}\]
+%
+An environment realises the context, associating a value for each type:
+%
+\[
+\stk{
+\SYMBEnv : \Ctxt \To \Set \\
+\begin{array}{@{}l@{\:\:=\:\:}l}
+\Env{\CtxtEmpty}        & \Unit \\
+\Env{(\CtxtSnoc{G}{S})} & \TIMES{\Env{G}}{\Val{S}}
+\end{array}
+}\]
+%
+In this setting, we define $\SYMBVar$ by:
+%
+\[\stk{
+\Var{}{} : \Ctxt \To \Ty \To \Set \\
+\begin{array}{@{}ll@{\:\:=\:\:}l}
+\Var{\CtxtEmpty}{& T}            & \Void \\
+\Var{(\CtxtSnoc{G}{S})}{& T}     & \SUM{(\Var{G}{T})}{(S \Eq T)}
+\end{array}
+}\]
+%
+While $\SYMBVal$ was mapping the type to the corresponding host type,
+$\SYMBVar$ indexes a value in the context, obtaining a proof that the
+types matches. The $\SYMBlookup$ function precisely follow this
+semantics:
+%
+\[\stk{
+\SYMBlookup : \PI{\V{G}}{\Ctxt} 
+          \Env{\V{G}} \To 
+          \PI{\V{T}}{\Ty} 
+          \Var{\V{G}}{\V{T}} \To
+          \Val{\V{T}} \\
+\begin{array}{@{}lll}
+\lookup{\CtxtEmpty}{               \_}{                   \V{T}}{()}                    &   & \\
+\lookup{(\CtxtSnoc{\V{G}}{.T})}{   \pair{\V{g}}{\V{t}}{}}{\V{T}}{(\SumRight{\C{refl}})} & = & \V{t} \\
+\lookup{(\CtxtSnoc{\V{G}}{\V{S}})}{\pair{\V{g}}{\V{t}}{}}{\V{T}}{(\SumLeft{\V{x}})}     & = & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
+\end{array}
+}\]
 
 Consequently, replacing $\Val{\V{ty}}$ by
-$(\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})$ in
+$(\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}})$ in
 Figure~\ref{fig:hexpr-full} turns the language of closed terms into a
@@
 readability, we abbreviate $\LAM{\V{ty}}
-\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ by
-$\SUM{\SYMBVal}{\Var{\V{dom}}{}}$. \note{Is it polite and
+\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}$ by
+$\SUM{\SYMBVal}{\Var{\V{G}}{}}$. \note{Is it polite and
   comprehensible to ask that?} This defines a new indexed description,
-called $\HExprVarD{\V{dom}}$.
+called $\HExprVarD{\V{G}}$.
 
@@
 \begin{array}{@{}ll}
-\discharge : & \PITEL{\V{ty}}{\Ty}
-               \PITEL{\V{dom}}{\EnumU} \\
-             & \PI{\V{\gamma}}{\spi{\V{dom}}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})}} \\
-             & (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To
-               \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}} 
+\discharge : & \PI{\V{G}}{\Ctxt}
+               \Env{G} \To
+               \PITEL{\V{ty}}{\Ty} \\
+             & (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
+               \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} 
 \end{array} \\
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumLeft \V{x})  & \Con{\pair{\SYMBEval}{\V{x}}{}} \\
-\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumRight \V{v}) &
-\end{array}\\
-\qquad\switch{\V{dom}}{(\LAM{\_}
-\IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})}{\V{\gamma}}{\V{v}}
+\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumLeft \V{x})  & 
+    \Con{\pair{\SYMBEval}{\V{x}}{}} \\
+\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumRight \V{v}) &
+    \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
+\end{array}
 }\]
@@
 \newcommand{\substH}{\F{substExpr}}
-\newcommand{\domNat}{dom_{\CN{nat}}}
-\newcommand{\domBool}{dom_{\CN{bool}}}
-\newcommand{\gammaNat}{\V{\ensuremath{\gamma_{\CN{nat}}}}}
-\newcommand{\gammaBool}{\V{\ensuremath{\gamma_{\CN{bool}}}}}
-\newcommand{\SYMBGammaSpi}{\F{\ensuremath{\Gamma_{\CN{ty}}}}\xspace}
-\newcommand{\GammaSpi}[2]{\SYMBGammaSpi\: #1\: #2}
 
@@
 \begin{array}{@{}ll}
-\substH  : & \PITEL{\V{dom}}{\EnumU} \\
-           & \PITEL{\V{\gammaNat}}{\GammaSpi{\V{dom}}{\Enat}} 
-             \PITEL{\V{\gammaBool}}{\GammaSpi{\V{dom}}{\Ebool}} \\
-           & \begin{array}{@{}ll}
-             \PI{\V{\sigma}}{& \PITEL{\V{dom}}{\EnumU} \\
-                             & \PITEL{\V{\gammaNat}}
-                                     {\GammaSpi{\V{dom}}{\Enat}} \\
-                             & \PITEL{\V{\gammaBool}}
-                                     {\GammaSpi{\V{dom}}{\Ebool}} \\
-                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To \\
-                             & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}}
-             \end{array}\\
-          & \PI{\V{ty}}{\Ty}
-            \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}} \To \\
-          & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
+\substH  : & \PITEL{\V{ty}}{\Ty}
+             \PI{\V{G}}{\Ctxt} \\
+           & \PITEL{\V{\sigma}}{\PI{\V{ty}}{\Ty}
+                                (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
+                                \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}} \\
+          & \To \IMu{\Ty}{\toIDesc{\HExprVarD{\V{G}}}}{\V{ty}} \To 
+            \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
 \end{array}
 \]
-%
-Where $\SYMBGammaSpi$ corresponds to a context, defined by:
-%
-\[\stk{
-\SYMBGammaSpi : \EnumU \To \Ty \To \Set                                             \\
-\GammaSpi{\V{dom}}{\V{ty}} = \spi{\V{dom}}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})} 
-}\]
 
@@
 facing a free monad, where $\SYMBEval$ is the \return introducing
-variables. For convenience, we wrap $\discharge$ in a $\V{\sigma}$
-function that picks the context of the right type:
-%
-\[
-\F{$\sigma$}\: \V{dom}\: \gammaNat\: \gammaBool\: \V{ty}\: \V{var} \mapsto
-    \F{$\discharge$}\: \V{ty}\: \V{dom}\: \F{$\gamma$}_{\V{ty}}\: \V{var} 
-\]
-%
-Where $\F{$\gamma$}_{\V{ty}}$ is short for
- $\caseB{\V{ty}}{\Enat  \To \gammaNat \\
-                 \Ebool \To \gammaBool}$
-
-Instead of implementing $\substH$ in this special case, we are now
-going to implement the \emph{free indexed-monad construction}.
+variables. Instead of implementing $\substH$ in this special case, we
+are now going to implement the \emph{free indexed-monad construction}.
 
@@
 In Section~\ref{sec:idesc-examples}, we had the intuition that our
-datatypes $\HExprD$ and $\HExprVarD{\V{dom}}$ enjoy a monadic
+datatypes $\HExprD$ and $\HExprVarD{\V{G}}$ enjoy a monadic
 structure. We had identified the variable substitution operation as
@@
 corresponds to the syntax of closed terms, $\HExprD$. Similarly,
-$\FreeIMonad{\HExprFreeD}{\Ty}{(\SUM{\SYMBVal}{\Var{\V{dom}}{}})}$
-corresponds to expressions with variables, $\HExprVarD{\V{dom}}$.
+$\FreeIMonad{\HExprFreeD}{\Ty}{(\SUM{\SYMBVal}{\Var{\V{G}}{}})}$
+corresponds to expressions with variables, $\HExprVarD{\V{G}}$.
 
@@
 \[\stk{
-\substH\: \V{dom}\:
-          \gammaNat\: \gammaBool\:
+\substH\: \V{ty}\:
+          \V{G}\:
+          \V{g}\:
+          \V{\sigma}\: 
+          \V{term} \mapsto  \\
+\quad \substI_{\Ty}\:
+               (\SUM{\SYMBVal}{\SYMBVar\: \V{G}})\: 
+               \SYMBVal\:
+               \HExprFreeD\: 
                \V{\sigma}\:
                \V{ty}\:
-          \V{term} \mapsto  \\
-\;\;\ \begin{array}{ll}
-       \substI_{\Ty}\: &  \HExprFreeD\: 
-                         (\SUM{\SYMBVal}{\Var{\V{dom}}{}})\: 
-                         \SYMBVal \\
-                      &
-                        (\V{\sigma}\: \V{dom}\: \gammaNat\: \gammaBool)\:
-                        \V{ty}\:
                \V{term}
-      \end{array}
 }\]

### 2010-07-05 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: motivate role of switchD
Reviewer 2 says:

"Also, in the implementation, I did not see anything resembling
"switchD", which makes me believe that the implementation does not
tell us anything about the soundness of the full system with "switchD"
in which levitation is possible."

To which we answered:

"Referee 23B asks how much is definable from a "vanilla" system and
advises us to bring our models into more light. As mentioned above,
Coq is problematic, but in Agda the news is rather better. We used
universe polymorphism to hardwire the Desc type at all levels, then
define the asso

@@
 As a consequence, our definition above now propagates without
-introducing recursion. Of course, by pasting together
-the declaration of \(\Desc\) and its internal copy, we have made
-it appear in its own type. Hardwired as a \emph{fait accompli},
-this creates no regress, although one must assume the definition
-to recheck it.
+introducing recursion. Of course, by pasting together the declaration
+of \(\Desc\) and its internal copy, we have made it appear in its own
+type. Hardwired as a \emph{fait accompli}, this creates no regress,
+although one must assume the definition to recheck it.\footnote{Our
+  Agda model does not formalize the $\SYMBswitchD$
+  construction. However, it proves the isomorphism between the defined
+  descriptions and the coded ones. The role of $\SYMBswitchD$ is to
+  collapse this isomorphism, hence identifiying defined and coded
+  descriptions.}
 

### 2010-07-06 papers/stratisfaction/paper.tex
Stratisfaction: massive load of CUM
Sorry, sorry, couldn't resist.

%% * Header

\documentclass[preprint
              , authoryear
              , onecolumn
              ]{sigplanconf}

%% ** Packages

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{verbatim}
\usepackage{url}
\usepackage{xspace}
\usepackage{subfigure}

\usepackage{pig}

\include{Macros}

%% ** Commands

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\ie}{\emph{ie.}\xspace}

\newcommand{\redb}[1]{\fcolorbox{red}{white}{$#1$}}
\newcommand{\blueb}[1]{\fcolorbox{blue}{white}{$#1$}}
\newcommand{\greenb}[3]{\fcolorbox{green}{white}{$#1$}_{\textcolor{green}{#2}}^{\textcolor{green}{#3}}}
\newcommand{\blackb}[1]{\fcolorbox{black}{white}{$#1$}}

\newcommand{\ruleName}[2]{(\mathrm{\textsc{#1}}^{#2})}
\newcommand{\ruleSet}[1]{\ruleName{Set}{#1}}
\newcommand{\rulePi}[1]{\ruleName{Pi}{#1}}
\newcommand{\ruleLam}[1]{\ruleName{Lam}{#1}}
\newcommand{\ruleConv}[1]{\ruleName{Conv}{#1}}
\newcommand{\ruleVar}[1]{\ruleName{Var}{#1}}
\newcommand{\ruleAsc}[1]{\ruleName{Asc}{#1}}
\newcommand{\ruleApp}[1]{\ruleName{App}{#1}}
\newcommand{\ruleSetEq}[1]{\ruleName{Set-Eq}{#1}}
\newcommand{\ruleVarEq}[1]{\ruleName{Var-Eq}{#1}}
\newcommand{\ruleSetKEq}[1]{\ruleName{Set-K-Eq}{#1}}
\newcommand{\ruleVarKEq}[1]{\ruleName{Var-K-Eq}{#1}}
\newcommand{\ruleSetVEq}[1]{\ruleName{Set-Var-Eq}{#1}}
\newcommand{\ruleVarVEq}[1]{\ruleName{Var-Var-Eq}{#1}}
\newcommand{\rulePiEq}[1]{\ruleName{Pi-Eq}{#1}}
\newcommand{\ruleLamEq}[1]{\ruleName{Lam-Eq}{#1}}
\newcommand{\ruleAppEq}[1]{\ruleName{App-Eq}{#1}}

\newcommand{\N}{\mathbb{N}}

\newcommand{\sublift}[3]{#2 \sqsubseteq_{#1} #3}

\begin{document}

%% ** Conference info

%\conferenceinfo{WXYZ '05}{date, City.} 
%\copyrightyear{2005} 
%\copyrightdata{[to be supplied]}

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

%% * Title

\title{Stratisfaction}

%% Alphabetical ordering.
\authorinfo{Pierre-\'{E}variste Dagand}
           {University of Strathclyde}
           {dagand@cis.strath.ac.uk}
\authorinfo{Conor McBride}
           {University of Strathclyde}
           {conor@cis.strath.ac.uk}

\maketitle

\ColourEpigram
%\MonochromeEpigram

%% * Abstract

\begin{abstract}
\end{abstract}

%\category{CR-number}{subcategory}{third-level}

%\terms
%term1, term2

%\keywords
%keyword1, keyword2

%% * Introduction
\section{Introduction}

%% ** Problem
%% *** <- Often, definitions are set polymorphic
%% **** <- Nat
%% **** <- Sigma type
%% **** <- Others?
%% *** -> Leads to code duplication
%% **** <- Define these objects at each level

%% ** Related work
%% *** <- Harper-Pollack: constraints too difficult
%% **** -> Lego, Coq, Agda experiments
%% ***** -> Leads to more or less powerful systems
%% **** -> Coq partial implementation, currently extended
%% **** -> Agda set polymorphism

%% *** -> Agda standard libray
%% *** -> Formalizations of CT, Voevoedsky

%% * Stratified type-theory
\section{A Universe Polymorphic Type Theory}

%% ** <- Presentation

\newcommand{\ECC}{ECCE\xspace}

%% *** <- Similar to ECC (FIND REF)
%% *** <- With CONV and extended CUML rule
%% **** -> Luo: subtyping: x :<:(N) A -> B ==> x :<:(O) A -> C with B =< C
%% **** -> Us: polymorphism: x :<:(N) A -> B ==> x :<:(O) Up A -> Up B

%% ***** Lifting operator

\begin{definition}[Lifting operator]

The lifting operator, denoted $\SYMBUp$, is defined by:
%%
$$
\begin{array}{l@{\:\:\mapsto\:\:}l}
\Up{k}{\Set{i}}                & \Set{i+k} \\
\Up{k}{\PI{\V{x}}{S}{T}}       & \PI{\V{x}}{\Up{k}{S}}{\Up{k}{T}} \\
\Up{k}{x_i}                    & x_{i+k} \\
\Up{k}{\LAM{x}{b}}             & \LAM{x}{\Up{k}{b}} \\
\Up{k}{f\: s}                  & (\Up{k}{f})\: s
\end{array}
$$

\note{$\Up{k}{(\Bhab{t}{T})}$ is embarrassing. In practice, lifts are
      applied on values, therefore ascriptions have disappeared.}

\end{definition}

%% ***** Lemma: lifting and substitution

\begin{lemma}[Lifting and substitution]
\label{lemma:up-bound-var}

Lifting commutes with substitution:
%%
$$
\Up{}{(T[s/x])} = (\Up{}{T})[s/x]
$$
%%
And with computation: for a term $a$ and $n$ its $\beta$-normal form,
\ie $a \leadsto n$, we have:
%%
$$
\Up{}{a} \leadsto \Up{}{n}
$$

\end{lemma}

Hence, we do not need to precise parenthesing: $\Up{}{(T[s/x])} =
(\Up{}{T})[s/x] = \Up{}{T}[s/x]$.

%% ****** Proof

\begin{proof}

The proof can be decomposed in two cases. First, if $x$ does not
appear in $T$, $T[s/x] = T$ and the proof is trivial. Second, without
loss of generality, if $x$ appears once in $T$. Being a variable, it
carries a number $j$ of pending liftings. There are two subsequent
cases: $x$ is at the head of a function application (possibly with an
empty spine), or it is in a spine of argument. In the latter case, by
definition of $\SYMBUp$, the lifting has no effect on $x$ therefore
both terms are equal.

In the former, we have to analyse the effect of lifting and
substitution on both side. On the right side, $x_j$ is updated to
$x_{j+k}$ when first computing $\Up{k}{T}$. When substitution happens,
the liftings are discharged on $s$: the variable $x$ is replaced by
$\Up{j+k}{s}$. We obtain the term $\Up{k}{T}$ where $x$ is replaced by
$\Up{j+k}{s}$. On the left side, we first compute $T[s/x]$: $x$ is
replaced by $\Up{j}{s}$ in $T$. Then, the lifting occurs on the very
same term at the exception of $\Up{j}{s}$ replacing $x_j$. Hence,
$\Up{k}{\cdotp}$ happens on $\Up{j}{s}$. We obtain the term
$\Up{k}{T}$ where $x$ is replaced by $\Up{j+k}{s}$, concluding the
proof.

\note{Do we consider that computation happens right after (during)
      substitution? If so, we are in trouble.}

\end{proof}

%% ***** Context lifting

\begin{definition}[Context lifting]

Let $\Gamma$ a valid context.

We define valid \emph{context liftings} by the following sequents:
%%
$$
\CAxiom{\sublift{i}{\epsilon}{\epsilon}}
\qquad
\RuleSide{\sublift{i}{\Gamma}{\Delta}}
         {j \leq i}
         {\sublift{i}{\Gamma, \Bhab{x}{A}}{\Delta, \Bhab{x}{\Up{j}{A}}}}
$$

\end{definition}

\note{There is an elephant in the room here: what kind of result do we
  have about this polymorphic type-system? Indeed, it's similar to
  Luo's subtyping presentation of ECC. But it cannot be translated to
  it (how do you translate our cumulativity rule?). Doing the full
  normalization proof (Tait and co.) is not super exciting.}

%% *** <- Figure: type theory

\begin{figure}[htb]

\begin{center}
\input{figure_ecce}
\end{center}

\caption{\ECC (Specification)}

\end{figure}

%% ** <- Full-reflection
%% *** <- Reflect equality as well as objects
%% **** <- Stratif. should be as unobtrusive as possible
%% ***** -> Dealt as much as possible definitionally
%% *** -> Early Martin-Lof papers
%% *** -> Palmgren
%% ** <- Meta-theoretical properties
%% *** ???
%% ** /> Just a spec
%% *** <- Flow of information unclear
%% *** <- CUML non algorithmic

%% * Bidirectionalized
\section{Bidirectional Presentation}

%% ** <- Push/pull distinction
\subsection{Introduction to bidirectionality}

%% *** -> Explicit flow of information
%% *** -> Algorithmic type-checking/synthesis
%% *** <- Push
%% **** -> Calculus of constants
%% *** <- Pull
%% **** -> Calculus of functions
%% *** <- Conversion
%% **** -> Calculus of conversion
%% *** <- Figure: bidirectional presentation (excepted CUML)

\begin{figure}[p]

\begin{center}
\input{figure_bidir}
\end{center}

\caption{Bidirectional presentation (excepted CUML)}

\end{figure}

%% *** <- Proof
%% **** Equational theory remains unchanged
%% ***** <- Same rules
%% **** Sound
%% ***** [OK] \Gamma \vdash A \ni^1 a \Rightarrow \Gamma \vdash a : A
%% ***** [OK] \Gamma \vdash a \in^1 A \Rightarrow \Gamma \vdash a : A
%% ***** Theorem

\begin{theorem}[Soundness]

Let $\Gamma$ a valid context.

\begin{itemize}
\item If \xspace $\Gamma \vdash \pushN{a}{1}{A}$, then $\Gamma \vdash \Bhab{a}{A}$
\item If \xspace $\Gamma \vdash \pullN{a}{1}{A}$, then $\Gamma \vdash \Bhab{a}{A}$
\end{itemize}

\end{theorem}

%% ***** Proof

\begin{proof}

The proof is by induction on the length of a derivation. The base
cases are $\ruleSet{1}$ and $\ruleVar{1}$ that translates (resp.) to
$\ruleSet{0}$ and $\ruleVar{0}$. We then proceed structurally for the
other rules: we simply forget the flow of information, turning $\in$
and $\ni$ into $:$. Because the equational theory is the same, the
$\ruleConv{1}$ rule translates directly.

\end{proof}

%% **** Complete
%% ***** [OK] \Gamma \vdash a : A \Rightarrow \Gamma \vdash A \ni^1 a
%% ***** Theorem

\begin{theorem}[Completeness]

Let $\Gamma$ a valid context. If \xspace$\Gamma \vdash \Bhab{a}{A}$, then
$\Gamma \vdash \push{a}{A}$.

\end{theorem}

%% ****** Proof

\begin{proof}

The proof is by induction on the length of a derivation. The base
cases are $\ruleSet{0}$ and $\ruleVar{0}$. The rule $\ruleSet{0}$
translates directly to its counterpart $\ruleSet{1}$. In the case of
the $\ruleVar{}$ rule, we use $\ruleVar{1}$ to get $\pull{x}{A}$,
followed by the conversion rule $\ruleConv{1}$ for $T = A$, asking for
$\equalN{A}{A}{1}{\SetTop}$ to hold (trivial).

The induction step is again direct for rules that have a counterpart
on the push-side. For the one on the pull-side, we proceed as for
$\ruleVar{0}$-$\ruleVar{1}$: we obtain a pull-oriented rule that we
trivially convert to its push-side equivalent.

\end{proof}

%% ** <- Fusing CUML push-side
\subsection{Fusing cumulativity on the push-side}

%% *** <- Types pushed in
%% **** -> No guess work needed
%% *** <- Figure: CUML fused on push-side

\begin{figure}

\begin{center}
\input{figure_infuse}

\subfigure[Algorithm: pull-side]{
$$
\stkc{
%% Form
\boxed{\Gamma \vdash \pullN{\CN{term}}{2}{\CN{type}}}
\\
\\
%% Variable extraction
\AxiomSide{\ruleName{Var}{2}}{\Bhab{x}{A} \vdash \pullN{x}{2}{A}}
\qquad
%% Type ascription
\RuleSide{\pushN{T}{2}{\SetTop} \quad
          \pushN{t}{2}{T}}
         {\ruleName{Asc}{2}}
         {\pullN{(\Bhab{t}{T})}{2}{T}}
\\
%% Function application
\RuleSide{\pullN{f}{2}{\PI{\V{x}}{S}{T}} \quad
          \pushN{s}{2}{A}}
         {\ruleName{App}{2}}
         {\pullN{f\: s}{2}{T[s/x]}}
}
$$
\label{fig:ccomega-algo-pull}
}
\subfigure[Algorithm: conversion]{
$$
\stkc{
%% Conversion rule
\RuleSide{\pullN{e}{2}{S} \quad
          \equalN{S}{T}{2}{\SetTop}}
         {\ruleName{Conv}{2}}
         {\pushN{e}{2}{T}}
\\
\\
%% Form
\boxed{\Gamma \vdash \equalN{\CN{term}}{\CN{term}}{2}{\CN{type}}}
\\
\\
%% Set equality
\AxiomSide{\ruleName{Set-Eq}{2}}{\equalN{\Set{i}}{\Set{i}}{2}{\SetTop}}
\\
%% Variable equality
\AxiomSide{\ruleName{Var-Eq}{2}}{\Bhab{x}{S} \vdash \equalN{x}{x}{2}{S}}
\\
%% Pi equality
\RuleSide{\equalN{S_1}{S_2}{2}{\SetTop} \quad
          \Bhab{x}{S_2} \vdash \equalN{T_1}{T_2}{2}{\SetTop}}
         {\ruleName{Pi-Eq}{2}}
         {\equalN{\PI{\V{x}}{S_1}{T_1}} 
                 {\PI{\V{x}}{S_2}{T_2}}{2}{\SetTop}}
\\
%% Function equality
\RuleSide{\Bhab{x}{S} \vdash \equalN{f\: x}{g\: x}{2}{T}}
         {\ruleName{Lam-Eq}{2}}
         {\equalN{f}{g}{2}{\PI{\V{x}}{S}{T}}}
\\
%% Application equality
\RuleSide{\equalN{f}{g}{2}{\PI{\V{x}}{S}{T}} \quad
          \equalN{a}{b}{2}{S}}
         {\ruleName{App-Eq}{2}}
         {\equalN{f\: a}{g\: b}{2}{T[b/x]}}
}
$$
\label{fig:ccomega-algo-conv}
}

\end{center}

\caption{Cumulativity fused push-side}

\end{figure}

%% ** <- Proof
%% *** Equational theory unchanged
%% *** Sound
%% **** \Gamma \vdash A \ni^2 a \Rightarrow \Gamma \vdash A \ni^1 a
%% **** \Gamma \vdash a \ni^2 A \Rightarrow \Gamma \vdash a \in^1 A
%% **** Theorem

\begin{theorem}[Soundness]

Let $\Gamma$ a valid context.

\begin{itemize}
\item If \xspace$\Gamma \vdash \pushN{a}{2}{A}$, then $\Gamma \vdash \pushN{a}{1}{A}$
\item If \xspace$\Gamma \vdash \pullN{a}{2}{A}$, then $\Gamma \vdash \pullN{a}{1}{A}$
\end{itemize}

\end{theorem}

%% ***** Proof

\begin{proof}

The proof is by induction on the height of a derivation. The base
cases are $\ruleSet{2}$ and $\ruleVar{2}$. The translation of
$\ruleVar{2}$ is direct. Concerning $\ruleSet{2}$, we have the
following judgment:
%%
$$
\AxiomSide{i < j}{\pushN{\Set{i}}{2}{\Set{j}}}
$$

This translates to the following derivation:
%%
$$
\stkc{
\AxiomSide{\ruleSet{1}}{\pushN{\Set{i}}{1}{\Set{i+1}}} \\
\AxiomSide{\ruleName{Cuml-pull}{1}}{\pushN{\Set{i}}{1}{\Set{i+2}}} \\
\ldots \\
\CAxiom{\pushN{\Set{i}}{1}{\Set{j}}}
}
$$

The induction step is trivial on the pull-side: the rules being the
same, they map directly to their counterpart. Similarly on the
push-side, the only rule that changed is $\ruleSet{2}$ that we have
already treated in the base case. Hence, all rules translates
directly.

\end{proof}

%% *** Lemma: lifting preserves equality

\begin{lemma}[Lifting preserves equality]
\label{lemma:lifting-equality}

Let $\Gamma$ a valid context. Let $k \in N$.

If $\equalN{S}{T}{2}{A}$, then $\equalN{\Up{k}{S}}{\Up{k}{T}}{2}{A}$.

\end{lemma}

%% **** Proof

\begin{proof}

The proof is by induction on the length of an equality judgement. The
base cases are $\ruleSetEq{2}$ and $\ruleVarEq{2}$. In the
$\ruleSetEq{2}$ rule, our goal is to prove that
$\equalN{\Set{i+k}}{\Set{i+k}}{2}{\SetTop}$ that is trivially
verified. Similarly, in the $\ruleVarEq{2}$ case, we have to verify
that $\equalN{x_{i+k}}{x_{i+k}}{2}{\SetTop}$, which is indeed true.

The induction step follows naturally, by the structure of the
judgements and the definition of $\SYMBUp$.

\end{proof}

%% *** Lemma: cumulativity fused push-side

\begin{lemma}[$\ruleName{Cuml-push}{1}$ fused]
\label{lemma:cuml-push-fused}

Let $\Gamma$ a valid context.

\begin{itemize}
\item If \xspace$\Gamma \vdash \pushN{M}{1}{A}$, then for all $k \geq 0$ and
  context $\Delta$ such that $\sublift{k}{\Gamma}{\Delta}$, we have
  $\Delta \vdash \pushN{M}{2}{\Up{k}{A}}$.
\item If \xspace$\Gamma \vdash \pullN{A}{1}{\Set{i}}$, then $\Gamma \vdash
  \pullN{A}{2}{\Set{i}}$
\end{itemize}

\end{lemma}

%% **** Proof

\begin{proof}

The proof is by induction on the height of a derivation. We focus on
the push-side rules, as the pull-side remains un-changed between both
systems. The base case of interest is $\ruleSet{1}$. In this
case, we have indeed for any $k \geq 0$:
%%
$$
\AxiomSide{i+k+1 > i}{\pushN{\Set{i}}{2}{\Set{i+k+1}}}
$$

The induction step is purely structural in the case of
$\rulePi{1}$. The interesting cases are $\ruleName{Cuml-push}{1}$,
$\ruleConv{1}$, and $\ruleLam{1}$. In the $\ruleName{Cuml-push}{1}$
case, the last rule is the following:
%%
$$
\Rule{\pushN{M}{1}{A}}{\pushN{M}{i}{\Up{}{A}}}
$$
%%
By induction hypothesis on the premise with $\Delta = \Gamma$, we have
that:
%%
$$
\forall k \geq 0, \Gamma \vdash \pushN{M}{2}{\Up{k}{A}}
$$
%%
From which we can easily conclude the proof of this case:
%%
$$
\forall k \geq 0, \Gamma \vdash \pushN{M}{2}{\Up{k+1}{A}}
$$

As for the $\ruleConv{1}$ case, the last derivation is the following:
%%
$$
\Rule{\pullN{e}{1}{S} \quad
      \equalN{S}{T}{1}{\SetTop}}
     {\pushN{e}{1}{T}}
$$
%%

The equational theories of both systems being the same, we can
transport the equality from the first to the second. By
Lemma~\ref{lemma:lifting-equality}, we deduce that $\Up{k}{S} \Eq
\Up{k}{T}$. Moreover, by induction hypothesis, we have that
$\pullN{e}{2}{S}$. Applying the $\ruleName{Cuml-pull}{2}$ rule
$k$-times, we obtain $\pullN{e}{2}{\Up{k}{S}}$.  Applying the
$\ruleConv{2}$ rule, we get the desired conclusion:
$\pushN{e}{2}{\Up{k}{T}}$.

Finally, the $\ruleLam{1}$ rule corresponds to the following
derivation:
%%
$$
\Rule{\Bhab{x}{S} \vdash \pushN{b}{1}{T}}
     {\pushN{\LAM{x}{b}}{1}{\PI{\V{x}}{S}{T}}}
$$
%%
Applying the induction hypothesis on the premises, with a carefully
chosen $\Delta$, we obtain:
%%
$$
\Bhab{x}{\Up{k}{S}} \vdash \pushN{b}{1}{\Up{k}{T}}
$$
%%
From $\ruleLam{2}$, we then deduce the desired conclusion.

\end{proof}

%% *** Complete
%% **** \Gamma \vdash A \ni^1 a \Rightarrow \Gamma \vdash A \ni^2 a
%% **** \Gamma \vdash a \ni^1 A \Rightarrow \Gamma \vdash a \in^2 A
%% **** Theorem

\begin{theorem}[Completeness]

Let $\Gamma$ a valid context.

\begin{itemize}
\item If \xspace$\Gamma \vdash \pushN{a}{1}{A}$, then $\Gamma \vdash \pushN{a}{2}{A}$
\item If \xspace$\Gamma \vdash \pullN{a}{1}{A}$, then $\Gamma \vdash \pullN{a}{2}{A}$
\end{itemize}

\end{theorem}

%% ***** Proof

\begin{proof}

The proof is by induction on the length of a derivation. The base
cases are $\ruleSet{1}$ and $\ruleVar{1}$. They both translate
directly to the other system.

The induction is trivial for all rules --- by existence of an
identical rule in the other system --- at the exception of the
$\ruleName{Cuml-push}{1}$ rule. In this case, the last derivation is
the following:
%%
$$
\Rule{\pushN{M}{1}{A}}
     {\pushN{M}{1}{\Up{}{A}}}
$$
%%
Applying Lemma \ref{lemma:cuml-push-fused} on the premise, we obtain:
%%
$$
\forall k \geq 0, \Gamma \vdash \pushN{M}{2}{\Up{k}{A}}
$$
From which we can conclude, by taking $k$ to be 1.

\end{proof}

%% * Annotated pull-side
\section{Manually Annotated Stratisfactor}

%% ** <- Generalizing CUML to Up
%% *** <- x :<: S for S = Set_i
%% *** <- x :<: A -> B ?
%% *** -> FTP/GFTP
%% *** -> x : Set(i) means x : Set(i+)
%% ** -> Annotating by the lifting required
%% *** -> Guess work in the hand of the annotator
%% *** -> Figure: annotated type system
%% ** <- Figure: Annotated pull-side

\begin{figure}

\begin{center}
\input{figure_annotated}
\end{center}

\caption{Annotated pull-side}

\end{figure}

%% ** <- Proof
%% *** GFTP
%% **** Theorem

\begin{theorem}[GFTP]
\label{thm:gftp}

Let $\Gamma$ a valid context.

If \xspace$\Gamma \vdash \pullN{a^{(j)}}{3}{\Up{j}{A}}$, then for any $i \geq j$ and
for all $\Delta$ such that $\sublift{i}{\Gamma}{\Delta}$, we have:
%%
$$
\RuleSide{\Gamma \vdash \pullN{a^{(j)}}{3}{\Up{j}{A}}}
         {k \in \N}
         {\Delta \vdash \pullN{a^{(k)}}{3}{\Up{i}{A}}}
$$

And similarly on the push-side, for if \xspace
$\Gamma \vdash \pushN{a}{3}{A}$.

\end{theorem}

%% ***** Proof

\begin{proof}

The proof is by induction on the length of derivations. The base cases
are $\ruleSet{3}$ and $\ruleVar{3}$. It is easily verified for the
$\ruleSet{3}$ rule, the interesting case being $\ruleVar{3}$. We have
that:
%%
$$
\CAxiom{\Bhab{x}{A} \vdash \pullN{x^{(l)}}{3}{\Up{l}{A}}}
$$
%%
We have to find the right annotation $k$ on $x$ so that
$\pullN{x^{(k)}}{\Up{i}{A}}$ in a context $\Delta$ containing
$\Bhab{x}{\Up{j}{A}}$ with $j \leq i$. Taking $l$ to be 
\xspace$i-j \geq 0$, we have the following valid derivation:
%%
$$
\CAxiom{\Bhab{x}{\Up{j}{A}} \vdash \pullN{x^{(i-j)}}{3}{\Up{i-j}{(\Up{j}{A})}}}
$$
%%
Which concludes this case.

The induction steps can be handled easily for all rules pushing a
$\Set{l}$ into a term. First, the push side is identical to the
previous one. Hence, Lemma \ref{lemma:cuml-push-fused} still applies
and gives:
%%
$$
\forall k \geq 0, \Gamma \vdash \pushN{A}{3}{\Set{l+k}}
$$
%%

By definiton of $\SYMBUp$, this is equivalent to our goal:
$\forall i \geq 0, \Gamma \vdash \pushN{a}{3}{\Up{i}{\Set{l}}}$.

There are two remaining rules on the push side: $\ruleConv{3}$ and
$\ruleLam{3}$. The premises of the $\ruleConv{3}$ rule tell us that
$\pullN{e^{(j)}}{3}{\Up{j}{S}}$ and
$\equalN{\Up{j}{S}}{T}{3}{\SetTop}$. By induction hypothesis, we
deduce that $\pullN{e^{(k)}}{3}{\Up{i+j}{S}}$ with $k \geq 0$. By
Lemma \ref{lemma:lifting-equality}, we also deduce that
$\equalN{\Up{i}{\Up{j}{S}}}{\Up{i}{T}}{3}{\SetTop}$. Hence, by $\ruleConv{3}$,
we conclude the desired judgement, \ie $\pushN{e}{3}{\Up{i}{T}}$.

The premise of the $\ruleLam{3}$ rule gives $\Bhab{x}{S} \vdash
\pushN{b}{3}{T}$. By applying the induction hypothesis with a suitably
lifted context, we get that: $\Bhab{x}{\Up{i}{S}} \vdash
\pushN{b}{3}{T}$. Applying rule $\ruleLam{3}$, we conclude that
$\pushN{\LAM{x}{b}}{3}{\Up{i}{\PI{\V{x}}{S}{T}}}$, by definition of
$\SYMBUp$ on a $\Pi$-type.

Finally, we handle the two remaining rules: $\ruleAsc{3}$ and
$\ruleApp{3}$. The ascription rule is easy as its type is governed by
the annotation we set: hence, to obtain an element in $\Up{i}{T}$, we
simply annotate $\Bhab{t}{T}$ by $(i)$. The premises of the
$\ruleApp{3}$ rule inform us that
$\pullN{f^{(j)}}{3}{\PI{\V{x}}{\Up{j}{S}}{\Up{j}{T}}}$ and
$\pushN{s}{3}{\Up{j}{S}}$. Applying the induction hypothesis on both,
we get that $\pullN{f^{(k)}}{3}{\PI{\V{x}}{\Up{i}{S}}{\Up{i}{T}}}$ and
$\pushN{s}{3}{\Up{i}{S}}$. We can apply $\ruleApp{3}$ straightaway and
conclude that $\pullN{{f\: s}^{(k)}}{3}{\Up{i}{T}[s/x]}$.

\end{proof}

%% **** FTP

\begin{theorem}[FTP]

In a context $\Gamma$ valid, if $\Gamma \vdash \pullN{a}{3}{A}$, then
the term $a$ checks all polymorphic instances of $A$, \ie $\forall i
\geq 0, \Gamma \vdash \pullN{a}{3}{\Up{i}{A}}$

\end{theorem}

%% ***** Proof

\begin{proof}

Corollary of Theorem~\ref{thm:gftp}, taking $\Delta$ to be $\Gamma$.

\end{proof}

%% *** Sound
%% **** \Gamma \vdash A \ni^3 a \Rightarrow \forall j \geq 0, \Gamma \vdash \Up{j}{A} \ni^2 a
%% **** \Gamma \vdash a \in^3 A \Rightarrow \forall j \geq 0, \Gamma \vdash a \in^2 \Up{j}{A}
%% **** Theorem

\begin{theorem}[Soundness]

Let $\Gamma$ a valid context.

\begin{itemize}
\item If \xspace$\Gamma \vdash \pullN{a^{(j)}}{3}{\Up{j}{A}}$, then for all 
      $k \in \N$, we have $\Gamma \vdash \pullN{a}{2}{\Up{k}{A}}$
\item If \xspace$\Gamma \vdash \pushN{a}{3}{A}$, then 
      $\Gamma \vdash \pushN{a}{2}{A}$
\end{itemize}

\end{theorem}

%% ***** Proof

\begin{proof}

Let $k \in \N$.

First, the push-side rules of both systems being identical, the
translation from one system to the other is trivial. Therefore, we
will focus on the pull-side.

The proof is by induction on the height of derivations. The base case
of interest here is $\ruleVar{3}$. We have the following:
%%
$$
\CAxiom{\Bhab{x}{A} \vdash \pullN{x^{(j)}}{3}{\Up{j}{A}}}
$$

This translates to an application of the $\ruleVar{2}$ rule followed
by as many application of the $\ruleName{Cuml-pull}{2}$ necessary to
reach $k$:
%%
$$
\stkc{
\CAxiom{\Bhab{x}{A} \vdash \pullN{x}{2}{A}} \\
\CAxiom{\Bhab{x}{A} \vdash \pullN{x}{2}{\Up{}{A}}} \\
\ldots \\
\CAxiom{\Bhab{x}{A} \vdash \pullN{x}{2}{\Up{k}{A}}}
}
$$

The induction step consists in treating the $\ruleAsc{3}$ and
$\ruleApp{3}$ rules. The $\ruleAsc{3}$ requires exactly the same
treatment than the $\ruleVar{3}$ rule: by induction hypothesis on the
premises, we can apply the corresponding rule in the previous
system. We then use $\ruleName{Cuml-pull}{2}$ to complete the missing
liftings to reach $\Up{k}{T}$.

The premises of the $\ruleApp{3}$ rule give us that
$\pullN{f^{(j)}}{3}{\PI{\V{x}}{\Up{j}{S}}{\Up{j}{T}}}$ and
$\pushN{s}{3}{\Up{j}{S}}$. By induction hypothesis, we get that
$\pullN{f}{2}{\PI{\V{x}}{\Up{k}{S}}{\Up{k}{T}}}$ and
$\pushN{s}{2}{S}$. Applying Lemma~\ref{lemma:cuml-push-fused} for
$\Delta = \Gamma$, we obtain that
$\pushN{s}{2}{\Up{k}{S}}$. Therefore, by application of $\ruleApp{2}$,
we deduce that $\pushN{f\: s}{2}{\Up{k}{T}[s/x]}$.

\end{proof}

%% *** Complete
%% **** \Gamma \vdash A \ni^2 a \Rightarrow \Gamma \vdash A \ni^3 a
%% **** \Gamma \vdash a \in^2 A \Rightarrow \Gamma \vdash a \in^3 A
%% **** Theorem

\begin{theorem}[Completeness]

Let $\Gamma$ be a valid context.

\begin{itemize}
\item If \xspace$\Gamma \vdash \pullN{a}{2}{A}$, then for any 
      $k \in \N$, there exists an annotation $j \in \N$ such that we have 
      $\Gamma \vdash \pullN{a^{(j)}}{3}{\Up{k}{A}}$
\item If \xspace$\Gamma \vdash \pushN{a}{2}{A}$, then $\Gamma \vdash
      \pushN{a}{3}{A}$
\end{itemize}

\end{theorem}

%% ***** Proof

\begin{proof}

Let $k \in \N$

First, the push-side is rather unexciting, as it is the same in both
systems. Hence, it translates directly from one to another. Therefore,
we will focus on the pull-side.

The proof is by induction on the height of derivations. The base case
of interest is $\ruleVar{2}$ here. Hence, we have $\Bhab{x}{A}$ in the
context. By direct application of $\ruleVar{3}$, we obtain the desired
result:
%%
$$
\CAxiom{\Bhab{x}{A} \vdash \pullN{x^{(k)}}{3}{\Up{k}{A}}}
$$

The induction steps consists of the rules $\ruleAsc{2}$,
$\ruleApp{2}$, and $\ruleName{Cuml-pull}{2}$. The $\ruleAsc{2}$ rule
translates the same way than the $\ruleVar{2}$ rule. The premises of
the $\ruleApp{2}$ give us that $\pullN{f}{2}{\PI{\V{x}}{S}{T}}$ and
$\pushN{s}{2}{S}$. By induction hypothesis, we deduce that
$\pullN{f^{(k)}}{3}{\Up{k}{(\PI{\V{x}}{S}{T})}}$ and
$\pushN{s}{3}{S}$. By application of the FTP, we get that
$\pushN{s}{3}{\Up{k}{S}}$. Applying the $\ruleApp{3}$ rule, we
conclude that $\pullN{{f\: s}^{(k)}}{3}{\Up{k}{T}[s/x]}$.

We conclude with the $\ruleName{Cuml-pull}{2}$ rule. The derivation
ends with the following judgment:
%%
$$
\Rule{\pullN{M}{2}{A}}
     {\pullN{M}{2}{\Up{}{A}}}
$$
%%
By induction hypothesis on the premise, we deduce that $\forall l \in
\N, \pullN{M^{(l)}}{3}{\Up{l}{A}}$. In particular, we have that
$\pullN{M^{(k+1)}}{3}{\Up{k}{(\Up{}{A})}}$, which is the expected
conclusion.

\end{proof}

%% * Semi-automatic
\section{Semi-automatic Stratisfactor}

%% ** <- State properties
%% *** /> Prove them a posteriori
%% ** <- Shape classification
%% *** <- Calculus of constants
%% **** <- They automatically lift
%% **** -> No need to adjust them
%% *** <- Calculus of *polymorphic* functions
%% **** <- We have to specify where they exist
%% ***** /> Cannot predict at which level on use-site
%% ***** -> Introduce variable lifting

The type $\greenb{\Up{j}{A}}{j}{k}$ is an object lifted by an unknown
amount $j \geq k$.

%% **** <- Use-site: check that a coherent assignment can be made
%% **** /> App-rule: green on red side
%% ***** -> Find the minimal assignment k such that A_j=k :>: s

\begin{definition}[Minimal assignment]

Let $k \in \N$.

We say that \emph{$k$ is a minimal assignement} in 
%%
$$
\Rule{\pi}
     {\pushN{s}{4}{\redb{\greenb{\Up{j}{S}}{j=k}{0}}}}
$$
%%
If, for all uses of the $\ruleSet{4}$ rule in $\pi$, the minimal
difference of index to a $j$-indexed set is 1.

Put otherwise, the bound on $j$ is tight: decrementing $k$ would lead
to a size issue.

\end{definition}

%% ***** /> Rely on annotation for now
%% *** <- Calculus of conversions
%% **** <- Conversion rule: use-site of a polymorphic object
%% ***** -> Set the variable
%% **** <- Push-side orders a given type
%% ***** -> Can we find an assignment to the variable on S?
%% ***** -> Conversion rule: S == T
%% ****** -> Solve variable
%% ****** -> Check for coherence
%% ** Figure: semi-automatic type system

\begin{figure}

\begin{center}
\subfigure[Push-side]{
\stkc{
%% Form
\boxed{\Gamma \vdash \pushN{\CN{term}}{4}{\redb{\CN{type}}}}
\\
\\
%% Set
\AxiomSide{i < j \quad \ruleSet{4}}{\pushN{\Set{i}}{4}{\redb{\Set{j}}}}
\qquad
%% Arrow
\RuleSide{\pushN{S}{4}{\redb{\Set{i}}} \quad
          \Bhab{x}{\redb{S}} \vdash \pushN{T}{4}{\redb{\Set{i}}}}
         {\rulePi{4}}
         {\pushN{\PI{\V{x}}{S}{T}}{4}{\redb{\Set{i}}}}
\\
%% Inhabitant of Arrow
\RuleSide{\Bhab{x}{\redb{S}} \vdash \pushN{b}{4}{\redb{T}} }
         {\ruleLam{4}}
         {\pushN{\LAM{\V{x}}{b}}{4}{\redb{\PI{\V{x}}{S}{T}}}} 
}
} \qquad
\subfigure[Pull-side]{
\stkc{
%% Form
\boxed{\Gamma \vdash \pullN{\CN{term}}{4}{\greenb{\CN{type}}{j}{k}}}
\\
\\
%% Variable extraction
\AxiomSide{\ruleVar{4}}
          {\Bhab{x}{\redb{A}} \vdash \pullN{x}{4}{\greenb{\Up{j}{\redb{A}}}{j}{0}}}
\qquad
%% Type ascription
\RuleSide{\pushN{T}{4}{\blueb{\SetTop}} \quad
          \pushN{t}{4}{\redb{T}}}
         {\ruleAsc{4}}
         {\pullN{(\Bhab{t}{\redb{T}})}{4}{\greenb{\Up{j}{T}}{j}{0}}}
\\
%% Function application
\RuleSide{\pullN{f}{4}{\greenb{\PI{\V{x}}{\greenb{\Up{j}{S}}{j}{k}}{\greenb{\Up{T}{j}}{j}{k}}}{j}{k}} \quad
          \pushN{s}{4}{\redb{\greenb{\Up{j}{S}}{j=l}{k}}}}
         {k \in \mathbb{N} \quad \ruleApp{4}}
         {\pullN{f\: s}{4}{\greenb{\Up{j}{T}[s/x]}{j}{k+l}}}
}
}

\subfigure[Conversion]{
$$
\stkc{
%% Conversion rule
\RuleSide{\pullN{e}{4}{\greenb{\Up{j}{S}}{j}{k}} \quad
          \equalN{\greenb{\Up{j}{S}}{j}{k}}
                 {\redb{T}}
                 {4}
                 {\blueb{\SetTop}}, j = k_1}
         {\ruleConv{4}}
         {\pushN{e}{4}{\redb{T}}}
\\
\\
%% Form
\boxed{\Gamma \vdash \equalN{\greenb{\CN{term}}{j}{k}}{\redb{\CN{term}}}{4}
                            {\blueb{\CN{type}}}, j = l }
\\
\\
%% Set equality
\AxiomSide{k_1 = l \quad 
           \ruleSetKEq{4}}
          {\equalN{\greenb{\Set{k_1}}{j}{k}}{\redb{\Set{l}}}{4}{\blueb{\SetTop}}, .}
\\
\AxiomSide{l - k_1 \geq 0 \quad
           \ruleSetVEq{4}}
          {\equalN{\greenb{\Set{j+k_1}}{j}{k}}{\redb{\Set{l}}}{4}{\blueb{\SetTop}}, j = l - k_1}
\\
%% Variable equality
\AxiomSide{k_1 = l \quad \ruleVarKEq{4}}
          {\Bhab{x}{\blueb{S}} \vdash \equalN{\greenb{x_{k_1}}{j}{k}}
                                             {\redb{x_l}}
                                             {4}
                                             {\blueb{S}}, .}
\\
\AxiomSide{l - k_1 \geq 0 \quad \ruleSetVEq{4}}
          {\Bhab{x}{\blueb{S}} \vdash \equalN{\greenb{x_{j+k_1}}{j}{k}}
                                             {\redb{x_l}}
                                             {4}
                                             {\blueb{S}}, j = l - k_1}
\\
%% Pi equality
\RuleSide{\equalN{\greenb{S_1}{j}{k}}
                 {\redb{S_2}}
                 {4}
                 {\blueb{\SetTop}}, j = k_1 \quad
          \Bhab{x}{\redb{S_2}} \vdash \equalN{\greenb{T_1}{j}{k}}
                                             {\redb{T_2}}
                                             {4}
                                             {\blueb{\SetTop}}, j = k_1}
         {\rulePiEq{4}}
         {\equalN{\greenb{\PI{\V{x}}{\greenb{S_1}{j}{k}}{\greenb{T_1}{j}{k}}}{j}{k}} 
                 {\redb{\PI{\V{x}}{\redb{S_2}}{\redb{T_2}}}}
                 {4}
                 {\blueb{\SetTop}}, j = k_1}
\\
%% Function equality
\RuleSide{\Bhab{x}{\blueb{S}} \vdash \equalN{\greenb{f}{j}{k}\: x}
                                            {\redb{g}\: x}
                                            {4}
                                            {\blueb{T}}, j = k_1 }
         {\ruleLamEq{4}}
         {\equalN{\greenb{f}{j}{k}}
                 {\redb{g}}
                 {4}
                 {\blueb{\PI{\V{x}}{S}{T}}}, j = k_1}
\\
%% Application equality
\RuleSide{\equalN{\greenb{f}{j}{k}}
                 {\redb{g}}
                 {4}
                 {\blueb{\PI{\V{x}}{S}{T}}}, j = k_1 \quad
          \equalN{\greenb{a}{j}{k}}
                 {\redb{b}}
                 {4}
                 {\blueb{S}}, j = k_1}
         {\ruleAppEq{4}}
         {\equalN{\greenb{\greenb{f}{j}{k}\: \greenb{a}{j}{k}}{j}{k}}
                 {\redb{\redb{g}\: \redb{b}}}
                 {4}
                 {\blueb{T [\redb{b}/x]}}, j = k_1}
}
}
\end{center}

\caption{Semi-automatic stratisfactor}
\label{fig:semi-automatic}

\end{figure}

%% ** <- Proofs
%% *** <- Color respected - only one live variable at a time
%% ***** <- Greens under control

\begin{lemma}[Color respect and variable uniqueness]

The following scheme is preserved by the judgments presented in
Figure~\ref{fig:semi-automatic}:
\begin{itemize}
\item A push rule is colored as:
      $\pushN{\CN{term}}{4}{\redb{\CN{type}}}$,
\item A pull rule is colored as:
      $\pullN{\CN{term}}{4}{\greenb{\CN{type}}{j}{k}}$,
\item An equality judgement is colored as:
      $\equalN{\greenb{\CN{term}}{j}{k}}{\redb{\CN{term}}}{4}{\blueb{\CN{term}}}$,
      and
\item Terms (left of the $\in$ symbol, right of the $\ni$ symbol) are
      red boxed.
\end{itemize}

And at any point in time, there exists only one live green variable.

\end{lemma}

%% ***** Proof

\begin{proof}

The proof is by induction on the height of a derivation. The base
cases are $\ruleSet{4}$, $\ruleVar{4}$, $\ruleSetEq{4}$, and
$\ruleVarEq{4}$. We simply check that the color of these rules follows
the color scheme. Note that in $\ruleVar{4}$, uniqueness of variable
comes from the fact that $A$ is red (containing no variable) from the
context. Hence, creating a new variable $j$ is safe.

The induction step then consists in considering each rule in turn:
assuming that the premises are well-colored, we have to check that the
conclusion is well-colored as well. This is trivial for most
rules. The only interesting case is $\ruleApp{4}$. Applying the
induction hypothesis on the premise, we have that
$\pullN{f}{4}{\greenb{\PI{\V{x}}{\greenb{S}{j}{k}}{\greenb{T}{j}{k}}}{j}{k}}$
with $f$ red. By setting $j$ to some $l$, we turn the green $S$ into a
the red term $\redb{\greenb{S}{j=l}{k}}$ hence being coherent with the
induction hypothesis that also tells us that $s$ is red. Hence, we
deduce that $f\: s$ is red, inhabiting
$\greenb{T[s/x]}{j}{k+l}$. Substituting $s$ in $T$ preserves
uniqueness of variable, as $s$ is red, hence containing no
variable. Uniqueness of variables also require some care in the
$\ruleAsc{4}$ rule where we introduce a fresh variable: this is again
justified by the fact that the type $T$ is red, hence containing no
variable.

\end{proof}

%% *** <- Variable assigmnent equivalent to lifting on base
%% **** \Up{l}{\greenb{A}{j=0}{k}} == \greenb{A}{j=l}{k}
%% **** Lemma

\begin{lemma}[Equivalence of variable assignment and lifting]
\label{lemma:lifting-var-assgnmt}

Let $l \geq k$. Let $A$ be a green type of variable $j$:
$\greenb{A}{j}{k}$. The following equality is always verified:
%%
$$
\greenb{\Up{j}{A}}{j=l}{k} = \Up{l}{\greenb{\Up{j}{A}}{j=0}{k}}
$$

\end{lemma}

%% ***** Proof

\begin{proof}

By manipulating the definitions:
%%
$$
\greenb{\Up{j}{A}}{j=l}{k} = \Up{l}{A} = \Up{l}{\greenb{\Up{j}{A}}{j=0}{k}}
$$

\end{proof}

%% *** <- No change to the definitional equality

\begin{lemma}[Equivalence of definitional equalities]
\label{lemma:equiv-def-eq-3-4}

Let $k \in \N$.

\begin{itemize}
\item If \xspace$\equalN{\Up{l}{S}}{T}{3}{A}$, then
      $\equalN{\greenb{\Up{j}{S}}{j}{k}}{\redb{T}}{4}{\blueb{A}},j=l$
\item If \xspace$\equalN{\greenb{\Up{j}{S}}{j}{k}}{\redb{T}}{4}{\blueb{A}},j=k_1$,
      then $\equalN{\Up{k_1}{S}}{T}{3}{A}$
\end{itemize}

\end{lemma}

%% **** Proof

\begin{proof}

Both proofs are by induction on the height of a derivation. The
induction step is purely structural in both cases. The only
interesting cases are the base cases, $\ruleSetEq{}$ and
$\ruleVarEq{}$. This is simply a matter of book-keeping.

\end{proof}

%% *** <- Sound
%% **** \Gamma \vdash A \ni^4 a \Rightarrow \Gamma \vdash A \ni^3 a
%% **** \Gamma \vdash a \in^4 \greenb{A}{j}{k} \Rightarrow \forall j \leq k, \Gamma \vdash a^{(j)} \in^3 A
%% **** Theorem

\begin{theorem}[Soundness]

Let $\Gamma$ a valid context. Let $k \in \N$.

\begin{itemize}
\item If \xspace$\Gamma \vdash \pushN{a}{4}{\redb{A}}$, then 
      $\Gamma \vdash \pushN{a}{3}{A}$
\item If \xspace$\Gamma \vdash \pullN{a}{4}{\greenb{\Up{j}{A}}{j}{k}}$, then for all
      $m \geq k$, we have that $\Gamma \vdash \pullN{a^{(m)}}{4}{\Up{m}{A}}$
\end{itemize}

\end{theorem}

%% ***** Proof

\begin{proof}

First, apart from the conversion rule, the push-side has not changed
between the previous system and the new one. Thefore, we will not
treat thoses cases, the translation being the obvious one.

The proof is by induction on the height of a derivation. The base case
of interest is $\ruleVar{4}$, which gives us that $\Bhab{x}{A}$
inhabits the context. By annotation $x$ by some $m \geq 0$, we obtain
the desired result:
%%
$$
\CAxiom{\Bhab{x}{A} \vdash \pullN{x^{(m)}}{3}{\Up{m}{A}}}
$$

The induction step consists in studying $\ruleAsc{4}$, $\ruleApp{4}$,
and $\ruleConv{4}$. The translation of the $\ruleAsc{4}$ rule follows
the pattern introduced in the proof of the $\ruleVar{4}$ rule.

The $\ruleApp{4}$ rule is the following:
%%
$$
\Rule{\pullN{f}{4}{\PI{\V{x}}{\greenb{\Up{j}{S}}{j}{k}}{\greenb{\Up{j}{T}}{j}{k}}} \quad
      \pushN{s}{4}{\redb{\greenb{\Up{j}{S}}{j=l}{k}}}}
     {\pullN{f\: s}{4}{\greenb{\Up{j}{T}}{j}{k+l}}}
$$
%%
By induction hypothesis on the premises, we obtain that $\forall j
\geq k, \pullN{f^{(j)}}{3}{\PI{\V{x}}{\Up{j}{S}}{\Up{j}{T}}}$ and
$\Gamma \vdash \pushN{s}{3}{\Up{l}{S}}$. For any $m \geq k+l \geq
k$, we therefore have that
$\pullN{f^{(j)}}{3}{\PI{\V{x}}{\Up{j}{S}}{\Up{j}{T}}}$. Moreover, by
the FTP in (3), we deduce that $\Gamma \vdash \pushN{s}{3}{\Up{m}{S}}$
holds, $m$ being greater or equal to $l$. Applying $\ruleApp{3}$
on these premises, we get the desired conclusion.

Finally, let us consider the conversion rule:
%%
$$
\RuleSide{\pullN{e}{4}{\greenb{\Up{j}{S}}{j}{k}} \quad
          \equalN{\greenb{\Up{j}{S}}{j}{k}}
                 {\redb{T}}
                 {4}
                 {\blueb{\SetTop}}, j = k_1}
         {\ruleConv{4}}
         {\pushN{e}{4}{\redb{T}}}
$$
%%
By induction hypothesis on the premise, we get that $\forall j \geq k,
\pullN{e}{3}{\Up{j}{S}}$. In particular, it also holds for $j = k_1$
($k_1 \geq k$, by construction). By Lemma~\ref{lemma:equiv-def-eq-3-4},
we also deduce that $\equalN{\Up{k_1}{S}}{T}{3}{\SetTop}$. Therefore, by
application of $\ruleConv{3}$, we conclude that $\pushN{e}{3}{T}$.

\end{proof}

%% *** <- Complete
%% **** \Gamma \vdash A \ni^3 a \Rightarrow \Gamma \vdash A \ni^4 a
%% **** \Gamma \vdash a^{(l)} \in^3 A \Rightarrow \Gamma \vdash a \in^4 \greenb{A}{j=l}{0}
%% **** Theorem

\begin{theorem}[Completeness]

Let $\Gamma$ a valid context. Let $k \in \N$.

\begin{itemize}
\item If \xspace$\Gamma \vdash \pushN{a}{3}{A}$, then 
      $\Gamma \vdash \pushN{a}{4}{A}$
\item If for any $l \geq k$, we have 
      \xspace$\Gamma \vdash \pullN{a^{(l)}}{3}{\Up{l}{A}}$, then 
      $\Gamma \vdash \pullN{a}{4}{\greenb{\Up{j}{A}}{j}{k}}$
\end{itemize}

\end{theorem}

%% ***** Proof

\begin{proof}

Again, the push-side is left unchanged between both systems, at the
exception of the conversion rule. Therefore, we will not explicitly
treat those cases, which translation is the obvious one.

The proof is by induction on the height of a derivation. The
interesting base case is $\ruleVar{3}$, which gives us that
$\Bhab{x}{A}$ is in the context. We simply apply $\ruleVar{4}$ and
deduce the desired consequence:
%%
$$
\CAxiom{\Bhab{x}{A} \vdash \pullN{x}{4}{\greenb{\Up{j}{A}}{j}{0}}}
$$

The induction step is reduced to the $\ruleAsc{3}$, $\ruleApp{3}$, and
$\ruleConv{3}$ rules. The translation of $\ruleAsc{3}$ follows the one
of $\ruleVar{3}$. The translation of $\ruleApp{3}$ is slightly more
involved. The premise tells us that $\forall l \geq k,
\pullN{f^{(l)}}{3}{\PI{\V{x}}{\Up{l}{S}}{\Up{l}{T}}}$ and
$\pushN{s}{3}{\Up{l}{S}}$. By induction hypothesis, we deduce that
$\pullN{f}{4}{\greenb{\PI{\V{x}}{\Up{j}{S}}{\Up{j}{T}}}{j}{k}}$ and
$\pushN{s}{4}{\redb{S}}$. The latter is the same as
$\pushN{s}{4}{\redb{\greenb{\Up{j}{S}}{j=0}{k}}}$. Hence, by
application of $\ruleApp{4}$, we deduce the desired conclusion:
$\pullN{f\: s}{4}{\greenb{\Up{j}{T}[s/x]}{j}{k}}$.

Finally, we treat the conversion rule:
%%
$$
\Rule{\pullN{e^{(l)}}{3}{\Up{l}{S}} \quad
      \equalN{\Up{l}{S}}{T}{3}{\SetTop}}
     {\pushN{e}{3}{T}}
$$

We cannot directly apply the induction hypothesis on
$\pullN{e^{(l)}}{3}{\Up{l}{S}}$, as it is true for a specific $l$
here. However, by the FTP, we obtain that $\forall j \geq l,
\pullN{e^{(j)}}{3}{\Up{j}{S}}$. Here induction hypothesis applies and
we get that $\pullN{e}{4}{\greenb{\Up{j}{S}}{j}{l}}$. By
Lemma~\ref{lemma:equiv-def-eq-3-4}, we also have that there exists
$k_1 \leq l$ such that
$\equalN{\greenb{S}{j}{l}}{\redb{T}}{4}{\blueb{A}},j=k_1$. Hence, by
application of $\ruleConv{4}$, we obtain that $\pushN{e}{4}{\redb{T}}$

\end{proof}

%% ** /> How to find k?
%% * Automated
\section{Automatic Stratisfactor}

%% ** <- Finding k
%% *** -> Distinction between type error and size issue
%% **** <- Type error: \forall k. \Up{k}{\greenb{A}{j=0}} \not \ni s
%% **** <- Size issue: \exists k. \Up{k}{\greenb{A}{j=0}} \ni s
%% ***** <- More precisely: \exists k. \forall k' \geq k. \Up{k'}{\greenb{A}{j=0}} \ni s
%% *** -> Modify push-side to return the differential
%% **** -> A \ni^5 a, k \Leftrightarrow \forall k' \geq k, \Up{k'}{A} \ni^4 a, \Up{k-1} \not \ni^4 a (or k = 0)
%% **** -> Proof a posteriori
%% ** <- Figure: automated type system

\begin{figure}

\begin{center}
\input{figure_strat}
\end{center}

\caption{Set polymorphic type theory}

\end{figure}

%% ** <- Proofs
%% *** <- Same equational theory
%% *** <- Push-side gives differential
%% **** A \ni^5 a, k \Rightarrow \forall k' \geq -\min\: k\: 0, \Up{k'}{A} \ni^4 a

\begin{lemma}[Push-side is not worst than the optimal bound]
\label{lemma:push-k-good-enough}

Let $\Gamma$ a valid context.

If \xspace$\Gamma \vdash \pushN{a}{5}{A}, k$, then forall $k'$ such
that $k' \geq -min(k,0)$ and $\Delta$ such that
$\sublift{\Gamma}{k'}{\Delta}$, we have $\Gamma \vdash
\pushN{a}{4}{\Up{k'}{A}}$. For if $\Gamma \vdash \pullN{a}{5}{A}$,
then $\Gamma \vdash \pullN{a}{4}{A}$.

\end{lemma}

%% ***** Proof

\begin{proof}

First, we do not cover the pull-side in depth: at the exception of the
$\ruleApp{5}$ rule, the judgements are identical in both systems,
hence they trivially translate from the one to the other.

We proceed by induction on the height of derivations. The base case is
the $\ruleSet{5}$ rule:
%%
$$
\CAxiom{\pushN{\Set{i}}{5}{\redb{\Set{j}}}, j - i - 1}
$$
%%
There are two cases to consider: first, if $j-i-1 \leq 0$; second, if
$j-i-1 > 0$. In the first case, we have to prove that $\forall k' \geq
i+1-j, \pushN{\Set{i}}{4}{\Up{k'}{\Set{j}}}$. This indeed the case as
$j + k' \geq j - (j - i - 1) \geq i + 1 > i$, hence satisfying the
rule $\ruleSet{4}$. In the second case, we have to prove that $\forall
k' \geq 0, \pushN{Set{i}}{4}{\Up{k'}{\Set{j}}}$. This is the case as
well, as $j + k' \geq k + i + 1 \geq i + 1 > i$.

The induction steps consists in studying the $\ruleLam{5}$,
$\ruleConv{5}$, and $\rulePi{5}$. The $\ruleLam{5}$ follows naturally
from the induction hypothesis on the premise, with $\Delta$ carefully
chosen to lift the type of $x$: $\Bhab{x}{\Up{k'}{A}}$. Applying
$\ruleLam{4}$, we get the desired outcome.

The conversion rule requires more work. We have:
%%
$$
\Rule{\pullN{e}{5}{\greenb{\Up{j}{S}}{j}{k}} \quad
      \equalN{\greenb{S}{j}{k}}{\redb{T}}{5}{\blueb{\SetTop}}, j = k_1}
     {\pushN{e}{5}{\redb{T}}, 0}
$$
%%
First, by induction hypothesis on the premise, we have that
$\pullN{e}{5}{\greenb{\Up{j}{S}}{j}{k}}$ for some $k \in \N$ and a
variable $j$. Moreover, both equational theories are the same, so we
also have that
$\equalN{\greenb{S}{j}{k}}{\redb{T}}{4}{\blueb{\SetTop}}, j = k_1$.
Applying $\ruleConv{4}$, we can readily conclude that
$\pushN{e}{4}{T}$. The FTP then gives us the expected result: 
$\forall k' \geq 0, \pushN{e}{4}{\Up{k'}{T}}$.

The last push-side case is $\rulePi{5}$. We have:
%%
$$
\Rule{\pushN{S}{5}{\redb{\Set{i}}}, k_1 \quad
      \Bhab{x}{\redb{S}} \vdash \pushN{T}{5}{\redb{\Set{i}}}, k_2}
     {\pushN{\PI{\V{x}}{S}{T}}{5}{\redb{\Set{i}}}, \min k_1 k_2}
$$
%%
By induction hypothesis on the premises, we obtain:
%%
$$
\begin{array}{l}
\forall k'_1 \geq -\min(k_1, 0), \Gamma \vdash \pushN{S}{4}{\Up{k'_1}{\Set{i}}} \\
\forall k'_2 \geq -\min(k_2, 0), \Gamma, \Bhab{x}{S} \vdash \pushN{T}{4}{\Up{k'_2}{\Set{i}}}
\end{array}
$$

Finally, we shall treat the only interesting pull-side rule:
$\ruleApp{5}$. We have that:
%%
$$
\RuleSide{\pullN{f}{5}{\greenb{\PI{\V{x}}{\greenb{\Up{j}{S}}{j}{k}}{\greenb{\Up{j}{T}}{j}{k}}}{j}{k}} \quad
          \pushN{s}{5}{\redb{\greenb{\Up{j}{S}}{j=0}{k}}}, l}
         {l' = \min l\: 0}
         {\pullN{f\: s}{5}{\greenb{\Up{j}{T}[\blackb{s}/x]}{j}{k-l'}}}
$$
%%

By induction hypothesis on the premises, we have that
$\pullN{f}{4}{\greenb{\PI{\V{x}}{\greenb{\Up{j}{S}}{j}{k}}{\greenb{\Up{j}{T}}{j}{k}}}{j}{k}}$
and $\pushN{s}{4}{\Up{-l'}{\redb{\greenb{\Up{j}{S}}{j=0}{k}}}}$. By
Lemma~\ref{lemma:lifting-var-assgnmt}, we get that
$\pushN{s}{4}{\redb{\greenb{\Up{j}{S}}{j=-l'}{k}}}$. Therefore, we can
apply $\ruleApp{4}$ and deduce the expected conclusion: 
$\pullN{f\: s}{4}{\greenb{\Up{j}{T}[\blackb{s}/x]}{j}{k-l'}}$.

\end{proof}

%% *** <- Tight bound gives difference of zero
%% **** \greenb{S}{j=k}{0} \ni^4 s \Rightarrow \greenb{S}{j=k}{0} \ni^5 s, 0

\begin{lemma}[Push-side captures the tight bound]
\label{lemma:push-k-tight}

Let $\Gamma$ a valid context.

If \xspace$\pushN{s}{4}{\redb{\greenb{\Up{j}{S}}{j=l}{k}}}$ where $l
\in \N$ is the minimal assignment, then
$\pushN{s}{5}{\redb{\greenb{\Up{j}{S}}{j=0}{k}}}, -l$. For if
$\pullN{s}{4}{S}$, then $\pullN{s}{5}{S}$.

\end{lemma}

%% ***** Proof

\begin{proof}

Again, at the exception of the $\ruleApp{4}$ rule, the pull-side is
identical in both systems, hence we take the trivial translation. The
proof is by induction on the height of derivations. The base case is
$\ruleSet{4}$. By definition of $l$, the derivation is the following:
%%
$$
\CAxiom{\pushN{\Set{i}}{4}{\Set{i+1}}}
$$
%%
Which translates to the desired judgement in (5):
%%
$$
\CAxiom{\pushN{\Set{i}}{5}{\Set{i+1}}, 0}
$$

The induction step consists in treating the cases $\ruleLam{4}$,
$\rulePi{4}$, and $\ruleConv{4}$ in turn. The treatment of the
$\ruleLam{4}$ and $\rulePi{4}$ rules is purely structural. Similarly,
the conversion rule follows by converting the pull-premise to itself
in (5) while the equality judgement carry over to the new system
unchanged. We then conclude by applying $\ruleConv{5}$ that directly
indicate a differential of zero. The $\ruleApp{4}$ rule follows
naturally, as the induction hypothesis on the premises give directly
the premises of the $\ruleApp{5}$ rule.

\end{proof}

%% *** <- Sound
%% **** \Gamma \vdash A \ni^5 a, k \wedge k \geq 0 \Rightarrow \Gamma \vdash A \ni^4 a
%% **** \Gamma \vdash a \in^5 A \Rightarrow \Gamma \vdash a \in^4 A
%% **** Theorem

\begin{theorem}[Soundness]

Let $\Gamma$ be a valid context.

\begin{itemize}
\item If \xspace$\Gamma \vdash \pushN{a}{5}{A}, k$ and $k \geq 0$, then $\Gamma \vdash \pushN{a}{4}{A}$
\item If \xspace$\Gamma \vdash \pullN{a}{5}{A}$, then $\Gamma \vdash \pullN{a}{4}{A}$
\end{itemize}

\end{theorem}

%% ***** Proof

\begin{proof}

The proof is by induction on the length of a derivation. The base
cases are $\ruleVar{5}$ and $\ruleSet{5}$. The $\ruleVar{}$ rule
translates directly to its counterpart. In the $\ruleSet{5}$ case, we
know that the differential $k$ is greater or equal to $0$. Therefore,
we can safely deduce:
%%
$$
\CAxiom{\pushN{\Set{i}}{4}{\Set{j}}}
$$

The induction step is trivial for most pull-side rules (they have
their equivalent in the target system), at the exception of the
$\ruleApp{5}$ rule. On the push-side, we have to treat the
$\ruleLam{5}$, $\ruleConv{5}$, and $\rulePi{5}$. The $\ruleLam{5}$ and
$\ruleConv{5}$ cases follow structurally. In the $\rulePi{5}$ case,
the hypothesis gives us that $\min(k_1, k_2) \geq 0$. We thus deduce
that $k_1 \geq 0$ and $k_2 \geq 0$ and can apply the induction
hypothesis on the premises and structurally obtain the desired
conclusion.

The application rule calls for some work. We have:
%%
$$
\RuleSide{\pullN{f}{5}{\greenb{\PI{\V{x}}{\greenb{\Up{j}{S}}{j}{k}}{\greenb{\Up{j}{T}}{j}{k}}}{j}{k}} \quad
          \pushN{s}{5}{\redb{\greenb{\Up{j}{S}}{j=0}{k}}}, l}
         {l' = \min(l,0)}
         {\pullN{f\: s}{5}{\greenb{\Up{j}{T}[\blackb{s}/x]}{j}{k-l'}}}
$$
%%
By induction hypothesis, we deduce that
$\pullN{f}{4}{\greenb{\PI{\V{x}}{\greenb{\Up{j}{S}}{j}{k}}{\greenb{\Up{j}{T}}{j}{k}}}{j}{k}}$. By
Lemma~\ref{lemma:push-k-good-enough}, we also have that
$\pushN{s}{5}{\Up{-l'}{\redb{\greenb{\Up{j}{S}}{j=0}{k}}}}$. By
Lemma~\ref{lemma:lifting-var-assgnmt}, we deduce that
$\pushN{s}{5}{\redb{\greenb{\Up{j}{S}}{j=-l'}{k}}}$. All conditions
are met to apply the $\ruleApp{4}$ rule.

\end{proof}

%% *** <- Complete
%% **** \Gamma \vdash A \ni^4 a \Rightarrow \Gamma \vdash A \ni^5 a, k \wedge k \geq 0
%% **** \Gamma \vdash a \in^4 A \Rightarrow \Gamma \vdash a \in^5 A
%% **** Theorem

\begin{theorem}[Completeness]

Let $\Gamma$ be a valid context.

\begin{itemize}
\item If \xspace$\Gamma \vdash \pushN{a}{4}{A}$, then $\Gamma \vdash \pushN{a}{5}{A}, k$ with $k \geq 0$
\item If \xspace$\Gamma \vdash \pullN{a}{4}{A}$, then $\Gamma \vdash \pullN{a}{5}{A}$
\end{itemize}

\end{theorem}

%% ***** Proof

\begin{proof}

Most pull-side rules are identical on the source and target side, at
the exception of the application rule. We will therefore ignore them,
assuming the naive translation.

The proof is by induction on the height of derivations. The base case
of interest is $\ruleSet{4}$. The conclusion is easily verified from
the requirement on levels.

The induction step is unsurprisingly structural for the $\rulePi{4}$,
$\ruleLam{4}$, and $\ruleConv{4}$ rules. The $\ruleApp{4}$ rule tells
us that:
%%
$$
\RuleSide{\pullN{f}{4}{\greenb{\PI{\V{x}}{\greenb{\Up{j}{S}}{j}{k}}{\greenb{\Up{T}{j}}{j}{k}}}{j}{k}} \quad
          \pushN{s}{4}{\redb{\greenb{\Up{j}{S}}{j=l}{k}}}}
         {k \in \mathbb{N} \quad \ruleApp{4}}
         {\pullN{f\: s}{4}{\greenb{\Up{j}{T}[s/x]}{j}{k+l}}}
$$
%%
By induction hypothesis, we can readily conclude that
$\pullN{f}{5}{\greenb{\PI{\V{x}}{\greenb{\Up{j}{S}}{j}{k}}{\greenb{\Up{T}{j}}{j}{k}}}{j}{k}}$. By
Lemma~\ref{lemma:push-k-tight}, we deduce that
$\pushN{s}{5}{\redb{\greenb{\Up{j}{S}}{j=0}{k}}}, -l$. Therefore, we
can apply $\ruleApp{5}$ and obtain the expected conclusion:
$\pullN{f\: s}{5}{\greenb{\Up{j}{T}[s/x]}{j}{k+l}}$.

\end{proof}

%% * Polymorphic definitions
%% ** <- Definitions: references
%% *** <- fully lambda-lifted representation
%% **** <- ???
%% *** -> Used through the Var rule
%% **** -> Automatically polymorphic
%% ** <- Examples
%% *** id
%% *** typeof
%% *** typeof/id (Eelis)
%% *** subtyping
%% ** <- Counter-examples (leading to paradoxes)
%% *** Define set : Set, typecheck set : set
%% * With data-types
%% ** <- Based on Levitation
%% *** <- (Brief sum up of levitation)
%% *** /> Work just as well with external presentation
%% **** /> More challenging in our case
%% ** <- Assign levels to things of Desc
%% *** <- data IDesc (I : Set1) : Set1
%% *** <- desc : {I : Set1} -> IDesc I -> (I -> Set0) -> Set0
%% *** <- data IMu {I : Set1}(R : I -> IDesc I)(i : I) : Set0
%% *** <- box : {I : Set1}(D : IDesc I)(P : I -> Set0) -> desc D P -> IDesc (Sigma I P)
%% *** <- induction : (...)
%% *** -> Remark that Mu decreases
%% ** <- Example: Nat : Set0
%% *** <- Canonical object
%% *** -> Get automatically Nat at all levels
%% ** <- Example: List : Set0 -> Set0
%% *** <- Parameterized object
%% *** -> Get List translating with its argument
%% *** -> [ Set0, Set1, Set2 ] : List Set3
%% ** <- Example: Vector : (n : N) -> Set0 -> Set0
%% *** <- Indexed object
%% *** -> Index, canonical, moves as well
%% ** <- Example: IDesc in IDesc
%% *** <- Levitation trick
%% *** -> Fall back to principal type!

%% * Related Work
%% ** Palmgren
%% ** Harper-Pollack
%% ** Courant
%% ** Agda experimental universe polymorphism
%% * Conclusion
\section{Conclusion}

%% ** We did
%% ** Future work

%% * Bibliography

\bibliography{stratisfaction}
\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.
%\begin{thebibliography}{}
%\softraggedright
%\end{thebibliography}

%% * Outro

\end{document}

%% Local Variables:
%% mode: outline-minor
%% outline-regexp: "%% [*\f]+"
%% outline-level: outline-level
%% End:

### 2010-07-12 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Explain and give ref. for mutually recursive datatypes
Desc@ICFP: pattern functors give rise to SPT
Reviewer 3 said:

"The paper does not discuss positivity and termination issues. As the
universe (seems to) describe only strictly positive datatypes, no
consistency issue can arise and all the recursive definitions are
structural (e.g. All(I) all(I)). An acknowledgment of this in the
paper would be good."

We said:

"The functors so computed are strictly positive by
construction---referees 23B and 23C are right that we should emphasize
this fact."

@@
   transformation} The code of $\SYMBIDesc$ is presented in
-Figure~\ref{fig:idesc}. Induction on indexed descriptions is defined
-by:
+Figure~\ref{fig:idesc}. By construction, $\SYMBIDesc$ can only
+describe \emph{strictly positive families}. Induction on indexed
+descriptions is defined by:
 %
@@
 types. Moreover, we gain the ability to write mutually recursive
-inductive types.
+inductive types by indexing with the choice of
+datatypes~\cite{paulin:habilitation, yakushev:mutual-def}.
 

### 2010-07-12 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: justify the adoption of a non-algebraic code for Desc
Reviewer 2 says:

"Fig. 3.  I couldn't figure out why 'indx takes a description argument
and is defined as X x [[D]] X instead of just being a constant
interpreted as X.  (Indeed, this is apparently done in the
implementation, and 'hindx is apparently replaced by pi).Similarly, in
sec. 4.4, it took me a second to see why you embed X as 'Sigma X '1 -
but presumably you could have a 'const description that embeds an
arbitrary X : Set into a Desc (just as you do for IDesc)."

And we said:

"Referee 23B also suggests that we use the

@@
 `pattern functor' on $\Set$, with $\SYMBdescop{\V{D}}$ its action on
-an object, \(\V{X}\), soon to be instantiated recursively.
+an object, \(\V{X}\), soon to be instantiated recursively. Note that
+the functors so described are \emph{strictly positive}, by
+construction.
 
-Descriptions are sequential structures, terminated by $\DUnit$, indicating
-the empty tuple. To build
-sigmas-of-sigmas, we define a $\SYMBDSigma$ code, interpreted as a
-$\Sigma$-type. To request a recursive component, we have $\DIndx{\V{D}}$,
-where \(\V{D}\) describes the rest of the node.
+Descriptions are sequential structures, terminated by $\DUnit$,
+indicating the empty tuple. To build sigmas-of-sigmas, we define a
+$\SYMBDSigma$ code, interpreted as a $\Sigma$-type. To request a
+recursive component, we have $\DIndx{\V{D}}$, where \(\V{D}\)
+describes the rest of the node. These codes directly captures the
+notion of sigmas-of-sigmas. An equivalent, more algebraic presentation
+could be given, as illustrated in Section~\ref{sec:indexing-desc}.
 
@@
 
+\newcommand{\SYMBind}{\F{ind}}
+
 However, iteration is inadequate for \emph{dependent} computation.
@@
 Here, $\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}$ states that
-$\Bhab{\V{P}}{\V{X} \To \Set}$ holds for every subobject $\V{x}:\V{X}$ in \(\V{D}\), and \(\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}\) is a
-`dependent map', applying
-some \(\Bhab{\V{p}}{\PIS{\Bhab{\x}{\V{X}}}\V{P}\,\V{x}}\) to each \(\x\)
-contained in \(\V{d}\). The full
-definition (including an extra case, introduced shortly) is presented in
-Figure~\ref{fig:all-predicates}. Note that
-\(\F{ind}\) is our first generic
-operation over descriptions, albeit a hardwired operator.
-Any datatype we define automatically
-comes with an induction principle.
+$\Bhab{\V{P}}{\V{X} \To \Set}$ holds for every subobject $\V{x}:\V{X}$
+in \(\V{D}\), and \(\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}\) is a
+`dependent map', applying some
+\(\Bhab{\V{p}}{\PIS{\Bhab{\x}{\V{X}}}\V{P}\,\V{x}}\) to each \(\x\)
+contained in \(\V{d}\). The full definition (including an extra case,
+introduced shortly) is presented in
+Figure~\ref{fig:all-predicates}.\footnote{To pass the termination
+  checker, we had to inline the definition of $\SYMBall$ into
+  $\SYMBind$ in our Agda model. A simulation argument shows that the
+  definition presented here terminates if the inlined version
+  does. Hence, although not directly structural, this definition is
+  indeed terminating.} Note that \(\F{ind}\) is our first generic
+operation over descriptions, albeit a hardwired operator.  Any
+datatype we define automatically comes with an induction principle.
 

### 2010-07-12 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix typo

@@
 representations are finitary by construction, and thus admit a richer,
-componentwise decidable equality than functions may in
+componentwise decidable equality that functions may in
 general possess.\footnote{E.g., extensionally, there is one inhabitant of

### 2010-07-12 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix typo

@@
 $\SYMBVar$ indexes a value in the context, obtaining a proof that the
-types matches. The $\SYMBlookup$ function precisely follow this
+types match. The $\SYMBlookup$ function precisely follow this
 semantics:

### 2010-07-12 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: explain subscript convention
Reviewer 2 says:

"p8.  Starting around page 8, some new notation is used where "(I:Set)
->" is written subscripted.  I guess this means "don't worry about
writing the index since we can infer it" but the notation is never
explained."
Desc@ICFP: uniformise notation for detagging and parenthesing
Reviewer 2 said:

"p5.  The mapping TagDesc to Desc, "de -", is replaced by "overline"
notation for indexed dataypes.  Neither notation helps me remember
what they mean - why not call it "TagDescToDesc"?"

'TagDescToDesc' is just impossible for space reason, but

@@
 
-We use the notation $\toIDesc{\V{TID}}$ to denote the
-indexed description computed from the tagged indexed description
-$\V{TID}$. Its expansion is similar to the definition of \(\SYMBtoDesc\)
-but more involved.
+We use the notation $\toIDesc{\V{\CN{TID}}}$ to denote the indexed
+description computed from the tagged indexed description
+$\V{TID}$. Its expansion is similar to the definition of
+\(\SYMBtoDesc\) for tagged descriptions.
 
@@
 \evalH : \PI{\V{ty}}{\Ty} 
-         \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
+         \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}} \To
          \Val{\V{ty}}
@@
 \evalH : \PI{\V{ty}}{\Ty} 
-           \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
+           \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}} \To
            \Val{\V{ty}} \\
@@
              & (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
-               \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} 
+               \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}} 
 \end{array} \\
@@
                                 (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
-                                \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}} \\
-          & \To \IMu{\Ty}{\toIDesc{\HExprVarD{\V{G}}}}{\V{ty}} \To 
-            \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
+                                \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}}} \\
+          & \To \IMu{\Ty}{(\toIDesc{\HExprVarD{\V{G}}})}{\V{ty}} \To 
+            \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}}
 \end{array}
@@
 \qquad    \cataI\: \toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}}\:
-                      (\IMu{}{\toIDesc{\FreeIMonad{\V{R}}{}{\V{Y}}}})\:
+                      (\IMu{}{(\toIDesc{\FreeIMonad{\V{R}}{}{\V{Y}}})})\:
                       (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
@@
             \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
-          & (\PI{\V{i}}{\V{I}} \V{X}\: \V{i} \To \IMu{\V{I}}{\toIDesc{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}}{\V{i}}) \To \\
+          & (\PI{\V{i}}{\V{I}} \V{X}\: \V{i} \To \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}) \To \\
           & \PI{\V{i}}{\V{I}} 
-            \idescop{\toIDesc{(\FreeIMonad{\V{R}}
+            \idescop{\toIDesc{\FreeIMonad{\V{R}}
                                          {\V{I}}
-                                         {\V{X}})}\: 
+                                         {\V{X}}}\: 
                      \V{i}}
                     {\V{I}}
-{\IMu{\V{I}}{\toIDesc{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}}} \To 
-           \IMu{\V{I}}{\toIDesc{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}}{\V{i}}
+{\IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}} \To 
+           \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}
 \end{array} \\
@@
 
-Let us now consider two examples of free indexed monads.
+The subscripted types corresponds to implicit arguments that can be
+automatically inferred, hence do not have to be typed in. Let us now
+consider two examples of free indexed monads.
 

### 2010-07-12 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: strive for space
Desc@ICFP: overload mu (de t) as mu t (elaboration problem).
Reviewer 2 said:

"Also, these functions are often used in expressions of the form mu
(de (D* X)) - perhaps an abbreviation such as

Fix D X == mu (de (D* X))

could help with this."

I think we could even ignore "de" inside "mu", turning this into an
elaboration problem. By the rule of "mu t", Desc is pushed in "t". If
"t" is "D * X", we would detag it first.

@@
 
-We use the notation $\toIDesc{\V{\CN{TID}}}$ to denote the indexed
-description computed from the tagged indexed description
-$\V{TID}$. Its expansion is similar to the definition of
-\(\SYMBtoDesc\) for tagged descriptions.
+We use the notation $\toIDesc{\V{D}}$ to denote the indexed
+description computed from the tagged indexed description $\V{D}$. Its
+expansion is similar to the definition of \(\SYMBtoDesc\) for tagged
+descriptions. For brevity, we shall also overload $\IMu{ty}{D}$ for
+$\IMu{ty}{(\toIDesc{D})}$, as the bidirectional presentation let us
+resolve the ambiguity.
 
@@
 \evalH : \PI{\V{ty}}{\Ty} 
-         \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}} \To
+         \IMu{\Ty}{\HExprD}{\V{ty}} \To
          \Val{\V{ty}}
@@
 \evalH : \PI{\V{ty}}{\Ty} 
-           \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}} \To
+           \IMu{\Ty}{\HExprD}{\V{ty}} \To
            \Val{\V{ty}} \\
@@
              & (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
-               \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}} 
+               \IMu{\Ty}{\HExprD}{\V{ty}} 
 \end{array} \\
@@
              \PI{\V{G}}{\Ctxt} \\
-           & \PITEL{\V{\sigma}}{\PI{\V{ty}}{\Ty}
+           & (\PI{\V{ty}}{\Ty}
               (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
-                                \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}}} \\
-          & \To \IMu{\Ty}{(\toIDesc{\HExprVarD{\V{G}}})}{\V{ty}} \To 
-            \IMu{\Ty}{(\toIDesc{\HExprD})}{\V{ty}}
+              \IMu{\Ty}{\HExprD}{\V{ty}}) \\
+          & \To \IMu{\Ty}{\HExprVarD{\V{G}}}{\V{ty}} \To 
+            \IMu{\Ty}{\HExprD}{\V{ty}}
 \end{array}
@@
           & (\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 
-                               \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}) \To \\
+                               \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}) \To \\
           & \PI{\V{i}}{\V{I}}
-            \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{X}}})}{\V{i}} \To
-            \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}
+            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})}{\V{i}} \To
+            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}
 \end{array} \\
@@
 \qquad    \cataI\: \toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}}\:
-                      (\IMu{}{(\toIDesc{\FreeIMonad{\V{R}}{}{\V{Y}}})})\:
+                      (\IMu{}{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
                       (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
@@
             \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
-          & (\PI{\V{i}}{\V{I}} \V{X}\: \V{i} \To \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}) \To \\
+          & (\PI{\V{i}}{\V{I}} \V{X}\: \V{i} \To \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}) \To \\
           & \PI{\V{i}}{\V{I}} 
@@
                     {\V{I}}
-{\IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}} \To 
-           \IMu{\V{I}}{(\toIDesc{\FreeIMonad{\V{R}}{\V{I}}{\V{Y}}})}{\V{i}}
+{\IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \To 
+           \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}
 \end{array} \\

### 2010-07-12 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: explain the coding of Hutton's razor on the Indexed part
Reviewer 2 said:

"p9, fig. 7: A reminder near ExprID that the order of the sublists of
IndexedD needs to line up with the definition of Ty would be helpful."
Desc@ICFP: use dotted arrow for (i : I) -> X i -> Y i
Reviewer 2 said:

"p10.  In the types of applyI and substI, maybe you could employ an
abbreviation

IFun X Y = (i:I) -> X i -> Y i
IFix R X =  mu_I \overline{R^*_I X}

Then the type of, say, substI becomes the much more readable:

(I : Set)(R : TagDesc I)(X,Y : I -> Set) ->
(IFun X (IFix R Y)) -> (IFun (IFix R X) (IFi

@@
 the type: the resulting description is indexed by $\Ty$. We observe
-that some constructors are always defined, namely $\SYMBEcond$
-and $\SYMBEval$. On the other hand, the $\SYMBEplus$ and
-$\SYMBEle$ constructors are index-dependent. $\SYMBEplus$ is
-defined if and only if the result type -- the index -- is $\Enat$,
-whereas $\SYMBEle$ is defined if and only if the index is
-$\Ebool$. The actual code precisely follows this intuition, as shown
-in Figure~\ref{fig:hexpr-full}.
+that some constructors are always defined, namely $\SYMBEcond$ and
+$\SYMBEval$. On the other hand, the $\SYMBEplus$ and $\SYMBEle$
+constructors are index-dependent. $\SYMBEplus$ is defined if and only
+if the result type -- the index -- is $\Enat$, whereas $\SYMBEle$ is
+defined if and only if the index is $\Ebool$. The actual code
+precisely follows this intuition, as shown in
+Figure~\ref{fig:hexpr-full}. The first element of $\HExprID$
+corresponds to this choice of constructor: the sublists corresponds to
+the constructors enabled by the corresponding type in $\Ty$.
 
@@
 
+\newcommand{\DotTo}{\mathop{\blue{\dot{\rightarrow}}}}
+
 The $\discharge$ function is relative to the required type and a
 context of the right type. Its action is to map values to themself,
-and variables to their value in context. This corresponds to the
-following function:
+and variables to their value in context. The pointwise indexed
+function type $\PI{\V{i}}{I}{X\: i \To Y\: i}$ is abbreviated  $X
+\DotTo Y$. This corresponds to the following function:
 %
@@
 \discharge : & \PI{\V{G}}{\Ctxt}
-               \Env{G} \To
-               \PITEL{\V{ty}}{\Ty} \\
-             & (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
-               \IMu{\Ty}{\HExprD}{\V{ty}} 
+               \Env{G} \To \\
+             & (\SUM{\Val{\V{}}}{\Var{\V{G}}{}}) \DotTo
+               \IMu{\Ty}{\HExprD}{}
 \end{array} \\
@@
 \begin{array}{@{}ll}
-\substH  : & \PITEL{\V{ty}}{\Ty}
-             \PI{\V{G}}{\Ctxt} \\
-           & (\PI{\V{ty}}{\Ty}
-              (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
-              \IMu{\Ty}{\HExprD}{\V{ty}}) \\
-          & \To \IMu{\Ty}{\HExprVarD{\V{G}}}{\V{ty}} \To 
-            \IMu{\Ty}{\HExprD}{\V{ty}}
+\substH  : & \PI{\V{G}}{\Ctxt} \\
+           & ((\SUM{\Val{}}{\Var{\V{G}}{}}) \DotTo
+              \IMu{\Ty}{\HExprD}{}) \DotTo \\
+          & \IMu{\Ty}{\HExprVarD{\V{G}}}{} \DotTo 
+            \IMu{\Ty}{\HExprD}{}
 \end{array}
@@
             \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
-          & (\PI{\V{i}}{\V{I}} \V{X}\:\V{i} \To 
-                               \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}) \To \\
-          & \PI{\V{i}}{\V{I}}
-            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})}{\V{i}} \To
-            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}
+          & (\V{X} \DotTo 
+             \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To 
+            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})}{} \DotTo
+            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
 \end{array} \\
@@
             \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
-          & (\PI{\V{i}}{\V{I}} \V{X}\: \V{i} \To \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}) \To \\
-          & \PI{\V{i}}{\V{I}} 
-            \idescop{\toIDesc{\FreeIMonad{\V{R}}
+          & (\V{X} \DotTo \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
+          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
                                          {\V{I}}
-                                         {\V{X}}}\: 
-                     \V{i}}
+                                         {\V{X}}}}
                     {\V{I}}
-{\IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \To 
-           \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{\V{i}}
+                    {\IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
+            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
 \end{array} \\
@@
 \[\stk{
-\substH\: \V{ty}\:
-          \V{G}\:
+\substH\: \V{G}\:
+          \V{ty}\:          
           \V{g}\:
@@
 presented the free monad construction, together with its monadic
-operations.
+operations. A summary of the components defining descriptions and
+their indexed counterparts is presented in
+Table~\ref{tab:sumup-operators}.
+
+\begin{table}
+
+{\small
+
+\begin{tabular}{|c|c|c|c|}
+\hline
+Desc   & IDesc & Role & Status \\
+\hline
+\hline
+\multicolumn{2}{|c|}{$\EnumU$}                  & Build finite sets           & Levitated \\
+\hline
+$\Desc$               & $\IDesc$               & Describe functors   & Levitated \\
+\hline
+$\SYMBdescop{\_}$         & $\SYMBidescop{\_}{\_}$         & Interpret descriptions      & Hardwired \\
+\hline
+$\SYMBMu$, $\SYMBCon$ & $\SYMBIMu_I$, $\SYMBCon$ & Define fixpoints & Hardwired \\
+\hline
+$\SYMBind$, $\SYMBAll$, $\SYMBall$ & $\inductionI$, $\SYMBAllI$, $\SYMBallI$ & Induction principle & Hardwired \\
+\hline
+\end{tabular}
+
+}
+
+\caption{Summary of constructions on Descriptions}
+\label{tab:sumup-operators}
+
+\end{table}

### 2010-07-21 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: Fix typos.
Desc@ICFP: develop our position for Set : Set
Reviewer 1 says:

"However, the idea of having a type of data type of encodings which
contains its own code is just too close to having a type of all types
to be taken lightly. It doesn't help that the authors "locally adopt
SET : SET, putting presentation before paradox", since this is related
to the key theoretical issue of the paper."

To which we said:

"As presented with Set:Set, our system is clearly open to
nontermination via paradox.  Choosing a particular stratification
discipline is an option but would introduce extr

@@
 for tuples and functions are also standard, save that we locally adopt
-\(\Set:\Set\), putting presentation before paradox~\cite{girard:set-in-set}.
-The usual remedies apply, \emph{stratifying}
-\(\Set\)~\cite{harper:implicit-universe, luo:utt,
-  courant:explicit-universe}.
+\(\Set:\Set\) for presentational purposes. Usual techniques to resolve
+this \emph{typical ambiguity} apply~\cite{harper:implicit-universe,
+  luo:utt, courant:explicit-universe}. A formal presentation of a
+stratified system is beyond the scope of this paper and is subject of
+ongoing work.
+%%  putting presentation before paradox~\cite{girard:set-in-set}.
+%% The usual remedies apply, \emph{stratifying}
+%% \(\Set\)~\cite{harper:implicit-universe, luo:utt,
+%%   courant:explicit-universe}.
+%%
 \input{figure_typing_judgements}
@@
 Our type propagation mechanism does no constraint
-solving, just copying, so it just the thin end of the elaboration wedge.
+solving, just copying, so it is just the thin end of the elaboration wedge.
 It can afford us this `assembly

### 2010-07-21 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: uniformise notation for detagging and parenthesing
Reviewer 2 said:

"p5.  The mapping TagDesc to Desc, "de -", is replaced by "overline"
notation for indexed dataypes.  Neither notation helps me remember
what they mean - why not call it "TagDescToDesc"?"

'TagDescToDesc' is just impossible for space reason, but at least we
are now uniform in our hieroglyphism.
Desc@ICFP: tentative: put level annotation in Section 4

@@
 We shall now fulfil our implementation promises, encoding the universe
-of descriptions. In and of itself, the codes, $\Desc$, is nothing but
-a datatype. We are in the same situation as with $\EnumU$: we ought to
-be able to describe the codes of $\Desc$ in $\Desc$ itself. Hence,
-this code would be a first-class citizen, born with the standard,
-generic equipment of datatypes.
+of descriptions. In and of itself, the codes, $\Desc_n$, is nothing
+but a datatype in $\Set_{n+1}$. We are in the same situation as with
+$\EnumU$: we ought to be able to describe the codes of $\Desc_n$ in
+$\Desc_{n+1}}$ itself. Hence, this code would be a first-class
+  citizen, born with the standard, generic equipment of datatypes.
 
@@
 \[\stk{
-\DescD : \Desc \\
-\DescD \mapsto
+\DescD_n : \Desc_{n+1} \\
+\DescD_n \mapsto
   \tada{\stkm{\DUnit\\ \SYMBDSigma\\ \SYMBDIndx}}
        {\stkm{ \DUnit                                \\
-         \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
+         \DSigma{\Set_{n}}{(\LAM{\V{S}} \SHED)}      \\
          \DIndx{\DUnit}    }                    } \\
@@
 \[
-\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
+\DSigma{(\Bhab{\V{S}}{\Set_n})}{(\Bhab{\V{D}}{\V{S} \To \Desc_n})} : \Desc_n\\
 \]
@@
 \[\stk{
-    \DHindx{(\Bhab{\M{H}}{\Set})}{(\Bhab{\M{D}}{\Desc})} : \Set \smallskip \\
-\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     \mapsto \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
+    \Bhab{\DHindx{(\Bhab{\M{H}}{\Set_n})}
+                 {(\Bhab{\M{D}}{\Desc_n})}}
+         {\Desc_n}
+\smallskip \\
+\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}} 
+    \mapsto 
+        \TIMES{(\V{H} \To \V{X})}
+              {\descop{\V{D}}{\V{X}}} \\
 }\]
@@
 \[\stk{
-\DescD : \Desc \\
-\DescD \mapsto\tada
+\DescD_n : \Desc_{n+1} \\
+\DescD_n \mapsto\tada
 {\stkm{ \DUnit\\ \SYMBDSigma\\ \SYMBDIndx\\ \SYMBDHindx}}
-{\stkm{
-\DUnit                                            \\
-                                   \DSigma{\Set}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
+{\stkm{ \DUnit  \\
+        \DSigma{\Set_n}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
         \DIndx{\DUnit}                                    \\
-                                   \DSigma{\Set}{\LAM{\_} \DIndx{\DUnit}}
+        \DSigma{\Set_n}{\LAM{\_} \DIndx{\DUnit}}
 }}
@@
 At a first glance, we have achieved our goal. We have described the
-codes of the universe of descriptions. Taking the fixpoint of \(\descop{\DescD}\)
-gives us a datatype exactly like $\Desc$. Might we be
-so bold as to take \(\Desc \mapsto \Mu{\DescD}\) as the levitating
+codes of the universe of descriptions. Taking the fixpoint of \(\descop{\DescD_n}\)
+gives us a datatype exactly like $\Desc_n$. Might we be
+so bold as to take \(\Desc_n \mapsto \Mu{\DescD_n}\) as the levitating
 definition? If we do, we shall come down with a bump! 
@@
 
-The definition \(\Desc \mapsto \Mu{\DescD}\) is circular,
+The definition \(\Desc_n \mapsto \Mu{\DescD_n}\) is circular,
 but the offensive recursion is concealed by a prestidigitation.
@@
 \[
-\Desc\mapsto
-\Mu{\stk{(\DSigma{\EnumT{\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}}\\}
-    {\;\SYMBswitch\:\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}\:(\LAM{\_}\Desc)\\
+\Desc_n \mapsto
+    \Mu{\stk{(
+        \DSigma{\EnumT{\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}}\\}
+               {\;\SYMBswitch\:\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}\:
+                                   (\LAM{\_}\Desc)\\
                              \;\sqr{\stkm{
                                    \DUnit \\
-                                   \DSigma{\Set}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
+                                   \DSigma{\Set_n}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}} \\
                                    \DIndx{\DUnit} \\
-                                   \DSigma{\Set}{\LAM{\_} \DIndx{\DUnit}}
-}}})}}
+                                   \DSigma{\Set_n}{\LAM{\_} \DIndx{\DUnit}}}}})}}
 \]
+%%
 The recursion shows up only because we must specify the return type
@@
 \Rule{\Gamma \Vdash
-  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{(\PLAM{x}{\EnumT{E}}\Desc)}}}
+  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{\PLAM{x}{\EnumT{E}}\Desc}}}
                            {\M{t'}}}
@@
 \(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
-\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\)
+\(\Bhab{\V{hs}}{\All{\V{D}}{\Mu{\V{D}}}{(\LAM{\_} \V{T})}{\V{xs}}}\)
 at hand. Our goal is to make an element of
@@
            \PI{\V{X}, \V{Y}}{\Set} 
-           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
-         & \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
-           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
+           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To \\
+         & \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
+           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
 \end{array} \\
 \subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
-  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
-          (\Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})})\: 
+  \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
+          \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}\: 
           \SHED

### 2010-07-21 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: (try to) be consistent on spacing, parenthesing, etc.
Reviewer 2 said:

"First, generally, parentheses and spacing are not used consistently,
and I had to guess what the precedence of various operations was too
often (or try to do the type inference in my head).  For example, D* X
is parenthesized in the type/definition of subst but not in apply (on
p. 7)."

I tried to be consistent. But Godel tells me that I cannot garantee my
own consistency, so someone will have to double check.
Desc@ICFP: fix spacing
Desc@ICFP: put level annotation in Section 3.
We said:

"At the very least, we

@@
 in Figure~\ref{fig:desc_universe}, with codes mimicking the grammar of
-datatype declarations. We can read a description $\V{D}:\Desc$ as a
-`pattern functor' on $\Set$, with $\SYMBdescop{\V{D}}$ its action on
-an object, \(\V{X}\), soon to be instantiated recursively. Note that
-the functors so described are \emph{strictly positive}, by
-construction.
+datatype declarations. We can read a description $\V{D}:\Desc^{n}$ as
+a `pattern functor' on $\Set^{n}$, with $\SYMBdescop{\V{D}}$ its
+action on an object, \(\V{X}\), soon to be instantiated
+recursively. The superscripts indicate the $\Set$-levels at which we
+\emph{expect} these objects in a stratified system. This is but an
+informal notation, to give a flavor of the stratified
+presentation. Note that the functors so described are \emph{strictly
+  positive}, by construction.
 
@@
 \begin{array}{l}
-\Desc : \Set\\
-\DUnit : \Desc\\
-\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
-\DIndx{(\Bhab{\V{D}}{\Desc})} : \Desc \\
+\Desc^n : \Set^{n+1}\\
+\DUnit : \Desc^{n}\\
+\DSigma{(\Bhab{\V{S}}{\Set^{n}})}{(\Bhab{\V{D}}{\V{S} \To \Desc^{n}})} : \Desc^{n}\\
+\DIndx{(\Bhab{\V{D}}{\Desc^{n}})} : \Desc^{n} \\
 \end{array}
@@
 \begin{array}{l@{\V{X}\:\mapsto\:}l}     
-\multicolumn{2}{l}{\descop{\_\:}{} : \Desc \To \Set \To \Set} \\
+\multicolumn{2}{l}{\descop{\_\:}{} : \Desc^{n} \To \Set^{n} \To \Set^{n}} \\
  \descop{\DUnit}{} &  \Unit \\
@@
 \[\stk{
-\NatD : \Desc \\
+\NatD : \Desc^n \\
 \NatD \mapsto \DSigma{\EnumT{\sqr{\NatZero\: \SYMBNatSuc }}}
@@
 \[\stk{
-\ListD : \Set \To \Desc \\
+\ListD : \Set^n \To \Desc^n \\
 \ListD \: \V{X} \mapsto
@@
 \[\stk{
-\TreeD : \Set \To \Desc \\
-\begin{array}{@{}ll}
-\TreeD \: \V{X} \mapsto \DSigma{ & \EnumT{\sqr{ \TreeLeaf\: \TreeNode }} \\}
+\TreeD : \Set^n \To \Desc^n \\
+\begin{array}{@{}l@{}l}
+\TreeD \: \V{X} \mapsto 
+    \DSigma{ & \EnumT{\sqr{ \TreeLeaf\: \TreeNode }} \\}
            { & \sqr{ \DUnit \quad
-       ( \DIndx{(\DSigma{\V{X}}{\LAM{\_}\DIndx{\DUnit}})})} }
+                     (\DIndx{(\DSigma{\V{X}}{\LAM{\_}\DIndx{\DUnit}})})} }
 \end{array}
@@
 \[\stk{
- \TagDesc : \Set \\
- \TagDesc \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{(\LAM{\_} \Desc)})
+ \TagDesc^n : \Set^{n+1} \\
+ \TagDesc^n \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{\LAM{\_} \Desc^n})
 \smallskip\\
-\SYMBtoDesc : \TagDesc\To\Desc \\
+\SYMBtoDesc : \TagDesc^n \To \Desc^n \\
 \SYMBtoDesc \mapsto
 \spl{\LAM{\V{E}}\LAM{\V{D}}
-\DSigma{\EnumT{\V{E}}}{(\F{switch}\:\V{E}\:(\LAM{\_}\Desc)\:\V{D})}}
+\DSigma{\EnumT{\V{E}}}{(\F{switch}\:\V{E}\:(\LAM{\_}\Desc^n)\:\V{D})}}
 }\]
@@
 \[
-\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc}}
-     {\Gamma \vdash \Bhab{\Mu{\M{D}}}{\Set}} \qquad
-\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc} \quad 
-      \Gamma \vdash \Bhab{\M{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}}
+\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc^n}}
+     {\Gamma \vdash \Bhab{\Mu{\M{D}}}{\Set^n}} \qquad
+\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc^n} \quad 
+      \Gamma \vdash \Bhab{\M{d}}{\descop{\M{D}}{\Mu{\M{D}}}}}
      {\Gamma \vdash \Bhab{\Con{\M{d}}}{\Mu{\M{D}}}}
@@
 \[\stk{
-\Nat\mapsto\Mu{(\toDesc{\pair{\sqr{\NatZero\: \SYMBNatSuc }}
-                       {\sqr{ \DUnit \quad (\DIndx{\DUnit})}}{}})}:\Set
+\Nat \mapsto \Bhab{\Mu{(\toDesc{\pair{\sqr{\NatZero\: \SYMBNatSuc }}
+                               {\sqr{\DUnit \quad (\DIndx{\DUnit})}}{}})}}
+                  {\Set^n}
 \\
@@
 \[
-\F{cata} : \PITEL{\V{D}}{\Desc}
-           \PI{\V{T}}{\Set}
+\F{cata} : \PITEL{\V{D}}{\Desc^n}
+           \PI{\V{T}}{\Set^n}
            (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
@@
 \[\stk{
-\F{ind} : \stk{ \PITEL{\V{D}}{\Desc}
-                    \PI{\V{P}}{\Mu{\V{D}} \To \Set}         \\
-               (\PI{\V{d}}{\descop{\V{D}}{(\Mu{\V{D}})}}       
-                \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{d}} \To \V{P} (\Con{\V{d}})) \To \\
+\F{ind} : \stk{ \PITEL{\V{D}}{\Desc^n}
+                \PI{\V{P}}{\Mu{\V{D}} \To \Set^k}         \\
+               (\PI{\V{d}}{\descop{\V{D}}{\Mu{\V{D}}}}       
+                \All{\V{D}}{\Mu{\V{D}}}{\V{P}}{\V{d}} \To \V{P} (\Con{\V{d}})) \To \\
                \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x} 
@@
 \F{ind}\, \V{D}\, \V{P}\, \V{m}\, (\Con{\V{d}}) = 
-    \V{m}\, \V{d}\, (\all{\V{D}}{(\Mu{\V{D}})}{\V{P}}
+    \V{m}\, \V{d}\, (\all{\V{D}}{\Mu{\V{D}}}{\V{P}}
                            {(\F{ind}\, \V{D}\, \V{P}\, \V{m})}
@@
 Here, $\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}$ states that
-$\Bhab{\V{P}}{\V{X} \To \Set}$ holds for every subobject $\V{x}:\V{X}$
+$\Bhab{\V{P}}{\V{X} \To \Set^k}$ holds for every subobject $\V{x}:\V{X}$
 in \(\V{D}\), and \(\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}\) is a
@@
   does. Hence, although not directly structural, this definition is
-  indeed terminating.} Note that \(\F{ind}\) is our first generic
+  indeed terminating.} Note that $\SYMBind$ is our first generic
 operation over descriptions, albeit a hardwired operator.  Any
@@
 \begin{array}{@{}ll}
-\SYMBAll : & \PITEL{\V{D}}{\Desc}
-             \PITEL{\V{X}}{\Set}
-             \PITEL{\V{P}}{\V{X} \To \Set} \\
+\SYMBAll : & \PITEL{\V{D}}{\Desc^n}
+             \PITEL{\V{X}}{\Set^n}
+             \PITEL{\V{P}}{\V{X} \To \Set^k} \\
            & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
-             \Set 
+             \Set^k 
 \end{array} \\
@@
 \begin{array}{@{}ll}
-\SYMBall : & \PITEL{\V{D}}{\Desc}
-             \PITEL{\V{X}}{\Set}
-             \PITEL{\V{P}}{\V{X} \To \Set} \\
+\SYMBall : & \PITEL{\V{D}}{\Desc^n}
+             \PITEL{\V{X}}{\Set^n}
+             \PITEL{\V{P}}{\V{X} \To \Set^k} \\
            & \PITEL{\V{p}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
@@
   \push{\sqr{\vec{t}}}
-   {\descop{\M{D}\: n}{(\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})})}}}
+   {\descop{\M{D}\: n}{\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})}}}}
         {\M{t'}}}
@@
 \propag{\push{\M{f}}
-             {\PI{\V{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}       
-                \All{\M{D}}{(\Mu{\M{D}})}{(\PLAM{\x}{\Mu{\M{D}}}\M{P})}{\V{d}} \To \M{P}[\Con{\V{d}}/\x]\\}}
+             {\PI{\V{d}}{\descop{\M{D}}{\Mu{\M{D}}}}       
+                \All{\M{D}}{\Mu{\M{D}}}{(\PLAM{\x}{\Mu{\M{D}}}\M{P})}{\V{d}} \To \M{P}[\Con{\V{d}}/\x]\\}}
        {\M{f'}}}}

### 2010-07-21 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: rollback level annotation in Section 4
Putting levels there is more confusing than anything. We are actually
collapsing the hierarchy, so the sudden mess up of indexes is not of
good effect.

rolling back:

Wed Jul 21 15:21:59 BST 2010  Pierre-Evariste Dagand <Pierre-Evariste.Dagand@cis.strath.ac.uk>
  * Desc@ICFP: tentative: put level annotation in Section 4

    M ./papers/icfp-2010-desc/paper_desc_levitation.tex -31 +38
Desc@ICFP: Skyhooks-all-the-way-up subsection
Reviewer 1 said:

"An implementation (available on the web) in Agda using Agda's
universe polymorphism is also insuf

--- 2010-07-12 papers/icfp-2010-desc/paper_desc_levitation.tex
@@
 %% functions.
+
+\subsection{Skyhooks all the way up?}
+
+%% * Desc in Desc
+%% ** /> Blur the lines between implementation and reflection
+%% ** /> Abusing the paradoxical nature of Set : Set
+
+In this section, we have seen how to \emph{levitate}
+descriptions. From this unusual situation, several questions naturally
+arise. First, we have blurred the line between the implementation of
+data-types and their reflection in themselves. Second, we have abused
+the paradoxical nature of $\Bhab{\Set}{\Set}$ to flatten the hierarchy
+of descriptions. We shall now be more precise about these points.
+
+%% * Summary 
+%% ** <- What is the equipment for making data-types
+%% ** <- What is reflected, what is implemented
+%% ** <- (table 1)
+
+Let us first clarify the status of the implementation. The kit for
+making datatypes is summed up in Table~\ref{tab:sumup-operators}. For
+each operation, we describe its role and its status, making clear
+which components are self-described and which ones are actually
+implemented.
+
+%% * Paradox
+%% ** <- How do we bottom out in a stratified setting?
+%% ** <- Spiral
+%% ***    <- Encoding of DescD_n : Desc_{n+1}
+%% ***    <- Desc_n = Mu DescD_n : Set_{n+1}
+%% ** <- Agda model: Desc_42
+%% ***    <- No dependent pattern-matching
+%% ***    <- No IR
+%% ***    <- Normal universes
+%% ***    -> Straight Agda (UTT)
+%% ** <- ``Self-encoding'' only in a level polymorphic sense
+%% ***    -> Agda model: set poly
+
+To tackle the apparent circularity of our construction, let us
+consider how it bottoms out in a stratified setting. Using the
+informal level annotations of Section~\ref{sec:universe-desc}, we have
+that \(\Bhab{\DescD_n}{\Desc_{n+1}}\), hence \(\Bhab{\Desc_n \mapsto
+  \Mu{\DescD_n}}{\Set_{n+1}}\). Therefore, there is no circularity in
+the description of descriptions: assuming an hardwired $\Desc_N$
+universe, we can build the tower of descriptions down to
+$\Desc_0$\footnote{We have modelled this process in Agda, hardwiring
+  $\Desc_{42}$ in $\Set_{43}$, then going down. This model does not
+  rely on dependent pattern-matching, induction-recursion, nor set
+  polymorphism.}. Hence, in a stratified system, the `self-encoded'
+nature of $\Desc$ appears only in a set polymorphic sense: the
+principal type of the levitating description generalizes to the type
+of $\Desc$ itself\footnote{This is witnessed in our set polymorphic
+  model in Agda. Set polymorphism is an experimental feature of
+  Agda. Although not backed by theoretical works, it is a conservative
+  extension of Agda that is generally accepted as free of paradox.}.
+
+\begin{table}
+
+{\small
+
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+\hline
+Object                & Role                        & Status \\
+\hline
+\hline
+$\EnumU$              & Build finite sets           & Levitated \\
+\hline
+$\Desc$               & Describe pattern functors   & Levitated \\
+\hline
+$\SYMBdescop{\_}$     & Interpret descriptions      & Hardwired \\
+\hline
+$\SYMBMu$, $\SYMBCon$ & Define, inhabit fixpoints   & Hardwired \\
+\hline
+$\SYMBind$, $\SYMBAll$, $\SYMBall$  
+                      & Induction principle         & Hardwired \\
+\hline
+\end{tabular}
+\end{center}
+}
+
+\caption{Summary of constructions on Descriptions}
+\label{tab:sumup-operators}
+
+\end{table}

### 2010-07-21 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: Skyhooks-all-the-way-up subsection
Reviewer 1 said:

"An implementation (available on the web) in Agda using Agda's
universe polymorphism is also insufficient. Agda's universe
polymorphism is itself an ad hoc feature which needs further
theoretical analysis."

And:

"The final sentence of the abstract is "Surprisingly, this apparently
self-supporting setup is achievable without paradox or infinite
regress:" How do we know this? On the other hand, in section 6.1 the
authors mention for the universe IDesc (for indexed data types) that
it "stratifies naturally". This I can believe, but

@@
 \label{sec:discussion}
-
-\subsection{Universe stratification}
-
-\begin{wstructure}
-!!! Need Help !!!
-<- Universe stratification
-    <- Stratified agda model
-        <- Fully stratified
-        <- Proof of iso between host and embedding
-    ???
-\end{wstructure}
-
-As such, our type theory suffers from an inconsistency. Indeed, the
-typing rule $\Bhab{\Set}{\Set}$ leads to Girard's paradox. We made
-that choice for presentational convenience, as universe stratification
-is orthogonal to our work. Nonetheless, our universe of description
-stratifies naturally. $\SYMBIDesc$ is self-encoding only in a
-level-polymorphic sense.  Unsurprisingly, $\SYMBIDesc$ at level $l$ is
-of type $\Set^{\blue{l+1}}$. Similarly, the interpretation of
-$\SYMBIDesc$ at level $l$ is an object of type $\Set^{\blue{l}}$:
-%
-\[\stk{\begin{array}{@{}ll}
-\SYMBIDesc^{\blue{l}} (\Bhab{\M{I}}{\Set^{\blue{l+1}}}) & :\Set^{\blue{l+1}} \\
-\DVar{(\Bhab{\M{i}}{\M{I}})} &: \SYMBIDesc^{\blue{l}}{\M{I}} \\
-\DConst{(\Bhab{\M{A}}{\Set^{\blue{l}}})} &:\SYMBIDesc^{\blue{l}}{\M{I}}       \\
-\DProd{(\Bhab{\M{D}}{\SYMBIDesc^{\blue{l}}{\M{I}}})}{(\Bhab{\M{D}}{\SYMBIDesc^{\blue{l}}{\M{I}}})}
-  & :\SYMBIDesc^{\blue{l}}{\M{I}}       \\
-\DSigma{(\Bhab{\M{S}}{\Set^{\blue{l}}})}{(\Bhab{\M{D}}{\M{S}\To\SYMBIDesc^{\blue{l}}{\M{I}}})}
-& : \SYMBIDesc^{\blue{l}}{\M{I}}  \\
-\DPi{(\Bhab{\M{S}}{\Set^{\blue{l}}})}{(\Bhab{\M{D}}{\M{S}\To\SYMBIDesc^{\blue{l}}{\M{I}}})}
-& : \SYMBIDesc^{\blue{l}}{\M{I}}  \\
-\end{array}\smallskip \\
-\idescop{\_\:}{}{}^{\blue{l}} : _{\PI{\V{I}}{\Set^{\blue{l+1}}}} \SYMBIDesc^{\blue{l}}\:\V{I} \To (\V{I} \To \Set^{\blue{l}}) \To \Set^{\blue{l}}    \\
-\ldots
-}
-\]
-Crucially, the types of data stored in an
-\(\SYMBIDesc^{\blue{l}}{\M{I}}\) all live no higher---we may store an
-\(I\) and a \(\Set^{\blue{l}}\) in a \(\Set^{\blue{l+1}}\).  The code
-for \(\SYMBIDesc^{\blue{l}}{\M{I}}\) is an element of
-\(\SYMBIDesc^{\blue{l+1}}{\Unit}\), so there is a spiral, not a cycle.
-We have checked the construction using Agda's universe polymorphism,
-coding $\SYMBIDesc$ in itself and have proved the isomorphism
-between the host and the embedded universes.
 

### 2010-07-21 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: (try to) be consistent on spacing, parenthesing, etc.
Reviewer 2 said:

"First, generally, parentheses and spacing are not used consistently,
and I had to guess what the precedence of various operations was too
often (or try to do the type inference in my head).  For example, D* X
is parenthesized in the type/definition of subst but not in apply (on
p. 7)."

I tried to be consistent. But Godel tells me that I cannot garantee my
own consistency, so someone will have to double check.
Desc@ICFP: Skyhooks-all-the-way-up subsection
Reviewer 1 said:

"An implementation (available on the w

@@
 
-So far, we have explored the well-known realm of inductive types. We
-have built upon our intuition of ML-like datatypes. In our dependent
-setting, we have provided these datatypes by the mean of $\Desc$, a
-universe of descriptions.
-
-Working with dependent types fosters new opportunities for
-datatypes. The typical example is bounded lists, also known as
-vectors. A vector is a list decorated by its length. Having this
-information prevents hazardous operations, such as taking the head of
-an empty vector: the head function only takes vectors of length
-$\NatSuc{\V{n}}$, as enforced by its type. This is made possible by
-the specificity of dependent types: a term -- the length -- can
-influence a type -- the vector type.
-
-However, these datatypes cannot be defined by mere induction. In the
-case of vectors, for instance, we have to define the whole
-\emph{family} of vectors in one go: vectors of all sizes need to be
-defined at the same time. In dependently-typed languages, the basic
-grammar of datatypes is that of inductive families. To capture this
-grammar, we rely on \emph{indexing}.
+So far, we have explored the realm of inductive types, building on
+intuition from ML-like datatypes, using type dependency as a
+descriptive tool in $\Desc$ and its interpretation. Let us now make
+dependent types the object as well as the means of our study.
+
+Dependent datatypes provide a way to work at higher level of
+precision \emph{a priori}, reducing the sources of failure we
+might otherwise need to manage. For the perennial
+example, consider \emph{vectors}---lists indexed by length. By
+making length explicit in the type, we can prevent hazardous
+operations (the type of `head' demands vectors of length
+$\NatSuc{\V{n}}$) and offer stronger guarantees (pointwise
+addition of $\V{n}$-vectors yields an $\V{n}$-vector).
+
+However, these datatypes are not \emph{individually} inductive. For
+instance, we have to define the whole \emph{family} of vectors
+mutually, in one go. In dependently typed languages, the basic grammar
+of datatypes is that of inductive families. To capture this grammar,
+we must account for \emph{indexing}.
 
@@
 
-In the previous section, we have presented the $\Desc$ universe as a
-grammar of functors in the category $\Set$. We have seen how to code
-inductive types in this setting. To describe an inductive family
-indexed by $\Bhab{\V{I}}{\Set}$, we use endofunctors on the category
-$\Set^{\V{I}}$. We call these \emph{indexed functors}. \note{Predicate
-  transformer anyone?} $\V{I} \To \IDesc{\V{I}}$ is our grammar for
-describing these functors. Hence, $\SYMBIDesc$ and its interpretation
-have the following types:
+\newcommand{\DotTo}{\mathop{\blue{\dot{\rightarrow}}}}
+
+We presented the $\Desc$ universe as a
+grammar of strictly positive endofunctors on $\Set$ and developed
+inductive types by taking a fixpoint. To describe inductive families
+indexed by some $\Bhab{\V{I}}{\Set}$, we play a similar game with
+endofunctors on the category $\Set^{\V{I}}$,
+families of sets \(\V{X},\V{Y}:\V{I}\To\Set\) for objects, and for morphisms,
+families of functions in \(\V{X}\DotTo\V{Y}\), defined pointwise:
+\[
+\V{X}\DotTo\V{Y} \mapsto \PI{\V{i}}{\V{I}}\V{X}\:\V{i}\To\V{Y}\:\V{i}
+\]
+
+An \emph{indexed functor} in $\Set^{\V{I}}\To\Set^{\V{J}}$ has the
+flavour of a device driver, characterising `responses' to a given
+request in \(\V{J}\) where we may in turn make `subrequests' at indices
+chosen from \(\V{I}\). When we use indexed functors to define inductive
+families of datatypes,
+\(\V{I}\) and \(\V{J}\) coincide: we explain how to make a node fit a given
+index, including subnodes at chosen indices. E.g., if we are asked for a
+vector of length 3, we choose to ask in turn for a tail of length 2.
+
+To code up valid notions of response to a given request, we introduce
+$\SYMBIDesc$ and its interpretation:
 %
@@
 
-Given these components, we may interpret a \emph{function}
-$\V{R}:\V{I} \To \IDesc{\V{I}}$ is interpreted as a function
-$\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
-\To \Set^{\V{I}}$, the type of endofunctors on \(\Set^{\V{I}}\).
-Inductive families are fixpoints defined over
-these indexed functors, hence computing a fixpoint of the entire
-\emph{family} of functors:
+An \(\IDesc{\V{I}}\) specifies just \emph{one} response, but a
+request-to-response \emph{function},
+$\V{R}:\V{I} \To \IDesc{\V{I}}$, yields a strictly positive endofunctor
+\[
+  \LAM{\V{X}} \LAM{\V{i}} \idescop{\V{R}\:\V{i}}{\V{I}}{\V{X}} :
+  \Set^{\V{I}} \To \Set^{\V{I}}
+\]
+whose fixpoint we then take:
 %
@@
 
+\newcommand{\upgrade}{\F{upgrade}}
 \newcommand{\inductionI}{\F{indI}}
@@
 
-However, we still have to define the actual grammar. We obtain it by
-evolving $\Desc$ to cope with indexing. \note{Say something about the
-  transformation} The code of $\SYMBIDesc$ is presented in
-Figure~\ref{fig:idesc}. By construction, $\SYMBIDesc$ can only
-describe \emph{strictly positive families}. Induction on indexed
-descriptions is defined by:
+We define the $\SYMBIDesc$ grammar in Figure~\ref{fig:idesc},
+delivering only \emph{strictly positive} families. As well as
+indexing our descriptions, we have refactored a little, adopting
+a more compositional algebra of codes, where $\Desc$ is rather
+biased towards the familiar right-nested tuples. We now have
+\(\DVar{i}\) for recursive `subrequests' at a chosen index \(i\),
+with tupling handled by \(\DProd{}{}\) and higher-order branching
+by \(\DPi{\!}{\!}\).  Upgrade your old $\Desc$
+to a trivially indexed $\IDesc{\Unit}$ as follows!
+\[\begin{array}{@{}ll}
+\upgrade :\Desc & \To\IDesc{\Unit} \\
+\upgrade\:\DUnit & = \DConst{\Unit} \\
+\upgrade\:(\DSigma{\V{S}}{\V{D}}) &
+   = \DSigma{\V{S}}{(\LAM{\V{s}}\upgrade\:(\V{D}\:\V{s}))} \\
+\upgrade\:(\DIndx{\V{D}}) & =
+  \DProd{\DVar{\Void}}{\upgrade\:\V{D}} \\
+\upgrade\:(\DHindx{\V{H}}{\V{D}}) & =
+  \DProd{(\DPi{\V{H}}{(\LAM{\_}\DVar{\Void})})}{\upgrade\:\V{D}} \\
+\end{array}\]
+
+To deliver induction for indexed datatypes, we need the `holds everywhere'
+machinery. We present $\SYMBAllI$ and $\SYMBallI$ in
+Figure~\ref{fig:allI-predicates}, with a twist---where
+$\Desc$ admits the $\SYMBall$ construction, $\SYMBIDesc$ is \emph{closed}
+under it! The $\SYMBAllI$
+operator for a description indexed on \(\V{I}\) is strictly positive in
+turn, and has a description indexed on
+ some \((\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}})\).
+Induction on indexed descriptions is then hardwired thus:
 %
@@
                                     {\V{R}\: \V{i}}
-                                    {\IMu{\V{I}}{\V{R}}{}}
+                                    {(\SYMBIMu_{\V{I}}{\V{R}})}
                                     {\V{P}}
@@
 %
-Where the operators $\SYMBAllI$ and $\SYMBallI$ are
-presented in Figure~\ref{fig:allI-predicates}. As for descriptions, we
-can compute a generic catamorphism, $\cataI$, from $\inductionI$.
+The generic catamorphism, $\cataI$, is constructed from $\inductionI$
+as before.
 
@@
 
-In order to gain some intuition of $\SYMBIDesc$, let us re-implement
-the pattern functor of natural numbers:
+In order to gain some intuition of $\SYMBIDesc$, let us revisit the
+natural numbers:
 %
@@
 
-Because $\Nat$ is just an inductive type, $\NatD$ is a $\Unit$-indexed
-functor. Therefore, the recursive argument is materialised by
-$\DVar{\Void}$, where we were using $\SYMBDIndx$ in the previous
-presentation. This transformation generalises to all inductive
-types. Moreover, we gain the ability to write mutually recursive
-inductive types by indexing with the choice of
-datatypes~\cite{paulin:habilitation, yakushev:mutual-def}.
+Of course, we can \(\upgrade\) any inductive type.
+Moreover, \(\IDesc{I}\) can now code a bunch of mutually
+inductive types, if \(I\) enumerates the
+bunch~\cite{paulin:habilitation, yakushev:mutual-def}.
 
@@
 
-Therefore, this universe is self-describing, hence can be
+Therefore, this universe is self-describing and can be
 levitated. As before, we rely on a special purpose $\F{switchID}$
@@
 
-So far, the examples we have seen live in $\IDesc{\Unit}$, hence are
-not using any indexing. We remedy this by encoding the vectors. Recall
-that the constructors $\VecNil$ and $\SYMBVecCons$ are only defined for an
-index $\NatZero$ and $\NatSuc$ respectively.:
+So far, our examples live in $\IDesc{\Unit}$, with no interesting
+indexing. Let us at least have vectors. Recall
+that the constructors $\VecNil$ and $\SYMBVecCons$ are defined only for
+$\NatZero$ and $\NatSuc$ respectively.:
 %
@@
 \[\stk{
-\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\:\V{ty}}{\Ty}{\SYMBVal}
+\evalOne : \PI{\V{ty}}{\Ty} \idescop{(\toIDesc{\HExprD})\:\V{ty}}{\Ty}{\SYMBVal}
            \To {\Val{\V{ty}}} \\
@@
            \Val{\V{ty}} \\
-\evalH\: \V{ty}\: \V{term} = \cataI_{\Ty}
-                                 \toIDesc{\HExprD}\: 
+\evalH\: \V{ty}\: \V{term} = \cataI_{\Ty} \:
+                                 (\toIDesc{\HExprD})\: 
                                  \SYMBVal\: 
@@
 
-\newcommand{\DotTo}{\mathop{\blue{\dot{\rightarrow}}}}
-
 The $\discharge$ function is relative to the required type and a
@@
 \discharge : & \PI{\V{G}}{\Ctxt}
-               \Env{G} \To \\
-             & (\SUM{\Val{\V{}}}{\Var{\V{G}}{}}) \DotTo
+               \Env{G} \To 
+              (\SUM{\Val{\V{}}}{\Var{\V{G}}{}}) \DotTo
                \IMu{\Ty}{\HExprD}{}
@@
           & (\V{X} \DotTo 
-             \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To 
-            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})}{} \DotTo
-            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
+             \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
+            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
+            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
 \end{array} \\
 \substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} = \\
-\qquad    \cataI\: \toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}}\:
-                      (\IMu{}{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
+\qquad    \cataI\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
+                      (\SYMBIMu{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
                       (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
@@
 presented the free monad construction, together with its monadic
-operations. A summary of the components defining descriptions and
-their indexed counterparts is presented in
-Table~\ref{tab:sumup-operators}.
-
-\begin{table}
-
-{\small
-
-\begin{tabular}{|c|c|c|c|}
-\hline
-Desc   & IDesc & Role & Status \\
-\hline
-\hline
-\multicolumn{2}{|c|}{$\EnumU$}                  & Build finite sets           & Levitated \\
-\hline
-$\Desc$               & $\IDesc$               & Describe functors   & Levitated \\
-\hline
-$\SYMBdescop{\_}$         & $\SYMBidescop{\_}{\_}$         & Interpret descriptions      & Hardwired \\
-\hline
-$\SYMBMu$, $\SYMBCon$ & $\SYMBIMu_I$, $\SYMBCon$ & Define fixpoints & Hardwired \\
-\hline
-$\SYMBind$, $\SYMBAll$, $\SYMBall$ & $\inductionI$, $\SYMBAllI$, $\SYMBallI$ & Induction principle & Hardwired \\
-\hline
-\end{tabular}
-
-}
-
-\caption{Summary of constructions on Descriptions}
-\label{tab:sumup-operators}
-
-\end{table}
+operations.

### 2010-07-21 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: 'intuitively' police

@@
 In order to capture a notion of higher-order induction, we add a code
-$\SYMBDHindx$ that takes an indexing set $\V{H}$. Intuitively,
-$\SYMBDHindx$ gives a recursive subobject for each element of $H$.
+$\SYMBDHindx$ that takes an indexing set $\V{H}$. This amounts to give
+a recursive subobject for each element of $H$.
 %
@@
 at hand. Our goal is to make an element of
-\(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of the right
-shape, but its sub-elements are of the wrong type. On the other hand,
-for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
-element in $\V{T}$.  Therefore, to construct an element of
+\(\descop{\V{D}}{\V{T}}\). $\V{xs}$ is of the right shape, but its
+sub-elements are of the wrong type. On the other hand, for each
+sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding element
+in $\V{T}$.  Therefore, to construct an element of
 \(\descop{\V{D}}{\V{T}}\), we must replace the recursive components of
 \(\V{xs}\) by their counterparts from \(\V{hs}\). Let us write a
-program to do that---please forgive us if we lapse to a pattern matching
-notation, for readability's sake.
+program to do that---please forgive us if we lapse to a pattern
+matching notation, for readability's sake.
 %
@@
 %
-We are left with implementing the algebra of the
-catamorphism. Intuitively, its role is to catch appearances of
-$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
-corresponds to the following definition:
+We are left with implementing the algebra of the catamorphism. Its
+role is to catch appearances of $\DVar{\V{x}}$ and replace them by
+$\V{\sigma}\: \V{x}$. This corresponds to the following definition:
 %

### 2010-07-22 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: 'replace' fits in single lines
Desc@ICFP: alternative to switchD: using fsigma

@@
 \[\stk{
-    \DHindx{(\Bhab{\M{H}}{\Set})}{(\Bhab{\M{D}}{\Desc})} : \Set \smallskip \\
-\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     \mapsto \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
+    \DHindx{(\Bhab{\M{H}}{\Set})}{(\Bhab{\M{D}}{\Desc})} : \Desc
+    \smallskip \\
+    \descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     
+        \mapsto 
+            \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
 }\]
@@
 type. Hardwired as a \emph{fait accompli}, this creates no regress,
-although one must assume the definition to recheck it.\footnote{Our
+although one must assume the definition to recheck it\footnote{Our
   Agda model does not formalize the $\SYMBswitchD$
@@
   collapse this isomorphism, hence identifiying defined and coded
-  descriptions.}
+  descriptions.}. Alternatively, we could have extended $\Desc$ with a
+code for \emph{finite} sums, internalizing the $\SYMBswitch$ in the
+interpretation function:
+%%
+\[\stk{
+    \Dsigma{(\Bhab{\V{E}}{\EnumU})}
+           {(\Bhab{\V{B}}{\spi{\V{E}}{\LAM{\_}{\Desc}}})} : \Desc
+    \smallskip \\
+    \descop{\Dsigma{\V{E}}{\V{B}}}{\V{X}}     
+        \mapsto 
+            \SIGMAS{\Bhab{\V{x}}{\EnumT{\V{E}}}}
+                   {\descop{\switch{\V{E}}{(\LAM{\_}{\Desc})}{\V{B}}{\V{x}}}{\V{X}}}    
+}\]
+%%
+Using $\SYMBDsigma$ to switch over the $\Desc$ constructors, we avoid
+mention of $\Desc$ while levitating it. The $\SYMBswitch$ operator is
+pushed into the interpretation function, at which point $\Desc$ indeed
+exists.
+
+%% * There exists alternative
+%% **   <- All we want is to avoid mention of Desc
+%% **   -> Add a code for finite sigma : $\sigma : (e : \EnumU) -> \pi e \Desc -> \Desc$
+%% **       <- Take a finite set in argument
+%% **       <- Interpreted as a switch
+%% **       -> No more need to mention Desc during its construction
+%% **           <- Pushed into intrepretation, at which time it exists
 
@@
                    \descop{\V{D}}{\V{Y}}} \\
-\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
+\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
+\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void   &
     \Void  \\
-\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}             \mapsto
+\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}  &
     \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
-\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} \mapsto        \\
-\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                      \\
-\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}\: \pair{\V{g}}{\V{d'}}{} \mapsto   \\
-\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
+\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} &
+    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
+\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}\: \pair{\V{g}}{\V{d'}}{} &
+ \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
+\end{array}
 }\]

### 2010-07-22 papers/icfp-2010-desc/paper_idesc.tex
gaol: sec 5 preamble tweaks
vector rejig

@@
 indexing our descriptions, we have refactored a little, adopting
-a more compositional algebra of codes, where $\Desc$ is rather
-biased towards the familiar right-nested tuples. We now have
+a more compositional algebra of codes, where $\Desc$ is
+biased towards the right-nested tuples. We now have
 \(\DVar{i}\) for recursive `subrequests' at a chosen index \(i\),
-with tupling handled by \(\DProd{}{}\) and higher-order branching
+with tupling by right-associative
+\(\DProd{}{}\) and higher-order branching
 by \(\DPi{\!}{\!}\).  Upgrade your old $\Desc$
@@
 \upgrade\:(\DSigma{\V{S}}{\V{D}}) &
-   = \DSigma{\V{S}}{(\LAM{\V{s}}\upgrade\:(\V{D}\:\V{s}))} \\
+   = \DSigma{\V{S}}{\LAM{\V{s}}\upgrade\:(\V{D}\:\V{s})} \\
 \upgrade\:(\DIndx{\V{D}}) & =
@@
 \upgrade\:(\DHindx{\V{H}}{\V{D}}) & =
-  \DProd{(\DPi{\V{H}}{(\LAM{\_}\DVar{\Void})})}{\upgrade\:\V{D}} \\
+  \DProd{(\DPi{\V{H}}{\LAM{\_}\DVar{\Void}})}{\upgrade\:\V{D}} \\
 \end{array}\]
@@
 turn, and has a description indexed on
- some \((\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}})\).
+ some \(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}\).
 Induction on indexed descriptions is then hardwired thus:
@@
 as before.
+\note{do this later?}
 
@@
 
-In order to gain some intuition of $\SYMBIDesc$, let us revisit the
-natural numbers:
-%
-\[\stk{
-\NatD : \IDesc{\Unit} \\
-\NatD \mapsto \DSigma{(\EnumT{[ \NatZero\: \SYMBNatSuc ]})}
-                     {\red{[} \DConst{\Unit} \quad \DVar{\Void} \red{]}}
-}\]
-
-Of course, we can \(\upgrade\) any inductive type.
+For basic reassurance, we \(\upgrade\:\NatD\):
+%
+\[\stk{
+\upgrade\:\NatD : \IDesc{\Unit} \\
+\upgrade\:\NatD = \DSigma{(\EnumT{[ \NatZero\: \SYMBNatSuc ]})}
+                     {\red{[} (\DConst{\Unit}) \; 
+   (\DProd{\DVar{\Void}}{\DConst{\Unit}}) \red{]}}
+}\]
+%
+Note that trailing \(\Unit\)'s keep our right-nested, \(\void\)-terminated
+tuple structure, and with it our elaboration machinery.
+We can similarly \(\upgrade\) any inductive type.
 Moreover, \(\IDesc{I}\) can now code a bunch of mutually
@@
 
-Note that $\IDesc{\V{I}}$ itself is merely an inductive type. Hence,
-we can describe it in $\IDesc{\Unit}$:
+Note that $\IDesc{\V{I}}$ is a plain inductive type, parametrized
+by \(\V{I}\), but indexed trivially.
 %
@@
 \IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
-\begin{array}{@{}ll}
-\IDescD\: \V{I} \mapsto \SYMBDSigma & (\EnumT \sqr{ 
-                                          \SYMBDVar\:\:
-                                          \SYMBDConst\:\:
-                                          \DProd\:\:
-                                          \SYMBDSigma\:\:
+\IDescD\: \V{I} \mapsto \SYMBDSigma \\
+\quad
+ \EnumT\bigRedBracket{\begin{array}{r}
+                                          \SYMBDVar\\
+                                          \SYMBDConst\\
+                                          \DProd{}{}\\
+                                          \SYMBDSigma\\
                                           \SYMBDPi
-                                             }) \\
-                              & \bigRedBracket{\begin{array}{l}
-                                      \DConst{\V{I}}                  \\
-                                      \DConst{\Set}               \\
-                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
-                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
-                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                        \end{array}}
-\end{array}
+         \; \bigRedBracket{\begin{array}{l@{}l}
+  (\DProd{\DConst{\V{I}} &}{\DConst{\Unit}})                  \\
+  (\DProd{\DConst{\Set}  &}{\DConst{\Unit}})                  \\
+  (\DProd{\DVar{\Void}}{\DProd{\DVar{\Void}&}{\DConst{\Unit}}})  \\
+  (\DSigma{\Set}{\LAM{\V{S}}
+     \DProd{( \DPi{\V{S}}{\LAM{\_} \DVar{\Void}}) &}{\DConst{\Unit}}})     \\
+  (\DSigma{\Set}{\LAM{\V{S}}
+     \DProd{( \DPi{\V{S}}{\LAM{\_} \DVar{\Void}}) &}{\DConst{\Unit}}})     \\
+                                   \end{array}}
 }\]
@@
 that the constructors $\VecNil$ and $\SYMBVecCons$ are defined only for
-$\NatZero$ and $\NatSuc$ respectively.:
+$\NatZero$ and $\NatSuc$ respectively:
 %
@@
 \stk{
-\data \D{Vec}\: \PITEL{\V{X}}{\Set} : \PI{\V{n}}{\Nat} \Set \where \\
+\data \D{Vec}\: \PITEL{\V{X}}{\Set} : \PI{\V{i}}{\Nat} \Set \where \\
 \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
@@
 
-One way to code constrained datatypes is to appeal to equality. The
-constraints are therefore captured by equations in the datatype. In
-this case, we obtain the following definition:
+One way to code constrained datatypes is to appeal to a suitable
+notion of propositional equality \(\PropEq\) on indices. The
+constraints are expressed as `Henry Ford' equations in the datatype.
+For vectors:
 %
@@
 \VecD : \Set \To \Nat \To \IDesc{\Nat} \\
-\begin{array}{@{}l@{}l}
-\VecD\: \V{X}\: \V{n} = \SYMBDSigma & (\EnumT{\sqr{ \VecNil\:\SYMBVecCons }}) \\
-                        & \bigRedBracket{
-                          \begin{array}{l}
-                            \DConst{(\V{n} \PropEq \NatZero)} \\
-                            \DSigma{\Nat}{\LAM{\V{m}}
-\stk{\DProd{\DConst{\V{X}}}{\DProd{\DVar{\V{m}}}{\\\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}}}
+\VecD\: \V{X}\: \V{i} = \SYMBDSigma\\
+\quad
+\EnumT{\bigRedBracket{\begin{array}{@{}r@{}}
+\VecNil\\ \SYMBVecCons
+\end{array}}}
+\; \bigRedBracket{
+                          \begin{array}{@{}r@{}}
+                            (\DConst{(\NatZero\PropEq\V{i})}) \\
+ ( \DSigma{\Nat}{\LAM{\V{n}}
+   \DProd{\DConst{\V{X}}}
+     {\DProd{\DVar{\V{n}}}{\DConst{(\NatSuc{\V{n}}\PropEq\V{i})}})}}
                           \end{array}
                           }
-\end{array}
-}\]
-
-In the $\VecNil$ case, a proof must be provided that the index is
-equal to $\NatZero$. In the $\SYMBVecCons$ case, we first store an element
-$\V{m}$ of $\Nat$. However, the constraint stipulates that $\V{m}$
-cannot be \emph{any} natural numbers: it must be ``the index minus
-one''. This translates into the constraint $\V{n} \PropEq
-\NatSuc{\V{m}}$, given a suitable presentation of propositional equality.
+}\]
+
+You may choose $\VecNil$ for any index you like as long as it is
+$\NatZero$; in the $\SYMBVecCons$ case, the length of the tail is
+given explicitly, and the index $\V{i}$ must be one more. Our previous
+\(\Unit\)-terminated tuple types can now be seen as the trivial case
+of constraint-terminated tuple types, with elaboration supplying the
+witnesses when trivial.
 
@@
 \VecD\:\V{X}\: \NatZero     & \mapsto & \DConst{\Unit} \\
-\VecD\:\V{X}\: (\NatSuc{\V{n}}) & \mapsto & \DProd{\DConst{\V{X}}}{\DVar{\V{n}}}
+\VecD\:\V{X}\: (\NatSuc{\V{n}}) & \mapsto &
+ \DProd{\DProd{\DConst{\V{X}}}{\DVar{\V{n}}}}{\DConst{\Unit}}
 \end{array}

### 2010-07-22 papers/icfp-2010-desc/paper_idesc.tex
gaol: boiling the indexed examples

@@
 \label{sec:indexing-desc}
+
+\newcommand{\vtup}[2]{\bigRedBracket{\begin{array}{@{}#1@{}}#2\end{array}}}
 
@@
 \quad
- \EnumT\bigRedBracket{\begin{array}{r}
-                                          \SYMBDVar\\
+ \EnumT\vtup{r}{\SYMBDVar\\
                 \SYMBDConst\\
@@
                 \SYMBDSigma\\
-                                          \SYMBDPi
-                       \end{array}}
-         \; \bigRedBracket{\begin{array}{l@{}l}
+                \SYMBDPi}
+         \; \vtup{l@{}l}{
   (\DProd{\DConst{\V{I}} &}{\DConst{\Unit}})                  \\
@@
      \DProd{( \DPi{\V{S}}{\LAM{\_} \DVar{\Void}}) &}{\DConst{\Unit}}})     \\
-                                   \end{array}}
+                                   }
 }\]
@@
 \quad
-\EnumT{\bigRedBracket{\begin{array}{@{}r@{}}
-\VecNil\\ \SYMBVecCons
-\end{array}}}
-\; \bigRedBracket{
-                          \begin{array}{@{}r@{}}
+\EnumT{\vtup{r}{\VecNil\\ \SYMBVecCons}}
+\; \vtup{r}{
                             (\DConst{(\NatZero\PropEq\V{i})}) \\
@@
      {\DProd{\DVar{\V{n}}}{\DConst{(\NatSuc{\V{n}}\PropEq\V{i})}})}}
-                          \end{array}
                           }
@@
 
-We have been careful to keep our setup agnostic with respect to
-notions of propositional equality. Any will do, according to your
-convictions, or for vectors, none---equality for \(\Nat\) is definable by
-recursion---and many variations are popular. The traditional
-homogeneous identity type used in Coq is not adequate to support
-dependent pattern matching, but its heterogeneous variant, allowing
-equations between elements of arbitrary types, is sufficient to
-allow the translation of structurally recursive pattern matching
-programs to \(\F{indI}\)~\cite{goguen:pattern-matching}. Our present
-inclination is towards the extensional equality proposed by
-\citet{altenkirch:ott}, which also sustains the translation.
+In this paper, we remain anxiously agnostic about
+propositional equality. Any will do, according to
+conviction; many variations are popular. The
+homogeneous identity type used in Coq is ill-suited to
+dependent types, but its heterogeneous variant (forming equations
+regardless of type) allows the translation of pattern
+matching with structural recursion to
+\(\F{indI}\)~\cite{goguen:pattern-matching}. The
+extensional equality of \citet{altenkirch:ott} also sustains the translation.
 
@@
 
-However, sometimes, we can actually remove these equations altogether. Let us
-look back at $\D{Vec}$. We note that the equations are introduced
-because we are \emph{storing} the index of the inductive family.
-However, \emph{inductive families need not store their
-  indices}~\cite{brady:index-inductive-families}.  By examining the
-incoming index, we can apply the \emph{forcing} and \emph{de-tagging}
-optimisations to our initial definition of $\D{Vec}$. This gives the
-following, equivalent definition:
+However, sometimes, the equations are redundant. 
+Looking back at $\D{Vec}$, we find that the equations constrain
+the choice of constructor and stored tail index retrospectively.
+But \emph{inductive families need not store their
+  indices}~\cite{brady:index-inductive-families}!  If we
+analyse the incoming index, we can tidy our description of $\D{Vec}$
+as follows:
 %
@@
 \VecD\:\V{X}\: (\NatSuc{\V{n}}) & \mapsto &
- \DProd{\DProd{\DConst{\V{X}}}{\DVar{\V{n}}}}{\DConst{\Unit}}
-\end{array}
-
-}\]
-%
-The equations (and constructors) have simply disappeared. A similar
-example is $\SYMBFin$, specified by:
+ \DProd{\DConst{\V{X}}}{\DVar{\V{n}}}
+\end{array}
+
+}\]
+%
+The constructors and equations have simply disappeared. A similar
+example is $\SYMBFin$ (bounded numbers), specified by:
 %
@@
 %
-In this case, we can apply forcing, but not detagging, since both
+In this case, we can eliminate equations but not constructors, since both
 $\FinZero$ and $\SYMBFinSuc$ both target $\SYMBNatSuc$:
@@
 \FinD : \Nat \To \IDesc{\Nat} \\
-\begin{array}{@{}lll@{}l}
-\FinD\: \NatZero         & \mapsto & \multicolumn{2}{l}{\DSigma{(\EnumT{\Void})}{\Void} } \\
-\FinD\: (\NatSuc{\V{n}}) & \mapsto & \SYMBDSigma & (\EnumT{\sqr{ \FinZero \: \SYMBFinSuc }}) \\
-                         &         &             & \bigRedBracket{
-                                                   \begin{array}{l}
-                                                       \DConst{\Unit} \\
-                                                       \DVar{\V{n}}
-                                                   \end{array}}
-\end{array}
-}\]
-
-Note that forcing a description is not guaranteed to remove all
-constraints. It is subject to future work to see if constraints can be
-entirely eradicated, or presented more conveniently to the developer.
-Finally, it is worth mentioning that these optimisations are
-\emph{source-to-source} transformations on descriptions.
+\begin{array}{@{}lll}
+\FinD\: \NatZero         & \mapsto & \DSigma{\EnumT{\Void}}{\Void} \\
+\FinD\: (\NatSuc{\V{n}}) & \mapsto & \DSigma
+ {\EnumT{\vtup{r}{\FinZero \\ \SYMBFinSuc }}}
+ {\vtup{l}{ (\DConst{\Unit}) \\
+            (\DVar{\V{n}})}}
+\end{array}
+}\]
+
+This technique of extracting information by case analysis on indices
+applies to descriptions exactly where Brady's `forcing' and
+`detagging' optimisations apply in compilation. They eliminate just
+those constructors, indices and constraints which are redundant even
+in \emph{open} computation. In \emph{closed} computation, where proofs
+can be trusted, all constraints are dropped.
 
@@
 
-When defining an indexed datatype, we have access to its
-index. Therefore, we can use this index to influence the choice of
-constructors. This captures the essence of dependent datatypes: a
-term -- the index -- has the ability to influence the datatype. We
-define tagged indexed descriptions to capture this specificity.
-
-We divide a tagged indexed description in two parts: first, the
+Let us reflect this index analysis technique.
+We can divide a description of tagged indexed data in two: first, the
 constructors that do not depend on the index; then, the constructors
@@
 constructors by $\V{I}$. Hence, by inspecting the index, it is
-possible to enable or disable constructors.
+possible to vary the `menu' of constructors.
 %
@@
  \TagIDesc{\V{I}}  & \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
- \ATagIDesc{\V{I}} & \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
+ \ATagIDesc{\V{I}} & \SIGMA{\V{E}}{\EnumU} \PI{\V{i}}{\V{I}} \spi{\V{E}}{\LAM{\_} \IDesc{\V{I}}} \\
  \ITagIDesc{\V{I}} & 
-     \SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
+     \SIGMA{\V{F}}{\V{I} \To \EnumU} \PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{\LAM{\_} \IDesc{\V{I}}} 
 \end{array}
@@
 expansion is similar to the definition of \(\SYMBtoDesc\) for tagged
-descriptions. For brevity, we shall also overload $\IMu{ty}{D}$ for
-$\IMu{ty}{(\toIDesc{D})}$, as the bidirectional presentation let us
+descriptions. For brevity, we shall also overload $\IMu{I}{D}$ for
+$\IMu{I}{(\toIDesc{D})}$, as the bidirectional presentation let us
 resolve the ambiguity.
@@
 \HExprAD : \ATagIDesc{\Ty} \\
-\HExprAD \mapsto \bigRedBracket{
-                 \begin{array}{l}
+\HExprAD \mapsto \vtup{l}{
                    {\sqr{\SYMBEval\: \SYMBEcond \,}} \red{,} \\
@@
                    }
-                 \end{array}
                  }

### 2010-07-23 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: uniform = vs. \mapsto: choose \mapsto
Desc@ICFP: helplessly stare at whitespaces and parenthesis
Desc@ICFP: remove duplicate definition of DotTo.
Desc@ICFP: this \Eq is a \PropEq
Desc@ICFP: uniform Desc and IDesc codes presentation
Desc@ICFP: life and section 5 in colour^W color.

@@
 \upgrade :\Desc & \To\IDesc{\Unit} \\
-\upgrade\:\DUnit & = \DConst{\Unit} \\
+\upgrade\:\DUnit & \mapsto \DConst{\Unit} \\
 \upgrade\:(\DSigma{\V{S}}{\V{D}}) &
-   = \DSigma{\V{S}}{\LAM{\V{s}}\upgrade\:(\V{D}\:\V{s})} \\
-\upgrade\:(\DIndx{\V{D}}) & =
+   \mapsto \DSigma{\V{S}}{\LAM{\V{s}}\upgrade\:(\V{D}\:\V{s})} \\
+\upgrade\:(\DIndx{\V{D}}) & \mapsto
   \DProd{\DVar{\Void}}{\upgrade\:\V{D}} \\
-\upgrade\:(\DHindx{\V{H}}{\V{D}}) & =
+\upgrade\:(\DHindx{\V{H}}{\V{D}}) & \mapsto
   \DProd{(\DPi{\V{H}}{\LAM{\_}\DVar{\Void}})}{\upgrade\:\V{D}} \\
@@
 \end{array} \\
-\inductionI\: \V{R}\: \V{P}\: \V{m}\: \V{i}\: (\Con{\V{xs}}) =  \\
-\qquad
+\inductionI\: \V{R}\: \V{P}\: \V{m}\: \V{i}\: (\Con{\V{xs}}) \mapsto 
     \V{m}\: \V{i}\: \V{xs}\: (\allI{}
@@
 \end{array} \\
-\begin{array}{@{}l@{}l@{\:=\:\:}l}
+\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
 \AllI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{x}} &
@@
 \AllI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{f}} &
-    \DPi{\V{S}}{(\LAM{\V{s}} \AllI{}{(\V{D}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})})}
+    \DPi{\V{S}}{\LAM{\V{s}} \AllI{}{(\V{D}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})}}
 \end{array}
@@
 \end{array} \\
-\begin{array}{@{}l@{}l@{\:=\:\:}l}
+\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
 \allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{P}}{\V{p}}{\V{x}} &
@@
 \[\stk{\begin{array}{@{}ll}
-\IDesc{(\Bhab{\M{I}}{\Set})} &: \Set \\
-\DVar{(\Bhab{\M{i}}{\M{I}})} &: \IDesc{\M{I}} \\
-\DConst{(\Bhab{\M{A}}{\Set})} &:\IDesc{\M{I}}       \\
-\DProd{(\Bhab{\M{D}}{\IDesc{\M{I}}})}{(\Bhab{\M{D}}{\IDesc{\M{I}}})}
-  & :\IDesc{\M{I}}       \\
-\DSigma{(\Bhab{\M{S}}{\Set})}{(\Bhab{\M{D}}{\M{S}\To\IDesc{\M{I}}})}
-& : \IDesc{\M{I}}  \\
-\DPi{(\Bhab{\M{S}}{\Set})}{(\Bhab{\M{D}}{\M{S}\To\IDesc{\M{I}}})}
-& : \IDesc{\M{I}}  \\
+\IDesc{(\Bhab{\V{I}}{\Set})} &: \Set \\
+\DVar{(\Bhab{\V{i}}{\V{I}})} &: \IDesc{\V{I}} \\
+\DConst{(\Bhab{\V{A}}{\Set})} &:\IDesc{\V{I}}       \\
+\DProd{(\Bhab{\V{D}}{\IDesc{\V{I}}})}{(\Bhab{\V{D}}{\IDesc{\V{I}}})}
+  & :\IDesc{\V{I}}       \\
+\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S}\To\IDesc{\V{I}}})}
+& : \IDesc{\V{I}}  \\
+\DPi{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S}\To\IDesc{\V{I}}})}
+& : \IDesc{\V{I}}  \\
 \end{array}\smallskip \\
@@
 \upgrade\:\NatD : \IDesc{\Unit} \\
-\upgrade\:\NatD = \DSigma{(\EnumT{[ \NatZero\: \SYMBNatSuc ]})}
-                     {\red{[} (\DConst{\Unit}) \; 
-   (\DProd{\DVar{\Void}}{\DConst{\Unit}}) \red{]}}
+\upgrade\:\NatD \mapsto \DSigma{(\EnumT{\sqr{\NatZero\: \SYMBNatSuc}})}
+                     {\sqr{(\DConst{\Unit}) \; 
+                           (\DProd{\DVar{\Void}}{\DConst{\Unit}})}}
 }\]
@@
 \VecD : \Set \To \Nat \To \IDesc{\Nat} \\
-\VecD\: \V{X}\: \V{i} = \SYMBDSigma\\
+\VecD\: \V{X}\: \V{i} \mapsto \SYMBDSigma\\
 \quad
@@
 \FinD\: \NatZero         & \mapsto & \DSigma{\EnumT{\Void}}{\Void} \\
-\FinD\: (\NatSuc{\V{n}}) & \mapsto & \DSigma
- {\EnumT{\vtup{r}{\FinZero \\ \SYMBFinSuc }}}
- {\vtup{l}{ (\DConst{\Unit}) \\
-            (\DVar{\V{n}})}}
+\FinD\: (\NatSuc{\V{n}}) & \mapsto & \DSigma{\EnumT{\sqr{\FinZero\: \SYMBFinSuc}}}
+                                            {\sqr{(\DConst{\Unit})\: (\DVar{\V{n}})}}
 \end{array}
@@
 expansion is similar to the definition of \(\SYMBtoDesc\) for tagged
-descriptions. For brevity, we shall also overload $\IMu{I}{D}$ for
-$\IMu{I}{(\toIDesc{D})}$, as the bidirectional presentation let us
+descriptions. For brevity, we shall also overload $\IMu{\V{I}}{\V{D}}$ for
+$\IMu{\V{I}}{(\toIDesc{\V{D}})}$, as the bidirectional presentation let us
 resolve the ambiguity.
@@
 \SYMBVal : \Ty \To \Set \\
-\begin{array}{@{}l@{\:=\:\:}l}
+\begin{array}{@{}l@{\:\mapsto\:\:}l}
 \Val{\Enat}   & \Nat \\
@@
            \To {\Val{\V{ty}}} \\
-\begin{array}{@{}l@{}c@{}l@{\:=\:\:}l}
+\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
 \evalOne\: & \_\: & \pair{\SYMBEval}{\V{x}}{}                                             & \V{x} \\
@@
            \Val{\V{ty}} \\
-\evalH\: \V{ty}\: \V{term} = \cataI_{\Ty} \:
+\evalH\: \V{ty}\: \V{term} \mapsto \cataI_{\Ty} \:
                                  (\toIDesc{\HExprD})\: 
@@
 \[
-\stk{
-\data \Ctxt : \Set \where \\
-\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
+\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
+\Ctxt           & \Set \\
 \SYMBCtxtEmpty  & \Ctxt \\
@@
 \end{array}
-}\]
+\]
 %
@@
 \SYMBEnv : \Ctxt \To \Set \\
-\begin{array}{@{}l@{\:\:=\:\:}l}
+\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
 \Env{\CtxtEmpty}                & \Unit \\
-\Env{(\CtxtSnoc{G}{S})} & \TIMES{\Env{G}}{\Val{S}}
+\Env{(\CtxtSnoc{\V{G}}{\V{S}})} & \TIMES{\Env{\V{G}}}{\Val{\V{S}}}
 \end{array}
@@
 \Var{}{} : \Ctxt \To \Ty \To \Set \\
-\begin{array}{@{}ll@{\:\:=\:\:}l}
-\Var{\CtxtEmpty}{& T}            & \Void \\
-\Var{(\CtxtSnoc{G}{S})}{& T}     & \SUM{(\Var{G}{T})}{(S \Eq T)}
+\begin{array}{@{}ll@{\:\:\mapsto\:\:}l}
+\Var{\CtxtEmpty}{& \V{T}}                & 
+    \Void \\
+\Var{(\CtxtSnoc{\V{G}}{\V{S}})}{& \V{T}} & 
+    \SUM{(\Var{\V{G}}{\V{T}})}{(\V{S} \PropEq \V{T})}
 \end{array}
@@
           \Val{\V{T}} \\
-\begin{array}{@{}lll}
-\lookup{\CtxtEmpty}{               \_}{                   \V{T}}{()}                    &   & \\
-\lookup{(\CtxtSnoc{\V{G}}{.T})}{   \pair{\V{g}}{\V{t}}{}}{\V{T}}{(\SumRight{\C{refl}})} & = & \V{t} \\
-\lookup{(\CtxtSnoc{\V{G}}{\V{S}})}{\pair{\V{g}}{\V{t}}{}}{\V{T}}{(\SumLeft{\V{x}})}     & = & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
+\begin{array}{@{}l@{}l@{}l@{}l@{}lll}
+%% \lookup{& \CtxtEmpty}{& \_}{& \V{T}}{& ()}                    &   & \\
+\lookup{& (\CtxtSnoc{\V{G}}{.T})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumRight{\C{refl}})} & \mapsto & \V{t} \\
+\lookup{& (\CtxtSnoc{\V{G}}{\V{S}})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumLeft{\V{x}})} & \mapsto & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
 \end{array}
@@
 context of the right type. Its action is to map values to themself,
-and variables to their value in context. The pointwise indexed
-function type $\PI{\V{i}}{I}{X\: i \To Y\: i}$ is abbreviated  $X
-\DotTo Y$. This corresponds to the following function:
+and variables to their value in context. This corresponds to the
+following function:
 %
@@
 \discharge : & \PI{\V{G}}{\Ctxt}
-               \Env{G} \To 
+               \Env{\V{G}} \To 
               (\SUM{\Val{\V{}}}{\Var{\V{G}}{}}) \DotTo
@@
 \end{array} \\
-\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} = \\
+\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
 \qquad    \cataI\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
-                      (\SYMBIMu{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
+                      \SYMBIMu{(\FreeIMonad{\V{R}}{}{\V{Y}})}\:
                       (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
@@
 \IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
-                                \sqr{\SYMBDConst\:
-                                     \DProd\:
-                                     \SYMBDSigma\: 
+                                \sqr{\SYMBDConst\:\:
+                                     \DProd\:\:
+                                     \SYMBDSigma\:\:
                                      \SYMBDPi} \red{,}\\
@@
 \[
-\idescop{\sigma D}{J}{X} = \idescop{D}{I}{\LAM{i}\idescop{\sigma i}{J}{X}}
-\quad \mbox{where}\;\sigma:I\To\IDesc{J}
+\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
+    \mapsto 
+        \idescop{\V{D}}
+                {\V{I}}
+                {\LAM{\V{i}}
+                     {\idescop{\V{\sigma} \V{i}}
+                              {\V{J}}
+                              {\V{X}}}}
+        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
 \]

### 2010-07-23 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: replace plusHost and leHost and +_H and \leq_H

@@
 
+\newcommand{\plusHost}{\mathop{\green{+_{\mathrm{H}}}}}
+\newcommand{\leHost}{\mathop{\green{\leq_{\mathrm{H}}}}}
+
 An expression of this language is either a value, a conditional
@@
 natural numbers and booleans in the host language, equipped with an
-addition operation $\F{plusHost}$ and a comparison function
-$\F{leHost}$. We define $\SYMBVal$ as follows:
+addition operation $\plusHost$ and a comparison function
+$\leHost$. We define $\SYMBVal$ as follows:
 %
@@
 \evalOne\: & \_\: & \pair{\SYMBEcond}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{y} \\
-\evalOne\: & \Enat\: & \pair{\SYMBEplus}{\pair{\V{x}}{\V{y}}{}}{}                     & \F{plusHost}\: \V{x}\: \V{y} \\
-\evalOne\: & \Ebool\: & \pair{\SYMBEle}{\pair{\V{x}}{\V{y}}{}}{}                      & \F{leHost}\: \V{x}\: \V{y} 
+\evalOne\: & \Enat\: & \pair{\SYMBEplus}{\pair{\V{x}}{\V{y}}{}}{}                     & \V{x} \plusHost \V{y} \\
+\evalOne\: & \Ebool\: & \pair{\SYMBEle}{\pair{\V{x}}{\V{y}}{}}{}                      & \V{x} \leHost \V{y} 
 \end{array} \\

### 2010-07-23 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: helplessly stare at whitespaces and parenthesis
Desc@ICFP: life and section 5 in colour^W color.
gaol: boiling the catamorphism

@@
        {\stkm{ \DUnit                                \\
-         \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
+         \DSigma{\Set}{\LAM{\V{S}} \SHED}      \\
          \DIndx{\DUnit}    }                    } \\
@@
 In Section~\ref{sec:desc-fix-point}, we hardwired a dependent
-$\F{ind}$unction principle, instead of the catamorphism. However, in
-some circumstances, the full power of a dependent elimination is not
-necessary. Let us now derive the catamorphism
-from $\F{ind}$ principle.
+$\F{ind}$uction principle, but sometimes,
+iteration suffices. Let us construct the catamorphism.
 
@@
 
-The catamorphism is defined by induction on the description $\V{D}$,
-with a readily propagated non-dependent return type $\V{T}$.
+We proceed by induction on the \emph{data} in \(\Mu{\V{D}}\):
+the non-dependent return type $\V{T}$ is readily propagated.
 Given a node $\V{xs}$
@@
 % 
-We are left with filling the hole. Recall that we have
+The hole remains: we have
 \(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
 \(\Bhab{\V{hs}}{\All{\V{D}}{\Mu{\V{D}}}{(\LAM{\_} \V{T})}{\V{xs}}}\)
-at hand. Our goal is to make an element of
-\(\descop{\V{D}}{\V{T}}\). $\V{xs}$ is of the right shape, but its
-sub-elements are of the wrong type. On the other hand, for each
-sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding element
-in $\V{T}$.  Therefore, to construct an element of
-\(\descop{\V{D}}{\V{T}}\), we must replace the recursive components of
-\(\V{xs}\) by their counterparts from \(\V{hs}\). Let us write a
-program to do that---please forgive us if we lapse to a pattern
-matching notation, for readability's sake.
+to hand, and we need a \(\descop{\V{D}}{\V{T}}\).
+Now, $\V{xs}$ has the right shape, but its
+components have the wrong type. However, for each such
+component, $\V{hs}$ holds the corresponding value
+in $\V{T}$.  We need a function to \(\F{replace}\) the former with the latter:
+this pattern
+matching sketch yields an induction on \(\V{D}\). We fill the hole
+with \(\F{replace}\: \V{D}\:(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\).
 %
@@
                    \descop{\V{D}}{\V{Y}}} \\
-\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
-\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void   &
+\begin{array}{@{}l@{\:}l@{\:}l@{\:\:\mapsto\:\:}l}
+\F{replace}\: \DUnit &          \V{X}\: \V{Y}\: \Void&          \Void   &
     \Void  \\
-\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}  &
+\F{replace}\: (\DSigma{\V{S}}{\V{D}})& \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}& \V{d'}  &
     \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
-\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} &
+\F{replace}\: (\DIndx{\V{D}})&     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}& \pair{\V{y}}{\V{d'}}{} &
     \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
-\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}\: \pair{\V{g}}{\V{d'}}{} &
+\F{replace}\: (\DHindx{\V{H}}{\V{D}})& \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}& \pair{\V{g}}{\V{d'}}{} &
  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
@@
 %
-Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
-(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the problem. In the
-type theory, we have built a generic catamorphism. Any datatype will
-now come equipped with this operation, for free.
 
@@
 
-With this example, we have shown how we can derive a generic
-operation, the catamorphism, from a pre-existing generic operation,
-the induction principle. This has been made possible by our ability to
-manipulate descriptions as first-class objects: the catamorphism is,
-basically, a function mapping a $\Desc$ to a datatype specific
-operation. This is a form of polytypic programming, as we learned from
-PolyP~\cite{jansson:polyp}.
+We have shown how to derive a generic
+operation, \(\cata\), from a pre-existing generic operation,
+\(\F{ind}\), by
+manipulating descriptions as data: the catamorphism is just a
+function taking each $\Desc$ value to a datatype specific
+operation. This is polytypic programming, as in
+PolyP~\cite{jansson:polyp}, made ordinary.
 
@@
 
-{\small
-
+{
 \begin{center}

### 2010-07-23 papers/icfp-2010-desc/paper_desc_levitation.tex
gaol: boiling 4.3 and 4.4

@@
 
-In this section, we will turn to a more ambitious generic operation on
-datatypes. Given a functor, represented as a tagged description, we
-build the free monad over this functor.
-
+In this section, we try a more ambitious generic operation. Given a functor,
+represented as a tagged description, we
+build its free monad.
+%
 \newcommand{\FMFreeMonad}{\D{FreeMonad}}
@@
 \newcommand{\FMComposite}{\C{Composite}}
-
-Let us recall the free monad construction in, say, Haskell. Given a
-functor \texttt{f}, the free monad over \texttt{f} is defined by the
-following datatype:
+%
+Let us recall this construction in, say, Haskell. Given a
+functor \texttt{f}, the free monad over \texttt{f} is given:
 %
@@
 %
-Being an inductive type, this $\FMFreeMonad$ datatype is itself
-defined by a pattern functor. It is given by:
+Provided \texttt{f} is an instance of \texttt{Functor}, we may
+take \texttt{Var} for \texttt{return} and use \texttt{f}'s
+\texttt{fmap} to define \texttt{>>=} as substitution.
+
+Being an inductive type, $\FMFreeMonad$ arises
+by a pattern functor:
 %
 \[
-\FMFreeMonadD\: \V{F}\: \V{X}\: \V{Z} \mapsto \V{X} \mathop{\D{+}} \V{F} \V{Z}
+\FMFreeMonadD\: \V{F}\: \V{X}\: \V{Z} \mapsto \V{X} \mathop{\D{+}} \V{F}\:\V{Z}
 \]
@@
 
-In our setting, the free monad construction will take the functor as a
-tagged description, a set $\V{X}$ of variables, and will compute the
-tagged description of the corresponding free monad. Implementing this
-function is surprisingly easy:
+Our free monad construction takes the functor as a
+tagged description, and given a set $\V{X}$ of variables, computes the
+tagged description of the free monad pattern functor.
 %
@@
 %
-We simply add a constructor, $\SYMBDVar$, and define its argument to
-be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
-$\V{E}$ and $\V{D}$ as they were, hence leaving the other constructors
+We simply add a constructor, $\SYMBDVar$, making its arguments
+$\DSigma{\V{X}}{\DUnit}$---just an element of $\V{X}$. 
+$\V{E}$ and $\V{D}$ stay put, leaving the other constructors
 unchanged. Unfolding the interpretation of this definition, we
-convince ourselves that this corresponds to the functor
-$\FMFreeMonadD$. The fixpoint operation ties the knot and gives us
-the full-blown free monad construction.
+find an extended sum, corresponding to the \(\V{X}\D{+}\) in
+$\FMFreeMonadD$. Taking the fixpoint ties the knot and we have our
+data.
 
@@
 
-Of course, we must equip the resulting datatypes with operations
-delivering a monadic interface. As expected, \(\LAM{\x}\DVar{\x}\)
-plays the r\^ole of \return, embedding variables into terms. The
-\bind operation corresponds to \emph{substitution}. We will now
-implement it, as a generic function.
-
 \newcommand{\subst}{\F{subst}}
@@
 
-Our implementation will appeal to the $\cata$ function developed
-previously. So, let us write down the types, and fill as many
-arguments to $\cata$ as possible:
+Now we need the operations. As expected, \(\LAM{\x}\DVar{\x}\)
+plays the r\^ole of \return, making variables terms. Meanwhile,
+\bind is indeed \emph{substitution}, which we now
+implement generically, making use of $\cata$. Let us write the type,
+and start filling in the blanks:
 %
@@
            \PI{\V{X}, \V{Y}}{\Set} 
-           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To \\
-         & \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
-           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
+           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
+         & \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
+           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
 \end{array} \\
 \subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
-  \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
-          \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}\: 
+  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
+          \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}\: 
           \SHED
@@
            \PI{\V{X}, \V{Y}}{\Set} 
-           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To \\
-         & \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
-           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
+           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
+         & \descop{\toDesc{(\FreeMonad{\V{D}}{\V{X}})}}{\Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}} \To
+           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}
 \\
@@
 
-Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
-\V{\sigma}$ completes the implementation. To sum up, we have
-implemented the free monad construction for an arbitrary tagged
-description. This gives the developer the ability, for any datatype,
-to extend it with a notion of variable. Then, we have equipped this
-structure with the corresponding monadic operations, \bind and
-\return. This construction is an example of type-indexed
-datatype~\cite{hinze:generic-haskell}, as found in Generic Haskell:
-from a datatype, we build a new datatype and equip it with its
-structure.
+We complete the hole with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
+\V{\sigma}$. Every tagged
+description can be seen as a signature of operations: we can uniformly
+add a notion of variable, building a new type from an old one, then
+providing the substitution structure.
 

### 2010-07-23 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: uniform = vs. \mapsto: choose \mapsto
Desc@ICFP: uniform Desc and IDesc codes presentation
a bit of boiling in 3; abbreviated mu.de

@@
 sum-of-products structure in each node of the tree-like values in a
-datatype, we need to account somehow for the recursive structure which
-ties these nodes together. Not for the first time, we do this by
+datatype, we need to account for the recursive structure which
+ties these nodes together. As ever, we do this by
 constructing a \emph{universe}~\cite{martin-lof:itt}. Universes
@@
 programming~\cite{benke:universe-generic-prog, oury:power-of-pi},
-but here we seek to exploit them as the foundation of our notion
+but here we take them as the foundation of our notion
 of datatypes.
@@
 
-Descriptions are sequential structures, terminated by $\DUnit$,
-indicating the empty tuple. To build sigmas-of-sigmas, we define a
+Descriptions are sequential structures ending in $\DUnit$,
+indicating the empty tuple. To build sigmas-of-sigmas, we provide a
 $\SYMBDSigma$ code, interpreted as a $\Sigma$-type. To request a
 recursive component, we have $\DIndx{\V{D}}$, where \(\V{D}\)
-describes the rest of the node. These codes directly captures the
-notion of sigmas-of-sigmas. An equivalent, more algebraic presentation
+describes the rest of the node. These codes give us
+sigmas-of-sigmas with recursive places. An equivalent, more algebraic presentation
 could be given, as illustrated in Section~\ref{sec:indexing-desc}.
@@
 \[\stk{
-\begin{array}{l}
-\Desc^n : \Set^{n+1}\\
-\DUnit : \Desc^{n}\\
-\DSigma{(\Bhab{\V{S}}{\Set^{n}})}{(\Bhab{\V{D}}{\V{S} \To \Desc^{n}})} : \Desc^{n}\\
-\DIndx{(\Bhab{\V{D}}{\Desc^{n}})} : \Desc^{n} \\
+\begin{array}{@{}ll}
+\Desc^n 
+    &: \Set^{n+1}\\
+\DUnit 
+    &: \Desc^{n}\\
+\DSigma{(\Bhab{\V{S}}{\Set^{n}})}{(\Bhab{\V{D}}{\V{S} \To \Desc^{n}})} 
+    &: \Desc^{n}\\
+\DIndx{(\Bhab{\V{D}}{\Desc^{n}})} 
+    &: \Desc^{n} \\
 \end{array}
@@
 
-You may have noticed that we are a little coy about this presentation,
+We admit to being a little coy,
 writing of `implementing a signature' without clarifying how. A viable
 approach would simply be to extend the theory with constants for the
-constructors and an operator for \(\descop{\V{D}}\). However, in
-Section~\ref{sec:desc-levitate}, you will see what we actually do.  In
-the meantime, let us first gain some intuition for its use by
-developing some examples.
+constructors and an operator for \(\descop{\V{D}}\). In
+Section~\ref{sec:desc-levitate}, you will see what we do instead.
+Meanwhile, let us gain some intuition by developing examples.
 
@@
 
-With a small change to this definition, we obtain the pattern functor
-for lists:
+The pattern functor
+for lists needs but a small change:
 %
@@
 %
-The $\SYMBNatSuc$ constructor is turned into a proper
-$\SYMBListCons$, taking an argument in $\V{X}$ followed by a
+The $\SYMBNatSuc$ constructor becomes
+$\SYMBListCons$, taking an $\V{X}$ followed by a
 recursive argument. This code describes the following functor:
@@
 
-Finally, we are not limited to one recursive argument. This is
-demonstrated by our description of node-labelled binary trees:
+Of course, we are not limited to one recursive argument. Here are
+the node-labelled binary trees:
 %
@@
 call codes fitting this pattern \emph{tagged} descriptions. Again,
-this is a clear reminder of the sum-of-products style. Every
+this is a clear reminder of the sum-of-products style. Any
 description can be forced into this style with a singleton constructor
-if necessary. We characterise tagged descriptions thus:
+set. We characterise tagged descriptions thus:
 \[\stk{
@@
 }\]
-It is not a great stretch to imagine that the traditional datatype
-declaration syntax might desugar to the \emph{definition} of a datatype
-via a tagged description.
+It is not such a stretch to expect that the familiar datatype
+declaration might desugar to the \emph{definitions} of a tagged description.
 
@@
 
+\newcommand{\mude}[1]{\D{\(\mu^{\!+}\)}\:#1}
+
 So far, we have built pattern functors with our \(\Desc\) universe.
@@
 \Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc^n} \quad 
-      \Gamma \vdash \Bhab{\M{d}}{\descop{\M{D}}{\Mu{\M{D}}}}}
+      \Gamma \vdash \Bhab{\M{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}}
      {\Gamma \vdash \Bhab{\Con{\M{d}}}{\Mu{\M{D}}}}
+\]
+Tagged descriptions are very common, so we abbreviate:
+\[
+  \mude{}:\TagDesc^n \To \Set^n \qquad \mude{\V{T}}\mapsto\Mu{(\toDesc{\V{T}})}
 \]
@@
 \[\stk{
-\Nat \mapsto \Bhab{\Mu{(\toDesc{\pair{\sqr{\NatZero\: \SYMBNatSuc }}
-                               {\sqr{\DUnit \quad (\DIndx{\DUnit})}}{}})}}
+\Nat \mapsto \Bhab{\mude{\pair{\sqr{\NatZero\: \SYMBNatSuc }}
+                               {\sqr{\DUnit \quad (\DIndx{\DUnit})}}{}}}
                   {\Set^n}
@@
                 \PI{\V{P}}{\Mu{\V{D}} \To \Set^k}         \\
-               (\PI{\V{d}}{\descop{\V{D}}{\Mu{\V{D}}}}       
-                \All{\V{D}}{\Mu{\V{D}}}{\V{P}}{\V{d}} \To \V{P} (\Con{\V{d}})) \To \\
+               (\PI{\V{d}}{\descop{\V{D}}{(\Mu{\V{D}})}}       
+                \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{d}} \To \V{P} (\Con{\V{d}})) \To \\
                \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x} 
 } \\
-\F{ind}\, \V{D}\, \V{P}\, \V{m}\, (\Con{\V{d}}) = 
-    \V{m}\, \V{d}\, (\all{\V{D}}{\Mu{\V{D}}}{\V{P}}
+\F{ind}\, \V{D}\, \V{P}\, \V{m}\, (\Con{\V{d}}) \mapsto 
+    \V{m}\, \V{d}\, (\all{\V{D}}{(\Mu{\V{D}})}{\V{P}}
                            {(\F{ind}\, \V{D}\, \V{P}\, \V{m})}
@@
 \(\Bhab{\V{p}}{\PIS{\Bhab{\x}{\V{X}}}\V{P}\,\V{x}}\) to each \(\x\)
-contained in \(\V{d}\). The full definition (including an extra case,
-introduced shortly) is presented in
+contained in \(\V{d}\). The definition (including an extra case,
+introduced soon) is in
 Figure~\ref{fig:all-predicates}.\footnote{To pass the termination
@@
   does. Hence, although not directly structural, this definition is
-  indeed terminating.} Note that $\SYMBind$ is our first generic
-operation over descriptions, albeit a hardwired operator.  Any
-datatype we define automatically comes with an induction principle.
+  indeed terminating.} So, $\SYMBind$ is our first
+operation generic over descriptions, albeit hardwired.  Any
+datatype we define comes with induction.
 
-We note that the very same functors \(\SYMBdescop{\M{D}}\) also admit greatest
-fixpoints, and we have indeed implemented coinductive types this way, but
-that is a story for another time.
+Note that the very same functors \(\SYMBdescop{\M{D}}\) also admit greatest
+fixpoints: we have indeed implemented coinductive types this way, but
+that is another story.
 
@@
 \end{array} \\
-\begin{array}{@{}l@{}l@{\:=\:\:}l}
+\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
 \All{\DUnit}{& \V{X}}{\V{P}}{\Void} &
@@
 \end{array} \\
-\begin{array}{@{}l@{}l@{\:=\:\:}l}
+\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
 \all{\DUnit}{& \V{X}}{\V{P}}{\V{p}}{\Void} &
@@
 
-We have now enough machinery to build and manipulate inductive
-types at a low level. Let us now apply cosmetic surgery
-to the syntactic overhead. We extend type checking of
+We now have low level machinery to build and manipulate inductive
+types. Let us apply cosmetic surgery
+to reduce the syntactic overhead. We extend type checking of
 expressions:
@@
   \push{\sqr{\vec{t}}}
-   {\descop{\M{D}\: n}{\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})}}}}
+   {\descop{\M{D}\: n}{(\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})})}}}
         {\M{t'}}}
@@
 \propag{\push{\M{f}}
-             {\PI{\V{d}}{\descop{\M{D}}{\Mu{\M{D}}}}       
-                \All{\M{D}}{\Mu{\M{D}}}{(\PLAM{\x}{\Mu{\M{D}}}\M{P})}{\V{d}} \To \M{P}[\Con{\V{d}}/\x]\\}}
+             {\PI{\V{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}       
+                \All{\M{D}}{(\Mu{\M{D}})}{(\PLAM{\x}{\Mu{\M{D}}}\M{P})}{\V{d}} \To \M{P}[\Con{\V{d}}/\x]\\}}
        {\M{f'}}}}

### 2010-07-23 papers/icfp-2010-desc/paper_desc_levitation.tex
gaol: mu.de free monad tweaks

@@
 
-In this section, we try a more ambitious generic operation. Given a functor,
-represented as a tagged description, we
-build its free monad.
+In this section, we try a more ambitious generic operation. Given a
+functor---a signature of operations represented as a tagged
+description---we build its free monad, extending the signature with
+variables and substitution.
 %
@@
 Let us recall this construction in, say, Haskell. Given a
-functor \texttt{f}, the free monad over \texttt{f} is given:
+functor \texttt{f}, the free monad over \texttt{f} is given
+thus:
 %
-\begin{code}
-data FreeMonad f x
-    = Var x
-    | Composite (f (FreeMonad f x))
-\end{code}
+\[
+\texttt{data FreeMonad f x = Var x | Op (f (FreeMonad f x))}
+\]
 %
@@
 
-Our free monad construction takes the functor as a
+Our construction takes the functor as a
 tagged description, and given a set $\V{X}$ of variables, computes the
@@
            \PI{\V{X}, \V{Y}}{\Set} 
-           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
-         & \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
-           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
+           (\V{X} \To \mude{(\FreeMonad{\V{D}}{\V{Y}})}) \To \\
+         & \mude{(\FreeMonad{\V{D}}{\V{X}})} \To
+           \mude{(\FreeMonad{\V{D}}{\V{Y}})} 
 \end{array} \\
@@
   \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
-          \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}\: 
+          (\mude{(\FreeMonad{\V{D}}{\V{Y}})})\: 
           \SHED
@@
            \PI{\V{X}, \V{Y}}{\Set} 
-           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
-         & \descop{\toDesc{(\FreeMonad{\V{D}}{\V{X}})}}{\Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}} \To
-           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}
+           (\V{X} \To \mude{(\FreeMonad{\V{D}}{\V{Y}})}) \To \\
+         & \descop{\toDesc{(\FreeMonad{\V{D}}{\V{X}})}}{(\mude{(\FreeMonad{\V{D}}{\V{Y}})})}
+           \To  \mude{(\FreeMonad{\V{D}}{\V{Y}})}
 \\

### 2010-07-23 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: a slightly less grungy Skyhook.

@@
 arise. First, we have blurred the line between the implementation of
-data-types and their reflection in themselves. Second, we have abused
-the paradoxical nature of $\Bhab{\Set}{\Set}$ to flatten the hierarchy
-of descriptions. We shall now be more precise about these points.
+data-types and their self-description. Second, we have abused the
+paradoxical nature of $\Bhab{\Set}{\Set}$ to flatten the hierarchy of
+descriptions. We shall now be more precise about these points.
 
@@
 Let us first clarify the status of the implementation. The kit for
-making datatypes is summed up in Table~\ref{tab:sumup-operators}. For
+making datatypes is presented in Table~\ref{tab:sumup-operators}. For
 each operation, we describe its role and its status, making clear
@@
 
-To tackle the apparent circularity of our construction, let us
+To demystify the apparent circularity of our construction, let us
 consider how it bottoms out in a stratified setting. Using the
 informal level annotations of Section~\ref{sec:universe-desc}, we have
-that \(\Bhab{\DescD_n}{\Desc_{n+1}}\), hence \(\Bhab{\Desc_n \mapsto
-  \Mu{\DescD_n}}{\Set_{n+1}}\). Therefore, there is no circularity in
-the description of descriptions: assuming an hardwired $\Desc_N$
-universe, we can build the tower of descriptions down to
-$\Desc_0$\footnote{We have modelled this process in Agda, hardwiring
-  $\Desc_{42}$ in $\Set_{43}$, then going down. This model does not
+that \(\Bhab{\DescD^n}{\Desc^{n+1}}\), hence \(\Bhab{\Desc^n \mapsto
+  \Mu{\DescD^n}}{\Set^{n+1}}\). Assuming an hardwired $\Desc^N$
+universe, we can build the tower of descriptions down to $\Desc^0$:
+description of descriptions forms a `spiral', rather than a
+circle \footnote{We have modelled this process in Agda, hardwiring
+  $\Desc^{42}$ in $\Set^{43}$, then going down. This model does not
   rely on dependent pattern-matching, induction-recursion, nor set

### 2010-07-23 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: set polymorphism in Coq model

@@
 of $\Desc$ itself\footnote{This is witnessed in our set polymorphic
-  model in Agda. Set polymorphism is an experimental feature of
-  Agda. Although not backed by theoretical works, it is a conservative
-  extension of Agda that is generally accepted as free of paradox.}.
+  model in Agda and in our Coq model, crucially relying on typical
+  ambiguity~\cite{harper:implicit-universe}.}.
 

### 2010-07-23 papers/icfp-2010-desc/paper_idesc.tex
gaol: interpreter shavings

@@
 
+\newcommand{\muide}[2]{\D{\({\mu}^{\!+}\)}\!\!_{#1}\:#2}
+
 Let us reflect this index analysis technique.
@@
 
-We use the notation $\toIDesc{\V{D}}$ to denote the indexed
-description computed from the tagged indexed description $\V{D}$. Its
-expansion is similar to the definition of \(\SYMBtoDesc\) for tagged
-descriptions. For brevity, we shall also overload $\IMu{\V{I}}{\V{D}}$ for
-$\IMu{\V{I}}{(\toIDesc{\V{D}})}$, as the bidirectional presentation let us
-resolve the ambiguity.
+We write $\toIDesc{\V{D}}\:\V{i}$ to denote the $\IDesc{\V{I}}$
+computed from the tagged indexed description $\V{D}$ at index
+$\V{i}$. Its expansion is similar to the definition of \(\SYMBtoDesc\)
+for tagged descriptions, except that it must also append the two parts.
+We again write $\muide{\V{I}}{\V{D}}$ for
+$\IMu{\V{I}}{(\toIDesc{\V{D}})}$.
 
@@
 %% Hutton expressions
-\newcommand{\HExprD}{\C{ExprD}}
-\newcommand{\HExprAD}{\C{ExprAD}}
-\newcommand{\HExprID}{\C{ExprID}}
+\newcommand{\HExprD}{\F{ExprD}}
+\newcommand{\HExprAD}{\F{ExprAD}}
+\newcommand{\HExprID}{\F{ExprID}}
 \newcommand{\HExprVarD}[1]{\C{ExprD}_{\F{Var},#1}}
@@
 
-We are going to define a syntax for a small typed language. We
-consider two types, natural numbers and booleans:
+We are going to define a syntax for a small language with
+two types, natural numbers and booleans:
 %
@@
 
-An expression of this language is either a value, a conditional
-expression, an addition of numbers, or a comparison of
-numbers. Informally, their type is the following:
+This language has values, conditional
+expression, addition and comparison. Informally, their types are:
 %
@@
 \begin{array}{l@{\::\:\:}l}
-\SYMBEcond           & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
+\SYMBEval            & \Val{\V{ty}} \To \V{ty} \\
+\SYMBEcond           & \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
+\end{array}
+\qquad
+\begin{array}{l@{\::\:\:}l}
 \SYMBEplus           & \Enat \To \Enat \To \Enat                           \\
 \SYMBEle             & \Enat \To \Enat \To \Ebool                          \\
-\SYMBEval            & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
 \end{array}
@@
 %
-The function $\SYMBVal$, used in the definition of $\SYMBEval$, simply
-maps a type $ty$ of the object language to the corresponding type in
-the host language. Hence, the arguments of $\SYMBEval$ are ensured to
-be of the expected type. We assume that $\Nat$ and $\Bool$ represent
-natural numbers and booleans in the host language, equipped with an
-addition operation $\plusHost$ and a comparison function
-$\leHost$. We define $\SYMBVal$ as follows:
-%
+The function $\SYMBVal$ interprets object language types in the
+host language, so that arguments to $\SYMBEval$ fit their
+expected type.
 \[\stk{
@@
 }\]
-
-In our universe of descriptions, the syntax of this language is
-described by a tagged indexed description. We use the index to carry
-the type: the resulting description is indexed by $\Ty$. We observe
-that some constructors are always defined, namely $\SYMBEcond$ and
-$\SYMBEval$. On the other hand, the $\SYMBEplus$ and $\SYMBEle$
-constructors are index-dependent. $\SYMBEplus$ is defined if and only
-if the result type -- the index -- is $\Enat$, whereas $\SYMBEle$ is
-defined if and only if the index is $\Ebool$. The actual code
-precisely follows this intuition, as shown in
-Figure~\ref{fig:hexpr-full}. The first element of $\HExprID$
-corresponds to this choice of constructor: the sublists corresponds to
-the constructors enabled by the corresponding type in $\Ty$.
-
-\begin{figure}
-
+We take $\Nat$ and $\Bool$ to represent
+natural numbers and Booleans in the host language, equipped with
+addition $\plusHost$ and comparison $\leHost$.
+
+We express our syntax as a tagged indexed description, indexing
+over obhect language types $\Ty$. We note
+that some constructors are always available, namely $\SYMBEval$ and
+$\SYMBEcond$. On the other hand, $\SYMBEplus$ and $\SYMBEle$
+constructors are index-dependent, with $\SYMBEplus$ available just when
+building a $\Enat$, $\SYMBEle$ just for $\Ebool$. The code,
+%%%in Figure~\ref{fig:hexpr-full},
+below,
+reflects this intuition, with the first
+component uniformly offering $\SYMBEval$ and $\SYMBEcond$, the second
+selectively offering $\SYMBEplus$ or $\SYMBEle$.
+%
+%%% \begin{figure}
+%
 \[\stk{
@@
 \HExprD : \TagIDesc{\Ty} \\
-\HExprD \mapsto ( \HExprAD , \HExprID ) \\
-} \\
-\\
+\HExprD \mapsto \sqr{ \HExprAD , \HExprID } \\
+} \smallskip\\
 \stk{
@@
 \HExprAD \mapsto \vtup{l}{
-                   {\sqr{\SYMBEval\: \SYMBEcond \,}} \red{,} \\
+   {\vtup{r}{\SYMBEval\\ \SYMBEcond \,}} \red{,} \;
       \LAM{\V{ty}}
-                   \bigRedBracket{
-                   \begin{array}{l}
-                   \DConst{(\Val{\V{ty}})} \\
-                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
-                   \end{array}
+      \vtup{l@{\:}l}{
+      \DProd{\DConst{(\Val{\V{ty}})}&}{\DConst{\Unit}} \\
+      \DProd{\DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}&}
+        {\DConst{\Unit}} \\
      }
    }
-\\
-\\
+\smallskip\\
 } 
@@
 \HExprID : \ITagIDesc{\Ty} \\
-\HExprID \mapsto \bigRedBracket{
-                 \begin{array}{l}
-                   \sqr{\sqr{\SYMBEplus} \quad \sqr{\SYMBEle}} \red{,} \\
-                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
-                   \end{array}
+\HExprID \mapsto \vtup{l}{
+                   \vtup{r}{\sqr{\SYMBEplus} \\ \sqr{\SYMBEle}} \red{,} \;
+  \LAM{\_} \sqr{\DProd{\DProd{\DVar{\Enat}}{\DVar{\Enat}}}{\DConst{\Unit}}}
                    }
@@
 
-\caption{Syntax of typed expressions}
-\label{fig:hexpr-full}
-
-\end{figure}
+%%%\caption{Syntax of typed expressions}
+%%%\label{fig:hexpr-full}
+
+%%%\end{figure}
 
@@
 
-Having implemented the syntax, we would like to describe its
-semantics. To do so, we implement an evaluator. The type of the
-evaluator is:
-%
-\[
-\evalH : \PI{\V{ty}}{\Ty} 
-         \IMu{\Ty}{\HExprD}{\V{ty}} \To
-         \Val{\V{ty}}
-\]
-%
-The type of $\evalH$ is strikingly similar to a
-catamorphism. Indeed, implementing a single step of evaluation -- the
-algebra -- is sufficient, as $\cataI$ gives, for free, the full
-evaluator. The implementation is as follows:
-%
-\[\stk{
-\evalOne : \PI{\V{ty}}{\Ty} \idescop{(\toIDesc{\HExprD})\:\V{ty}}{\Ty}{\SYMBVal}
-           \To {\Val{\V{ty}}} \\
-\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
-\evalOne\: & \_\: & \pair{\SYMBEval}{\V{x}}{}                                             & \V{x} \\
-\evalOne\: & \_\: & \pair{\SYMBEcond}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
-\evalOne\: & \_\: & \pair{\SYMBEcond}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{y} \\
-\evalOne\: & \Enat\: & \pair{\SYMBEplus}{\pair{\V{x}}{\V{y}}{}}{}                     & \V{x} \plusHost \V{y} \\
-\evalOne\: & \Ebool\: & \pair{\SYMBEle}{\pair{\V{x}}{\V{y}}{}}{}                      & \V{x} \leHost \V{y} 
-\end{array} \\
-\\
+Given the syntax, let us supply the
+semantics. We implement an evaluator as a catamorphism:
+%
+\[\stk{
 \evalH : \PI{\V{ty}}{\Ty} 
@@
 }\]
+%
+To finish the job, we must supply the algebra which implements a single
+step of evaluation, given subexpressions evaluated already.
+%
+\[\stk{
+\evalOne : \PI{\V{ty}}{\Ty}
+ \idescop{(\toIDesc{\HExprD})\:\V{ty}}{\Ty}{\SYMBVal}
+           \To {\Val{\V{ty}}} \\
+\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
+\evalOne\: & \_\: & (\SYMBEval\;\V{x})                                            & \V{x} \\
+\evalOne\: & \_\: & (\SYMBEcond\:\BoolTrue\:\V{x}\:\_)   & \V{x} \\
+\evalOne\: & \_\: & (\SYMBEcond\:\BoolFalse\:\_\:\V{y})  & \V{y} \\
+\evalOne\: & \Enat\: & (\SYMBEplus\:\V{x}\:\V{y})   & \V{x} \plusHost \V{y} \\
+\evalOne\: & \Ebool\: & (\SYMBEle\:\V{x}\:\V{y})  & \V{x} \leHost \V{y} 
+\end{array}
+}\]
 
@@
 
-Hence, we have defined the syntax of a typed language of arithmetic
-and boolean expressions. We have given its semantics through an
-evaluation function. Provided a one step semantics of the language, the
-big step interpreter is granted without effort thanks to the generic
-catamorphism.
+Hence, we have a type-safe syntax and a tagless interpreter for our
+language, in the spirit
+of~\citet{augustsson.carlsson:dependent.interpreter}, with
+help from the generic catamorphism.
 

### 2010-07-23 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: add type of cataI
Desc@ICFP: refactor Free monad story / Hutton's example
The presentation of the Hutton's razor is now completely done after
the Free Indexed Monad section, as an example of such
construction. So, roughly, we start off saying 'how it would be great
to have a small language'. Let's starts with the operation, that's the
pattern functor of a free monad. Then, we make this damn free monad:
with Empty for closed terms, with Var for open terms. 

So far, so good, we have a syntax. So, we ask 'what about the
semantics'. Semantics of closed terms is rather standard evaluato

@@
 The generic catamorphism, $\cataI$, is constructed from $\inductionI$
-as before.
-\note{do this later?}
+as before. Its type becomes more elaborated, to deal with the
+indexing:
+%
+\[
+\begin{array}{@{}l@{}l}
+\cataI :& \PITEL{\V{I}}{\Set}
+          \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
+        & \PI{\V{T}}{\V{I} \To \Set}
+          (\PI{\V{i}}{\V{I}}{\idescop{\V{R}\: \V{i}}{}{\V{T}} \To \V{T}\: \V{i}}) \To
+          \SYMBIMu_{\V{I}}{\V{R}} \DotTo \V{T}
+\end{array}
+\]
 
@@
 
-%% \[\stk{
-%% \data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
-%% \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
-%%     \DVar{}         & I \To \IDesc{I}                                   \\
-%%     \ldots          & \ldots
-%% \end{array} \\
-%% \\
-%% \idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{I} \To (\V{I} \To \Set) \To \Set        \\
-%% \begin{array}{@{}l@{\:=\:\:}ll}
-%% \idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
-%% \ldots                        &  \ldots
-%% \end{array}
-%% }\]
-
 \begin{figure}
@@
 
-%\subsection{Tagged indexed descriptions}
-
-% This discussion seemed out of place to me, so i've moved it a bit later (pwm).
-
-%\begin{wstructure}
-%<- Tagged constructor choice
-%    <- Index available when defining the datatype
-%        -> Can influence the choice of constructors
-%        -> Dependently-typed datatypes
-%            <- term (index) influence types
-%    -> Two parts
-%        <- Always on the menu
-%            <- E : EnumU 
-%            <- ED : E -> IDesc I
-%        <- Index-dependent
-%            <- F : I -> EnumU 
-%            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
-%        -> taggedIDesc I == Sigma E ED x Sigma F FD
-%        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
-%\end{wstructure}
-%
-%In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
-%form for descriptions. This format follows the usual presentation of
-%inductive types as sum-of-product. With indexed description, this
-%definition can be generalized. Indeed, when defining an indexed
-%datatype, we have access to this index. Therefore, we can use this
-%index to influence the choice of constructors. This captures the
-%essence of dependent datatypes: a term -- the index -- has the
-%ability to influence the datatype.
-%
-%For convenience, we divide a tagged indexed description in two parts:
-%first, the constructors that do not depend on the index; then, the
-%constructors that do. The non-dependent part mirrors the definition
-%for non-indexed descriptions: we are provided a finite choice of
-%constructors. The index-depend part simply indexes the choice of
-%constructors by $I$. Hence, by inspecting the index, it is possible to
-%enable or disable the constructors. 
-%
-%\[\stk{
-% \TagIDesc{\V{I}}  \mapsto \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
-% \ATagIDesc{\V{I}} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
-% \ITagIDesc{\V{I}} \mapsto \\
-%\qquad\SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
-%}\]
-%
-%
-%\begin{wstructure}
-%<- Vectors
-%    Do we treat them in the end? 
-%    What can we say here we haven't with Fin?
-%\end{wstructure}
-%
-%In the case of vectors, for instance, for the index $\NatZero$, we
-%would only propose the constructor $\ListNil$. Similarly, for
-%$\NatSuc{n}$, we only propose the constructor $\ListCons{\!}{\!}$. Hence,
-%we obtain the following definition:
-%
-%\newcommand{\VecD}{\C{VecD}}
-%\newcommand{\VecNil}{\etag{\CN{vnil}}}
-%\newcommand{\VecCons}[2]{\etag{\CN{vcons}}\:#1\:#2}
-%
-%\[\stk{
-%\VecD : \Set \To \TagIDesc{\Nat} \\
-%\begin{array}{@{}ll}
-%\VecD\: \V{X} = \\ \quad\bigRedBracket{ 
-%            \begin{array}{l}
-%                \pair{[]}{\_}{} 
-%                \\\\
-%                \bigRedBracket{
-%                \begin{array}{l} 
-%                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
-%                                   (\NatSuc{\V{n}}) \mapsto [\VecCons{\!}{\!}]} \\
-%                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
-%                                   (\NatSuc{\V{n}}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
-%                \end{array}}
-%            \end{array}}
-%\end{array}
-%}\]
-%
-%We will see more instances of this pattern in the following.
-
 \subsection{Examples}
@@
 
+\subsection{Free indexed monad}
+
+\begin{wstructure}
+<- Variation on a theme: free imonad construction
+    <- Recall existence of generic free monad construction
+    -> Present its generalisation to IDesc [equation]
+        <- \I -> IDesc I as describing an indexed endofunctor
+        <- Free monad construction
+    -> Still a suitable, generic notion of substitution
+        <- show type signature
+        <- show implementation?? (space! space!)
+\end{wstructure}
+
+In Section~\ref{sec:desc-free-monad}, we have built a free monad
+operation for simple descriptions. The process is similar in the
+indexed world. Namely, given an indexed functor, we derive the indexed
+functor coding its free monad: \note{pwm: Whoa there. Maybe we should
+  say something about IMonads in general before we get to this point?}
+%
+\[\stk{
+\begin{array}{ll}
+\FreeIMonad{\_}{} : & _{\PI{\V{I}}{\Set}}
+                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
+                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
+                      \TagIDesc{\V{I}}
+\end{array} \\
+\FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto \\ \;\;
+    \pair{\pair{\ListCons{\SYMBDVar}{(\fst{\V{E}})}} 
+               {\LAM{\V{i}}
+                \pair{\DConst{(\V{R}\: \V{i})}}
+                     {(\snd{\V{E}})\: \V{i}}{}}{}}
+         {\V{F}}{}
+}\]
+
+\newcommand{\substI}{\F{substI}}
+
+Just as in the universe of descriptions, this construction comes with
+an obvious \return and a substitution operation, the \bind. Its
+definition is the following:
+%
+\[\stk{
+\begin{array}{@{}ll}
+\substI : & _{\PI{\V{I}}{\Set}}
+            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
+            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
+          & (\V{X} \DotTo 
+             \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
+            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
+            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
+\end{array} \\
+\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
+\qquad    \cataI\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
+                      \SYMBIMu{(\FreeIMonad{\V{R}}{}{\V{Y}})}\:
+                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
+                      \V{i}\:
+                      \V{t} 
+}\]
+% 
+Where  $\F{applyI}$ is defined as follow:
+%
+\[\stk{
+\begin{array}{@{}ll}
+\F{applyI} : & _{\PI{\V{I}}{\Set}}
+            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
+            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
+          & (\V{X} \DotTo \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
+          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
+                                         {\V{I}}
+                                         {\V{X}}}}
+                    {\V{I}}
+                    {\IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
+            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
+\end{array} \\
+\begin{array}{@{}l@{\:\mapsto\:\:}l}
+\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
+\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
+\end{array}
+}\]
+
+The subscripted types corresponds to implicit arguments that can be
+automatically inferred, hence do not have to be typed in. Let us now
+consider two examples of free indexed monads.
+
+\paragraph{Indexed descriptions:}
+
+An interesting instance of free monad is $\SYMBIDesc$ itself. Indeed,
+$\SYMBDVar$ is nothing but the \return. The remaining constructors form
+the carrier functor, trivially indexed by $\Unit$. The carrier functor
+is described as follow:
+%
+\[\stk{
+\IDescFreeD : \ATagIDesc{\Unit} \\
+\begin{array}{@{}ll}
+\IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
+                                \sqr{\SYMBDConst\:\:
+                                     \DProd\:\:
+                                     \SYMBDSigma\:\:
+                                     \SYMBDPi} \red{,}\\
+                                  \LAM{\_}\bigRedBracket{\begin{array}{l}
+                                        \DConst{\Set}               \\
+                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
+                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
+                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
+                                    \end{array}}\end{array}}
+\end{array}
+}\]
+%
+Then, we get $\IDesc$ by building its free monad:
+%
+\[\stk{
+\IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
+\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}\sqr{}\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
+}\]
+
+The fact that indexed descriptions are closed under substitution
+is potentially of considerable utility, if we can exploit this fact:
+\[
+\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
+    \mapsto 
+        \idescop{\V{D}}
+                {\V{I}}
+                {\LAM{\V{i}}
+                     {\idescop{\V{\sigma} \V{i}}
+                              {\V{J}}
+                              {\V{X}}}}
+        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
+\]
+By observing that a description can be decomposed via substitution, we
+split its meaning into a superstructure of substructures, e.g. a
+`database containing salaries', ready for traversal operations
+preserving the former and targeting the latter. We shall now turn to
+some examples of indexed descriptions, as well as a practical use of
+free monad construction.
+
 \paragraph{Typed expressions:}
@@
 \newcommand{\SYMBEval}{\etag{\CN{val}}\xspace}
+\newcommand{\SYMBEvar}{\etag{\CN{var}}\xspace}
 \newcommand{\Eval}[1]{\SYMBEval\:#1}
@@
 
-We express our syntax as a tagged indexed description, indexing
-over obhect language types $\Ty$. We note
-that some constructors are always available, namely $\SYMBEval$ and
-$\SYMBEcond$. On the other hand, $\SYMBEplus$ and $\SYMBEle$
-constructors are index-dependent, with $\SYMBEplus$ available just when
-building a $\Enat$, $\SYMBEle$ just for $\Ebool$. The code,
-%%%in Figure~\ref{fig:hexpr-full},
-below,
-reflects this intuition, with the first
+We describe our syntax in two steps. First, we give the pattern
+functor of the free monad, defining the operators of the
+language. Second, we apply the free monad construction to obtain a
+language of closed terms and a language of open terms. The operations
+are described by a tagged indexed description, indexing over object
+language types $\Ty$. We note that some constructors are always
+available, namely $\SYMBEval$ and $\SYMBEcond$. On the other hand,
+$\SYMBEplus$ and $\SYMBEle$ constructors are index-dependent, with
+$\SYMBEplus$ available just when building a $\Enat$, $\SYMBEle$ just
+for $\Ebool$. The code, below, reflects this intuition, with the first
 component uniformly offering $\SYMBEval$ and $\SYMBEcond$, the second
@@
 \begin{wstructure}
-    -> evaluation: IMu TermD -> Val
-        -> it is a catamorphism
-            <- Look closer at the type
-        -> implementation [code]
-            <- Just define one reduction step
-            -> cata does the rest
-                /> cata is for free!
-\end{wstructure}
-
-\newcommand{\evalH}{\F{eval}_{\green{\Downarrow}}}
-\newcommand{\evalOne}{\F{eval}_{\green{\downarrow}}}
-
-Given the syntax, let us supply the
-semantics. We implement an evaluator as a catamorphism:
-%
-\[\stk{
-\evalH : \PI{\V{ty}}{\Ty} 
-           \IMu{\Ty}{\HExprD}{\V{ty}} \To
-           \Val{\V{ty}} \\
-\evalH\: \V{ty}\: \V{term} \mapsto \cataI_{\Ty} \:
-                                 (\toIDesc{\HExprD})\: 
-                                 \SYMBVal\: 
-                                 \evalOne\: 
-                                 \V{ty}\: 
-                                 \V{term}
-}\]
-%
-To finish the job, we must supply the algebra which implements a single
-step of evaluation, given subexpressions evaluated already.
-%
-\[\stk{
-\evalOne : \PI{\V{ty}}{\Ty}
- \idescop{(\toIDesc{\HExprD})\:\V{ty}}{\Ty}{\SYMBVal}
-           \To {\Val{\V{ty}}} \\
-\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
-\evalOne\: & \_\: & (\SYMBEval\;\V{x})                                            & \V{x} \\
-\evalOne\: & \_\: & (\SYMBEcond\:\BoolTrue\:\V{x}\:\_)   & \V{x} \\
-\evalOne\: & \_\: & (\SYMBEcond\:\BoolFalse\:\_\:\V{y})  & \V{y} \\
-\evalOne\: & \Enat\: & (\SYMBEplus\:\V{x}\:\V{y})   & \V{x} \plusHost \V{y} \\
-\evalOne\: & \Ebool\: & (\SYMBEle\:\V{x}\:\V{y})  & \V{x} \leHost \V{y} 
-\end{array}
-}\]
-
-\begin{wstructure}
     /> Closed term
@@
 \end{wstructure}
-
-Hence, we have a type-safe syntax and a tagless interpreter for our
-language, in the spirit
-of~\citet{augustsson.carlsson:dependent.interpreter}, with
-help from the generic catamorphism.
 
@@
 However, so far, we are only able to define and manipulate
-\emph{closed} terms. By abstracting over $\SYMBVal$, it is possible to
-build and manipulate \emph{open} terms, that is, terms in a
-context. The context is represented by a snoc-list of types:
+\emph{closed} terms. Adding variables, it is possible to build and
+manipulate \emph{open} terms, that is, terms in a context. The context
+is represented by a snoc-list of types:
 %
@@
 %
-While $\SYMBVal$ was mapping the type to the corresponding host type,
+While $\SYMBVal$ maps the type to the corresponding host type,
 $\SYMBVar$ indexes a value in the context, obtaining a proof that the
@@
 \begin{array}{@{}l@{}l@{}l@{}l@{}lll}
-%% \lookup{& \CtxtEmpty}{& \_}{& \V{T}}{& ()}                    &   & \\
 \lookup{& (\CtxtSnoc{\V{G}}{.T})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumRight{\C{refl}})} & \mapsto & \V{t} \\
@@
 
-Consequently, replacing $\Val{\V{ty}}$ by
-$(\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}})$ in
-Figure~\ref{fig:hexpr-full} turns the language of closed terms into a
-language of opened terms with variables and constants. For
-readability, we abbreviate $\LAM{\V{ty}}
-\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}$ by
-$\SUM{\SYMBVal}{\Var{\V{G}}{}}$. \note{Is it polite and
-  comprehensible to ask that?} This defines a new indexed description,
-called $\HExprVarD{\V{G}}$.
-
-\begin{wstructure}
-        <- evaluator, with a context
-            -> First, close variables
-                -> Perform assignment
-                <- subst [code]
-\end{wstructure}
+\newcommand{\SYMBEmpty}{\F{Empty}\xspace}
+\newcommand{\Empty}[1]{\SYMBEmpty\:#1}
+
+\newcommand{\SYMBopenTerm}{\F{openTm}\xspace}
+\newcommand{\openTerm}[1]{\SYMBopenTerm\: #1}
+\newcommand{\closeTerm}{\F{closeTm}}
+
+Consequently, taking the free monad of \(\HExprD\) by \(\SYMBVar\:
+\V{G}\), we obtain the language of open terms in a context \(\V{G}\):
+%
+\[
+\openTerm{\V{G}} \mapsto \FreeIMonad{\HExprD}{\Ty}{(\SYMBVar\:\V{G})}
+\]
+%
+The language of closed terms corresponds to the free monad assigning
+an empty set of values to variables
+%
+\[
+\closeTerm \mapsto \FreeIMonad{\HExprD}{\Ty}{\SYMBEmpty}
+\]
+%
+With
+%
+\[\stk{
+\SYMBEmpty : \Ty \To \Set \\
+\begin{array}{@{}l@{\:\mapsto\:\:}l}
+\Empty{\_}   & \Zero \\
+\end{array}
+}\]
+%
+
+\newcommand{\evalH}{\F{eval}_{\green{\Downarrow}}}
+\newcommand{\evalOne}{\F{eval}_{\green{\downarrow}}}
+
+Given the syntax, let us supply the semantics of closed terms. We
+implement an evaluator as a catamorphism:
+%
+\[\stk{
+\evalH : \PI{\V{ty}}{\Ty} 
+           \IMu{\Ty}{\closeTerm}{\V{ty}} \To
+           \Val{\V{ty}} \\
+\evalH\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty} \:
+                                 (\toIDesc{\closeTerm})\: 
+                                 \SYMBVal\: 
+                                 \evalOne\: 
+                                 \V{ty}\: 
+                                 \V{tm}
+}\]
+%
+To finish the job, we must supply the algebra which implements a single
+step of evaluation, given subexpressions evaluated already.
+%
+\[\stk{
+\evalOne : \PI{\V{ty}}{\Ty}
+ \idescop{(\toIDesc{\closeTerm})\:\V{ty}}{\Ty}{\SYMBVal}
+           \To {\Val{\V{ty}}} \\
+\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
+\evalOne\: & \_\: & (\SYMBEval\;\V{x})                   & \V{x} \\
+\evalOne\: & \_\: & (\SYMBEcond\:\BoolTrue\:\V{x}\:\_)   & \V{x} \\
+\evalOne\: & \_\: & (\SYMBEcond\:\BoolFalse\:\_\:\V{y})  & \V{y} \\
+\evalOne\: & \Enat\: & (\SYMBEplus\:\V{x}\:\V{y})   & \V{x} \plusHost \V{y} \\
+\evalOne\: & \Ebool\: & (\SYMBEle\:\V{x}\:\V{y})  & \V{x} \leHost \V{y} 
+\end{array}
+}\]
+
+Hence, we have a type-safe syntax and a tagless interpreter for our
+closed term language, in the spirit
+of~\citet{augustsson.carlsson:dependent.interpreter}, with help from
+the generic catamorphism.
 
@@
 
-Again, we would like to give a semantics to this extended language. We
-proceed in two steps: first, we replace the variables by their value
-in the context; then, we evaluate the resulting closed term. Thanks to
+Now we would like to give a semantics to the open term language. We
+proceed in two steps: first, we substitute variables by their value in
+the context; then, we evaluate the resulting closed term. Thanks to
 $\evalH$, the second problem is already solved. Let us focus on
-discharging variables from the context. Again, we can subdivide this
+substituting variables from the context. Again, we can subdivide this
 problem: first, discharging a single variable from the context; then,
@@
                \Env{\V{G}} \To 
-              (\SUM{\Val{\V{}}}{\Var{\V{G}}{}}) \DotTo
-               \IMu{\Ty}{\HExprD}{}
+               \Var{\V{G}}{} \DotTo
+               \IMu{\Ty}{\closeTerm}{}
 \end{array} \\
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumLeft \V{x})  & 
-    \Con{\pair{\SYMBEval}{\V{x}}{}} \\
-\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumRight \V{v}) &
+\discharge\: \V{G}\: \V{g}\: \V{ty}\: \V{v} &
     \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
@@
 We are now left with applying $\discharge$ over all variables of the
-term. The type of this operation is the following:
-%
-\[
+term.  We simply have to fill in the right arguments to $\substI$, the
+type guiding us:
+%
+\[
+\stk{
 \begin{array}{@{}ll}
 \substH  : & \PI{\V{G}}{\Ctxt} \\
-           & ((\SUM{\Val{}}{\Var{\V{G}}{}}) \DotTo
-              \IMu{\Ty}{\HExprD}{}) \DotTo \\
-          & \IMu{\Ty}{\HExprVarD{\V{G}}}{} \DotTo 
-            \IMu{\Ty}{\HExprD}{}
-\end{array}
-\]
-
-Abstracting away the book-keeping introduced by contexts, this
-definition looks familiar. It is similar to a monadic \bind. This is
-not surprising as we are defining a first-order syntax with variables:
-our datatype enjoys more structure than what we are given. We are
-facing a free monad, where $\SYMBEval$ is the \return introducing
-variables. Instead of implementing $\substH$ in this special case, we
-are now going to implement the \emph{free indexed-monad construction}.
-
-\subsection{Free indexed monad}
-
-\begin{wstructure}
-<- Variation on a theme: free imonad construction
-    <- Recall existence of generic free monad construction
-    -> Present its generalisation to IDesc [equation]
-        <- \I -> IDesc I as describing an indexed endofunctor
-        <- Free monad construction
-    -> Still a suitable, generic notion of substitution
-        <- show type signature
-        <- show implementation?? (space! space!)
-\end{wstructure}
-
-In Section~\ref{sec:desc-free-monad}, we have built a free monad
-operation for simple descriptions. The process is similar in the
-indexed world. Namely, given an indexed functor, we derive the indexed
-functor coding its free monad: \note{pwm: Whoa there. Maybe we should
-  say something about IMonads in general before we get to this point?}
-%
-\[\stk{
-\begin{array}{ll}
-\FreeIMonad{\_}{} : & _{\PI{\V{I}}{\Set}}
-                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
-                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
-                      \TagIDesc{\V{I}}
+           & (\Var{\V{G}}{} \DotTo
+              \IMu{\Ty}{\closeTerm}{}) \DotTo \\
+          & \IMu{\Ty}{(\openTerm{\V{G}})}{} \DotTo 
+            \IMu{\Ty}{\closeTerm}{}
 \end{array} \\
-\FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto \\ \;\;
-    \pair{\pair{\ListCons{\SYMBDVar}{(\fst{\V{E}})}} 
-               {\LAM{\V{i}}
-                \pair{\DConst{(\V{R}\: \V{i})}}
-                     {(\snd{\V{E}})\: \V{i}}{}}{}}
-         {\V{F}}{}
-}\]
-
-\newcommand{\substI}{\F{substI}}
-
-Just as in the universe of descriptions, this construction comes with
-an obvious \return and a substitution operation, the \bind. Its
-definition is the following:
-%
-\[\stk{
-\begin{array}{@{}ll}
-\substI : & _{\PI{\V{I}}{\Set}}
-            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
-            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
-          & (\V{X} \DotTo 
-             \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
-            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
-            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
-\end{array} \\
-\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
-\qquad    \cataI\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
-                      \SYMBIMu{(\FreeIMonad{\V{R}}{}{\V{Y}})}\:
-                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
-                      \V{i}\:
-                      \V{t} 
-}\]
-% 
-Where  $\F{applyI}$ is defined as follow:
-%
-\[\stk{
-\begin{array}{@{}ll}
-\F{applyI} : & _{\PI{\V{I}}{\Set}}
-            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
-            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
-          & (\V{X} \DotTo \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
-          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
-                                         {\V{I}}
-                                         {\V{X}}}}
-                    {\V{I}}
-                    {\IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
-            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
-\end{array} \\
-\begin{array}{@{}l@{\:\mapsto\:\:}l}
-\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
-\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
-\end{array}
-}\]
-
-The subscripted types corresponds to implicit arguments that can be
-automatically inferred, hence do not have to be typed in. Let us now
-consider two examples of free indexed monads.
-
-\paragraph{Typed expressions:}
-
-\begin{wstructure}
-<- Hutton's razor is a free monad
-    <- substI was our candidate bind
-        -> Massage the definition of expr to get it for free
-    <- Finding the functor
-        <- 'val is the return
-        <- The other components are the action
-        -> Updated tagged description [figure]
-    -> Compute the free monad Hutton * X
-        -> With X = Val: We get back our closed terms
-        -> With X = Val + Var vars: We get back our open terms
-\end{wstructure}
-
-In Section~\ref{sec:idesc-examples}, we had the intuition that our
-datatypes $\HExprD$ and $\HExprVarD{\V{G}}$ enjoy a monadic
-structure. We had identified the variable substitution operation as
-the \bind of a free monad. To exhibit its monadic structure, we first
-have to massage the definition of our datatype.
-
-As previously mentioned, we identify $\SYMBEval$ with the \return of
-the free monad, while the other components are the actions of the
-monad. As a result, the definition is similar to $\HExprD$ presented
-in Figure~\ref{fig:hexpr-full}, replacing $\HExprAD$ by
-$\HExprAFreeD$:
-%
-\[\stk{
-\HExprAFreeD : \ATagIDesc{\Ty} \\
-\HExprAFreeD \mapsto \bigRedBracket{
-                 \begin{array}{l}
-                   \EnumT{\sqr{\SYMBEcond}} \red{,} \\
-                   \LAM{\V{ty}}
-                   \bigRedBracket{
-                   \begin{array}{l}
-                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
-                   \end{array}
-                   }
-                 \end{array}
-                 }
-}\]
-
-We call this datatype $\HExprFreeD$. By a simple unfolding of
-definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}$
-corresponds to the syntax of closed terms, $\HExprD$. Similarly,
-$\FreeIMonad{\HExprFreeD}{\Ty}{(\SUM{\SYMBVal}{\Var{\V{G}}{}})}$
-corresponds to expressions with variables, $\HExprVarD{\V{G}}$.
-
-\begin{wstructure}
-    /> On open terms, we get a substitution
-        -> Apply substI on assgnmt
-            -> Have a well-typed language
-            -> Get a safe evaluator 
-                <- for well-typed terms 
-                <- in well-typed contexts
-\end{wstructure}
-
-The evaluator for closed terms we implemented in
-Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
-terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}\: \V{ty}$ to values in
-$\Val{\V{ty}}$. We are left with implementing $\substH$. We simply
-have to fill in the right arguments to $\substI$, the type guiding us:
-%
-\[\stk{
 \substH\: \V{G}\:
@@
           \V{\sigma}\: 
-          \V{term} \mapsto  \\
-\quad \substI_{\Ty}\:
-               (\SUM{\SYMBVal}{\SYMBVar\: \V{G}})\: 
-               \SYMBVal\:
-               \HExprFreeD\: 
+          \V{tm} \mapsto  
+\substI_{\Ty}\:
+               (\SYMBVar\: \V{G})\: 
+               \SYMBEmpty\:
+               \HExprD\: 
                \V{\sigma}\:
                \V{ty}\:
-               \V{term}
-}\]
-%
+               \V{tm}
+}\]
+
 Hence completing our implementation of the open terms interpreter.
-
 We have defined a well-typed language of arithmetical expressions,
@@
 semantics.
-
-\begin{wstructure}
-<- IDesc
-    <- Another instance of free monad
-        <- Var: Return
-        <- Remaining: (not even indexed) functor
-    -> Madness just starts
-        <- map operation
-        <- [ subst sigma D ] X = [D] (\x -> [subst sigma x] X)
-        ???
-\end{wstructure}
-
-\paragraph{Indexed descriptions:}
-
-Another instance of free monad is $\SYMBIDesc$ itself. Indeed,
-$\SYMBDVar$ is nothing but the \return. The remaining constructors are
-the carrier functor, trivially indexed by $\Unit$. The carrier functor
-is described as follow:
-%
-\[\stk{
-\IDescFreeD : \ATagIDesc{\Unit} \\
-\begin{array}{@{}ll}
-\IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
-                                \sqr{\SYMBDConst\:\:
-                                     \DProd\:\:
-                                     \SYMBDSigma\:\:
-                                     \SYMBDPi} \red{,}\\
-                                  \LAM{\_}\bigRedBracket{\begin{array}{l}
-                                        \DConst{\Set}               \\
-                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
-                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
-                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
-                                    \end{array}}\end{array}}
-\end{array}
-}\]
-%
-Then, we get $\IDesc$ by building its free monad:
-%
-\[\stk{
-\IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
-\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}\sqr{}\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
-}\]
-
-The fact that indexed descriptions are closed under substitution
-is potentially of considerable utility, if we can exploit this fact:
-\[
-\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
-    \mapsto 
-        \idescop{\V{D}}
-                {\V{I}}
-                {\LAM{\V{i}}
-                     {\idescop{\V{\sigma} \V{i}}
-                              {\V{J}}
-                              {\V{X}}}}
-        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
-\]
-By observing that a description can be decomposed via substitution, we
-split its meaning into a superstructure of substructures, e.g. a
-`database containing salaries', ready for traversal operations
-preserving the former and targeting the latter.
-
-In this section, we have presented the universe of indexed
-description. It embraces indexed families of types and, as such,
-allows us to write dependent datatypes. Hence, we have presented
-several example of indexed datatypes. In this context, we have
-presented the free monad construction, together with its monadic
-operations.

### 2010-07-25 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: 'Discussion' is the 'Related Work' section now.
Discussion rejigged

@@
 \section{Discussion}
-\label{sec:discussion}
 
-\subsection{Related work}
+In this paper, we have presented a universe of datatypes for a
+dependent type theory. We started from an unremarkable type theory
+with dependent functions and tuples, but relying on few other
+assumptions, especially where propositional equality is concerned.  We
+added finite enumeration sufficient to account for constructor choice,
+and then we built coding systems, first (as a learning experience) for
+simple ML-like inductive types, then for the indexed inductive
+families which dependently typed programmers in Agda, Coq and Epigram
+take for granted. We adopt a bidirectional type propagation mechanism
+to conceal artefacts of the encoding, giving a familiar and practicable
+constructor-based presentation to data.
+
+Crucially to our approach, we ensure that the codes describing
+datatypes inhabit a datatype with a code. In a stratified setting, we
+avoid paradox by ensuring that this type of codes lives uniformly one
+level above the types the codes describe. The adoption of ordinary
+data to describe types admits datatype-generic operations implemented
+just by ordinary programming. In working this way, we make
+considerable use of type equality modulo open computation, silently
+specialising the types of generic operations as far as the datatype
+code for any given usage is known.
+
+\subsection{Related work in Generic Programming}
 
@@
 \citet{garcia:generic-comparative-study} for a broad overview of
-generic programming in various languages. In the sole context of
-Haskell, there is a myriad of proposals. These approaches are compared
-in \citet{hinze:generic-approach-comparative} and
-\citet{rodriguez:generic-libs-comparative}.
+generic programming in various languages. For Haskell alone, there is
+a myriad of proposals: 
+\citet{hinze:generic-approach-comparative} and
+\citet{rodriguez:generic-libs-comparative} provide useful comparative
+surveys.
 
@@
 by PolyP~\cite{jansson:polyp}. Indeed, we build generic functions by
-induction on pattern functors. Unlike PolyP, we do not have to resort
-to preprocessing: our datatypes are, natively, nothing but codes.
+induction on pattern functors, exploiting type-level computation
+to avoid the preprocessing phase: our
+datatypes are, natively, nothing but codes.
 
-We share with Generic Haskell the \emph{type-indexed datatype}
-approach~\cite{hinze:generic-haskell}, as exemplified by the free
-monad construction: from a datatype, we can compute new datatypes and
-equip them with their structure. Generic Haskell also features
-\emph{generic views}~\cite{holdermans:generic-view}, transparently
-transforming the structure of datatype definitions. An example is the
-tagged descriptions, presenting datatypes under a sum-of-sigmas
-angle. Unlike Generic Haskell, we do not have to modify the compiler
-to obtain views on datatypes: we can massage descriptions from inside
-our language.
+We have the \emph{type-indexed datatypes} of Generic
+Haskell~\cite{hinze:generic-haskell} for free.
+From one datatype, we can compute
+others and equip them with relevant structure: the free monad
+construction provides one example. Our approach to encoding datatypes
+as data also sustains \emph{generic
+  views}~\cite{holdermans:generic-view}, allowing us to rebias the
+presentation of datatypes conveniently. Tagged descriptions,
+giving us a sum-of-sigmas view, are a natural example.
 
@@
 programming~\cite{hinze:polytypic-polykinded}. Our descriptions are
-limited to endofunctors on $\Set$ and $\Set^I$. While we could
-\emph{encode} higher-kinded datatypes, we do not plan to adopt this
-strategy. As future work, we plan to extend our universe to capture
-higher-kinded definitions and generic functions over them. For the
-same reason, arity-generic programming~\cite{weirich:arity-generic} is
-out of reach of our current presentation.
+limited to endofunctors on $\Set^I$. Whilst indexing is known
+to be sufficient to \emph{encode} a large class of higher-kinded
+datatypes~\citep{DBLP:conf/ifip2-1/AltenkirchM02}, we should rather
+hope to work in a more compositional style. We are free to write
+higher-order programs manipulating codes, but is not yet clear whether
+that is sufficient to deliver abstraction at higher kinds.
+Similarly, it will be interesting to see whether
+arity-generic programming~\cite{weirich:arity-generic} arises
+just by computing with our codes, or whether a richer abstraction is
+called for.
 
-Another generic programming paradigm is Scrap Your
-Boilerplate~\cite{spj:syb} (SYB). Our proposal is different in various
-ways. The corner stone of SYB is the \emph{spine} view of datatype
-constructors. A piece of data is a spine composed by a constructor
-applied to some arguments. SYB provides a combinator library to write
-generic functions over spines. This relies on a $\CN{Typeable}$
-type-class, allowing dynamic dispatch to datatype-specific
-operations. As a result, SYB is not reflexive: it is restricted to
-datatypes instantiating $\CN{Typeable}$. Moreover, it is limited to
-building generic functions, hence type-indexed datatypes cannot be
-implemented in this framework.
+The Scrap Your Boilerplate~\cite{spj:syb} (SYB) approach to generic
+programming offers a way to construct generic functions, based on
+dynamic type-testing via the $\CN{Typeable}$
+type class. SYB cannot compute types from codes, but its dynamic character
+does allow a more flexible \emph{ad hoc} approach to generic data
+traversal. By maintaining the correspondence between codes and types
+whilst supporting arbitrary inspection of codes, we pursue the same
+flexibility statically. The substitutive character of $\SYMBIDesc$
+may allow us to observe and exploit \emph{ad hoc} substructural
+relationships in data, but again, further work is needed if we
+are to make a proper comparison.
 
@@
 
-Generic programming in dependent types is not new
-either. \citet{norell:msc-thesis} has given a formalization of
-polytypic programming in Alfa, a precursor of Agda. Similarly,
-\citet{verbruggen:polytype-prog-coq, verbruggen:polytype-coq} have
-developed a framework for polytypic programming in the Coq theorem
+\subsection{Generic Programming with Dependent Types}
+
+Generic programming is not new to dependent
+types. \citet{DBLP:conf/ifip2-1/AltenkirchM02} developed a universe
+of polykinded types in Lego;
+\citet{norell:msc-thesis} gave a formalization of
+polytypic programming in Alfa, a precursor to Agda;
+\citet{verbruggen:polytype-prog-coq, verbruggen:polytype-coq} provided
+a framework for polytypic programming in the Coq theorem
 prover. However, these works aim at \emph{modelling} PolyP or Generic
@@
 
-However, we differ in several ways. First, their approach is
-generative: each universe extends the base type theory with both type
-formers and elimination rules. Thanks to levitation, we only rely on a
-generic induction and a specialised $\SYMBswitchD$. Second,
-the authors do not tackle the issue of \emph{programming} with
-codes. We have shown how to abstract away codes and give a convenient
-presentation to the developer. The authors often resort to an
-extensional equality, while we have given an equality-agnostic
-presentation. Beside, our universes are arranged so as to use
-definitional equality as much as possible. Hence, in practice, the
-developer is relieved from many proof obligations.
+However, we feel ready to offer a more radical prospectus. Their
+approach is generative: each universe extends the base type theory
+with both type formers and elimination rules. Thanks to levitation, we
+rely only on a generic induction and a specialised $\SYMBswitchD$,
+closing the type theory. We explore \emph{programming} with codes, but
+also how to conceal the encoding when writing `ordinary' programs.
+
+\subsection{Metatheoretical Status}
+
+The \(\Set:\Set\) approach we have taken in this paper is convenient
+from an experimental perspective, and it has allowed us to focus
+primarily on the encoding of universes, leaving the question
+of stratification (and with it, consistency, totality,
+and decidability of type checking) to one side. However, we must
+surely face up to the latter, especially since we have taken up the
+habit of constructing `the set of all sets'. A proper account requires
+a concrete proposal for a system of stratified universes which allows
+us to make `level-polymorphic' constructions, and we are actively
+pursuing such a proposal. We hope soon to have something to prove.
+
+In the meantime, we can gain some confidence by systematically
+embedding predicative fragments of our theory within
+systems which already offer a universe hierarchy. We can, at the
+very least, confirm that in UTT-style theories with conventional
+inductive families of types~\citep{luo:utt}, as found in Coq
+(and in Agda if one avoids experimental extensions), we build the
+tower of universes we propose, cut off at an arbitrary height.
+It is correspondingly clear that some such system can be made to
+work, or else that other, longer-standing tools are troubled.
+
+A metatheoretical issue open at time of writing concerns the
+size of the \emph{index} set \(\V{I}\) in \(\IDesc{\V{I}}\). Both Agda
+and recent versions of Coq allow inductive families with \emph{large}
+indices, effectively allowing `higher-kind' fixpoints on
+\(\Set^{\Set}\) and more. They retain the safeguard that the types of
+\emph{substructures} must be as small as the inductively
+defined superstructure. This liberalisation allows us large index sets
+in our models, but whilst it offers no obvious route to paradox by
+smuggling a large universe inside a small type, it is not yet known
+to be safe. We can restrict \(\V{I}\) as necessary to avoid
+paradox, provided \(\Unit\), used to index \(\SYMBIDesc\) itself, is
+`small'.
+
+\subsection{Further Work}
+
+Apart from the need to nail down a stratified version of the system and
+its metatheory, we face plenty of further problems and opportunities.
+Although we have certainly covered Luo's criteria for inductive
+families~\cite{luo:utt}, there are several dimensions in which to
+consider expanding our universe.
+
+Firstly, we seek to encompass
+\emph{inductive-recursive} datatype families~\cite{dybjer:iir}, allowing
+us to interleave the definition and interpretation of data in intricate
+and powerful ways. This interleaving seems particularly useful when
+reflecting the syntax of dependent type systems.
+
+Secondly, we should very much like to extend our
+universe with a codes for internal fixpoints, as
+in~\cite{DBLP:conf/types/MorrisAM04}. The external knot-tying
+approach we have taken here makes types like `trees with lists of subtrees'
+more trouble than they should be. Moreover, if we allow the alternation
+of least and greatest fixpoints, we should expect to gain types which are
+not readily encoded with one external \(\SYMBIMu\).
+
+Thirdly, it would be fascinating to extend our universe with dedicated
+support for syntax with \emph{binding}, not least because a universe
+with internal fixpoints has such a syntax. Harper and Licata have
+demonstrated the potential for and of such an
+encoding~\citep{DBLP:conf/icfp/LicataH09}, boldly encoding the invalid
+definitions along with the valid. A more conservative strategy might
+be to offer improved support for datatypes indexed by an extensible
+context of free variables, with the associated free monad structure
+avoiding capture as shown by~\citet{altenkirch:monadic-lambda}.
+
+Lastly, we must ask how our new presentation of datatypes
+should affect the tools we use to build software. It is not enough to
+change the game: we must enable better play. If datatypes are data,
+what is design?

### 2010-07-25 papers/icfp-2010-desc/paper_desc_levitation.tex
level annotations in levitations

@@
 the signatures, first for the enumerations, and then for the codes of
-the $\Desc$ universe.  Persuading this to perform was a perilous
+the $\Desc^n$ universe.  Persuading these programs to perform was a perilous
 pedagogical peregrination for the protagonist.  Our method was indeed to
@@
 \emph{definitions}: ``Is \(2+2\) still \(4\)?'', ``No, it's a loop!''.
-But we did find a way, so now we hope to convey to the reader the
+But we did find a way, so now we hope to convey to you the
 dizzy feeling of levitation, without the falling.
@@
 In Section~\ref{sec:finite-sets}, we specified the
-finite sets of tags. We are going to implement the $\EnumU$ type former and
-its constructors. Recall:
+finite sets of tags. We are going to implement (at every universe level)
+the $\EnumU$ type former and its constructors. Recall:
 %
-\[\Type{\EnumU}\qquad\Bhab{\NilE}{\EnumU}\qquad
+\[\Type{\EnumU}^n\qquad\Bhab{\NilE}{\EnumU}\qquad
 \Bhab{\ConsE{(\Bhab{\M{t}}{\UId})}{(\Bhab{\M{E}}{\EnumU})}}{\EnumU}
@@
 
-Let us consider the consequences. We discover that the type
-theory does not need to be extended with a special type former $\EnumU$,
+Let us consider the consequences. We find that the type
+theory does not need a special type former $\EnumU$,
 or special constructors $\NilE$ and $\SYMBConsE$. 
@@
 We shall now fulfil our implementation promises, encoding the universe
-of descriptions. In and of itself, the codes, $\Desc$, is nothing but
-a datatype. We are in the same situation as with $\EnumU$: we ought to
-be able to describe the codes of $\Desc$ in $\Desc$ itself. Hence,
-this code would be a first-class citizen, born with the standard,
-generic equipment of datatypes.
+of descriptions. $\Desc$ is already some sort of datatype, but we are
+in the same situation as with $\EnumU$: we ought to be able to
+describe it, coding of $\Desc^n$ in $\Desc^{n+1}$, spiralling
+upwards. Hence, this code would be a first-class citizen, born with
+the generic equipment of datatypes.
 
@@
 \[\stk{
-\DescD : \Desc \\
-\DescD \mapsto
+\DescD^n : \Desc^{n+1} \\
+\DescD^n \mapsto
   \tada{\stkm{\DUnit\\ \SYMBDSigma\\ \SYMBDIndx}}
        {\stkm{ \DUnit                                \\
-         \DSigma{\Set}{\LAM{\V{S}} \SHED}      \\
+         \DSigma{\Set^n}{\LAM{\V{S}} \SHED}      \\
          \DIndx{\DUnit}    }                    } \\
@@
 we have done so far, we first offer a constructor choice from
-$\DUnit$, $\SYMBDSigma$, and $\SYMBDIndx$. The reader will notice
-that the `tagged' notation we have used for the $\Desc$ constructors now
-fully makes sense: these were actually the tags we are defining.
+$\DUnit$, $\SYMBDSigma$, and $\SYMBDIndx$. You may notice
+that the `tagged' notation we have used for the $\Desc^n$ constructors now
+fits the facts: these were actually the tags we are defining.
 For $\DUnit$, we immediately reach the end of the description. For
@@
 \[
-\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
+\DSigma{(\Bhab{\V{S}}{\Set^n})}{(\Bhab{\V{D}}{\V{S} \To \Desc^n})} : \Desc^n\\
 \]
 %
-So, we first pack a  $\Set$, $\V{S}$. We should then like
+So, we first pack a  $\Set^n$, $\V{S}$, as well we might when working in
+$\Desc^{n+1}$. We should then like
 a recursive argument \emph{indexed} by $\V{S}$, but
-that is an \emph{exponential}, and our presentation is entirely
-first-order so far, delivering only sums-of-products. To code our
-universe, we must first enlarge it!
+that is an \emph{exponential}, and our presentation so far delivers only
+sums-of-products. To code our universe, we must first enlarge it!
 
@@
 \[\stk{
-    \DHindx{(\Bhab{\M{H}}{\Set})}{(\Bhab{\M{D}}{\Desc})} : \Desc
+    \DHindx{(\Bhab{\M{H}}{\Set^n})}{(\Bhab{\M{D}}{\Desc^n})} : \Desc^n
     \smallskip \\
@@
 representations are finitary by construction, and thus admit a richer,
-componentwise decidable equality that functions may in
+componentwise decidable equality than functions may in
 general possess.\footnote{E.g., extensionally, there is one inhabitant of
@@
 \[\stk{
-\DescD : \Desc \\
-\DescD \mapsto\tada
+\DescD^n : \Desc^{n+1} \\
+\DescD^n \mapsto\tada
 {\stkm{ \DUnit\\ \SYMBDSigma\\ \SYMBDIndx\\ \SYMBDHindx}}
@@
 \DUnit                                            \\
-                                   \DSigma{\Set}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
+                                   \DSigma{\Set^n}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
                                    \DIndx{\DUnit}                                    \\
-                                   \DSigma{\Set}{\LAM{\_} \DIndx{\DUnit}}
+                                   \DSigma{\Set^n}{\LAM{\_} \DIndx{\DUnit}}
 }}
@@
 The $\DUnit$ and $\SYMBDIndx$ cases remain unchanged, as expected. We
-successfully describe the $\SYMBDSigma$ case, by a simple appeal to
-the higher-order induction on $\V{S}$. The $\SYMBDHindx$ case
-consists in packing a $\Set$ with a recursive argument.
+successfully describe the $\SYMBDSigma$ case via
+the higher-order induction, branching on $\V{S}$. The $\SYMBDHindx$ case
+just packs a $\Set^n$ with a recursive argument.
 
 At a first glance, we have achieved our goal. We have described the
-codes of the universe of descriptions. Taking the fixpoint of \(\descop{\DescD}\)
-gives us a datatype exactly like $\Desc$. Might we be
-so bold as to take \(\Desc \mapsto \Mu{\DescD}\) as the levitating
+codes of the universe of descriptions. The fixpoint of \(\descop{\DescD^n}\)
+is a datatype just like $\Desc^n$, in $\Set^{n+1}$. Might we be
+so bold as to take \(\Desc^n \mapsto \Mu{\DescD^n}\) as the levitating
 definition? If we do, we shall come down with a bump! 
@@
 
-The definition \(\Desc \mapsto \Mu{\DescD}\) is circular,
+The definition \(\Desc^n \mapsto \Mu{\DescD^n}\) is circular,
 but the offensive recursion is concealed by a prestidigitation.
@@
 \[
-\Desc\mapsto
+\Desc^n\mapsto
 \Mu{\stk{(\DSigma{\EnumT{\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}}\\}
-    {\;\SYMBswitch\:\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}\:(\LAM{\_}\Desc)\\
+    {\;\SYMBswitch\:\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}\:(\LAM{\_}\Desc^{n+1})\\
 \;\sqr{\stkm{
 \DUnit                                            \\
-                                   \DSigma{\Set}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
+                                   \DSigma{\Set^n}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
                                    \DIndx{\DUnit}                                    \\
-                                   \DSigma{\Set}{\LAM{\_} \DIndx{\DUnit}}
+                                   \DSigma{\Set^n}{\LAM{\_} \DIndx{\DUnit}}
 }}})}}
@@
 The recursion shows up only because we must specify the return type
-of the general-purpose \(\F{switch}\), and it is computing a \(\Desc\)!
+of the general-purpose \(\F{switch}\), and it is computing a \(\Desc^{n+1}\)!
 Although type propagation
@@
 We are too close to give up now. If only we did not need to
-supply that return type, especially when we know what it must be.
+supply that return type, especially when we know what it must be!
 We eliminate the recursion by \emph{specialising} \(\F{switch}\):
@@
 \F{switchD} : \PITEL{\V{E}}{\EnumU}   \To
-                (\spi{\V{E}}{\LAM{\_} \Desc}) \To
-                \EnumT{\V{E}} \To \Desc
+                (\spi{\V{E}}{\LAM{\_} \Desc^m}) \To
+                \EnumT{\V{E}} \To \Desc^m
 \]
@@
 \Rule{\Gamma \Vdash
-  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{\PLAM{x}{\EnumT{E}}\Desc}}}
+  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{\PLAM{x}{\EnumT{E}}\Desc^m}}}
                            {\M{t'}}}
      {\Gamma \Vdash
-\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc}}
+\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc^m}}
  {\switchD{\M{E}}{\M{t'}}}}
@@
 introducing recursion. Of course, by pasting together the declaration
-of \(\Desc\) and its internal copy, we have made it appear in its own
+of \(\Desc^n\) and its internal copy, we have made it appear in its own
 type. Hardwired as a \emph{fait accompli}, this creates no regress,
-although one must assume the definition to recheck it\footnote{Our
-  Agda model does not formalize the $\SYMBswitchD$
-  construction. However, it proves the isomorphism between the defined
-  descriptions and the coded ones. The role of $\SYMBswitchD$ is to
-  collapse this isomorphism, hence identifiying defined and coded
-  descriptions.}. Alternatively, we could have extended $\Desc$ with a
-code for \emph{finite} sums, internalizing the $\SYMBswitch$ in the
-interpretation function:
+although one must assume the definition to recheck it.
+
+Our Agda model does not formalize the $\SYMBswitchD$
+construction. Instead, we exhibit the isomorphism between declared
+and encoded descriptions. Here, $\SYMBswitchD$ lets us collapse
+this isomorphism, operationally identifying defined and coded
+descriptions.
+
+There are other ways to achieve a sufficient specialisation to avoid a
+recursive code, e.g., extending $\Desc^n$ with specialised codes for
+\emph{finite} sums and products, pushing the $\SYMBswitch$ into the
+interpretation of codes, rather than the code itself. Here, we prefer
+not to add codes to $\Desc^n$ which are otherwise unmotivated.
 %%
-\[\stk{
-    \Dsigma{(\Bhab{\V{E}}{\EnumU})}
-           {(\Bhab{\V{B}}{\spi{\V{E}}{\LAM{\_}{\Desc}}})} : \Desc
-    \smallskip \\
-    \descop{\Dsigma{\V{E}}{\V{B}}}{\V{X}}     
-        \mapsto 
-            \SIGMAS{\Bhab{\V{x}}{\EnumT{\V{E}}}}
-                   {\descop{\switch{\V{E}}{(\LAM{\_}{\Desc})}{\V{B}}{\V{x}}}{\V{X}}}    
-}\]
+%\[\stk{
+%    \Dsigma{(\Bhab{\V{E}}{\EnumU})}
+%           {(\Bhab{\V{B}}{\spi{\V{E}}{\LAM{\_}{\Desc^n}}})} : \Desc^n
+%    \smallskip \\
+%    \descop{\Dsigma{\V{E}}{\V{B}}}{\V{X}}     
+%        \mapsto 
+%            \SIGMAS{\Bhab{\V{x}}{\EnumT{\V{E}}}}
+%                   %{\descop{\switch{\V{E}}{(\LAM{\_}{\Desc})}{\V{B}}{\V{x}}}{\V{X}}}    
+%}\]
 %%
-Using $\SYMBDsigma$ to switch over the $\Desc$ constructors, we avoid
-mention of $\Desc$ while levitating it. The $\SYMBswitch$ operator is
-pushed into the interpretation function, at which point $\Desc$ indeed
-exists.
+%Using $\SYMBDsigma$ to switch over the $\Desc$ constructors, we avoid
+%mention of $\Desc$ while levitating it. The $\SYMBswitch$ operator is
+%pushed into the interpretation function, at which point $\Desc$ indeed
+%exists.
 
@@
 
-We have levitated \(\Desc\). Beyond its pedagogical value, this
-exercise has several practical outcomes. First of all, it reveals that
-the $\Desc$ universe is just plain data. As any piece of data, it can
-therefore be inspected and manipulated. Moreover, it is expressed in
-the $\Desc$ universe. As a consequence, it is equipped, for free, with
-an induction principle. So, our ability to inspect and program with
-$\Desc$ is not restricted to a meta-language: we now have all the
-necessary equipment in the theory to \emph{program} over
-datatypes. \emph{Generic programming is just
-  programming}.
+We have levitated \(\Desc\) at every level. Beyond its pedagogical
+value, this exercise has several practical outcomes. First, it
+confirms that each $\Desc$ universe is just plain data. As any piece of
+data, it can therefore be inspected and manipulated. Moreover, it is
+expressed in a $\Desc$ universe. As a consequence, it is equipped,
+for free, with an induction principle. So, our ability to inspect and
+program with $\Desc$ is not restricted to a meta-language: we have
+the necessary equipment to program with \emph{data}, so we can program over
+datatypes. \emph{Generic programming is just programming}.
 
@@
 In this section, we have seen how to \emph{levitate}
-descriptions. From this unusual situation, several questions naturally
-arise. First, we have blurred the line between the implementation of
-data-types and their self-description. Second, we have abused the
-paradoxical nature of $\Bhab{\Set}{\Set}$ to flatten the hierarchy of
-descriptions. We shall now be more precise about these points.
+descriptions. Although our theory, as presented here, takes
+$\Bhab{\Set}{\Set}$, our annotations indicate how a stratified theory
+could code each level from above.  We do not rely on the paradoxical
+nature of $\Bhab{\Set}{\Set}$ to flatten the hierarchy of descriptions
+and fit large inside small.  We shall now be more precise about what
+we have done.
 
@@
 
-To demystify the apparent circularity of our construction, let us
-consider how it bottoms out in a stratified setting. Using the
-informal level annotations of Section~\ref{sec:universe-desc}, we have
-that \(\Bhab{\DescD^n}{\Desc^{n+1}}\), hence \(\Bhab{\Desc^n \mapsto
-  \Mu{\DescD^n}}{\Set^{n+1}}\). Assuming an hardwired $\Desc^N$
-universe, we can build the tower of descriptions down to $\Desc^0$:
-description of descriptions forms a `spiral', rather than a
-circle \footnote{We have modelled this process in Agda, hardwiring
-  $\Desc^{42}$ in $\Set^{43}$, then going down. This model does not
-  rely on dependent pattern-matching, induction-recursion, nor set
-  polymorphism.}. Hence, in a stratified system, the `self-encoded'
-nature of $\Desc$ appears only in a set polymorphic sense: the
-principal type of the levitating description generalizes to the type
-of $\Desc$ itself\footnote{This is witnessed in our set polymorphic
-  model in Agda and in our Coq model, crucially relying on typical
-  ambiguity~\cite{harper:implicit-universe}.}.
+In a stratified system, the `self-encoded' nature of $\Desc$ appears
+only in a set polymorphic sense: the principal type of the encoded
+description generalizes to the type of $\Desc$ itself. We encode this
+much in our set polymorphic model in Agda and in our Coq model,
+crucially relying on typical ambiguity~\cite{harper:implicit-universe}.
+We step outside current technology only to replace the declared $\Desc$
+with its encoding.
+
+Even this last step, we can approximate within a standard predicative
+hierarchy. Fix a top level, perhaps \(42\). We may start by declaring
+$\Desc^{42}:\Set^{43}$. We can then construct $\DescD^{41}:\Desc^{42}$
+and thus acquire an encoded $\Desc^{41}$. Although $\Desc^{41}$ is
+encoded, not declared, it includes the relevant descriptions,
+including $\DescD^{40}$. We can thus build the tower of descriptions
+down to $\Desc^0$, encoding every level below the top.  Description of
+descriptions forms a `spiral', rather than a circle. We have modelled
+this process exactly in Agda, without any appeal to dependent
+pattern-matching, induction-recursion, or set polymorphism. All it
+takes to build such a sawn-off model of encodings is inductive
+definition and a cumulative predicative hierarchy of set levels.
 

### 2010-07-25 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: freemonad fits on one line
Desc@ICFP: use \muide for tagged stuffs
Desc@ICFP: fix spacing, fix redundance.
Desc@ICFP: re-re-shaping of freemo

@@
 
-\newcommand{\muide}[2]{\D{\({\mu}^{\!+}\)}\!\!_{#1}\:#2}
+\newcommand{\SYMBmuide}{\D{\({\mu}^{\!+}\)}\xspace}
+\newcommand{\muide}[2]{\SYMBmuide\!\!_{#1}\:#2}
 
@@
 $\IMu{\V{I}}{(\toIDesc{\V{D}})}$.
-
-\subsection{Free indexed monad}
-
-\begin{wstructure}
-<- Variation on a theme: free imonad construction
-    <- Recall existence of generic free monad construction
-    -> Present its generalisation to IDesc [equation]
-        <- \I -> IDesc I as describing an indexed endofunctor
-        <- Free monad construction
-    -> Still a suitable, generic notion of substitution
-        <- show type signature
-        <- show implementation?? (space! space!)
-\end{wstructure}
-
-In Section~\ref{sec:desc-free-monad}, we have built a free monad
-operation for simple descriptions. The process is similar in the
-indexed world. Namely, given an indexed functor, we derive the indexed
-functor coding its free monad: \note{pwm: Whoa there. Maybe we should
-  say something about IMonads in general before we get to this point?}
-%
-\[\stk{
-\begin{array}{ll}
-\FreeIMonad{\_}{} : & _{\PI{\V{I}}{\Set}}
-                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
-                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
-                      \TagIDesc{\V{I}}
-\end{array} \\
-\FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto \\ \;\;
-    \pair{\pair{\ListCons{\SYMBDVar}{(\fst{\V{E}})}} 
-               {\LAM{\V{i}}
-                \pair{\DConst{(\V{R}\: \V{i})}}
-                     {(\snd{\V{E}})\: \V{i}}{}}{}}
-         {\V{F}}{}
-}\]
-
-\newcommand{\substI}{\F{substI}}
-
-Just as in the universe of descriptions, this construction comes with
-an obvious \return and a substitution operation, the \bind. Its
-definition is the following:
-%
-\[\stk{
-\begin{array}{@{}ll}
-\substI : & _{\PI{\V{I}}{\Set}}
-            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
-            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
-          & (\V{X} \DotTo 
-             \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
-            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
-            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
-\end{array} \\
-\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
-\qquad    \cataI\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
-                      \SYMBIMu{(\FreeIMonad{\V{R}}{}{\V{Y}})}\:
-                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
-                      \V{i}\:
-                      \V{t} 
-}\]
-% 
-Where  $\F{applyI}$ is defined as follow:
-%
-\[\stk{
-\begin{array}{@{}ll}
-\F{applyI} : & _{\PI{\V{I}}{\Set}}
-            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
-            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
-          & (\V{X} \DotTo \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
-          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
-                                         {\V{I}}
-                                         {\V{X}}}}
-                    {\V{I}}
-                    {\IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
-            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
-\end{array} \\
-\begin{array}{@{}l@{\:\mapsto\:\:}l}
-\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
-\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
-\end{array}
-}\]
-
-The subscripted types corresponds to implicit arguments that can be
-automatically inferred, hence do not have to be typed in. Let us now
-consider two examples of free indexed monads.
-
-\paragraph{Indexed descriptions:}
-
-An interesting instance of free monad is $\SYMBIDesc$ itself. Indeed,
-$\SYMBDVar$ is nothing but the \return. The remaining constructors form
-the carrier functor, trivially indexed by $\Unit$. The carrier functor
-is described as follow:
-%
-\[\stk{
-\IDescFreeD : \ATagIDesc{\Unit} \\
-\begin{array}{@{}ll}
-\IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
-                                \sqr{\SYMBDConst\:\:
-                                     \DProd\:\:
-                                     \SYMBDSigma\:\:
-                                     \SYMBDPi} \red{,}\\
-                                  \LAM{\_}\bigRedBracket{\begin{array}{l}
-                                        \DConst{\Set}               \\
-                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
-                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
-                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
-                                    \end{array}}\end{array}}
-\end{array}
-}\]
-%
-Then, we get $\IDesc$ by building its free monad:
-%
-\[\stk{
-\IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
-\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}\sqr{}\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
-}\]
-
-The fact that indexed descriptions are closed under substitution
-is potentially of considerable utility, if we can exploit this fact:
-\[
-\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
-    \mapsto 
-        \idescop{\V{D}}
-                {\V{I}}
-                {\LAM{\V{i}}
-                     {\idescop{\V{\sigma} \V{i}}
-                              {\V{J}}
-                              {\V{X}}}}
-        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
-\]
-By observing that a description can be decomposed via substitution, we
-split its meaning into a superstructure of substructures, e.g. a
-`database containing salaries', ready for traversal operations
-preserving the former and targeting the latter. We shall now turn to
-some examples of indexed descriptions, as well as a practical use of
-free monad construction.
 
@@
 
-This language has values, conditional
-expression, addition and comparison. Informally, their types are:
+This language has values, conditional expression, addition and
+comparison. Informally, their types are:
 %
@@
 expected type.
+%
 \[\stk{
@@
 }\]
-We take $\Nat$ and $\Bool$ to represent
-natural numbers and Booleans in the host language, equipped with
-addition $\plusHost$ and comparison $\leHost$.
-
-We describe our syntax in two steps. First, we give the pattern
-functor of the free monad, defining the operators of the
-language. Second, we apply the free monad construction to obtain a
-language of closed terms and a language of open terms. The operations
-are described by a tagged indexed description, indexing over object
-language types $\Ty$. We note that some constructors are always
+%
+We take $\Nat$ and $\Bool$ to represent natural numbers and Booleans
+in the host language, equipped with addition $\plusHost$ and
+comparison $\leHost$.
+
+We express our syntax as a tagged indexed description, indexing over
+object language types $\Ty$. We note that some constructors are always
 available, namely $\SYMBEval$ and $\SYMBEcond$. On the other hand,
@@
 
+\newcommand{\evalH}{\F{eval}_{\green{\Downarrow}}}
+\newcommand{\evalOne}{\F{eval}_{\green{\downarrow}}}
+
+Given the syntax, let us supply the semantics. We implement an
+evaluator as a catamorphism:
+%
+\[\stk{
+\evalH : \PI{\V{ty}}{\Ty} 
+         \muide{\Ty}{\HExprD}\: \V{ty} \To
+         \Val{\V{ty}} \\
+\evalH\: \V{ty}\: \V{term} \mapsto \cataI_{\Ty} \:
+                                 (\toIDesc{\HExprD})\: 
+                                 \SYMBVal\: 
+                                 \evalOne\: 
+                                 \V{ty}\: 
+                                 \V{term}
+}\]
+%
+To finish the job, we must supply the algebra which implements a single
+step of evaluation, given subexpressions evaluated already.
+%
+\[\stk{
+\evalOne : \PI{\V{ty}}{\Ty}
+ \idescop{(\toIDesc{\HExprD})\:\V{ty}}{\Ty}{\SYMBVal}
+           \To {\Val{\V{ty}}} \\
+\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
+\evalOne\: & \_\: & (\SYMBEval\;\V{x})                                            & \V{x} \\
+\evalOne\: & \_\: & (\SYMBEcond\:\BoolTrue\:\V{x}\:\_)   & \V{x} \\
+\evalOne\: & \_\: & (\SYMBEcond\:\BoolFalse\:\_\:\V{y})  & \V{y} \\
+\evalOne\: & \Enat\: & (\SYMBEplus\:\V{x}\:\V{y})   & \V{x} \plusHost \V{y} \\
+\evalOne\: & \Ebool\: & (\SYMBEle\:\V{x}\:\V{y})  & \V{x} \leHost \V{y} 
+\end{array}
+}\]
+
+\begin{wstructure}
+    /> Closed term
+        <- only constants and operations on them
+        -> Extend Val with Var : Ty -> Set, mapping to EnumU
+            -> Open term
+            -> Language of well-typed terms
+                <- By construction
+\end{wstructure}
+
+Hence, we have a type-safe syntax and a tagless interpreter for our
+language, in the spirit
+of~\citet{augustsson.carlsson:dependent.interpreter}, with help from
+the generic catamorphism. However, so far, we are only able to define
+and manipulate \emph{closed} terms. Adding variables, it is possible
+to build and manipulate \emph{open} terms, that is, terms in a
+context. We shall get this representation, for free, thanks to the
+\emph{free indexed monad} construction.
+
+\subsection{Free indexed monad}
+
+\begin{wstructure}
+<- Variation on a theme: free imonad construction
+    <- Recall existence of generic free monad construction
+    -> Present its generalisation to IDesc [equation]
+        <- \I -> IDesc I as describing an indexed endofunctor
+        <- Free monad construction
+    -> Still a suitable, generic notion of substitution
+        <- show type signature
+        <- show implementation?? (space! space!)
+\end{wstructure}
+
+In Section~\ref{sec:desc-free-monad}, we have built a free monad
+operation for simple descriptions. The process is similar in the
+indexed world. Namely, given an indexed functor, we derive the indexed
+functor coding its free monad: \note{pwm: Whoa there. Maybe we should
+  say something about IMonads in general before we get to this point?}
+%
+\[\stk{
+\begin{array}{ll}
+\FreeIMonad{\_}{} : & _{\PI{\V{I}}{\Set}}
+                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
+                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
+                      \TagIDesc{\V{I}}
+\end{array} \\
+\FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto
+    \pair{\pair{\ListCons{\SYMBDVar}{(\fst{\V{E}})}} 
+               {\LAM{\V{i}}
+                \pair{\DConst{(\V{R}\: \V{i})}}
+                     {(\snd{\V{E}})\: \V{i}}{}}{}}
+         {\V{F}}{}
+}\]
+
+\newcommand{\substI}{\F{substI}}
+
+Just as in the universe of descriptions, this construction comes with
+an obvious \return and a substitution operation, the \bind. Its
+definition is the following:
+%
+\[\stk{
+\begin{array}{@{}ll}
+\substI : & _{\PI{\V{I}}{\Set}}
+            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
+            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
+          & (\V{X} \DotTo 
+             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
+             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
+             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
+\end{array} \\
+\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
+\qquad    \cataI_{\V{I}}\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
+                      (\muide{\V{Y}}{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
+                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
+                      \V{i}\:
+                      \V{t} 
+}\]
+% 
+Where  $\F{applyI}$ is defined as follow:
+%
+\[\stk{
+\begin{array}{@{}ll}
+\F{applyI} : & _{\PI{\V{I}}{\Set}}
+            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
+            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
+          & (\V{X} \DotTo \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
+          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
+                                         {\V{I}}
+                                         {\V{X}}}}
+                    {\V{I}}
+                    {\muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
+            \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
+\end{array} \\
+\begin{array}{@{}l@{\:\mapsto\:\:}l}
+\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
+\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
+\end{array}
+}\]
+
+The subscripted types corresponds to implicit arguments that can be
+automatically inferred, hence do not have to be typed in. Let us now
+consider two examples of free indexed monads.
+
+\paragraph{Typed expressions:}
+
 \begin{wstructure}
@@
 
-However, so far, we are only able to define and manipulate
-\emph{closed} terms. Adding variables, it is possible to build and
-manipulate \emph{open} terms, that is, terms in a context. The context
-is represented by a snoc-list of types:
+In the previous section, we have presented a language of closed
+arithmetic expressions. Using the free monad construction, we are
+going to extend this construction to open terms. An open term is
+defined with respect to a context, represented by a snoc-list of
+types:
 %
@@
 
+\newcommand{\update}{\F{update}}
+
 Consequently, taking the free monad of \(\HExprD\) by \(\SYMBVar\:
@@
 %
-The language of closed terms corresponds to the free monad assigning
-an empty set of values to variables
+In this setting, the language of closed terms corresponds to the free
+monad assigning an empty set of values to variables
 %
@@
 \SYMBEmpty : \Ty \To \Set \\
-\begin{array}{@{}l@{\:\mapsto\:\:}l}
+\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
 \Empty{\_}   & \Zero \\
@@
 %
-
-\newcommand{\evalH}{\F{eval}_{\green{\Downarrow}}}
-\newcommand{\evalOne}{\F{eval}_{\green{\downarrow}}}
-
-Given the syntax, let us supply the semantics of closed terms. We
-implement an evaluator as a catamorphism:
-%
-\[\stk{
-\evalH : \PI{\V{ty}}{\Ty} 
-           \IMu{\Ty}{\closeTerm}{\V{ty}} \To
-           \Val{\V{ty}} \\
-\evalH\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty} \:
-                                 (\toIDesc{\closeTerm})\: 
-                                 \SYMBVal\: 
-                                 \evalOne\: 
+Obviously, \(\closeTerm\) and \(\HExprD\) describe isomorphic
+datatypes. For instance, you can update your old \(\HExprD\) to a
+shiny \(\closeTerm\) with \(\update\):
+%
+\[\stk{
+\update : \muide{\Ty}{\HExprD} \DotTo \muide{\Ty}{\closeTerm} \\
+\begin{array}{@{}l@{}l}
+\update\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty}\: & \Ty\:
+                                                  (\toIDesc{\HExprD})\:
+                                                  (\muide{\Ty}{\closeTerm})\: \\
+                                                & (\LAM{\_} \LAM{\pair{\V{tag}}{\V{tm}}{}} \Con{\pair{\Su{\V{tag}}}{\V{tm}}{}})\:
                                                   \V{ty}\:
                                                   \V{tm}
-}\]
-%
-To finish the job, we must supply the algebra which implements a single
-step of evaluation, given subexpressions evaluated already.
-%
-\[\stk{
-\evalOne : \PI{\V{ty}}{\Ty}
- \idescop{(\toIDesc{\closeTerm})\:\V{ty}}{\Ty}{\SYMBVal}
-           \To {\Val{\V{ty}}} \\
-\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
-\evalOne\: & \_\: & (\SYMBEval\;\V{x})                   & \V{x} \\
-\evalOne\: & \_\: & (\SYMBEcond\:\BoolTrue\:\V{x}\:\_)   & \V{x} \\
-\evalOne\: & \_\: & (\SYMBEcond\:\BoolFalse\:\_\:\V{y})  & \V{y} \\
-\evalOne\: & \Enat\: & (\SYMBEplus\:\V{x}\:\V{y})   & \V{x} \plusHost \V{y} \\
-\evalOne\: & \Ebool\: & (\SYMBEle\:\V{x}\:\V{y})  & \V{x} \leHost \V{y} 
-\end{array}
-}\]
-
-Hence, we have a type-safe syntax and a tagless interpreter for our
-closed term language, in the spirit
-of~\citet{augustsson.carlsson:dependent.interpreter}, with help from
-the generic catamorphism.
+\end{array}
+}\]
+% 
+The other direction of the isomorphism is straightforward, the
+$\SYMBDVar$ case being impossible. Therefore, we are entitled to
+reuse the $\evalH$ function to define the semantics of $\closeTerm$.
 
@@
                \Var{\V{G}}{} \DotTo
-               \IMu{\Ty}{\closeTerm}{}
+               \muide{\Ty}{\closeTerm}
 \end{array} \\
@@
            & (\Var{\V{G}}{} \DotTo
-              \IMu{\Ty}{\closeTerm}{}) \DotTo \\
-          & \IMu{\Ty}{(\openTerm{\V{G}})}{} \DotTo 
-            \IMu{\Ty}{\closeTerm}{}
+              \muide{\Ty}{\closeTerm}) \DotTo \\
+          & \muide{\Ty}{(\openTerm{\V{G}})} \DotTo 
+            \muide{\Ty}{\closeTerm}
 \end{array} \\
@@
 open terms. Using its monadic structure, we have implemented the
-interpreter for open terms in context. Hence, without much effort, we
+interpreter for open terms in context. Without much effort, we
 have described the syntax of a well-typed language, together with its
 semantics.
+
+\paragraph{Indexed descriptions:}
+
+An interesting instance of free monad is $\SYMBIDesc$ itself. Indeed,
+$\SYMBDVar$ is nothing but the \return. The remaining constructors form
+the carrier functor, trivially indexed by $\Unit$. The carrier functor
+is described as follow:
+%
+\[\stk{
+\IDescFreeD : \ATagIDesc{\Unit} \\
+\begin{array}{@{}ll}
+\IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
+                                \sqr{\SYMBDConst\:\:
+                                     \DProd\:\:
+                                     \SYMBDSigma\:\:
+                                     \SYMBDPi} \red{,}\\
+                                  \LAM{\_}\bigRedBracket{\begin{array}{l}
+                                        \DConst{\Set}               \\
+                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
+                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
+                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
+                                    \end{array}}\end{array}}
+\end{array}
+}\]
+%
+Then, we get $\IDesc$ by building its free monad:
+%
+\[\stk{
+\IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
+\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}\sqr{}\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
+}\]
+
+The fact that indexed descriptions are closed under substitution
+is potentially of considerable utility, if we can exploit this fact:
+\[
+\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
+    \mapsto 
+        \idescop{\V{D}}
+                {\V{I}}
+                {\LAM{\V{i}}
+                     {\idescop{\V{\sigma} \V{i}}
+                              {\V{J}}
+                              {\V{X}}}}
+        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
+\]
+By observing that a description can be decomposed via substitution, we
+split its meaning into a superstructure of substructures, e.g. a
+`database containing salaries', ready for traversal operations
+preserving the former and targeting the latter. We shall now turn to
+some practical use of the free monad construction.

### 2010-07-25 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix \update

@@
 \begin{array}{@{}l@{}l}
-\update\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty}\: & \Ty\:
-                                                  (\toIDesc{\HExprD})\:
+\update\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty}\: & (\toIDesc{\HExprD})\:
                                                   (\muide{\Ty}{\closeTerm})\: \\

### 2010-07-25 papers/icfp-2010-desc/paper_idesc.tex
gaol: no Obviously

@@
 \closeTerm \mapsto \FreeIMonad{\HExprD}{\Ty}{\SYMBEmpty}
-\]
-%
-With
-%
-\[\stk{
+\quad
+\mbox{where}
+\quad
+\stk{
 \SYMBEmpty : \Ty \To \Set \\
@@
 %
-Obviously, \(\closeTerm\) and \(\HExprD\) describe isomorphic
-datatypes. For instance, you can update your old \(\HExprD\) to a
+Allowing variables from an empty set is much like forbidding variables,
+so \(\closeTerm\) and \(\HExprD\) describe isomorphic
+datatypes. Correspndingly, you can update your old \(\HExprD\) to a
 shiny \(\closeTerm\) with \(\update\):

### 2010-07-25 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: cut on the conclusive US instructor speech. Add a \newspace so that Discussion is on its own page.

@@
 
-Hence completing our implementation of the open terms interpreter.
-We have defined a well-typed language of arithmetical expressions,
-taking advantage of indexing. Then, we have implemented an evaluator
-for closed term, based on the generic catamorphism function. Using the
-free monad construction, we have automatically derived the language of
-open terms. Using its monadic structure, we have implemented the
-interpreter for open terms in context. Without much effort, we
-have described the syntax of a well-typed language, together with its
-semantics.
+Hence completing our implementation of the open terms
+interpreter. Without much effort, we have described the syntax of a
+well-typed language, together with its semantics.
 
@@
 some practical use of the free monad construction.
+
+\newpage

### 2010-07-26 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: Ispell -brit

@@
 
-Our Agda model does not formalize the $\SYMBswitchD$
+Our Agda model does not formalise the $\SYMBswitchD$
 construction. Instead, we exhibit the isomorphism between declared
@@
 only in a set polymorphic sense: the principal type of the encoded
-description generalizes to the type of $\Desc$ itself. We encode this
+description generalises to the type of $\Desc$ itself. We encode this
 much in our set polymorphic model in Agda and in our Coq model,

### 2010-07-26 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: Ispell -brit

@@
 take for granted. We adopt a bidirectional type propagation mechanism
-to conceal artefacts of the encoding, giving a familiar and practicable
+to conceal artifacts of the encoding, giving a familiar and practicable
 constructor-based presentation to data.
@@
 of polykinded types in Lego;
-\citet{norell:msc-thesis} gave a formalization of
+\citet{norell:msc-thesis} gave a formalisation of
 polytypic programming in Alfa, a precursor to Agda;

### 2010-07-26 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Ispell -brit

@@
 
-Note that $\IDesc{\V{I}}$ is a plain inductive type, parametrized
+Note that $\IDesc{\V{I}}$ is a plain inductive type, parametrised
 by \(\V{I}\), but indexed trivially.
@@
 so \(\closeTerm\) and \(\HExprD\) describe isomorphic
-datatypes. Correspndingly, you can update your old \(\HExprD\) to a
+datatypes. Correspondingly, you can update your old \(\HExprD\) to a
 shiny \(\closeTerm\) with \(\update\):
@@
 The $\discharge$ function is relative to the required type and a
-context of the right type. Its action is to map values to themself,
+context of the right type. Its action is to map values to themselves,
 and variables to their value in context. This corresponds to the

### 2010-07-26 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix the Env widow line
Desc@ICFP: re-shaping inconsistency
Desc@ICFP: Jack the Ripper at work. Another victim.

@@
 %
-An environment realises the context, associating a value for each type:
+An environment realises the context, packing a value for each type:
 %
@@
 so \(\closeTerm\) and \(\HExprD\) describe isomorphic
-datatypes. Correspondingly, you can update your old \(\HExprD\) to a
-shiny \(\closeTerm\) with \(\update\):
+datatypes. Correspondingly, you can update an \(\HExprD\) to a
+\(\closeTerm\):
 %
@@
 `database containing salaries', ready for traversal operations
-preserving the former and targeting the latter. We shall now turn to
-some practical use of the free monad construction.
+preserving the former and targeting the latter.
 

### 2010-07-26 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: clean line break
gaol: minor last-minute tweakage

@@
 In this section, we describe an implementation of inductive types, as
-we know them in ML-like languages. By working with familiar datatypes,
+we know them from ML-like languages. By working with familiar datatypes,
 we hope to focus on the delivery mechanism, warming up gently to the
@@
 
-On the other hand, our choice of notation $\SIGMA{\x}{\M{A}}
-(\M{B}\:\x)$ emphasises that $\Sigma$-types generalise products,
+On the other hand, our choice, $\SIGMA{\x}{\M{A}}
+(\M{B}\:\x)$, emphasises that $\Sigma$-types generalise products,
 with the type of the second component depending on the value of the
-first, where simply typed languages do not express such relative validity.
+first. Simply typed languages do not express such relative validity.
 
@@
 datatype, we need to account for the recursive structure which
-ties these nodes together. As ever, we do this by
+ties these nodes together. We do this by
 constructing a \emph{universe}~\cite{martin-lof:itt}. Universes
@@
 \emph{expect} these objects in a stratified system. This is but an
-informal notation, to give a flavor of the stratified
+informal notation, to give a flavour of the stratified
 presentation. Note that the functors so described are \emph{strictly
@@
 matching to which we are entitled in a high-level language built on top
-of this theory~\cite{goguen:pattern-matching}, but it does at least make
-`assembly language' programming mercifully brief, if hieroglyphic.
+of this theory~\cite{goguen:pattern-matching}. It does at least make
+`assembly language' programming mercifully brief, albeit hieroglyphic.
 \[\stk{

### 2010-07-26 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: chop off the widow!
gaol: minor last-minute tweakage

@@
 
-We shall now fulfil our implementation promises, encoding the universe
-of descriptions. $\Desc$ is already some sort of datatype, but we are
-in the same situation as with $\EnumU$: we ought to be able to
-describe it, coding of $\Desc^n$ in $\Desc^{n+1}$, spiralling
+The set of codes, $\Desc$, is already some sort of datatype; as with $\EnumU$,
+we ought to be able to describe it, coding of $\Desc^n$ in $\Desc^{n+1}$, spiralling
 upwards. Hence, this code would be a first-class citizen, born with
@@
 componentwise decidable equality than functions may in
-general possess.\footnote{E.g., extensionally, there is one inhabitant of
+general possess.\footnote{E.g., extensionally, there is one function in
 \(\EnumT{\void}\To\Nat\); intensionally, there is a countable infinitude
-which it is not safe to collapse.}
+which it is dangerous to identify definitionally.}
 
@@
 but the offensive recursion is concealed by a prestidigitation.
+
 Expanding \(\toDesc{-}\) and propagating types as in
@@
 of \(\Desc^n\) and its internal copy, we have made it appear in its own
-type. Hardwired as a \emph{fait accompli}, this creates no regress,
+type. Hardwired as a trusted \emph{fait accompli}, this creates no regress,
 although one must assume the definition to recheck it.
@@
 \begin{array}{@{}l@{\:\mapsto\:\:}l}
-\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
+\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \sqr{\DVar{\V{x}}}   & \V{\sigma}\: \V{x}                   \\
 \apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
@@
 this process exactly in Agda, without any appeal to dependent
-pattern-matching, induction-recursion, or set polymorphism. All it
+pattern matching, induction-recursion, or set polymorphism. All it
 takes to build such a sawn-off model of encodings is inductive

### 2010-07-26 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: we can even afford some fanciness on that line
Desc@ICFP: remove odd \newpage break
gaol: minor last-minute tweakage

@@
 % 
-Where  $\F{applyI}$ is defined as follow:
+where  $\F{applyI}$ is defined as follows:
 %
@@
 
-In the previous section, we have presented a language of closed
+In the previous section, we presented a language of closed
 arithmetic expressions. Using the free monad construction, we are
@@
 %
-In this setting, we define $\SYMBVar$ by:
+In this setting, we define type variables, $\SYMBVar$ by:
 %
@@
 so \(\closeTerm\) and \(\HExprD\) describe isomorphic
-datatypes. Correspondingly, you can update an \(\HExprD\) to a
+datatypes. Correspondingly, you can update an old \(\HExprD\) to a shiny
 \(\closeTerm\):
@@
 $\SYMBDVar$ is nothing but the \return. The remaining constructors form
-the carrier functor, trivially indexed by $\Unit$. The carrier functor
+the carrier functor, trivially indexed by $\Unit$. The signature functor
 is described as follow:
@@
 %
-Then, we get $\IDesc$ by building its free monad:
+We get $\IDesc{\V{I}}$ by extending the signature with variables from \(\V{I}\):
 %
@@
 
-\newpage
+%\newpage

### 2010-07-26 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: helplessly stare at whitespaces and parenthesis
gaol: minor last-minute tweakage

@@
 this \emph{typical ambiguity} apply~\cite{harper:implicit-universe,
-  luo:utt, courant:explicit-universe}. A formal presentation of a
-stratified system is beyond the scope of this paper and is subject of
+  luo:utt, courant:explicit-universe}. A formal treatment of
+stratification for our system is a matter of
 ongoing work.
@@
 It is time for our first example of a \emph{universe}. You might
-want to offer a choice of named constructors in your datatypes, we shall
+want to offer a choice of named constructors in your datatypes: we shall
 equip you with sets of tags to choose from. Our plan
@@
 \spi{}{}: \PITEL{\V{E}}{\EnumU}\PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To \Set \\
-\begin{array}{@{}l@{\:}l@{\:\mapsto\:\:}l}
+\begin{array}{@{}l@{\:}l@{\;\;\mapsto\;\;}l}
 \spi{\NilE}{& \V{P}}        & \Unit \\
-\spi{(\ConsE{\V{t}}{\V{E}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{E}}{(\LAM{\V{x}} \V{P}\: (\Su{\V{x}}))}}
+\spi{(\ConsE{\V{t}}{\V{E}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{E}}{\LAM{\V{x}} \V{P}\: (\Su{\V{x}})}}
 \end{array}
@@
 This builds a right-nested
-tuple type, demanding an object of $\V{P}\:\V{i}$ for each $\V{i}$ in the given finite
-domain. We can see these tuples as `jump tables' tabulating dependently
-typed functions from the domain. We give this functional
+tuple type, packing a $\V{P}\:\V{i}$ value for each $\V{i}$ in the given
+domain. The step case exposes our notational convention that
+binders scope rightwards as far as possible. These tuples are `jump tables', tabulating dependently
+typed functions.
+We give this functional
 interpretation---the eliminator we need---by the

### 2010-07-26 papers/icfp-2010-desc/icfp23l-dagand.tex
Desc@ICFP: submitted papers

\documentclass[authoryear]{sigplanconf}

\usepackage[T1]{fontenc}
\usepackage{pslatex}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{verbatim}
\usepackage{pig}
\usepackage{url}
\usepackage{xspace}
\usepackage{listings}
\usepackage{flushend}

\include{macros}

%% Haskell-mode for listings
\lstnewenvironment{code}{\lstset{language=Haskell,
                                 basicstyle=\small\ttfamily,
                                 xleftmargin=1cm}}{}

%% Structure
%\newenvironment{structure}{\footnotesize\verbatim}{\endverbatim}
\newenvironment{structure}{\comment}{\endcomment}

%% Written bits of Structure
\newenvironment{wstructure}{\comment}{\endcomment}

%% Comments
%% \setlength{\marginparwidth}{0.7in}
%% \newcommand{\note}[1]{\-\marginpar[\raggedright\footnotesize #1]%
%%                                   {\raggedright\footnotesize #1}}
\newcommand{\note}[1]{}

%% Syntax
\newcommand{\bind}{\emph{bind}\xspace}
\newcommand{\return}{\emph{return}\xspace}

\begin{document}

%\ColourEpigram
\MonochromeEpigram

\conferenceinfo{ICFP'10,} {September 27--29, 2010, Baltimore, Maryland, USA.}
\CopyrightYear{2010}
\copyrightdata{978-1-60558-794-3/10/09}

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{The Gentle Art of Levitation}

%% Alphabetical ordering.
\authorinfo{James Chapman}
           {Institute of Cybernetics, Tallinn University of Technology}
           {james@cs.ioc.ee}
\authorinfo{Pierre-\'{E}variste Dagand \\ Conor McBride}
           {University of Strathclyde}
           {\{dagand,conor\}@cis.strath.ac.uk}
\authorinfo{Peter Morris}
           {University of Nottingham}
           {pwm@cs.nott.ac.uk}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Abstract
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
  We present a closed dependent type theory whose inductive types
  are given not by a scheme for generative declarations,
  but by encoding in a \emph{universe}. Each inductive datatype arises
  by interpreting its \emph{description}---a
  first-class value in a datatype of descriptions. Moreover, the
  latter itself has a description. Datatype-generic programming thus
  becomes ordinary programming. We show some of the resulting generic
  operations and deploy them in particular, useful ways on the
  datatype of datatype descriptions itself. Simulations in existing
  systems suggest that this
  apparently self-supporting setup is achievable without paradox or
  infinite regress.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}
\category{D.1.1}{Programming Techniques}{Applicative (Functional) Programming}
\category{D.3.3}{Language Constructs and Features}{Data types and structures}

\terms
Design, Languages, Theory

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

Dependent datatypes, such as the ubiquitous vectors (lists indexed by
length) express \emph{relative} notions of data validity. They allow
us to function in a complex world with a higher standard of basic
hygiene than is practical with the context-free datatypes of ML-like
languages. Dependent type systems, as found in
Agda~\cite{norell:agda}, Coq~\cite{coq},
Epigram~\cite{mcbride.mckinna:view-from-the-left}, and contemporary
Haskell~\cite{cheney:gadt, xi:gadt}, are beginning to make themselves useful. As
with rope, the engineering benefits of type indexing sometimes
outweigh the difficulties you can arrange with enough of it.

%Dependent types are an appealing technique for building safer and more
%reliable software. By giving types more expressive power, the
%developer is able to encode more precise invariants in the types. As a
%result, more bugs are caught automatically, during
%type-checking. Because of this benefit, dependently-typed systems have
%flourished, such as Generalized Abstract Data-Types (GADT) in
%Haskell~\cite{spj:gadt}, Agda~\cite{norell:agda},
%Ynot~\cite{morrisett:ynot}, or Epigram~\cite{pigs:epigram}, to name
%but a few.

\begin{wstructure}
<- Describe the problem
    <- Data-types in dependent-type theory
        <- Much more precise
            <- More powerful type-system
            -> Stronger safety guarantees
\end{wstructure}

%In this paper, we will focus on data-types in such systems. Indeed,
%the expressive power of the type-system has a direct impact on
%data-types. Because types can \emph{depend} on terms, our data-types
%can be made more precise. The typical example is vectors, which type
%depend on the size of the vector. Having more precision about
%data-types, we can write safer code: taking the $\CN{head}$ of a
%vector is \emph{ensured} to succeed whenever its type states that it
%is a non-empty vector. This property is automatically enforced by the
%type-checker.

\begin{wstructure}
        <- Equipped with elimination principle
            <- Defining functions over them
            <- Making proofs over them
\end{wstructure}

%Moreover, in total programming systems, such as Agda, Ynot, or
%Epigram, data-types come equipped with an elimination principle: while
%a data-type definition introduces new type formers in the theory, we
%need an eliminator to dispose of them. Looking through the
%Curry-Howard lenses, the elimination principle corresponds to an
%induction principle associated with the data-type. To program over our
%data-types, we rely on their induction principle, guaranteeing the
%well-foundedness of our definition.

\begin{wstructure}
    <- Agda standard library [Nisse file]
        <- x implementations of natural numbers
        <- y implementations of lists
        -> Painful duplication of code and functionality
            <- Types are (slightly) different
                -> Same functions need to be re-implemented 
        -> Crucial need for ``genericity''
\end{wstructure}

The blessing of expressing just the right type for the job can also be
a curse. Where once we might have had a small collection of basic
datatypes and a large library, we now must cope with a cornucopia of
finely confected structures, subtly designed, subtly different. The
basic vector equipment is much like that for lists, but we implement
it separately, often retyping the same code. The Agda standard
library~\cite{nisse:asl}, for example, sports a writhing mass of
list-like structures, including vectors, bounded-length lists,
difference lists, reflexive-transitive closures---the list is
petrifying. Here, we seek equipment to tame this gorgon's head with
\emph{reflection}.

\begin{wstructure}
        /> Coq, Agda: external notion
            <- Not first-class citizen
            -> Cannot "compute" with them
            -> No reflection for data-types (?)
\end{wstructure}

The business of belonging to a datatype is itself a notion
relative to the type's \emph{declaration}. Most typed functional
languages, including those with dependent types, feature a datatype
declaration construct, external to and extending the language for
defining values and programs. However, dependent type systems also
allow us to reflect types as the image of a function from a set of
`codes'---a \emph{universe construction}~\cite{martin-lof:itt}. 
Computing with codes, we expose operations on and
relationships between the types they reflect. Here, we adopt
the universe as our guiding design principle. We abolish the
datatype declaration construct, by reflecting it as a datatype of
datatype descriptions which, moreover, \emph{describes itself}. This
apparently self-supporting construction is a trick, of course, but
we shall show the art of it. We contribute

%As in the simply-typed world, the definition of
%data-types is processed by a meta-theoretical engine, before being
%reifed by extending the type theory with the corresponding type
%formers and elimination principle. Because of this external apparatus,
%data-type definition is not \emph{first-class}: we cannot compute with
%them, such as making new data-types from previous data-types. 
%
%This is a rather harsh limitation, in particular in a
%dependently-typed system. Indeed, reflection~\cite{allen:reflection,
%  gregoire:ring-solver} is at the heart of many dependently-typed
%programming techniques. Not having first-class data-type definitions,
%we have to give up reflection for data-types.

\begin{wstructure}
    <- Dependent types offer new programming techniques
        <- Eg.: universe construction
        /> State of the art haunted by the simply-typed paradigm
            -> Generative
            -> Non reflective
\end{wstructure}

%However, we do not think that we are condemned to such fate. The
%external presentation of data-types is an heritage of the simply-typed
%paradigm. Dependently-typed systems have more to offer. Indeed, new
%programming techniques, unavailable in a simply-typed setting,
%arises. One of them is \emph{universe
%  construction}~\cite{martin-lof:itt}. We shall see how this technique
%help us overcoming the limitations of the standard, non reflective and
%generative presentation of data-types.

\begin{wstructure}
<- State contributions
    <- Closed presentation of data-types 
        -> No generativity requires
        -> Subsuming standard inductive families 
            /> Some popular extensions excluded for now
    <- Descriptions of data-types are first-class 
        <- Self-encoded [Section sec:desc-levitate]
    <- ``generic programming is just programming''
        <- Ability to inspect data-type definition
            -> Write program over them
        <- A generic program works over a class of data-types (???)
            -> Capture this class by common structure
            -> Write a program over this common code
    <- Design a language for generic programming
        -> First serious attempt
            /> except possibly Lisp
                <- ???
\end{wstructure}

%In this paper, we propose a new approach to building data-types in a
%dependent-type theory. Our contributions are the following:

\begin{itemize}
\item a \emph{closed} type theory, extensible only \emph{definitionally},
  nonetheless equipped with a universe of inductive families of datatypes;
\item a \emph{self-encoding} of the universe codes as a datatype in the
  universe---datatype generic programming is just programming;
\item a bidirectional \emph{type propagation} mechanism to conceal
  artefacts of the encoding, restoring
  a convenient presentation of data;
\item examples of generic operations and constructions over our universe,
  notably the \emph{free monad} construction;
\item datatype generic programming delivered \emph{directly},
  not via some isomorphic
  model or `view' of declared types.
\end{itemize}

We study two universes as a means to explore this novel way to equip a
programming language with its datatypes. We warm up with a universe of
\emph{simple} datatypes, just sufficient to describe itself. Once we
have learned this art, we scale up to \emph{indexed} datatypes,
encompassing the inductive families~\cite{dybjer:families,luo:utt}
found in Coq and Epigram, and delivering experiments in generic
programming with applications to the datatype of codes itself.

We aim to deliver proof of concept, showing that a closed theory with
a self-encoding universe of datatypes can be made practicable, but we
are sure there are bigger and better universes waiting for a similar
treatment. Benke, Dybjer and
Jansson~\cite{benke:universe-generic-prog} provide a useful survey of
the possibilities, including extension to inductive-recursive
definition, whose closed-form presentation~\cite{dybjer:axiom-ir,
  dybjer:ir-initial-algebra} is both an inspiration for the present
enterprise, and a direction for future study.

The work of Morris, Altenkirch and
Ghani~\cite{morris:PhD,morris:spf,alti:lics09} on
(indexed) containers has informed our style of encoding and the
equipment we choose to develop, but the details here reflect pragmatic
concerns about intensional properties which demand care in
practice. We have thus been able to implement our work as the basis
for datatypes in the Epigram 2 prototype~\cite{pigs:epigram}. We
have also developed a \emph{stratified} model of our coding scheme
in Agda and Coq\footnote{This model is available at \\
\url{http://personal.cis.strath.ac.uk/~dagand/levitate.tar.gz}}.

%\item We present a basic type-theory and extend it with a universe of
%  finite sets (Section~\ref{sec:type-theory}). We show how coding can
%  be made practical by putting types at work ;
%\item We give a closed presentation of inductive data-types, through a
%  universe of descriptions (Section~\ref{sec:universe-desc}). This
%  first universe has the expressive power of simple inductive
%  types. Being closed, this presentation does not require
%  generativity, hence the type theory remains unchanged when
%  data-types are introduced ;
%\item We present a self-encoding of the universe of description inside
%  itself (Section~\ref{sec:desc-levitate}). As a consequence,
%  description of data-types appears as first-class object in the type
%  theory. We illustrate the benefit of a first-order presentation by
%  implementing a generic catamorphism as well as a generic free monad
%  construction, together with its monadic operations ;
%\item We index the universe of descriptions, to subsume standard
%  inductive families (Section~\ref{sec:indexing-desc}). In this
%  setting, we develop several examples of dependently-typed
%  data-structure and some generic operations over them ;
%\item We have implemented this technology in the Epigram programming
%  language. This is, we believe, the first attempt to design a
%  language for generic programming, Lisp having opened the way. We
%  propose and demonstrate with several examples that generic
%  programming is just programming. Because data-types are described by
%  code, we can finally program with them. As a consequence, generic
%  programs are implemented as functions built from the data-type
%  definition.
%\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The Type Theory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The Type Theory}
\label{sec:type-theory}

One challenge in writing this paper is to extricate our account of datatypes from what else is new in Epigram 2. In fact, we demand relatively little from the setup, so we shall start with a `vanilla' theory and add just what we need. The reader accustomed to dependent types will recognise the basis of her favourite system; for those less familiar, we try to keep the presentation self-contained.

\subsection{Base theory}

\begin{wstructure}
<- Presentation of the formalism
    <- Standard presentation
        -> No novelty here
    <- 3 judgments [equation]
        -> Context validity
        -> Typing judgements
        -> Equality judgements
\end{wstructure}

We adopt a traditional presentation for our type
theory, with three mutually defined systems of judgments:
\emph{context validity}, \emph{typing}, and \emph{equality},
with the following forms:
%
\[
\begin{array}{ll}
\G \vdash \Valid                & \mbox{\(\G\) is a valid context, giving types to variables} \\
\G \vdash \Bhab{\M{t}}{\M{T}}           & \mbox{term \(\M{t}\) has type \(\M{T}\) in context \(\G\)} \\
\G \vdash \Bhab{\M{s} \equiv \M{t}}{\M{T}}  & \mbox{\(\M{s}\) and \(\M{t}\) are equal at type \(\M{T}\) in context \(\G\)} \\
\end{array}
\]

\begin{wstructure}
    <- Invariants [equation]
        -> By induction on derivations
\end{wstructure}

The rules are formulated to ensure that the
following `sanity checks' hold by induction on derivations
%
\[
\begin{array}{l@{\;\Rightarrow\;\;}l}
\G            \vdash \Bhab{\M{t}}{\M{T}}            
    & \G \vdash \Valid\; \wedge\; \G\vdash\Type{\M{T}} \\
\G            \vdash s \equiv \Bhab{\M{t}}{\M{T}}   
    & \G \vdash \Bhab{s}{T} \;\wedge\; \G\vdash \Bhab{\M{t}}{\M{T}}
\end{array} \]
%
and that judgments \(\M{J}\) are preserved by well-typed instantiation.
%
\[
\G ; \xS ; \Delta \vdash \M{J}            \;\Rightarrow\;      
    \G \vdash \Bhab{\M{s}}{\M{S}} \;\Rightarrow\; 
          \G ; \Delta[\M{s}/\x] \vdash \M{J}[\M{s}/\x] 
\]

%We are not going to prove the validity of these invariants. They
%follow rather straightforwardly from the induction rules. For formal
%proofs, we refer the reader to standard presentations of type theory,
%such as Luo's seminal work \cite{luo:utt}.

\begin{wstructure}
    <- Judgemental equality
        <- Presentation independent of particular implementation choice
        -> Model in Agda, intensional
        -> Used in Epigram, OTT
\end{wstructure}

We specify equality as a judgment, leaving open the details of its implementation, requiring only a congruence including ordinary computation (\(\beta\)-rules), decided, e.g., by testing \(\alpha\)-equivalence of \(\beta\)-normal forms~\cite{DBLP:journals/jfp/Adams06}. Coquand and Abel feature prominently in a literature of richer equalities, involving \(\eta\)-expansion, proof-irrelevance and other attractions~\cite{DBLP:journals/scp/Coquand96,DBLP:conf/tlca/AbelCP09}. Agda and Epigram 2 support such features, Coq currently does not, but they are surplus to requirements here.

% Therefore, we are not tied to a particular implementation
%choice. In particular, our system has been modelled in Agda, which
%features an intensional equality. On the other hand, it is used in
%Epigram, whose equality has a slightly extensional
%flavor~\cite{altenkirch:ott}. We expect users of fully extensional
%systems to also find their way through this presentation.

\begin{wstructure}
<- Context validity [figure no longer]
    <- Not much to be said
\end{wstructure}

Context validity ensures that variables inhabit well-formed
sets.
%
\[
%% Empty context validity
\Axiom{\vdash \Valid}
\qquad
%% Extend context
\Rule{\G       \vdash \Type{\M{S}}}
     {\G ; \xS \vdash \Valid}\;\x\not\in\G
\]
%
\begin{wstructure}
<- Typing judgements [figure]
    <- Set in Set
        -> For simplicity of presentation
        -> Assume that a valid stratification can be inferred
            <- Harper-Pollack, Luo, Courant
        -> See later discussion
    <- Standard presentation of Pi and Sigma types
\end{wstructure}
%
The basic typing rules
% (Fig.~\ref{fig:typing-judgements})
for tuples and functions are also standard, save that we locally adopt
\(\Set:\Set\) for presentational purposes. Usual techniques to resolve
this \emph{typical ambiguity} apply~\cite{harper:implicit-universe,
  luo:utt, courant:explicit-universe}. A formal treatment of
stratification for our system is a matter of
ongoing work.
%%  putting presentation before paradox~\cite{girard:set-in-set}.
%% The usual remedies apply, \emph{stratifying}
%% \(\Set\)~\cite{harper:implicit-universe, luo:utt,
%%   courant:explicit-universe}.
%%
\input{figure_typing_judgements}

\paragraph{Notation.} We subscript information needed for type synthesis
but not type checking, e.g., the domain of a \(\LAMBINDER\)-abstraction,
and suppress it informally where clear. Square brackets denote tuples,
with a LISP-like right-nesting convention: \(\sqr{a\;b}\) abbreviates
\(\pair{a}{\pair{b}{\void}{}}{}\).

%We recognise the standard presentation of $\Pi$ and $\Sigma$
%types, respectively inhabited by lambda terms and dependent
%pairs. Naturally, there are rules for function application and
%projections of $\Sigma$-types. Equal types can be substituted, thanks
%to the conversion rule.

%For the sake of presentation, we postulate a $\Set$ in $\Set$
%rule. Having this rule makes our type theory inconsistent, by Girard's
%paradox~\cite{girard:set-in-set}. However, it has been
%shown~\cite{harper:implicit-universe, luo:utt,
%  courant:explicit-universe} that a valid stratification can be
%inferred, automatically or semi-automatically. In the remaining of our
%presentation, we will assume that such a stratification exists, even
%though we will keep it implicit. We shall discuss this assumption in
%Section~\ref{sec:discussion}.

%\begin{figure}
%
%\input{figure_typing_judgements}
%
%\caption{Typing judgements}
%\label{fig:typing-judgements}
%
%\end{figure}

\begin{wstructure}
<- Judgemental equality [figure]
    <- symmetry, reflexivity, and transitivity
    <- beta-rules for lambda and pair
    <- xi-rule for functions
    -> Agnostic in the notion of equality
        <- Doesn't rely on a ``propositional'' equality
        -> Key: wide applicability of our proposal
\end{wstructure}

The judgmental equality comprises the computational rules below,
closed under reflexivity,
symmetry, transitivity and structural congruence, even under binders.
We omit the mundane rules which ensure these closure properties for
reasons of space.
\input{figure_judgemental_equality}
Given a suitable stratification of \(\Set\), the
computation rules yield a terminating evaluation procedure, ensuring
the decidability of equality and thence type checking.

%Finally, we define the rules governing judgemental equality in
%Figure~\ref{fig:judgemental-equality}. We implicitly assume that
%judgemental equality respects symmetry, reflexivity, and
%transitivity. We capture the computational behavior of the language
%through the $\beta$-rules for function application and pairs. Finally,
%we implicitly assume that it respects purely syntactic and structural
%equality. This includes equality under lambda ($\xi$-rule).

%Crucially, being judgemental, this presentation is agnostic in the
%notion of equality actually implemented. Indeed, our typing and
%equality judgements do not rely on a ``propositional'' equality. This
%freedom is a key point in favour of the wide applicability of our
%proposal. This judgemental presentation must be read as a
%\emph{specification}: our proposal works with any propositional
%equality satisfying this specification. Moreover, our lightweight
%requirements do not endanger decidability of equality-checking.
%Obviously, when implementing our technology in an existing
%type-theory, some opportunities arise. We will present some of them
%along the course of the paper.

%\begin{figure}

%\input{figure_judgemental_equality}
%
%\caption{Judgemental equality}
%\label{fig:judgemental-equality}
%
%\end{figure}

\begin{wstructure}
!!! Need Help !!!
<- Meta-theoretical properties
    <- Assuming a stratified discipline
    <> The point here is to reassert that dependent types are not evil, 
       there is no non-terminating type-checker, or such horrible lies <>
    -> Strongly normalising
        -> Every program terminates
    -> Type-checking terminates
    ???
\end{wstructure}

%This completes our presentation of the type theory. Assuming a
%stratified discipline of universe, the system we have described enjoy
%some very strong meta-theoretical properties. Unlike simply typed
%languages, such as Haskell, dependently-typed systems are
%\emph{strongly normalising}: every program that type-checks
%terminates. Moreover, type-checking is decidable and can therefore be
%implemented by a terminating algorithm.
%\note{Need some care here. Expansion would be good too. I wanted to
%  carry the intuition that we are not the bad guys with a
%  non-terminating type-checker.}

\subsection{Finite enumerations of tags}
\label{sec:finite-sets}

\begin{wstructure}
<- Motivation
    <- Finite sets could be encoded with Unit and Bool
        /> Hinder the ability to name things
    <- W-types considered harmful?
        ???
    -> For convenience
        <- Named elements
        <- Referring by name instead of code
        -> Types as coding presentation
            /> Also as coding representation!
\end{wstructure}

It is time for our first example of a \emph{universe}. You might
want to offer a choice of named constructors in your datatypes: we shall
equip you with sets of tags to choose from. Our plan
is to implement (by extending the theory, or by encoding) the signature
%
\[
  \Type{\EnumU}\qquad \Type{\EnumT{(\Bhab{\M{E}}{\EnumU})}}
\]
%
where some value \(E:\EnumU\) in the `enumeration universe' describes
a type of tag choices \(\EnumT{E}\). We shall need
some tags---valid identifiers, marked to indicate that
they are data, not variables scoped and substitutable---so we hardwire these
rules:
%
\[
%% UId
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Type{\UId}}
\qquad
%% Tag
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\Tag{\V{s}}}{\UId}}\;\V{s}\: \mbox{a valid identifier}
\]
%
Let us describe enumerations as lists of tags, with signature:
%
\[
\Bhab{\NilE}{\EnumU}\qquad
\Bhab{\ConsE{(\Bhab{\M{t}}{\UId})}{(\Bhab{\M{E}}{\EnumU})}}{\EnumU}
\]
%
What are the \emph{values} in \(\EnumT{E}\)? Formally, we represent
the choice of a tag as a numerical index into \(E\), via new rules:
%
\[
%% Ze
\Rule{\Gamma \vdash \Valid}
     {\Gamma \vdash \Bhab{\Ze}{\EnumT{(\ConsE{\M{t}}{\M{E}})}}} 
\qquad
%% Su
\Rule{\Gamma \vdash \Bhab{\M{n}}{\EnumT{\M{E}}}}
     {\Gamma \vdash \Bhab{\Su{\M{n}}}{\EnumT{(\ConsE{\M{t}}{\M{E}})}}}
\]
%
However, we expect that in practice, you might rather refer to these
values \emph{by tag}, and we shall ensure that this is possible in due course.

%As a motivating example, we are now going to extend the type theory
%with a notion of finite set. One could argue that there is no need for
%such an extension: finite sets, just as any data-structure, can be
%encoded inside the type theory. A well-known example of such encoding
%is the Church encoding of natural numbers, which is isomorphic to
%finite sets. \note{Shall we talk about W-types encoding?}

%However, using encodings is impractical. In the case of finite sets,
%for instance, we would like to name the elements of the sets. Then, we
%need to be able to manipulate these elements by their name, instead of
%their encoding. While we are able to give names to encodings, it is
%extremely tedious to map the encodings back to a name. Whereas these
%objects have a structure, the structure is lost during the encoding,
%when they become anonymous inhabitants of a $\Pi$ or $\Sigma$-type.

%In the simply-typed world, we are used to see types as a coding
%presentation -- presentation of invariants, presentation of
%properties. In the dependently-typed world, we also learn to use types
%as a coding representation: finite sets being good citizens, they
%ought to be democratically represented at the type level. As we will
%see, this gives us the ability to name and manipulate them (this is
%were the democracy analogy goes crazy, I think).
%\note{Did I got the coding presentation vs. coding representation
%  story right? No.}

\begin{wstructure}
<- Implementation [figure]      
    <- Tags
        -> Purely informational token
    <- EnumU
        -> Universe of finite sets
    <- EnumT e
        -> Elements of finite set e
\end{wstructure}

%The specification of finite sets is presented in
%Figure~\ref{fig:typing-finite-set}. It is composed of three
%components. First, we define tags as inhabitants of the $\UId$ type. A
%tag is solely an informative token, used for diagnostic
%purposes. Finite sets inhabits the $\EnumU$ type. Unfolding the
%definition, we get that a finite set is a list of tags. Finally,
%elements of a finite set $\V{u}$ belong to the corresponding $\EnumT{\V{u}}$
%type. Intuitively, it corresponds to an index -- a number -- pointing
%to an element of $\V{u}$.

%\begin{figure}

%\input{figure_finite_sets}

%\caption{Typing rules for finite sets}
%\label{fig:typing-finite-set}

%\end{figure}

\begin{wstructure}
<- Equipment
    <- \spi operator
        <- Equivalent of Pi on finite sets
        <- First argument: (finite) domain
        <- Second argument: for each element of the domain, a co-domain
        -> Inhabitant of \spi: right-nested tuple of solutions
            <- Skip code for space reasons
    <- switch operator
        <- case analyses over x
        <- index into the \spi tuple to retrieve the corresponding result
\end{wstructure}

Enumerations come with further machinery. Each \(\EnumT{E}\) needs an
eliminator, allowing us to branch according to a tag choice. Formally,
whenever we need such new computational facilities, we add primitive
operators to the type theory and extend the judgmental equality with
their computational behavior. However, for compactness and readability, we
shall write these operators as functional programs (much as we
model them in Agda).

We first define the `small product' $\SYMBspi$ operator:
%
\[\stk{
%% spi
\spi{}{}: \PITEL{\V{E}}{\EnumU}\PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To \Set \\
\begin{array}{@{}l@{\:}l@{\;\;\mapsto\;\;}l}
\spi{\NilE}{& \V{P}}        & \Unit \\
\spi{(\ConsE{\V{t}}{\V{E}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{E}}{\LAM{\V{x}} \V{P}\: (\Su{\V{x}})}}
\end{array}
}\]
%
This builds a right-nested
tuple type, packing a $\V{P}\:\V{i}$ value for each $\V{i}$ in the given
domain. The step case exposes our notational convention that
binders scope rightwards as far as possible. These tuples are `jump tables', tabulating dependently
typed functions.
We give this functional
interpretation---the eliminator we need---by the
$\SYMBswitch$ operator, which, unsurprisingly, iterates projection:
%
\[\stk{
%% switch
\begin{array}{@{}ll}
\SYMBswitch : \PITEL{\V{E}}{\EnumU}
               \PITEL{\V{P}}{\EnumT{\V{E}} \To \Set} \To
              \spi{\V{E}}{\V{P}} \To
               \PITEL{\V{x}}{\EnumT{\V{E}}} \To \V{P}\: \x
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\switch{(\ConsE{\V{t}}{\V{E}})}{\V{P}}{\V{b}}{\Ze}      & \fst{\V{b}} \\
\switch{(\ConsE{\V{t}}{\V{E}})}{\V{P}}{\V{b}}{(\Su{\V{x}})} & \switch{\V{E}}{(\LAM{\V{x}} \V{P}
  (\Su{\V{x}}))}{(\snd{\V{b}})}{\V{x}}
\end{array}
}\]

%Again, there is a clear equivalent in the full-$\Set$ world: function
%application. The operational behaviour of $\F{switch}$ is
%straightforward: $\V{x}$ is peeled off as we move deeper inside the nested
%tuple $\V{b}$. When $\V{x}$ equals $\Ze$, we simply return the value we are
%pointing to.

\begin{wstructure}
<- Equivalent to having a function space over finite sets
    /> Made non-obvious by low-level encodings
        <- General issue with codes
             -> Need to provide an attractive presentation to the user
    -> Types seem to obfuscate our reading
        <- Provide ``too much'' information
        /> False impression: information is actually waiting to be used more widely
        -> See next Section
\end{wstructure}

The $\SYMBspi$ and $\SYMBswitch$ operators deliver dependent
elimination for finite enumerations, but are rather awkward to
use directly. We do not write the range for a \(\LAMBINDER\)-abstraction, so
it is galling to supply \(\V{P}\) for functions defined by $\SYMBswitch$.
Let us therefore find a way to recover the tedious details of the
encoding from types.

%, they also
%come with a notion of finite function space. However, we had to
%extract that intuition from the type, by a careful reading. This seems
%to contradict our argument in favour of types for coding
%representation. Here, we are overflown by low-level, very precise type
%information.

%However, our situation is significantly different from the one we
%faced with encoded data: while we were suffering from a crucial lack
%of information, we are now facing too much information, hence losing
%focus. This is a general issue with the usage of codes, as they convey
%much more information than what the developer is willing to see.

%As we will see in the following section, there exists a cure to this
%problem. In a nutshell, instead of being overflown by typing
%information, we will put it at work, automatically. The consequence is
%that, in such system, working with codes is \emph{practical}: one
%should not be worried by information overload, but how to use it as
%much as possible. Therefore, we should not be afraid of using codes for
%practical purposes.

\subsection{Type propagation}
\label{sec:type-propagation}

\begin{wstructure}
<- Bidirectional type-checking [ref. Turner,Pierce]
    -> Separating type-checking from type synthesis
    <- Type checking: push terms into types
        <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
    <- Type synthesis: pull types out of terms
        <- Example: |x : S l- x :<: S| gives us the type of x
\end{wstructure}

Our approach to tidying the coding cruft is deeply rooted in the
bidirectional presentation of type checking from Pierce and
Turner~\cite{pierce:bidirectional-tc}. They divide
type inference into two communicating components. In
\emph{type synthesis}, types are \emph{pulled} out of terms. A
typical example is a variable in the context:
%
\[
\Rule{\G ; \xS ; \Delta \vdash \Valid}
     {\G ; \xS ; \Delta \vdash \Bhab{\V{x}}{\M{S}}}
\]
%
Because the context stores the type of the variable, we can extract the
type whenever the variable is used.

On the other hand, in the \emph{type checking} phase, types are
\emph{pushed} into terms. We are handed a type together with a term,
our task consists of checking that the type admits the term. In doing
so, we can and should use the information
provided by the type. Therefore, we can relax our requirements on the
term. Consider \(\LAMBINDER\)-abstraction:
%
\[
\Rule{\G       \vdash \Type{\M{S}} \quad
      \G ; \xS \vdash \Bhab{\M{t}}{\M{T}}}
     {\G \vdash \Bhab{\PLAM{\x}{\M{S}} \M{t}}{\PIS{\xS} \M{T}}}
\]
%
The official rules require an annotation specifying the domain.
However, in type \emph{checking}, the \(\Pi\)-type we push in determines
the domain, so we can drop the annotation.

\begin{wstructure}
<- Formalisation: type propagation
    <- Motivation
        -> High-level syntax
            -> exprIn: types are pushed in
                <- Subject to type *checking*
            -> exprEx: types are pulled from
                <- Subject to type *synthesis*
        -> Translated into our low-level type theory
        -> Presented as judgements
    -> Presentation mirrors typing rule of [figure] 
        -> Ignore identical judgements
\end{wstructure}

We adapt this idea, yielding a \emph{type
propagation} system, whose purpose is to elaborate compact
\emph{expressions} into the terms of our underlying type theory, much
as in the definition of Epigram
1~\cite{mcbride.mckinna:view-from-the-left}.  We divide expressions
into two syntactic categories: $\exprIn$ into which types are pushed,
and $\exprEx$ from which types are extracted. In the
bidirectional spirit, the $\exprIn$ are subject to type
\emph{checking}, while the $\exprEx$---variables and elimination
forms---admit type \emph{synthesis}. We embed $\exprEx$ into
$\exprIn$, demanding that the synthesised type coincides with the type
proposed. The other direction---only necessary to apply
abstractions or project from pairs---takes a type annotation.

% As the presentation largely
%mirrors the inference rules of the type theory, we will ignore the
%judgments that are identical. We refer our reader to the associated
%technical report~\cite{chapman:desc-tech-report} for the complete
%system of rules.

\begin{wstructure}
<- Type synthesis [figure]
    <- Pull a type out of an exprEx
    <- Result in a full term, together with its type
    -> Do *not* need to specify types
        -> Extracting a term from the context
        -> Function application
        -> Projections
\end{wstructure}

Type synthesis (Fig.~\ref{fig:type-synthesis}) is the \emph{source} of
types. It follows the \(\exprEx\) syntax, delivering both the
elaborated term and its type. Terms and expressions never mix: e.g.,
for application, we instantiate the range with the \emph{term}
delivered by checking the argument \emph{expression}. Hardwired
operators are checked as variables.

\begin{figure}
\input{figure_type_synthesis}
\caption{Type synthesis}
\label{fig:type-synthesis}
\end{figure}

\begin{wstructure}
<- Type checking [figure]
    <- Push a type in an exprIn
    <- Result in a full term
    -> *Use* the type to build the term!
        -> Domain and co-domain propagation for Pi and Sigma
        -> Translation of 'tags into EnumTs
        -> Translation of ['tags ...] into EnumUs
        -> Finite function space into switch
\end{wstructure}

Dually, type checking judgments (Fig.~\ref{fig:type-checking})
are \emph{sinks} for types. From an
$\exprIn$ and a type pushed into it, they elaborate a low-level
term, extracting information from the type. Note that we inductively ensure the following `sanity checks':
%
\[\stkc{
\Gamma\Vdash\propag{e}{\pull{t}{T}} \Rightarrow
  \Gamma\vdash t:T \\
\Gamma\Vdash\push{\propag{e}{t}}{T} \Rightarrow
  \Gamma\vdash t:T
}\]

Canonical set-formers are \emph{checked}: we could exploit
\(\Set:\Set\) to give them synthesis rules, but this would prejudice
our future stratification plans. Note that abstraction and pairing are
free of annotation, as promised. Most of the propagation rules are
unremarkably structural: we have omitted some mundane rules which just
follow the pattern, e.g., for \(\UId\).

\begin{figure}
\input{figure_type_checking}
\caption{Type checking}
\label{fig:type-checking}
\end{figure}

However, we also add abbreviations. We write \(\spl{\M{f}}\),
pronounced `uncurry \(\M{f}\)' for the function which takes a pair and
feeds it to \(\M{f}\) one component at a time, letting us name
them individually. Now, for the finite enumerations, we go to work.

Firstly, we present the codes for enumerations as right-nested tuples
which, by our LISP convention, we write as unpunctuated lists of tags
\(\sqr{\etag{t_0}\ldots\etag{t_n}}\).
Secondly, we can denote an element \emph{by its
name}: the type pushed in allows us to recover the numerical
index. We retain the numerical forms to facilitate
\emph{generic} operations and ensure that shadowing is punished
fittingly, not fatally.
Finally, we express functions from enumerations as tuples.
Any tuple-form, \(\void\) or \(\pair{\_}{\_}{}\), is accepted by the
function space---the generalised product---if it is accepted by the
small product. Propagation fills in the appeal to $\SYMBswitch$,
copying the range information.

Our interactive development tools also perform the
reverse transformation for intelligible output.
The encoding of any specific enumeration is thus hidden by these
translations. Only, and rightly, in enumeration-generic programs is the
encoding exposed.

\begin{wstructure}
<- Summary
    -> Not a novel technique [refs?]
        /> Used as a boilerplate scrapper
    -> Make dealing with codes *practical*
        <- Example: Finite sets/finite function space
        -> We should not restrain our self in using codes
            <- We know how to present them to the user
-> Will extend this machinery in further sections
\end{wstructure}

%In this section, we have developed a type propagation system based on
%bidirectional type-checking. Using bidirectional type-checking as a
%boilerplate scrapper is a well-known
%technique~\cite{pierce:bidirectional-tc,
%  xi:bidirectional-tc-bound-array, chlipala:strict-bidirectional-tc}
%\note{Everybody ok with the citations?}. In our case, we have shown
%how to instrument bidirectionality to rationalise the expressivity of
%dependent types. We have illustrated our approach with finite sets. We
%have abstracted away the low-level presentation of finite sets,
%offering a convenient syntax instead.

%This example teaches us that we should not be afraid of codes, as soon
%as type information is available. We have shown how to rationalise
%this information in a formal presentation. Hence, we have shown that
%programming with such objects is practical. As we introduce more codes
%in our theory, we will show how to extend the framework we have
%developed so far.

Our type propagation mechanism does no constraint
solving, just copying, so it is just the thin end of the elaboration wedge.
It can afford us this `assembly
language' level of civilisation as \(\EnumU\) universe
specifies not only the \emph{representation} of the
low-level values in each set as bounded numbers, but also the
\emph{presentation} of these values as high-level tags. To encode only
the former, we should merely need the \emph{size} of enumerations, but
we extract more work from these types by making them more informative.
We have also, \emph{en passant}, distinguished enumerations which have
the same cardinality but describe distinct notions:
\(\EnumT{\sqr{\etag{\CN{red}}\,\etag{\CN{blue}}}}\) is not
\(\EnumT{\sqr{\etag{\CN{green}}\,\etag{\CN{orange}}}}\).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% A Universe of simple data-types
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{A Universe of Inductive Datatypes}
\label{sec:universe-desc}

\begin{wstructure}
<- Why starting with simple datatypes
    <- For pedagogical purposes
        <- datatypes as we know them every day
        /> Target dependent types
    -> Cut down version of Induction Recursion
        -> Presentation evolves independently as we go to dependent types
\end{wstructure}

In this section, we describe an implementation of inductive types, as
we know them from ML-like languages. By working with familiar datatypes,
we hope to focus on the delivery mechanism, warming up gently to the
indexed datatypes we really want.  % Our encoding is inspired by
Dybjer and Setzer's closed formulation of
induction-recursion~\cite{dybjer:axiom-ir}, but without the
`-recursion'.  An impredicative Church-style encoding of datatypes is
not adequate for dependently typed programming, as although such
encodings present data as non-dependent eliminators, they do not
support dependent \emph{induction}~\cite{geuvers:induction-not-derivable}.
Whilst the \(\LAMBINDER\)-calculus captures all that data can \emph{do},
it cannot ultimately delimit all that data can \emph{be}.

\subsection{The power of $\Sigma$}

\begin{wstructure}
<- The duality of Sigma
    <- Sigma generalises sum over arbitrary arities
        -> \Sigma A B == \Sigma_{x : A} B x
    <- Sigma generalises product to have a dependant second component
        -> \Sigma A B == (x : A) \times (B x)
\end{wstructure}

In dependently typed languages, $\Sigma$-types can be interpreted as
two different generalisations. This duality is reflected in the
notation we can find in the literature. The notation
$\blue{\Sigma}_{\x : \M{A}} (\M{B}\: \x)$ stresses that
$\Sigma$-types are `dependent sums', generalising sums over
arbitrary arities, where simply typed languages have finite sums.

On the other hand, our choice, $\SIGMA{\x}{\M{A}}
(\M{B}\:\x)$, emphasises that $\Sigma$-types generalise products,
with the type of the second component depending on the value of the
first. Simply typed languages do not express such relative validity.

\begin{wstructure}
<- datatypes in the simply-typed world
    -> "sums-of-product"
        <- Sum of constructors
        <- Product of arguments
<- datatypes in the dependently-typed world
    -> "sigmas-of-sigmas"
    /> Need ability to manipulate these sigmas
        -> Define a Code for datatypes
        -> Together with a sigma-based Interpretation
\end{wstructure}

In ML-like languages, datatypes are presented as a
\emph{sum-of-products}. A datatype is defined by a finite sum of
constructors, each carrying a product of arguments. To embrace
these datatypes, we have to capture this grammar.
With dependent types, the notion of sum-of-products translates into
\emph{sigmas-of-sigmas}.

\subsection{The universe of descriptions}
\label{sec:desc-universe}

\begin{wstructure}
<- Introduction to Universe construction
    <- Define a Code
        -> Name objects
    <- Define an Interpretation of codes into the type theory
        -> Give a semantics to objects
    -> Ability to manipulate code
    -> Ability to compute with these objects
\end{wstructure}

While sigmas-of-sigmas can give a \emph{semantics} for the
sum-of-products structure in each node of the tree-like values in a
datatype, we need to account for the recursive structure which
ties these nodes together. We do this by
constructing a \emph{universe}~\cite{martin-lof:itt}. Universes
are ubiquitous in dependently typed
programming~\cite{benke:universe-generic-prog, oury:power-of-pi},
but here we take them as the foundation of our notion
of datatypes.

%The key idea behind universe construction is our ability to make names
%by defining new types. These names are called \emph{codes}. By
%defining a set of codes, we define the syntax of a language. To give
%it a semantics, we \emph{interpret} these codes back into the type
%theory. Hence, codes act as labels, while the type theory provides the
%computational machinery. Codes being mere labels, we can inspect them,
%hence regaining structural information. We can also compute over them:
%deriving new codes from others, or even functions on them.

\begin{wstructure}
<- Justification of the code 
    <- [both figures]: cannot be read separately
    <- Mimic the standard grammar of datatypes description
        -> Just as we already know it
        <- '\Sigma for making sigmas-of-sigmas
        <- 'indx for exhibiting the functoriality
            -> For recursive arguments
        <- '1 for end of description
\end{wstructure}

To add inductive types to our type theory, we build a universe of
datatype \emph{descriptions} by implementing the signature presented
in Figure~\ref{fig:desc_universe}, with codes mimicking the grammar of
datatype declarations. We can read a description $\V{D}:\Desc^{n}$ as
a `pattern functor' on $\Set^{n}$, with $\SYMBdescop{\V{D}}$ its
action on an object, \(\V{X}\), soon to be instantiated
recursively. The superscripts indicate the $\Set$-levels at which we
\emph{expect} these objects in a stratified system. This is but an
informal notation, to give a flavour of the stratified
presentation. Note that the functors so described are \emph{strictly
  positive}, by construction.

Descriptions are sequential structures ending in $\DUnit$,
indicating the empty tuple. To build sigmas-of-sigmas, we provide a
$\SYMBDSigma$ code, interpreted as a $\Sigma$-type. To request a
recursive component, we have $\DIndx{\V{D}}$, where \(\V{D}\)
describes the rest of the node. These codes give us
sigmas-of-sigmas with recursive places. An equivalent, more algebraic presentation
could be given, as illustrated in Section~\ref{sec:indexing-desc}.

\begin{figure}
\[\stk{
\begin{array}{@{}ll}
\Desc^n 
    &: \Set^{n+1}\\
\DUnit 
    &: \Desc^{n}\\
\DSigma{(\Bhab{\V{S}}{\Set^{n}})}{(\Bhab{\V{D}}{\V{S} \To \Desc^{n}})} 
    &: \Desc^{n}\\
\DIndx{(\Bhab{\V{D}}{\Desc^{n}})} 
    &: \Desc^{n} \\
\end{array}
\smallskip\\
\begin{array}{l@{\V{X}\:\mapsto\:}l}     
\multicolumn{2}{l}{\descop{\_\:}{} : \Desc^{n} \To \Set^{n} \To \Set^{n}} \\
 \descop{\DUnit}{} &  \Unit \\
 \descop{\DSigma{\V{S}}{\V{D}}}{} &
     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{X}}}  \\
\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
\end{array}
}\]

\caption{Universe of Descriptions}
\label{fig:desc_universe}

\end{figure}

We admit to being a little coy,
writing of `implementing a signature' without clarifying how. A viable
approach would simply be to extend the theory with constants for the
constructors and an operator for \(\descop{\V{D}}\). In
Section~\ref{sec:desc-levitate}, you will see what we do instead.
Meanwhile, let us gain some intuition by developing examples.

\subsection{Examples}
\label{sec:desc-examples}

\begin{wstructure}
<- Nat
    <- Sum of zero, suc
    <- zero case: done
    <- suc case: leave open and done
    -> NatD Z = 1 + Z
\end{wstructure}

We begin with the natural numbers, now working in the high-level
expression language of Section~\ref{sec:type-propagation}, exploiting
type propagation.
%
\[\stk{
\NatD : \Desc^n \\
\NatD \mapsto \DSigma{\EnumT{\sqr{\NatZero\: \SYMBNatSuc }}}
                     {\sqr{ \DUnit \quad (\DIndx{\DUnit}) }}
}\]
%
Let us explain its construction. First, we use $\SYMBDSigma$ to
give a choice between the $\NatZero$ and $\SYMBNatSuc$ constructors.
What follows depends on this choice, so we write the function
computing the rest of the description in tuple notation.  In the
$\NatZero$ case, we reach the end of the description. In the
$\SYMBNatSuc$ case, we attach one recursive argument and close the
description. Translating the \(\Sigma\) to a binary sum, we have
effectively described the functor:
%
\[    \NatD\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{Z}    \]
Correspondingly, we can see the injections to the sum:
\[
\sqr{\NatZero}:\descop{\NatD}Z \qquad
\sqr{\NatSuc{(\Bhab{z}{Z})}}:\descop{\NatD}Z
\]

\begin{wstructure}
<- List
    <- Sum of nil, cons
    <- nil case: done
    <- cons case: product of X with leave open and done
    -> ListD X Z = 1 + X * Z
\end{wstructure}

The pattern functor
for lists needs but a small change:
%
\[\stk{
\ListD : \Set^n \To \Desc^n \\
\ListD \: \V{X} \mapsto
 \DSigma{\EnumT{\sqr{ \ListNil\: \SYMBListCons }}}
         {\sqr{ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) }}
}\]
%
The $\SYMBNatSuc$ constructor becomes
$\SYMBListCons$, taking an $\V{X}$ followed by a
recursive argument. This code describes the following functor:
%
\[    \ListD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \D{\ensuremath{\times}} \V{Z}     \]

\begin{wstructure}
<- Tree
    <- sum of leaf, node
    <- leaf case: done
    <- node case: product of X with two leave open and done
    -> TreeD X Z = 1 + X * Z * Z
\end{wstructure}

Of course, we are not limited to one recursive argument. Here are
the node-labelled binary trees:
%
\[\stk{
\TreeD : \Set^n \To \Desc^n \\
\begin{array}{@{}l@{}l}
\TreeD \: \V{X} \mapsto 
    \DSigma{ & \EnumT{\sqr{ \TreeLeaf\: \TreeNode }} \\}
           { & \sqr{ \DUnit \quad
                     (\DIndx{(\DSigma{\V{X}}{\LAM{\_}\DIndx{\DUnit}})})} }
\end{array}
}\]
%
Again, we are one evolutionary step away from $\ListD$. However,
instead of a single call to the induction code, we add another. The
interpretation of this code corresponds to the following functor:
%
\[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} 
          \V{Z} \Prod \V{X}  \Prod \V{Z}     \]

\begin{wstructure}
<- Tagged description
    <- Form TDesc = List (UId x Desc) [equation]
    <- Follow usual sums-of-product presentation of datatype
        <- Finite set of constructors
        <- Then whatever you want
    -> Any Desc datatype can be turned into this form
        -> No loss of expressive power
        /> Guarantee a ``constructor form''
\end{wstructure}

From the examples above, we observe that datatypes are defined by a
$\SYMBDSigma$ whose first argument enumerates the constructors. We
call codes fitting this pattern \emph{tagged} descriptions. Again,
this is a clear reminder of the sum-of-products style. Any
description can be forced into this style with a singleton constructor
set. We characterise tagged descriptions thus:
\[\stk{
 \TagDesc^n : \Set^{n+1} \\
 \TagDesc^n \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{\LAM{\_} \Desc^n})
\smallskip\\
\SYMBtoDesc : \TagDesc^n \To \Desc^n \\
\SYMBtoDesc \mapsto
\spl{\LAM{\V{E}}\LAM{\V{D}}
\DSigma{\EnumT{\V{E}}}{(\F{switch}\:\V{E}\:(\LAM{\_}\Desc^n)\:\V{D})}}
}\]
It is not such a stretch to expect that the familiar datatype
declaration might desugar to the \emph{definitions} of a tagged description.

\begin{wstructure}
<- Fictive object [figure 'data Desc']
    -> Must be read as a type signature
    -> See further for its actual implementation
        <- Subject to our levitation exercise
\end{wstructure}

\subsection{The least fixpoint}
\label{sec:desc-fix-point}

\begin{wstructure}
<- Build the fixpoint of functors
    <- See examples: need to build their initial algebra
    -> Extend the type theory with Mu/Con [figure]
        <- Straightforward definition of a fixpoint
            <- Interpret D with (Mu D) as sub-objects
\end{wstructure}

\newcommand{\mude}[1]{\D{\(\mu^{\!+}\)}\:#1}

So far, we have built pattern functors with our \(\Desc\) universe.
Being polynomial functors, they all admit a least fixpoint, which we
now construct by \emph{tying the knot}: the element type abstracted
by the functor is now instantiated recursively:
%
\[
\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc^n}}
     {\Gamma \vdash \Bhab{\Mu{\M{D}}}{\Set^n}} \qquad
\Rule{\Gamma \vdash \Bhab{\M{D}}{\Desc^n} \quad 
      \Gamma \vdash \Bhab{\M{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}}
     {\Gamma \vdash \Bhab{\Con{\M{d}}}{\Mu{\M{D}}}}
\]
Tagged descriptions are very common, so we abbreviate:
\[
  \mude{}:\TagDesc^n \To \Set^n \qquad \mude{\V{T}}\mapsto\Mu{(\toDesc{\V{T}})}
\]

\begin{wstructure}
<- Elimination on Mu
    <- We are used to foldD : \forall X. (desc D X -> X) -> mu D -> X
        /> Not dependent
        -> Cannot express some (which one again?) properties
    -> Develop a dependent induction
        <- Everywhere/All
        <- Induction
    -> *Generic*
    ???
\end{wstructure}

We can now build datatypes and their elements, e.g.:
\[\stk{
\Nat \mapsto \Bhab{\mude{\pair{\sqr{\NatZero\: \SYMBNatSuc }}
                               {\sqr{\DUnit \quad (\DIndx{\DUnit})}}{}}}
                  {\Set^n}
\\
\Con{\sqr{\NatZero}}:\Nat \qquad
\Con{\sqr{\NatSuc{(\Bhab{n}{\Nat})}}}:\Nat
}\]

But how shall we compute with our data?  We should expect an
elimination principle. Following a categorical intuition, we might
provide the `fold', or `iterator', or `catamorphism':
%
\[
\F{cata} : \PITEL{\V{D}}{\Desc^n}
           \PI{\V{T}}{\Set^n}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} 
\]

\newcommand{\SYMBind}{\F{ind}}

However, iteration is inadequate for \emph{dependent} computation.
We need \emph{induction} to write functions whose type depends on inductive
data. Following \citet{benke:universe-generic-prog}, we adopt the following:
%
\[\stk{
\F{ind} : \stk{ \PITEL{\V{D}}{\Desc^n}
                \PI{\V{P}}{\Mu{\V{D}} \To \Set^k}         \\
               (\PI{\V{d}}{\descop{\V{D}}{(\Mu{\V{D}})}}       
                \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{d}} \To \V{P} (\Con{\V{d}})) \To \\
               \PI{\V{x}}{\Mu{\V{D}}} \V{P} \V{x} 
} \\
\F{ind}\, \V{D}\, \V{P}\, \V{m}\, (\Con{\V{d}}) \mapsto 
    \V{m}\, \V{d}\, (\all{\V{D}}{(\Mu{\V{D}})}{\V{P}}
                           {(\F{ind}\, \V{D}\, \V{P}\, \V{m})}
                           {\V{d}})
}\]
%
Here, $\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}$ states that
$\Bhab{\V{P}}{\V{X} \To \Set^k}$ holds for every subobject $\V{x}:\V{X}$
in \(\V{D}\), and \(\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}\) is a
`dependent map', applying some
\(\Bhab{\V{p}}{\PIS{\Bhab{\x}{\V{X}}}\V{P}\,\V{x}}\) to each \(\x\)
contained in \(\V{d}\). The definition (including an extra case,
introduced soon) is in
Figure~\ref{fig:all-predicates}.\footnote{To pass the termination
  checker, we had to inline the definition of $\SYMBall$ into
  $\SYMBind$ in our Agda model. A simulation argument shows that the
  definition presented here terminates if the inlined version
  does. Hence, although not directly structural, this definition is
  indeed terminating.} So, $\SYMBind$ is our first
operation generic over descriptions, albeit hardwired.  Any
datatype we define comes with induction.

Note that the very same functors \(\SYMBdescop{\M{D}}\) also admit greatest
fixpoints: we have indeed implemented coinductive types this way, but
that is another story.

\begin{figure*}

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\SYMBAll : & \PITEL{\V{D}}{\Desc^n}
             \PITEL{\V{X}}{\Set^n}
             \PITEL{\V{P}}{\V{X} \To \Set^k} \\
           & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
             \Set^k 
\end{array} \\
\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
\All{\DUnit}{& \V{X}}{\V{P}}{\Void} &
    \Unit \\
\All{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{s}}{\V{d}}{}} &
    \All{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{d}} \\
\All{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{x}}{\V{d}}{}} &
    \TIMES{\V{P}\: \V{x}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}} \\
\All{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\pair{\V{f}}{\V{d}}{}} &
    \TIMES{(\PI{\V{h}}{\V{H}} \V{P}\: (\V{f}\: \V{h}))}
          {\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}
\end{array}
}
&
%%
\stk{
\begin{array}{@{}ll}
\SYMBall : & \PITEL{\V{D}}{\Desc^n}
             \PITEL{\V{X}}{\Set^n}
             \PITEL{\V{P}}{\V{X} \To \Set^k} \\
           & \PITEL{\V{p}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
             \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
             \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
\end{array} \\
\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
\all{\DUnit}{& \V{X}}{\V{P}}{\V{p}}{\Void} &
    \Void \\
\all{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{s}}{\V{d}}{}} &
    \all{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{p}}{\V{d}} \\
\all{(\DIndx{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{x}}{\V{d}}{}} &
    \pair{\V{p}\: \V{x}}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}{} \\
\all{(\DHindx{\V{H}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{f}}{\V{d}}{}} &
    \pair{\LAM{\V{h}} \V{p}\: (\V{f}\: \V{h})}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}{}
\end{array}
\end{array}
}
\]

\caption{Defining and collecting inductive hypotheses}
\label{fig:all-predicates}

\end{figure*}

\subsection{Extending type propagation}

\begin{wstructure}
<- Extending type propagation
    <- datatype declaration turns into definitions
        -> Straightforward translation to Desc
        -> Creation of a variable referring to the structure
    <- Labelled Mu
        /> Just mention the possibility of labelling, no details required
        -> For the user, objects have names rather than Mu of codes
    <- Push Mu to an applied name [figure]
        -> Direct integration into the type propagation machinery
    -> Coded presentation is practical
        <- The user never see a code
\end{wstructure}

We now have low level machinery to build and manipulate inductive
types. Let us apply cosmetic surgery
to reduce the syntactic overhead. We extend type checking of
expressions:
%
\[
\Rule{\Gamma\Vdash\propag{\push{\etag{c}}{\EnumT{\M{E}}}}{n} \quad
\Gamma\Vdash  
\propag{
  \push{\sqr{\vec{t}}}
   {\descop{\M{D}\: n}{(\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})})}}}
        {\M{t'}}}
     {\Gamma\Vdash\propag{\push{\etag{c}\: \vec{\M{t}}}
                   {\Mu{(\DSigma{\EnumT{\M{E}}}{\M{D}})}}}
             {\Con{\pair{n}{\M{t'}}{}}}}
\]
%
Here $\etag{c}\:\vec{t}$ denotes a tag `applied' to a sequence
of arguments, and \(\sqr{\vec{t}}\) that sequence's repackaging as
a right-nested tuple. Now we can just write data directly.
\[
\NatZero:\Nat\qquad \NatSuc{(\Bhab{n}{\Nat})}:\Nat
\]
Once again, the type explains the legible presentation, as well as
the low-level representation.

We may also simplify appeals to induction by type propagation, as we
have done with functions from pairs and enumerations.
\[
\Rule{\Gamma\Vdash\stk{
\propag{\push{\M{f}}
             {\PI{\V{d}}{\descop{\M{D}}{(\Mu{\M{D}})}}       
                \All{\M{D}}{(\Mu{\M{D}})}{(\PLAM{\x}{\Mu{\M{D}}}\M{P})}{\V{d}} \To \M{P}[\Con{\V{d}}/\x]\\}}
       {\M{f'}}}}
{\Gamma\Vdash
\propag{\push{\sind{\M{f}}}
             {\PIS{\Bhab{\x}{\Mu{\M{D}}}}\M{P} }}
       {\F{ind}\:\M{D}\:(\PLAM{\x}{\Mu{\M{D}}}\M{P})\:\M{f'}}}
\]
This abbreviation is no substitute for the dependent pattern
matching to which we are entitled in a high-level language built on top
of this theory~\cite{goguen:pattern-matching}. It does at least make
`assembly language' programming mercifully brief, albeit hieroglyphic.
\[\stk{
\F{plus}:\Nat\To\Nat\To\Nat \\
\F{plus}\mapsto\sind{\spl{\sqr{(\LAM{\_}\LAM{\_}\LAM{\V{y}}\V{y})
  \quad(\LAM{\_}\spl{\LAM{\V{h}}\LAM{\_}\LAM{\V{y}}
    \NatSuc{(\V{h}\:\V{y})}})}}}
}\]

This concludes our introduction to the universe of datatype descriptions.
We have encoded sum-of-products datatypes from the simply-typed world
as data and equipped them with computation. We have also made sure
to hide the details by type propagation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Levitating the universe of descriptions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Levitating the Universe of Descriptions}
\label{sec:desc-levitate}

In this section, we will fulfil our promises and show how we implement
the signatures, first for the enumerations, and then for the codes of
the $\Desc^n$ universe.  Persuading these programs to perform was a perilous
pedagogical peregrination for the protagonist.  Our method was indeed to
hardwire constants implementing the signatures specified above, in the
first instance, but then attempt to replace them, step by step, with
\emph{definitions}: ``Is \(2+2\) still \(4\)?'', ``No, it's a loop!''.
But we did find a way, so now we hope to convey to you the
dizzy feeling of levitation, without the falling.

\subsection{Implementing finite enumerations}

\begin{wstructure}
<- Recall typing rules of 1st section
    -> Make clear they were just promises
    -> Can be implemented now
        <- Simply List UId
\end{wstructure}

In Section~\ref{sec:finite-sets}, we specified the
finite sets of tags. We are going to implement (at every universe level)
the $\EnumU$ type former and its constructors. Recall:
%
\[\Type{\EnumU}^n\qquad\Bhab{\NilE}{\EnumU}\qquad
\Bhab{\ConsE{(\Bhab{\M{t}}{\UId})}{(\Bhab{\M{E}}{\EnumU})}}{\EnumU}
\]
%
The $\NilE$ and $\SYMBConsE$ constructors are just the `nil' and
`cons' or ordinary lists, with elements from \(\UId\).
Therefore, we
implement:
%
\[
\EnumU \mapsto \Mu{(\ListD\: \UId)}
\qquad
\NilE \mapsto \ListNil
\qquad
\ConsE{\M{t}}{\M{E}} \mapsto \ListCons{\M{t}}{\M{E}}
\]

\begin{wstructure}
<- Consequences
    -> Type theory doesn't need to be extended with EnumU, NilE, and ConsE
        <- EnumU == Mu EnumUD
        <- NilE, ConsE are just tags
    -> Do not need a specific \spi eliminator
        <- \spi is an instance of the generic eliminator
            <- Code?
    -> Anything else remains the same (switch, EnumT, 0, 1+)
\end{wstructure}

Let us consider the consequences. We find that the type
theory does not need a special type former $\EnumU$,
or special constructors $\NilE$ and $\SYMBConsE$. 
Moreover, the $\spi{\M{E}}{\M{P}}$ operator, computing tuple types
of \(\M{P}\)s by recursion on \(\M{E}\) need not be hardwired: we can
just use the generic $\F{ind}$ operator, as we would for any ordinary
program.

Note, however, that the universe decoder \(\EnumT{\M{E}}\) \emph{is}
hardwired, as are the primitive \(\Ze\) and \(\Su\) that
we use for low-level values, and indeed the \(\SYMBswitch\) operator.
We cannot dispose of data altogether! We have, however, gained
the ordinariness of the enumeration \emph{codes}, and hence of generic
programs which manipulate them. Our next step is similar: we are going to
condense the entire naming scheme of datatypes \emph{into itself}.

%% Apart from these changes, we are left with implementing the other
%% components of finite sets. This consists of $\EnumT$, $\Ze$, $\Su$, as
%% well as the $\F{switch}$ operator. Note that the actual implementation
%% of $\EnumU$ does not influence our implementation of $\EnumT$: be they
%% hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects
%% behave the same. We have witnessed this effect when carrying the
%% operation in Epigram, moving from an hard-coded presentation to a
%% self-hosted one. Absolutely no change to the $\EnumT$ objects was
%% required.

\begin{wstructure}
<- Summary of the operation
    <- The content of the type theory is exactly the same
        <- before == after
    /> type naming scheme condenses
        <- Replace named constructors by codes in the universe of datatypes
    -> Our next step is a similar move (in essence)
        /> Condenses the entire naming scheme of datatypes
\end{wstructure}

%In this section, we have replaced a low-level presentation of finite
%sets by a self-hosted one, expressed in the universe of
%descriptions. However, formally, the content of the type theory
%remains unchanged: objects that were present before the modification
%are still there. Conversely, we have not introduced any spurious
%object.

%If not on the content, this modification had an effect on the
%names. The type naming scheme of the type theory has condensed: named
%type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
%now replaced by codes and their fixpoint in the universe of
%descriptions. In essence,

\subsection{Implementing descriptions}

\begin{wstructure}
<- Realising our promises
    <- We are going to implement Desc
    /> Desc is itself a datatype
        <- Same situation as EnumU
            <- We want to benefit from generic operations
        -> It ought to be encoded in itself
\end{wstructure}

The set of codes, $\Desc$, is already some sort of datatype; as with $\EnumU$,
we ought to be able to describe it, coding of $\Desc^n$ in $\Desc^{n+1}$, spiralling
upwards. Hence, this code would be a first-class citizen, born with
the generic equipment of datatypes.

\subsubsection{First attempt}

\begin{wstructure}
<- A partial implementation
    <- '1 and 'indx are easy
    <- 'sigma is partially doable
        /> lack the ability to do an higher-order inductive call
    -> Show partial code [figure]
\end{wstructure}

Our first attempt gets stuck quite quickly:
%
\[\stk{
\DescD^n : \Desc^{n+1} \\
\DescD^n \mapsto
  \tada{\stkm{\DUnit\\ \SYMBDSigma\\ \SYMBDIndx}}
       {\stkm{ \DUnit                                \\
         \DSigma{\Set^n}{\LAM{\V{S}} \SHED}      \\
         \DIndx{\DUnit}    }                    } \\
%\begin{array}{@{}ll}
%\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
%                               & \left[\begin{array}{l}
%                                   \DUnit                                \\
%                                   \DSigma{\Set}{\LAM{\V{S}} \SHED}      \\
%                                   \DIndx{\DUnit}                        \\
%                                 \end{array}
%                                 \right]
%\end{array}
}\]
%
Let us explain where we stand. Much as
we have done so far, we first offer a constructor choice from
$\DUnit$, $\SYMBDSigma$, and $\SYMBDIndx$. You may notice
that the `tagged' notation we have used for the $\Desc^n$ constructors now
fits the facts: these were actually the tags we are defining.
For $\DUnit$, we immediately reach the end of the description. For
$\SYMBDIndx$, there is a single recursive argument. Describing
$\SYMBDSigma$ is problematic. Recall the specification of
$\SYMBDSigma$:
%
\[
\DSigma{(\Bhab{\V{S}}{\Set^n})}{(\Bhab{\V{D}}{\V{S} \To \Desc^n})} : \Desc^n\\
\]
%
So, we first pack a  $\Set^n$, $\V{S}$, as well we might when working in
$\Desc^{n+1}$. We should then like
a recursive argument \emph{indexed} by $\V{S}$, but
that is an \emph{exponential}, and our presentation so far delivers only
sums-of-products. To code our universe, we must first enlarge it!

\subsubsection{Second attempt}

\begin{wstructure}
<- Extending the universe of description
    -> With higher-order induction
    <- Intuition: index elements in X by H, and go on reading
        -> indx is isomorph to hindx for H = 1
    /> Keep indx
        <- First order!
        -> Extensionally equal to hindx 1
        /> Practically, definitional equality on Sigma/Pi cannot cope with it
    -> Show DescD code
\end{wstructure}

In order to capture a notion of higher-order induction, we add a code
$\SYMBDHindx$ that takes an indexing set $\V{H}$. This amounts to give
a recursive subobject for each element of $H$.
%
\[\stk{
    \DHindx{(\Bhab{\M{H}}{\Set^n})}{(\Bhab{\M{D}}{\Desc^n})} : \Desc^n
    \smallskip \\
    \descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     
        \mapsto 
            \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
}\]

Note that up to isomorphism, $\SYMBDIndx$ is subsumed by
$\DHindx{\Unit}{\!}$. However, the apparent duplication has some
value.  Unlike its counterpart, $\SYMBDIndx$ is first-order: we prefer
not to demand dummy functions from \(\Unit\) in ordinary data,
e.g. \(\NatSuc{(\LAM{\_}n)}\). It is na{\"\i}ve to imagine that up to
isomorphism, any representation of data will do.  First-order
representations are finitary by construction, and thus admit a richer,
componentwise decidable equality than functions may in
general possess.\footnote{E.g., extensionally, there is one function in
\(\EnumT{\void}\To\Nat\); intensionally, there is a countable infinitude
which it is dangerous to identify definitionally.}

We are now able to describe our universe of datatypes:
%
\[\stk{
\DescD^n : \Desc^{n+1} \\
\DescD^n \mapsto\tada
{\stkm{ \DUnit\\ \SYMBDSigma\\ \SYMBDIndx\\ \SYMBDHindx}}
{\stkm{
\DUnit                                            \\
                                   \DSigma{\Set^n}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set^n}{\LAM{\_} \DIndx{\DUnit}}
}}
}\]
%
The $\DUnit$ and $\SYMBDIndx$ cases remain unchanged, as expected. We
successfully describe the $\SYMBDSigma$ case via
the higher-order induction, branching on $\V{S}$. The $\SYMBDHindx$ case
just packs a $\Set^n$ with a recursive argument.

At a first glance, we have achieved our goal. We have described the
codes of the universe of descriptions. The fixpoint of \(\descop{\DescD^n}\)
is a datatype just like $\Desc^n$, in $\Set^{n+1}$. Might we be
so bold as to take \(\Desc^n \mapsto \Mu{\DescD^n}\) as the levitating
definition? If we do, we shall come down with a bump! 
To complete our levitation, just as in the magic trick, requires
hidden assistance. Let us explain the problem and reveal the `invisible
cable' which fixes it.

\subsubsection{Final move}

\begin{wstructure}
<- Subtlety: translation of [ ... ]
    -> Let us do it manually
        -> Code with problem for the motive of switch
\end{wstructure}

The definition \(\Desc^n \mapsto \Mu{\DescD^n}\) is circular,
but the offensive recursion is concealed by a prestidigitation.

Expanding \(\toDesc{-}\) and propagating types as in
Figure~\ref{fig:type-checking} reveals the
awful truth:
\[
\Desc^n\mapsto
\Mu{\stk{(\DSigma{\EnumT{\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}}\\}
    {\;\SYMBswitch\:\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}\:(\LAM{\_}\Desc^{n+1})\\
\;\sqr{\stkm{
\DUnit                                            \\
                                   \DSigma{\Set^n}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set^n}{\LAM{\_} \DIndx{\DUnit}}
}}})}}
\]
The recursion shows up only because we must specify the return type
of the general-purpose \(\F{switch}\), and it is computing a \(\Desc^{n+1}\)!
Although type propagation
allows us to hide this detail \emph{when defining a function}, we cannot
readily suppress this information and check types
when \(\F{switch}\) is fully applied.

\begin{wstructure}
<- The magician trick
    <- Our problem is to give a motive for switch
        /> We perfectly know what it ought to be: \_ -> DescD
    -> Solution: extend the type theory with a special purpose switchD
        -> Only extension required to the type theory!
        -> Hidden away to the user by the syntactic sugar
            -> Sufficient to ensure unavailability as a raw operator
            <- Another instance of type propagation
\end{wstructure}

%What happens if we unfold the definition? We ought to build the following term:
%
%\[
%\PLAM{\V{x}}{(\EnumT{\V{E}})} \switch{\V{E}}{(\LAM{\_} %\Mu{\DescD})}{\V{\pi^f}}{\V{x}}
%\]
%
%But this is quite problematic. We are still in the process of
%constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
%begging for this very same $\DescD$. Despite our willingness, we
%cannot materialise such motive. However, we perfectly know what the
%motive is.

We are too close to give up now. If only we did not need to
supply that return type, especially when we know what it must be!
We eliminate the recursion by \emph{specialising} \(\F{switch}\):
%
\[
\F{switchD} : \PITEL{\V{E}}{\EnumU}   \To
                (\spi{\V{E}}{\LAM{\_} \Desc^m}) \To
                \EnumT{\V{E}} \To \Desc^m
\]
%
The magician's art rests here, in this extension. We conceal it
behind a type propagation rule for \(\F{switchD}\) which we apply
with higher priority than for \(\F{switch}\) in general.
%
\[
\Rule{\Gamma \Vdash
  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{\PLAM{x}{\EnumT{E}}\Desc^m}}}
                           {\M{t'}}}
     {\Gamma \Vdash
\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc^m}}
 {\switchD{\M{E}}{\M{t'}}}}
\]
As a consequence, our definition above now propagates without
introducing recursion. Of course, by pasting together the declaration
of \(\Desc^n\) and its internal copy, we have made it appear in its own
type. Hardwired as a trusted \emph{fait accompli}, this creates no regress,
although one must assume the definition to recheck it.

Our Agda model does not formalise the $\SYMBswitchD$
construction. Instead, we exhibit the isomorphism between declared
and encoded descriptions. Here, $\SYMBswitchD$ lets us collapse
this isomorphism, operationally identifying defined and coded
descriptions.

There are other ways to achieve a sufficient specialisation to avoid a
recursive code, e.g., extending $\Desc^n$ with specialised codes for
\emph{finite} sums and products, pushing the $\SYMBswitch$ into the
interpretation of codes, rather than the code itself. Here, we prefer
not to add codes to $\Desc^n$ which are otherwise unmotivated.
%%
%\[\stk{
%    \Dsigma{(\Bhab{\V{E}}{\EnumU})}
%           {(\Bhab{\V{B}}{\spi{\V{E}}{\LAM{\_}{\Desc^n}}})} : \Desc^n
%    \smallskip \\
%    \descop{\Dsigma{\V{E}}{\V{B}}}{\V{X}}     
%        \mapsto 
%            \SIGMAS{\Bhab{\V{x}}{\EnumT{\V{E}}}}
%                   %{\descop{\switch{\V{E}}{(\LAM{\_}{\Desc})}{\V{B}}{\V{x}}}{\V{X}}}    
%}\]
%%
%Using $\SYMBDsigma$ to switch over the $\Desc$ constructors, we avoid
%mention of $\Desc$ while levitating it. The $\SYMBswitch$ operator is
%pushed into the interpretation function, at which point $\Desc$ indeed
%exists.

%% * There exists alternative
%% **   <- All we want is to avoid mention of Desc
%% **   -> Add a code for finite sigma : $\sigma : (e : \EnumU) -> \pi e \Desc -> \Desc$
%% **       <- Take a finite set in argument
%% **       <- Interpreted as a switch
%% **       -> No more need to mention Desc during its construction
%% **           <- Pushed into intrepretation, at which time it exists

\begin{wstructure}
<- Generic programming now!
    <- Desc is just data
        -> Can be manipulated
    <- Free induction scheme on Desc
        -> Ability to inspect datatypes
        -> Ability to program on datatypes
\end{wstructure}

We have levitated \(\Desc\) at every level. Beyond its pedagogical
value, this exercise has several practical outcomes. First, it
confirms that each $\Desc$ universe is just plain data. As any piece of
data, it can therefore be inspected and manipulated. Moreover, it is
expressed in a $\Desc$ universe. As a consequence, it is equipped,
for free, with an induction principle. So, our ability to inspect and
program with $\Desc$ is not restricted to a meta-language: we have
the necessary equipment to program with \emph{data}, so we can program over
datatypes. \emph{Generic programming is just programming}.

\subsection{The generic catamorphism}

\begin{wstructure}
<- Making cata
    <- Present the type signature
    <- Starts with a call to generic induction
        <- induction on Desc!
        /> Show types at hand
        -> Explain how to use inductive hypothesis
    <- Implement the 'replace' function
    -> Dependent-typeless catamorphism 
\end{wstructure}

In Section~\ref{sec:desc-fix-point}, we hardwired a dependent
$\F{ind}$uction principle, but sometimes,
iteration suffices. Let us construct the catamorphism.

\newcommand{\cata}{\F{cata}}

We proceed by induction on the \emph{data} in \(\Mu{\V{D}}\):
the non-dependent return type $\V{T}$ is readily propagated.
Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:
%
\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f} \mapsto
  \sind \LAM{\V{xs}}\LAM{\V{hs}} \V{f}\: \SHED
}\]
% 
The hole remains: we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{\Mu{\V{D}}}{(\LAM{\_} \V{T})}{\V{xs}}}\)
to hand, and we need a \(\descop{\V{D}}{\V{T}}\).
Now, $\V{xs}$ has the right shape, but its
components have the wrong type. However, for each such
component, $\V{hs}$ holds the corresponding value
in $\V{T}$.  We need a function to \(\F{replace}\) the former with the latter:
this pattern
matching sketch yields an induction on \(\V{D}\). We fill the hole
with \(\F{replace}\: \V{D}\:(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\).
%
\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\begin{array}{@{}l@{\:}l@{\:}l@{\:\:\mapsto\:\:}l}
\F{replace}\: \DUnit &          \V{X}\: \V{Y}\: \Void&          \Void   &
    \Void  \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})& \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}& \V{d'}  &
    \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
\F{replace}\: (\DIndx{\V{D}})&     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}& \pair{\V{y}}{\V{d'}}{} &
    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})& \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}& \pair{\V{g}}{\V{d'}}{} &
 \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
\end{array}
}\]
%

%% The astute reader will have been struck by the type of $\F{replace}$:
%% it is \emph{almost} the morphism part -- sometimes called \emph{map}
%% -- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
%% the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
%% is the dependent version of the map, which uses the induction
%% hypotheses. For space reason, we will not present the non-dependent
%% map. It can be found in the Agda model.

\begin{wstructure}
<- Deriving generic functions
    <- Taking a Desc and computing a function
        <- Desc comes equipped with an induction principle
        -> Ability to compute more functions from it
            -> More generic functions
    <- Inspecting datatypes
        <- All described byu a Desc code
        -> Ability to explore the code
            <- Desc equipped with an induction principle
            -> Build new objects based on that structure
\end{wstructure}

We have shown how to derive a generic
operation, \(\cata\), from a pre-existing generic operation,
\(\F{ind}\), by
manipulating descriptions as data: the catamorphism is just a
function taking each $\Desc$ value to a datatype specific
operation. This is polytypic programming, as in
PolyP~\cite{jansson:polyp}, made ordinary.

%% Moreover, the $\F{replace}$ function demonstrates the benefit of an
%% approach based on universes. The datatypes living in the universe of
%% descriptions, we are able to \emph{inspect} them. As shown by
%% $\F{replace}$, it is easy to explore these structures, as well as
%% building new ones.

\subsection{The generic free monad}
\label{sec:desc-free-monad}

\begin{wstructure}
<- A generic program: the free monad construction
    <- Recall free monad construction in Haskell
        -> Based on a functor F
    <- Note that the free monad construction is itself defined by a functor
        -> Extract it
\end{wstructure}

In this section, we try a more ambitious generic operation. Given a
functor---a signature of operations represented as a tagged
description---we build its free monad, extending the signature with
variables and substitution.
%
\newcommand{\FMFreeMonad}{\D{FreeMonad}}
\newcommand{\FMFreeMonadD}{\D{FreeMonadD}}
\newcommand{\FMVar}{\C{Var}}
\newcommand{\FMComposite}{\C{Composite}}
%
Let us recall this construction in, say, Haskell. Given a
functor \texttt{f}, the free monad over \texttt{f} is given
thus:
%
\[
\texttt{data FreeMonad f x = Var x | Op (f (FreeMonad f x))}
\]
%
Provided \texttt{f} is an instance of \texttt{Functor}, we may
take \texttt{Var} for \texttt{return} and use \texttt{f}'s
\texttt{fmap} to define \texttt{>>=} as substitution.

Being an inductive type, $\FMFreeMonad$ arises
by a pattern functor:
%
\[
\FMFreeMonadD\: \V{F}\: \V{X}\: \V{Z} \mapsto \V{X} \mathop{\D{+}} \V{F}\:\V{Z}
\]

\begin{wstructure}
    <- Encode it in the Desc world [equation]
        <- F is the Desc we start with
        <- The free monad functor is what we have just defined
        <- [\_]* : Desc -> Set -> Desc
           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
        -> Mu does the fixpoint
\end{wstructure}

Our construction takes the functor as a
tagged description, and given a set $\V{X}$ of variables, computes the
tagged description of the free monad pattern functor.
%
\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{\V{E}}{\V{D}}{}}\:\V{X} \mapsto
    \pair{\pair{\DVar{}}{\V{E}}{}}
         {\pair{\DSigma{\V{X}}{\DUnit}}{\V{D}}{}}{}
}\]
%
We simply add a constructor, $\SYMBDVar$, making its arguments
$\DSigma{\V{X}}{\DUnit}$---just an element of $\V{X}$. 
$\V{E}$ and $\V{D}$ stay put, leaving the other constructors
unchanged. Unfolding the interpretation of this definition, we
find an extended sum, corresponding to the \(\V{X}\D{+}\) in
$\FMFreeMonadD$. Taking the fixpoint ties the knot and we have our
data.

\begin{wstructure}
<- A generic program: monadic substitution [equation]
    <- subst : \forall T X Y. mu ([T]* X) -> (X -> mu ([T]* Y)) -> mu ([T]* Y)
        -> Using Fold
\end{wstructure}

\newcommand{\subst}{\F{subst}}
\newcommand{\apply}{\F{apply}}

Now we need the operations. As expected, \(\LAM{\x}\DVar{\x}\)
plays the r\^ole of \return, making variables terms. Meanwhile,
\bind is indeed \emph{substitution}, which we now
implement generically, making use of $\cata$. Let us write the type,
and start filling in the blanks:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \mude{(\FreeMonad{\V{D}}{\V{Y}})}) \To \\
         & \mude{(\FreeMonad{\V{D}}{\V{X}})} \To
           \mude{(\FreeMonad{\V{D}}{\V{Y}})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
          (\mude{(\FreeMonad{\V{D}}{\V{Y}})})\: 
          \SHED
}\]
%
We are left with implementing the algebra of the catamorphism. Its
role is to catch appearances of $\DVar{\V{x}}$ and replace them by
$\V{\sigma}\: \V{x}$. This corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \mude{(\FreeMonad{\V{D}}{\V{Y}})}) \To \\
         & \descop{\toDesc{(\FreeMonad{\V{D}}{\V{X}})}}{(\mude{(\FreeMonad{\V{D}}{\V{Y}})})}
           \To  \mude{(\FreeMonad{\V{D}}{\V{Y}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \sqr{\DVar{\V{x}}}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
\end{array}
}\]

\begin{wstructure}
    -> Consequences
        <- We have free monad datatype
            <- Term + variables
        <- We have monad operations
            <- Return / var
            <- Substitution / bind
\end{wstructure}

We complete the hole with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$. Every tagged
description can be seen as a signature of operations: we can uniformly
add a notion of variable, building a new type from an old one, then
providing the substitution structure.

\begin{wstructure}
<- Deriving new data-structure and functions on them
    <- Computing the Free Monad of a datatype
        <- Derive new data-structure from previous one
            <- It is just code
        /> New data-structure comes with some equipment
    <- Computing new functions on computed datatypes
        <- If data comes with structure, we ought to be able to capture it
            <- Induction on Desc
            -> Ability to compute over data
\end{wstructure}

%% Candidate for removal:
%% With the free monad construction, we have seen two kinds of generic
%% operations. Firstly, we have derived a new data-structure from another
%% one: we make the free monad from its underlying functor. To do so, we
%% crucially rely on the fact that datatypes are nothing but codes. We
%% are therefore entitled to modify this code and, in this case, extend
%% it. Extending a datatype might give rise to a more structured object,
%% as was the case here.  So, secondly, we have equipped this new
%% datatype with its inherent structure: the \bind\ and
%% \return\ operations. We have been able to build them as generic
%% functions.

\subsection{Skyhooks all the way up?}

%% * Desc in Desc
%% ** /> Blur the lines between implementation and reflection
%% ** /> Abusing the paradoxical nature of Set : Set

In this section, we have seen how to \emph{levitate}
descriptions. Although our theory, as presented here, takes
$\Bhab{\Set}{\Set}$, our annotations indicate how a stratified theory
could code each level from above.  We do not rely on the paradoxical
nature of $\Bhab{\Set}{\Set}$ to flatten the hierarchy of descriptions
and fit large inside small.  We shall now be more precise about what
we have done.

%% * Summary 
%% ** <- What is the equipment for making data-types
%% ** <- What is reflected, what is implemented
%% ** <- (table 1)

Let us first clarify the status of the implementation. The kit for
making datatypes is presented in Table~\ref{tab:sumup-operators}. For
each operation, we describe its role and its status, making clear
which components are self-described and which ones are actually
implemented.

%% * Paradox
%% ** <- How do we bottom out in a stratified setting?
%% ** <- Spiral
%% ***    <- Encoding of DescD_n : Desc_{n+1}
%% ***    <- Desc_n = Mu DescD_n : Set_{n+1}
%% ** <- Agda model: Desc_42
%% ***    <- No dependent pattern-matching
%% ***    <- No IR
%% ***    <- Normal universes
%% ***    -> Straight Agda (UTT)
%% ** <- ``Self-encoding'' only in a level polymorphic sense
%% ***    -> Agda model: set poly

In a stratified system, the `self-encoded' nature of $\Desc$ appears
only in a set polymorphic sense: the principal type of the encoded
description generalises to the type of $\Desc$ itself. We encode this
much in our set polymorphic model in Agda and in our Coq model,
crucially relying on typical ambiguity~\cite{harper:implicit-universe}.
We step outside current technology only to replace the declared $\Desc$
with its encoding.

Even this last step, we can approximate within a standard predicative
hierarchy. Fix a top level, perhaps \(42\). We may start by declaring
$\Desc^{42}:\Set^{43}$. We can then construct $\DescD^{41}:\Desc^{42}$
and thus acquire an encoded $\Desc^{41}$. Although $\Desc^{41}$ is
encoded, not declared, it includes the relevant descriptions,
including $\DescD^{40}$. We can thus build the tower of descriptions
down to $\Desc^0$, encoding every level below the top.  Description of
descriptions forms a `spiral', rather than a circle. We have modelled
this process exactly in Agda, without any appeal to dependent
pattern matching, induction-recursion, or set polymorphism. All it
takes to build such a sawn-off model of encodings is inductive
definition and a cumulative predicative hierarchy of set levels.

\begin{table}

{
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
Object                & Role                        & Status \\
\hline
\hline
$\EnumU$              & Build finite sets           & Levitated \\
\hline
$\Desc$               & Describe pattern functors   & Levitated \\
\hline
$\SYMBdescop{\_}$     & Interpret descriptions      & Hardwired \\
\hline
$\SYMBMu$, $\SYMBCon$ & Define, inhabit fixpoints   & Hardwired \\
\hline
$\SYMBind$, $\SYMBAll$, $\SYMBall$  
                      & Induction principle         & Hardwired \\
\hline
\end{tabular}
\end{center}
}

\caption{Summary of constructions on Descriptions}
\label{tab:sumup-operators}

\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Indexing descriptions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{A Universe of Inductive Families}
\label{sec:indexing-desc}

\newcommand{\vtup}[2]{\bigRedBracket{\begin{array}{@{}#1@{}}#2\end{array}}}

\begin{wstructure}
!!! Need Help !!!
<- Motivation
    <- Desc: expressivity of simply-typed datatypes: inductive types
        <- Values do not influence types
    /> Example: Vectors
        <- Cannot be defined by just induction
            <- Vectors of all size need to be defined at the *same* time
            -> Defined as a *family* of types
                -> Index
        -> I -> IDesc I: Inductive family
    ???
\end{wstructure}

So far, we have explored the realm of inductive types, building on
intuition from ML-like datatypes, using type dependency as a
descriptive tool in $\Desc$ and its interpretation. Let us now make
dependent types the object as well as the means of our study.

Dependent datatypes provide a way to work at higher level of
precision \emph{a priori}, reducing the sources of failure we
might otherwise need to manage. For the perennial
example, consider \emph{vectors}---lists indexed by length. By
making length explicit in the type, we can prevent hazardous
operations (the type of `head' demands vectors of length
$\NatSuc{\V{n}}$) and offer stronger guarantees (pointwise
addition of $\V{n}$-vectors yields an $\V{n}$-vector).

However, these datatypes are not \emph{individually} inductive. For
instance, we have to define the whole \emph{family} of vectors
mutually, in one go. In dependently typed languages, the basic grammar
of datatypes is that of inductive families. To capture this grammar,
we must account for \emph{indexing}.

%% \subsection{Desc, atomically}
%% \label{sec:idesc-atomic-desc}

%% \begin{wstructure}
%% [Outdated: type former presentation instead]
%% <- Adding hindx have introduced some duplication
%%     <- indx == hindx 1
%%     -> We can factor out commonalities 
%%         /> Obtain an equivalent presentation
%%         /> Still embeddable (refer to the Agda model)
%% \end{wstructure}

%% \begin{wstructure}
%% <- Also replacing '1 by 'const  [figure]
%%     <- For convenience
%%         <- 'const X equivalent to 'sigma X (\_ -> '1)
%%         /> Easier to abstract
%%             <- Extensionally same
%%             /> 'const more useful in practice
%% \end{wstructure}

%% Before moving on to indexed descriptions, we have to carry out some
%% maintenance work on descriptions. We presented $\Desc$ as the grammar
%% of inductive types. Hence, the codes closely follow this grammar. In
%% the following, we adopt an alternative presentation. With
%% $\DSigma{\!}{\!}$, we are actually \emph{quoting} a standard
%% type-former, namely

%% $$\Bhab{\Sigma}{\PI{\V{S}}{\Set} \PI{\V{S}}{\Set} \Set}$$

%% In the alternative presentation, we go further and present all our
%% codes as quotations of standard type-formers. This presentation is
%% shown in Figure~\ref{fig:type-former-desc}. The reader will notice
%% that we replace $\DUnit$ by a more general $\DConst{\!}$ code. Whereas
%% $\DUnit$ was interpreted as the unit set, $\DConst{\V{X}}$ is
%% interpreted as $\V{X}$, for any $\Bhab{\V{X}}{\Set}$. Extensionally,
%% $\DConst{\V{X}}$ and $\DSigma{\V{X}}{\DUnit}$ are equivalent. However,
%% $\DConst{\!}$ is more succinct. More importantly, $\DConst$ is
%% \emph{first-order}, unlike its equivalent encoding. From a
%% definitional perspective, we are giving more opportunities to the
%% type-system, hence reducing the burden on the programmer. For the same
%% reason, we introduce $\DProd{\!}{\!}$ that overlaps with
%% $\DSigma{\!}{\!}$.

%% This reorganisation is strictly equivalent to the previous one
%% (Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
%% self-descriptive. We refer the reader to the companion technical
%% report for details. In this finer-grained presentation, we can define
%% $\DIndx{\!}$ and $\DHindx{\!}{\!}$ as follow:

%% \[\begin{array}{l@{\:\mapsto\:\:}l}
%% \DIndx{\V{D}}         & \DProd{\DId}{\V{D}}                      \\
%% \DHindx{\V{H}}{\V{D}}     & \DProd{(\DPi{\V{H}}{(\LAM{\_} \DId)})}{\V{D}}
%% \end{array}
%% \]

%% Consequently, the examples previously developed can be
%% straightforwardly translated into this new presentation. For example,
%% here is the new definition of $\NatD$:

%% \[\stk{
%% \NatD : \Desc \\
%% \NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
%%                      {[ \DUnit \quad \DId ]}
%% }\]

%% In the following, we adopt this last version as our de
%% facto universe of inductive types. In particular, we are going to
%% evolve this presentation into an indexed one.

%% \note{Shall we talk about the Type Theory being Desc Zero? or such story?}

%% \begin{figure}

%% \[\stk{
%% \begin{array}{ll}
%% \stk{
%% \data \Desc : \Set \where                                      \\
%% \;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
%%     \DId            & \Desc                                    \\
%%     \DConst{\!}     & \Set \To \Desc                           \\
%%     \DProd{\!}{\!}  & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
%%     \DSigma{\!}{\!} & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
%%     \DPi{\!}{\!}    & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc 
%% \end{array}
%% }
%% \vspace{0.2in}
%% \\
%% \stk{
%% \descop{\_\:}{} : \Desc \To \Set \To \Set \\
%% \begin{array}{@{}l@{\:=\:\:}ll}
%% \descop{\DId}{\V{X}}          &  \V{X}                                           \\
%% \descop{\DConst{\V{Z}}}{\V{X}}    &  \V{Z}                                           \\
%% \descop{\DProd{\V{D}}{\V{D'}}}{\V{X}} &  \TIMES{\descop{\V{D}}{\V{X}}}{\descop{\V{D}\V{'}}{\V{X}}}       \\
%% \descop{\DSigma{\V{S}}{\V{D}}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}          \\
%% \descop{\DPi{\V{S}}{\V{D}}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}            
%% \end{array}
%% }
%% \end{array}
%% }\]

%% \caption{Universe of descriptions based on Type-formers}
%% \label{fig:type-former-desc}

%% \end{figure}

\subsection{The universe of indexed descriptions}

\begin{wstructure}
<- Labelling Id
    <- We had: data Desc : Set -> Set
    -> We want: data IDesc : (I -> Set) -> Set
        <- Indexed functor (?)
        -> It is sufficient to label Id
            <- Where the functor is built
\end{wstructure}

\newcommand{\DotTo}{\mathop{\blue{\dot{\rightarrow}}}}

We presented the $\Desc$ universe as a
grammar of strictly positive endofunctors on $\Set$ and developed
inductive types by taking a fixpoint. To describe inductive families
indexed by some $\Bhab{\V{I}}{\Set}$, we play a similar game with
endofunctors on the category $\Set^{\V{I}}$,
families of sets \(\V{X},\V{Y}:\V{I}\To\Set\) for objects, and for morphisms,
families of functions in \(\V{X}\DotTo\V{Y}\), defined pointwise:
\[
\V{X}\DotTo\V{Y} \mapsto \PI{\V{i}}{\V{I}}\V{X}\:\V{i}\To\V{Y}\:\V{i}
\]

An \emph{indexed functor} in $\Set^{\V{I}}\To\Set^{\V{J}}$ has the
flavour of a device driver, characterising `responses' to a given
request in \(\V{J}\) where we may in turn make `subrequests' at indices
chosen from \(\V{I}\). When we use indexed functors to define inductive
families of datatypes,
\(\V{I}\) and \(\V{J}\) coincide: we explain how to make a node fit a given
index, including subnodes at chosen indices. E.g., if we are asked for a
vector of length 3, we choose to ask in turn for a tail of length 2.

To code up valid notions of response to a given request, we introduce
$\SYMBIDesc$ and its interpretation:
%
\[\stk{
\IDesc{(\Bhab{\V{I}}{\Set})} : \Set \smallskip \\
\idescop{\_}{}{} : _{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
}\]

An \(\IDesc{\V{I}}\) specifies just \emph{one} response, but a
request-to-response \emph{function},
$\V{R}:\V{I} \To \IDesc{\V{I}}$, yields a strictly positive endofunctor
\[
  \LAM{\V{X}} \LAM{\V{i}} \idescop{\V{R}\:\V{i}}{\V{I}}{\V{X}} :
  \Set^{\V{I}} \To \Set^{\V{I}}
\]
whose fixpoint we then take:
%
\[\stkl{
\Rule{\Gamma \vdash \Bhab{\V{I}}{\Set} \qquad
      \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}}}
     {\Gamma \vdash \Bhab{\SYMBIMu_{\V{I}}{\V{R}}}{\V{I}\To\Set}} \qquad
\\
\Rule{\begin{array}{l@{\qquad}l}
          \Gamma \vdash \Bhab{\V{I}}{\Set} &
          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
          \Gamma \vdash \Bhab{\V{i}}{\V{I}} &
          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\:\V{i}}{\V{I}}{(\SYMBIMu_{\V{I}}{\V{R}})}}
      \end{array}}
     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}}
}\]

\newcommand{\upgrade}{\F{upgrade}}
\newcommand{\inductionI}{\F{indI}}
\newcommand{\cataI}{\F{cataI}}

We define the $\SYMBIDesc$ grammar in Figure~\ref{fig:idesc},
delivering only \emph{strictly positive} families. As well as
indexing our descriptions, we have refactored a little, adopting
a more compositional algebra of codes, where $\Desc$ is
biased towards the right-nested tuples. We now have
\(\DVar{i}\) for recursive `subrequests' at a chosen index \(i\),
with tupling by right-associative
\(\DProd{}{}\) and higher-order branching
by \(\DPi{\!}{\!}\).  Upgrade your old $\Desc$
to a trivially indexed $\IDesc{\Unit}$ as follows!
\[\begin{array}{@{}ll}
\upgrade :\Desc & \To\IDesc{\Unit} \\
\upgrade\:\DUnit & \mapsto \DConst{\Unit} \\
\upgrade\:(\DSigma{\V{S}}{\V{D}}) &
   \mapsto \DSigma{\V{S}}{\LAM{\V{s}}\upgrade\:(\V{D}\:\V{s})} \\
\upgrade\:(\DIndx{\V{D}}) & \mapsto
  \DProd{\DVar{\Void}}{\upgrade\:\V{D}} \\
\upgrade\:(\DHindx{\V{H}}{\V{D}}) & \mapsto
  \DProd{(\DPi{\V{H}}{\LAM{\_}\DVar{\Void}})}{\upgrade\:\V{D}} \\
\end{array}\]

To deliver induction for indexed datatypes, we need the `holds everywhere'
machinery. We present $\SYMBAllI$ and $\SYMBallI$ in
Figure~\ref{fig:allI-predicates}, with a twist---where
$\Desc$ admits the $\SYMBall$ construction, $\SYMBIDesc$ is \emph{closed}
under it! The $\SYMBAllI$
operator for a description indexed on \(\V{I}\) is strictly positive in
turn, and has a description indexed on
 some \(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}\).
Induction on indexed descriptions is then hardwired thus:
%
\[\stk{
\begin{array}{@{}ll}
\inductionI : & _{\PI{\V{I}}{\Set}}
                   \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
                   \PI{\V{P}}{(\SIGMA{\V{i}}{\V{I}}{\IMu{\V{I}}{\V{R}}{\V{i}}}) \To \Set} \\
                 & (      \PITEL{\V{i}}{\V{I}} 
                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\SYMBIMu_{\V{I}}{\V{R}})}} \\
                 & \   \idescop{\AllI{}
                                     {(\V{R}\: \V{i})}
                                     {(\SYMBIMu_{\V{I}}{\V{R}})}
                                     {\V{xs}}}
                               {}
                               {\V{P}} \To
                       \V{P}\: \pair{\V{i}}{\Con{\V{xs}}}{}) \To \\
                 & \PITEL{\V{i}}{\V{I}}
                   \PI{\V{x}}{\IMu{\V{I}}{\V{R}}{\V{i}}}
                   \V{P}\: \pair{\V{i}}{\V{x}}{}
\end{array} \\
\inductionI\: \V{R}\: \V{P}\: \V{m}\: \V{i}\: (\Con{\V{xs}}) \mapsto 
    \V{m}\: \V{i}\: \V{xs}\: (\allI{}
                                    {\V{R}\: \V{i}}
                                    {(\SYMBIMu_{\V{I}}{\V{R}})}
                                    {\V{P}}
                                    {(\spl{\LAM{\V{i}}\LAM{\V{xs}} \inductionI\: \V{R}\: \V{P}\: \V{m}})}
                                    {\V{xs}})
}\]
%
The generic catamorphism, $\cataI$, is constructed from $\inductionI$
as before. Its type becomes more elaborated, to deal with the
indexing:
%
\[
\begin{array}{@{}l@{}l}
\cataI :& \PITEL{\V{I}}{\Set}
          \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
        & \PI{\V{T}}{\V{I} \To \Set}
          (\PI{\V{i}}{\V{I}}{\idescop{\V{R}\: \V{i}}{}{\V{T}} \To \V{T}\: \V{i}}) \To
          \SYMBIMu_{\V{I}}{\V{R}} \DotTo \V{T}
\end{array}
\]

\begin{figure*}

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\SYMBAllI : & _{\PI{\V{I}}{\Set}}
              \PITEL{\V{D}}{\IDesc{\V{I}}}
              \PI{\V{X}}{\V{I} \To \Set} \\
            & \idescop{\V{D}}{\V{I}}{\V{X}} \To
              \IDesc{(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}})}
\end{array} \\
\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
\AllI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{x}} &
    \DVar{\pair{\V{i}}{\V{x}}{}} \\
\AllI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{k}} &
    \DConst{\Unit} \\
\AllI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\pair{\V{d}}{\V{d'}}{}} &
    \DProd{\AllI{}{\V{D}}{\V{X}}{\V{d}}}{\AllI{}{\V{D'}}{\V{X}}{\V{d'}}} \\
\AllI{\:}{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\pair{\V{s}}{\V{d}}{}} &
    \AllI{}{(\V{D}\: \V{s})}{\V{X}}{\V{d}} \\
\AllI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{f}} &
    \DPi{\V{S}}{\LAM{\V{s}} \AllI{}{(\V{D}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})}}
\end{array}
}
&
%%
\stk{
\begin{array}{@{}ll}
\SYMBallI : & _{\PI{\V{I}}{\Set}}
              \PITEL{\V{D}}{\IDesc{\V{I}}}
              \PITEL{\V{X}}{\V{I} \To \Set} 
              \PI{\V{P}}{(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}) \To \Set} \\
            & (\PI{\V{x}}{\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}} \V{P}\: \V{x}) \To
              \PI{\V{xs}}{\idescop{\V{D}}{\V{I}}{\V{X}}} 
              \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{}{\V{P}}
\end{array} \\
\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
\allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{P}}{\V{p}}{\V{x}} &
    \V{p}\: \pair{\V{i}}{\V{x}}{} \\
\allI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{P}}{\V{p}}{\V{k}} &
    \void \\
\allI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{d}}{\V{d'}}{}} &
    \pair{\allI{}{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}
         {\allI{}{\V{D'}}{\V{X}}{\V{P}}{\V{p}}{\V{d'}}}{} \\
\allI{\:}{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{s}}{\V{d}}{}} &
    \allI{}{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{p}}{\V{d}} \\
\allI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\V{f}} &
    \LAM{\V{a}}\allI{}{(\V{D}\: \V{a})}{\V{X}}{\V{P}}{\V{p}}{(\V{f}\: \V{a})}
\end{array}
\end{array}
}
\]

\caption{Indexed induction predicates}
\label{fig:allI-predicates}

\end{figure*}

\begin{figure}

\[\stk{\begin{array}{@{}ll}
\IDesc{(\Bhab{\V{I}}{\Set})} &: \Set \\
\DVar{(\Bhab{\V{i}}{\V{I}})} &: \IDesc{\V{I}} \\
\DConst{(\Bhab{\V{A}}{\Set})} &:\IDesc{\V{I}}       \\
\DProd{(\Bhab{\V{D}}{\IDesc{\V{I}}})}{(\Bhab{\V{D}}{\IDesc{\V{I}}})}
  & :\IDesc{\V{I}}       \\
\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S}\To\IDesc{\V{I}}})}
& : \IDesc{\V{I}}  \\
\DPi{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S}\To\IDesc{\V{I}}})}
& : \IDesc{\V{I}}  \\
\end{array}\smallskip \\
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\V{X}}@{\:\mapsto\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{&}      &  \V{X}\: \V{i}                                           \\
\idescop{\DConst{\V{K}}}{\V{I}}{&}    &  \V{K}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{&} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{X}}}{\idescop{\V{D'}}{\V{I}}{\V{X}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{&} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{&}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}            
\end{array}
}
\]

\caption{Universe of indexed descriptions}
\label{fig:idesc}

\end{figure}

\subsection{Examples}
\label{sec:idesc-examples}

\paragraph{Natural numbers:}

\begin{wstructure}
<- Nat
    -> [equation]
    <- Non-indexed types lives in IDesc 1
        -> This applies to all previous examples
\end{wstructure}

For basic reassurance, we \(\upgrade\:\NatD\):
%
\[\stk{
\upgrade\:\NatD : \IDesc{\Unit} \\
\upgrade\:\NatD \mapsto \DSigma{(\EnumT{\sqr{\NatZero\: \SYMBNatSuc}})}
                     {\sqr{(\DConst{\Unit}) \; 
                           (\DProd{\DVar{\Void}}{\DConst{\Unit}})}}
}\]
%
Note that trailing \(\Unit\)'s keep our right-nested, \(\void\)-terminated
tuple structure, and with it our elaboration machinery.
We can similarly \(\upgrade\) any inductive type.
Moreover, \(\IDesc{I}\) can now code a bunch of mutually
inductive types, if \(I\) enumerates the
bunch~\cite{paulin:habilitation, yakushev:mutual-def}.

\paragraph{Indexed descriptions:}

\begin{wstructure}
<- Levitation [figure]
    <- Following Desc encoding
        /> Note: simple datatype
            -> Live in IDesc 1
    -> Behind the scene, relies on the special purpose switchD
\end{wstructure}

Note that $\IDesc{\V{I}}$ is a plain inductive type, parametrised
by \(\V{I}\), but indexed trivially.
%
\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\IDescD\: \V{I} \mapsto \SYMBDSigma \\
\quad
 \EnumT\vtup{r}{\SYMBDVar\\
                \SYMBDConst\\
                \DProd{}{}\\
                \SYMBDSigma\\
                \SYMBDPi}
         \; \vtup{l@{}l}{
  (\DProd{\DConst{\V{I}} &}{\DConst{\Unit}})                  \\
  (\DProd{\DConst{\Set}  &}{\DConst{\Unit}})                  \\
  (\DProd{\DVar{\Void}}{\DProd{\DVar{\Void}&}{\DConst{\Unit}}})  \\
  (\DSigma{\Set}{\LAM{\V{S}}
     \DProd{( \DPi{\V{S}}{\LAM{\_} \DVar{\Void}}) &}{\DConst{\Unit}}})     \\
  (\DSigma{\Set}{\LAM{\V{S}}
     \DProd{( \DPi{\V{S}}{\LAM{\_} \DVar{\Void}}) &}{\DConst{\Unit}}})     \\
                                   }
}\]

Therefore, this universe is self-describing and can be
levitated. As before, we rely on a special purpose $\F{switchID}$
operator to build the finite function $\bigRedBracket{\ldots}$
without mentioning \(\SYMBIDesc\).

\paragraph{Vectors:}

\newcommand{\VecD}{\F{VecD}}
\newcommand{\VecNil}{\etag{\CN{vnil}}}
\newcommand{\SYMBVecCons}{\etag{\CN{vcons}}\xspace}
\newcommand{\VecCons}[2]{\SYMBVecCons\:#1\:#2}

So far, our examples live in $\IDesc{\Unit}$, with no interesting
indexing. Let us at least have vectors. Recall
that the constructors $\VecNil$ and $\SYMBVecCons$ are defined only for
$\NatZero$ and $\NatSuc$ respectively:
%
\[
\stk{
\data \D{Vec}\: \PITEL{\V{X}}{\Set} : \PI{\V{i}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \VecNil          & \D{Vec}\:\V{X}\:{\NatZero}   \\
    \SYMBVecCons & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
\end{array}
}
\]

One way to code constrained datatypes is to appeal to a suitable
notion of propositional equality \(\PropEq\) on indices. The
constraints are expressed as `Henry Ford' equations in the datatype.
For vectors:
%
\[\stk{
\VecD : \Set \To \Nat \To \IDesc{\Nat} \\
\VecD\: \V{X}\: \V{i} \mapsto \SYMBDSigma\\
\quad
\EnumT{\vtup{r}{\VecNil\\ \SYMBVecCons}}
\; \vtup{r}{
                            (\DConst{(\NatZero\PropEq\V{i})}) \\
 ( \DSigma{\Nat}{\LAM{\V{n}}
   \DProd{\DConst{\V{X}}}
     {\DProd{\DVar{\V{n}}}{\DConst{(\NatSuc{\V{n}}\PropEq\V{i})}})}}
                          }
}\]

You may choose $\VecNil$ for any index you like as long as it is
$\NatZero$; in the $\SYMBVecCons$ case, the length of the tail is
given explicitly, and the index $\V{i}$ must be one more. Our previous
\(\Unit\)-terminated tuple types can now be seen as the trivial case
of constraint-terminated tuple types, with elaboration supplying the
witnesses when trivial.

In this paper, we remain anxiously agnostic about
propositional equality. Any will do, according to
conviction; many variations are popular. The
homogeneous identity type used in Coq is ill-suited to
dependent types, but its heterogeneous variant (forming equations
regardless of type) allows the translation of pattern
matching with structural recursion to
\(\F{indI}\)~\cite{goguen:pattern-matching}. The
extensional equality of \citet{altenkirch:ott} also sustains the translation.

\begin{wstructure}
!!! Need Help !!!
<- Brady optimisation: forcing
    <- Source to source translation
    <- Able to remove some constraints
    -> Example: Fin [figure]
    ??? More technical detail needed
\end{wstructure}

However, sometimes, the equations are redundant. 
Looking back at $\D{Vec}$, we find that the equations constrain
the choice of constructor and stored tail index retrospectively.
But \emph{inductive families need not store their
  indices}~\cite{brady:index-inductive-families}!  If we
analyse the incoming index, we can tidy our description of $\D{Vec}$
as follows:
%
\[\stk{
\VecD \:\PITEL{\V{X}}{\Set} : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}lll}
\VecD\:\V{X}\: \NatZero     & \mapsto & \DConst{\Unit} \\
\VecD\:\V{X}\: (\NatSuc{\V{n}}) & \mapsto &
 \DProd{\DConst{\V{X}}}{\DVar{\V{n}}}
\end{array}

}\]
%
The constructors and equations have simply disappeared. A similar
example is $\SYMBFin$ (bounded numbers), specified by:
%
\[
\stk{
\data \SYMBFin : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
    \SYMBFinSuc   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
}\]
%
In this case, we can eliminate equations but not constructors, since both
$\FinZero$ and $\SYMBFinSuc$ both target $\SYMBNatSuc$:
%
\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}lll}
\FinD\: \NatZero         & \mapsto & \DSigma{\EnumT{\Void}}{\Void} \\
\FinD\: (\NatSuc{\V{n}}) & \mapsto & \DSigma{\EnumT{\sqr{\FinZero\: \SYMBFinSuc}}}
                                            {\sqr{(\DConst{\Unit})\: (\DVar{\V{n}})}}
\end{array}
}\]

This technique of extracting information by case analysis on indices
applies to descriptions exactly where Brady's `forcing' and
`detagging' optimisations apply in compilation. They eliminate just
those constructors, indices and constraints which are redundant even
in \emph{open} computation. In \emph{closed} computation, where proofs
can be trusted, all constraints are dropped.

\paragraph{Tagged indexed descriptions:}

\newcommand{\SYMBmuide}{\D{\({\mu}^{\!+}\)}\xspace}
\newcommand{\muide}[2]{\SYMBmuide\!\!_{#1}\:#2}

Let us reflect this index analysis technique.
We can divide a description of tagged indexed data in two: first, the
constructors that do not depend on the index; then, the constructors
that do. The non-dependent part mirrors the definition for non-indexed
descriptions. The index-dependent part simply indexes the choice of
constructors by $\V{I}$. Hence, by inspecting the index, it is
possible to vary the `menu' of constructors.
%
\[
\begin{array}{@{}l@{\:\mapsto\:\:}l}
 \TagIDesc{\V{I}}  & \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
 \ATagIDesc{\V{I}} & \SIGMA{\V{E}}{\EnumU} \PI{\V{i}}{\V{I}} \spi{\V{E}}{\LAM{\_} \IDesc{\V{I}}} \\
 \ITagIDesc{\V{I}} & 
     \SIGMA{\V{F}}{\V{I} \To \EnumU} \PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{\LAM{\_} \IDesc{\V{I}}} 
\end{array}
\]

\begin{wstructure}
<- Vectors
    Do we treat them in the end? 
    What can we say here we haven't with Fin?
\end{wstructure}

In the case of a tagged $\D{Vec}$, for instance, for the index
$\NatZero$, we would only propose the constructor
$\ListNil$. Similarly, for $\NatSuc{n}$, we would only propose the
constructor $\SYMBListCons$.

We write $\toIDesc{\V{D}}\:\V{i}$ to denote the $\IDesc{\V{I}}$
computed from the tagged indexed description $\V{D}$ at index
$\V{i}$. Its expansion is similar to the definition of \(\SYMBtoDesc\)
for tagged descriptions, except that it must also append the two parts.
We again write $\muide{\V{I}}{\V{D}}$ for
$\IMu{\V{I}}{(\toIDesc{\V{D}})}$.

\paragraph{Typed expressions:}

\begin{wstructure}
<- Hutton's razor
    <- Types
        <- 'Nat
        <- 'Bool
    <- Term [figure]
        <- val : Val 'a -> 'a  for Val : Ty -> Set, mapping to Nat and Bool
        <- cond : 'Bool -> a -> a -> a
        <- plus : 'Nat -> 'Nat -> 'Nat
        <- le : 'Nat -> 'Nat -> 'Bool
\end{wstructure}

%% Types
\newcommand{\Ty}{\D{Ty}}
\newcommand{\Ebool}{\etag{\CN{bool}}}
\newcommand{\Enat}{\etag{\CN{nat}}}

%% Constructors
\newcommand{\SYMBEval}{\etag{\CN{val}}\xspace}
\newcommand{\SYMBEvar}{\etag{\CN{var}}\xspace}
\newcommand{\Eval}[1]{\SYMBEval\:#1}
\newcommand{\SYMBEcond}{\etag{\CN{cond}}\xspace}
\newcommand{\Econd}[3]{\SYMBEcond\:#1\:#2\:#3}
\newcommand{\SYMBEplus}{\etag{\CN{plus}}\xspace}
\newcommand{\Eplus}[2]{\SYMBEplus\:#1\:#2}
\newcommand{\SYMBEle}{\etag{\CN{le}}\xspace}
\newcommand{\Ele}[2]{\SYMBEle\:#1\:#2}

%% Index mapper (terminology?)
\newcommand{\SYMBVal}{\F{Val}\xspace}
\newcommand{\Val}[1]{\SYMBVal\:#1}
\newcommand{\SYMBVar}{\F{Var}\xspace}
\newcommand{\Var}[2]{\SYMBVar\: #1\: #2}

%% Hutton expressions
\newcommand{\HExprD}{\F{ExprD}}
\newcommand{\HExprAD}{\F{ExprAD}}
\newcommand{\HExprID}{\F{ExprID}}
\newcommand{\HExprVarD}[1]{\C{ExprD}_{\F{Var},#1}}
\newcommand{\HExprFreeD}{\C{ExprD}^{\C{Free}}}
\newcommand{\HExprAFreeD}{\C{ExprAD}^{\C{Free}}}

We are going to define a syntax for a small language with
two types, natural numbers and booleans:
%
\[
\Ty \mapsto \EnumT{\sqr{\Enat\: \Ebool}}
\]

\newcommand{\plusHost}{\mathop{\green{+_{\mathrm{H}}}}}
\newcommand{\leHost}{\mathop{\green{\leq_{\mathrm{H}}}}}

This language has values, conditional expression, addition and
comparison. Informally, their types are:
%
\[
\begin{array}{l@{\::\:\:}l}
\SYMBEval            & \Val{\V{ty}} \To \V{ty} \\
\SYMBEcond           & \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
\end{array}
\qquad
\begin{array}{l@{\::\:\:}l}
\SYMBEplus           & \Enat \To \Enat \To \Enat                           \\
\SYMBEle             & \Enat \To \Enat \To \Ebool                          \\
\end{array}
\]
%
The function $\SYMBVal$ interprets object language types in the
host language, so that arguments to $\SYMBEval$ fit their
expected type.
%
\[\stk{
\SYMBVal : \Ty \To \Set \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\Val{\Enat}   & \Nat \\
\Val{\Ebool}  & \Bool
\end{array}
}\]
%
We take $\Nat$ and $\Bool$ to represent natural numbers and Booleans
in the host language, equipped with addition $\plusHost$ and
comparison $\leHost$.

We express our syntax as a tagged indexed description, indexing over
object language types $\Ty$. We note that some constructors are always
available, namely $\SYMBEval$ and $\SYMBEcond$. On the other hand,
$\SYMBEplus$ and $\SYMBEle$ constructors are index-dependent, with
$\SYMBEplus$ available just when building a $\Enat$, $\SYMBEle$ just
for $\Ebool$. The code, below, reflects this intuition, with the first
component uniformly offering $\SYMBEval$ and $\SYMBEcond$, the second
selectively offering $\SYMBEplus$ or $\SYMBEle$.
%
%%% \begin{figure}
%
\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto \sqr{ \HExprAD , \HExprID } \\
} \smallskip\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \vtup{l}{
   {\vtup{r}{\SYMBEval\\ \SYMBEcond \,}} \red{,} \;
      \LAM{\V{ty}}
      \vtup{l@{\:}l}{
      \DProd{\DConst{(\Val{\V{ty}})}&}{\DConst{\Unit}} \\
      \DProd{\DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}&}
        {\DConst{\Unit}} \\
     }
   }
\smallskip\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \vtup{l}{
                   \vtup{r}{\sqr{\SYMBEplus} \\ \sqr{\SYMBEle}} \red{,} \;
  \LAM{\_} \sqr{\DProd{\DProd{\DVar{\Enat}}{\DVar{\Enat}}}{\DConst{\Unit}}}
                   }
}
}\]

%%%\caption{Syntax of typed expressions}
%%%\label{fig:hexpr-full}

%%%\end{figure}

\newcommand{\evalH}{\F{eval}_{\green{\Downarrow}}}
\newcommand{\evalOne}{\F{eval}_{\green{\downarrow}}}

Given the syntax, let us supply the semantics. We implement an
evaluator as a catamorphism:
%
\[\stk{
\evalH : \PI{\V{ty}}{\Ty} 
         \muide{\Ty}{\HExprD}\: \V{ty} \To
         \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{term} \mapsto \cataI_{\Ty} \:
                                 (\toIDesc{\HExprD})\: 
                                 \SYMBVal\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{term}
}\]
%
To finish the job, we must supply the algebra which implements a single
step of evaluation, given subexpressions evaluated already.
%
\[\stk{
\evalOne : \PI{\V{ty}}{\Ty}
 \idescop{(\toIDesc{\HExprD})\:\V{ty}}{\Ty}{\SYMBVal}
           \To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
\evalOne\: & \_\: & (\SYMBEval\;\V{x})                                            & \V{x} \\
\evalOne\: & \_\: & (\SYMBEcond\:\BoolTrue\:\V{x}\:\_)   & \V{x} \\
\evalOne\: & \_\: & (\SYMBEcond\:\BoolFalse\:\_\:\V{y})  & \V{y} \\
\evalOne\: & \Enat\: & (\SYMBEplus\:\V{x}\:\V{y})   & \V{x} \plusHost \V{y} \\
\evalOne\: & \Ebool\: & (\SYMBEle\:\V{x}\:\V{y})  & \V{x} \leHost \V{y} 
\end{array}
}\]

\begin{wstructure}
    /> Closed term
        <- only constants and operations on them
        -> Extend Val with Var : Ty -> Set, mapping to EnumU
            -> Open term
            -> Language of well-typed terms
                <- By construction
\end{wstructure}

Hence, we have a type-safe syntax and a tagless interpreter for our
language, in the spirit
of~\citet{augustsson.carlsson:dependent.interpreter}, with help from
the generic catamorphism. However, so far, we are only able to define
and manipulate \emph{closed} terms. Adding variables, it is possible
to build and manipulate \emph{open} terms, that is, terms in a
context. We shall get this representation, for free, thanks to the
\emph{free indexed monad} construction.

\subsection{Free indexed monad}

\begin{wstructure}
<- Variation on a theme: free imonad construction
    <- Recall existence of generic free monad construction
    -> Present its generalisation to IDesc [equation]
        <- \I -> IDesc I as describing an indexed endofunctor
        <- Free monad construction
    -> Still a suitable, generic notion of substitution
        <- show type signature
        <- show implementation?? (space! space!)
\end{wstructure}

In Section~\ref{sec:desc-free-monad}, we have built a free monad
operation for simple descriptions. The process is similar in the
indexed world. Namely, given an indexed functor, we derive the indexed
functor coding its free monad: \note{pwm: Whoa there. Maybe we should
  say something about IMonads in general before we get to this point?}
%
\[\stk{
\begin{array}{ll}
\FreeIMonad{\_}{} : & _{\PI{\V{I}}{\Set}}
                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
                      \TagIDesc{\V{I}}
\end{array} \\
\FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto
    \pair{\pair{\ListCons{\SYMBDVar}{(\fst{\V{E}})}} 
               {\LAM{\V{i}}
                \pair{\DConst{(\V{R}\: \V{i})}}
                     {(\snd{\V{E}})\: \V{i}}{}}{}}
         {\V{F}}{}
}\]

\newcommand{\substI}{\F{substI}}

Just as in the universe of descriptions, this construction comes with
an obvious \return and a substitution operation, the \bind. Its
definition is the following:
%
\[\stk{
\begin{array}{@{}ll}
\substI : & _{\PI{\V{I}}{\Set}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
          & (\V{X} \DotTo 
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
\end{array} \\
\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
\qquad    \cataI_{\V{I}}\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
                      (\muide{\V{Y}}{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
                      \V{i}\:
                      \V{t} 
}\]
% 
where  $\F{applyI}$ is defined as follows:
%
\[\stk{
\begin{array}{@{}ll}
\F{applyI} : & _{\PI{\V{I}}{\Set}}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
          & (\V{X} \DotTo \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
                                         {\V{I}}
                                         {\V{X}}}}
                    {\V{I}}
                    {\muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
            \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
\end{array}
}\]

The subscripted types corresponds to implicit arguments that can be
automatically inferred, hence do not have to be typed in. Let us now
consider two examples of free indexed monads.

\paragraph{Typed expressions:}

\begin{wstructure}
    /> Closed term
        <- only constants and operations on them
        -> Extend Val with Var : Ty -> Set, mapping to EnumU
            -> Open term
            -> Language of well-typed terms
                <- By construction
\end{wstructure}

\newcommand{\Ctxt}{\D{Context}}
\newcommand{\SYMBCtxtEmpty}{\C{[]}\xspace}
\newcommand{\CtxtEmpty}{\SYMBCtxtEmpty}
\newcommand{\SYMBCtxtSnoc}{\C{snoc}\xspace}
\newcommand{\CtxtSnoc}[2]{\SYMBCtxtSnoc\:#1\:#2}
\newcommand{\SYMBEnv}{\F{Env}}
\newcommand{\Env}[1]{\SYMBEnv\: #1}
\newcommand{\SYMBlookup}{\F{lookup}}
\newcommand{\lookup}[4]{\SYMBlookup\: #1\: #2\: #3\: #4}

In the previous section, we presented a language of closed
arithmetic expressions. Using the free monad construction, we are
going to extend this construction to open terms. An open term is
defined with respect to a context, represented by a snoc-list of
types:
%
\[
\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
\Ctxt           & \Set \\
\SYMBCtxtEmpty  & \Ctxt \\
\SYMBCtxtSnoc   & \Ctxt \To \Ty \To \Ctxt
\end{array}
\]
%
An environment realises the context, packing a value for each type:
%
\[
\stk{
\SYMBEnv : \Ctxt \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Env{\CtxtEmpty}                & \Unit \\
\Env{(\CtxtSnoc{\V{G}}{\V{S}})} & \TIMES{\Env{\V{G}}}{\Val{\V{S}}}
\end{array}
}\]
%
In this setting, we define type variables, $\SYMBVar$ by:
%
\[\stk{
\Var{}{} : \Ctxt \To \Ty \To \Set \\
\begin{array}{@{}ll@{\:\:\mapsto\:\:}l}
\Var{\CtxtEmpty}{& \V{T}}                & 
    \Void \\
\Var{(\CtxtSnoc{\V{G}}{\V{S}})}{& \V{T}} & 
    \SUM{(\Var{\V{G}}{\V{T}})}{(\V{S} \PropEq \V{T})}
\end{array}
}\]
%
While $\SYMBVal$ maps the type to the corresponding host type,
$\SYMBVar$ indexes a value in the context, obtaining a proof that the
types match. The $\SYMBlookup$ function precisely follow this
semantics:
%
\[\stk{
\SYMBlookup : \PI{\V{G}}{\Ctxt} 
          \Env{\V{G}} \To 
          \PI{\V{T}}{\Ty} 
          \Var{\V{G}}{\V{T}} \To
          \Val{\V{T}} \\
\begin{array}{@{}l@{}l@{}l@{}l@{}lll}
\lookup{& (\CtxtSnoc{\V{G}}{.T})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumRight{\C{refl}})} & \mapsto & \V{t} \\
\lookup{& (\CtxtSnoc{\V{G}}{\V{S}})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumLeft{\V{x}})} & \mapsto & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
\end{array}
}\]

\newcommand{\SYMBEmpty}{\F{Empty}\xspace}
\newcommand{\Empty}[1]{\SYMBEmpty\:#1}

\newcommand{\SYMBopenTerm}{\F{openTm}\xspace}
\newcommand{\openTerm}[1]{\SYMBopenTerm\: #1}
\newcommand{\closeTerm}{\F{closeTm}}

\newcommand{\update}{\F{update}}

Consequently, taking the free monad of \(\HExprD\) by \(\SYMBVar\:
\V{G}\), we obtain the language of open terms in a context \(\V{G}\):
%
\[
\openTerm{\V{G}} \mapsto \FreeIMonad{\HExprD}{\Ty}{(\SYMBVar\:\V{G})}
\]
%
In this setting, the language of closed terms corresponds to the free
monad assigning an empty set of values to variables
%
\[
\closeTerm \mapsto \FreeIMonad{\HExprD}{\Ty}{\SYMBEmpty}
\quad
\mbox{where}
\quad
\stk{
\SYMBEmpty : \Ty \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Empty{\_}   & \Zero \\
\end{array}
}\]
%
Allowing variables from an empty set is much like forbidding variables,
so \(\closeTerm\) and \(\HExprD\) describe isomorphic
datatypes. Correspondingly, you can update an old \(\HExprD\) to a shiny
\(\closeTerm\):
%
\[\stk{
\update : \muide{\Ty}{\HExprD} \DotTo \muide{\Ty}{\closeTerm} \\
\begin{array}{@{}l@{}l}
\update\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty}\: & (\toIDesc{\HExprD})\:
                                                  (\muide{\Ty}{\closeTerm})\: \\
                                                & (\LAM{\_} \LAM{\pair{\V{tag}}{\V{tm}}{}} \Con{\pair{\Su{\V{tag}}}{\V{tm}}{}})\:
                                                  \V{ty}\:
                                                  \V{tm}
\end{array}
}\]
% 
The other direction of the isomorphism is straightforward, the
$\SYMBDVar$ case being impossible. Therefore, we are entitled to
reuse the $\evalH$ function to define the semantics of $\closeTerm$.

\newcommand{\discharge}{\F{discharge}}

Now we would like to give a semantics to the open term language. We
proceed in two steps: first, we substitute variables by their value in
the context; then, we evaluate the resulting closed term. Thanks to
$\evalH$, the second problem is already solved. Let us focus on
substituting variables from the context. Again, we can subdivide this
problem: first, discharging a single variable from the context; then,
applying this $\discharge$ function on every variables in the term.

The $\discharge$ function is relative to the required type and a
context of the right type. Its action is to map values to themselves,
and variables to their value in context. This corresponds to the
following function:
%
\[\stk{
\begin{array}{@{}ll}
\discharge : & \PI{\V{G}}{\Ctxt}
               \Env{\V{G}} \To 
               \Var{\V{G}}{} \DotTo
               \muide{\Ty}{\closeTerm}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{G}\: \V{g}\: \V{ty}\: \V{v} &
    \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
\end{array}
}\]

\begin{wstructure}
            /> Then, perform subst everywhere in the term
                -> Show type [code]
                /> This is a bind!?
                -> There is some more structure 
                    -> We should try to get it
\end{wstructure}

\newcommand{\substH}{\F{substExpr}}

We are now left with applying $\discharge$ over all variables of the
term.  We simply have to fill in the right arguments to $\substI$, the
type guiding us:
%
\[
\stk{
\begin{array}{@{}ll}
\substH  : & \PI{\V{G}}{\Ctxt} \\
           & (\Var{\V{G}}{} \DotTo
              \muide{\Ty}{\closeTerm}) \DotTo \\
          & \muide{\Ty}{(\openTerm{\V{G}})} \DotTo 
            \muide{\Ty}{\closeTerm}
\end{array} \\
\substH\: \V{G}\:
          \V{ty}\:          
          \V{g}\:
          \V{\sigma}\: 
          \V{tm} \mapsto  
\substI_{\Ty}\:
               (\SYMBVar\: \V{G})\: 
               \SYMBEmpty\:
               \HExprD\: 
               \V{\sigma}\:
               \V{ty}\:
               \V{tm}
}\]

Hence completing our implementation of the open terms
interpreter. Without much effort, we have described the syntax of a
well-typed language, together with its semantics.

\paragraph{Indexed descriptions:}

An interesting instance of free monad is $\SYMBIDesc$ itself. Indeed,
$\SYMBDVar$ is nothing but the \return. The remaining constructors form
the carrier functor, trivially indexed by $\Unit$. The signature functor
is described as follow:
%
\[\stk{
\IDescFreeD : \ATagIDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD \mapsto \bigRedBracket{\begin{array}{l}
                                \sqr{\SYMBDConst\:\:
                                     \DProd\:\:
                                     \SYMBDSigma\:\:
                                     \SYMBDPi} \red{,}\\
                                  \LAM{\_}\bigRedBracket{\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}}\end{array}}
\end{array}
}\]
%
We get $\IDesc{\V{I}}$ by extending the signature with variables from \(\V{I}\):
%
\[\stk{
\IDescD : \PI{\V{I}}{\Set} \TagIDesc{\Unit} \\
\IDescD\: \V{I} \mapsto \FreeIMonad{\red{[}\IDescFreeD\red{,[}\LAM{\_}\sqr{}\red{,}\LAM{\_}\red{[]]]}}{\Unit}\LAM{\_}\V{I}
}\]

The fact that indexed descriptions are closed under substitution
is potentially of considerable utility, if we can exploit this fact:
\[
\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
    \mapsto 
        \idescop{\V{D}}
                {\V{I}}
                {\LAM{\V{i}}
                     {\idescop{\V{\sigma} \V{i}}
                              {\V{J}}
                              {\V{X}}}}
        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
\]
By observing that a description can be decomposed via substitution, we
split its meaning into a superstructure of substructures, e.g. a
`database containing salaries', ready for traversal operations
preserving the former and targeting the latter.

%\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Discussion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discussion}

In this paper, we have presented a universe of datatypes for a
dependent type theory. We started from an unremarkable type theory
with dependent functions and tuples, but relying on few other
assumptions, especially where propositional equality is concerned.  We
added finite enumeration sufficient to account for constructor choice,
and then we built coding systems, first (as a learning experience) for
simple ML-like inductive types, then for the indexed inductive
families which dependently typed programmers in Agda, Coq and Epigram
take for granted. We adopt a bidirectional type propagation mechanism
to conceal artifacts of the encoding, giving a familiar and practicable
constructor-based presentation to data.

Crucially to our approach, we ensure that the codes describing
datatypes inhabit a datatype with a code. In a stratified setting, we
avoid paradox by ensuring that this type of codes lives uniformly one
level above the types the codes describe. The adoption of ordinary
data to describe types admits datatype-generic operations implemented
just by ordinary programming. In working this way, we make
considerable use of type equality modulo open computation, silently
specialising the types of generic operations as far as the datatype
code for any given usage is known.

\subsection{Related work in Generic Programming}

\begin{wstructure}
!!! Need Help !!!
<- Comparison with Induction Recursion
    ???
\end{wstructure}

\begin{wstructure}
!!! Need Help !!!
<- Related Work
    <- Generic in simply-typed functional languages
        <- PolyP \cite{jansson:polyp}
        <- Generic Haskell \cite{hinze:generic-haskell}
        <- Scratch your boilerplate \cite{spj:syb}
\end{wstructure}

Generic programming is a vast topic. We refer our reader to
\citet{garcia:generic-comparative-study} for a broad overview of
generic programming in various languages. For Haskell alone, there is
a myriad of proposals: 
\citet{hinze:generic-approach-comparative} and
\citet{rodriguez:generic-libs-comparative} provide useful comparative
surveys.

Our approach follows the polytypic programming style, as initiated
by PolyP~\cite{jansson:polyp}. Indeed, we build generic functions by
induction on pattern functors, exploiting type-level computation
to avoid the preprocessing phase: our
datatypes are, natively, nothing but codes.

We have the \emph{type-indexed datatypes} of Generic
Haskell~\cite{hinze:generic-haskell} for free.
From one datatype, we can compute
others and equip them with relevant structure: the free monad
construction provides one example. Our approach to encoding datatypes
as data also sustains \emph{generic
  views}~\cite{holdermans:generic-view}, allowing us to rebias the
presentation of datatypes conveniently. Tagged descriptions,
giving us a sum-of-sigmas view, are a natural example.

Unlike Generic Haskell, we do not support polykinded
programming~\cite{hinze:polytypic-polykinded}. Our descriptions are
limited to endofunctors on $\Set^I$. Whilst indexing is known
to be sufficient to \emph{encode} a large class of higher-kinded
datatypes~\citep{DBLP:conf/ifip2-1/AltenkirchM02}, we should rather
hope to work in a more compositional style. We are free to write
higher-order programs manipulating codes, but is not yet clear whether
that is sufficient to deliver abstraction at higher kinds.
Similarly, it will be interesting to see whether
arity-generic programming~\cite{weirich:arity-generic} arises
just by computing with our codes, or whether a richer abstraction is
called for.

The Scrap Your Boilerplate~\cite{spj:syb} (SYB) approach to generic
programming offers a way to construct generic functions, based on
dynamic type-testing via the $\CN{Typeable}$
type class. SYB cannot compute types from codes, but its dynamic character
does allow a more flexible \emph{ad hoc} approach to generic data
traversal. By maintaining the correspondence between codes and types
whilst supporting arbitrary inspection of codes, we pursue the same
flexibility statically. The substitutive character of $\SYMBIDesc$
may allow us to observe and exploit \emph{ad hoc} substructural
relationships in data, but again, further work is needed if we
are to make a proper comparison.

\begin{wstructure}
    <- Generic in dependent types
        <- Norell \cite{norell:msc-thesis}
        <- Polytypic prog in Coq \cite{verbruggen:polytype-coq}
        <- Universes for generic prog \cite{benke:universe-generic-prog}
\end{wstructure}

\subsection{Generic Programming with Dependent Types}

Generic programming is not new to dependent
types. \citet{DBLP:conf/ifip2-1/AltenkirchM02} developed a universe
of polykinded types in Lego;
\citet{norell:msc-thesis} gave a formalisation of
polytypic programming in Alfa, a precursor to Agda;
\citet{verbruggen:polytype-prog-coq, verbruggen:polytype-coq} provided
a framework for polytypic programming in the Coq theorem
prover. However, these works aim at \emph{modelling} PolyP or Generic
Haskell in a dependently-typed setting for the purpose of proving
correctness properties of Haskell code. Our approach is different in
that we aim at building a foundation for datatypes, in a
dependently-typed system, for a dependently-typed system.

Closer to us is the work of \citet{benke:universe-generic-prog}. This
seminal work introduced the usage of universes for developing generic
programs. Our universes share similarities to theirs: our universe of
descriptions is similar to their universe of iterated induction, and
our universe of indexed descriptions is equivalent to their universe
of finitary indexed induction. This is not surprising, as we share the
same source of inspiration, namely induction-recursion.

However, we feel ready to offer a more radical prospectus. Their
approach is generative: each universe extends the base type theory
with both type formers and elimination rules. Thanks to levitation, we
rely only on a generic induction and a specialised $\SYMBswitchD$,
closing the type theory. We explore \emph{programming} with codes, but
also how to conceal the encoding when writing `ordinary' programs.

\subsection{Metatheoretical Status}

The \(\Set:\Set\) approach we have taken in this paper is convenient
from an experimental perspective, and it has allowed us to focus
primarily on the encoding of universes, leaving the question
of stratification (and with it, consistency, totality,
and decidability of type checking) to one side. However, we must
surely face up to the latter, especially since we have taken up the
habit of constructing `the set of all sets'. A proper account requires
a concrete proposal for a system of stratified universes which allows
us to make `level-polymorphic' constructions, and we are actively
pursuing such a proposal. We hope soon to have something to prove.

In the meantime, we can gain some confidence by systematically
embedding predicative fragments of our theory within
systems which already offer a universe hierarchy. We can, at the
very least, confirm that in UTT-style theories with conventional
inductive families of types~\citep{luo:utt}, as found in Coq
(and in Agda if one avoids experimental extensions), we build the
tower of universes we propose, cut off at an arbitrary height.
It is correspondingly clear that some such system can be made to
work, or else that other, longer-standing tools are troubled.

A metatheoretical issue open at time of writing concerns the
size of the \emph{index} set \(\V{I}\) in \(\IDesc{\V{I}}\). Both Agda
and recent versions of Coq allow inductive families with \emph{large}
indices, effectively allowing `higher-kind' fixpoints on
\(\Set^{\Set}\) and more. They retain the safeguard that the types of
\emph{substructures} must be as small as the inductively
defined superstructure. This liberalisation allows us large index sets
in our models, but whilst it offers no obvious route to paradox by
smuggling a large universe inside a small type, it is not yet known
to be safe. We can restrict \(\V{I}\) as necessary to avoid
paradox, provided \(\Unit\), used to index \(\SYMBIDesc\) itself, is
`small'.

\subsection{Further Work}

Apart from the need to nail down a stratified version of the system and
its metatheory, we face plenty of further problems and opportunities.
Although we have certainly covered Luo's criteria for inductive
families~\cite{luo:utt}, there are several dimensions in which to
consider expanding our universe.

Firstly, we seek to encompass
\emph{inductive-recursive} datatype families~\cite{dybjer:iir}, allowing
us to interleave the definition and interpretation of data in intricate
and powerful ways. This interleaving seems particularly useful when
reflecting the syntax of dependent type systems.

Secondly, we should very much like to extend our
universe with a codes for internal fixpoints, as
in~\cite{DBLP:conf/types/MorrisAM04}. The external knot-tying
approach we have taken here makes types like `trees with lists of subtrees'
more trouble than they should be. Moreover, if we allow the alternation
of least and greatest fixpoints, we should expect to gain types which are
not readily encoded with one external \(\SYMBIMu\).

Thirdly, it would be fascinating to extend our universe with dedicated
support for syntax with \emph{binding}, not least because a universe
with internal fixpoints has such a syntax. Harper and Licata have
demonstrated the potential for and of such an
encoding~\citep{DBLP:conf/icfp/LicataH09}, boldly encoding the invalid
definitions along with the valid. A more conservative strategy might
be to offer improved support for datatypes indexed by an extensible
context of free variables, with the associated free monad structure
avoiding capture as shown by~\citet{altenkirch:monadic-lambda}.

Lastly, we must ask how our new presentation of datatypes
should affect the tools we use to build software. It is not enough to
change the game: we must enable better play. If datatypes are data,
what is design?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Acknowledgments
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\acks

We are grateful to Jos\'{e} Pedro Magalh\~{a}es for his helpful
comments on a draft of this paper. We are also grateful to the Agda
team, without which levitation would have been a much more perilous
exercise. J. Chapman was supported by the Estonian Centre of
Excellence in Computer Science, EXCS, financed by the European
Regional Development Fund. P.-\'{E}. Dagand, C. McBride and P. Morris
are supported by the Engineering and Physical Sciences Research
Council, Grants EP/G034699/1 and EP/G034109/1.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{thebibliography}{47}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Abel et~al.()Abel, Coquand, and Pagano]{DBLP:conf/tlca/AbelCP09}
A.~Abel, T.~Coquand, and M.~Pagano.
\newblock A modular type-checking algorithm for type theory with singleton
  types and proof irrelevance.
\newblock In \emph{TLCA}.

\bibitem[Adams(2006)]{DBLP:journals/jfp/Adams06}
R.~Adams.
\newblock Pure type systems with judgemental equality.
\newblock \emph{JFP}, 2006.

\bibitem[Altenkirch and McBride(2002)]{DBLP:conf/ifip2-1/AltenkirchM02}
T.~Altenkirch and C.~McBride.
\newblock Generic programming within dependently typed programming.
\newblock In \emph{Generic Programming}, 2002.

\bibitem[Altenkirch and Reus(1999)]{altenkirch:monadic-lambda}
T.~Altenkirch and B.~Reus.
\newblock Monadic presentations of lambda terms using generalized inductive
  types.
\newblock In \emph{Computer Science Logic}. 1999.

\bibitem[Altenkirch et~al.(2007)Altenkirch, McBride, and
  Swierstra]{altenkirch:ott}
T.~Altenkirch, C.~McBride, and W.~Swierstra.
\newblock Observational equality, now!
\newblock In \emph{PLPV}, 2007.

\bibitem[Augustsson and
  Carlsson(1999)]{augustsson.carlsson:dependent.interpreter}
L.~Augustsson and M.~Carlsson.
\newblock An exercise in dependent types: A well-typed interpreter.
\newblock {Available at
  \url{http://www.cs.chalmers.se/~augustss/cayenne/interp.ps}}, 1999.

\bibitem[Benke et~al.(2003)Benke, Dybjer, and
  Jansson]{benke:universe-generic-prog}
M.~Benke, P.~Dybjer, and P.~Jansson.
\newblock Universes for generic programs and proofs in dependent type theory.
\newblock \emph{Nordic Journal of Computing}, 2003.

\bibitem[Brady et~al.()Brady, Chapman, Dagand, Gundry, McBride, Morris, and
  Norell]{pigs:epigram}
E.~Brady, J.~Chapman, P.-E. Dagand, A.~Gundry, C.~McBride, P.~Morris, and
  U.~Norell.
\newblock An {E}pigram implementation.

\bibitem[Brady et~al.(2003)Brady, McBride, and
  McKinna]{brady:index-inductive-families}
E.~Brady, C.~McBride, and J.~McKinna.
\newblock Inductive families need not store their indices.
\newblock In \emph{TYPES}, 2003.

\bibitem[Cheney and Hinze(2003)]{cheney:gadt}
J.~Cheney and R.~Hinze.
\newblock First-class phantom types.
\newblock Technical report, Cornell University, 2003.

\bibitem[Coquand(1996)]{DBLP:journals/scp/Coquand96}
T.~Coquand.
\newblock An algorithm for type-checking dependent types.
\newblock \emph{SCP}, 1996.

\bibitem[Courant(2002)]{courant:explicit-universe}
J.~Courant.
\newblock Explicit universes for the calculus of constructions.
\newblock In \emph{TPHOLs}, 2002.

\bibitem[Danielsson(2010)]{nisse:asl}
N.~A. Danielsson.
\newblock The {A}gda standard library, 2010.

\bibitem[Dybjer(1991)]{dybjer:families}
P.~Dybjer.
\newblock Inductive sets and families in {M}artin-{L}{\"o}f's type theory.
\newblock In \emph{Logical {F}rameworks}. 1991.

\bibitem[Dybjer and Setzer(1999)]{dybjer:axiom-ir}
P.~Dybjer and A.~Setzer.
\newblock A finite axiomatization of inductive-recursive definitions.
\newblock In \emph{TLCA}, 1999.

\bibitem[Dybjer and Setzer(2000)]{dybjer:ir-initial-algebra}
P.~Dybjer and A.~Setzer.
\newblock Induction-recursion and initial algebras.
\newblock In \emph{Annals of Pure and Applied Logic}, 2000.

\bibitem[Dybjer and Setzer(2001)]{dybjer:iir}
P.~Dybjer and A.~Setzer.
\newblock Indexed induction-recursion.
\newblock In \emph{Proof Theory in Computer Science}. 2001.

\bibitem[Garcia et~al.(2003)Garcia, Jarvi, Lumsdaine, Siek, and
  Willcock]{garcia:generic-comparative-study}
R.~Garcia, J.~Jarvi, A.~Lumsdaine, J.~Siek, and J.~Willcock.
\newblock A comparative study of language support for generic programming.
\newblock In \emph{OOPSLA}, 2003.

\bibitem[Geuvers(2001)]{geuvers:induction-not-derivable}
H.~Geuvers.
\newblock Induction is not derivable in second order dependent type theory.
\newblock In \emph{TLCA}, 2001.

\bibitem[Goguen et~al.(2006)Goguen, McBride, and
  McKinna]{goguen:pattern-matching}
H.~Goguen, C.~McBride, and J.~McKinna.
\newblock Eliminating dependent pattern matching.
\newblock In \emph{Algebra, Meaning and Computation}. 2006.

\bibitem[Harper and Pollack()]{harper:implicit-universe}
R.~Harper and R.~Pollack.
\newblock Type checking with universes.
\newblock In \emph{TAPSOFT'89}.

\bibitem[Hinze(2000)]{hinze:polytypic-polykinded}
R.~Hinze.
\newblock Polytypic values possess polykinded types.
\newblock In \emph{MPC}. 2000.

\bibitem[Hinze et~al.(2002)Hinze, Jeuring, and L\"{o}h]{hinze:generic-haskell}
R.~Hinze, J.~Jeuring, and A.~L\"{o}h.
\newblock Type-indexed data types.
\newblock In \emph{MPC}, 2002.

\bibitem[Hinze et~al.(2007)Hinze, Jeuring, and
  L\"{o}h]{hinze:generic-approach-comparative}
R.~Hinze, J.~Jeuring, and A.~L\"{o}h.
\newblock Comparing approaches to generic programming in {H}askell.
\newblock In \emph{Datatype-Generic Programming}. 2007.

\bibitem[Holdermans et~al.(2006)Holdermans, Jeuring, L\"{o}h, and
  Rodriguez]{holdermans:generic-view}
S.~Holdermans, J.~Jeuring, A.~L\"{o}h, and A.~Rodriguez.
\newblock Generic views on data types.
\newblock In \emph{MPC}. 2006.

\bibitem[Jansson and Jeuring(1997)]{jansson:polyp}
P.~Jansson and J.~Jeuring.
\newblock Poly{P}---a polytypic programming language extension.
\newblock In \emph{POPL}, 1997.

\bibitem[L{\"a}mmel and {Peyton Jones}(2003)]{spj:syb}
R.~L{\"a}mmel and S.~{Peyton Jones}.
\newblock Scrap your boilerplate: a practical design pattern for generic
  programming.
\newblock In \emph{TLDI}, 2003.

\bibitem[Licata and Harper(2009)]{DBLP:conf/icfp/LicataH09}
D.~R. Licata and R.~Harper.
\newblock A universe of binding and computation.
\newblock In \emph{ICFP}, 2009.

\bibitem[Luo(1994)]{luo:utt}
Z.~Luo.
\newblock \emph{Computation and Reasoning}.
\newblock Oxford University Press, 1994.

\bibitem[Martin-L{{\"o}f}(1984)]{martin-lof:itt}
P.~Martin-L{{\"o}f}.
\newblock \emph{Intuitionistic Type Theory}.
\newblock Bibliopolis$\cdot$Napoli, 1984.

\bibitem[McBride and McKinna(2004)]{mcbride.mckinna:view-from-the-left}
C.~McBride and J.~McKinna.
\newblock The view from the left.
\newblock \emph{JFP}, 2004.

\bibitem[Morris(2007)]{morris:PhD}
P.~Morris.
\newblock \emph{Constructing Universes for Generic Programming}.
\newblock PhD thesis, University of Nottingham, 2007.

\bibitem[Morris and Altenkirch(2009)]{alti:lics09}
P.~Morris and T.~Altenkirch.
\newblock Indexed containers.
\newblock In \emph{LICS}, 2009.

\bibitem[Morris et~al.(2004)Morris, Altenkirch, and
  McBride]{DBLP:conf/types/MorrisAM04}
P.~Morris, T.~Altenkirch, and C.~McBride.
\newblock Exploring the regular tree types.
\newblock In \emph{TYPES}, 2004.

\bibitem[Morris et~al.(2009)Morris, Altenkirch, and Ghani]{morris:spf}
P.~Morris, T.~Altenkirch, and N.~Ghani.
\newblock A universe of strictly positive families.
\newblock \emph{IJCS}, 2009.

\bibitem[Norell(2002)]{norell:msc-thesis}
U.~Norell.
\newblock Functional generic programming and type theory.
\newblock Master's thesis, Chalmers University of Technology, 2002.

\bibitem[Norell(2007)]{norell:agda}
U.~Norell.
\newblock \emph{Towards a practical programming language based on dependent
  type theory}.
\newblock PhD thesis, Chalmers University of Technology, 2007.

\bibitem[Oury and Swierstra(2008)]{oury:power-of-pi}
N.~Oury and W.~Swierstra.
\newblock The power of {P}i.
\newblock In \emph{ICFP}, 2008.

\bibitem[Paulin-Mohring(1996)]{paulin:habilitation}
C.~Paulin-Mohring.
\newblock \emph{D{\'e}finitions inductives en th{\'e}orie des types d'ordre
  sup{\'e}rieur}.
\newblock th{\`e}se d'habilitation, ENS Lyon, 1996.

\bibitem[Pierce and Turner(1998)]{pierce:bidirectional-tc}
B.~C. Pierce and D.~N. Turner.
\newblock Local type inference.
\newblock In \emph{POPL}, 1998.

\bibitem[Rodriguez et~al.(2008)Rodriguez, Jeuring, Jansson, Gerdes, Kiselyov,
  and d.~S.~Oliveira]{rodriguez:generic-libs-comparative}
A.~Rodriguez, J.~Jeuring, P.~Jansson, A.~Gerdes, O.~Kiselyov, and B.~C.
  d.~S.~Oliveira.
\newblock Comparing libraries for generic programming in {H}askell.
\newblock In \emph{Haskell Symposium}, 2008.

\bibitem[{The Coq Development Team}()]{coq}
{The Coq Development Team}.
\newblock \emph{The Coq Proof Assistant Reference Manual}.

\bibitem[Verbruggen et~al.(2008)Verbruggen, de~Vries, and
  Hughes]{verbruggen:polytype-prog-coq}
W.~Verbruggen, E.~de~Vries, and A.~Hughes.
\newblock Polytypic programming in {C}oq.
\newblock In \emph{WGP}, 2008.

\bibitem[Verbruggen et~al.(2009)Verbruggen, de~Vries, and
  Hughes]{verbruggen:polytype-coq}
W.~Verbruggen, E.~de~Vries, and A.~Hughes.
\newblock Polytypic properties and proofs in {C}oq.
\newblock In \emph{WGP}, 2009.

\bibitem[Weirich and Casinghino(2010)]{weirich:arity-generic}
S.~Weirich and C.~Casinghino.
\newblock Arity-generic datatype-generic programming.
\newblock In \emph{PLPV}, 2010.

\bibitem[Xi et~al.(2003)Xi, Chen, and Chen]{xi:gadt}
H.~Xi, C.~Chen, and G.~Chen.
\newblock Guarded recursive datatype constructors.
\newblock In \emph{POPL}, 2003.

\bibitem[Yakushev et~al.(2009)Yakushev, Holdermans, L\"{o}h, and
  Jeuring]{yakushev:mutual-def}
A.~R. Yakushev, S.~Holdermans, A.~L\"{o}h, and J.~Jeuring.
\newblock Generic programming with fixed points for mutually recursive
  datatypes.
\newblock In \emph{ICFP}, 2009.

\end{thebibliography}

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.
%\begin{thebibliography}{}
%\softraggedright
%\end{thebibliography}

\end{document}

### 2010-08-26 notes/Epigram2/Epigram2.tex
design: yet another barely started document Now that we have quite a lot of the machinery needed to elaborate programs and proofs, it's time we had a language to elaborate.
I've started a document (not for the first time) whose purpose
is to collect ideas about the high level language that Epigram
2 will be. Opportunities for Wadler's Law abound! And indeed,
the lexical structure of comments is likely to prove a matter of
considerable fascination.

\documentclass{article}
\usepackage{palatino}
\usepackage{a4}

\begin{document}

\title{Epigram 2\\
       more design thoughts}
\author{Conor McBride}
\maketitle

\section{Introduction}

With the underlying type theory beginning to settle, it's time to
think a bit more about how the high-level Epigram language might
work. This document contains some ideas, and delineates some gaps.

I shall write mainly in the present tense, intending that the
assertions I make become true at some point.

\section{The Coauthor Transducer Model}

As an experience, Epigram programming is interactive. As a document,
an Epigram program is the `minute' of an interaction. To load and
check an Epigram program is to replay the minuted interaction,
checking its plausibility.

Epigram operates as a \textbf{transducer}, consuming a source
file, and producing a modified source file, marked up with error
reports and responses to requests for help. The transducer is
idempotent. Epigram thus acts like a mechanical coauthor, offering
opinions and responding to requests for contribution.

Epigram has a high-level \textbf{source} language, in which
\texttt{.epi} documents are written, and a low-level \textbf{evidence}
language, into which documents are \textbf{elaborated}. The system
thus works by constructing a sequence of definitions in the evidence
language elaborating the definitions found in its input. The
elaboration process is, by its nature, incremental. The system
maintains the correspondence between regions of the source document
and definitions at the evidence layer: the output file may thus be
generated from the final \textbf{proof state} of the elaboration
process.

In addition to generating the output file, the system will (on
request) dump the final proof state in a textual format---a
\texttt{.gram} file. From a \texttt{.gram} file, both input and output
\texttt{.epi} files are recoverable, hence the patch from one to the
other is computable. Given a \texttt{.gram} file and
a patch to its output \texttt{.epi}, a new input \texttt{.epi} can be
constructed, which can be processed in turn, resulting in a new output
\texttt{.epi}, and hence a further patch. Editor integration is thus
unnecessary, but in any case easy.

A separate (or at least separable) component, based on Edwin Brady's
\textsc{epic} compiler, can generate a \texttt{.gram} file from an
executable. The latter, if executed, offers a straightforward
read-eval-print loop for first-order expressions constructed over the
signature of definitions provided by the original source. Incomplete
definitions in the source may result in run-time errors: these are the
only run-time errors.  Such an executable, invoked with an argument,
treats that argument as a single command session.

Markdown is king.

In the event that Epigram programs ever become large enough to
necessitate modular development, \texttt{.gram} files will be generated
on a per-module basis. We may thus become interested in patches
between states of these \texttt{.gram} files as a means to transmit
changes downstream.

\section{Interactive Problem Solving}

Epigram documents show the declaration and partial solution of problems.
Problem solving is mediated as an interactive process. Every problem
in the source language can be encoded as a type in the evidence
language. Solving a problem in the source language amounts to finding
an inhabitant of the type which encodes it, by a process of
hierarchical refinement.

A \textbf{document} is a series of \textbf{developments}. A
development comprises a \textbf{declaration} introducing a problem,
and a \textbf{refinement} reducing the problem to a hopefully
empty collection of subproblems. Too much abstraction: example, please!
\newcommand{\capbox}[2]{\raisebox{0.12in}{\begin{array}[t]{|l|}\hline
      #1 \vspace*{-0.05in}\\
    \hfill\mbox{\scriptsize{#2}} \\ \hline\end{array}}}
\newcommand{\Ts}[1]{\texttt{#1}\;}
\newcommand{\co}{\;:\;}
\newcommand{\sco}{;\;\;}
\newcommand{\cm}{,\;}
\[
\capbox{
  \Ts{let}
    \capbox{
      \capbox{\Ts{x} \cm \Ts{y}\co \Ts{Nat}}{declaration}\\
      \Ts{--------------}\\
      \capbox{\Ts{x} \Ts{+} \Ts{y}}{template} \co \capbox{\Ts{Nat}}{type}
      }
      {declaration}\\ \\
  \capbox{
     \capbox{\Ts{x}\Ts{+}\Ts{y}}{problem}
       \capbox{\Ts{<=}\capbox{\Ts{induction}\Ts{x}}{eliminator}}{tactic}\\
     \capbox{
      \{\;\capbox{\capbox{\Ts{'zero}\Ts{+}\Ts{y}}{problem}
                  \capbox{\Ts{=} \capbox{\Ts{y}}{term}}{tactic}
                  \capbox{}{block}
          }{refinement}\\
      ;\;\capbox{\capbox{\Ts{'suc}\Ts{x}\Ts{+}\Ts{y}}{problem}
                 \capbox{\Ts{=} \capbox{\Ts{'suc}
                   \capbox{\{!\;!\}}{shed}}{term}}{tactic}
                  \capbox{}{block}
         }{refinement}\\
      \}
      }{block}
    }
    {refinement}}
    {development}
\]

\end{document}

### 2010-08-31 notes/Epigram2/Epigram2.tex
notes: fixed reversal, finished description of example Those of you who guessed that I had a lapse of concentration when I wrote that you could extract a .gram from an executable can pat yourselves on the back. I meant the other way around and have repaired the damage. I've also added some superficial description of the example program. Tighter definitions are yet to come. The purpose of this section is to give an impression of the interaction.

@@
 A separate (or at least separable) component, based on Edwin Brady's
-\textsc{epic} compiler, can generate a \texttt{.gram} file from an
-executable. The latter, if executed, offers a straightforward
+\textsc{epic} compiler, can generate an executable from a \texttt{.gram}
+file. The former, if executed, offers a straightforward
 read-eval-print loop for first-order expressions constructed over the
@@
           }{refinement}\\
-      ;\;\capbox{\capbox{\Ts{'suc}\Ts{x}\Ts{+}\Ts{y}}{problem}
+      ;\;\capbox{\capbox{\Ts{'suc}\Ts{x'}\Ts{+}\Ts{y}}{problem}
                  \capbox{\Ts{=} \capbox{\Ts{'suc}
@@
 
+Lots of points show up from this example. I'll skim through them,
+starting with the declaration: we give a template for addition,
+hypothesizing declarations for the fields of the template, then giving
+a type for the whole thing. This template determines the initial
+problem. You can tell which tokens in the template are punctuation
+rather than placeholders, because they don't have declarations!
+Note that here, we can see value declarations, but we shall also have
+propositional declarations, \(\texttt{:-}\;P\), where a proposition is
+asserted anonymously---\(P\) then gives the shapes of problems for
+refinement, i.e., propositional goals.
+
+Refinements go `\textbf{problem} \textbf{tactic} \textbf{block}'. A
+refinement is an offer to solve a problem of a particular form.
+Correspondingly, the elaborator needs to check that the refinement
+offered suits the problem or problems at hand: the rigidity of this
+check will soften with time, but the minimum acceptable level is that
+the refinement's problem freely determines the programmer's choice of
+names for the variables in the problem. Working interactively, of course,
+the problem parts of refinements are generated mechanically from the
+problems at hand, ensuring a match.
+
+A tactic provides the means to split a problem into zero or more
+subproblems. The \texttt{<=} (`by') tactic invokes an \textbf{eliminator}
+in the elimination-with-a-motive style. The \texttt{=} (`return') tactic
+takes a term explaining how to compute the return value, and leaves no
+subproblems. There will doubtless be others, not least the notorious
+\texttt{with}.
+
+The \textbf{shed}, \texttt{\{! !\}}, stands in place of a piece of program
+(most syntactic categories, not just terms) to indicate that the programmer
+has yet to choose something, thus causing elaboration to suspend. Note
+that a tactic \texttt{= \{! !\}} does commit to return, where a
+shed directly in a tactic place does not.
+
 \end{document}

### 2010-08-31 notes/Epigram2/Epigram2.tex
notes: stuff on interaction and where-clauses I tweaked the notation for sheds to be {. .} in order to allow {? ?} and {! !} for query and response. I've written a little about the uses of query and response in the coauthor transducer model.
I've added a bit about where-clauses, for helper-functions and
lemmas. None of this is terribly precise yet, and the notation
is very much up for grabs.

@@
 \usepackage{a4}
+\usepackage{alltt}
 
@@
 an inhabitant of the type which encodes it, by a process of
-hierarchical refinement.
+hierarchical refinement. This section is not supposed to be a definition:
+it's an explorations.
 
@@
 \capbox{
-  \Ts{let}
   \capbox{
@@
                  \capbox{\Ts{=} \capbox{\Ts{'suc}
-                   \capbox{\{!\;!\}}{shed}}{term}}{tactic}
+                   \capbox{\{.\;.\}}{shed}}{term}}{tactic}
                   \capbox{}{block}
@@
 
-The \textbf{shed}, \texttt{\{! !\}}, stands in place of a piece of program
+The \textbf{shed}, \texttt{\{. .\}}, stands in place of a piece of program
 (most syntactic categories, not just terms) to indicate that the programmer
 has yet to choose something, thus causing elaboration to suspend. Note
-that a tactic \texttt{= \{! !\}} does commit to return, where a
+that a tactic \texttt{= \{. .\}} does commit to return, where a
 shed directly in a tactic place does not.
 
+Now, in the coauthor transducer model, there is no trigger to fill in
+a shed: one simply replaces the shed with code and reloads. Of course,
+an editor could be customized to give this illusion, and the
+patch-to-patch optimization would yield very similar results. What,
+then, is the interaction?
+
+The basic refinement interaction simply involves the system replacing
+a null block with subproblems which are not accounted for.
+Correspondingly,
+
+\begin{alltt}
+ x , y : Nat
+-------------
+ x + y : Nat
+
+x + y <= induction x
+\end{alltt}
+%
+gets fleshed out to
+%
+\begin{alltt}
+let   x , y : Nat
+     -------------
+      x + y : Nat
+
+x + y <= induction x
+\{ 'zero + y    \{. .\}
+; 'suc x' + y  \{. .\} 
+\}
+\end{alltt}
+
+One might imagine going further, reporting coverage failure by inserting
+subproblems into nonempty blocks. Such a circumstance might readily
+arise if one added constructors to a datatype.
+
+\subsection{Query and Response}
+
+In addition to basic refinement functionality, we shall need to equip
+programmers with the means to query the system and receive responses
+in situ. Moreover, we shall need to equip the machine with the means
+to insert remarks in a file. We need anchored comments. The machine
+writes
+\[
+\Ts{\{!}\; \mathit{thing}\; \Ts{!!} \mathit{comment} \Ts{!\}}
+\]
+to indicate that the \(\mathit{comment}\) concerns the \(\mathit{thing}\).
+Such a textual pattern is replaced by \(\mathit{thing}\) on reloading
+(unless \(\mathit{thing}\) is blank, in which case the replacement is an
+empty shed). Machine remarks are ephemeral responses, and should not be
+put into the machine's mouth by the programmer. Meanwhile, an anchored query,
+\[
+\Ts{\{?}\; \mathit{thing}\; \Ts{??} \mathit{query} \Ts{?\}}
+\]
+means the same as \(\mathit{thing}\) (unless empty, in which case it means
+the same thing as an empty shed) but attaches a query; on return, we can
+expect a response in the form given above.
+
+Your basic empty \texttt{\{? ?\}} means
+``What goes here and what's in local scope?''. The basic use of
+\(\Ts{\{!}\; \mathit{thing}\; \Ts{!!} \mathit{comment} \Ts{!\}}\)
+is to report an error. A more amusing possibility would be to offer
+a menu of choices for \(\mathit{thing}\).
+
+It does not take a great leap of imagination to imagine how a fancier
+interface might mediate point-and-click queries via this mechanism.
+Again, one would need the patch-based transducer to support such a
+rapid dialogue.
+
+\subsection{Subproblems versus Side-Problems}
+
+The hierarchical refinement of problems delivers the basic necessary
+functionality, but it forces all sorts of auxiliary functions and lemmas
+to be developed globally, in advance (spatially, if not temporally).
+We might prefer to localize the development of helpers. I suggest that
+we allow each refinement to carry a \texttt{where}-clause bearing a local
+module of (partially) solved side-problems. Here's a simple example.
+
+\begin{alltt}
+ f : X -> Y; xs : List X
+-------------------------
+ map f xs : List Y
+
+map f xs = hit xs where
+\{  xs : List X
+  -----------------
+   hit xs : List X
+
+  hit xs <= induction xs
+  \{ hit 'nil = 'nil
+  ; hit ('cons x xs') zs = 'cons (f x) (hit xs') \}\}
+\end{alltt}
+
+Scope-wise, the \texttt{where}-clause comes after the problem statement
+but before tactic and block; textually, it can come before the tactic,
+between tactic and block, or after block.
+
+Note that on a good day with a following wind, the types of helper
+functions can be inferred. One might hope to write
+\begin{alltt}
+ f : X -> Y; xs : List X
+-------------------------
+ map f xs : List Y
+
+map f xs = \{? hit xs ?? helper ?\}
+\end{alltt}
+might deliver
+\begin{alltt}
+ f : X -> Y; xs : List X
+-------------------------
+ map f xs : List Y
+
+map f xs = hit xs where
+\{  xs : List X
+  -----------------
+   hit xs : List X
+
+  hit xs \{! !\} \}
+\end{alltt}
+
+on the basis that the type \texttt{List X} is being pushed into
+the helper application, comprising the function \texttt{hit} applied
+(Miller-pattern-style) to distinct local variables (here just \texttt{xs})
+whose types are known, thus determining \texttt{hit}'s declaration!
+Where the helper function has need of implicit syntax, a little more
+care will be needed to ensure control over what gets abstracted.
+
+Propositional requirements arising in the course of programming might
+also be discharged as side-problems in \texttt{where}-clauses. We do not
+write proofs in terms: we rather expect proofs to be picked up from the
+context, so side-problems allow us to seed the context appropriately.
+Moreover, elaboration models type constraints as propositional equations,
+so you can fix type-glitches (where provably equal types mismatch
+definitionally) by adding the proofs to a \texttt{where}-clause.
 \end{document}

### 2010-09-05 src/Documentation/Introduction.tex
a start at adding some opeing waffle to the Epitome Intro

@@
+\emph{''I believe that the time is ripe for significantly better
+  documentation of programs, and that we can best achieve this by
+  considering programs to be works of literature.''} Donald Knuth,
+1992.
+
+This document is intended to realise this lofty aim for the
+implementation of Epigram. Although, given that it is being written by
+several different people at once, the plot changes constantly, and
+it's currently pushing 300 pages even though it has barely been started,
+I don't expect it'll be a best seller.
+
+Epigram is a many faceted beast whose facets are constructed roughly
+one on top of the other. This document starts by describing its
+bowels: the core type theory which serves as Epigram's language of
+evidence and into which programs are elaborated. We then proceed
+outwards and upwards until we reach its snorting, wild eyed, and rabid
+face, a.k.a. the high-level language.
+
 \section{In the beginning}
+In the beginning there was Epigram 1. The Epigram 1 language was
+designed by Conor McBride and James McKinna in their weighty paper
+``The view from the left''\cite{conor.james:viewfromleft}. It descibes
+a programming language with support for inductive families and
+dependent pattern matching. Notably pattern matching is an added extra
+to the type theory.  Instead, pattern matching programs in the high
+level language are \emph{elaborated} into type theoretic expressions
+made of traditional eliminators in the underlying theory. An
+implementation of Epigram 1 was written by Conor and with its
+idiosyncratic emacs interface with 2d syntax, one could make small
+experiments. Unfortunately this prototype implementation proved
+unscalable and unmaintainable. Since then Agda 2 and Coq's Russell
+language have assimilated many of Epigram 1's features and many of
+Epigram's programs can be written in these two other languages.
 
-We really should write something informative here.
+Epigram 2 is not just an attempt at a better implementation of Epigram
+1; it has a radically different underlying type theory which supports
+interesting features in the high-level language. Firstly, it supports
+extensional equality of functions in an intentional theory by using
+observational equality. Secondly, it is a closed type theory where new
+data definitions are new codes in a universe of datatype descriptions
+whose validity is internally guaranteed rather than decreed to be
+acceptable by an external checker.
 

### 2010-10-26 papers/ydtm-rev/ydtm-rev.tex
ydtm-rev: fundam.bst (ugly but compulsory?)
james-auctex
ydtm: fix typos

@@
 which the code must \emph{respect} dynamically.  What if we wanted to
-guarantee the the output of our sorting algorithm is in order? The
+guarantee that the output of our sorting algorithm is in order? The
 order is not a prior static notion---that is why we need a sorting
@@
 types, which specify the values of some of their fields, can be used
-to express sharing between records, and between the the inputs and
+to express sharing between records, and between the inputs and
 outputs of record-transforming operations~\cite{pollack:records}.
@@
 
-\bibliographystyle{alpha}
+\bibliographystyle{fundam}
 \bibliography{handy,obsdec,alti,local}
@@
 \end{document}
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2011-02-16 notes/Epigram2/Epigram2.tex
NatFinVec in Kit

@@
 \usepackage{palatino}
-\usepackage{a4}
+\usepackage{a4wide}
 \usepackage{alltt}
+\usepackage{bbold}
+
+%%%%%%%%% INFERENCE RULES
+
+\newlength{\rulevgap}
+\setlength{\rulevgap}{0.05in}
+\newlength{\ruleheight}
+\newlength{\ruledepth}
+\newsavebox{\rulebox}
+\newlength{\GapLength}
+\newcommand{\gap}[1]{\settowidth{\GapLength}{#1} \hspace*{\GapLength}}
+\newcommand{\dotstep}[2]{\begin{tabular}[b]{@{}c@{}}
+                         #1\\$\vdots$\\#2
+                         \end{tabular}}
+\newlength{\fracwid}
+\newcommand{\dotfrac}[2]{\settowidth{\fracwid}{$\frac{#1}{#2}$}
+                         \addtolength{\fracwid}{0.1in}
+                         \begin{tabular}[b]{@{}c@{}}
+                         $#1$\\
+                         \parbox[c][0.02in][t]{\fracwid}{\dotfill} \\
+                         $#2$\\
+                         \end{tabular}}
+\newcommand{\Rule}[2]{\savebox{\rulebox}[\width][b]                         %
+                              {\( \frac{\raisebox{0in} {\( #1 \)}}       %
+                                       {\raisebox{-0.03in}{\( #2 \)}} \)}   %
+                      \settoheight{\ruleheight}{\usebox{\rulebox}}          %
+                      \addtolength{\ruleheight}{\rulevgap}                  %
+                      \settodepth{\ruledepth}{\usebox{\rulebox}}            %
+                      \addtolength{\ruledepth}{\rulevgap}                   %
+                      \raisebox{0in}[\ruleheight][\ruledepth]               %
+                               {\usebox{\rulebox}}}
+\newcommand{\Case}[2]{\savebox{\rulebox}[\width][b]                         %
+                              {\( \dotfrac{\raisebox{0in} {\( #1 \)}}       %
+                                       {\raisebox{-0.03in}{\( #2 \)}} \)}   %
+                      \settoheight{\ruleheight}{\usebox{\rulebox}}          %
+                      \addtolength{\ruleheight}{\rulevgap}                  %
+                      \settodepth{\ruledepth}{\usebox{\rulebox}}            %
+                      \addtolength{\ruledepth}{\rulevgap}                   %
+                      \raisebox{0in}[\ruleheight][\ruledepth]               %
+                               {\usebox{\rulebox}}}
+\newcommand{\Axiom}[1]{\savebox{\rulebox}[\width][b]                        %
+                               {$\frac{}{\raisebox{-0.03in}{$#1$}}$}        %
+                      \settoheight{\ruleheight}{\usebox{\rulebox}}          %
+                      \addtolength{\ruleheight}{\rulevgap}                  %
+                      \settodepth{\ruledepth}{\usebox{\rulebox}}            %
+                      \addtolength{\ruledepth}{\rulevgap}                   %
+                      \raisebox{0in}[\ruleheight][\ruledepth]               %
+                               {\usebox{\rulebox}}}
+\newcommand{\RuleSide}[3]{\gap{\mbox{$#2$}} \hspace*{0.1in}               %
+                            \Rule{#1}{#3}                          %
+                          \hspace*{0.1in}\mbox{$#2$}}
+\newcommand{\AxiomSide}[2]{\gap{\mbox{$#1$}} \hspace*{0.1in}              %
+                             \Axiom{#2}                            %
+                           \hspace*{0.1in}\mbox{$#1$}}
+\newcommand{\RULE}[1]{\textbf{#1}}
+\newcommand{\hg}{\quad}
 
@@
 definitionally) by adding the proofs to a \texttt{where}-clause.
+
+\section{Kernel Mustered}
+
+\newcommand{\SC}{\textsc}
+\newcommand{\Set}[1]{\mathrm{Set}^{#1}}
+\newcommand{\PI}[2]{(#1\!:\!#2)\to}
+\newcommand{\SG}[2]{(#1\!:\!#2)\times}
+\newcommand{\ONE}{\mathrm{1}}
+\newcommand{\ifr}{\underline}
+\newcommand{\lam}[1]{\lambda#1\mapsto}
+\newcommand{\pr}[2]{[#1 | #2]}
+\newcommand{\vd}{[]}
+\newcommand{\cum}{\preceq}
+\newcommand{\car}[1]{#1\:!}
+\newcommand{\cdr}[1]{#1\:-}
+\newcommand{\nm}{\Downarrow}
+\newcommand{\SJ}[4]{#1\:\{#2 \mapsto #3\}\nm#4}
+\newcommand{\letin}[3]{\mathrm{let}\: #1 = #2 \:\mathrm{in}\: #3}
+
+It's a moving target, but not a bad plan to write it down once in a
+while.
+
+\[\begin{array}{rcll}
+\SC{ChkTm} & ::= &  \Set{n} & n\in\mathbb{N} \\
+  &|& \PI{v}{\SC{ChkTm}}\SC{ChkTm} & \mbox{dependent function type} \\
+  &|& \SG{v}{\SC{ChkTm}}\SC{ChkTm} & \mbox{dependent pair type} \\
+  &|& \ONE \\
+  &|& \lam{v}\SC{ChkTm} \\
+  &|& \pr{\SC{ChkTm}}{\SC{ChkTm}} & \mbox{pair}\\
+  &|& \vd \\
+  &|& \letin{v}{\SC{InfTm}}{\SC{ChkTm}} \\
+  &|& \ifr{\SC{InfTm}} & \mbox{explicit direction change}
+\medskip \\
+\SC{InfTm} & ::= & v & \mbox{bound variable, untyped}\\
+  &|& x_{:\SC{Nm}} & \mbox{free variable, typed}\\
+  &|& f_{\mapsto\SC{Nm}:\SC{Nm}} & \mbox{definition, with value
+    and type}\\
+  &|& \SC{InfTm}\:\SC{ChkTm} & \mbox{application} \\
+  &|& \car{\SC{InfTm}} & \mbox{car}\\
+  &|& \cdr{\SC{InfTm}} & \mbox{cdr}\\
+  &|& (:\SC{ChkTm}) & \mbox{identity cast at given type}
+\end{array}\]
+
+If you eliminate \(v\)'s with no binder,
+definitions, defined variables, and identity casts, \(\SC{ChkTm}\)
+becomes \(\SC{Nm}\) and \(\SC{InfTm}\) becomes \(\SC{Ne}\), the
+\emph{normal} and \emph{neutral} terms, respectively. Let us define
+untyped computation relations of the following forms:
+\[\begin{array}{c}
+\SJ{\SC{Nm}}{v}{\SC{Nm}}{\SC{Nm}}\\
+\SJ{\SC{Ne}}{v}{\SC{Nm}}{\SC{Nm}}\\
+\SC{Nm}\:\SC{Nm} \mapsto \SC{Nm}\\
+\car{\SC{Nm}} \mapsto \SC{Nm}\\
+\cdr{\SC{Nm}} \mapsto \SC{Nm}\\
+\end{array}\]
+
+Note that substituting a \SC{Ne} for a \(v\) preserves its \SC{Ne}-ness
+and requires no worry about variable capture. Similarly, we may substitute
+a \SC{ChkTm} for a \(v\) and get a \SC{ChkTm}.
+
+The typing judgments are then as follows:
+\[\begin{array}{lr}
+(\{\Set{\omega}\}\cup\SC{Nm})\ni\SC{ChkTm}\nm\SC{Nm} & \mbox{type checking}\\
+\SC{InfTm}\nm{\SC{Nm}}\in\SC{Nm} & \mbox{type inference}\\
+\SC{Nm}\cum\SC{Nm} & \mbox{cumulativity}\\
+\SC{Nm}\ni\SC{Nm}\equiv\SC{Nm} &
+\mbox{equality} \\
+\SC{Ne}\equiv\SC{Ne}\in\SC{Nm} & \mbox{equality} \\
+\end{array}\]
+
+Type checking. Where \(x\) or \(f\) appears in premises but not
+conclusion, it is chosen fresh.
+\[\begin{array}{c}
+\Axiom{\Set{j}\ni\Set{i}\nm\Set{i}}\;j>i \quad
+\Axiom{\Set{i}\ni\ONE\nm\ONE} \quad
+\Rule{s\nm s'\in S\hg S\cum T}
+  {T\ni\ifr{s}\nm s'}
+\\
+
+\Rule{\Set{i}\ni S\nm S' \hg \Set{i}\ni T[x_{:S}/v]\nm T'[x_{:S}/v]}
+  {\Set{i}\ni\PI{v}{S}T\nm \PI{v}{S'}{T'}}
+\quad
+\Rule{\Set{i}\ni S\nm S' \hg \Set{i}\ni T[x_{:S}/v]\nm T'[x_{:S}/v]}
+  {\Set{i}\ni\SG{v}{S}T\nm \SG{v}{S'}{T'}}
+\\
+
+\Rule{T[x_{:S}/u]\ni t[x_{:S}/v]\nm t'[x_{:S}/v]}
+  {\PI{u}{S}{T}\ni\lam{v}t\nm\lam{v}t'}
+\quad
+\Rule{S\ni s\nm s'\hg \SJ{T}{v}{s'}{T'}\hg T' \ni t\nm t'}
+  {\SG{v}{S}T \ni \pr{s}{t} \nm \pr{s'}{t'}}
+\quad
+\Axiom{\ONE\ni\vd\nm\vd}\\
+\Rule{s\nm s'\in S\hg T\ni t[f_{\mapsto s':S}/v]\nm t'}
+  {T\ni\letin{v}{s}{t}\nm t'}
+
+\end{array}\]
+
+Type inference. Note no rule for bound vars.
+\[\begin{array}{c}
+\Axiom{x_{:S}\nm\ifr{x_{:S}}\in S} \quad \Axiom{f_{\mapsto s : S}\nm s \in S}
+\quad
+\Rule{\Set{\omega}\ni T}
+  {(:T)\nm\lam{v}v\in \PI{v}{T}T}
+ \\
+\Rule{f\nm f'\in\PI{v}{S}T \hg S\ni s\nm s'
+   \hg \SJ{T}{v}{s'}{T'} \hg f'\:s'\mapsto t'}
+  {f\:s\nm t' \in T'}
+\\
+\Rule{p\nm p'\in\SG{v}{S}T\hg \car{p'}\mapsto s'}
+  {s' \in S}
+\quad
+\Rule{p\nm p'\in\SG{v}{S}T\hg \car{p'}\mapsto s' \hg \SJ{T}{v}{s'}{T'}
+   \hg \cdr{p'}\mapsto t'}
+  {\cdr{p}\nm t' \in T'}
+\end{array}\]
+
+Cumulativity.
+\[\begin{array}{c}
+\Axiom{\Set{i}\cum\Set{j}}\;i\le j
+\quad
+\Rule{S'\cum S \quad T[x_{S'}/u]\cum T'[x_{S'}/v]}
+  {\PI{u}{S}T \cum \PI{v}{S'}T'}
+\quad
+\Rule{S\cum S' \quad T[x_{S}/u]\cum T'[x_{S}/v]}
+  {\SG{u}{S}T \cum \SG{v}{S'}T'}
+\quad
+\Axiom{\ONE\cum\ONE}
+\\
+\Rule{\ifr{S} \equiv \ifr{T}\in\Set{n}}
+  {S \cum T}
+\end{array}\]
+
+Equality for checkable terms.
+\[\begin{array}{c}
+\Axiom{\Set{j}\ni\Set{i}\equiv\Set{i}}\;j>i
+\\
+\Rule{\Set{i}\ni S'\equiv S \hg \Set{i}\ni T[x_{S'}/u] \equiv T'[x_{s'}/v]}
+  {\Set{i}\ni \PI{u}{S}T\equiv\PI{v}{S'}{T'}}
+\\
+\Rule{\Set{i}\ni S\equiv S' \hg \Set{i}\ni T[x_{S}/u] \equiv T'[x_{s'}/v]}
+  {\Set{i}\ni \SG{u}{S}T\equiv\SG{v}{S'}{T'}}
+\\
+\Rule{\PI{v}{S}T \ni f \hg \PI{v}{S}T \ni f' \hg
+ f\:\ifr{x_{:S}}\mapsto t\hg f'\:\ifr{x_{:S}}\mapsto t' \hg T[x_{:S}/v]\ni t \equiv t'}
+  {\PI{v}{S}T \ni f \equiv f'}
+\\
+\Rule{\begin{array}{c}
+\SG{v}{S}T \ni p \hg \SG{v}{S}T \ni p' \\
+\car{p}\mapsto s \hg \cdr{p}\mapsto t \hg 
+\car{p'}\mapsto s' \hg \cdr{p'}\mapsto t' \\
+S\ni s\equiv s' \hg \SJ{T}{v}{s}{T'} \hg T'\ni t\equiv t'
+\end{array}}
+  {\SG{v}{S}T \ni p \equiv p'}
+\quad
+\Rule{\ONE\ni t \hg \ONE \ni t'}
+  {\ONE\ni t \equiv t'}
+\\
+\Rule{s\equiv t \in S \hg S\cum T}
+  {T\ni\ifr{s}\equiv\ifr{t}}
+\end{array}\]
+
+Equality for inferrable terms.
+\[\begin{array}{c}
+\Axiom{x_{:S}\equiv x_{:S}\in S}
+\quad
+\Rule{f\equiv f'\in\PI{v}{S}T \hg S\ni s\equiv s' \hg \SJ{T}{v}{s}{T'}}
+  {f\:s\equiv f'\:s'\in T'}
+\\
+\Rule{p\equiv p'\in\SG{v}{S}T}
+  {\car{p}\equiv \car{p'}\in S}
+\quad
+\Rule{p\equiv p'\in\SG{v}{S}T}
+  {\cdr{p}\equiv \cdr{p'}\in T[\car{p}/v]}
+\end{array}\]
+
+Computation.
+\[\begin{array}{c}
+\Rule{\SJ{t}{v}{s}{t'}}
+  {(\lam{v}t)\:s\mapsto t'} \quad
+\Axiom{\ifr{f}\:s \mapsto \ifr{f\:s}}
+\quad
+\Axiom{\car{\pr{s}{t}}\mapsto s}\quad
+\Axiom{\car{\ifr{p}} \mapsto \ifr{\car{p}}}\quad
+\Axiom{\cdr{\pr{s}{t}}\mapsto t}\quad
+\Axiom{\cdr{\ifr{p}} \mapsto \ifr{\cdr{p}}}
+\end{array}\]
+
+Normal substitution.
+\[\begin{array}{c}
+\Axiom{\SJ{\Set{i}}{v}{r}{\Set{i}}}
+\\
+\Rule{\SJ{S}{v}{r}{S'}\quad\SJ{T[x_{S'}/u]}{v}{r}{T'[x_{S'}/u]}}
+  {\SJ{(\PI{u}{S}T)}{v}{r}{\PI{u}{S'}T'}}
+\\
+\Rule{\SJ{S}{v}{r}{S'}\quad\SJ{T[x_{S'}/u]}{v}{r}{T'[x_{S'}/u]}}
+  {\SJ{(\SG{u}{S}T)}{v}{r}{\SG{u}{S'}T'}}
+\\
+\Rule{\SJ{t[x_{S'}/u]}{v}{r}{t'[x_{S'}/u]}}
+  {\SJ{(\lam{u}{t})}{v}{r}{\lam{u}t'}}
+\quad
+\Rule{\SJ{s}{v}{r}{s'}\hg\SJ{t}{v}{r}{t'}}
+  {\SJ{\pr{s}{t}}{v}{r}{\pr{s'}{t'}}}
+\quad
+\Axiom{\SJ{\vd}{v}{r}{\vd}}
+\\
+\Rule{\SJ{t}{v}{r}{t'}}
+  {\SJ{\ifr{t}}{v}{r}{t'}}
+\end{array}\]
+
+Neutral substitution.
+\[\begin{array}{c}
+\Axiom{\SJ{v}{v}{r}{r}} \quad
+\Axiom{\SJ{x_{:S}}{v}{r}{\ifr{x_{:S}}}} \\
+\Rule{\SJ{f}{v}{r}{f'}\hg\SJ{s}{v}{r}{s'}\hg f'\:s'\mapsto t'}
+  {\SJ{(f\:s)}{v}{r}{t'}} \\
+\Rule{\SJ{p}{v}{r}{p'}\hg\car{p'}\mapsto s'}
+  {\SJ{(\car{p})}{v}{r}{s'}} \quad
+\Rule{\SJ{p}{v}{r}{p'}\hg\cdr{p'}\mapsto t'}
+  {\SJ{(\cdr{p})}{v}{r}{t'}}
+\end{array}\]
+
 \end{document}
